---
title: "ROC Curve Analysis"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",  echo = TRUE,
  eval = FALSE)
```

This vignette shows how to perform Receiver Operating Characteristic
(ROC) analysis with **meddecide** using the `psychopdaroc()` function.

## Loading the Data

```{r, eval=FALSE}
df_roc <- read.csv(system.file("extdata", "roc_example.csv", package = "meddecide"))
head(df_roc)
```

## Creating the ROC Curve

```{r, eval=FALSE}
roc_res <- psychopdaroc(data = df_roc, class = df_roc$class, value = df_roc$value)
roc_res$plot
```

The resulting plot shows the ROC curve along with the area under the curve (AUC).
You can extract the AUC value and other statistics from the result object.

```{r, eval=FALSE}
roc_res$AUC
```

## Bootstrapping and Confidence Intervals

The `pROC` package provides robust methods for computing confidence intervals for diagnostic metrics.

```{r, eval=FALSE}
# Calculate AUC confidence intervals using pROC
library(pROC)
roc_obj <- roc(response = outcome, predictor = biomarker)
pROC::ci.auc(roc_obj, method = "bootstrap")

# Calculate threshold-specific confidence intervals
pROC::ci.thresholds(roc_obj)
```

These methods provide statistically sound confidence intervals for your ROC analysis.
