---
title: "Clinical Research Examples with jSurvival"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This vignette provides practical examples of using jSurvival for common clinical research scenarios. Each example demonstrates a complete workflow from data preparation to interpretation of results.

# Example 1: Cancer Treatment Comparison Study

## Background

A clinical trial comparing two cancer treatments (Treatment A vs. Treatment B) with overall survival as the primary endpoint.

## Research Question

Is there a significant difference in overall survival between Treatment A and Treatment B?

## Data Structure

```{r cancer-data}
# Example data structure
cancer_data <- data.frame(
  PatientID = 1:200,
  Treatment = rep(c("Treatment A", "Treatment B"), each = 100),
  OS_months = c(rnorm(100, 24, 8), rnorm(100, 30, 10)),
  Death = rbinom(200, 1, 0.6),
  Age = rnorm(200, 65, 10),
  Stage = sample(c("I", "II", "III", "IV"), 200, replace = TRUE),
  ECOG = sample(0:2, 200, replace = TRUE)
)
```

## Analysis Steps

### Step 1: Single Arm Analysis (Overall Population)

```{r cancer-singlearm}
# First, examine overall survival characteristics
singlearm_result <- singlearm(
  data = cancer_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  cutp = "12, 24, 36",
  timetypeoutput = "months"
)
```

**Interpretation**: 
- Median overall survival for the entire cohort
- 1, 2, and 3-year survival rates
- 95% confidence intervals for all estimates

### Step 2: Treatment Comparison

```{r cancer-treatment}
# Compare survival between treatments
survival_result <- survival(
  data = cancer_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Treatment,
  analysistype = "overall",
  ph_cox = TRUE,
  ci95 = TRUE,
  risktable = TRUE
)
```

**Key Outputs**:
- Log-rank test p-value for treatment difference
- Hazard ratio comparing Treatment B vs. Treatment A
- Kaplan-Meier curves with risk tables
- Median survival times by treatment group

### Step 3: Multivariable Analysis

```{r cancer-multivariable}
# Adjust for potential confounders
multi_result <- multisurvival(
  data = cancer_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = c(Treatment, Age, Stage, ECOG)
)
```

**Interpretation**:
- Adjusted hazard ratio for treatment effect
- Independent prognostic factors
- Model fit statistics (concordance index)

## Clinical Interpretation

**If HR = 0.75 (95% CI: 0.60-0.95), p = 0.02**:
- Treatment B reduces the risk of death by 25% compared to Treatment A
- This result is statistically significant
- The confidence interval excludes 1.0, supporting the treatment benefit

# Example 2: Biomarker Cut-point Analysis

## Background

Investigating whether a continuous biomarker can predict patient outcomes and finding the optimal threshold for clinical decision-making.

## Research Question

What is the optimal cut-point for Biomarker X to predict overall survival?

## Analysis Workflow

### Step 1: Continuous Variable Analysis

```{r biomarker-continuous}
# Find optimal cut-point for biomarker
biomarker_result <- survivalcont(
  data = biomarker_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  contexpl = Biomarker_X,
  findcut = TRUE,
  cutp = "12, 36, 60"
)
```

**Outputs**:
- Optimal cut-point value
- Hazard ratio for high vs. low biomarker groups
- ROC analysis for cut-point validation
- Sensitivity and specificity at optimal cut-point

### Step 2: Validation Analysis

```{r biomarker-validation}
# Validate cut-point using dichotomized variable
# (Assuming cut-point = 15.2 was identified)
biomarker_data$Biomarker_High <- ifelse(biomarker_data$Biomarker_X >= 15.2, "High", "Low")

validation_result <- survival(
  data = biomarker_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Biomarker_High,
  ph_cox = TRUE
)
```

## Clinical Translation

**If optimal cut-point = 15.2 ng/mL, HR = 2.1 (95% CI: 1.4-3.2)**:
- Patients with biomarker ≥15.2 ng/mL have 2.1 times higher risk of death
- This threshold could be used for treatment stratification
- Consider validation in independent cohort

# Example 3: Time-to-Event Analysis with Competing Risks

## Background

Studying cancer-specific mortality while accounting for deaths from other causes.

## Research Question

What factors predict cancer-specific death, considering other-cause mortality as a competing risk?

## Analysis Approach

```{r competing-risks}
# For competing risks, use cause-specific outcomes
# Death_cancer: 1 = cancer death, 0 = alive or other death
# Death_other: 1 = other death, 0 = alive or cancer death

# Cancer-specific survival analysis
cancer_specific <- survival(
  data = competing_data,
  elapsedtime = OS_months,
  outcome = Death_cancer,
  outcomeLevel = "1",
  explanatory = Treatment,
  analysistype = "overall",
  ph_cox = TRUE
)

# Other-cause mortality analysis
other_cause <- survival(
  data = competing_data,
  elapsedtime = OS_months,
  outcome = Death_other,
  outcomeLevel = "1",
  explanatory = Treatment,
  analysistype = "overall",
  ph_cox = TRUE
)
```

**Interpretation**:
- Separate hazard ratios for cancer-specific and other-cause mortality
- Treatment effects may differ between competing outcomes
- Consider cumulative incidence functions for formal competing risk analysis

# Example 4: Landmark Analysis

## Background

Addressing guarantee-time bias when analyzing post-treatment biomarkers or response markers.

## Research Question

Does tumor response at 3 months predict long-term survival?

## Analysis Strategy

```{r landmark-analysis}
# Create landmark dataset (patients alive at 3 months)
landmark_data <- subset(original_data, OS_months > 3 | (OS_months <= 3 & Death == 0))

# Adjust survival times for landmark point
landmark_data$OS_months_landmark <- landmark_data$OS_months - 3
landmark_data$OS_months_landmark[landmark_data$OS_months_landmark < 0] <- 0

# Analyze survival from landmark time
landmark_result <- survival(
  data = landmark_data,
  elapsedtime = OS_months_landmark,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Response_3month,
  uselandmark = TRUE,
  landmark = 3
)
```

**Key Points**:
- Only includes patients who survived to landmark time
- Eliminates guarantee-time bias
- Useful for post-treatment predictors

# Example 5: Dose-Response Analysis

## Background

Evaluating survival outcomes across different dose levels of a treatment.

## Research Question

Is there a dose-response relationship for treatment efficacy?

## Analysis Steps

### Step 1: Categorical Dose Analysis

```{r dose-categorical}
# Treat dose as categorical variable
dose_categorical <- survival(
  data = dose_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Dose_category,  # Low, Medium, High
  analysistype = "pairwise",
  padjustmethod = "holm"
)
```

### Step 2: Trend Analysis

```{r dose-trend}
# Test for linear trend across dose levels
# Create ordinal dose variable (1, 2, 3)
dose_data$Dose_ordinal <- as.numeric(factor(dose_data$Dose_category, 
                                           levels = c("Low", "Medium", "High")))

trend_result <- survival(
  data = dose_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Dose_ordinal,
  ph_cox = TRUE
)
```

**Interpretation**:
- Test for overall dose effect (3-group comparison)
- Test for linear trend (ordinal dose variable)
- HR per dose level increase

# Example 6: Propensity Score Matching

## Background

Comparing treatments in observational data while controlling for selection bias.

## Research Question

What is the treatment effect after accounting for baseline differences between groups?

## Analysis Workflow

```{r propensity-matching}
# Step 1: Create matched dataset (outside jSurvival)
# matched_data <- created using propensity score matching

# Step 2: Analyze matched cohort
matched_result <- survival(
  data = matched_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Treatment,
  analysistype = "overall",
  ph_cox = TRUE
)

# Step 3: Sensitivity analysis with additional adjustment
sensitivity_result <- multisurvival(
  data = matched_data,
  elapsedtime = OS_months,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = c(Treatment, remaining_confounders)
)
```

# Reporting Guidelines

## Essential Elements

1. **Study Design**: Clearly describe cohort and follow-up
2. **Sample Size**: Total patients, events, median follow-up
3. **Statistical Methods**: Analysis approach and software used
4. **Results Presentation**:
   - Median survival with 95% CI
   - Hazard ratios with 95% CI
   - P-values and statistical tests used
   - Kaplan-Meier curves with risk tables

## Example Results Section

> "Among 200 patients (Treatment A: n=100, Treatment B: n=100), there were 120 deaths during a median follow-up of 18.5 months (IQR: 12.2-24.8). Median overall survival was 22.1 months (95% CI: 18.5-25.8) for Treatment A versus 28.4 months (95% CI: 24.1-32.7) for Treatment B. Treatment B was associated with a 25% reduction in the risk of death compared to Treatment A (HR=0.75, 95% CI: 0.60-0.95, p=0.02). The 2-year survival rates were 45% (95% CI: 35-55%) and 60% (95% CI: 50-70%) for Treatment A and B, respectively."

## Figure Legends

> "Figure 1. Kaplan-Meier survival curves comparing Treatment A versus Treatment B. The table below shows the number of patients at risk at each time point. P-value from log-rank test. HR: hazard ratio, CI: confidence interval."

# Common Pitfalls and Solutions

## Pitfall 1: Immortal Time Bias

**Problem**: Including post-baseline events as baseline predictors
**Solution**: Use landmark analysis or time-dependent covariates

## Pitfall 2: Multiple Testing

**Problem**: Testing multiple endpoints without adjustment
**Solution**: Use appropriate p-value correction methods

## Pitfall 3: Proportional Hazards Violation

**Problem**: HR changes over time
**Solution**: Use stratified Cox models or time-dependent effects

## Pitfall 4: Insufficient Follow-up

**Problem**: High censoring rates affect reliability
**Solution**: Report median follow-up and censoring patterns

# Conclusion

These examples demonstrate the versatility of jSurvival for clinical research applications. The key to successful survival analysis is:

1. **Clear research questions** with appropriate study design
2. **Careful data preparation** and quality control
3. **Appropriate statistical methods** for the research context
4. **Comprehensive reporting** following established guidelines
5. **Clinical interpretation** that considers both statistical and practical significance

For additional support and examples, visit the [jSurvival website](https://www.serdarbalci.com/jsurvival/) and [ClinicoPath documentation](https://sbalci.github.io/ClinicoPathJamoviModule/).
