---
title: "Grouped Hazard Forest Plots for Subgroup Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction to Grouped Forest Plots

The Grouped Hazard Forest Plot module provides a powerful visualization method for comparing treatment effects across different patient subgroups. This analysis addresses a critical need in precision medicine and clinical research by performing separate Cox proportional hazards regression analyses for each subgroup and presenting the results in a unified forest plot.

This module was developed to address **GitHub Issue #88**: Create grouped forest plots showing treatment vs control for each variant.

## Key Features

- **Subgroup-Specific Analysis**: Separate Cox regression for each group
- **Visual Comparison**: Side-by-side hazard ratios in forest plot format
- **Statistical Testing**: Individual p-values and interaction tests
- **Clinical Applications**: Perfect for biomarker studies, precision medicine, and clinical trials
- **Flexible Grouping**: Support for any categorical grouping variable

## Clinical Applications

### Precision Medicine
- Treatment efficacy by genetic variants
- Biomarker-stratified therapy selection
- Molecular subtype analysis

### Clinical Trials
- Subgroup efficacy analysis
- Patient population identification
- Treatment interaction detection

### Biomarker Studies
- Predictive biomarker evaluation
- Treatment-biomarker interactions
- Patient stratification strategies

---

# Getting Started

## Required Data Structure

Your dataset should contain:

1. **Time Variable**: Continuous variable for follow-up duration
2. **Event Variable**: Binary indicator (0=censored, 1=event)
3. **Treatment Variable**: Factor variable for treatment comparison
4. **Grouping Variable**: Categorical variable defining subgroups

## Example Dataset

Let's examine the structure of our test data:


```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
    echo = TRUE,
  eval = FALSE
)
```


```{r setup}
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Load example data
data("groupedforest_comprehensive_data")

# Examine data structure
str(groupedforest_comprehensive_data)
```

```{r data-preview}
# Preview the data
head(groupedforest_comprehensive_data) %>%
  knitr::kable(caption = "Sample of Comprehensive Grouped Forest Data")
```

---

# Basic Grouped Forest Analysis

## Simple Two-Group Comparison

Let's start with a basic analysis comparing treatment effects between biomarker-positive and biomarker-negative patients:

```{r basic-analysis}
# Basic grouped forest plot by biomarker status
result_biomarker <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "biomarker_status",
  show_statistics = TRUE,
  show_overall = TRUE
)

# The result object contains the analysis
class(result_biomarker)
```

## Tumor Stage Stratification

Compare treatment effects between early and advanced tumor stages:

```{r stage-analysis}
# Grouped analysis by tumor stage
result_stage <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "tumor_stage",
  confidence_level = 0.95,
  sort_by_hr = TRUE,
  plot_title = "Treatment Effects by Tumor Stage"
)
```

## Gender-Based Subgroup Analysis

Evaluate treatment effects across gender groups:

```{r gender-analysis}
# Analysis by gender
result_gender <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "gender",
  show_counts = TRUE,
  plot_theme = "clinical"
)
```

---

# Advanced Analysis Features

## Covariate Adjustment

Include covariates to adjust for confounding factors:

```{r covariate-adjustment}
# Analysis with covariate adjustment
result_adjusted <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "biomarker_status",
  covariates = c("age", "performance_status"),
  reference_treatment = "Control",
  show_statistics = TRUE
)
```

## Interaction Testing

Test for treatment-by-subgroup interactions:

```{r interaction-testing}
# Test for interaction effects
result_interaction <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "tumor_stage",
  interaction_test = TRUE,
  show_overall = TRUE
)
```

## Custom Display Options

Customize the appearance and range of the forest plot:

```{r custom-display}
# Custom forest plot settings
result_custom <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "biomarker_status",
  plot_title = "Biomarker-Stratified Treatment Analysis",
  plot_theme = "publication",
  hr_range = "wide",
  sort_by_hr = TRUE,
  confidence_level = 0.95
)
```

---

# Multiple Subgroup Analysis

## Multi-Subgroup Dataset

Let's work with a dataset containing multiple molecular subtypes:

```{r multi-subgroup-data}
# Load multi-subgroup data
data("groupedforest_multi_subgroups")

# Examine the subgroups
table(groupedforest_multi_subgroups$molecular_subtype)
```

## Molecular Subtype Analysis

Analyze treatment effects across molecular subtypes:

```{r molecular-analysis}
# Molecular subtype forest plot
result_molecular <- groupedforest(
  data = groupedforest_multi_subgroups,
  time_var = "time_to_event",
  event_var = "event_occurred",
  treatment_var = "intervention",
  grouping_var = "molecular_subtype",
  covariates = "patient_age",
  show_statistics = TRUE,
  show_counts = TRUE
)
```

## Risk Category Stratification

Compare treatment effects across risk categories:

```{r risk-analysis}
# Risk category analysis
result_risk <- groupedforest(
  data = groupedforest_multi_subgroups,
  time_var = "time_to_event",
  event_var = "event_occurred",
  treatment_var = "intervention",
  grouping_var = "risk_category",
  interaction_test = TRUE,
  plot_theme = "minimal"
)
```

---

# Precision Medicine Applications

## Genomic Variant Analysis

Demonstrate precision medicine applications with genomic data:

```{r precision-medicine-data}
# Load precision medicine dataset
data("groupedforest_precision_medicine")

# Examine genomic variants
table(groupedforest_precision_medicine$genomic_variant)
```

```{r genomic-analysis}
# Genomic variant forest plot
result_genomic <- groupedforest(
  data = groupedforest_precision_medicine,
  time_var = "progression_free_months",
  event_var = "progression_event",
  treatment_var = "therapy_type",
  grouping_var = "genomic_variant",
  reference_treatment = "Chemotherapy",
  plot_title = "Precision Therapy by Genomic Variant"
)
```

## Expression Level Stratification

Analyze treatment effects by biomarker expression levels:

```{r expression-analysis}
# Expression level analysis
result_expression <- groupedforest(
  data = groupedforest_precision_medicine,
  time_var = "progression_free_months",
  event_var = "progression_event",
  treatment_var = "therapy_type",
  grouping_var = "expression_level",
  covariates = c("age_at_diagnosis", "tumor_size"),
  show_overall = TRUE
)
```

---

# Biomarker Stratification Studies

## Biomarker Level Analysis

Work with biomarker stratification data:

```{r biomarker-data}
# Load biomarker data
data("groupedforest_biomarker_data")

# Examine biomarker levels
table(groupedforest_biomarker_data$biomarker_level)
```

```{r biomarker-analysis}
# Biomarker level forest plot
result_biomarker_level <- groupedforest(
  data = groupedforest_biomarker_data,
  time_var = "overall_survival_months",
  event_var = "death_indicator",
  treatment_var = "treatment_arm",
  grouping_var = "biomarker_level",
  plot_title = "Biomarker-Targeted Therapy by Expression Level",
  hr_range = "custom",
  custom_hr_min = 0.2,
  custom_hr_max = 5.0
)
```

## Pathway Status Analysis

Analyze treatment effects by pathway activation status:

```{r pathway-analysis}
# Pathway status analysis
result_pathway <- groupedforest(
  data = groupedforest_biomarker_data,
  time_var = "overall_survival_months",
  event_var = "death_indicator",
  treatment_var = "treatment_arm",
  grouping_var = "pathway_status",
  plot_theme = "publication",
  show_counts = TRUE
)
```

---

# Clinical Trial Applications

## Multi-Factor Clinical Trial

Demonstrate complex clinical trial analysis:

```{r clinical-trial-data}
# Load clinical trial data
data("groupedforest_interaction_data")

# Examine trial characteristics
table(groupedforest_interaction_data$genetic_profile)
table(groupedforest_interaction_data$disease_severity)
```

```{r clinical-trial-analysis}
# Genetic profile analysis
result_clinical <- groupedforest(
  data = groupedforest_interaction_data,
  time_var = "event_free_survival",
  event_var = "event_status",
  treatment_var = "randomized_treatment",
  grouping_var = "genetic_profile",
  interaction_test = TRUE,
  export_data = TRUE,
  plot_title = "Clinical Trial: Treatment by Genetic Profile"
)
```

## Disease Severity Stratification

Analyze treatment effects across disease severity levels:

```{r severity-analysis}
# Disease severity analysis
result_severity <- groupedforest(
  data = groupedforest_interaction_data,
  time_var = "event_free_survival",
  event_var = "event_status",
  treatment_var = "randomized_treatment",
  grouping_var = "disease_severity",
  covariates = c("age_years", "baseline_severity_score"),
  confidence_level = 0.90
)
```

---

# Customization and Visualization Options

## Plot Themes

Demonstrate different plot themes:

```{r theme-comparison}
# Load simple dataset for theme demonstration
data("groupedforest_simple_data")

# Clinical theme (default)
result_theme_clinical <- groupedforest(
  data = groupedforest_simple_data,
  time_var = "time",
  event_var = "event",
  treatment_var = "treatment",
  grouping_var = "subgroup",
  plot_theme = "clinical",
  plot_title = "Clinical Theme"
)

# Publication theme
result_theme_pub <- groupedforest(
  data = groupedforest_simple_data,
  time_var = "time",
  event_var = "event",
  treatment_var = "treatment",
  grouping_var = "subgroup",
  plot_theme = "publication",
  plot_title = "Publication Theme"
)
```

## Hazard Ratio Ranges

Demonstrate different HR range options:

```{r hr-ranges}
# Wide range
result_wide <- groupedforest(
  data = groupedforest_simple_data,
  time_var = "time",
  event_var = "event",
  treatment_var = "treatment",
  grouping_var = "subgroup",
  hr_range = "wide",
  plot_title = "Wide HR Range (0.1-10)"
)

# Narrow range
result_narrow <- groupedforest(
  data = groupedforest_simple_data,
  time_var = "time",
  event_var = "event",
  treatment_var = "treatment",
  grouping_var = "subgroup",
  hr_range = "narrow",
  plot_title = "Narrow HR Range (0.5-2)"
)
```

---

# Statistical Considerations

## Sample Size and Power

Understanding the impact of subgroup sample sizes:

```{r sample-size-demo}
# Examine sample sizes by subgroup
groupedforest_comprehensive_data %>%
  group_by(biomarker_status, treatment) %>%
  summarise(
    n = n(),
    events = sum(death_event),
    event_rate = round(mean(death_event), 3),
    .groups = 'drop'
  ) %>%
  knitr::kable(caption = "Sample Sizes and Event Rates by Subgroup")
```

## Confidence Intervals

Demonstrate different confidence levels:

```{r confidence-levels}
# Different confidence levels
conf_levels <- c(0.80, 0.90, 0.95, 0.99)

for (conf_level in conf_levels) {
  result <- groupedforest(
    data = groupedforest_simple_data,
    time_var = "time",
    event_var = "event", 
    treatment_var = "treatment",
    grouping_var = "subgroup",
    confidence_level = conf_level,
    plot_title = paste0(conf_level * 100, "% Confidence Intervals")
  )
}
```

---

# Interpretation Guidelines

## Reading Forest Plots

### Key Elements:
- **Vertical line at HR = 1**: No treatment effect
- **Points to the left (HR < 1)**: Treatment reduces hazard (beneficial)
- **Points to the right (HR > 1)**: Treatment increases hazard (harmful)
- **Horizontal lines**: Confidence intervals
- **Wider intervals**: Less precise estimates (smaller sample sizes)

### Statistical Significance:
- **Confidence interval excludes 1**: Statistically significant effect
- **Confidence interval includes 1**: Non-significant effect
- **P-values**: Individual subgroup significance tests

## Clinical Interpretation

### Treatment Heterogeneity:
- **Consistent effects**: Similar HRs across subgroups
- **Treatment interactions**: Varying effects between subgroups
- **Precision medicine opportunities**: Subgroups with differential benefit

### Clinical Decision Making:
- **Patient selection**: Identify subgroups most likely to benefit
- **Treatment recommendation**: Consider subgroup-specific effects
- **Future research**: Hypothesis generation for subsequent studies

---

# Advanced Statistical Methods

## Interaction Testing

Understanding treatment-by-subgroup interactions:

```{r interaction-interpretation}
# Comprehensive interaction analysis
result_full_interaction <- groupedforest(
  data = groupedforest_comprehensive_data,
  time_var = "survival_months",
  event_var = "death_event",
  treatment_var = "treatment",
  grouping_var = "biomarker_status",
  interaction_test = TRUE,
  covariates = c("age", "performance_status"),
  show_overall = TRUE,
  show_statistics = TRUE
)
```

## Multiple Comparisons

Considerations for multiple subgroup testing:

```{r multiple-comparisons}
# Multiple subgroup analysis
result_multiple <- groupedforest(
  data = groupedforest_multi_subgroups,
  time_var = "time_to_event",
  event_var = "event_occurred",
  treatment_var = "intervention",
  grouping_var = "molecular_subtype",  # 4 subgroups = multiple comparisons
  show_statistics = TRUE,
  plot_title = "Multiple Subgroup Analysis (Consider Adjustment)"
)
```

---

# Clinical Research Applications

## Regulatory Submissions

Guidelines for regulatory use:

### Best Practices:
1. **Pre-specified analyses**: Define subgroups before data analysis
2. **Clinical rationale**: Justify subgroup selections
3. **Statistical plan**: Document analysis methodology
4. **Interaction testing**: Formally test for treatment interactions
5. **Multiple comparisons**: Consider adjustment methods

## Publication Guidelines

### Reporting Standards:
- **CONSORT guidelines**: Subgroup analysis reporting
- **Transparency**: Report all analyzed subgroups
- **Effect sizes**: Include confidence intervals
- **Clinical significance**: Interpret beyond statistical significance

---

# Troubleshooting and Common Issues

## Data Requirements

### Minimum Sample Sizes:
- **Per subgroup**: At least 10-15 events recommended
- **Treatment arms**: Balanced representation preferred
- **Event rates**: Sufficient events for Cox regression

### Data Quality:
```{r data-quality}
# Check data quality
check_data_quality <- function(data, time_var, event_var, treatment_var, grouping_var) {
  cat("Data Quality Checks:\n")
  cat("Total sample size:", nrow(data), "\n")
  cat("Complete cases:", sum(complete.cases(data[, c(time_var, event_var, treatment_var, grouping_var)])), "\n")
  cat("Event rate:", round(mean(data[[event_var]], na.rm = TRUE), 3), "\n")
  
  # Subgroup sizes
  cat("\nSubgroup sizes:\n")
  print(table(data[[grouping_var]], data[[treatment_var]]))
  
  # Event rates by subgroup
  cat("\nEvent rates by subgroup:\n")
  event_rates <- data %>%
    group_by(.data[[grouping_var]], .data[[treatment_var]]) %>%
    summarise(event_rate = round(mean(.data[[event_var]], na.rm = TRUE), 3), .groups = 'drop')
  print(event_rates)
}

check_data_quality(
  groupedforest_comprehensive_data,
  "survival_months",
  "death_event", 
  "treatment",
  "biomarker_status"
)
```

## Common Error Messages

### Solutions:
1. **"No valid Cox regression results"**: Check subgroup sample sizes
2. **"Insufficient events"**: Verify event rates per subgroup
3. **"Model convergence failed"**: Consider removing small subgroups
4. **"Missing data"**: Address missing values in key variables

---

# Summary

The Grouped Hazard Forest Plot module provides a comprehensive solution for subgroup analysis in survival studies. Key benefits include:

## Clinical Value:
- **Precision Medicine**: Identify patients most likely to benefit
- **Treatment Personalization**: Guide individual treatment decisions
- **Research Planning**: Inform future study designs

## Statistical Rigor:
- **Formal Testing**: Cox regression for each subgroup
- **Interaction Analysis**: Test for differential treatment effects
- **Visualization**: Clear presentation of complex results

## Practical Applications:
- **Biomarker Studies**: Evaluate predictive markers
- **Clinical Trials**: Subgroup efficacy analysis
- **Drug Development**: Support regulatory submissions

This module addresses the critical need for systematic subgroup analysis in clinical research, providing both statistical rigor and practical clinical insights for precision medicine applications.

---

# References and Further Reading

## Statistical Methods:
- Cox Proportional Hazards Models
- Forest Plot Methodology
- Subgroup Analysis Guidelines
- Treatment-Biomarker Interactions

## Clinical Applications:
- Precision Medicine Frameworks
- Biomarker Validation Studies
- Clinical Trial Design
- Regulatory Guidelines

## Software Implementation:
- survival R package
- ggplot2 visualization
- jamovi integration
- ClinicoPath module ecosystem
