---
title: "jSurvival 15: Competing Survival Analysis"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

Competing survival analysis is essential when patients can experience multiple types of events that prevent the occurrence of the primary event of interest. The **competingsurvival** function in jSurvival provides comprehensive tools for analyzing overall survival, cause-specific survival, and competing risks scenarios using advanced statistical methods.

**Learning Objectives:**

- Understand the differences between overall, cause-specific, and competing risks survival analysis
- Master the application of Fine-Gray subdistribution hazard models
- Interpret cumulative incidence functions in competing risks scenarios
- Apply competing survival methods to real-world clinical datasets
- Create publication-ready competing risks plots and tables
- Understand when to use each type of survival analysis

## Module Overview

The **competingsurvival** function provides three complementary analysis approaches:

### 1. **Overall Survival Analysis**
- **Purpose**: Analyze all-cause mortality regardless of specific cause
- **Event Definition**: Any death (disease-related or other causes) vs. alive
- **Use Cases**: Primary endpoint in many clinical trials, population health studies
- **Statistical Method**: Standard Kaplan-Meier and Cox regression

### 2. **Cause-Specific Survival Analysis**  
- **Purpose**: Focus exclusively on disease-specific deaths
- **Event Definition**: Disease death vs. censored (including other deaths)
- **Use Cases**: When other causes of death are considered nuisance events
- **Limitation**: May overestimate survival if competing risks are substantial

### 3. **Competing Risks Analysis**
- **Purpose**: Account for multiple competing events simultaneously
- **Event Definition**: Disease death vs. other death vs. alive outcomes
- **Statistical Method**: Fine-Gray subdistribution hazard model with cumulative incidence functions
- **Advantage**: Provides realistic probability estimates accounting for competition

```{r load-packages}
library(ClinicoPath)
library(survival)
library(dplyr)
```

# Getting Started

## Sample Data Overview

We'll use a comprehensive competing survival dataset designed specifically for this analysis type.

```{r sample-data}
# Load competing survival data
data(competing_survival_data)

# Dataset overview
cat("Dataset dimensions:", nrow(competing_survival_data), "×", ncol(competing_survival_data), "\n")

# Key survival variables
survival_vars <- c("Overall_Time", "Outcome")
cat("Survival variables:", paste(survival_vars, collapse = ", "), "\n")

# Explanatory variables for stratification
explanatory_vars <- c("Treatment", "Tumor_Stage", "EGFR_Status", "Grade")
cat("Explanatory variables:", paste(explanatory_vars, collapse = ", "), "\n")

# Outcome distribution
outcome_summary <- competing_survival_data %>%
  group_by(Outcome) %>%
  summarise(
    n = n(),
    percentage = round(n() / nrow(competing_survival_data) * 100, 1),
    median_time = round(median(Overall_Time), 1),
    .groups = 'drop'
  )

cat("\nOutcome Distribution:\n")
print(outcome_summary)

# Treatment group summary
treatment_outcome <- competing_survival_data %>%
  group_by(Treatment, Outcome) %>%
  summarise(n = n(), .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = Outcome, values_from = n, values_fill = 0)

cat("\nTreatment by Outcome:\n")
print(treatment_outcome)
```

# Core Competing Survival Analysis

## 1. Overall Survival Analysis

Analyze all-cause mortality treating any death as an event.

```{r overall-survival}
# Overall Survival Analysis Example
# In jamovi: Survival > Competing Survival > Analysis Type: Overall

cat("Overall Survival Analysis\n")
cat("========================\n\n")

if(requireNamespace("survival", quietly = TRUE) && 
   requireNamespace("finalfit", quietly = TRUE)) {
  
  # Create overall survival indicator
  os_data <- competing_survival_data %>%
    mutate(
      death_overall = ifelse(Outcome %in% c("Dead_of_Disease", "Dead_of_Other"), 1, 0)
    )
  
  # Kaplan-Meier analysis by treatment
  surv_obj <- Surv(os_data$Overall_Time, os_data$death_overall)
  km_overall <- survfit(surv_obj ~ Treatment, data = os_data)
  
  cat("Overall Survival Summary by Treatment:\n")
  os_summary <- summary(km_overall, times = c(6, 12, 24, 36))$table
  print(os_summary)
  
  # Log-rank test
  logrank_test <- survdiff(surv_obj ~ Treatment, data = os_data)
  p_value <- 1 - pchisq(logrank_test$chisq, 1)
  
  cat("\nLog-rank Test Results:\n")
  cat("Chi-square statistic:", round(logrank_test$chisq, 3), "\n")
  cat("P-value:", round(p_value, 4), "\n")
  
  if(p_value < 0.05) {
    cat("Conclusion: Significant difference in overall survival by treatment\n")
  } else {
    cat("Conclusion: No significant difference in overall survival by treatment\n")
  }
  
  # Cox regression for hazard ratio
  cox_model <- coxph(surv_obj ~ Treatment, data = os_data)
  hr <- exp(cox_model$coefficients)
  hr_ci <- exp(confint(cox_model))
  
  cat("\nHazard Ratio Analysis:\n")
  cat("Hazard Ratio:", round(hr, 2), "\n")
  cat("95% Confidence Interval:", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), "\n")
  
  treatment_levels <- levels(os_data$Treatment)
  cat("Interpretation:", treatment_levels[2], "vs", treatment_levels[1], "\n")
}
```

## 2. Cause-Specific Survival Analysis

Focus exclusively on disease-specific deaths, treating other deaths as censored.

```{r cause-specific-survival}
# Cause-Specific Survival Analysis Example
# In jamovi: jSurvival > Competing Survival > Analysis Type: Cause Specific

cat("\nCause-Specific Survival Analysis\n")
cat("=================================\n\n")

if(requireNamespace("survival", quietly = TRUE) && 
   requireNamespace("finalfit", quietly = TRUE)) {
  
  # Create cause-specific survival indicator
  css_data <- competing_survival_data %>%
    mutate(
      death_disease = ifelse(Outcome == "Dead_of_Disease", 1, 0)
    )
  
  # Kaplan-Meier analysis for disease-specific survival
  surv_css <- Surv(css_data$Overall_Time, css_data$death_disease)
  km_css <- survfit(surv_css ~ Treatment, data = css_data)
  
  cat("Cause-Specific Survival Summary by Treatment:\n")
  css_summary <- summary(km_css, times = c(6, 12, 24, 36))$table
  print(css_summary)
  
  # Log-rank test for cause-specific survival
  logrank_css <- survdiff(surv_css ~ Treatment, data = css_data)
  p_css <- 1 - pchisq(logrank_css$chisq, 1)
  
  cat("\nCause-Specific Log-rank Test:\n")
  cat("Chi-square statistic:", round(logrank_css$chisq, 3), "\n")
  cat("P-value:", round(p_css, 4), "\n")
  
  # Cox regression for cause-specific hazard ratio
  cox_css <- coxph(surv_css ~ Treatment, data = css_data)
  hr_css <- exp(cox_css$coefficients)
  hr_ci_css <- exp(confint(cox_css))
  
  cat("\nCause-Specific Hazard Ratio:\n")
  cat("Hazard Ratio:", round(hr_css, 2), "\n")
  cat("95% Confidence Interval:", round(hr_ci_css[1], 2), "-", round(hr_ci_css[2], 2), "\n")
  
  # Compare with overall survival
  cat("\nComparison with Overall Survival:\n")
  cat("Disease-specific deaths:", sum(css_data$death_disease), "of", nrow(css_data), 
      "(", round(mean(css_data$death_disease) * 100, 1), "%)\n")
  
  other_deaths <- sum(competing_survival_data$Outcome == "Dead_of_Other")
  cat("Other deaths:", other_deaths, "(", 
      round(other_deaths / nrow(css_data) * 100, 1), "%)\n")
  
  if(other_deaths / nrow(css_data) > 0.1) {
    cat("Note: Substantial competing mortality present - consider competing risks analysis\n")
  }
}
```

## 3. Competing Risks Analysis

Account for multiple competing events using Fine-Gray subdistribution hazard models.

```{r competing-risks}
# Competing Risks Analysis Example  
# In jamovi: jSurvival > Competing Survival > Analysis Type: Competing Risk

cat("\nCompeting Risks Analysis\n")
cat("=======================\n\n")

if(requireNamespace("survival", quietly = TRUE) && 
   requireNamespace("finalfit", quietly = TRUE) &&
   requireNamespace("cmprsk", quietly = TRUE)) {
  
  # Create competing risks status variable
  crr_data <- competing_survival_data %>%
    mutate(
      status_crr = case_when(
        Outcome %in% c("Alive_w_Disease", "Alive_wo_Disease") ~ 0,  # censored
        Outcome == "Dead_of_Disease" ~ 1,                          # disease death
        Outcome == "Dead_of_Other" ~ 2,                            # competing death
        TRUE ~ 0
      ),
      Treatment_numeric = as.numeric(Treatment) - 1  # for cmprsk
    )
  
  cat("Competing Events Summary:\n")
  event_summary <- crr_data %>%
    group_by(status_crr) %>%
    summarise(
      n = n(),
      percentage = round(n() / nrow(crr_data) * 100, 1),
      .groups = 'drop'
    ) %>%
    mutate(
      event_type = case_when(
        status_crr == 0 ~ "Censored/Alive",
        status_crr == 1 ~ "Disease Death", 
        status_crr == 2 ~ "Other Death"
      )
    )
  print(event_summary[, c("event_type", "n", "percentage")])
  
  # Fine-Gray competing risks regression
  library(cmprsk)
  cuminc_result <- cuminc(
    ftime = crr_data$Overall_Time,
    fstatus = crr_data$status_crr,
    group = crr_data$Treatment
  )
  
  cat("\nCumulative Incidence Function Results:\n")
  
  # Extract cumulative incidence estimates at key time points
  time_points <- c(12, 24, 36)
  
  for (tp in time_points) {
    cat(paste0("\nAt ", tp, " months:\n"))
    
    # Find closest time point in results
    for (i in 1:length(cuminc_result)) {
      group_name <- names(cuminc_result)[i]
      times <- cuminc_result[[i]]$time
      est <- cuminc_result[[i]]$est
      
      closest_idx <- which.min(abs(times - tp))
      if (length(closest_idx) > 0) {
        cat(paste0(group_name, ": ", round(est[closest_idx] * 100, 1), "%\n"))
      }
    }
  }
  
  # Fine-Gray regression for treatment effect
  crr_fit <- crr(
    ftime = crr_data$Overall_Time,
    fstatus = crr_data$status_crr, 
    cov1 = crr_data$Treatment_numeric,
    failcode = 1  # disease death
  )
  
  cat("\nFine-Gray Regression Results (Disease Death):\n")
  cat("Subdistribution Hazard Ratio:", round(exp(crr_fit$coef), 2), "\n")
  cat("95% Confidence Interval:", round(exp(crr_fit$coef - 1.96*sqrt(crr_fit$var)), 2), 
      "-", round(exp(crr_fit$coef + 1.96*sqrt(crr_fit$var)), 2), "\n")
  cat("P-value:", round(2 * (1 - pnorm(abs(crr_fit$coef / sqrt(crr_fit$var)))), 4), "\n")
  
  # Clinical interpretation
  cat("\nClinical Interpretation:\n")
  if (exp(crr_fit$coef) > 1) {
    cat("Experimental therapy associated with higher subdistribution hazard of disease death\n")
  } else {
    cat("Experimental therapy associated with lower subdistribution hazard of disease death\n")
  }
}
```

# Advanced Applications

## 4. Stratified Analysis by Disease Stage

Evaluate competing survival patterns across different disease stages.

```{r stage-analysis}
# Stratified Competing Risks Analysis by Tumor Stage
cat("\nStratified Analysis by Tumor Stage\n")
cat("==================================\n\n")

# Stage distribution
stage_dist <- competing_survival_data %>%
  group_by(Tumor_Stage) %>%
  summarise(
    n = n(),
    disease_deaths = sum(Outcome == "Dead_of_Disease"),
    other_deaths = sum(Outcome == "Dead_of_Other"),
    alive = sum(Outcome %in% c("Alive_w_Disease", "Alive_wo_Disease")),
    .groups = 'drop'
  ) %>%
  mutate(
    disease_death_rate = round(disease_deaths / n * 100, 1),
    other_death_rate = round(other_deaths / n * 100, 1)
  )

cat("Disease and Other Death Rates by Stage:\n")
print(stage_dist[, c("Tumor_Stage", "n", "disease_death_rate", "other_death_rate")])

# Competing risks analysis by stage
if(requireNamespace("cmprsk", quietly = TRUE)) {
  stage_data <- competing_survival_data %>%
    mutate(
      status_crr = case_when(
        Outcome %in% c("Alive_w_Disease", "Alive_wo_Disease") ~ 0,
        Outcome == "Dead_of_Disease" ~ 1,
        Outcome == "Dead_of_Other" ~ 2,
        TRUE ~ 0
      )
    )
  
  # Cumulative incidence by stage
  stage_cuminc <- cuminc(
    ftime = stage_data$Overall_Time,
    fstatus = stage_data$status_crr,
    group = stage_data$Tumor_Stage
  )
  
  cat("\nCumulative Incidence at 24 months by Stage:\n")
  for (i in 1:length(stage_cuminc)) {
    group_name <- names(stage_cuminc)[i]
    times <- stage_cuminc[[i]]$time
    est <- stage_cuminc[[i]]$est
    
    # Find 24-month estimate
    closest_24m <- which.min(abs(times - 24))
    if (length(closest_24m) > 0) {
      cat(paste0(group_name, ": ", round(est[closest_24m] * 100, 1), "%\n"))
    }
  }
  
  # Trend analysis across stages
  stage_numeric <- as.numeric(factor(stage_data$Tumor_Stage, 
                                    levels = c("I", "II", "III", "IV")))
  
  stage_crr <- crr(
    ftime = stage_data$Overall_Time,
    fstatus = stage_data$status_crr,
    cov1 = stage_numeric,
    failcode = 1
  )
  
  cat("\nTrend Analysis Across Stages (per stage increase):\n")
  cat("Subdistribution HR:", round(exp(stage_crr$coef), 2), "\n")
  cat("95% CI:", round(exp(stage_crr$coef - 1.96*sqrt(stage_crr$var)), 2), 
      "-", round(exp(stage_crr$coef + 1.96*sqrt(stage_crr$var)), 2), "\n")
  
  trend_p <- 2 * (1 - pnorm(abs(stage_crr$coef / sqrt(stage_crr$var))))
  cat("P for trend:", round(trend_p, 4), "\n")
  
  if (trend_p < 0.05) {
    cat("Conclusion: Significant trend of increasing disease death risk with advancing stage\n")
  }
}
```

## 5. Biomarker-Based Competing Risks Analysis

Evaluate prognostic biomarkers in the context of competing risks.

```{r biomarker-analysis}
# Biomarker Analysis in Competing Risks Context
cat("\nBiomarker Analysis: EGFR Status\n")
cat("===============================\n\n")

# EGFR status distribution
egfr_summary <- competing_survival_data %>%
  group_by(EGFR_Status, Outcome) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(EGFR_Status) %>%
  mutate(
    total = sum(n),
    percentage = round(n / total * 100, 1)
  )

cat("Outcome Distribution by EGFR Status:\n")
print(egfr_summary[, c("EGFR_Status", "Outcome", "n", "percentage")])

# Competing risks analysis by EGFR status
if(requireNamespace("cmprsk", quietly = TRUE)) {
  egfr_data <- competing_survival_data %>%
    mutate(
      status_crr = case_when(
        Outcome %in% c("Alive_w_Disease", "Alive_wo_Disease") ~ 0,
        Outcome == "Dead_of_Disease" ~ 1,
        Outcome == "Dead_of_Other" ~ 2,
        TRUE ~ 0
      ),
      EGFR_numeric = ifelse(EGFR_Status == "Positive", 1, 0)
    )
  
  # Cumulative incidence by EGFR status
  egfr_cuminc <- cuminc(
    ftime = egfr_data$Overall_Time,
    fstatus = egfr_data$status_crr,
    group = egfr_data$EGFR_Status
  )
  
  cat("\nCumulative Incidence Comparison (EGFR Positive vs Negative):\n")
  
  time_points <- c(12, 24, 36)
  for (tp in time_points) {
    cat(paste0("\nAt ", tp, " months:\n"))
    
    for (i in 1:length(egfr_cuminc)) {
      group_name <- names(egfr_cuminc)[i]
      times <- egfr_cuminc[[i]]$time
      est <- egfr_cuminc[[i]]$est
      
      closest_idx <- which.min(abs(times - tp))
      if (length(closest_idx) > 0) {
        cat(paste0("  ", group_name, ": ", round(est[closest_idx] * 100, 1), "%\n"))
      }
    }
  }
  
  # Fine-Gray regression for EGFR effect
  egfr_crr <- crr(
    ftime = egfr_data$Overall_Time,
    fstatus = egfr_data$status_crr,
    cov1 = egfr_data$EGFR_numeric,
    failcode = 1
  )
  
  cat("\nEGFR Status Effect on Disease Death (Positive vs Negative):\n")
  cat("Subdistribution HR:", round(exp(egfr_crr$coef), 2), "\n")
  cat("95% CI:", round(exp(egfr_crr$coef - 1.96*sqrt(egfr_crr$var)), 2), 
      "-", round(exp(egfr_crr$coef + 1.96*sqrt(egfr_crr$var)), 2), "\n")
  
  egfr_p <- 2 * (1 - pnorm(abs(egfr_crr$coef / sqrt(egfr_crr$var))))
  cat("P-value:", round(egfr_p, 4), "\n")
  
  if (egfr_p < 0.05) {
    if (exp(egfr_crr$coef) > 1) {
      cat("Interpretation: EGFR-positive tumors have significantly higher disease death risk\n")
    } else {
      cat("Interpretation: EGFR-positive tumors have significantly lower disease death risk\n")
    }
  } else {
    cat("Interpretation: No significant association between EGFR status and disease death risk\n")
  }
}
```

# Statistical Considerations

## Model Assumptions and Validation

Important considerations for competing risks analysis.

```{r assumptions}
# Statistical Assumptions and Model Validation
cat("Statistical Considerations for Competing Risks Analysis\n")
cat("======================================================\n\n")

# Sample size considerations
total_events <- sum(competing_survival_data$Outcome %in% c("Dead_of_Disease", "Dead_of_Other"))
disease_events <- sum(competing_survival_data$Outcome == "Dead_of_Disease")
other_events <- sum(competing_survival_data$Outcome == "Dead_of_Other")

cat("Sample Size Assessment:\n")
cat("Total patients:", nrow(competing_survival_data), "\n")
cat("Total events:", total_events, "(", round(total_events/nrow(competing_survival_data)*100, 1), "%)\n")
cat("Disease deaths:", disease_events, "(", round(disease_events/nrow(competing_survival_data)*100, 1), "%)\n")
cat("Other deaths:", other_events, "(", round(other_events/nrow(competing_survival_data)*100, 1), "%)\n")

# Power considerations
min_events_per_group <- 10
groups <- length(unique(competing_survival_data$Treatment))
cat("\nPower Considerations:\n")
cat("Minimum events per group recommended:", min_events_per_group, "\n")
cat("Number of treatment groups:", groups, "\n")

disease_events_per_group <- disease_events / groups
cat("Disease events per group:", round(disease_events_per_group, 1), "\n")

if (disease_events_per_group >= min_events_per_group) {
  cat("Assessment: Adequate power for disease death analysis\n")
} else {
  cat("Assessment: Limited power - consider longer follow-up or larger sample size\n")
}

# Competing risk proportion
competing_proportion <- other_events / total_events
cat("\nCompeting Risk Assessment:\n")
cat("Proportion of deaths from other causes:", round(competing_proportion * 100, 1), "%\n")

if (competing_proportion >= 0.2) {
  cat("Recommendation: Competing risks analysis strongly recommended\n")
  cat("Rationale: Substantial competing mortality may bias cause-specific analysis\n")
} else if (competing_proportion >= 0.1) {
  cat("Recommendation: Consider competing risks analysis\n")
  cat("Rationale: Moderate competing mortality present\n")
} else {
  cat("Recommendation: Cause-specific analysis may be adequate\n")
  cat("Rationale: Limited competing mortality\n")
}

# Follow-up adequacy
median_followup <- median(competing_survival_data$Overall_Time)
max_followup <- max(competing_survival_data$Overall_Time)

cat("\nFollow-up Assessment:\n")
cat("Median follow-up:", round(median_followup, 1), "months\n")
cat("Maximum follow-up:", round(max_followup, 1), "months\n")

if (median_followup >= 24) {
  cat("Assessment: Adequate follow-up for survival analysis\n")
} else {
  cat("Assessment: Limited follow-up - interpret results cautiously\n")
}
```

## Interpretation Guidelines

Framework for interpreting competing risks results.

```{r interpretation}
# Interpretation Guidelines for Competing Risks Analysis
cat("\nInterpretation Guidelines\n")
cat("========================\n\n")

cat("1. Overall vs Cause-Specific vs Competing Risks:\n")
cat("   • Overall Survival: Most clinically relevant for patients\n")
cat("   • Cause-Specific: Useful for understanding disease-specific effects\n") 
cat("   • Competing Risks: Most appropriate when competing events are common\n\n")

cat("2. Subdistribution Hazard Ratio Interpretation:\n")
cat("   • HR > 1: Increased risk of the event of interest\n")
cat("   • HR < 1: Decreased risk of the event of interest\n")
cat("   • Accounts for competing events in the calculation\n\n")

cat("3. Cumulative Incidence Function (CIF):\n")
cat("   • Probability of experiencing the event by time t\n")
cat("   • Accounts for competing events\n")
cat("   • Sum of all CIFs ≤ 1 (remaining probability is survival)\n\n")

cat("4. Clinical Decision Making:\n")
cat("   • Use overall survival for prognosis discussions\n")
cat("   • Use cause-specific for treatment effect evaluation\n")
cat("   • Use competing risks for realistic risk prediction\n\n")

cat("5. Reporting Requirements:\n")
cat("   • Report number at risk and events for each time point\n")
cat("   • Include confidence intervals for all estimates\n")
cat("   • Specify the statistical method used\n")
cat("   • Justify choice between cause-specific and competing risks\n")
```

# Integration with jamovi

## Using competingsurvival in jamovi

Step-by-step guide for jamovi users.

```{r jamovi-guide}
# jamovi Integration Guide for Competing Survival
cat("Using competingsurvival in jamovi\n")
cat("=================================\n\n")

cat("Step-by-Step Instructions:\n\n")

cat("1. Data Preparation:\n")
cat("   • Ensure survival time variable (continuous, numeric)\n")
cat("   • Create outcome variable with 4 levels:\n")
cat("     - Dead of Disease\n")
cat("     - Dead of Other Causes\n") 
cat("     - Alive with Disease\n")
cat("     - Alive without Disease\n")
cat("   • Prepare explanatory variable (categorical)\n\n")

cat("2. Access the Function:\n")
cat("   • Navigate to: jSurvival → Competing Survival\n")
cat("   • Or use: Analyses → ClinicoPath → jSurvival → Competing Survival\n\n")

cat("3. Variable Assignment:\n")
cat("   • Explanatory Variable: Select grouping variable\n")
cat("   • Overall Time (in months): Select survival time column\n")
cat("   • Outcome: Select outcome variable\n")
cat("   • Configure outcome levels:\n")
cat("     - Dead of Disease: Select appropriate level\n")
cat("     - Dead of Other: Select appropriate level\n")
cat("     - Alive w Disease: Select appropriate level\n")
cat("     - Alive w/o Disease: Select appropriate level\n\n")

cat("4. Analysis Selection:\n")
cat("   • Overall: All-cause mortality analysis\n")
cat("   • Cause Specific: Disease-specific mortality only\n")
cat("   • Competing Risk: Full competing risks analysis\n\n")

cat("5. Output Interpretation:\n")
cat("   • Review analysis summary for methods used\n")
cat("   • Examine survival table for hazard ratios\n")
cat("   • For competing risks: review cumulative incidence table\n")
cat("   • For competing risks: examine competing risks plot\n")
cat("   • Read clinical interpretation section\n\n")

cat("6. Troubleshooting:\n")
cat("   • Ensure no missing values in key variables\n")
cat("   • Verify outcome variable has exactly 4 levels\n")
cat("   • Check that follow-up times are positive\n")
cat("   • Ensure adequate sample sizes per group\n")
cat("   • Verify appropriate level assignments\n")
```

# Conclusion

The **competingsurvival** function provides comprehensive tools for analyzing survival data when multiple competing events can occur. The choice between overall survival, cause-specific survival, and competing risks analysis depends on the research question, the proportion of competing events, and the clinical context.

## Key Takeaways

1. **Method Selection**: Choose the appropriate analysis type based on research objectives and competing event frequency
2. **Clinical Relevance**: Overall survival is most clinically relevant; competing risks provides most realistic probabilities
3. **Statistical Rigor**: Use Fine-Gray models when competing events comprise >10-20% of total events
4. **Interpretation**: Subdistribution hazard ratios account for competing events in their calculation
5. **Visualization**: Cumulative incidence plots provide intuitive displays of competing risks

## Best Practices

- Always report the proportion of competing events
- Justify the choice of analysis method
- Present confidence intervals for all estimates
- Consider sensitivity analyses using different methods
- Validate findings with clinical domain experts

## Next Steps

- Explore multivariable competing risks models
- Consider time-dependent covariates in competing risks
- Integrate findings with other ClinicoPath survival modules
- Develop prognostic models incorporating competing risks

---

*This vignette demonstrates comprehensive competing survival analysis methodology using the competingsurvival function. Apply these principles to your own clinical datasets for robust survival analysis accounting for competing events.*
