---
title: "jsurvival: Introduction to Survival Analysis for Clinical Research"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

jsurvival provides comprehensive survival analysis tools specifically designed for clinical research and oncology studies. This module offers user-friendly interfaces for time-to-event analysis, from basic Kaplan-Meier curves to advanced competing risks and multivariate Cox regression models.

**Learning Objectives:**

- Master fundamental survival analysis concepts and methods
- Learn Kaplan-Meier survival curve construction and interpretation
- Apply Cox proportional hazards regression for multivariable analysis
- Understand competing risks analysis for complex clinical scenarios
- Implement molecular biomarker survival analysis workflows
- Create publication-ready survival plots and tables

## Module Overview

jsurvival encompasses four main areas of survival analysis:

### 1. **Basic Survival Analysis**
- **Kaplan-Meier Curves**: Non-parametric survival estimation
- **Log-rank Tests**: Compare survival between groups
- **Median Survival**: Time-to-event summary statistics
- **Confidence Intervals**: Uncertainty quantification for survival estimates

### 2. **Regression Models**
- **Cox Proportional Hazards**: Multivariable time-to-event modeling
- **Stratified Cox Models**: Account for non-proportional hazards
- **Time-dependent Covariates**: Dynamic risk factor modeling
- **Model Diagnostics**: Assumption testing and validation

### 3. **Advanced Methods**
- **Competing Risks**: Multiple event types analysis
- **Multistate Models**: Disease progression modeling
- **Parametric Survival**: Weibull, exponential, and other distributions
- **Frailty Models**: Unobserved heterogeneity modeling

### 4. **Clinical Applications**
- **Biomarker Analysis**: Molecular predictor evaluation
- **Treatment Comparisons**: Therapeutic intervention assessment
- **Prognostic Models**: Risk stratification tools
- **Real-World Evidence**: Population-based survival studies

```{r load-packages}
library(ClinicoPath)
library(survival)
library(dplyr)
```

# Getting Started

## Sample Data Overview

```{r sample-data}
# Load molecular pathology survival data
data(ihc_molecular_comprehensive)

# Survival data structure
cat("Survival dataset dimensions:", nrow(ihc_molecular_comprehensive), "×", ncol(ihc_molecular_comprehensive), "\n")

# Key survival variables
survival_vars <- c("Overall_Survival_Months", "Death_Event", "PFS_Months", "Progression_Event")
cat("Survival variables:", paste(survival_vars, collapse = ", "), "\n")

# Molecular biomarkers for stratification
biomarker_vars <- c("EGFR_Mutation", "MSI_Status", "HER2_IHC", "PD_L1_TPS")
cat("Biomarker variables:", paste(biomarker_vars, collapse = ", "), "\n")

# Basic survival summary
survival_summary <- ihc_molecular_comprehensive %>%
  summarise(
    n = n(),
    median_os = median(Overall_Survival_Months),
    death_rate = round(mean(Death_Event) * 100, 1),
    median_pfs = median(PFS_Months),
    progression_rate = round(mean(Progression_Event) * 100, 1)
  )

cat("\nSurvival Summary:\n")
cat("Patients:", survival_summary$n, "\n")
cat("Median OS:", survival_summary$median_os, "months\n")
cat("Death rate:", survival_summary$death_rate, "%\n")
cat("Median PFS:", survival_summary$median_pfs, "months\n")
cat("Progression rate:", survival_summary$progression_rate, "%\n")
```

# Core Survival Analysis

## 1. Kaplan-Meier Survival Analysis

Foundation of survival analysis for time-to-event data.

```{r kaplan-meier}
# Kaplan-Meier Analysis Example
# In jamovi: jsurvival > Survival Analysis

cat("Kaplan-Meier Survival Analysis\n")
cat("==============================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  library(survival)
  
  # Create survival object
  surv_obj <- Surv(ihc_molecular_comprehensive$Overall_Survival_Months,
                   ihc_molecular_comprehensive$Death_Event)
  
  # Overall survival curve
  km_overall <- survfit(surv_obj ~ 1, data = ihc_molecular_comprehensive)
  
  cat("Overall Survival Analysis:\n")
  print(summary(km_overall, times = c(12, 24, 36, 60))$table)
  
  # Survival by EGFR mutation status
  km_egfr <- survfit(surv_obj ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  
  cat("\nSurvival by EGFR Status:\n")
  egfr_summary <- summary(km_egfr)$table
  print(egfr_summary)
  
  # Log-rank test
  logrank_egfr <- survdiff(surv_obj ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  p_value <- 1 - pchisq(logrank_egfr$chisq, 1)
  
  cat("\nLog-rank test p-value:", round(p_value, 4), "\n")
  
  if(p_value < 0.05) {
    cat("Conclusion: Significant difference in survival by EGFR status\n")
  } else {
    cat("Conclusion: No significant difference in survival by EGFR status\n")
  }
}
```

## 2. Cox Proportional Hazards Regression

Multivariable analysis accounting for multiple prognostic factors.

```{r cox-regression}
# Cox Regression Analysis Example
# In jamovi: jsurvival > Cox Diagnostics

cat("\nCox Proportional Hazards Regression\n")
cat("===================================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  # Multivariable Cox model
  cox_model <- coxph(Surv(Overall_Survival_Months, Death_Event) ~ 
                     EGFR_Mutation + MSI_Status + Age + Grade + Tumor_Mutational_Burden,
                     data = ihc_molecular_comprehensive)
  
  cat("Cox Regression Results:\n")
  cox_summary <- summary(cox_model)
  print(cox_summary$coefficients[, c("coef", "exp(coef)", "Pr(>|z|)")])
  
  # Model performance
  cat("\nModel Performance:\n")
  cat("Concordance:", round(cox_summary$concordance[1], 3), "\n")
  cat("R-squared:", round(cox_summary$rsq[1], 3), "\n")
  cat("Likelihood ratio test p-value:", round(cox_summary$logtest[3], 4), "\n")
  
  # Hazard ratios with confidence intervals
  cat("\nHazard Ratios (95% CI):\n")
  hr_ci <- exp(confint(cox_model))
  hr_estimates <- exp(cox_model$coefficients)
  
  for(i in 1:length(hr_estimates)) {
    var_name <- names(hr_estimates)[i]
    hr <- round(hr_estimates[i], 2)
    ci_lower <- round(hr_ci[i, 1], 2)
    ci_upper <- round(hr_ci[i, 2], 2)
    p_val <- round(cox_summary$coefficients[i, "Pr(>|z|)"], 3)
    
    cat(paste0(var_name, ": HR = ", hr, " (", ci_lower, "-", ci_upper, "), p = ", p_val, "\n"))
  }
}
```

## 3. Competing Risks Analysis

Handle situations where multiple types of events can occur.

```{r competing-risks}
# Competing Risks Analysis Example
# In jamovi: jsurvival > Competing Survival

cat("\nCompeting Risks Analysis\n")
cat("=======================\n\n")

# Create competing events data
# Death from disease vs death from other causes
competing_data <- ihc_molecular_comprehensive %>%
  mutate(
    # Simulate competing events
    death_cause = case_when(
      Death_Event == 0 ~ "Alive",
      Death_Event == 1 & Grade == 3 ~ "Disease",
      Death_Event == 1 & Age > 75 ~ "Other",
      Death_Event == 1 ~ sample(c("Disease", "Other"), 1, prob = c(0.7, 0.3)),
      TRUE ~ "Alive"
    ),
    
    # Competing event indicator
    event_type = case_when(
      death_cause == "Disease" ~ 1,
      death_cause == "Other" ~ 2,
      TRUE ~ 0
    )
  )

# Competing risks summary
competing_summary <- competing_data %>%
  group_by(death_cause) %>%
  summarise(
    n = n(),
    percentage = round(n() / nrow(competing_data) * 100, 1),
    median_time = median(Overall_Survival_Months),
    .groups = 'drop'
  )

cat("Competing Events Summary:\n")
print(competing_summary)

# Event rates by biomarker status
biomarker_events <- competing_data %>%
  group_by(EGFR_Mutation, death_cause) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(EGFR_Mutation) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))

cat("\nEvent Rates by EGFR Status:\n")
print(biomarker_events)

# Clinical interpretation
disease_death_rate <- mean(competing_data$death_cause == "Disease") * 100
cat("\nClinical Interpretation:\n")
cat("Disease-specific death rate:", round(disease_death_rate, 1), "%\n")

if(disease_death_rate > 50) {
  cat("Conclusion: Disease is the predominant cause of death\n")
} else {
  cat("Conclusion: Competing causes of death are significant\n")
}
```

# Biomarker Survival Analysis

## Molecular Stratification

Evaluate prognostic impact of molecular biomarkers.

```{r molecular-stratification}
cat("Molecular Biomarker Survival Analysis\n")
cat("====================================\n\n")

# Multiple biomarker analysis
biomarkers <- c("EGFR_Mutation", "MSI_Status", "HER2_IHC")

for(biomarker in biomarkers) {
  cat(paste0("Survival Analysis for ", biomarker, ":\n"))
  cat(paste0(rep("-", nchar(biomarker) + 25), collapse = ""), "\n")
  
  # Create survival object
  surv_obj <- Surv(ihc_molecular_comprehensive$Overall_Survival_Months,
                   ihc_molecular_comprehensive$Death_Event)
  
  # Survival by biomarker
  formula_str <- paste("surv_obj ~", biomarker)
  km_biomarker <- survfit(as.formula(formula_str), data = ihc_molecular_comprehensive)
  
  # Summary statistics
  biomarker_summary <- summary(km_biomarker)$table
  print(biomarker_summary)
  
  # Log-rank test
  logrank_test <- survdiff(as.formula(formula_str), data = ihc_molecular_comprehensive)
  p_value <- 1 - pchisq(logrank_test$chisq, length(unique(ihc_molecular_comprehensive[[biomarker]])) - 1)
  
  cat("Log-rank test p-value:", round(p_value, 4), "\n")
  
  if(p_value < 0.05) {
    cat("Result: Significant prognostic biomarker\n")
  } else {
    cat("Result: Not a significant prognostic biomarker\n")
  }
  cat("\n")
}
```

## Continuous Biomarker Analysis

Evaluate continuous biomarkers like PD-L1 expression or tumor mutational burden.

```{r continuous-biomarkers}
cat("Continuous Biomarker Analysis\n")
cat("=============================\n\n")

# PD-L1 TPS analysis
pdl1_quartiles <- quantile(ihc_molecular_comprehensive$PD_L1_TPS, 
                          probs = c(0, 0.25, 0.5, 0.75, 1.0))

ihc_with_quartiles <- ihc_molecular_comprehensive %>%
  mutate(
    PD_L1_Quartile = cut(PD_L1_TPS, 
                        breaks = pdl1_quartiles,
                        labels = c("Q1 (Low)", "Q2", "Q3", "Q4 (High)"),
                        include.lowest = TRUE),
    
    PD_L1_Clinical = case_when(
      PD_L1_TPS < 1 ~ "Negative (<1%)",
      PD_L1_TPS < 50 ~ "Low Positive (1-49%)",
      TRUE ~ "High Positive (≥50%)"
    )
  )

cat("PD-L1 TPS Distribution:\n")
pdl1_dist <- table(ihc_with_quartiles$PD_L1_Clinical)
print(pdl1_dist)

# Survival by PD-L1 clinical categories
if(requireNamespace("survival", quietly = TRUE)) {
  surv_obj <- Surv(ihc_with_quartiles$Overall_Survival_Months,
                   ihc_with_quartiles$Death_Event)
  
  km_pdl1 <- survfit(surv_obj ~ PD_L1_Clinical, data = ihc_with_quartiles)
  
  cat("\nSurvival by PD-L1 Expression Level:\n")
  pdl1_summary <- summary(km_pdl1)$table
  print(pdl1_summary)
  
  # Test for trend
  logrank_pdl1 <- survdiff(surv_obj ~ PD_L1_Clinical, data = ihc_with_quartiles)
  p_value_pdl1 <- 1 - pchisq(logrank_pdl1$chisq, 2)
  
  cat("Log-rank test p-value:", round(p_value_pdl1, 4), "\n")
  
  # Tumor mutational burden analysis
  tmb_median <- median(ihc_molecular_comprehensive$Tumor_Mutational_Burden)
  tmb_high <- ihc_molecular_comprehensive$Tumor_Mutational_Burden >= tmb_median
  
  cat("\nTumor Mutational Burden Analysis:\n")
  cat("Median TMB:", tmb_median, "mutations/Mb\n")
  cat("TMB-High cases:", sum(tmb_high), "(", round(mean(tmb_high) * 100, 1), "%)\n")
  
  # TMB survival analysis
  km_tmb <- survfit(surv_obj ~ tmb_high, data = ihc_molecular_comprehensive)
  tmb_summary <- summary(km_tmb)$table
  
  cat("Survival by TMB Status:\n")
  print(tmb_summary)
}
```

# Advanced Applications

## Multivariable Prognostic Models

Develop comprehensive prognostic models incorporating multiple biomarkers.

```{r prognostic-models}
cat("Multivariable Prognostic Model Development\n")
cat("=========================================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  # Comprehensive prognostic model
  prog_model <- coxph(Surv(Overall_Survival_Months, Death_Event) ~ 
                      Age + Grade + 
                      EGFR_Mutation + MSI_Status + 
                      PD_L1_TPS + Tumor_Mutational_Burden,
                      data = ihc_molecular_comprehensive)
  
  # Model summary
  prog_summary <- summary(prog_model)
  
  cat("Comprehensive Prognostic Model:\n")
  print(prog_summary$coefficients[, c("coef", "exp(coef)", "Pr(>|z|)")])
  
  # Model performance metrics
  cat("\nModel Performance:\n")
  cat("Concordance index:", round(prog_summary$concordance[1], 3), "\n")
  cat("Standard error:", round(prog_summary$concordance[2], 3), "\n")
  
  # Risk score calculation
  risk_scores <- predict(prog_model, type = "risk")
  risk_tertiles <- cut(risk_scores, 
                      breaks = quantile(risk_scores, probs = c(0, 1/3, 2/3, 1)),
                      labels = c("Low Risk", "Intermediate Risk", "High Risk"),
                      include.lowest = TRUE)
  
  # Risk group validation
  risk_validation <- ihc_molecular_comprehensive %>%
    mutate(Risk_Group = risk_tertiles) %>%
    group_by(Risk_Group) %>%
    summarise(
      n = n(),
      events = sum(Death_Event),
      event_rate = round(mean(Death_Event) * 100, 1),
      median_survival = median(Overall_Survival_Months),
      .groups = 'drop'
    )
  
  cat("\nRisk Group Validation:\n")
  print(risk_validation)
  
  # Test risk stratification
  surv_risk <- Surv(ihc_molecular_comprehensive$Overall_Survival_Months,
                    ihc_molecular_comprehensive$Death_Event)
  
  km_risk <- survfit(surv_risk ~ risk_tertiles)
  logrank_risk <- survdiff(surv_risk ~ risk_tertiles)
  p_value_risk <- 1 - pchisq(logrank_risk$chisq, 2)
  
  cat("Risk stratification p-value:", round(p_value_risk, 4), "\n")
  
  if(p_value_risk < 0.001) {
    cat("Conclusion: Excellent risk stratification achieved\n")
  } else if(p_value_risk < 0.05) {
    cat("Conclusion: Significant risk stratification achieved\n")
  } else {
    cat("Conclusion: Risk stratification needs improvement\n")
  }
}
```

# Integration with Other Modules

jsurvival works seamlessly with other ClinicoPath modules:

## Connection to ClinicoPathDescriptives
- Baseline characteristics inform survival model covariates
- Quality metrics validate survival endpoint reliability
- Descriptive statistics support survival study design

## Connection to meddecide
- Survival analysis provides time-to-event data for decision trees
- Hazard ratios inform Markov model transition probabilities
- Prognostic models support treatment decision algorithms

## Connection to jjstatsplot
- Survival curves and forest plots enhance presentation
- Statistical visualization supports survival analysis interpretation
- Publication-ready graphics for survival studies

# Best Practices

## Study Design Considerations

1. **Follow-up Time**: Adequate duration to observe events of interest
2. **Censoring**: Understand and appropriately handle censored observations
3. **Sample Size**: Power calculations for survival studies
4. **Endpoint Definition**: Clear, objective, and clinically meaningful endpoints
5. **Covariate Collection**: Comprehensive baseline and time-dependent variables

## Statistical Analysis Guidelines

1. **Proportional Hazards**: Test and address violations when necessary
2. **Model Building**: Use principled variable selection approaches
3. **Validation**: Internal and external validation of prognostic models
4. **Missing Data**: Appropriate handling of incomplete observations
5. **Multiple Testing**: Adjust for multiple comparisons when appropriate

## Reporting Standards

1. **STROBE Guidelines**: Follow reporting guidelines for observational studies
2. **Complete Reporting**: Include all relevant model diagnostics
3. **Clinical Interpretation**: Translate statistical findings to clinical relevance
4. **Limitations**: Discuss study limitations and potential biases
5. **Reproducibility**: Provide sufficient detail for study replication

# Conclusion

jsurvival provides comprehensive tools for modern survival analysis in clinical research, from basic Kaplan-Meier curves to sophisticated prognostic modeling. The integration with molecular biomarker data and other ClinicoPath modules makes it essential for contemporary oncology research and precision medicine applications.

## Next Steps

To explore specific survival analysis techniques:

- **Advanced Methods**: See detailed survival analysis vignettes
- **Biomarker Analysis**: Explore molecular prognostic factor evaluation
- **Model Development**: Learn comprehensive prognostic model building
- **Clinical Applications**: Study real-world survival analysis examples

---

*This introduction provides the foundation for using jsurvival in clinical research. Explore the detailed vignettes for specific techniques and advanced applications.*
