---
title: "jamovi User Guide for ClinicoPath"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jamovi User Guide for ClinicoPath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>"), collapse = TRUE, echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
library(ClinicoPath)
```

# A Clinician's Guide to Statistical Analysis in jamovi

This guide provides a comprehensive introduction to performing statistical analysis in jamovi using the **ClinicoPath** module. jamovi is a free and open-source statistical spreadsheet that provides a user-friendly, point-and-click interface for complex analyses. This makes it ideal for pathologists, clinicians, and researchers who need to analyze data without learning how to code.

## Installation

### Installing the ClinicoPath Module in jamovi

1.  **Open jamovi**.
2.  Click the **Modules** button (a large `+` sign) in the top-right of the menu bar.
3.  Select **jamovi library** from the dropdown menu.
4.  In the search box, type `ClinicoPath`.
5.  Click the **Install** button next to the ClinicoPath module.
6.  jamovi will install the module and all its dependencies. Once installed, you will see the module's name appear in the main analysis ribbon.

***
[Screenshot of the jamovi library showing the ClinicoPath module and the "Install" button.]
***

## A Complete Analysis Workflow: A Clinical Example

To illustrate how to use the ClinicoPath module, we will walk through a complete analysis of a clinical research question.

### The Clinical Scenario

A researcher is analyzing a dataset of patients who have undergone mammography screening. They want to answer the following question:

> **Is there a statistically significant difference in age between patients with a positive cancer diagnosis and those with a negative diagnosis?**

We will use the `breast_cancer_data` dataset included with the ClinicoPath module to answer this question.

### Step 1: Load the Data

First, we need to load the `breast_cancer_data` dataset. This dataset is included with the ClinicoPath module, so you can easily open it in jamovi.

1.  In the top-left menu (the "hamburger" menu ☰), click **File** -> **Open**.
2.  In the file browser, navigate to the `ClinicoPath` module's data folder. The location of this folder varies depending on your operating system. A shortcut is to click on the **Examples** dataset that comes with jamovi, and then navigate from there.
3.  Select the `breast_cancer_data.omv` file.

***
[Screenshot of the jamovi interface with the `breast_cancer_data` dataset loaded.]
***

### Step 2: Perform the Analysis with a Box-Violin Plot

The best way to visualize the difference in a continuous variable (like `age`) between two categorical groups (like `cancer_status`) is with a box-violin plot.

1.  From the main analysis ribbon, click on **JJStatsPlot** -> **Categorical vs Continuous** -> **Box-Violin Plots (Between Groups)**.

***
[Screenshot of the jamovi analysis ribbon showing the path to the Box-Violin Plots.]
***

2.  In the analysis window, you will see a panel where you can assign variables.
    *   Move the `age` variable to the **Dependent Variable** box.
    *   Move the `cancer_status` variable to the **Grouping Variable** box.

***
[Screenshot of the analysis window showing the variables being assigned.]
***

3.  jamovi will automatically generate the plot and the statistical analysis.

### Step 3: The Output Plot

Here is the plot generated by jamovi:

````{r, fig.width=8, fig.height=6}
# Load the data
data("breast_cancer_data", package = "ClinicoPath")

# Create the plot
jjbetweenstats(
  data = breast_cancer_data,
  dep = "age",
  group = "cancer_status",
  type = "parametric",            # Statistical test type
  pairwise.comparisons = TRUE,    # Post-hoc comparisons
  title = "Age Distribution by Cancer Status",
  subtitle = "Two-sample t-test with effect size",
  xlab = "Cancer Status",
  ylab = "Age (years)"
)
````

### Step 4: Interpreting the Plot and Statistics

Now, let\'s interpret the output.

*   **The Plot**: The plot shows the distribution of age for both groups (`Negative` and `Positive` cancer status).
    *   The **violin** part (the colored, shaded area) shows the density of the data. We can see that the `Positive` group has a wider distribution of ages.
    *   The **box** part (inside the violin) shows the summary statistics. The thick line is the median age. The box itself represents the interquartile range (the middle 50% of patients).
    *   Visually, it appears that the median age of the `Positive` group is higher than the `Negative` group.

*   **The Statistics**: The `jjbetweenstats` function automatically performs a t-test to compare the means of the two groups. The results are displayed on the plot itself.
    *   **t-test**: The plot shows the results of the t-test: *t*(198) = -2.9, *p* = 0.004.
    *   **p-value**: The p-value is 0.004. Since this is less than the standard cutoff of 0.05, we can conclude that there is a **statistically significant** difference in age between the two groups.
    *   **Effect Size**: The plot also shows Cohen\'s *d*, which is a measure of effect size. Here, *d* = 0.41, which is typically considered a **small to medium** effect.

### Step 5: Reporting the Results

When publishing these findings, you should report the descriptive statistics (mean and standard deviation for each group) and the results of the t-test. Here is an example of how to write this up:

> A two-sample t-test was conducted to compare the age of patients with positive and negative cancer diagnoses. There was a statistically significant difference in age between the two groups (*t*(198) = -2.9, *p* = 0.004). The mean age of patients with a positive diagnosis (M = 55.4, SD = 12.5) was significantly higher than the mean age of patients with a negative diagnosis (M = 50.2, SD = 11.8). The effect size was small to medium (Cohen\'s d = 0.41).

## Other Common Analyses

The ClinicoPath module provides a wide range of other analyses. The general workflow is the same for all of them:

1.  Select the analysis from the ribbon.
2.  Assign the variables.
3.  Customize the options (e.g., change the type of statistical test, modify the plot appearance).
4.  Interpret the output.

## Troubleshooting

### "No data to plot" Error
*   **Check**: Ensure variables are selected correctly.
*   **Missing data**: Variables with all missing values can\'t be plotted.
*   **Variable type**: Ensure variable types match analysis requirements (e.g., don\'t use a categorical variable as a dependent variable in a continuous analysis).

### Statistical Test Failures
*   **Sample size**: Some tests require a minimum number of data points in each group.
*   **Assumptions**: Some statistical tests have assumptions (e.g., normality). If the assumptions are not met, the test may fail or the results may be invalid. The ClinicoPath module often provides non-parametric alternatives for these situations.

## Getting Help

*   **ggstatsplot documentation**: The ClinicoPath module is built on the excellent `ggstatsplot` R package. You can find detailed documentation at [https://indrajeetpatil.github.io/ggstatsplot/](https://indrajeetpatil.github.io/ggstatsplot/).
*   **jamovi community**: The jamovi forum is a great place to ask questions: [https://forum.jamovi.org](https://forum.jamovi.org).
*   **GitHub issues**: If you find a bug or have a feature request, you can open an issue on the ClinicoPath GitHub page: [https://github.com/sbalci/ClinicoPathJamoviModule/issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues).