---
title: "Correlations and Scatter Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlations and Scatter Plots}
  %\VignetteEngine{rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, eval = FALSE)
```


```{r setup 2, include=FALSE}
library(ClinicoPath)
data(mtcars)
```

This vignette covers `jjcorrmat()` for creating correlation matrices and
`jjscatterstats()` for scatter plots.

## Correlation matrices with `jjcorrmat()`

`jjcorrmat()` visualises pairwise correlations between numeric variables
and reports the associated tests. Here we look at the relationships
between `mpg`, `hp` and `wt` in the `mtcars` data.

```{r}
jjcorrmat(data = mtcars, dep = c(mpg, hp, wt), grvar = NULL)
```

## Scatter plots with `jjscatterstats()`

`jjscatterstats()` produces a scatter plot with a regression line and
textual output describing the correlation and regression statistics.

```{r}
jjscatterstats(data = mtcars, dep = mpg, group = hp, grvar = NULL)
```


