---
title: "Complete Analysis Gallery - All jjstatsplot Functions"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, eval = TRUE, 
                      fig.width = 8, fig.height = 6, warning = FALSE, message = FALSE)
```

# Complete jjstatsplot Analysis Gallery

This comprehensive guide demonstrates every analysis type available in jjstatsplot with practical examples, use cases, and statistical interpretations.

```{r libraries}
library(ClinicoPath)
library(dplyr)

# Load datasets
data(mtcars)
data(iris)

# Prepare example datasets
mtcars_clean <- mtcars %>%
  mutate(
    cyl = factor(cyl, labels = c("4-cyl", "6-cyl", "8-cyl")),
    am = factor(am, labels = c("Automatic", "Manual")),
    vs = factor(vs, labels = c("V-shaped", "Straight"))
  )

iris_clean <- iris %>%
  mutate(Species = factor(Species))
```

## 1. Histogram Analysis - `jjhistostats()`

### Purpose
Explore the distribution of continuous variables with automatic normality testing and descriptive statistics.

### When to Use
- Examine variable distributions before analysis
- Check normality assumptions
- Identify outliers and skewness
- Compare distributions across groups

### Basic Usage

```{r histogram-basic}
# Single variable distribution
hist_basic <- jjhistostats(
  data = mtcars_clean,
  dep = "mpg",
  grvar = NULL  # No grouping variable
)

hist_basic$plot
```

### Advanced: Multiple Variables

```{r histogram-multiple}
# Multiple variables
hist_multi <- jjhistostats(
  data = iris,
  dep = c("Sepal.Length", "Sepal.Width", "Petal.Length"),
  grvar = NULL  # No grouping variable
)

hist_multi$plot
```

### Grouped Analysis

```{r histogram-grouped}
# Separate histograms by group
hist_grouped <- jjhistostats(
  data = mtcars_clean,
  dep = "mpg",
  grvar = "cyl"
)

hist_grouped$plot2
```

### Statistical Output
- Shapiro-Wilk normality test
- Descriptive statistics (mean, median, SD)
- Sample size information

### Interpretation Tips
- **Normal distribution**: p > 0.05 in normality test
- **Skewness**: Asymmetric distribution patterns
- **Outliers**: Points far from main distribution

---

## 2. Scatter Plots - `jjscatterstats()`

### Purpose
Examine relationships between two continuous variables with correlation analysis and regression fitting.

### When to Use
- Explore bivariate relationships
- Test correlation hypotheses
- Visualize regression relationships
- Identify influential points

### Basic Usage

```{r scatter-basic}
# Basic scatter plot with correlation
scatter_basic <- jjscatterstats(
  data = mtcars_clean,
  dep = "mpg",
  group = "hp",
  grvar = NULL  # No grouping variable
)

scatter_basic$plot
```

### Grouped Analysis

```{r scatter-grouped}
# Separate scatter plots by group
scatter_grouped <- jjscatterstats(
  data = mtcars_clean,
  dep = "mpg",
  group = "hp", 
  grvar = "cyl"
)

scatter_grouped$plot2
```

### Statistical Output
- Pearson/Spearman correlation coefficient
- Confidence intervals
- Regression equation
- R-squared value

### Interpretation Tips
- **Strong correlation**: |r| > 0.7
- **Moderate correlation**: 0.3 < |r| < 0.7
- **Weak correlation**: |r| < 0.3
- **Linear relationship**: Points follow straight line pattern

---

## 3. Box-Violin Plots (Between Groups) - `jjbetweenstats()`

### Purpose
Compare continuous variables across independent groups with appropriate statistical tests.

### When to Use
- Compare means/medians between groups
- Test group differences
- Visualize data distribution by group
- Assess homogeneity of variance

### Basic Usage

```{r between-basic}
# Compare groups
between_basic <- jjbetweenstats(
  data = iris_clean,
  dep = "Sepal.Length",
  group = "Species"
)

between_basic$plot
```

### Multiple Dependent Variables

```{r between-multiple}
# Multiple variables comparison
between_multi <- jjbetweenstats(
  data = iris_clean,
  dep = c("Sepal.Length", "Petal.Length"),
  group = "Species"
)

between_multi$plot
```

### Statistical Output
- ANOVA or Kruskal-Wallis test
- Post-hoc pairwise comparisons
- Effect size measures (eta-squared, Cohen's d)
- Homogeneity of variance tests

### Interpretation Tips
- **Significant difference**: p < 0.05
- **Large effect**: η² > 0.14 or d > 0.8
- **Medium effect**: η² > 0.06 or d > 0.5
- **Small effect**: η² > 0.01 or d > 0.2

---

## 4. Correlation Matrix - `jjcorrmat()`

### Purpose
Visualize correlations among multiple continuous variables simultaneously.

### When to Use
- Explore multivariate relationships
- Identify redundant variables
- Screen variables for analysis
- Data reduction decisions

### Basic Usage

```{r corrmat-basic}
# Correlation matrix
corrmat_basic <- jjcorrmat(
  data = mtcars,
  dep = c("mpg", "hp", "wt", "qsec", "disp"),
  grvar = NULL  # No grouping variable
)

corrmat_basic$plot
```

### Grouped Analysis

```{r corrmat-grouped}
# Separate correlation matrices by group
corrmat_grouped <- jjcorrmat(
  data = iris,
  dep = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
  grvar = "Species"
)

corrmat_grouped$plot2
```

### Statistical Output
- Correlation coefficients matrix
- Significance testing for each correlation
- Confidence intervals

### Interpretation Tips
- **High correlation**: |r| > 0.8 (consider multicollinearity)
- **Pattern recognition**: Look for clusters of correlated variables
- **Significance**: Starred values indicate p < 0.05

---

## 5. Dot Charts - `jjdotplotstats()`

### Purpose
Display group means with confidence intervals and statistical comparisons.

### When to Use
- Present group means clearly
- Show uncertainty (confidence intervals)
- Compare multiple groups
- Publication-ready group comparisons

### Basic Usage

```{r dot-basic}
# Dot chart with group means
dot_basic <- jjdotplotstats(
  data = mtcars_clean,
  dep = "mpg",
  group = "cyl",
  grvar = NULL  # No grouping variable
)

dot_basic$plot
```

### Multiple Variables

```{r dot-multiple}
# Multiple dependent variables
dot_multi <- jjdotplotstats(
  data = iris_clean,
  dep = c("Sepal.Length"),
  group = "Species",
  grvar = NULL  # No grouping variable
)

dot_multi$plot
```

### Statistical Output
- Group means and standard errors
- Confidence intervals
- Statistical test results
- Effect sizes

### Interpretation Tips
- **Non-overlapping CIs**: Likely significant difference
- **Large separation**: Meaningful group differences
- **Consistent patterns**: Across multiple variables

---

## 6. Bar Charts - `jjbarstats()`

### Purpose
Analyze categorical variables and their associations with frequency tests.

### When to Use
- Display frequency distributions
- Test independence between categorical variables
- Visualize contingency tables
- Report categorical data summaries

### Basic Usage

```{r bar-basic}
# Frequency bar chart
bar_basic <- jjbarstats(
  data = mtcars_clean,
  dep = "cyl",
  group = NULL,  # No grouping variable
  grvar = NULL  # No grouping variable
)

bar_basic$plot
```

### Two-Way Analysis

```{r bar-twoway}
# Two categorical variables
bar_twoway <- jjbarstats(
  data = mtcars_clean,
  dep = "cyl",
  group = "am",
  grvar = NULL  # No additional grouping variable
)

bar_twoway$plot
```

### Statistical Output
- Chi-square test of independence
- Cramér's V (effect size)
- Expected vs observed frequencies
- Standardized residuals

### Interpretation Tips
- **Significant association**: p < 0.05 in chi-square test
- **Strong association**: Cramér's V > 0.5
- **Moderate association**: Cramér's V > 0.3
- **Weak association**: Cramér's V > 0.1

---

## 7. Pie Charts - `jjpiestats()`

### Purpose
Display proportional data with statistical testing for categorical variables.

### When to Use
- Show proportions of categories
- Test goodness-of-fit
- Display survey results
- Present percentage distributions

### Basic Usage

```{r pie-basic}
# Basic pie chart
pie_basic <- jjpiestats(
  data = mtcars_clean,
  dep = "cyl",
  group = NULL,  # No grouping variable
  grvar = NULL  # No grouping variable
)

pie_basic$plot1
```

### Grouped Analysis

```{r pie-grouped}
# Separate pie charts by group
pie_grouped <- jjpiestats(
  data = mtcars_clean,
  dep = "cyl",
  group = NULL,  # No grouping variable
  grvar = "am"
)

pie_grouped$plot2
```

### Statistical Output
- Goodness-of-fit tests
- Frequency and percentage tables
- Chi-square statistics

### Interpretation Tips
- **Equal proportions**: Uniform distribution
- **Dominant category**: One slice much larger
- **Balanced distribution**: Similar slice sizes

---

## 8. Within-Subjects Analysis - `jjwithinstats()`

### Purpose
Compare repeated measures or matched observations with appropriate paired tests.

### When to Use
- Pre-post comparisons
- Repeated measures designs
- Matched pairs analysis
- Longitudinal data

### Example with Paired Data

```{r within-example}
# Create example paired data
paired_data <- data.frame(
  Subject = rep(1:20, 2),
  Time = rep(c("Pre", "Post"), each = 20),
  Score = c(rnorm(20, 50, 10), rnorm(20, 55, 10)),
  Group = rep(c("Treatment", "Control"), 20)
)

# Reshape for within-subjects analysis
paired_wide <- paired_data %>%
  tidyr::pivot_wider(names_from = Time, values_from = Score)

within_basic <- jjwithinstats(
  data = paired_wide,
  dep1 = "Pre",
  dep2 = "Post",
  dep3 = NULL,  # No third dependent variable
  dep4 = NULL,  # No fourth dependent variable
  )

within_basic$plot
```


### Statistical Output
- Paired t-test or Wilcoxon signed-rank test
- Effect size (Cohen's d for paired data)
- Confidence intervals for differences

### Interpretation Tips
- **Significant change**: p < 0.05
- **Practical significance**: Consider effect size magnitude
- **Direction of change**: Positive vs negative differences

---

## 9. Waffle Charts - `jwaffle()`

### Purpose
Display proportional data using a grid of squares for intuitive visualization.

### When to Use
- Show parts of a whole
- Display survey results
- Present demographic breakdowns
- Alternative to pie charts

### Basic Usage

```{r waffle-basic}
# Waffle chart
waffle_basic <- jwaffle(
  data = mtcars_clean,
  counts = "hp",
  groups = "cyl",
  facet = "am"
)

waffle_basic$plot
```

### Interpretation Tips
- **Visual proportion**: Each square represents equal units
- **Easy comparison**: Squares easier to compare than pie slices
- **Pattern recognition**: Spatial arrangement shows patterns

---

## Choosing the Right Analysis

### Decision Tree

```{r decision-tree, eval=FALSE}
# Pseudocode decision tree
if (data_type == "one_continuous") {
  use_jjhistostats()
} else if (data_type == "two_continuous") {
  use_jjscatterstats() # or jjcorrmat() for multiple variables
} else if (data_type == "continuous_by_groups") {
  if (groups_independent) {
    use_jjbetweenstats()
  } else {
    use_jjwithinstats()
  }
} else if (data_type == "categorical") {
  if (display_preference == "bars") {
    use_jjbarstats()
  } else if (display_preference == "pie") {
    use_jjpiestats()
  } else if (display_preference == "waffle") {
    use_waffle()
  }
} else if (data_type == "group_means") {
  use_jjdotplotstats()
}
```

### Data Type Requirements

| Analysis | Dependent Variable | Grouping Variable | Sample Size |
|----------|-------------------|-------------------|-------------|
| Histogram | Continuous | Optional categorical | n ≥ 10 |
| Scatter | 2 × Continuous | Optional categorical | n ≥ 10 |
| Box-Violin | Continuous | Categorical (2+ levels) | n ≥ 5 per group |
| Correlation Matrix | 2+ × Continuous | Optional categorical | n ≥ 10 |
| Dot Chart | Continuous | Categorical (2+ levels) | n ≥ 3 per group |
| Bar Chart | Categorical | Optional categorical | n ≥ 5 |
| Pie Chart | Categorical | Optional categorical | n ≥ 5 |
| Within-Subjects | 2+ × Continuous | Optional categorical | n ≥ 5 pairs |
| Waffle Chart | Categorical | Optional categorical | n ≥ 10 |

## Best Practices Summary

### Statistical Considerations
1. **Check assumptions** before interpreting results
2. **Report effect sizes** alongside p-values
3. **Consider multiple comparisons** when doing many tests
4. **Validate findings** with appropriate diagnostics

### Visualization Guidelines
1. **Choose appropriate plot** for your data type
2. **Use consistent color schemes** across related plots
3. **Include informative titles** and labels
4. **Consider your audience** when selecting plot complexity

### Workflow Recommendations
1. **Start with exploratory** plots (histograms, scatter plots)
2. **Progress to confirmatory** analyses (between-groups, correlations)
3. **Use grouped analyses** to explore interactions
4. **Combine multiple approaches** for comprehensive understanding

This gallery provides a complete reference for all jjstatsplot analyses. Each function offers extensive customization options - explore the jamovi interface or function documentation for additional parameters and settings.
