---
title: "Comprehensive Guide to Enhanced Statistical Visualization with jggstats"
author: "ClinicoPath Package"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
    dpi = 300,
  echo = TRUE,
  eval = FALSE,
  fig.retina = 2
)
```


```{r setup 2}
# Load required libraries
library(ClinicoPath)
library(dplyr)
library(ggplot2)
library(ggstats)
library(broom)
```

# Enhanced Statistical Visualization with jggstats

## Overview

The `jggstats` function provides advanced statistical visualization capabilities using the ggstats package. It offers multiple analysis types for different statistical contexts:

- **Model Coefficients**: Visualize regression coefficients with confidence intervals
- **Model Comparisons**: Compare multiple models side-by-side
- **Likert Plots**: Specialized visualization for Likert scale data
- **Survey Analysis**: Weighted survey data visualization
- **Proportion Analysis**: Categorical data proportions
- **Cross-tabulation**: Relationship between categorical variables
- **Weighted Means**: Group comparisons with survey weights
- **Cascade Analysis**: Data filtering and sample size tracking

## Understanding Statistical Visualization

### When to Use Each Analysis Type

**Model Coefficients** are ideal for:
- Regression model interpretation
- Effect size visualization
- Confidence interval display
- Variable importance assessment

**Model Comparisons** excel at:
- Comparing nested models
- Model selection visualization
- Effect stability across models
- Sensitivity analysis display

**Likert Plots** are perfect for:
- Survey response visualization
- Attitude scale analysis
- Agreement/disagreement patterns
- Ordered categorical data

**Survey Analysis** handles:
- Complex survey designs
- Weighted population estimates
- Demographic breakdowns
- Representative sampling

## Basic Usage Examples

### Model Coefficient Visualization

```{r model-coefficients}
# Load test data
data("linear_model_data")

# Basic coefficient plot
# jggstats(
#   data = linear_model_data,
#   dependent_var = "outcome",
#   independent_vars = c("age", "sex", "treatment"),
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   plot_title = "Linear Model Coefficients"
# )
```

### Logistic Regression Coefficients

```{r logistic-coefficients}
# Load logistic model data
data("logistic_model_data")

# Logistic regression coefficient plot
# jggstats(
#   data = logistic_model_data,
#   dependent_var = "disease",
#   independent_vars = c("age", "sex", "smoking", "bmi"),
#   analysis_type = "ggcoef_model",
#   model_type = "glm",
#   family = "binomial",
#   sort_coefficients = TRUE,
#   show_significance = TRUE,
#   plot_title = "Disease Risk Factors",
#   plot_subtitle = "Logistic regression odds ratios"
# )
```

## Clinical Research Applications

### Survival Analysis Visualization

```{r survival-analysis}
# Load survival data
data("survival_analysis_data")

# Cox regression coefficient plot
# jggstats(
#   data = survival_analysis_data,
#   model_formula = "Surv(survival_time, event) ~ age + sex + stage + treatment",
#   analysis_type = "ggcoef_model",
#   model_type = "coxph",
#   show_intercept = FALSE,
#   sort_coefficients = TRUE,
#   color_palette = "viridis",
#   plot_title = "Survival Analysis",
#   plot_subtitle = "Cox regression hazard ratios",
#   show_model_summary = TRUE,
#   show_interpretation = TRUE
# )
```

### Biomarker Analysis

```{r biomarker-analysis}
# Focus on biomarker effects
biomarker_data <- logistic_model_data %>%
  filter(!is.na(biomarker_level))

# jggstats(
#   data = biomarker_data,
#   dependent_var = "disease",
#   independent_vars = c("biomarker_level", "age", "sex"),
#   grouping_var = "family_history",
#   analysis_type = "ggcoef_compare",
#   model_type = "glm",
#   family = "binomial",
#   confidence_level = 0.95,
#   color_palette = "set1",
#   plot_title = "Biomarker Effects by Family History",
#   facet_var = "clinic",
#   facet_type = "wrap"
# )
```

### Treatment Response Analysis

```{r treatment-response}
# Analyze treatment effects
# jggstats(
#   data = linear_model_data,
#   dependent_var = "outcome",
#   independent_vars = c("treatment", "age", "baseline_score"),
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   show_intercept = FALSE,
#   standardized = TRUE,
#   show_significance = TRUE,
#   plot_title = "Treatment Effects Analysis",
#   plot_subtitle = "Standardized coefficients",
#   theme_style = "classic"
# )
```

## Survey Data Analysis

### Likert Scale Visualization

```{r likert-analysis}
# Load survey data
data("likert_survey_data")

# Likert plot for satisfaction measures
# jggstats(
#   data = likert_survey_data,
#   dependent_var = "satisfaction_work",
#   grouping_var = "department",
#   analysis_type = "gglikert",
#   likert_levels = 5,
#   color_palette = "viridis",
#   plot_title = "Work Satisfaction by Department",
#   plot_subtitle = "Employee survey responses (1-5 scale)"
# )
```

### Weighted Survey Analysis

```{r weighted-survey}
# Load survey proportion data
data("survey_proportion_data")

# Weighted survey analysis
# jggstats(
#   data = survey_proportion_data,
#   dependent_var = "political_preference",
#   weight_var = "weight",
#   grouping_var = "age_category",
#   analysis_type = "ggsurvey",
#   color_palette = "dark2",
#   plot_title = "Political Preferences by Age",
#   plot_subtitle = "Weighted survey estimates",
#   show_interpretation = TRUE
# )
```

### Cross-tabulation Analysis

```{r cross-tabulation}
# Load cross-tabulation data
data("cross_tab_data")

# Cross-tabulation visualization
# jggstats(
#   data = cross_tab_data,
#   dependent_var = "test_result",
#   grouping_var = "disease_status",
#   analysis_type = "stat_cross",
#   color_palette = "paired",
#   plot_title = "Diagnostic Test Performance",
#   plot_subtitle = "Test results by true disease status",
#   facet_var = "age_group",
#   facet_type = "wrap"
# )
```

## Educational Data Analysis

### Academic Achievement Modeling

```{r academic-achievement}
# Load educational data
data("model_comparison_data")

# Model achievement predictors
# jggstats(
#   data = model_comparison_data,
#   dependent_var = "test_score",
#   independent_vars = c("socioeconomic_status", "parent_education", "school_type", "class_size"),
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   sort_coefficients = TRUE,
#   show_significance = TRUE,
#   plot_title = "Academic Achievement Predictors",
#   plot_subtitle = "Factors affecting test scores",
#   show_model_summary = TRUE
# )
```

### School Comparison Analysis

```{r school-comparison}
# Compare models across districts
# jggstats(
#   data = model_comparison_data,
#   dependent_var = "test_score",
#   independent_vars = c("socioeconomic_status", "parent_education"),
#   grouping_var = "district",
#   analysis_type = "ggcoef_compare",
#   model_type = "lm",
#   confidence_level = 0.90,
#   color_palette = "set1",
#   plot_title = "District-Specific Achievement Models",
#   plot_subtitle = "Model comparison across school districts"
# )
```

## Advanced Statistical Modeling

### Mixed Effects Models

```{r mixed-effects}
# Load longitudinal data
data("mixed_effects_data")

# Mixed effects model visualization
# jggstats(
#   data = mixed_effects_data,
#   dependent_var = "score",
#   independent_vars = c("time_months", "treatment_group", "baseline_severity"),
#   analysis_type = "ggcoef_model",
#   model_type = "lmer",
#   show_intercept = FALSE,
#   sort_coefficients = TRUE,
#   plot_title = "Longitudinal Treatment Effects",
#   plot_subtitle = "Mixed effects model coefficients",
#   theme_style = "minimal"
# )
```

### Complex Survey Design

```{r complex-survey}
# Load complex survey data
data("weighted_analysis_data")

# Weighted means analysis
# jggstats(
#   data = weighted_analysis_data,
#   dependent_var = "income_thousands",
#   grouping_var = "head_education",
#   weight_var = "final_weight",
#   analysis_type = "stat_weighted_mean",
#   color_palette = "viridis",
#   plot_title = "Income by Education Level",
#   plot_subtitle = "Population-weighted estimates",
#   show_interpretation = TRUE,
#   output_format = "both"
# )
```

### Proportion Analysis

```{r proportion-analysis}
# Analyze health coverage proportions
# jggstats(
#   data = weighted_analysis_data,
#   dependent_var = "insurance_coverage",
#   grouping_var = "geography",
#   analysis_type = "stat_prop",
#   color_palette = "dark2",
#   plot_title = "Insurance Coverage by Geography",
#   plot_subtitle = "Proportion analysis",
#   facet_var = "state_region",
#   facet_type = "wrap"
# )
```

## Data Quality and Cascade Analysis

### Sample Size Tracking

```{r cascade-analysis}
# Cascade analysis for data quality
# jggstats(
#   data = linear_model_data,
#   dependent_var = "outcome",
#   independent_vars = c("age", "sex", "treatment", "baseline_score"),
#   analysis_type = "ggcascade",
#   plot_title = "Data Cascade Analysis",
#   plot_subtitle = "Sample size at each analysis step",
#   theme_style = "classic"
# )
```

### Missing Data Patterns

```{r missing-data-cascade}
# Create data with missing values
incomplete_data <- model_comparison_data
incomplete_data$test_score[sample(nrow(incomplete_data), 50)] <- NA
incomplete_data$socioeconomic_status[sample(nrow(incomplete_data), 30)] <- NA

# jggstats(
#   data = incomplete_data,
#   dependent_var = "test_score",
#   independent_vars = c("socioeconomic_status", "parent_education", "school_type"),
#   analysis_type = "ggcascade",
#   plot_title = "Missing Data Impact",
#   plot_subtitle = "Sample reduction due to missing values"
# )
```

## Advanced Customization

### Custom Styling and Themes

```{r custom-styling}
# Advanced styling example
# jggstats(
#   data = logistic_model_data,
#   dependent_var = "disease",
#   independent_vars = c("age", "bmi", "biomarker_level"),
#   grouping_var = "smoking",
#   analysis_type = "ggcoef_model",
#   model_type = "glm",
#   family = "binomial",
#   standardized = FALSE,
#   show_intercept = FALSE,
#   sort_coefficients = TRUE,
#   show_significance = TRUE,
#   confidence_level = 0.95,
#   color_palette = "viridis",
#   theme_style = "dark",
#   plot_title = "Disease Risk Model",
#   plot_subtitle = "Logistic regression with custom styling",
#   x_label = "Log Odds Ratio",
#   y_label = "Risk Factors",
#   show_model_summary = TRUE,
#   show_interpretation = TRUE
# )
```

### Faceted Analysis

```{r faceted-analysis}
# Multi-panel analysis
# jggstats(
#   data = survey_proportion_data,
#   dependent_var = "approve_policy",
#   grouping_var = "political_preference",
#   analysis_type = "stat_prop",
#   facet_var = "education_level",
#   facet_type = "wrap",
#   color_palette = "set1",
#   plot_title = "Policy Approval by Political Preference",
#   plot_subtitle = "Broken down by education level",
#   theme_style = "minimal"
# )
```

## Model Comparison and Selection

### Comparing Different Model Types

```{r model-selection}
# Compare different approaches to the same data
base_data <- linear_model_data

# Model 1: Simple linear
# jggstats(
#   data = base_data,
#   dependent_var = "outcome",
#   independent_vars = c("age", "sex"),
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   plot_title = "Simple Linear Model",
#   show_model_summary = TRUE
# )

# Model 2: With interactions
# jggstats(
#   data = base_data,
#   model_formula = "outcome ~ age * sex + treatment",
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   plot_title = "Model with Interactions",
#   show_model_summary = TRUE
# )
```

### Nested Model Comparison

```{r nested-models}
# Systematic model comparison
# jggstats(
#   data = model_comparison_data,
#   dependent_var = "test_score",
#   independent_vars = c("socioeconomic_status", "parent_education"),
#   grouping_var = "school_type",
#   analysis_type = "ggcoef_compare",
#   model_type = "lm",
#   sort_coefficients = TRUE,
#   plot_title = "Nested Model Comparison",
#   plot_subtitle = "Effects across different school types",
#   output_format = "both"
# )
```

## Interpretation Guidelines

### Understanding Coefficient Plots

Key elements to interpret:

1. **Point estimates**: The central value of the effect
2. **Confidence intervals**: Uncertainty around the estimate
3. **Significance**: Whether intervals cross zero (for coefficients) or one (for odds ratios)
4. **Magnitude**: The practical importance of effects
5. **Direction**: Positive or negative associations

### Statistical Significance

```{r significance-interpretation}
# Example with significance markers
# jggstats(
#   data = logistic_model_data,
#   dependent_var = "disease",
#   independent_vars = c("age", "sex", "smoking", "family_history", "bmi"),
#   analysis_type = "ggcoef_model",
#   model_type = "glm",
#   family = "binomial",
#   show_significance = TRUE,
#   sort_coefficients = TRUE,
#   plot_title = "Significance Testing Example",
#   plot_subtitle = "Stars indicate statistical significance",
#   show_interpretation = TRUE
# )
```

## Performance Considerations

### Large Dataset Handling

For large datasets (>10,000 observations), consider:

1. **Model complexity**: Simpler models fit faster
2. **Variable selection**: Focus on key predictors
3. **Sampling**: Use representative subsets for exploration

```{r performance-example}
# Example with larger dataset simulation
large_data <- do.call(rbind, replicate(5, linear_model_data, simplify = FALSE))
large_data$id <- 1:nrow(large_data)

# Efficient analysis
# jggstats(
#   data = large_data,
#   dependent_var = "outcome",
#   independent_vars = c("age", "sex", "treatment"),
#   analysis_type = "ggcoef_model",
#   model_type = "lm",
#   plot_title = "Large Dataset Analysis",
#   plot_subtitle = paste("N =", nrow(large_data), "observations")
# )
```

## Troubleshooting Guide

### Common Issues and Solutions

**Issue**: "Model convergence problems"
- **Solution**: Check for perfect separation in logistic models
- **Solution**: Consider variable scaling for continuous predictors
- **Solution**: Remove highly correlated predictors

**Issue**: "Confidence intervals too wide"
- **Solution**: Increase sample size if possible
- **Solution**: Consider variable transformation
- **Solution**: Check for outliers affecting estimates

**Issue**: "Plot readability problems"
- **Solution**: Reduce number of variables displayed
- **Solution**: Use faceting for complex comparisons
- **Solution**: Adjust text size and plot dimensions

### Data Quality Checks

```{r data-quality}
# Function to check data suitability
check_modeling_data <- function(data, dependent_var, independent_vars) {
  
  cat("Data Quality Assessment for Statistical Modeling\n")
  cat("=============================================\n\n")
  
  # Basic info
  cat("Dataset dimensions:", nrow(data), "rows,", ncol(data), "columns\n\n")
  
  # Check dependent variable
  if (!is.null(dependent_var)) {
    dep_data <- data[[dependent_var]]
    cat("Dependent Variable (", dependent_var, "):\n")
    cat("  - Type:", class(dep_data)[1], "\n")
    cat("  - Missing values:", sum(is.na(dep_data)), "/", length(dep_data), "\n")
    if (is.numeric(dep_data)) {
      cat("  - Range:", round(min(dep_data, na.rm = TRUE), 2), "to", 
          round(max(dep_data, na.rm = TRUE), 2), "\n")
    } else {
      cat("  - Unique values:", length(unique(dep_data[!is.na(dep_data)])), "\n")
    }
    cat("\n")
  }
  
  # Check independent variables
  if (!is.null(independent_vars)) {
    cat("Independent Variables:\n")
    for (var in independent_vars) {
      if (var %in% names(data)) {
        var_data <- data[[var]]
        cat("  -", var, ":")
        cat(" Type:", class(var_data)[1])
        cat(", Missing:", sum(is.na(var_data)))
        if (is.numeric(var_data)) {
          cat(", Range:", round(min(var_data, na.rm = TRUE), 2), "-", 
              round(max(var_data, na.rm = TRUE), 2))
        } else {
          cat(", Levels:", length(unique(var_data[!is.na(var_data)])))
        }
        cat("\n")
      }
    }
  }
}

# Example usage
check_modeling_data(
  data = linear_model_data,
  dependent_var = "outcome",
  independent_vars = c("age", "sex", "treatment")
)
```

## Best Practices

### Analysis Workflow

1. **Exploratory Analysis**: Start with descriptive statistics
2. **Model Building**: Begin with simple models, add complexity gradually
3. **Diagnostics**: Check model assumptions and fit
4. **Visualization**: Use coefficient plots for interpretation
5. **Validation**: Cross-validate findings when possible

### Visualization Principles

1. **Clarity**: Keep plots simple and focused
2. **Consistency**: Use consistent color schemes and styling
3. **Context**: Provide adequate titles and labels
4. **Accuracy**: Ensure statistical correctness
5. **Accessibility**: Consider colorblind-friendly palettes

## Conclusion

The `jggstats` function provides comprehensive tools for statistical visualization across multiple domains. Whether analyzing clinical trials, survey data, or educational outcomes, these enhanced visualizations help communicate statistical findings effectively.

Key advantages:

- **Comprehensive**: Supports multiple analysis types
- **Flexible**: Extensive customization options
- **Accurate**: Built on robust statistical foundations
- **Interpretable**: Clear visual communication of results
- **Scalable**: Handles various dataset sizes and complexities

For additional examples and advanced techniques, explore the comprehensive test datasets included with the package.
