---
title: "jjdotplotstats: Comprehensive Dot Plot Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

The `jjdotplotstats` function provides a powerful interface for creating statistical dot plots that compare continuous variables across different groups. This function is a wrapper around `ggstatsplot::ggdotplotstats` and `ggstatsplot::grouped_ggdotplotstats`, offering both single and grouped dot plot visualizations with comprehensive statistical analysis.

## Key Features

- **Multiple statistical methods**: Parametric (t-tests), nonparametric (Mann-Whitney U), robust, and Bayesian
- **Grouped analysis**: Create separate dot plots for different subgroups
- **Performance optimized**: Uses internal caching to eliminate redundant computations
- **Effect size calculations**: Multiple effect size options (Cohen's d, Hedge's g, eta-squared, omega-squared)
- **Centrality measures**: Optional display of mean, median, or robust centrality measures
- **Comprehensive customization**: Titles, labels, themes, and statistical results display

## Installation and Setup

```{r setup}
#| echo: true
#| eval: false

# Install ClinicoPath if not already installed
if (!require("ClinicoPath")) {
  devtools::install_github("sbalci/ClinicoPathJamoviModule")
}

library(ClinicoPath)
library(ggplot2)
```

## Quick Start

### Basic Dot Plot

```{r basic-example}
#| echo: true
#| eval: false

# Load test data
data(jjdotplotstats_test_data)

# Basic dot plot comparing CRP levels across disease severity
result <- jjdotplotstats(
  data = jjdotplotstats_test_data,
  dep = "crp_level",
  group = "disease_severity",
  typestatistics = "parametric"
)

# View the plot
result$plot
```

### Grouped Dot Plot

```{r grouped-example}
#| echo: true
#| eval: false

# Grouped dot plot by treatment center
result_grouped <- jjdotplotstats(
  data = jjdotplotstats_test_data,
  dep = "esr_level",
  group = "disease_severity",
  grvar = "treatment_center",
  typestatistics = "nonparametric",
  centralityplotting = TRUE
)

# View the grouped plot
result_grouped$plot2
```

## Function Parameters

### Core Parameters

- **`data`**: Input data frame containing the variables to analyze
- **`dep`**: Continuous numeric variable for the dot plot (dependent variable)
- **`group`**: Categorical variable defining the groups for comparison
- **`grvar`**: Optional grouping variable for creating separate plots
- **`typestatistics`**: Type of statistical analysis to perform

### Statistical Methods

The `typestatistics` parameter supports four different approaches:

1. **"parametric"** (default): Independent t-tests or one-way ANOVA
   - Assumes normality and equal variances
   - Most powerful when assumptions are met
   - Effect sizes: Cohen's d or Hedge's g

2. **"nonparametric"**: Mann-Whitney U test or Kruskal-Wallis test
   - Distribution-free method
   - Robust to outliers and non-normality
   - Based on rank comparisons

3. **"robust"**: Robust statistical tests
   - Robust to outliers and violations of normality
   - Uses trimmed means and robust estimators
   - Good compromise between parametric and nonparametric

4. **"bayes"**: Bayesian statistical analysis
   - Provides Bayes factors for evidence assessment
   - Incorporates prior beliefs
   - Gives probabilistic interpretation of results

### Effect Size Options

The `effsizetype` parameter controls effect size calculations:

- **"biased"**: Cohen's d (default)
- **"unbiased"**: Hedge's g (small sample correction)
- **"eta"**: Eta-squared (proportion of variance explained)
- **"omega"**: Omega-squared (less biased than eta-squared)

### Centrality Measures

When `centralityplotting = TRUE`, the `centralitytype` parameter controls the measure displayed:

- **"parametric"**: Mean
- **"nonparametric"**: Median
- **"robust"**: Trimmed mean
- **"bayes"**: MAP (Maximum A Posteriori) estimator

## Performance Optimizations

### Version History

The function has been significantly optimized for performance:

**Previous Issues:**
- Redundant data processing in `.plot()` and `.plot2()` methods
- No caching infrastructure
- Repeated variable conversion and formula construction
- Duplicated options processing

**Current Optimizations:**
- **Data Caching**: Uses `.prepareData()` method to cache processed data
- **Options Caching**: Uses `.prepareOptions()` method to cache option processing
- **Eliminated Redundancy**: Both plot methods now use cached results
- **Better Progress Feedback**: Clear user messaging during processing
- **Efficient Variable Conversion**: Numeric conversion done once and cached

### Performance Benefits

```{r performance-comparison}
#| echo: true
#| eval: false

# Performance comparison (conceptual)
# Before optimization: 
# - Data processed twice (once for each plot)
# - Variable conversion repeated
# - Options processing duplicated

# After optimization:
# - Data processed once and cached
# - Variable conversion done once
# - Significant speedup for large datasets
```

## Advanced Usage Examples

### Multiple Statistical Methods Comparison

```{r multiple-methods}
#| echo: true
#| eval: false

# Compare different statistical methods
methods <- c("parametric", "nonparametric", "robust", "bayes")

results <- list()
for (method in methods) {
  results[[method]] <- jjdotplotstats(
    data = jjdotplotstats_test_data,
    dep = "crp_level",
    group = "disease_severity",
    typestatistics = method
  )
}

# Each result contains the dot plot with method-specific statistics
# results$parametric$plot
# results$nonparametric$plot
# etc.
```

### Clinical Biomarker Analysis

```{r clinical-example}
#| echo: true
#| eval: false

# Comprehensive biomarker analysis
biomarkers <- c("crp_level", "esr_level", "platelet_count", "hemoglobin_level")

# Analyze each biomarker across disease severity groups
biomarker_results <- list()

for (biomarker in biomarkers) {
  biomarker_results[[biomarker]] <- jjdotplotstats(
    data = jjdotplotstats_test_data,
    dep = biomarker,
    group = "disease_severity",
    typestatistics = "parametric",
    effsizetype = "unbiased",
    centralityplotting = TRUE,
    centralitytype = "parametric",
    mytitle = paste("Distribution of", biomarker, "by Disease Severity"),
    xtitle = biomarker,
    ytitle = "Disease Severity"
  )
}
```

### Gender-Stratified Analysis

```{r gender-stratified}
#| echo: true
#| eval: false

# Analyze hemoglobin levels by disease severity, stratified by gender
hgb_gender_analysis <- jjdotplotstats(
  data = jjdotplotstats_test_data,
  dep = "hemoglobin_level",
  group = "disease_severity",
  grvar = "gender",
  typestatistics = "parametric",
  effsizetype = "unbiased",
  centralityplotting = TRUE,
  mytitle = "Hemoglobin Levels by Disease Severity and Gender",
  xtitle = "Hemoglobin Level (g/dL)",
  ytitle = "Disease Severity"
)
```

### Treatment Response Analysis

```{r treatment-response}
#| echo: true
#| eval: false

# Analyze biomarker levels by treatment response
# (excluding healthy patients who don't have treatment)
treated_patients <- subset(jjdotplotstats_test_data, 
                          treatment_response != "N/A")

response_analysis <- jjdotplotstats(
  data = treated_patients,
  dep = "crp_level",
  group = "treatment_response",
  grvar = "comorbidity_status",
  typestatistics = "nonparametric",
  centralityplotting = TRUE,
  centralitytype = "nonparametric",
  mytitle = "CRP Levels by Treatment Response and Comorbidity Status",
  xtitle = "CRP Level (mg/L)",
  ytitle = "Treatment Response"
)
```

## Data Requirements

### Input Data Structure

The input data should be a data frame with:

- **Continuous variable**: Numeric column for the dot plot (dependent variable)
- **Grouping variable**: Factor or character column defining comparison groups
- **Optional grouping variable**: Additional factor for stratified analysis
- **Complete cases**: Missing values are automatically excluded

### Example Data Structure

```{r data-structure}
#| echo: true
#| eval: false

# Structure of test data
str(jjdotplotstats_test_data)

# Key variables:
# - crp_level: C-reactive protein levels (mg/L)
# - esr_level: Erythrocyte sedimentation rate (mm/hr)  
# - platelet_count: Platelet count (×10³/μL)
# - hemoglobin_level: Hemoglobin levels (g/dL)
# - disease_severity: Ordered factor (Healthy, Mild, Moderate, Severe)
# - treatment_center: Factor (Center A, B, C, D)
# - gender: Factor (Male, Female)
```

## Best Practices

### Variable Selection and Preparation

1. **Choose appropriate variables**: 
   - Dependent variable should be continuous and numeric
   - Grouping variables should be categorical with meaningful levels
   
2. **Check distributions**: 
   - Examine histograms and Q-Q plots
   - Consider transformations for highly skewed data
   
3. **Handle missing data**: 
   - Decide on complete case analysis vs. imputation
   - Document missing data patterns

4. **Sample size considerations**:
   - Ensure adequate sample size within each group
   - Consider power analysis for effect size detection

### Statistical Method Selection

- **Use parametric** when data is approximately normal with equal variances
- **Use nonparametric** for ordinal data, non-normal distributions, or small samples
- **Use robust** when there are outliers but parametric interpretation is desired
- **Use Bayesian** when you want to quantify evidence and incorporate prior knowledge

### Effect Size Interpretation

- **Cohen's d / Hedge's g**:
  - 0.2: Small effect
  - 0.5: Medium effect  
  - 0.8: Large effect

- **Eta-squared / Omega-squared**:
  - 0.01: Small effect (1% variance explained)
  - 0.06: Medium effect (6% variance explained)
  - 0.14: Large effect (14% variance explained)

### Customization Tips

```{r customization-example}
#| echo: true
#| eval: false

# Highly customized dot plot
custom_plot <- jjdotplotstats(
  data = jjdotplotstats_test_data,
  dep = "crp_level",
  group = "disease_severity",
  typestatistics = "parametric",
  effsizetype = "unbiased",
  centralityplotting = TRUE,
  centralitytype = "parametric",
  mytitle = "C-Reactive Protein Levels Across Disease Severity Groups",
  xtitle = "CRP Level (mg/L)",
  ytitle = "Disease Severity",
  originaltheme = FALSE,  # Use jamovi theme
  resultssubtitle = TRUE  # Show statistical results
)
```

## Troubleshooting

### Common Issues

1. **"Data contains no (complete) rows"**
   - Check for missing values in your variables
   - Ensure at least some complete cases exist
   - Consider imputation or subset analysis

2. **No plot generated**
   - Verify that you have specified both `dep` and `group`
   - Check that variables exist in your dataset
   - Ensure proper variable types (numeric for dep, factor for group)

3. **Slow performance**
   - The optimized version should be much faster
   - For very large datasets, consider sampling
   - Check for memory constraints

### Error Handling

```{r error-handling}
#| echo: true
#| eval: false

# Example error handling
tryCatch({
  result <- jjdotplotstats(
    data = my_data,
    dep = "continuous_var",
    group = "group_var",
    typestatistics = "parametric"
  )
}, error = function(e) {
  message("Error in dot plot analysis: ", e$message)
  message("Check your data structure and variable types")
})
```

## Technical Details

### Underlying Functions

The `jjdotplotstats` function is built on:

- **ggstatsplot::ggdotplotstats**: For single dot plots
- **ggstatsplot::grouped_ggdotplotstats**: For grouped analyses
- **jmvcore**: For data handling and option processing

### Caching Implementation

```{r caching-details}
#| echo: true
#| eval: false

# Internal caching structure (conceptual)
# private$.processedData: Cached cleaned data with converted variables
# private$.processedOptions: Cached option processing and titles
# 
# Benefits:
# - Eliminates redundant jmvcore::naOmit() calls
# - Avoids repeated variable conversion
# - Shares processed data between plot methods
# - Reduces option processing overhead
```

## Clinical Applications

### Biomarker Studies

Dot plots are particularly useful for:

- **Diagnostic biomarker evaluation**: Comparing levels across disease groups
- **Treatment monitoring**: Assessing biomarker changes over time
- **Cohort comparisons**: Analyzing differences between study populations
- **Quality control**: Identifying outliers and batch effects

### Research Scenarios

1. **Drug Development**: Compare efficacy biomarkers across treatment arms
2. **Epidemiology**: Analyze exposure-outcome relationships
3. **Clinical Trials**: Assess treatment effects on continuous endpoints
4. **Laboratory Medicine**: Compare reference ranges across populations

## Conclusion

The optimized `jjdotplotstats` function provides:

- **High performance**: Significant speed improvements through caching
- **Statistical rigor**: Multiple statistical methods with effect sizes
- **Flexibility**: Extensive customization and grouping options
- **Clinical relevance**: Designed for biomarker and clinical research applications
- **Usability**: Clear documentation and comprehensive examples

The function is well-suited for clinical research, biomarker analysis, and any scenario requiring robust comparison of continuous variables across groups.

## Session Information

```{r session-info}
sessionInfo()
```
