---
title: "Comprehensive Descriptive Statistics with jsummarytools"
subtitle: "Professional Data Summaries using summarytools Integration"
date: last-modified
author: "ClinicoPath"
output: html_document
---

# Introduction

The `jsummarytools` function provides comprehensive descriptive statistics using the powerful `summarytools` package. This function is designed specifically for clinical researchers, epidemiologists, and data analysts who need professional-quality descriptive statistics with publication-ready output.

## Key Features

- **Four Analysis Types**: Data frame summaries (dfSummary), frequency tables (freq), descriptive statistics (descr), and cross-tabulations (ctable)
- **Professional Output**: HTML-formatted tables and summaries ready for reports and publications
- **Comprehensive Statistics**: Complete descriptive statistics with missing data reporting
- **Grouping Support**: Stratified analyses by grouping variables
- **Performance Optimized**: Intelligent caching system for faster repeated analyses
- **Flexible Customization**: Extensive options for formatting and statistical content

## Installation and Setup

```{r setup}
# Load required libraries
library(ClinicoPath)
library(dplyr)
library(knitr)

# Set options for better output
options(digits = 3)
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
    dpi = 300,
  echo = TRUE,
  eval = FALSE,
  out.width = "100%"
)

# Check if summarytools is available
if (!requireNamespace("summarytools", quietly = TRUE)) {
  message("Note: summarytools package not available. Install with: install.packages('summarytools')")
}
```

# Data Preparation

Let's create realistic datasets for different types of descriptive analyses:

## Clinical Research Data

```{r clinical-data}
# Create comprehensive clinical dataset
set.seed(123)
clinical_data <- data.frame(
  # Patient identifiers
  patient_id = paste0("PT_", sprintf("%03d", 1:300)),
  
  # Demographics
  age = round(rnorm(300, 65, 12)),
  gender = factor(sample(c("Male", "Female"), 300, replace = TRUE)),
  race = factor(sample(c("White", "Black", "Hispanic", "Asian", "Other"), 
                      300, replace = TRUE, prob = c(0.65, 0.15, 0.12, 0.06, 0.02))),
  
  # Clinical measurements
  height_cm = round(rnorm(300, 170, 12)),
  weight_kg = round(rnorm(300, 75, 15)),
  systolic_bp = round(rnorm(300, 140, 25)),
  diastolic_bp = round(rnorm(300, 85, 15)),
  heart_rate = round(rnorm(300, 75, 12)),
  
  # Laboratory values
  glucose_mg_dl = round(rnorm(300, 110, 30)),
  cholesterol_mg_dl = round(rnorm(300, 200, 40)),
  creatinine_mg_dl = round(rnorm(300, 1.0, 0.3), 2),
  hemoglobin_g_dl = round(rnorm(300, 13.5, 2.0), 1),
  
  # Medical history
  diabetes = factor(sample(c("No", "Yes"), 300, replace = TRUE, prob = c(0.7, 0.3))),
  hypertension = factor(sample(c("No", "Yes"), 300, replace = TRUE, prob = c(0.6, 0.4))),
  smoking_status = factor(sample(c("Never", "Former", "Current"), 
                                300, replace = TRUE, prob = c(0.5, 0.35, 0.15))),
  
  # Treatment information
  treatment_group = factor(sample(c("Control", "Treatment_A", "Treatment_B"), 
                                 300, replace = TRUE)),
  study_site = factor(sample(c("Site_1", "Site_2", "Site_3"), 300, replace = TRUE)),
  
  # Outcomes
  treatment_response = factor(sample(c("No Response", "Partial", "Complete"), 
                                    300, replace = TRUE, prob = c(0.3, 0.5, 0.2))),
  quality_of_life = round(rnorm(300, 70, 18)),
  adverse_events = rpois(300, 1.2)
) %>%
  mutate(
    # Calculate BMI
    bmi = round(weight_kg / (height_cm/100)^2, 1),
    
    # Create realistic medical relationships
    systolic_bp = pmax(90, systolic_bp + 0.3 * age + 
                      ifelse(diabetes == "Yes", 15, 0) +
                      ifelse(smoking_status == "Current", 10, 0)),
    
    glucose_mg_dl = pmax(70, glucose_mg_dl + 
                        ifelse(diabetes == "Yes", 50, 0) + 
                        0.1 * age),
    
    cholesterol_mg_dl = pmax(120, cholesterol_mg_dl + 0.5 * age + 0.8 * bmi),
    
    # Blood pressure category
    bp_category = factor(case_when(
      systolic_bp < 120 ~ "Normal",
      systolic_bp < 130 ~ "Elevated", 
      systolic_bp < 140 ~ "Stage 1",
      TRUE ~ "Stage 2"
    ), levels = c("Normal", "Elevated", "Stage 1", "Stage 2")),
    
    # BMI category
    bmi_category = factor(case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight", 
      TRUE ~ "Obese"
    ), levels = c("Underweight", "Normal", "Overweight", "Obese"))
  )

# Display data structure
str(clinical_data)
```

## Survey Research Data

```{r survey-data}
# Create survey/questionnaire dataset
survey_data <- data.frame(
  respondent_id = 1:500,
  
  # Demographics
  age_group = factor(sample(c("18-25", "26-35", "36-50", "51-65", "65+"), 
                           500, replace = TRUE, prob = c(0.15, 0.25, 0.3, 0.2, 0.1))),
  gender = factor(sample(c("Male", "Female", "Non-binary"), 
                        500, replace = TRUE, prob = c(0.48, 0.5, 0.02))),
  education = factor(sample(c("High School", "Some College", "Bachelor", "Graduate"), 
                           500, replace = TRUE, prob = c(0.25, 0.25, 0.35, 0.15))),
  income_bracket = factor(sample(c("Under $30k", "$30k-$60k", "$60k-$100k", "Over $100k"), 
                                500, replace = TRUE, prob = c(0.25, 0.35, 0.25, 0.15))),
  employment = factor(sample(c("Full-time", "Part-time", "Unemployed", "Student", "Retired"), 
                            500, replace = TRUE, prob = c(0.55, 0.15, 0.1, 0.1, 0.1))),
  
  # Likert scale responses (1-5: Strongly Disagree to Strongly Agree)
  satisfaction_service = sample(1:5, 500, replace = TRUE, prob = c(0.05, 0.1, 0.2, 0.45, 0.2)),
  satisfaction_price = sample(1:5, 500, replace = TRUE, prob = c(0.1, 0.15, 0.25, 0.35, 0.15)),
  satisfaction_quality = sample(1:5, 500, replace = TRUE, prob = c(0.03, 0.07, 0.15, 0.5, 0.25)),
  
  # 7-point agreement scales
  recommend_likelihood = sample(1:7, 500, replace = TRUE, 
                               prob = c(0.05, 0.05, 0.1, 0.15, 0.2, 0.25, 0.2)),
  value_perception = sample(1:7, 500, replace = TRUE, 
                           prob = c(0.08, 0.07, 0.12, 0.18, 0.25, 0.2, 0.1)),
  
  # Frequency responses
  usage_frequency = factor(sample(c("Never", "Rarely", "Sometimes", "Often", "Always"), 
                                 500, replace = TRUE, prob = c(0.05, 0.15, 0.3, 0.35, 0.15))),
  
  # Binary choices
  will_recommend = factor(sample(c("No", "Yes"), 500, replace = TRUE, prob = c(0.25, 0.75))),
  return_customer = factor(sample(c("No", "Yes"), 500, replace = TRUE, prob = c(0.3, 0.7))),
  
  # Continuous measures
  years_customer = round(runif(500, 0, 10), 1),
  monthly_spending = round(exp(rnorm(500, 4.5, 0.8)), 2),
  overall_rating = round(rnorm(500, 4.2, 0.8), 1)
) %>%
  mutate(
    # Constrain ratings to valid ranges
    overall_rating = pmax(1, pmin(5, overall_rating))
  )

str(survey_data)
```

# Basic Usage

## Data Frame Summary (dfSummary)

The most comprehensive overview of your dataset:

```{r dfsummary-basic}
# Basic data frame summary
summary_result <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary"
)

# The result contains comprehensive information about each variable
print(names(summary_result))
```

### Customized dfSummary

```{r dfsummary-custom}
# Customized data frame summary with specific options
custom_summary <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  vars = c("age", "gender", "bmi", "systolic_bp", "diabetes", "treatment_group"),
  show_graphs = TRUE,
  show_labels = TRUE,
  show_variable_numbers = FALSE,
  round_digits = 2
)
```

### Grouped dfSummary

```{r dfsummary-grouped}
# Data frame summary stratified by treatment group
grouped_summary <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary", 
  vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl"),
  group_var = "treatment_group",
  show_graphs = TRUE
)
```

## Frequency Tables (freq)

Detailed frequency distributions for categorical variables:

```{r freq-basic}
# Basic frequency analysis
freq_result <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("satisfaction_service", "usage_frequency", "education")
)
```

### Frequency Tables with Options

```{r freq-options}
# Frequency analysis with cumulative percentages
freq_cumulative <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("recommend_likelihood", "value_perception"),
  include_cumulative = TRUE,
  report_missing = TRUE
)
```

### Grouped Frequency Analysis

```{r freq-grouped}
# Frequency analysis by demographic groups
freq_grouped <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("satisfaction_service", "will_recommend"),
  group_var = "age_group"
)
```

## Descriptive Statistics (descr)

Comprehensive descriptive statistics for numeric variables:

```{r descr-basic}
# Basic descriptive statistics
descr_result <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp", "diastolic_bp", "glucose_mg_dl", "cholesterol_mg_dl")
)
```

### Custom Statistics Selection

```{r descr-custom}
# Custom selection of statistics
descr_custom <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "quality_of_life"),
  stats_to_include = "custom",
  include_mean = TRUE,
  include_median = TRUE,
  include_sd = TRUE,
  include_quartiles = TRUE,
  include_skewness = TRUE,
  round_digits = 3
)
```

### Descriptive Statistics by Group

```{r descr-grouped}
# Descriptive statistics stratified by treatment group
descr_grouped <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("systolic_bp", "glucose_mg_dl", "quality_of_life"),
  group_var = "treatment_group",
  stats_to_include = "all"
)
```

### Different Statistics Sets

```{r descr-sets}
# Basic statistics only
descr_basic <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi"),
  stats_to_include = "basic"
)

# Central tendency measures
descr_central <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr", 
  vars = c("systolic_bp", "diastolic_bp"),
  stats_to_include = "central"
)

# Dispersion measures
descr_dispersion <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("glucose_mg_dl", "cholesterol_mg_dl"),
  stats_to_include = "dispersion"
)
```

## Cross-tabulation (ctable)

Professional cross-tabulation tables with statistical tests:

```{r ctable-basic}
# Basic cross-tabulation
ctable_result <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "treatment_group",
  cross_var2 = "treatment_response"
)
```

### Cross-tabulation with Proportions

```{r ctable-proportions}
# Cross-tabulation with row proportions
ctable_row <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "gender",
  cross_var2 = "bp_category",
  cross_proportions = "row"
)

# Cross-tabulation with column proportions  
ctable_col <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "smoking_status",
  cross_var2 = "diabetes",
  cross_proportions = "col"
)

# Cross-tabulation with total proportions
ctable_total <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "study_site", 
  cross_var2 = "treatment_group",
  cross_proportions = "total"
)
```

### Weighted Cross-tabulation

```{r ctable-weighted}
# Add sample weights to data
clinical_weighted <- clinical_data %>%
  mutate(sample_weight = runif(nrow(clinical_data), 0.5, 2.0))

# Weighted cross-tabulation
ctable_weighted <- jsummarytools(
  data = clinical_weighted,
  analysis_type = "ctable",
  cross_var1 = "treatment_group",
  cross_var2 = "treatment_response", 
  weights_var = "sample_weight",
  cross_proportions = "row"
)
```

# Advanced Features

## Output Styling and Formatting

### Different Output Styles

```{r styling-options}
# Grid style (default)
style_grid <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("satisfaction_service"),
  output_style = "grid"
)

# Simple style
style_simple <- jsummarytools(
  data = survey_data,
  analysis_type = "freq", 
  vars = c("satisfaction_service"),
  output_style = "simple"
)

# Bootstrap style
style_bootstrap <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("satisfaction_service"), 
  output_style = "bootstrap"
)

# Minimal style
style_minimal <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("satisfaction_service"),
  output_style = "minimal"
)
```

### HTML vs ASCII Output

```{r output-formats}
# HTML output (default)
html_output <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  vars = c("age", "gender", "bmi"),
  plain_ascii = FALSE,
  bootstrap_css = TRUE
)

# Plain ASCII output
ascii_output <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  vars = c("age", "gender", "bmi"),
  plain_ascii = TRUE
)
```

### Customization Options

```{r customization}
# Highly customized output
customized_output <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  vars = c("age", "systolic_bp", "treatment_group"),
  show_labels = TRUE,
  show_variable_numbers = TRUE,
  show_graphs = TRUE,
  show_valid_counts = TRUE,
  show_na_counts = TRUE,
  round_digits = 3,
  max_distinct_values = 15,
  headings = TRUE,
  escape_pipe = TRUE
)
```

## Missing Data Analysis

### Dataset with Missing Values

```{r missing-data}
# Create dataset with missing values
clinical_missing <- clinical_data
set.seed(456)

# Introduce different missing patterns
clinical_missing$age[sample(1:300, 25)] <- NA  # MCAR
clinical_missing$weight_kg[sample(1:300, 20)] <- NA  # MCAR

# MAR: older patients less likely to report income-related variables
older_patients <- which(clinical_missing$age > 70)
clinical_missing$quality_of_life[sample(older_patients, length(older_patients) * 0.3)] <- NA

# MNAR: patients with diabetes less likely to report glucose
diabetic_patients <- which(clinical_missing$diabetes == "Yes")
clinical_missing$glucose_mg_dl[sample(diabetic_patients, length(diabetic_patients) * 0.2)] <- NA

# Analysis of missing data patterns
missing_summary <- jsummarytools(
  data = clinical_missing,
  analysis_type = "dfsummary",
  show_na_counts = TRUE,
  show_valid_counts = TRUE
)
```

### Missing Data in Different Analysis Types

```{r missing-analysis}
# Descriptive statistics with missing data
missing_descr <- jsummarytools(
  data = clinical_missing,
  analysis_type = "descr",
  vars = c("age", "weight_kg", "glucose_mg_dl", "quality_of_life")
)

# Frequency analysis with missing data reporting
missing_freq <- jsummarytools(
  data = clinical_missing,
  analysis_type = "freq",
  vars = c("diabetes", "treatment_group"),
  report_missing = TRUE
)

# Cross-tabulation handling missing values
missing_ctable <- jsummarytools(
  data = clinical_missing,
  analysis_type = "ctable",
  cross_var1 = "diabetes",
  cross_var2 = "treatment_group"
)
```

# Real-World Applications

## Clinical Trial Analysis

```{r clinical-trial}
# Comprehensive baseline characteristics table
baseline_characteristics <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  vars = c("age", "gender", "race", "bmi", "systolic_bp", "diastolic_bp", 
           "diabetes", "hypertension", "smoking_status"),
  group_var = "treatment_group",
  show_graphs = FALSE,
  round_digits = 1
)

# Treatment group comparisons
treatment_comparison <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl", "quality_of_life"),
  group_var = "treatment_group",
  stats_to_include = "all"
)

# Outcome analysis
outcome_analysis <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "treatment_group",
  cross_var2 = "treatment_response",
  cross_proportions = "row"
)
```

## Survey Research Analysis

```{r survey-analysis}
# Demographics overview
demographics_overview <- jsummarytools(
  data = survey_data,
  analysis_type = "freq",
  vars = c("age_group", "gender", "education", "income_bracket", "employment")
)

# Satisfaction analysis
satisfaction_analysis <- jsummarytools(
  data = survey_data,
  analysis_type = "descr",
  vars = c("satisfaction_service", "satisfaction_price", "satisfaction_quality", 
           "recommend_likelihood", "overall_rating"),
  stats_to_include = "all"
)

# Customer segmentation
customer_segments <- jsummarytools(
  data = survey_data,
  analysis_type = "ctable",
  cross_var1 = "usage_frequency",
  cross_var2 = "return_customer",
  cross_proportions = "row"
)
```

## Laboratory Data Analysis

```{r lab-analysis}
# Laboratory values overview
lab_overview <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("glucose_mg_dl", "cholesterol_mg_dl", "creatinine_mg_dl", "hemoglobin_g_dl"),
  stats_to_include = "all",
  round_digits = 2
)

# Laboratory values by patient characteristics
lab_by_diabetes <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("glucose_mg_dl", "cholesterol_mg_dl"),
  group_var = "diabetes"
)

lab_by_gender <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("hemoglobin_g_dl", "creatinine_mg_dl"),
  group_var = "gender"
)
```

# Performance and Optimization

## Caching Benefits

```{r performance-demo}
# Large dataset for performance testing
large_dataset <- do.call(rbind, replicate(3, clinical_data, simplify = FALSE))
large_dataset$patient_id <- paste0("PT_", sprintf("%04d", 1:nrow(large_dataset)))

# First run (will be slower)
system.time({
  perf_result1 <- jsummarytools(
    data = large_dataset,
    analysis_type = "descr",
    vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl")
  )
})

# Second run with same parameters (should be faster due to caching)
system.time({
  perf_result2 <- jsummarytools(
    data = large_dataset,
    analysis_type = "descr", 
    vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl")
  )
})
```

## Memory Efficiency

```{r memory-efficiency}
# Check memory usage with different dataset sizes
datasets <- list(
  small = clinical_data[1:50, ],
  medium = clinical_data[1:150, ],
  large = clinical_data
)

memory_usage <- sapply(names(datasets), function(size) {
  gc()  # Garbage collection
  start_memory <- gc()[2,2]  # Memory before
  
  result <- jsummarytools(
    data = datasets[[size]],
    analysis_type = "dfsummary"
  )
  
  end_memory <- gc()[2,2]  # Memory after
  return(end_memory - start_memory)
})

print(memory_usage)
```

# Advanced Statistical Options

## Comprehensive Statistics Selection

```{r advanced-stats}
# All available statistics
all_stats <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp"),
  stats_to_include = "all",
  transpose_output = FALSE
)

# Custom statistics with advanced measures
custom_advanced <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("glucose_mg_dl", "cholesterol_mg_dl"),
  stats_to_include = "custom",
  include_mean = TRUE,
  include_median = TRUE,
  include_mode = FALSE,
  include_sd = TRUE,
  include_var = TRUE,
  include_range = TRUE,
  include_quartiles = TRUE,
  include_skewness = TRUE,
  include_kurtosis = TRUE
)

# Transposed output for publication
transposed_stats <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl"),
  transpose_output = TRUE,
  round_digits = 2
)
```

## Advanced Cross-tabulation

```{r advanced-crosstab}
# Complex cross-tabulation with multiple variables
multi_var_data <- clinical_data %>%
  mutate(
    age_category = cut(age, breaks = c(0, 50, 65, 100), 
                      labels = c("Under 50", "50-65", "Over 65")),
    bmi_category_simple = cut(bmi, breaks = c(0, 25, 30, 50),
                             labels = c("Normal", "Overweight", "Obese"))
  )

# Age and BMI relationship
age_bmi_crosstab <- jsummarytools(
  data = multi_var_data,
  analysis_type = "ctable",
  cross_var1 = "age_category",
  cross_var2 = "bmi_category_simple",
  cross_proportions = "total"
)

# Treatment and outcome by study site
site_treatment_outcome <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "study_site",
  cross_var2 = "treatment_response",
  cross_proportions = "col"
)
```

# Error Handling and Edge Cases

## Small Sample Sizes

```{r small-samples}
# Very small dataset
tiny_dataset <- clinical_data[1:10, ]

small_summary <- jsummarytools(
  data = tiny_dataset,
  analysis_type = "dfsummary"
)

small_descr <- jsummarytools(
  data = tiny_dataset,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp")
)
```

## Single Variable Analysis

```{r single-variable}
# Single numeric variable
single_numeric <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age")
)

# Single categorical variable
single_categorical <- jsummarytools(
  data = clinical_data,
  analysis_type = "freq",
  vars = c("gender")
)
```

## Unusual Data Distributions

```{r unusual-distributions}
# Create data with unusual characteristics
unusual_data <- data.frame(
  id = 1:100,
  constant_var = rep(5, 100),  # No variation
  binary_rare = c(rep(0, 95), rep(1, 5)),  # Very rare event
  skewed_var = exp(rnorm(100, 0, 2)),  # Highly skewed
  outlier_var = c(rnorm(95, 0, 1), rnorm(5, 100, 1)),  # With outliers
  normal_var = rnorm(100, 50, 10)
)

unusual_summary <- jsummarytools(
  data = unusual_data,
  analysis_type = "dfsummary"
)

unusual_descr <- jsummarytools(
  data = unusual_data,
  analysis_type = "descr",
  vars = c("skewed_var", "outlier_var", "normal_var")
)
```

# Best Practices and Guidelines

## Analysis Type Selection Guide

```{r analysis-guide}
# Create guidance table
analysis_guide <- data.frame(
  Analysis_Type = c("dfsummary", "freq", "descr", "ctable"),
  Best_For = c(
    "Initial data exploration and quality assessment",
    "Categorical variable distributions and patterns",
    "Comprehensive numeric variable summaries",
    "Relationships between categorical variables"
  ),
  When_To_Use = c(
    "First look at new dataset, data cleaning",
    "Survey responses, demographic analysis",
    "Baseline characteristics, outcome measures",
    "Treatment comparisons, association testing"
  ),
  Output_Type = c(
    "HTML tables with graphs and summaries",
    "Frequency tables with percentages",
    "Statistical summary tables",
    "Cross-tabulation with chi-square tests"
  )
)

kable(analysis_guide, caption = "jsummarytools Analysis Type Selection Guide")
```

## Statistical Interpretation Guidelines

```{r interpretation-guide}
# Create interpretation guide
interpretation_guide <- data.frame(
  Statistic = c("Mean", "Median", "Standard Deviation", "IQR", "Skewness", "Kurtosis"),
  Interpretation = c(
    "Average value, sensitive to outliers",
    "Middle value, robust to outliers", 
    "Average distance from mean",
    "Range of middle 50% of data",
    "Asymmetry of distribution",
    "Peakedness of distribution"
  ),
  Normal_Range = c(
    "Depends on variable",
    "Should be close to mean if normal",
    "About 1/6 of range for normal data",
    "About 1.35 * SD for normal data", 
    "0 for normal distribution",
    "3 for normal distribution"
  )
)

kable(interpretation_guide, caption = "Statistical Interpretation Guide")
```

## Publication-Ready Tables

```{r publication-tables}
# Table 1: Baseline characteristics
table1 <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp", "diastolic_bp", "glucose_mg_dl"),
  group_var = "treatment_group",
  stats_to_include = "basic",
  round_digits = 1
)

# Table 2: Categorical variables summary
table2 <- jsummarytools(
  data = clinical_data,
  analysis_type = "freq", 
  vars = c("gender", "race", "diabetes", "hypertension", "smoking_status")
)

# Table 3: Treatment outcomes
table3 <- jsummarytools(
  data = clinical_data,
  analysis_type = "ctable",
  cross_var1 = "treatment_group",
  cross_var2 = "treatment_response",
  cross_proportions = "row"
)
```

# Integration with Research Workflows

## Reproducible Research

```{r reproducible-workflow}
# Set parameters for reproducible analysis
analysis_params <- list(
  variables_continuous = c("age", "bmi", "systolic_bp", "glucose_mg_dl"),
  variables_categorical = c("gender", "diabetes", "treatment_group"),
  grouping_variable = "treatment_group",
  decimal_places = 2,
  include_missing = TRUE
)

# Systematic analysis using parameters
systematic_analysis <- function(data, params) {
  results <- list()
  
  # Data overview
  results$overview <- jsummarytools(
    data = data,
    analysis_type = "dfsummary",
    show_na_counts = params$include_missing
  )
  
  # Continuous variables
  results$continuous <- jsummarytools(
    data = data,
    analysis_type = "descr",
    vars = params$variables_continuous,
    group_var = params$grouping_variable,
    round_digits = params$decimal_places
  )
  
  # Categorical variables
  results$categorical <- jsummarytools(
    data = data,
    analysis_type = "freq",
    vars = params$variables_categorical,
    report_missing = params$include_missing
  )
  
  return(results)
}

# Run systematic analysis
study_results <- systematic_analysis(clinical_data, analysis_params)
```

## Quality Control and Validation

```{r quality-control}
# Data quality assessment
quality_check <- jsummarytools(
  data = clinical_data,
  analysis_type = "dfsummary",
  show_na_counts = TRUE,
  show_valid_counts = TRUE,
  max_distinct_values = 20
)

# Outlier detection through descriptive statistics
outlier_detection <- jsummarytools(
  data = clinical_data,
  analysis_type = "descr",
  vars = c("age", "bmi", "systolic_bp", "glucose_mg_dl", "cholesterol_mg_dl"),
  stats_to_include = "all"
)

# Completeness assessment
completeness_check <- clinical_data %>%
  summarise(
    total_observations = n(),
    complete_cases = sum(complete.cases(.)),
    completeness_rate = round(100 * complete_cases / total_observations, 1)
  )

print(completeness_check)
```

# Summary

The `jsummarytools` function provides a comprehensive solution for descriptive statistics in clinical and survey research. Key advantages include:

## Core Capabilities

- **Four Analysis Types**: dfSummary for data exploration, freq for categorical analysis, descr for numeric summaries, ctable for cross-tabulations
- **Professional Output**: HTML-formatted tables ready for reports and publications
- **Grouping Support**: Stratified analyses for treatment comparisons and subgroup analyses
- **Missing Data Handling**: Comprehensive reporting and appropriate handling of missing values

## Performance Features

- **Intelligent Caching**: Automatic caching of results for faster repeated analyses
- **Memory Efficiency**: Optimized for large datasets with smart data preparation
- **Error Handling**: Robust error handling for edge cases and unusual data
- **Validation**: Comprehensive input validation and informative error messages

## Customization Options

- **Output Formatting**: Multiple styles including HTML, ASCII, and Bootstrap CSS
- **Statistical Control**: Custom selection of statistics and precision control
- **Visual Elements**: Optional graphs and charts in dfSummary output
- **Publication Ready**: Professional formatting suitable for academic publications

## Use Cases

- **Clinical Research**: Baseline characteristics, outcome analysis, safety reporting
- **Survey Research**: Demographics, satisfaction analysis, response patterns
- **Quality Control**: Data validation, completeness assessment, outlier detection
- **Regulatory Submissions**: Standardized statistical summaries for regulatory bodies

## Next Steps

- Explore different analysis types with your own data
- Experiment with grouping variables for comparative analyses
- Try different output styles for various reporting needs
- Consider the summarytools package documentation for advanced features

```{r session-info}
sessionInfo()
```
