---
title: "Modern Flowcharts with ggFlowchart"
subtitle: "Creating Publication-Ready Flowcharts with ggplot2 Aesthetics"
author: "ClinicoPath Module"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction to ggFlowchart

The `ggflowchart` function in ClinicoPath provides a modern alternative to traditional flowcharts by leveraging the power and aesthetics of ggplot2. This function complements the existing DiagrammeR-based flowchart module by offering contemporary styling and seamless integration with the ggplot2 ecosystem.

## Key Features

- **ggplot2 Integration**: Native ggplot2 aesthetics and theming support
- **Modern Styling**: Clean, contemporary flowchart appearance
- **Simple Data Structure**: Requires only 'from' and 'to' node connections
- **Customizable Colors**: Multiple color palette options for different contexts
- **Group Support**: Optional node grouping for categorical visualization
- **Publication Ready**: High-quality output suitable for academic publications

## When to Use ggFlowchart vs. DiagrammeR Flowchart

### Choose ggFlowchart for:
- **Modern Aesthetics**: When you need contemporary, ggplot2-style visualization
- **Simple Workflows**: Basic process flows and decision trees
- **ggplot2 Integration**: When working within the ggplot2 ecosystem
- **Quick Prototyping**: Simple edge-based flowchart creation

### Choose DiagrammeR Flowchart for:
- **CONSORT Diagrams**: Complex clinical trial reporting with participant counts
- **Detailed Statistics**: When you need percentage calculations and exclusion tracking
- **Complex Layouts**: Multi-level hierarchical structures with detailed annotations

# Package Setup and Data Loading

```{r setup}
library(ClinicoPath)
library(dplyr)
library(knitr)

# Load comprehensive ggflowchart test datasets
data("simple_process_flow")
data("clinical_decision_flow")
data("research_pipeline_flow")
data("lab_workflow_flow")
data("data_analysis_pipeline")
data("patient_care_pathway")
data("software_dev_flow")
data("minimal_ggflow")
data("complex_decision_tree")
data("manufacturing_process")
data("multi_group_workflow")
data("ggflowchart_comprehensive_data")
```

# Basic ggFlowchart Creation

## Simple Process Flow

Let's start with a basic process flowchart:

```{r basic-process-flow, eval=FALSE}
# Display the simple process flow data structure
knitr::kable(simple_process_flow, 
             caption = "Simple Process Flow Data Structure")

# Create a basic ggflowchart
basic_ggflow <- ggflowchart(
  data = simple_process_flow,
  from_var = "from_node",
  to_var = "to_node",
  group_var = "process_type",
  node_fill_palette = "clinical_blue",
  plot_title = "Basic Process Workflow",
  show_interpretation = TRUE
)
```

This creates a modern flowchart showing:
- **Start**: Initial process entry point
- **Process A & B**: Sequential analysis steps
- **Decision**: Decision point with branching
- **Process C**: Final processing step
- **End**: Process termination

## Clinical Decision Flow

```{r clinical-decision-flow, eval=FALSE}
# Display clinical decision data
knitr::kable(clinical_decision_flow,
             caption = "Clinical Decision Flow Data Structure")

# Create clinical decision flowchart
clinical_ggflow <- ggflowchart(
  data = clinical_decision_flow,
  from_var = "from_step",
  to_var = "to_step",
  group_var = "decision_type",
  node_fill_palette = "modern_gray",
  plot_title = "Clinical Trial Decision Flow",
  show_interpretation = TRUE
)
```

This demonstrates a clinical trial workflow with:
- **Screening Process**: Initial patient assessment
- **Eligibility Decisions**: Inclusion/exclusion criteria
- **Randomization**: Treatment allocation
- **Follow-up**: Patient monitoring and analysis

# Advanced ggFlowchart Features

## Research Pipeline Workflow

```{r research-pipeline, eval=FALSE}
# Display research pipeline data
knitr::kable(research_pipeline_flow,
             caption = "Research Pipeline Flow Structure")

# Create research pipeline flowchart with viridis colors
research_ggflow <- ggflowchart(
  data = research_pipeline_flow,
  from_var = "pipeline_from",
  to_var = "pipeline_to",
  group_var = "phase",
  node_fill_palette = "viridis",
  plot_title = "Complete Research Pipeline",
  show_interpretation = TRUE
)
```

This shows the complete research workflow from concept to publication:
- **Planning Phase**: Literature review and study design
- **Approval Phase**: Ethics review and regulatory compliance
- **Execution Phase**: Data collection and cleaning
- **Analysis Phase**: Primary and secondary analysis
- **Dissemination Phase**: Manuscript preparation and publication

## Laboratory Workflow

```{r lab-workflow, eval=FALSE}
# Display lab workflow data
knitr::kable(lab_workflow_flow,
             caption = "Laboratory Processing Workflow")

# Create lab workflow flowchart
lab_ggflow <- ggflowchart(
  data = lab_workflow_flow,
  from_var = "lab_from",
  to_var = "lab_to",
  group_var = "workflow_type",
  node_fill_palette = "pastel",
  plot_title = "Laboratory Sample Processing",
  show_interpretation = FALSE
)
```

# Color Palette Options

## Available Color Schemes

```{r color-palettes, eval=FALSE}
# Test all available color palettes
palettes <- c("clinical_blue", "modern_gray", "viridis", "set1", "pastel")

for (palette in palettes) {
  cat("Testing palette:", palette, "\n")
  
  palette_flow <- ggflowchart(
    data = simple_process_flow,
    from_var = "from_node",
    to_var = "to_node",
    group_var = "process_type",
    node_fill_palette = palette,
    plot_title = paste("Flowchart with", palette, "palette"),
    show_interpretation = FALSE
  )
}
```

### Palette Descriptions

- **clinical_blue**: Professional medical/clinical appearance with blue tones
- **modern_gray**: Contemporary neutral appearance with gray gradients
- **viridis**: Colorblind-friendly viridis color scale
- **set1**: Bright, distinct colors from RColorBrewer Set1
- **pastel**: Soft, gentle colors for subtle visualization

# Specialized Applications

## Data Analysis Pipeline

```{r data-analysis-pipeline, eval=FALSE}
# Display data analysis pipeline
knitr::kable(data_analysis_pipeline,
             caption = "Data Science Workflow")

# Create data analysis flowchart
analysis_ggflow <- ggflowchart(
  data = data_analysis_pipeline,
  from_var = "analysis_from",
  to_var = "analysis_to",
  group_var = "analysis_phase",
  node_fill_palette = "set1",
  plot_title = "Data Science Pipeline",
  show_interpretation = TRUE
)
```

## Patient Care Pathway

```{r patient-care-pathway, eval=FALSE}
# Display patient care pathway
knitr::kable(patient_care_pathway,
             caption = "Healthcare Process Flow")

# Create patient care flowchart
care_ggflow <- ggflowchart(
  data = patient_care_pathway,
  from_var = "care_from",
  to_var = "care_to",
  group_var = "care_category",
  node_fill_palette = "clinical_blue",
  plot_title = "Patient Care Pathway",
  show_interpretation = TRUE
)
```

This healthcare workflow demonstrates:
- **Intake Process**: Patient admission and initial assessment
- **Assessment Phase**: Diagnosis and treatment planning
- **Implementation**: Treatment delivery and monitoring
- **Review Phase**: Evaluation and outcome assessment

## Software Development Lifecycle

```{r software-development, eval=FALSE}
# Display software development flow
knitr::kable(software_dev_flow,
             caption = "Software Development Lifecycle")

# Create software development flowchart
dev_ggflow <- ggflowchart(
  data = software_dev_flow,
  from_var = "dev_from",
  to_var = "dev_to",
  group_var = "dev_stage",
  node_fill_palette = "modern_gray",
  plot_title = "Software Development Process",
  show_interpretation = TRUE
)
```

# Complex Workflow Examples

## Manufacturing Process Flow

```{r manufacturing-process, eval=FALSE}
# Display manufacturing process
knitr::kable(manufacturing_process,
             caption = "Industrial Manufacturing Workflow")

# Create manufacturing flowchart
manufacturing_ggflow <- ggflowchart(
  data = manufacturing_process,
  from_var = "process_from",
  to_var = "process_to",
  group_var = "manufacturing_type",
  node_fill_palette = "viridis",
  plot_title = "Manufacturing Process Flow",
  show_interpretation = TRUE
)
```

## Complex Decision Tree

```{r complex-decision-tree, eval=FALSE}
# Display complex decision structure
knitr::kable(complex_decision_tree,
             caption = "Multi-Level Decision Tree")

# Create complex decision flowchart
decision_ggflow <- ggflowchart(
  data = complex_decision_tree,
  from_var = "decision_from",
  to_var = "decision_to",
  group_var = "decision_level",
  node_fill_palette = "pastel",
  plot_title = "Complex Decision Tree Structure",
  show_interpretation = TRUE
)
```

# Multi-Group Workflows

## Group-Based Visualization

```{r multi-group-workflow, eval=FALSE}
# Display multi-group workflow
knitr::kable(multi_group_workflow,
             caption = "Multi-Group Workflow Structure")

# Create multi-group flowchart
multigroup_ggflow <- ggflowchart(
  data = multi_group_workflow,
  from_var = "workflow_from",
  to_var = "workflow_to",
  group_var = "group_category",
  node_fill_palette = "clinical_blue",
  plot_title = "Multi-Group Workflow Visualization",
  show_interpretation = TRUE
)
```

This example demonstrates how to visualize parallel workflows:
- **Group A**: Independent workflow stream
- **Group B**: Parallel processing path
- **Group C**: Concurrent operation flow

# Data Structure Requirements

## Basic Requirements

For ggFlowchart to work properly, your data must contain:

1. **From Variable**: Source nodes for connections
2. **To Variable**: Target nodes for connections
3. **Optional Grouping Variable**: Categories for color coding

```{r data-requirements, eval=FALSE}
# Example of minimal data structure
minimal_example <- data.frame(
  source = c("Start", "Process"),
  target = c("Process", "End"),
  category = c("Begin", "Finish")
)

knitr::kable(minimal_example,
             caption = "Minimal Data Structure for ggFlowchart")

# Create minimal flowchart
minimal_ggflow_example <- ggflowchart(
  data = minimal_example,
  from_var = "source",
  to_var = "target",
  group_var = "category",
  node_fill_palette = "clinical_blue",
  plot_title = "Minimal ggFlowchart Example"
)
```

## Data Preparation Guidelines

```{r data-preparation, eval=FALSE}
# Function to validate ggflowchart data
validate_ggflowchart_data <- function(data, from_var, to_var, group_var = NULL) {
  issues <- c()
  
  # Check required variables exist
  if (!from_var %in% names(data)) {
    issues <- c(issues, paste("Missing 'from' variable:", from_var))
  }
  
  if (!to_var %in% names(data)) {
    issues <- c(issues, paste("Missing 'to' variable:", to_var))
  }
  
  # Check optional grouping variable
  if (!is.null(group_var) && !group_var %in% names(data)) {
    issues <- c(issues, paste("Missing 'group' variable:", group_var))
  }
  
  # Check for missing values
  if (from_var %in% names(data)) {
    missing_from <- sum(is.na(data[[from_var]]))
    if (missing_from > 0) {
      issues <- c(issues, paste("Missing values in 'from' variable:", missing_from))
    }
  }
  
  if (to_var %in% names(data)) {
    missing_to <- sum(is.na(data[[to_var]]))
    if (missing_to > 0) {
      issues <- c(issues, paste("Missing values in 'to' variable:", missing_to))
    }
  }
  
  # Check for self-loops
  if (from_var %in% names(data) && to_var %in% names(data)) {
    self_loops <- sum(data[[from_var]] == data[[to_var]], na.rm = TRUE)
    if (self_loops > 0) {
      issues <- c(issues, paste("Self-loops detected:", self_loops))
    }
  }
  
  return(issues)
}

# Example validation
validation_result <- validate_ggflowchart_data(
  simple_process_flow,
  "from_node", 
  "to_node", 
  "process_type"
)

if (length(validation_result) == 0) {
  cat("✓ Data validation passed successfully\n")
} else {
  cat("✗ Data validation issues:\n")
  for (issue in validation_result) {
    cat("  -", issue, "\n")
  }
}
```

# Comprehensive Dataset Analysis

## Multi-Type Workflow Comparison

```{r comprehensive-analysis, eval=FALSE}
# Display comprehensive dataset structure
knitr::kable(ggflowchart_comprehensive_data,
             caption = "Comprehensive Multi-Workflow Dataset")

# Analyze each workflow type
workflow_types <- unique(ggflowchart_comprehensive_data$workflow_type)

for (workflow_type in workflow_types) {
  cat("\n### ", workflow_type, " Analysis\n")
  
  # Subset data for this workflow type
  subset_data <- ggflowchart_comprehensive_data[
    ggflowchart_comprehensive_data$workflow_type == workflow_type, 
  ]
  
  # Calculate workflow statistics
  n_connections <- nrow(subset_data)
  unique_nodes <- length(unique(c(subset_data$from_var, subset_data$to_var)))
  unique_groups <- length(unique(subset_data$group_var))
  
  cat("Connections:", n_connections, "\n")
  cat("Unique nodes:", unique_nodes, "\n")
  cat("Group categories:", unique_groups, "\n")
  
  # Create flowchart for this workflow type
  workflow_ggflow <- ggflowchart(
    data = subset_data,
    from_var = "from_var",
    to_var = "to_var",
    group_var = "group_var",
    node_fill_palette = "clinical_blue",
    plot_title = paste("ggFlowchart:", workflow_type),
    show_interpretation = FALSE
  )
}
```

# Integration with jamovi

## Using ggFlowchart in jamovi

The ggflowchart function integrates seamlessly with jamovi:

1. **Open jamovi** and load your dataset
2. **Navigate** to Exploration → ClinicoPath Reports → ggFlowchart
3. **Configure your flowchart**:
   - Select 'From Node' variable (source connections)
   - Select 'To Node' variable (target connections)
   - Optionally select 'Node Grouping' variable for color coding
   - Choose color palette and customize title
4. **Review outputs**:
   - Modern ggplot2-styled flowchart
   - Optional usage guide and interpretation

## Data Preparation for jamovi

```{r jamovi-data-prep, eval=FALSE}
# Example of preparing data for jamovi jflowchart ggflowchart mode
jamovi_workflow_data <- data.frame(
  process_from = c("Data Import", "Data Cleaning", "Analysis", "Reporting"),
  process_to = c("Data Cleaning", "Analysis", "Reporting", "Publication"),
  stage_type = c("Input", "Preparation", "Analysis", "Output"),
  workflow_id = 1:4
)

knitr::kable(jamovi_workflow_data,
             caption = "Data Structure Ready for jamovi jFlowchart ggplot2 mode")
```

# Comparison with Other Flowchart Methods

## ggFlowchart vs. DiagrammeR Flowchart

```{r comparison-table, eval=FALSE}
comparison_features <- data.frame(
  Feature = c("Aesthetic Style", "Data Structure", "Complexity Support", 
              "Color Options", "Statistics", "Use Case", "Learning Curve"),
  ggFlowchart = c("Modern ggplot2", "Simple edges", "Basic to moderate", 
                  "Multiple palettes", "Basic counts", "Simple workflows", "Easy"),
  DiagrammeR_Flowchart = c("Traditional", "Node + count pairs", "High complexity", 
                           "Custom schemes", "Percentages + exclusions", "CONSORT diagrams", "Moderate"),
  stringsAsFactors = FALSE
)

knitr::kable(comparison_features,
             caption = "Feature Comparison: ggFlowchart vs. DiagrammeR Flowchart")
```

## When to Choose Each Method

### Use ggFlowchart for:
- Modern, publication-ready aesthetics
- Simple process documentation
- Integration with ggplot2 ecosystem
- Quick workflow visualization
- Contemporary design requirements

### Use DiagrammeR Flowchart for:
- CONSORT trial reporting
- Complex hierarchical structures
- Detailed statistical annotations
- Traditional flowchart formats
- Comprehensive participant tracking

# Best Practices and Guidelines

## Design Principles

1. **Keep It Simple**: ggFlowchart works best with clear, straightforward workflows
2. **Use Grouping Wisely**: Color coding should enhance understanding, not complicate it
3. **Choose Appropriate Palettes**: Match colors to your context (clinical, technical, etc.)
4. **Clear Node Names**: Use concise, descriptive labels for nodes

## Common Patterns

```{r common-patterns, eval=FALSE}
# Linear workflow pattern
linear_pattern <- data.frame(
  from = c("Step 1", "Step 2", "Step 3"),
  to = c("Step 2", "Step 3", "Step 4"),
  type = "Sequential"
)

# Branching pattern
branching_pattern <- data.frame(
  from = c("Decision", "Decision", "Option A", "Option B"),
  to = c("Option A", "Option B", "Outcome A", "Outcome B"),
  type = c("Branch", "Branch", "Result", "Result")
)

# Convergent pattern
convergent_pattern <- data.frame(
  from = c("Input A", "Input B", "Process A", "Process B"),
  to = c("Process A", "Process B", "Final Result", "Final Result"),
  type = c("Input", "Input", "Convergence", "Convergence")
)

# Display patterns
knitr::kable(linear_pattern, caption = "Linear Workflow Pattern")
knitr::kable(branching_pattern, caption = "Branching Decision Pattern")
knitr::kable(convergent_pattern, caption = "Convergent Process Pattern")
```

# Troubleshooting Common Issues

## Edge Case Handling

```{r edge-cases, eval=FALSE}
# Test various edge cases
edge_case_tests <- list(
  # Case 1: Self-loops
  self_loop = data.frame(
    from = c("A", "B", "B"),
    to = c("B", "C", "B"),
    category = "Test"
  ),
  
  # Case 2: Disconnected components
  disconnected = data.frame(
    from = c("A", "C", "E"),
    to = c("B", "D", "F"),
    category = c("Group1", "Group1", "Group2")
  ),
  
  # Case 3: Multiple edges between same nodes
  multiple_edges = data.frame(
    from = c("Start", "Start", "Process"),
    to = c("Process", "Process", "End"),
    category = c("Type1", "Type2", "Type1")
  )
)

# Test each edge case
for (case_name in names(edge_case_tests)) {
  cat("\nTesting edge case:", case_name, "\n")
  
  tryCatch({
    result <- ggflowchart(
      data = edge_case_tests[[case_name]],
      from_var = "from",
      to_var = "to",
      group_var = "category",
      plot_title = paste("Edge Case:", case_name)
    )
    cat("✓ Edge case handled successfully\n")
  }, error = function(e) {
    cat("✗ Error:", e$message, "\n")
  })
}
```

## Data Quality Checks

```{r data-quality, eval=FALSE}
# Function to assess flowchart data quality
assess_flowchart_quality <- function(data, from_var, to_var) {
  quality_report <- list()
  
  # Basic statistics
  quality_report$n_edges <- nrow(data)
  quality_report$n_nodes <- length(unique(c(data[[from_var]], data[[to_var]])))
  quality_report$connectivity <- quality_report$n_edges / quality_report$n_nodes
  
  # Missing data
  quality_report$missing_from <- sum(is.na(data[[from_var]]))
  quality_report$missing_to <- sum(is.na(data[[to_var]]))
  
  # Complexity metrics
  from_counts <- table(data[[from_var]])
  to_counts <- table(data[[to_var]])
  quality_report$max_out_degree <- max(from_counts)
  quality_report$max_in_degree <- max(to_counts)
  
  # Self-loops
  quality_report$self_loops <- sum(data[[from_var]] == data[[to_var]], na.rm = TRUE)
  
  return(quality_report)
}

# Example quality assessment
quality_result <- assess_flowchart_quality(
  simple_process_flow, 
  "from_node", 
  "to_node"
)

cat("Flow Quality Assessment:\n")
cat("Edges:", quality_result$n_edges, "\n")
cat("Nodes:", quality_result$n_nodes, "\n")
cat("Connectivity ratio:", round(quality_result$connectivity, 2), "\n")
cat("Self-loops:", quality_result$self_loops, "\n")
```

# Conclusion

The ClinicoPath ggflowchart function provides a powerful, modern alternative for creating flowcharts with contemporary aesthetics. Key advantages include:

✅ **Modern Design**: ggplot2-native styling for contemporary appearance  
✅ **Simple Data Structure**: Requires only 'from' and 'to' connections  
✅ **Flexible Grouping**: Optional color coding for categorical workflows  
✅ **Multiple Palettes**: Professional color schemes for different contexts  
✅ **jamovi Integration**: Seamless interface with interactive parameter adjustment  
✅ **Publication Ready**: High-quality output suitable for academic publications  

## Key Takeaways

1. **Choose the Right Tool**: Use ggflowchart for modern aesthetics, DiagrammeR for complex CONSORT diagrams
2. **Prepare Data Properly**: Ensure clean 'from' and 'to' connections with appropriate grouping
3. **Select Appropriate Colors**: Match palette to context and audience
4. **Keep It Simple**: ggflowchart excels with straightforward workflows
5. **Validate Your Data**: Check for missing values and logical flow patterns

## Next Steps

1. **Experiment** with different color palettes and grouping options
2. **Integrate** with other ClinicoPath functions for comprehensive reporting
3. **Customize** styling using ggplot2 ecosystem extensions
4. **Export** flowcharts for presentations and publications

## Additional Resources

- [ggflowchart Package Documentation](https://nrennie.rbind.io/ggflowchart/)
- [ggplot2 Documentation](https://ggplot2.tidyverse.org/)
- [ClinicoPath Documentation](https://serdarbalci.github.io/ClinicoPathJamoviModule/)

For advanced workflows and integration with other analysis functions, explore the complete ClinicoPath vignette series.

---

*This vignette demonstrates the modern capabilities of the ClinicoPath ggflowchart function for contemporary workflow visualization with ggplot2 aesthetics.*
