---
title: "Publication-Ready Tables with gtsummary"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

The `gtsummary` function in ClinicoPath provides a comprehensive interface to the powerful gtsummary R package for creating publication-ready tables. This function is designed specifically for clinical and pathological research, offering multiple table types, statistical testing, and professional formatting options.

## Key Features

- **Multiple Table Types**: Summary tables, cross-tabulations, regression tables, and survival tables
- **Statistical Testing**: Automatic or manual selection of appropriate statistical tests
- **Professional Formatting**: Bold labels, italics, spanning headers, and custom styling
- **Publication Ready**: HTML, LaTeX, Word, and Markdown output formats
- **Missing Value Handling**: Flexible options for displaying missing data
- **Comprehensive Statistics**: Mean/SD, median/IQR, frequencies, ranges, and more

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  eval = FALSE
)
```


```{r setup 2, include=FALSE}

library(ClinicoPath)
library(dplyr)
library(knitr)
library(gtsummary)
```

# Loading Test Datasets

ClinicoPath provides comprehensive test datasets specifically designed for gtsummary functionality:

```{r load-data}
# Load all gtsummary test datasets
data(gtsummary_clinical_trial, package = "ClinicoPath")
data(gtsummary_survey_data, package = "ClinicoPath")
data(gtsummary_laboratory_data, package = "ClinicoPath")
data(gtsummary_manufacturing_data, package = "ClinicoPath")
data(gtsummary_cross_data, package = "ClinicoPath")

# Display dataset dimensions
cat("Dataset Dimensions:\n")
cat("Clinical Trial:", nrow(gtsummary_clinical_trial), "×", ncol(gtsummary_clinical_trial), "\n")
cat("Survey Data:", nrow(gtsummary_survey_data), "×", ncol(gtsummary_survey_data), "\n")
cat("Laboratory Data:", nrow(gtsummary_laboratory_data), "×", ncol(gtsummary_laboratory_data), "\n")
cat("Manufacturing Data:", nrow(gtsummary_manufacturing_data), "×", ncol(gtsummary_manufacturing_data), "\n")
cat("Cross-tabulation Data:", nrow(gtsummary_cross_data), "×", ncol(gtsummary_cross_data), "\n")
```

# Basic Summary Tables

## Simple Summary Table

Let's start with a basic summary table using clinical trial data:

```{r basic-summary}
# Create a basic summary table
result1 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi", "diabetes"),
  tableType = "summary",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  includeMissing = "ifany"
)

# Display basic information about the table
cat("Basic Summary Table Created Successfully\n")
cat("Variables included:", length(c("age", "gender", "bmi", "diabetes")), "\n")
cat("Missing values:", ifelse(any(is.na(gtsummary_clinical_trial[c("age", "gender", "bmi", "diabetes")])), "Present", "None"), "\n")
```

## Grouped Summary Table (Table 1)

The most common use case is creating a "Table 1" for clinical papers - a summary table grouped by treatment or exposure:

```{r grouped-summary}
# Create a grouped summary table (classic "Table 1")
result2 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "race", "bmi", "diabetes", "disease_stage", "hemoglobin"),
  byvar = "treatment_group",
  tableType = "summary",
  includeOverall = TRUE,
  addPValue = TRUE,
  testMethod = "auto",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  pValueThreshold = 0.05,
  showNHeader = TRUE,
  tableTitle = "Baseline Characteristics by Treatment Group",
  footnote = "Data presented as mean (SD), median (IQR), or n (%). P-values from appropriate statistical tests."
)

cat("Grouped Summary Table Created Successfully\n")
cat("Treatment groups:", length(unique(gtsummary_clinical_trial$treatment_group)), "\n")
cat("P-values included for group comparisons\n")
```

# Cross-Tabulation Tables

Cross-tabulation tables are essential for analyzing relationships between categorical variables:

```{r cross-tabulation}
# Create a cross-tabulation table
result3 <- ClinicoPath::gtsummary(
  data = gtsummary_cross_data,
  vars = c("treatment_response", "drug_dosage"),
  tableType = "cross",
  addPValue = TRUE,
  percentType = "row",
  boldLabels = TRUE,
  tableTitle = "Treatment Response by Drug Dosage",
  footnote = "Row percentages shown. P-value from chi-square test."
)

cat("Cross-tabulation Table Created Successfully\n")
cat("Row variable: treatment_response (", length(unique(gtsummary_cross_data$treatment_response)), " levels)\n")
cat("Column variable: drug_dosage (", length(unique(gtsummary_cross_data$drug_dosage)), " levels)\n")
```

## Complex Cross-Tabulation

```{r complex-cross}
# Create a more complex cross-tabulation
result4 <- ClinicoPath::gtsummary(
  data = gtsummary_cross_data,
  vars = c("side_effects", "compliance"),
  tableType = "cross",
  addPValue = TRUE,
  percentType = "column",
  boldLabels = TRUE,
  boldPValues = TRUE,
  tableTitle = "Side Effects by Treatment Compliance",
  footnote = "Column percentages shown. Higher compliance associated with fewer side effects."
)

cat("Complex Cross-tabulation Created Successfully\n")
cat("Analysis of side effects by compliance level\n")
```

# Regression Tables

Regression tables are useful for presenting model results:

```{r regression-table}
# Create a regression table
result5 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi", "diabetes"),
  tableType = "regression",
  boldLabels = TRUE,
  boldPValues = TRUE,
  tableTitle = "Linear Regression: Age as Outcome",
  footnote = "Model: age ~ gender + bmi + diabetes. Coefficients with 95% CI shown."
)

cat("Regression Table Created Successfully\n")
cat("Model includes demographic and clinical predictors\n")
```

# Laboratory Data Analysis

Clinical laboratory data requires special consideration for reference ranges and flagged values:

```{r laboratory-analysis}
# Create laboratory summary table
result6 <- ClinicoPath::gtsummary(
  data = gtsummary_laboratory_data,
  vars = c("glucose", "creatinine", "cholesterol_flag", "glucose_flag"),
  byvar = "department",
  tableType = "summary",
  addPValue = TRUE,
  testMethod = "auto",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  showNHeader = TRUE,
  tableTitle = "Laboratory Results by Department",
  footnote = "Normal ranges: Glucose 70-125 mg/dL, Creatinine <1.3 mg/dL, Cholesterol <200 mg/dL"
)

cat("Laboratory Analysis Created Successfully\n")
cat("Departments analyzed:", length(unique(gtsummary_laboratory_data$department)), "\n")
cat("Laboratory parameters with reference ranges included\n")
```

# Survey Data Analysis

Survey data often requires special handling for Likert scales and categorical responses:

```{r survey-analysis}
# Create survey data summary
result7 <- ClinicoPath::gtsummary(
  data = gtsummary_survey_data,
  vars = c("age_group", "income_level", "education_level", "overall_satisfaction"),
  byvar = "region",
  tableType = "summary",
  addPValue = TRUE,
  statistics = c("n_percent", "median_iqr"),
  boldLabels = TRUE,
  showNHeader = TRUE,
  includeMissing = "ifany",
  tableTitle = "Survey Demographics by Geographic Region",
  footnote = "Satisfaction scored 1-7 (Very Dissatisfied to Very Satisfied)"
)

cat("Survey Analysis Created Successfully\n")
cat("Geographic regions:", length(unique(gtsummary_survey_data$region)), "\n")
cat("Ordinal variables handled appropriately\n")
```

# Manufacturing Quality Control

Manufacturing data demonstrates the versatility of gtsummary for quality control applications:

```{r manufacturing-qc}
# Create manufacturing quality control table
result8 <- ClinicoPath::gtsummary(
  data = gtsummary_manufacturing_data,
  vars = c("actual_temperature", "pressure_psi", "overall_grade", "within_specification"),
  byvar = "shift",
  tableType = "summary",
  addPValue = TRUE,
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  showNHeader = TRUE,
  tableTitle = "Quality Control Metrics by Production Shift",
  footnote = "Temperature in °F, Pressure in PSI. Specification limits monitored."
)

cat("Manufacturing QC Analysis Created Successfully\n")
cat("Production shifts analyzed:", length(unique(gtsummary_manufacturing_data$shift)), "\n")
cat("Process control parameters included\n")
```

# Advanced Formatting Options

## Custom Styling and Formatting

```{r advanced-formatting}
# Create a highly formatted table
result9 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi_category", "hypertension", "diabetes"),
  byvar = "treatment_group",
  tableType = "summary",
  includeOverall = TRUE,
  addPValue = TRUE,
  addQ = TRUE,  # Add FDR-adjusted p-values
  testMethod = "auto",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldLevels = FALSE,
  boldPValues = TRUE,
  italicizeLabels = FALSE,
  pValueThreshold = 0.05,
  digitsOverall = 2,
  digitsByGroup = 2,
  digitsPValue = 3,
  showNHeader = TRUE,
  outputFormat = "html",
  tableTitle = "Comprehensive Clinical Characteristics",
  footnote = "Q-values shown for multiple comparison correction (FDR method)"
)

cat("Advanced Formatting Applied Successfully\n")
cat("Features: FDR correction, custom digits, styling options\n")
```

## Multiple Statistics Display

```{r multiple-stats}
# Show comprehensive statistics
result10 <- ClinicoPath::gtsummary(
  data = gtsummary_laboratory_data,
  vars = c("glucose", "creatinine", "hemoglobin", "total_cholesterol"),
  byvar = "patient_sex",
  tableType = "summary",
  statistics = c("mean_sd", "median_iqr", "range", "n_percent"),
  includeMissing = "always",
  addPValue = TRUE,
  boldLabels = TRUE,
  tableTitle = "Laboratory Values: Comprehensive Statistics",
  footnote = "All available statistics shown: mean (SD), median (IQR), range, and missing values"
)

cat("Comprehensive Statistics Display Created Successfully\n")
cat("Statistics included: mean, median, range, frequencies, missing values\n")
```

# R Code Generation

The gtsummary function can generate the equivalent R code for transparency and reproducibility:

```{r code-generation}
# Create table with R code output
result11 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi"),
  byvar = "treatment_group",
  tableType = "summary",
  addPValue = TRUE,
  boldLabels = TRUE,
  showCode = TRUE,  # Generate R code
  exportTable = TRUE,  # Export information
  tableTitle = "Example with R Code Generation"
)

cat("R Code Generation Enabled\n")
cat("Generated code can be used for reproducible research\n")
```

# Statistical Testing Options

## Automatic Test Selection

```{r auto-testing}
# Demonstrate automatic test selection
result12 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi", "diabetes", "disease_stage"),
  byvar = "treatment_group",
  tableType = "summary",
  addPValue = TRUE,
  testMethod = "auto",  # Automatic test selection
  pairedTest = FALSE,
  boldPValues = TRUE,
  pValueThreshold = 0.05,
  tableTitle = "Automatic Statistical Test Selection",
  footnote = "Appropriate tests selected automatically: t-test/ANOVA for continuous, chi-square/Fisher for categorical"
)

cat("Automatic Test Selection Applied\n")
cat("Tests selected based on variable type and group structure\n")
```

## Manual Test Selection

```{r manual-testing}
# Demonstrate manual test selection
result13 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "hemoglobin", "creatinine"),
  byvar = "treatment_group",
  tableType = "summary",
  addPValue = TRUE,
  testMethod = "kruskal.test",  # Manual test selection
  boldPValues = TRUE,
  tableTitle = "Manual Test Selection: Kruskal-Wallis",
  footnote = "Kruskal-Wallis test used for all continuous variables (non-parametric)"
)

cat("Manual Test Selection Applied\n")
cat("Kruskal-Wallis test specified for non-parametric analysis\n")
```

# Missing Value Handling

## Different Missing Value Strategies

```{r missing-values}
# Demonstrate different missing value handling
result14 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "education", "biomarker_b", "qol_baseline"),
  byvar = "treatment_group",
  tableType = "summary",
  includeMissing = "always",  # Always show missing values
  addPValue = TRUE,
  boldLabels = TRUE,
  tableTitle = "Missing Value Handling: Always Show",
  footnote = "Missing values shown for all variables, including those with no missing data"
)

cat("Missing Value Handling Demonstrated\n")
cat("Strategy: Always show missing values\n")
```

```{r missing-values-conditional}
# Show missing values only when present
result15 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "education", "biomarker_b", "qol_baseline"),
  byvar = "treatment_group",
  tableType = "summary",
  includeMissing = "ifany",  # Show missing only if present
  addPValue = TRUE,
  boldLabels = TRUE,
  tableTitle = "Missing Value Handling: If Any",
  footnote = "Missing values shown only for variables with missing data"
)

cat("Conditional Missing Value Display\n")
cat("Strategy: Show missing values only when present\n")
```

# Output Formats

## HTML Output (Default)

```{r html-output}
# HTML output format
result16 <- ClinicoPath::gtsummary(
  data = gtsummary_survey_data,
  vars = c("age_group", "income_level", "overall_satisfaction"),
  byvar = "region",
  tableType = "summary",
  outputFormat = "html",
  addPValue = TRUE,
  boldLabels = TRUE,
  tableTitle = "HTML Output Format Example"
)

cat("HTML Output Format Applied\n")
cat("Best for interactive viewing and web-based reports\n")
```

## Markdown Output

```{r markdown-output}
# Markdown output format
result17 <- ClinicoPath::gtsummary(
  data = gtsummary_survey_data,
  vars = c("age_group", "income_level", "overall_satisfaction"),
  byvar = "region",
  tableType = "summary",
  outputFormat = "markdown",
  addPValue = TRUE,
  boldLabels = TRUE,
  tableTitle = "Markdown Output Format Example"
)

cat("Markdown Output Format Applied\n")
cat("Ideal for reproducible reports and documentation\n")
```

# Real-World Applications

## Clinical Trial Table 1

```{r clinical-table1}
# Create a comprehensive clinical trial Table 1
result18 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "race", "education", "bmi_category", "hypertension", 
           "diabetes", "disease_stage", "hemoglobin", "creatinine", "cholesterol"),
  byvar = "treatment_group",
  tableType = "summary",
  includeOverall = TRUE,
  addPValue = TRUE,
  testMethod = "auto",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  showNHeader = TRUE,
  includeMissing = "ifany",
  tableTitle = "Patient Characteristics by Treatment Assignment",
  footnote = "Continuous variables: mean (SD) or median (IQR); Categorical variables: n (%). P-values from t-test, ANOVA, chi-square, or Fisher exact test as appropriate."
)

cat("Clinical Trial Table 1 Created\n")
cat("Publication-ready format with comprehensive baseline characteristics\n")
```

## Biomarker Analysis

```{r biomarker-analysis}
# Biomarker analysis table
result19 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("biomarker_a", "biomarker_b", "hemoglobin", "creatinine", "cholesterol"),
  byvar = "response",
  tableType = "summary",
  addPValue = TRUE,
  testMethod = "wilcox.test",  # Non-parametric for biomarkers
  statistics = c("median_iqr", "range"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  tableTitle = "Biomarker Levels by Treatment Response",
  footnote = "Biomarker values shown as median (IQR) and range. Wilcoxon rank-sum test used for comparisons."
)

cat("Biomarker Analysis Table Created\n")
cat("Non-parametric analysis appropriate for biomarker data\n")
```

# Best Practices and Tips

## Variable Selection Guidelines

1. **Continuous Variables**: Best displayed with mean (SD) and/or median (IQR)
2. **Categorical Variables**: Show frequencies and percentages
3. **Ordinal Variables**: Consider median (IQR) for analysis
4. **Missing Values**: Always document how missing data is handled

## Statistical Testing

1. **Automatic Selection**: Usually appropriate for exploratory analysis
2. **Manual Selection**: Use when you have specific hypotheses or assumptions
3. **Multiple Testing**: Consider FDR correction (addQ = TRUE) when testing many variables
4. **Non-parametric**: Use Kruskal-Wallis or Wilcoxon for non-normal data

## Formatting for Publication

1. **Bold Labels**: Always use for professional appearance
2. **P-value Formatting**: Use 3 decimal places, bold significant values
3. **Table Titles**: Be descriptive and specific
4. **Footnotes**: Explain statistics, tests, and abbreviations

```{r best-practices}
# Demonstrate best practices
result20 <- ClinicoPath::gtsummary(
  data = gtsummary_clinical_trial,
  vars = c("age", "gender", "bmi", "diabetes", "disease_stage", "hemoglobin"),
  byvar = "treatment_group",
  tableType = "summary",
  includeOverall = TRUE,
  addPValue = TRUE,
  addQ = TRUE,  # FDR correction
  testMethod = "auto",
  statistics = c("mean_sd", "median_iqr", "n_percent"),
  boldLabels = TRUE,
  boldPValues = TRUE,
  pValueThreshold = 0.05,
  showNHeader = TRUE,
  includeMissing = "ifany",
  digitsOverall = 2,
  digitsByGroup = 2,
  digitsPValue = 3,
  tableTitle = "Best Practices Example: Complete Clinical Table",
  footnote = "Data: mean (SD), median (IQR), or n (%). Tests: t-test/ANOVA (continuous), chi-square/Fisher (categorical). Q-values: FDR-adjusted p-values for multiple comparisons. Missing values shown when present."
)

cat("Best Practices Example Created\n")
cat("Professional formatting with comprehensive documentation\n")
```

# Summary

The `gtsummary` function in ClinicoPath provides a powerful and flexible interface for creating publication-ready tables. Key advantages include:

1. **Comprehensive Statistics**: Multiple options for displaying data appropriately
2. **Flexible Testing**: Automatic or manual selection of statistical tests
3. **Professional Formatting**: Bold labels, italics, and custom styling
4. **Multiple Output Formats**: HTML, Markdown, LaTeX, and Word compatibility
5. **Missing Value Handling**: Flexible options for displaying incomplete data
6. **R Code Generation**: Transparency and reproducibility
7. **Specialized Datasets**: Comprehensive test data for various research domains

This function is particularly valuable for:
- Clinical trial reports (Table 1)
- Laboratory data analysis
- Survey research
- Quality control applications
- Biomarker studies
- Cross-sectional studies

The generated tables meet publication standards for major medical and scientific journals, making it an essential tool for clinical and pathological research.

---

*This vignette demonstrates the comprehensive capabilities of the gtsummary function in ClinicoPath. For more information about specific options and advanced features, refer to the function documentation and the gtsummary R package documentation.*
