---
title: "Hull Plot Visualization for Group Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

The `hullplot` function in ClinicoPath provides powerful cluster visualization capabilities using hull polygons to show group boundaries in scatter plots. This function is built on the ggforce package and is perfect for identifying customer segments, patient subgroups, experimental conditions, and quality control clusters.

## Key Features

- **Hull Polygons**: Create convex or concave boundaries around grouped data points
- **Multiple Variables**: Support for X/Y axes, grouping, coloring, and sizing variables
- **Customizable Appearance**: Multiple color palettes, themes, and styling options
- **Statistical Analysis**: Group statistics, outlier detection, and confidence ellipses
- **Professional Output**: Publication-ready plots for research and presentations

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  eval = FALSE
)
```


```{r setup 2, include=FALSE}

library(ClinicoPath)
library(dplyr)
library(knitr)
```

# Loading Test Datasets

ClinicoPath provides comprehensive test datasets specifically designed for hull plot visualization:

```{r load-data}
# Load all hullplot test datasets
data(hullplot_customer_data, package = "ClinicoPath")
data(hullplot_clinical_data, package = "ClinicoPath")
data(hullplot_experimental_data, package = "ClinicoPath")
data(hullplot_survey_data, package = "ClinicoPath")
data(hullplot_quality_data, package = "ClinicoPath")

# Display dataset dimensions
cat("Dataset Dimensions:\n")
cat("Customer Data:", nrow(hullplot_customer_data), "×", ncol(hullplot_customer_data), "\n")
cat("Clinical Data:", nrow(hullplot_clinical_data), "×", ncol(hullplot_clinical_data), "\n")
cat("Experimental Data:", nrow(hullplot_experimental_data), "×", ncol(hullplot_experimental_data), "\n")
cat("Survey Data:", nrow(hullplot_survey_data), "×", ncol(hullplot_survey_data), "\n")
cat("Quality Data:", nrow(hullplot_quality_data), "×", ncol(hullplot_quality_data), "\n")
```

# Basic Hull Plots

## Simple Hull Plot

Let's start with a basic hull plot showing customer segmentation:

```{r basic-hull}
# Create a basic hull plot
result1 <- ClinicoPath::hullplot(
  data = hullplot_customer_data,
  x_var = "annual_spending",
  y_var = "purchase_frequency",
  group_var = "segment"
)

# Display basic information about the plot
cat("Basic Hull Plot Created Successfully\n")
cat("Customer segments:", length(unique(hullplot_customer_data$segment)), "\n")
cat("Data points:", nrow(hullplot_customer_data), "\n")
```

## Enhanced Hull Plot with Color Mapping

Add color mapping for better visualization:

```{r enhanced-hull}
# Create an enhanced hull plot with color mapping
result2 <- ClinicoPath::hullplot(
  data = hullplot_customer_data,
  x_var = "annual_spending",
  y_var = "purchase_frequency",
  group_var = "segment",
  color_var = "preferred_category",
  size_var = "satisfaction",
  hull_alpha = 0.3,
  show_labels = TRUE,
  plot_title = "Customer Segmentation Analysis",
  x_label = "Annual Spending ($)",
  y_label = "Purchase Frequency (per year)"
)

cat("Enhanced Hull Plot Created\n")
cat("Color mapping by preferred category\n")
cat("Point size represents customer satisfaction\n")
```

# Clinical Data Visualization

## Biomarker Clustering

Hull plots are excellent for visualizing patient subgroups based on biomarker data:

```{r clinical-hull}
# Create clinical biomarker hull plot
result3 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  color_var = "treatment_response",
  show_statistics = TRUE,
  outlier_detection = TRUE,
  confidence_ellipses = TRUE,
  color_palette = "clinical",
  plot_title = "Patient Biomarker Clustering",
  x_label = "Biomarker X (ng/mL)",
  y_label = "Biomarker Y (units/L)"
)

cat("Clinical Hull Plot Created\n")
cat("Disease subtypes:", length(unique(hullplot_clinical_data$disease_subtype)), "\n")
cat("Includes statistics and outlier detection\n")
```

## Advanced Clinical Analysis

```{r advanced-clinical}
# Advanced clinical analysis with multiple features
result4 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  color_var = "hospital",
  size_var = "severity_score",
  hull_concavity = 1.5,
  hull_alpha = 0.4,
  show_labels = TRUE,
  show_statistics = TRUE,
  color_palette = "viridis",
  plot_theme = "clinical",
  plot_title = "Multi-Hospital Biomarker Analysis",
  x_label = "Primary Biomarker",
  y_label = "Secondary Biomarker"
)

cat("Advanced Clinical Analysis Created\n")
cat("Multi-dimensional visualization with hospital comparison\n")
```

# Experimental Data Analysis

## Treatment Response Visualization

Visualize experimental treatment responses with hull plots:

```{r experimental-hull}
# Create experimental response hull plot
result5 <- ClinicoPath::hullplot(
  data = hullplot_experimental_data,
  x_var = "response_x",
  y_var = "response_y",
  group_var = "condition",
  color_var = "timepoint",
  size_var = "dose_level",
  confidence_ellipses = TRUE,
  hull_alpha = 0.25,
  color_palette = "set1",
  plot_title = "Treatment Response Analysis",
  x_label = "Primary Response Measure",
  y_label = "Secondary Response Measure"
)

cat("Experimental Hull Plot Created\n")
cat("Treatment conditions:", length(unique(hullplot_experimental_data$condition)), "\n")
cat("Time points analyzed:", length(unique(hullplot_experimental_data$timepoint)), "\n")
```

## Quality Control Visualization

```{r experimental-qc}
# Quality control analysis
result6 <- ClinicoPath::hullplot(
  data = hullplot_experimental_data,
  x_var = "response_x",
  y_var = "response_y",
  group_var = "batch",
  color_var = "condition",
  outlier_detection = TRUE,
  show_statistics = TRUE,
  color_palette = "dark2",
  plot_title = "Experimental Batch Quality Control",
  x_label = "Response Variable 1",
  y_label = "Response Variable 2"
)

cat("Quality Control Analysis Created\n")
cat("Batch variation assessment with outlier detection\n")
```

# Survey Data Analysis

## Political Orientation Mapping

Hull plots can effectively visualize survey data and political orientations:

```{r survey-hull}
# Create political orientation hull plot
result7 <- ClinicoPath::hullplot(
  data = hullplot_survey_data,
  x_var = "economic_conservatism",
  y_var = "social_conservatism",
  group_var = "political_orientation",
  color_var = "region",
  size_var = "institutional_trust",
  hull_concavity = 1.8,
  show_labels = TRUE,
  color_palette = "viridis",
  plot_title = "Political Orientation by Geographic Region",
  x_label = "Economic Conservatism (0-100)",
  y_label = "Social Conservatism (0-100)"
)

cat("Political Survey Hull Plot Created\n")
cat("Political orientations:", length(unique(hullplot_survey_data$political_orientation)), "\n")
cat("Geographic regions:", length(unique(hullplot_survey_data$region)), "\n")
```

## Demographic Analysis

```{r demographic-analysis}
# Demographic clustering analysis
result8 <- ClinicoPath::hullplot(
  data = hullplot_survey_data,
  x_var = "economic_conservatism",
  y_var = "social_conservatism",
  group_var = "education",
  color_var = "income_level",
  show_statistics = TRUE,
  hull_alpha = 0.35,
  color_palette = "set2",
  plot_theme = "minimal",
  plot_title = "Political Views by Education and Income",
  x_label = "Economic Conservatism",
  y_label = "Social Conservatism"
)

cat("Demographic Analysis Created\n")
cat("Education levels analyzed with income stratification\n")
```

# Manufacturing Quality Control

## Quality Group Visualization

Hull plots are valuable for quality control and manufacturing analysis:

```{r quality-hull}
# Create quality control hull plot
result9 <- ClinicoPath::hullplot(
  data = hullplot_quality_data,
  x_var = "dimensional_accuracy",
  y_var = "surface_finish",
  group_var = "quality_group",
  color_var = "material_grade",
  show_statistics = TRUE,
  outlier_detection = TRUE,
  color_palette = "clinical",
  plot_title = "Manufacturing Quality Control Analysis",
  x_label = "Dimensional Accuracy (%)",
  y_label = "Surface Finish Quality (%)"
)

cat("Quality Control Hull Plot Created\n")
cat("Quality groups:", length(unique(hullplot_quality_data$quality_group)), "\n")
cat("Material grades:", length(unique(hullplot_quality_data$material_grade)), "\n")
```

## Production Line Analysis

```{r production-analysis}
# Production line comparison
result10 <- ClinicoPath::hullplot(
  data = hullplot_quality_data,
  x_var = "dimensional_accuracy",
  y_var = "surface_finish",
  group_var = "production_line",
  color_var = "shift",
  size_var = "defect_count",
  hull_alpha = 0.3,
  confidence_ellipses = TRUE,
  color_palette = "dark2",
  plot_title = "Production Line Performance by Shift",
  x_label = "Dimensional Accuracy",
  y_label = "Surface Finish"
)

cat("Production Line Analysis Created\n")
cat("Production shift comparison with defect tracking\n")
```

# Customization Options

## Hull Appearance Customization

```{r hull-customization}
# Demonstrate hull appearance options
result11 <- ClinicoPath::hullplot(
  data = hullplot_customer_data,
  x_var = "annual_spending",
  y_var = "purchase_frequency",
  group_var = "segment",
  hull_concavity = 0.5,    # More concave hulls
  hull_alpha = 0.6,        # More opaque
  hull_expand = 0.15,      # Larger hulls
  point_size = 3,          # Larger points
  point_alpha = 0.8,       # More opaque points
  show_labels = TRUE,
  plot_title = "Customized Hull Appearance"
)

cat("Hull Customization Demonstrated\n")
cat("Features: concave hulls, custom transparency, expanded boundaries\n")
```

## Color Palette Comparison

```{r color-palettes}
# Demonstrate different color palettes
palettes_to_test <- c("viridis", "clinical", "set1", "set2", "dark2")

for (palette in palettes_to_test[1:3]) {  # Test first 3 for brevity
  result <- ClinicoPath::hullplot(
    data = hullplot_clinical_data,
    x_var = "biomarker_x",
    y_var = "biomarker_y",
    group_var = "disease_subtype",
    color_palette = palette,
    plot_title = paste("Color Palette:", palette)
  )
  cat("Tested palette:", palette, "\n")
}

cat("Color palette comparison completed\n")
```

## Theme Variations

```{r theme-variations}
# Demonstrate different plot themes
themes_to_test <- c("minimal", "classic", "clinical")

for (theme in themes_to_test) {
  result <- ClinicoPath::hullplot(
    data = hullplot_survey_data,
    x_var = "economic_conservatism",
    y_var = "social_conservatism",
    group_var = "political_orientation",
    plot_theme = theme,
    plot_title = paste("Theme:", theme)
  )
  cat("Applied theme:", theme, "\n")
}

cat("Theme variations demonstrated\n")
```

# Statistical Features

## Group Statistics Analysis

```{r group-statistics}
# Comprehensive group statistics
result12 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  show_statistics = TRUE,
  plot_title = "Biomarker Analysis with Group Statistics"
)

cat("Group Statistics Analysis Created\n")
cat("Includes mean, SD, and sample sizes for each group\n")
```

## Outlier Detection

```{r outlier-detection}
# Outlier detection analysis
result13 <- ClinicoPath::hullplot(
  data = hullplot_quality_data,
  x_var = "dimensional_accuracy",
  y_var = "surface_finish",
  group_var = "quality_group",
  outlier_detection = TRUE,
  plot_title = "Quality Control with Outlier Detection"
)

cat("Outlier Detection Analysis Created\n")
cat("Uses IQR method to identify potential outliers within groups\n")
```

## Confidence Ellipses

```{r confidence-ellipses}
# Confidence ellipses analysis
result14 <- ClinicoPath::hullplot(
  data = hullplot_experimental_data,
  x_var = "response_x",
  y_var = "response_y",
  group_var = "condition",
  confidence_ellipses = TRUE,
  hull_alpha = 0.2,  # Lower alpha to see ellipses better
  plot_title = "Treatment Effects with Confidence Ellipses"
)

cat("Confidence Ellipses Analysis Created\n")
cat("95% confidence ellipses show statistical boundaries\n")
```

# Advanced Applications

## Multi-Variable Analysis

```{r multi-variable}
# Complex multi-variable analysis
result15 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  color_var = "treatment_response",
  size_var = "severity_score",
  show_statistics = TRUE,
  outlier_detection = TRUE,
  confidence_ellipses = TRUE,
  hull_concavity = 1.8,
  hull_alpha = 0.3,
  color_palette = "viridis",
  plot_theme = "clinical",
  plot_title = "Comprehensive Biomarker Analysis",
  x_label = "Primary Biomarker (ng/mL)",
  y_label = "Secondary Biomarker (units/L)"
)

cat("Multi-Variable Analysis Created\n")
cat("Integrates grouping, coloring, sizing, statistics, and outliers\n")
```

## Time Series Clustering

```{r time-series}
# Time-based experimental analysis
result16 <- ClinicoPath::hullplot(
  data = hullplot_experimental_data,
  x_var = "response_x",
  y_var = "response_y",
  group_var = "timepoint",
  color_var = "condition",
  size_var = "dose_level",
  hull_alpha = 0.25,
  show_labels = TRUE,
  color_palette = "set1",
  plot_title = "Treatment Response Over Time",
  x_label = "Primary Response",
  y_label = "Secondary Response"
)

cat("Time Series Clustering Created\n")
cat("Shows treatment evolution across timepoints\n")
```

# Real-World Applications

## Customer Segmentation for Marketing

```{r marketing-application}
# Professional customer segmentation
result17 <- ClinicoPath::hullplot(
  data = hullplot_customer_data,
  x_var = "annual_spending",
  y_var = "purchase_frequency",
  group_var = "segment",
  color_var = "loyalty_member",
  size_var = "satisfaction",
  show_statistics = TRUE,
  hull_alpha = 0.3,
  color_palette = "clinical",
  plot_theme = "minimal",
  plot_title = "Customer Segmentation Strategy",
  x_label = "Annual Spending ($USD)",
  y_label = "Purchase Frequency (per year)"
)

cat("Marketing Application Created\n")
cat("Professional customer segmentation for business strategy\n")
```

## Clinical Research Publication

```{r clinical-publication}
# Publication-ready clinical plot
result18 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  color_var = "treatment_response",
  show_statistics = TRUE,
  confidence_ellipses = TRUE,
  hull_alpha = 0.25,
  color_palette = "clinical",
  plot_theme = "clinical",
  plot_title = "Biomarker Expression in Disease Subtypes",
  x_label = "Biomarker A Expression (ng/mL)",
  y_label = "Biomarker B Expression (units/L)"
)

cat("Clinical Publication Plot Created\n")
cat("Publication-ready format for medical journals\n")
```

## Quality Assurance Dashboard

```{r qa-dashboard}
# Quality assurance visualization
result19 <- ClinicoPath::hullplot(
  data = hullplot_quality_data,
  x_var = "dimensional_accuracy",
  y_var = "surface_finish",
  group_var = "quality_group",
  color_var = "production_line",
  outlier_detection = TRUE,
  show_statistics = TRUE,
  hull_alpha = 0.4,
  color_palette = "dark2",
  plot_title = "Production Quality Dashboard",
  x_label = "Dimensional Accuracy (%)",
  y_label = "Surface Finish Quality (%)"
)

cat("Quality Assurance Dashboard Created\n")
cat("Real-time quality monitoring visualization\n")
```

# Best Practices and Tips

## Variable Selection Guidelines

1. **X and Y Variables**: Choose continuous variables that represent meaningful dimensions
2. **Grouping Variable**: Use categorical variables with 2-8 groups for best visualization
3. **Color Variable**: Optional, but useful for additional categorical information
4. **Size Variable**: Best with continuous variables representing importance or magnitude

## Hull Customization

1. **Concavity**: Lower values (0-1) create more detailed, concave hulls; higher values (1-2) create simpler shapes
2. **Alpha**: 0.2-0.4 typically works well for overlapping groups
3. **Expansion**: 0.05-0.1 provides good boundary padding without excessive whitespace

## Color and Theme Selection

1. **Clinical Data**: Use "clinical" or "viridis" palettes for professional appearance
2. **Business Data**: "set1" or "set2" palettes work well for presentations
3. **Publications**: "clinical" theme with "viridis" palette recommended

```{r best-practices}
# Demonstrate best practices
result20 <- ClinicoPath::hullplot(
  data = hullplot_clinical_data,
  x_var = "biomarker_x",
  y_var = "biomarker_y",
  group_var = "disease_subtype",
  color_var = "treatment_response",
  size_var = "severity_score",
  hull_concavity = 1.5,      # Balanced shape
  hull_alpha = 0.3,          # Good transparency
  hull_expand = 0.08,        # Appropriate padding
  show_labels = TRUE,
  show_statistics = TRUE,
  confidence_ellipses = TRUE,
  color_palette = "viridis",  # Professional colors
  plot_theme = "clinical",    # Clean theme
  point_alpha = 0.7,         # Visible but not overwhelming
  plot_title = "Best Practices Example: Clinical Biomarker Analysis",
  x_label = "Primary Biomarker Expression",
  y_label = "Secondary Biomarker Expression"
)

cat("Best Practices Example Created\n")
cat("Professional formatting with comprehensive features\n")
```

# Summary

The `hullplot` function in ClinicoPath provides powerful visualization capabilities for group analysis and cluster identification. Key advantages include:

1. **Visual Clarity**: Hull polygons clearly show group boundaries and overlaps
2. **Multi-Dimensional**: Support for grouping, coloring, and sizing variables
3. **Statistical Features**: Group statistics, outlier detection, and confidence intervals
4. **Customization**: Extensive options for colors, themes, and appearance
5. **Professional Output**: Publication-ready plots for research and business
6. **Versatile Applications**: Customer segmentation, clinical research, quality control, survey analysis

This function is particularly valuable for:
- Customer segmentation and market research
- Clinical patient subgroup identification
- Experimental condition comparison
- Quality control and manufacturing analysis
- Survey data visualization
- Scientific research publications

The hull plot visualization method is excellent for presentations and publications as it clearly communicates group structures and relationships in complex datasets.

---

*This vignette demonstrates the comprehensive capabilities of the hullplot function in ClinicoPath. Hull plots provide an intuitive and powerful way to visualize group structures in multidimensional data, making them essential tools for data analysis and presentation.*
