---
title: "Using jamovi for the Common Statistical Analysis in Pathology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using jamovi for the Common Statistical Analysis in Pathology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(ClinicoPath)
```



In this page I am going to collect the most common statistical methods used in pathology research and how can one do them in jamovi.


- Recommendations for reporting histopathology studies: a proposal

Virchows Arch (2015) 466:611–615 DOI 10.1007/s00428-015-1762-3



- Statistical Literacy Among Academic Pathologists: A Survey Study to Gauge Knowledge of Frequently Used Statistical Tests Among Trainees and Faculty.

Arch Pathol Lab Med. 2017 Feb;141(2):279-287. doi: 10.5858/arpa.2016-0200-OA

https://www.ncbi.nlm.nih.gov/pubmed/27959584

https://www.archivesofpathology.org/doi/pdf/10.5858/arpa.2016-0200-OA

- Pathologists should probably forget about kappa. Percent agreement, diagnostic specificity and related metrics provide more clinically applicable measures of interobserver variability

Annals of Diagnostic Pathology, Volume 47, August 2020

https://www.sciencedirect.com/science/article/abs/pii/S1092913420301040





# Student t test


# Regression/ANOVA


# Chi-square test


# Mann-Whitney test (rank sum)


# Fisher exact test


# Survival analysis


## Kaplan-Meier/log-rank


## Cox regression


# Multiple comparison adjustment

## Tukey

## Bonferroni

## Newman-Keuls


# Kappa Statistic


# ROC analysis


# Logistic regression


# Spearman rank correlation


# Kruskal-Wallis test

<iframe width="560" height="315" src="https://www.youtube.com/embed/qfs6T311_88" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


# Pearson correlation statistic


# Normality test


# McNemar test




