---
title: "ClinicoPath Descriptives 04: Data Quality Assessment for Clinical Research"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ClinicoPath Descriptives 04: Data Quality Assessment for Clinical Research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Data quality assessment is fundamental to reliable clinical research. This vignette demonstrates comprehensive data exploration and quality assessment techniques using ClinicoPathDescriptives, including missing data analysis, distribution assessment, outlier detection, and fraud detection using Benford's law.

## Data Quality Framework

### Quality Dimensions

1. **Completeness**: Presence of missing data
2. **Consistency**: Logical relationships between variables
3. **Accuracy**: Conformance to expected patterns
4. **Validity**: Adherence to defined formats and ranges
5. **Integrity**: Referential and business rule compliance

```{r setup}
library(ClinicoPath)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# Load the dataset
data(histopathology)

# Initial data overview
cat("Dataset Dimensions:", dim(histopathology), "\n")
cat("Variables:", ncol(histopathology), "\n")
cat("Observations:", nrow(histopathology), "\n\n")

# Variable types overview
str(histopathology, give.attr = FALSE)
```

## Phase 1: Completeness Assessment

### Missing Data Analysis

```{r}
# Calculate missing data percentages
missing_summary <- histopathology %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(
    Missing_Percentage = round(Missing_Count / nrow(histopathology) * 100, 1),
    Data_Quality = case_when(
      Missing_Percentage == 0 ~ "Complete",
      Missing_Percentage <= 5 ~ "Excellent", 
      Missing_Percentage <= 10 ~ "Good",
      Missing_Percentage <= 20 ~ "Acceptable",
      TRUE ~ "Poor"
    )
  ) %>%
  arrange(desc(Missing_Percentage))

# Display missing data summary
kable(missing_summary %>% filter(Missing_Count > 0),
      caption = "Missing Data Summary",
      col.names = c("Variable", "Missing Count", "Missing %", "Quality Rating"))
```

### Completeness by Variable Type

```{r}
# Analyze completeness by variable groups
demographic_vars <- c("Age", "Sex", "Race")
pathology_vars <- c("Grade", "TStage", "LVI", "PNI", "LymphNodeMetastasis")
biomarker_vars <- c("MeasurementA", "MeasurementB", "Anti-X-intensity", "Anti-Y-intensity")
outcome_vars <- c("Outcome", "Death", "OverallTime", "Mortality5yr")

# Function to calculate completeness for variable groups
assess_completeness <- function(data, vars, group_name) {
  if(all(vars %in% names(data))) {
    completeness <- data %>%
      select(all_of(vars)) %>%
      summarise_all(~sum(!is.na(.))) %>%
      pivot_longer(everything(), names_to = "Variable", values_to = "Complete_Count") %>%
      mutate(
        Group = group_name,
        Completeness_Rate = round(Complete_Count / nrow(data) * 100, 1)
      )
    return(completeness)
  }
  return(NULL)
}

# Assess each variable group
completeness_by_group <- bind_rows(
  assess_completeness(histopathology, demographic_vars, "Demographics"),
  assess_completeness(histopathology, pathology_vars, "Pathology"),
  assess_completeness(histopathology, biomarker_vars, "Biomarkers"),
  assess_completeness(histopathology, outcome_vars, "Outcomes")
) %>%
  filter(!is.null(.))

# Summary by group
group_summary <- completeness_by_group %>%
  group_by(Group) %>%
  summarise(
    Variables = n(),
    Mean_Completeness = round(mean(Completeness_Rate), 1),
    Min_Completeness = round(min(Completeness_Rate), 1),
    .groups = 'drop'
  )

kable(group_summary,
      caption = "Data Completeness by Variable Group",
      col.names = c("Variable Group", "Variables", "Mean Completeness (%)", "Min Completeness (%)"))
```

## Phase 2: Distribution Assessment

### Continuous Variables Analysis

```{r}
# Analyze distributions of continuous variables
continuous_vars <- c("Age", "OverallTime", "MeasurementA", "MeasurementB")

# Check which variables exist in the dataset
available_vars <- continuous_vars[continuous_vars %in% names(histopathology)]
if(length(available_vars) > 0) {
  tryCatch({
    summarydata(
      data = histopathology,
      vars = available_vars,
      date_vars = NULL,
      grvar = NULL
    )
  }, error = function(e) {
    cat("Error with summarydata:", conditionMessage(e), "\n")
    cat("Available continuous variables:", paste(available_vars, collapse = ", "), "\n")
    cat("Variable types:\n")
    print(sapply(histopathology[available_vars], class))
  })
} else {
  cat("No continuous variables found in dataset\n")
}
```

### Categorical Variables Analysis

```{r}
# Analyze categorical variable distributions
categorical_vars <- c("Sex", "Grade", "TStage", "Group", "LVI", "PNI")

# Check which categorical variables exist in the dataset
available_cat_vars <- categorical_vars[categorical_vars %in% names(histopathology)]
if(length(available_cat_vars) > 0) {
  tryCatch({
    reportcat(
      data = histopathology,
      vars = available_cat_vars
    )
  }, error = function(e) {
    cat("Error with reportcat:", conditionMessage(e), "\n")
    cat("Available categorical variables:", paste(available_cat_vars, collapse = ", "), "\n")
  })
} else {
  cat("No categorical variables found in dataset\n")
}
```

### Variable Relationship Overview

```{r}
# Hierarchical overview of key variables
vartree_vars <- c("Group", "Grade", "TStage", "LVI", "PNI", "LymphNodeMetastasis", "Outcome")
available_vartree_vars <- vartree_vars[vartree_vars %in% names(histopathology)]

if(length(available_vartree_vars) > 0) {
  tryCatch({
    vartree(
      data = histopathology,
      vars = available_vartree_vars,
      percvar = NULL,
      percvarLevel = NULL,
      summaryvar = NULL,
      prunebelow = NULL,
      pruneLevel1 = NULL,
      pruneLevel2 = NULL,
      follow = NULL,
      followLevel1 = NULL,
      followLevel2 = NULL,
      excl = FALSE,
      vp = TRUE,
      horizontal = FALSE,
      sline = TRUE,
      varnames = FALSE,
      nodelabel = TRUE,
      pct = FALSE
    )
  }, error = function(e) {
    cat("Error with vartree:", conditionMessage(e), "\n")
    cat("Available variables for tree:", paste(available_vartree_vars, collapse = ", "), "\n")
  })
} else {
  cat("No variables found for variable tree\n")
}
```

## Phase 3: Consistency and Logic Checks

### Temporal Consistency

```{r}
# Check date consistency (if dates are properly formatted)
date_vars <- histopathology %>%
  select(contains("Date")) %>%
  names()

if(length(date_vars) > 0) {
  cat("Date variables found:", paste(date_vars, collapse = ", "), "\n")
  
  # Basic date analysis
  tryCatch({
    reportcat(
      data = histopathology,
      vars = date_vars[1:min(2, length(date_vars))]
    )
  }, error = function(e) {
    cat("Error with date variables reportcat:", conditionMessage(e), "\n")
    cat("Date variables:", paste(date_vars, collapse = ", "), "\n")
  })
}
```

### Logical Relationship Checks

```{r}
# Cross-tabulation to check logical relationships
crosstable(
  data = histopathology,
  vars = "Death",
  group = "Mortality5yr"
)
```

```{r}
# Outcome consistency check
crosstable(
  data = histopathology,
  vars = "Outcome",
  group = "Death"
)
```

### Grade-Stage Relationship Assessment

```{r}
# Check pathological grade vs stage relationships
crosstable(
  data = histopathology,
  vars = "Grade",
  group = "TStage"
)
```

## Phase 4: Benford's Law Analysis

Benford's law states that in naturally occurring datasets, the leading digit 1 appears about 30.1% of the time, digit 2 about 17.6%, and so on. Deviation from this pattern may indicate data fabrication.

### Measurement A Analysis

```{r}
# Benford's law analysis for Measurement A
benford(
  data = histopathology,
  var = "MeasurementA"
)
```

### Measurement B Analysis

```{r}
# Benford's law analysis for Measurement B
benford(
  data = histopathology, 
  var = "MeasurementB"
)
```

### Overall Time Analysis

```{r}
# Benford's law analysis for follow-up time
benford(
  data = histopathology,
  var = "OverallTime"
)
```

## Phase 5: Inter-rater Reliability Assessment

### Rater Agreement Analysis

```{r}
# Analyze inter-rater reliability data
rater_vars <- histopathology %>%
  select(starts_with("Rater")) %>%
  names()

if(length(rater_vars) > 0) {
  # Binary rater variables
  binary_raters <- rater_vars[grepl("[0-9]$", rater_vars)]
  if(length(binary_raters) > 0) {
    tryCatch({
      reportcat(
        data = histopathology,
        vars = binary_raters[1:min(3, length(binary_raters))]
      )
    }, error = function(e) {
      cat("Error with binary raters reportcat:", conditionMessage(e), "\n")
      cat("Binary rater variables:", paste(binary_raters, collapse = ", "), "\n")
    })
  } else {
    cat("No binary rater variables found (ending with digits)\n")
  }
  
  # Categorical rater variables  
  categorical_raters <- rater_vars[grepl("[A-Z]$", rater_vars)]
  if(length(categorical_raters) > 0) {
    tryCatch({
      reportcat(
        data = histopathology,
        vars = categorical_raters
      )
    }, error = function(e) {
      cat("Error with categorical raters reportcat:", conditionMessage(e), "\n")
      cat("Categorical rater variables:", paste(categorical_raters, collapse = ", "), "\n")
    })
  } else {
    cat("No categorical rater variables found (ending with letters)\n")
  }
} else {
  cat("No rater variables found in dataset\n")
}
```

### Agreement Cross-tabulation

```{r}
# Cross-tabulate raters for agreement assessment
if(length(rater_vars) >= 2) {
  crosstable(
    data = histopathology,
    vars = rater_vars[1],
    group = rater_vars[2]
  )
}
```

## Phase 6: Outlier Detection

### Statistical Outlier Analysis

```{r}
# Identify potential outliers in continuous variables
outlier_analysis <- histopathology %>%
  select(Age, OverallTime, MeasurementA, MeasurementB) %>%
  summarise_all(
    list(
      Mean = ~round(mean(., na.rm = TRUE), 2),
      SD = ~round(sd(., na.rm = TRUE), 2),
      Q1 = ~round(quantile(., 0.25, na.rm = TRUE), 2),
      Q3 = ~round(quantile(., 0.75, na.rm = TRUE), 2),
      IQR = ~round(IQR(., na.rm = TRUE), 2),
      Lower_Fence = ~round(quantile(., 0.25, na.rm = TRUE) - 1.5 * IQR(., na.rm = TRUE), 2),
      Upper_Fence = ~round(quantile(., 0.75, na.rm = TRUE) + 1.5 * IQR(., na.rm = TRUE), 2)
    )
  ) %>%
  pivot_longer(everything(), names_to = "Stat", values_to = "Value") %>%
  separate(Stat, into = c("Variable", "Statistic"), sep = "_", extra = "merge") %>%
  pivot_wider(names_from = Statistic, values_from = Value)

kable(outlier_analysis,
      caption = "Outlier Detection Statistics",
      digits = 2)
```

### Extreme Value Assessment

```{r}
# Count extreme values beyond fences
extreme_values <- histopathology %>%
  summarise(
    Age_Low_Outliers = sum(Age < (quantile(Age, 0.25, na.rm = TRUE) - 1.5 * IQR(Age, na.rm = TRUE)), na.rm = TRUE),
    Age_High_Outliers = sum(Age > (quantile(Age, 0.75, na.rm = TRUE) + 1.5 * IQR(Age, na.rm = TRUE)), na.rm = TRUE),
    Time_Low_Outliers = sum(OverallTime < (quantile(OverallTime, 0.25, na.rm = TRUE) - 1.5 * IQR(OverallTime, na.rm = TRUE)), na.rm = TRUE),
    Time_High_Outliers = sum(OverallTime > (quantile(OverallTime, 0.75, na.rm = TRUE) + 1.5 * IQR(OverallTime, na.rm = TRUE)), na.rm = TRUE),
    MeasA_Low_Outliers = sum(MeasurementA < (quantile(MeasurementA, 0.25, na.rm = TRUE) - 1.5 * IQR(MeasurementA, na.rm = TRUE)), na.rm = TRUE),
    MeasA_High_Outliers = sum(MeasurementA > (quantile(MeasurementA, 0.75, na.rm = TRUE) + 1.5 * IQR(MeasurementA, na.rm = TRUE)), na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Variable_Type", values_to = "Outlier_Count") %>%
  separate(Variable_Type, into = c("Variable", "Outlier_Type"), sep = "_", extra = "merge")

kable(extreme_values,
      caption = "Extreme Value Counts by Variable")
```

## Phase 7: Diagnostic Test Validation

### Test Performance Analysis

```{r}
# Analyze diagnostic test performance if available
test_vars <- histopathology %>%
  select(contains("Test"), contains("Standard")) %>%
  names()

if(length(test_vars) >= 2) {
  # Cross-tabulate new test vs gold standard
  crosstable(
    data = histopathology,
    vars = test_vars[1],
    group = test_vars[2]
  )
}
```

### Validity Assessment

```{r}
# Analyze data validity flags
validity_vars <- histopathology %>%
  select(contains("Valid")) %>%
  names()

if(length(validity_vars) > 0) {
  tryCatch({
    reportcat(
      data = histopathology,
      vars = validity_vars[1]
    )
  }, error = function(e) {
    cat("Error with validity variables reportcat:", conditionMessage(e), "\n")
    cat("Validity variables:", paste(validity_vars, collapse = ", "), "\n")
  })
} else {
  cat("No validity variables found in dataset\n")
}
```

## Phase 8: Data Quality Dashboard

### Overall Quality Metrics

```{r}
# Create comprehensive quality summary
quality_metrics <- data.frame(
  Metric = c(
    "Total Observations",
    "Total Variables", 
    "Complete Cases",
    "Variables with No Missing Data",
    "Variables with <5% Missing",
    "Variables with >20% Missing",
    "Mean Completeness Rate"
  ),
  Value = c(
    nrow(histopathology),
    ncol(histopathology),
    sum(complete.cases(histopathology)),
    sum(missing_summary$Missing_Count == 0),
    sum(missing_summary$Missing_Percentage <= 5),
    sum(missing_summary$Missing_Percentage > 20),
    paste0(round(mean(100 - missing_summary$Missing_Percentage), 1), "%")
  )
)

kable(quality_metrics,
      caption = "Data Quality Dashboard")
```

### Quality Recommendations

```{r}
# Generate quality recommendations
high_missing <- missing_summary %>%
  filter(Missing_Percentage > 20) %>%
  pull(Variable)

moderate_missing <- missing_summary %>%
  filter(Missing_Percentage > 5 & Missing_Percentage <= 20) %>%
  pull(Variable)

recommendations <- data.frame(
  Priority = c("High", "Medium", "Low"),
  Action = c(
    ifelse(length(high_missing) > 0, 
           paste("Investigate high missing data in:", paste(high_missing, collapse = ", ")),
           "No high-priority missing data issues"),
    ifelse(length(moderate_missing) > 0,
           paste("Consider imputation for:", paste(moderate_missing, collapse = ", ")),
           "No moderate missing data issues"), 
    "Continue routine quality monitoring"
  )
)

kable(recommendations,
      caption = "Data Quality Recommendations")
```

## Phase 9: Reporting Quality Issues

### Missing Data Patterns

```{r}
# Analyze missing data patterns
if(any(missing_summary$Missing_Count > 0)) {
  # Create missing data profile
  missing_profile <- histopathology %>%
    summarise_all(~ifelse(is.na(.), "Missing", "Present")) %>%
    pivot_longer(everything(), names_to = "Variable", values_to = "Status") %>%
    count(Variable, Status) %>%
    pivot_wider(names_from = Status, values_from = n, values_fill = 0) %>%
    mutate(
      Total = Missing + Present,
      Missing_Rate = round(Missing / Total * 100, 1)
    ) %>%
    filter(Missing > 0) %>%
    arrange(desc(Missing_Rate))
  
  kable(missing_profile,
        caption = "Missing Data Profile")
}
```

### Benford's Law Compliance Summary

```{r}
# Summary of Benford's law results
benford_summary <- data.frame(
  Variable = c("MeasurementA", "MeasurementB", "OverallTime"),
  Expected_Pattern = "Natural logarithmic distribution",
  Assessment = c(
    "Evaluate first digit distribution",
    "Evaluate first digit distribution", 
    "Evaluate first digit distribution"
  ),
  Recommendation = "Review any significant deviations from expected pattern"
)

kable(benford_summary,
      caption = "Benford's Law Assessment Summary")
```

## Implementation in jamovi

Data quality assessment functions are available in jamovi:

1. **Benford's Law**: `Exploration > ClinicoPath Descriptives > Benford Analysis`
2. **Variable Summary**: `Exploration > ClinicoPath Descriptives > Summary of Continuous/Categorical Variables`
3. **Cross-tabulations**: `Exploration > ClinicoPath Comparisons > Cross Tables`
4. **Variable Tree**: `Exploration > ClinicoPath Descriptive Plots > Variable Tree`

## Conclusions

This comprehensive data quality assessment provides:

1. **Completeness Evaluation**: Systematic missing data analysis
2. **Consistency Checks**: Logical relationship validation
3. **Accuracy Assessment**: Statistical distribution analysis and outlier detection
4. **Fraud Detection**: Benford's law analysis for data integrity
5. **Reliability Testing**: Inter-rater agreement evaluation

Regular data quality assessment using these techniques ensures reliable and trustworthy clinical research results.

## Best Practices

- Perform quality assessment before any statistical analysis
- Document and investigate any quality issues found
- Establish quality thresholds appropriate for your research context
- Implement ongoing quality monitoring throughout data collection
- Use multiple quality assessment methods for comprehensive evaluation
