---
title: "Time-Dependent Covariates in Multisurvival Analysis"
author: "ClinicoPath Team"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    highlight-style: github
    code-fold: false
---

# Introduction {#introduction}

Time-dependent covariates are variables that change their values over the course of follow-up time. In clinical research, many important variables can change during the study period, such as:

- Treatment regimens (switching from standard to experimental therapy)
- Disease status (remission, progression)
- Performance status
- Biomarker levels
- Comorbidities

The `multisurvival` function in ClinicoPath now supports time-dependent covariates in both wide and long data formats, making it easier to analyze survival data where covariates change over time.

# Data Format Options {#data-formats}

## Wide Format Data {#wide-format}

In wide format, each subject has one row, with time-dependent variables represented as separate columns for different time points.

### Example Wide Format Structure

```{r wide-example}
# Example wide format data structure
wide_data <- data.frame(
  id = 1:3,
  age = c(65, 72, 58),
  sex = factor(c("Male", "Female", "Male")),
  time = c(24, 18, 30),
  status = c(1, 0, 1),
  
  # Baseline treatment (time 0)
  treatment_baseline = factor(c("Standard", "Standard", "Experimental")),
  
  # Treatment at 6 months
  treatment_t6 = factor(c("Standard", "Experimental", "Experimental")),
  
  # Treatment at 12 months  
  treatment_t12 = factor(c("Experimental", "Experimental", "Stopped")),
  
  # Treatment at 18 months
  treatment_t18 = factor(c("Experimental", "Stopped", "Stopped"))
)

print(wide_data)
```

**Wide Format Advantages:**
- One row per patient (easier to understand)
- Common format in clinical databases
- Easy to create from existing data

**Wide Format Requirements:**
- Baseline variables: `variable_baseline` or just `variable`
- Time-specific variables: `variable_t6`, `variable_t12`, etc.
- Change times specified in options

## Long Format Data {#long-format}

In long format, each subject can have multiple rows, with one row for each time interval where covariates remain constant.

### Example Long Format Structure

```{r long-example}
# Example long format data structure  
long_data <- data.frame(
  id = c(1, 1, 1, 2, 2, 3, 3),
  age = c(65, 65, 65, 72, 72, 58, 58),
  sex = factor(c("Male", "Male", "Male", "Female", "Female", "Male", "Male")),
  tstart = c(0, 6, 12, 0, 6, 0, 12),
  tstop = c(6, 12, 24, 6, 18, 12, 30),
  status = c(0, 0, 1, 0, 0, 0, 1),
  treatment = factor(c("Standard", "Standard", "Experimental", 
                      "Standard", "Experimental", "Experimental", "Stopped"))
)

print(long_data)
```

**Long Format Advantages:**
- More flexible for complex time patterns
- Standard format for survival analysis packages
- Can handle irregular time intervals

**Long Format Requirements:**
- `tstart` and `tstop` variables for time intervals
- Status = 1 only in final interval if event occurs
- Time-dependent variables change between rows

# Using Time-Dependent Covariates {#usage}

## Basic Wide Format Analysis {#basic-wide}

```{r eval=FALSE}
library(ClinicoPath)

# Load example wide format data
load("test_wide_time_dependent.rda")

result <- multisurvival(
  data = test_wide_time_dependent,
  elapsedtime = "time",
  outcome = "status", 
  outcomeLevel = "1",
  explanatory = c("sex", "stage"),
  contexpl = "age",
  
  # Time-dependent covariate options
  use_time_dependent = TRUE,
  td_format = "wide",
  time_dep_vars = c("treatment_baseline"),  # Base variable name
  change_times = "6, 12, 18",               # When changes occur
  td_suffix_pattern = "_t{time}",           # Naming pattern
  
  timetypeoutput = "months"
)
```

## Basic Long Format Analysis {#basic-long}

```{r eval=FALSE}
# Load example long format data
load("test_long_time_dependent.rda")

result <- multisurvival(
  data = test_long_time_dependent,
  
  # No elapsedtime needed - using tstart/tstop
  outcome = "status",
  outcomeLevel = "1", 
  explanatory = c("sex", "stage"),
  contexpl = "age",
  
  # Time-dependent covariate options
  use_time_dependent = TRUE,
  td_format = "long", 
  start_time_var = "tstart",
  stop_time_var = "tstop",
  time_dep_vars = c("treatment"),  # Time-varying variable
  
  timetypeoutput = "months"
)
```

# Advanced Examples {#advanced-examples}

## Multiple Time-Dependent Variables {#multiple-td}

```{r eval=FALSE}
# Wide format with multiple time-dependent variables
result <- multisurvival(
  data = clinical_data,
  elapsedtime = "followup_months",
  outcome = "event_status",
  outcomeLevel = "Event",
  explanatory = c("sex", "initial_stage"),
  contexpl = "baseline_age",
  
  # Multiple time-dependent variables
  use_time_dependent = TRUE,
  td_format = "wide",
  time_dep_vars = c("treatment_baseline", "ps_baseline"),  # Multiple variables
  change_times = "3, 6, 12, 18",           # More frequent assessments
  td_suffix_pattern = "_month{time}",       # Different naming pattern
  
  # Additional analysis options
  calculateRiskScore = TRUE,
  numRiskGroups = "three",
  hr = TRUE
)
```

## Combining Time-Dependent Covariates with Other Features {#combined}

```{r eval=FALSE}
# Complex analysis with multiple advanced features
result <- multisurvival(
  data = oncology_data,
  elapsedtime = "survival_months", 
  outcome = "vital_status",
  outcomeLevel = "Dead",
  explanatory = c("histology", "grade"),
  contexpl = c("age_at_diagnosis", "tumor_size"),
  
  # Time-dependent covariates
  use_time_dependent = TRUE,
  td_format = "wide",
  time_dep_vars = c("treatment_baseline", "response_baseline"),
  change_times = "6, 12, 24",
  td_suffix_pattern = "_m{time}",
  
  # Frailty model for hospital clustering
  use_frailty = TRUE,
  frailty_var = "treatment_center",
  frailty_distribution = "gamma",
  
  # Splines for non-proportional hazards
  use_splines = TRUE,
  spline_vars = "age_at_diagnosis",
  spline_df = 3,
  spline_type = "pspline",
  
  # Model selection
  use_modelSelection = TRUE,
  modelSelection = "backward",
  selectionCriteria = "aic",
  
  # Risk stratification
  calculateRiskScore = TRUE,
  numRiskGroups = "four",
  plotRiskGroups = TRUE
)
```

# Data Preparation Guidelines {#data-prep}

## Converting Long to Wide Format {#long-to-wide}

```{r eval=FALSE}
# Example conversion from long to wide format
library(dplyr)
library(tidyr)

# Starting with long format data
long_format <- data.frame(
  patient_id = c(rep(1, 3), rep(2, 2), rep(3, 4)),
  visit_month = c(0, 6, 12, 0, 6, 0, 6, 12, 18),
  treatment = c("A", "A", "B", "A", "B", "A", "A", "A", "B"),
  performance_status = c(0, 1, 1, 0, 0, 0, 1, 2, 2)
)

# Convert to wide format
wide_format <- long_format %>%
  pivot_wider(
    id_cols = patient_id,
    names_from = visit_month,
    values_from = c(treatment, performance_status),
    names_sep = "_t"
  ) %>%
  # Rename baseline variables
  rename(
    treatment_baseline = treatment_t0,
    ps_baseline = performance_status_t0
  )
```

## Converting Wide to Long Format {#wide-to-long}

```{r eval=FALSE}
# Example conversion from wide to long format
wide_data <- data.frame(
  id = 1:100,
  age = rnorm(100, 65, 10),
  sex = factor(sample(c("M", "F"), 100, replace = TRUE)),
  survival_time = rexp(100, 0.01),
  status = rbinom(100, 1, 0.3),
  treatment_baseline = factor(sample(c("A", "B"), 100, replace = TRUE)),
  treatment_t6 = factor(sample(c("A", "B", "C"), 100, replace = TRUE)),
  treatment_t12 = factor(sample(c("A", "B", "C", "Stop"), 100, replace = TRUE))
)

# This conversion would typically be done by the multisurvival function
# when td_format = "wide" is specified
```

# Clinical Applications {#clinical-applications}

## Cancer Treatment Switching {#cancer-treatment}

Time-dependent covariates are particularly useful in oncology:

```{r eval=FALSE}
# Analyze impact of treatment switching in cancer patients
cancer_analysis <- multisurvival(
  data = cancer_cohort,
  elapsedtime = "overall_survival_months",
  outcome = "death_status", 
  outcomeLevel = "Dead",
  explanatory = c("cancer_type", "stage_at_diagnosis"),
  contexpl = c("age_at_diagnosis", "baseline_performance_status"),
  
  # Treatment changes over time
  use_time_dependent = TRUE,
  td_format = "wide",
  time_dep_vars = c("primary_treatment_baseline"),
  change_times = "3, 6, 12, 18, 24",  # Quarterly assessments
  td_suffix_pattern = "_quarter{time}",
  
  # Account for treatment center effects
  use_frailty = TRUE,
  frailty_var = "treatment_center",
  
  # Test proportional hazards assumption
  ph_cox = TRUE,
  
  # Generate comprehensive outputs
  hr = TRUE,
  calculateRiskScore = TRUE,
  person_time = TRUE
)
```

## Cardiovascular Risk Factors {#cardiovascular}

```{r eval=FALSE}
# Analyze changing cardiovascular risk factors
cv_analysis <- multisurvival(
  data = cardiovascular_study,
  tint = TRUE,
  dxdate = "enrollment_date",
  fudate = "last_contact_date",
  timetypedata = "ymd",
  timetypeoutput = "years",
  outcome = "cv_event",
  outcomeLevel = "Yes",
  explanatory = c("gender", "diabetes_baseline"),
  contexpl = c("baseline_age", "baseline_bmi"),
  
  # Time-varying medication and lab values
  use_time_dependent = TRUE,
  td_format = "wide", 
  time_dep_vars = c("statin_use_baseline", "bp_medication_baseline"),
  change_times = "0.5, 1, 2, 3, 4, 5",  # Annual visits
  td_suffix_pattern = "_year{time}",
  
  # Account for clinic clustering
  use_frailty = TRUE,
  frailty_var = "clinic_id",
  frailty_distribution = "gamma"
)
```

# Interpretation Guidelines {#interpretation}

## Understanding Time-Dependent Effects {#td-effects}

When interpreting results with time-dependent covariates:

1. **Hazard Ratios**: Represent the instantaneous risk at any given time, accounting for the current value of time-dependent variables.

2. **Model Fit**: Time-dependent models often fit better than static models but require more data and assumptions.

3. **Causality**: Time-dependent covariates help address time-varying confounding but don't guarantee causal interpretation.

## Common Pitfalls {#pitfalls}

1. **Immortal Time Bias**: Ensure time-dependent variables are properly aligned with risk periods.

2. **Missing Data**: Handle missing time-dependent measurements appropriately.

3. **Measurement Timing**: Consider whether variable changes are known at the time of measurement or retrospectively.

# Technical Considerations {#technical}

## Wide Format Requirements {#wide-requirements}

- Consistent naming pattern for time-dependent variables
- Complete specification of change times
- Baseline values for all time-dependent variables

## Long Format Requirements {#long-requirements}

- Proper `tstart`/`tstop` intervals with no gaps
- Status = 1 only in final interval for events
- Consistent covariate values within intervals

## Performance Considerations {#performance}

- Wide format is generally faster for simple patterns
- Long format is more flexible but computationally intensive
- Consider data size when choosing format

# Advanced Features Integration {#integration}

Time-dependent covariates work seamlessly with other advanced features:

```{r eval=FALSE}
# Comprehensive analysis combining all advanced features
comprehensive_analysis <- multisurvival(
  data = complex_dataset,
  elapsedtime = "time_to_event",
  outcome = "event_indicator",
  outcomeLevel = "1",
  explanatory = c("treatment_arm", "biomarker_status"),
  contexpl = c("age", "baseline_score"),
  
  # Time-dependent covariates
  use_time_dependent = TRUE,
  td_format = "wide",
  time_dep_vars = c("drug_dose_baseline", "toxicity_grade_baseline"),
  change_times = "4, 8, 12, 16, 20, 24",
  
  # Frailty for center effects
  use_frailty = TRUE,
  frailty_var = "study_center",
  frailty_distribution = "gamma",
  
  # Splines for non-linear age effects
  use_splines = TRUE,
  spline_vars = "age",
  spline_df = 4,
  spline_type = "ns",
  
  # Stratification for non-proportional hazards
  use_stratify = TRUE,
  stratvar = "biomarker_status",
  
  # Model selection
  use_modelSelection = TRUE,
  modelSelection = "both",
  selectionCriteria = "aic",
  
  # Risk assessment
  calculateRiskScore = TRUE,
  numRiskGroups = "four",
  plotRiskGroups = TRUE,
  
  # Comprehensive outputs
  hr = TRUE,
  km = TRUE,
  ac = TRUE,
  adjexplanatory = "treatment_arm",
  person_time = TRUE,
  ph_cox = TRUE,
  showNomogram = TRUE
)
```

# Conclusion {#conclusion}

Time-dependent covariates in `multisurvival` provide a powerful tool for analyzing survival data where important variables change over time. The support for both wide and long data formats makes this functionality accessible to researchers working with different data structures.

Key benefits include:

- **Realistic modeling** of clinical scenarios where treatments and patient status change
- **Flexible data input** supporting both wide and long formats  
- **Integration** with other advanced survival analysis features
- **Comprehensive output** including plots, tables, and model diagnostics

For most clinical applications, wide format provides the most intuitive approach, while long format offers maximum flexibility for complex time patterns.

## Further Reading {#further-reading}

- Therneau, T.M. and Grambsch, P.M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer.
- Kleinbaum, D.G. and Klein, M. (2012). *Survival Analysis: A Self-Learning Text*. Springer.
- ClinicoPath package documentation and examples.