---
title: "Getting Started with ClinicoPathDescriptives"
author: "ClinicoPathDescriptives Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ClinicoPathDescriptives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`ClinicoPathDescriptives` provides a suite of functions for summarising and visualising clinicopathological data. This vignette showcases the main tools in the package with reproducible examples. The package ships with example data sets `histopathology` and `treatmentResponse` to illustrate typical workflows.

```{r load-package, eval=FALSE}
# install.packages("ClinicoPathDescriptives")
library(ClinicoPath)

# Example data
data(histopathology)
head(histopathology)
```

# Summaries of Continuous Variables

Use `summarydata()` to obtain descriptive statistics for numeric variables.

```{r summarydata, eval=FALSE}
summarydata(data = histopathology, vars = vars(Age, TumorSize))
```

# Summaries of Categorical Variables

`reportcat()` creates frequency tables and counts for categorical variables.

```{r reportcat, eval=FALSE}
reportcat(data = histopathology, vars = vars(Sex, Grade))
```

# Cross Tables

Generate cross tabulations with chi-square tests using `crosstable()`.

```{r crosstable, eval=FALSE}
crosstable(
  data = histopathology,
  vars = vars(Sex, Grade),
  group = "PreinvasiveComponent",
  sty = "nejm",
  excl = TRUE
)
```

# Table One

`tableone()` produces formatted baseline characteristic tables for reports.

```{r tableone, eval=FALSE}
tableone(
  data = histopathology,
  vars = vars(Sex, PreinvasiveComponent, LVI, PNI, Grade, Age),
  sty = "t3",
  excl = TRUE
)
```

# Visualisations

The package includes multiple plotting functions to explore data.

## Age Pyramid

```{r agepyramid, eval=FALSE}
agepyramid(
  data = histopathology,
  age = "Age",
  gender = "Sex",
  female = "Female"
)
```

## Alluvial Diagrams

```{r alluvial, eval=FALSE}
alluvial(
  data = histopathology,
  vars = vars(Grade, LVI, PNI),
  fill = "first_variable"
)
```

## Benford Analysis

```{r benford, eval=FALSE}
benford(data = histopathology, var = "TumorSize")
```

## Venn and Variable Trees

```{r venn-vartree, eval=FALSE}
venn(data = histopathology, vars = vars(LVI, PNI, PreinvasiveComponent))
vartree(
  data = histopathology,
  vars = vars(Grade, LVI, PNI),
  percvar = "Grade",
  percvarLevel = "High"
)
```

## Waterfall Plots

`waterfall()` visualises tumor response over time.

```{r waterfall, eval=FALSE}
waterfall(
  data = treatmentResponse,
  patientID = "PatientID",
  responseVar = "Response",
  timeVar = "Month",
  inputType = "percentage",
  showWaterfallPlot = TRUE,
  showSpiderPlot = TRUE
)
```

# Conclusion

These examples demonstrate the versatility of `ClinicoPathDescriptives` for summarising and visualising clinicopathological data. Explore the help pages of each function for additional options and customisation.

