---
title: "Vignette: Creating an Age Pyramid with ClinicoPathDescriptives"
author: "Analysis by Gemini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette: Creating an Age Pyramid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# Create a sample data frame
set.seed(123) # for reproducibility
cohort_data <- data.frame(
  patient_id = 1:150,
  age = sample(20:85, 150, replace = TRUE),
  gender = sample(c("F", "M"), 150, replace = TRUE, prob = c(0.6, 0.4))
)

# View the first few rows
head(cohort_data)
```


```{r}
# This code simulates how the jamovi module would be called in an R environment.
# You would need the ClinicoPathDescriptives package installed.

# Load the library
library(ClinicoPath)

# Example 1: Basic age pyramid analysis
results1 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F", # Specify that 'F' represents females
    bin_width = 10, # Use 10-year age bins
    plot_title = "Age Distribution of Study Cohort"
)

# To view the results:
# View the summary table
print(results1$pyramidTable)

# View the plot
print(results1$plot)
```

## Example 2: Customizing Colors

You can customize the colors of the age pyramid bars using the `color1` and `color2` parameters:

```{r}
# Age pyramid with custom colors
results2 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F",
    bin_width = 5, # Use 5-year age bins for more detail
    plot_title = "Age Distribution with Custom Colors",
    color1 = "#E74C3C", # Red for first group
    color2 = "#3498DB"  # Blue for second group
)

# View the customized plot
print(results2$plot)
```

## Example 3: Different Age Bin Sizes

```{r}
# Compare different bin widths
# Fine granularity with 2-year bins
results3 <- agepyramid(
    data = cohort_data,
    age = "age",
    gender = "gender",
    female = "F",
    bin_width = 2,
    plot_title = "Detailed Age Distribution (2-year bins)",
    color1 = "#2ECC71", # Green
    color2 = "#F39C12"  # Orange
)

print(results3$plot)
```

## Example 4: Using Real Clinical Data

```{r}
# Load the histopathology dataset
data(histopathology)

# Create age pyramid for clinical cohort
clinical_results <- agepyramid(
    data = histopathology,
    age = "Age",
    gender = "Sex",
    female = "Female",
    bin_width = 10,
    plot_title = "Clinical Cohort Age Distribution",
    color1 = "#9B59B6", # Purple
    color2 = "#1ABC9C"  # Teal
)

# View results
print(clinical_results$pyramidTable)
print(clinical_results$plot)
```

## Color Options

The `color1` and `color2` parameters accept any valid color specification:

- **Hex codes**: `"#FF5733"`, `"#3498DB"`
- **Named colors**: `"red"`, `"blue"`, `"darkgreen"`
- **RGB values**: `"rgb(255, 87, 51)"`

## Interpretation

The age pyramid plot shows:
- **Population distribution** by age groups and gender
- **Age structure** of your study cohort
- **Gender balance** across different age ranges
- **Data completeness** for demographic analysis

The accompanying table provides exact counts for each age group and gender category.

