---
title: "Automated Exploratory Data Analysis with AutoEDA"
subtitle: "Comprehensive automated EDA using DataExplorer integration"
date: "`r Sys.Date()`"
author: "ClinicoPath"
output: html_document
---

# Introduction to AutoEDA

The AutoEDA module provides comprehensive automated exploratory data analysis capabilities specifically designed for clinical and pathological research. Built on the DataExplorer package framework, it offers intelligent automation while maintaining clinical relevance and interpretation.

## Key Features

- **7 Analysis Types**: From basic overview to comprehensive reporting
- **Clinical Intelligence**: Specialized pattern detection for medical data
- **Reference Range Validation**: Automatic detection of physiologically implausible values
- **Quality Assessment**: Clinical data quality scoring system
- **Target Variable Analysis**: Supervised EDA for outcome prediction
- **Missing Data Strategy**: Clinical-aware missing value analysis

## Research Foundation

Based on research from *R Journal 2019: The Landscape of R Packages for Automated Exploratory Data Analysis* (Staniak & Biecek), implementing best practices for automated statistical exploration.

# Getting Started

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Use the histopathology dataset for comprehensive examples
data("histopathology")
mydata <- histopathology

# Display basic dataset information
cat("Dataset dimensions:", nrow(mydata), "rows ×", ncol(mydata), "columns\n")
cat("Sample variables:", paste(head(names(mydata), 8), collapse = ", "), "...\n")
```

## Basic AutoEDA Workflow

The AutoEDA analysis follows a systematic approach:

1. **Variable Selection**: Choose variables for analysis
2. **Analysis Type**: Select from 7 analysis types
3. **Configuration**: Set parameters (target variable, correlation method, etc.)
4. **Execution**: Run automated analysis
5. **Interpretation**: Review insights and recommendations

# Analysis Types Overview

## 1. Dataset Overview Analysis

Comprehensive introduction to your dataset using DataExplorer's `introduce()` function:

```{r overview-example, eval=FALSE}
# Dataset Overview Analysis
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", 
           "Death", "Recurrence", "OverallTime"),
  analysis_type = "overview",
  include_plots = TRUE,
  plot_theme = "clinical"
)
```

**What you get:**
- Dataset dimensions and completeness
- Variable type breakdown (continuous vs discrete)
- Data quality assessment with scoring
- Memory usage analysis
- Recommendations for next steps

## 2. Missing Value Analysis

Comprehensive missing data pattern analysis:

```{r missing-example, eval=FALSE}
# Missing Value Analysis with Clinical Insights
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", 
           "MarginDistance", "MeasurementA", "MeasurementB"),
  analysis_type = "missing",
  missing_threshold = 10,  # Flag variables with >10% missing
  plot_theme = "clinical"
)
```

**Features:**
- Variable-by-variable missing value summary
- Percentage missing with threshold flagging
- Clinical recommendations for handling missing data
- MCAR/MAR/MNAR considerations for clinical data

## 3. Variable Distribution Analysis

Automated univariate analysis for both numeric and categorical variables:

```{r distribution-example, eval=FALSE}
# Distribution Analysis
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "OverallTime", 
           "MeasurementA", "MeasurementB"),
  analysis_type = "distributions",
  include_plots = TRUE,
  categorical_limit = 15,
  plot_theme = "clinical"
)
```

**Analyzes:**
- Continuous variables: histograms, summary statistics, normality
- Categorical variables: frequency distributions, level analysis
- Distribution characteristics and outlier detection
- Recommendations for statistical test selection

## 4. Correlation Analysis

Relationship analysis between numeric variables:

```{r correlation-example, eval=FALSE}
# Correlation Analysis with Multiple Methods
autoeda(
  data = mydata,
  vars = c("Age", "OverallTime", "MeasurementA", "MeasurementB", 
           "MarginDistance"),
  analysis_type = "correlation",
  correlation_method = "pearson",  # or "spearman", "kendall"
  include_plots = TRUE,
  plot_theme = "clinical"
)
```

**Provides:**
- Correlation matrix calculation
- Strong correlation detection (|r| > 0.7)
- Multicollinearity warnings
- Interpretation guidelines for correlation strength

## 5. Principal Component Analysis (PCA)

Dimensionality reduction and feature importance:

```{r pca-example, eval=FALSE}
# PCA Analysis
autoeda(
  data = mydata,
  vars = c("Age", "OverallTime", "MeasurementA", "MeasurementB", 
           "MarginDistance"),
  analysis_type = "pca",
  pca_components = 5,
  include_plots = TRUE,
  plot_theme = "clinical"
)
```

**Results:**
- Principal component variance explained
- Cumulative variance analysis
- Dimensionality reduction assessment
- Recommendations for component retention

## 6. Target Variable Analysis

Supervised EDA with outcome variable relationships:

```{r target-example, eval=FALSE}
# Target Variable Analysis for Survival Outcome
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", 
           "MeasurementA", "MeasurementB"),
  target_var = "Death",  # Outcome variable
  analysis_type = "target",
  include_plots = TRUE,
  plot_theme = "clinical"
)

# Target Analysis for Continuous Outcome
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI"),
  target_var = "OverallTime",  # Time to event
  analysis_type = "target",
  include_plots = TRUE,
  plot_theme = "clinical"
)
```

**Features:**
- Target variable characterization (continuous vs categorical)
- Relationship analysis with predictors
- Statistical approach recommendations
- Feature importance insights

## 7. Comprehensive Report

Complete automated EDA report combining all analyses:

```{r comprehensive-example, eval=FALSE}
# Comprehensive AutoEDA Report
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", 
           "Death", "Recurrence", "OverallTime", "MeasurementA", "MeasurementB"),
  analysis_type = "comprehensive",
  target_var = "Death",
  correlation_method = "spearman",
  missing_threshold = 5,
  pca_components = 5,
  include_plots = TRUE,
  advanced_options = TRUE,
  plot_theme = "clinical",
  output_format = "combined"
)
```

**Comprehensive Output:**
- All analysis types in one report
- Executive summary with key findings
- Data quality assessment
- Clinical insights and patterns
- Next steps recommendations

# Advanced Clinical Features

## Clinical Pattern Detection

The AutoEDA module includes specialized detection for clinical variable patterns:

```{r clinical-patterns, eval=FALSE}
# Advanced Clinical Analysis
autoeda(
  data = mydata,
  vars = names(mydata),  # Analyze all variables
  analysis_type = "comprehensive",
  advanced_options = TRUE,  # Enable clinical insights
  plot_theme = "clinical"
)
```

**Detects:**
- **Age Variables**: Age-related patterns and distributions
- **Vital Signs**: Blood pressure, temperature, heart rate patterns
- **Laboratory Values**: Glucose, cholesterol, HbA1c ranges
- **Anthropometric**: Weight, height, BMI analysis
- **Diagnostic Variables**: Stage, grade, diagnosis hierarchies
- **Treatment Variables**: Therapy, medication, surgery patterns
- **Outcome Variables**: Survival, recurrence, progression analysis

## Reference Range Validation

Automatic detection of physiologically implausible values:

```{r reference-ranges, eval=FALSE}
# Create sample clinical data with potential outliers
clinical_data <- data.frame(
  patient_id = 1:100,
  age = c(rnorm(95, 65, 15), 150, 200, -5, 250, 300),  # Some impossible ages
  weight = c(rnorm(95, 70, 15), 600, 700, -10, 0.1, 800),  # Impossible weights
  systolic_bp = c(rnorm(95, 120, 20), 300, 400, 20, 500, 10),  # Impossible BP
  temperature = c(rnorm(95, 36.5, 0.5), 50, 60, 10, 70, 0)  # Impossible temp
)

# Analyze with reference range checking
autoeda(
  data = clinical_data,
  vars = c("age", "weight", "systolic_bp", "temperature"),
  analysis_type = "comprehensive",
  advanced_options = TRUE,
  plot_theme = "clinical"
)
```

**Validates Against:**
- Age: 0-120 years
- Weight: 0.5-500 kg
- Height: 30-250 cm
- Temperature: 35-42°C
- Blood pressure: Systolic 70-250, Diastolic 40-150 mmHg
- Heart rate: 30-200 bpm
- Common lab values: Glucose, cholesterol ranges

## Clinical Data Quality Scoring

Automated assessment of clinical data quality:

```{r quality-scoring, eval=FALSE}
# Data Quality Assessment
autoeda(
  data = mydata,
  vars = c("Age", "Sex", "Grade", "TStage", "Death", "OverallTime"),
  analysis_type = "overview",
  advanced_options = TRUE
)
```

**Quality Score Factors:**
- Sample size adequacy (penalty for <30 cases)
- Variables with high missing values (>20%)
- Critical variables missing (outcomes, diagnoses)
- Overall completeness percentage

**Scoring:**
- **80-100**: Excellent - Ready for analysis
- **60-79**: Good - Minor issues to address  
- **0-59**: Needs Improvement - Significant data quality concerns

# Parameter Reference

## Core Parameters

```{r params-demo, eval=FALSE}
autoeda(
  data = mydata,
  
  # Variable selection
  vars = c("Age", "Sex", "Grade"),  # Required: variables to analyze
  
  # Analysis configuration  
  analysis_type = "overview",  # "overview", "missing", "distributions", 
                              # "correlation", "pca", "target", "comprehensive"
  target_var = "Death",       # Optional: for supervised analysis
  
  # Missing value settings
  missing_threshold = 5,      # Threshold % for flagging missing values
  
  # Correlation settings
  correlation_method = "pearson",  # "pearson", "spearman", "kendall"
  
  # PCA settings
  pca_components = 5,         # Number of components to analyze
  
  # Visualization settings
  include_plots = TRUE,       # Include automated plots
  plot_theme = "clinical",    # "default", "minimal", "classic", "clinical"
  
  # Output settings
  output_format = "combined", # "html", "tables", "combined"
  
  # Advanced features
  advanced_options = TRUE,    # Enable clinical insights
  categorical_limit = 15,     # Max levels for categorical variables
  generate_report = FALSE     # Generate comprehensive report file
)
```

## Analysis Type Details

| Type | Description | Best For |
|------|-------------|----------|
| `overview` | Dataset introduction and summary | Initial data exploration |
| `missing` | Missing value patterns | Data quality assessment |
| `distributions` | Univariate analysis | Variable characterization |
| `correlation` | Relationship analysis | Feature selection prep |
| `pca` | Dimensionality reduction | High-dimensional data |
| `target` | Supervised EDA | Outcome prediction prep |
| `comprehensive` | Complete automated report | Full analysis workflow |

# Clinical Research Applications

## Pathology Research Example

```{r pathology-research, eval=FALSE}
# Comprehensive analysis for pathology research
autoeda(
  data = histopathology,
  vars = c("Age", "Sex", "Race", "Grade", "TStage", "NStage", 
           "LVI", "PNI", "Death", "Recurrence", "OverallTime"),
  analysis_type = "comprehensive",
  target_var = "Death",
  correlation_method = "spearman",  # More robust for ordinal data
  missing_threshold = 5,
  advanced_options = TRUE,
  plot_theme = "clinical"
)
```

## Biomarker Discovery Workflow

```{r biomarker-workflow, eval=FALSE}
# Focus on measurement variables for biomarker analysis
autoeda(
  data = histopathology,
  vars = c("MeasurementA", "MeasurementB", "Age", "Grade", "TStage", "Death"),
  analysis_type = "target",
  target_var = "Death",
  correlation_method = "pearson",
  include_plots = TRUE,
  advanced_options = TRUE,
  plot_theme = "clinical"
)
```

## Survival Analysis Preparation

```{r survival-prep, eval=FALSE}
# Prepare data for survival analysis
autoeda(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", 
           "OverallTime", "Death"),
  analysis_type = "comprehensive",
  target_var = "OverallTime",  # Time to event
  missing_threshold = 3,  # Strict threshold for survival data
  advanced_options = TRUE,
  plot_theme = "clinical"
)
```

# Best Practices

## Variable Selection Strategy

1. **Start Broad**: Include all potentially relevant variables
2. **Focus Analysis**: Use specific analysis types for targeted exploration
3. **Target-Driven**: Always specify target variable when appropriate
4. **Iterative Approach**: Use insights to refine variable selection

## Clinical Data Considerations

1. **Missing Data Strategy**: Consider clinical reasons for missingness
2. **Reference Ranges**: Review flagged outliers for data entry errors
3. **Variable Hierarchies**: Understand staging and grading systems
4. **Temporal Relationships**: Consider treatment sequences and follow-up

## Analysis Workflow

1. **Overview First**: Start with dataset overview
2. **Quality Assessment**: Check missing values and data quality
3. **Target Analysis**: Focus on outcome relationships
4. **Comprehensive Report**: Generate final automated report

# Interpretation Guidelines

## Data Quality Indicators

- **Excellent (80-100)**: Ready for immediate analysis
- **Good (60-79)**: Address minor data quality issues
- **Poor (<60)**: Significant preprocessing required

## Missing Value Thresholds

- **<5%**: Generally acceptable for most analyses
- **5-15%**: Consider imputation strategies
- **15-30%**: Evaluate variable importance vs completeness
- **>30%**: Consider excluding variable unless critical

## Correlation Strength

- **|r| > 0.7**: Strong correlation, check multicollinearity
- **0.3 < |r| < 0.7**: Moderate correlation, meaningful relationship
- **|r| < 0.3**: Weak correlation, limited linear relationship

## Sample Size Guidelines

- **n ≥ 100**: Adequate for most analyses
- **30 ≤ n < 100**: Acceptable with careful interpretation
- **n < 30**: Limited power, consider descriptive analysis only

# Advanced Topics

## Custom Clinical Patterns

The module can be extended to detect custom clinical patterns by modifying the pattern detection logic.

## Integration with Other Modules

AutoEDA results inform:
- **Survival Analysis**: Variable selection and preprocessing
- **Decision Analysis**: Feature importance for decision trees
- **Statistical Plots**: Variable characterization for visualization
- **Cross-tabulation**: Categorical variable relationships

## Computational Considerations

- **Large Datasets**: Consider variable subset for initial exploration
- **Memory Usage**: Monitor memory usage displayed in overview
- **Processing Time**: Comprehensive analysis may take time with many variables

# Troubleshooting

## Common Issues

**DataExplorer not installed:**
```r
install.packages("DataExplorer")
```

**Insufficient variables for correlation/PCA:**
- Ensure at least 2 numeric variables selected
- Check for complete cases after missing value removal

**No target variable detected:**
- Verify target variable name spelling
- Ensure target variable exists in selected variables

**Plot generation errors:**
- Try different plot themes
- Disable plots with `include_plots = FALSE`

# References

1. Staniak, M., & Biecek, P. (2019). The landscape of R packages for automated exploratory data analysis. *The R Journal*, 11(2), 249-269.

2. DataExplorer package documentation: https://cran.r-project.org/package=DataExplorer

3. ClinicoPath package documentation: Clinical and pathological research analysis tools

---

*This vignette demonstrates the comprehensive AutoEDA capabilities in ClinicoPath. The module provides intelligent automation while maintaining clinical relevance and interpretability for medical research applications.*
