---
title: "CONSORT Flow Charts for Clinical Trial Reporting"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 10
vignette: >
  %\VignetteIndexEntry{CONSORT Flow Charts for Clinical Trial Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/",
  warning = FALSE,
  message = FALSE
)

# Load required libraries
# library(ClinicoPath)  # Commented out for vignette building
library(DiagrammeR)
```

# Introduction to CONSORT Flow Charts

The **CONSORT (Consolidated Standards of Reporting Trials)** statement provides evidence-based recommendations for reporting randomized controlled trials (RCTs). A key component is the flow diagram that tracks participants through each stage of the trial, from initial assessment through final analysis.

## Why CONSORT Flow Charts Matter

CONSORT flow diagrams are essential for:

- **Transparency**: Showing exactly how many participants entered and completed the study
- **Quality Assessment**: Allowing readers to evaluate potential bias from dropout patterns
- **Reproducibility**: Providing clear documentation of study conduct
- **Journal Requirements**: Most medical journals require CONSORT-compliant reporting
- **Regulatory Compliance**: FDA and EMA expect CONSORT-style reporting

## The ClinicoPath CONSORT Function

The `consort()` function in ClinicoPath generates publication-ready CONSORT flow diagrams automatically. Simply input your trial numbers, and it creates both visual flowcharts and summary statistics.

# Basic CONSORT Flow Chart

Let's start with a simple two-arm trial example:

```{r basic-example, eval=FALSE}
# Basic cardiovascular trial
basic_consort <- consort(
  data = data.frame(),  # consort doesn't analyze patient data
  initialN = 1000,
  notEligibleN = 300,
  notEligibleText = "Age >80 years (n=150), Prior MI <6 months (n=100), Other (n=50)",
  randomizedN = 700,
  arm1Label = "ACE Inhibitor",
  arm1N = 350,
  arm1ReceivedN = 340,
  arm1LostN = 45,
  arm1AnalyzedN = 295,
  arm2Label = "Placebo",
  arm2N = 350,
  arm2ReceivedN = 345,
  arm2LostN = 40,
  arm2AnalyzedN = 305,
  excludedText = "Lost to follow-up, protocol violations"
)
```

The function produces:
1. **Summary Table**: Key numbers with percentages
2. **Flow Diagram**: Visual CONSORT chart
3. **Text Summary**: Detailed narrative description

# Real-World Examples Using Sample Data

ClinicoPath includes the `consort_examples_data` dataset with realistic trial scenarios:

```{r load-data, eval=FALSE}
# Load the example data
data(consort_examples_data)
str(consort_examples_data)

# View the different trial types
consort_examples_data$trial_name
```

## Example 1: Large Cardiovascular Outcomes Trial

```{r cv-trial, eval=FALSE}
# Extract cardiovascular trial data
cv_trial <- consort_examples_data[1, ]

# Create CONSORT flowchart
cv_consort <- consort(
  data = data.frame(),
  initialN = cv_trial$initial_assessed,
  notEligibleN = cv_trial$not_eligible,
  notEligibleText = cv_trial$not_eligible_reasons,
  randomizedN = cv_trial$randomized,
  arm1Label = cv_trial$arm1_label,
  arm1N = cv_trial$arm1_allocated,
  arm1ReceivedN = cv_trial$arm1_received,
  arm1LostN = cv_trial$arm1_lost_followup,
  arm1AnalyzedN = cv_trial$arm1_analyzed,
  arm2Label = cv_trial$arm2_label,
  arm2N = cv_trial$arm2_allocated,
  arm2ReceivedN = cv_trial$arm2_received,
  arm2LostN = cv_trial$arm2_lost_followup,
  arm2AnalyzedN = cv_trial$arm2_analyzed,
  excludedText = cv_trial$exclusion_reasons
)
```

This represents a typical large-scale cardiovascular outcomes trial with:
- 5,000 patients initially screened
- 30% exclusion rate (typical for cardiovascular studies)
- Balanced randomization
- ~10% dropout rate (acceptable for long-term studies)

## Example 2: Cancer Immunotherapy Trial

```{r cancer-trial, eval=FALSE}
# Extract cancer trial data
cancer_trial <- consort_examples_data[2, ]

cancer_consort <- consort(
  data = data.frame(),
  initialN = cancer_trial$initial_assessed,
  notEligibleN = cancer_trial$not_eligible,
  notEligibleText = cancer_trial$not_eligible_reasons,
  randomizedN = cancer_trial$randomized,
  arm1Label = cancer_trial$arm1_label,
  arm1N = cancer_trial$arm1_allocated,
  arm1ReceivedN = cancer_trial$arm1_received,
  arm1LostN = cancer_trial$arm1_lost_followup,
  arm1AnalyzedN = cancer_trial$arm1_analyzed,
  arm2Label = cancer_trial$arm2_label,
  arm2N = cancer_trial$arm2_allocated,
  arm2ReceivedN = cancer_trial$arm2_received,
  arm2LostN = cancer_trial$arm2_lost_followup,
  arm2AnalyzedN = cancer_trial$arm2_analyzed,
  excludedText = cancer_trial$exclusion_reasons
)
```

Cancer trials often have:
- Strict inclusion criteria (29% exclusion here)
- Higher dropout rates due to disease progression
- Time-to-event endpoints requiring careful follow-up

## Example 3: Small Pilot Study

```{r pilot-trial, eval=FALSE}
# Extract pilot study data
pilot_trial <- consort_examples_data[5, ]

pilot_consort <- consort(
  data = data.frame(),
  initialN = pilot_trial$initial_assessed,
  notEligibleN = pilot_trial$not_eligible,
  notEligibleText = pilot_trial$not_eligible_reasons,
  randomizedN = pilot_trial$randomized,
  arm1Label = pilot_trial$arm1_label,
  arm1N = pilot_trial$arm1_allocated,
  arm1ReceivedN = pilot_trial$arm1_received,
  arm1LostN = pilot_trial$arm1_lost_followup,
  arm1AnalyzedN = pilot_trial$arm1_analyzed,
  arm2Label = pilot_trial$arm2_label,
  arm2N = pilot_trial$arm2_allocated,
  arm2ReceivedN = pilot_trial$arm2_received,
  arm2LostN = pilot_trial$arm2_lost_followup,
  arm2AnalyzedN = pilot_trial$arm2_analyzed,
  excludedText = pilot_trial$exclusion_reasons
)
```

Pilot studies typically have:
- Smaller sample sizes (150 assessed, 100 randomized)
- Focus on feasibility and safety
- Higher proportional exclusions due to strict criteria

# Advanced Usage Scenarios

## Unequal Randomization (2:1 Ratio)

Some trials use unequal randomization to get more safety data on the experimental treatment:

```{r unequal-randomization, eval=FALSE}
unequal_consort <- consort(
  data = data.frame(),
  initialN = 900,
  notEligibleN = 300,
  notEligibleText = "Failed eligibility criteria",
  randomizedN = 600,
  arm1Label = "Experimental Drug (2:1)",
  arm1N = 400,  # 2/3 of randomized patients
  arm1ReceivedN = 390,
  arm1LostN = 60,
  arm1AnalyzedN = 330,
  arm2Label = "Standard Care (2:1)",
  arm2N = 200,  # 1/3 of randomized patients
  arm2ReceivedN = 195,
  arm2LostN = 25,
  arm2AnalyzedN = 170,
  excludedText = "Lost to follow-up, adverse events"
)
```

## High Dropout Scenario

Some studies face high dropout rates that need careful documentation:

```{r high-dropout, eval=FALSE}
high_dropout_consort <- consort(
  data = data.frame(),
  initialN = 500,
  notEligibleN = 100,
  notEligibleText = "High-risk population with multiple comorbidities",
  randomizedN = 400,
  arm1Label = "Intensive Intervention",
  arm1N = 200,
  arm1ReceivedN = 185,
  arm1LostN = 75,  # 40% dropout
  arm1AnalyzedN = 110,
  arm2Label = "Standard Care",
  arm2N = 200,
  arm2ReceivedN = 195,
  arm2LostN = 45,  # 23% dropout
  arm2AnalyzedN = 150,
  excludedText = "High burden intervention, patient withdrawal, adverse events"
)
```

# CONSORT Guidelines and Best Practices

## Essential Elements

Every CONSORT flow diagram should include:

1. **Enrollment**:
   - Number assessed for eligibility
   - Number excluded (with reasons)
   - Number randomized

2. **Allocation**:
   - Number allocated to each intervention
   - Number receiving intended treatment

3. **Follow-up**:
   - Number lost to follow-up (with reasons)
   - Number discontinuing intervention

4. **Analysis**:
   - Number analyzed for primary outcome
   - Exclusions from analysis

## Common Patterns by Study Type

### Phase I Dose Escalation
```{r phase1-example, eval=FALSE}
# Small numbers, safety focus
phase1_consort <- consort(
  data = data.frame(),
  initialN = 60,
  notEligibleN = 15,
  notEligibleText = "Strict safety criteria",
  randomizedN = 45,
  arm1Label = "Dose Level 1",
  arm1N = 15,
  arm1ReceivedN = 15,
  arm1LostN = 1,
  arm1AnalyzedN = 14,
  arm2Label = "Dose Level 2", 
  arm2N = 15,
  arm2ReceivedN = 14,
  arm2LostN = 2,
  arm2AnalyzedN = 12
)
```

### Phase II Efficacy Studies
```{r phase2-example, eval=FALSE}
# Moderate size, efficacy focus
phase2_consort <- consort(
  data = data.frame(),
  initialN = 300,
  notEligibleN = 80,
  randomizedN = 220,
  arm1Label = "Experimental",
  arm1N = 110,
  arm1AnalyzedN = 95,
  arm2Label = "Control",
  arm2N = 110,
  arm2AnalyzedN = 98
)
```

### Phase III Confirmatory Trials
```{r phase3-example, eval=FALSE}
# Large scale, definitive evidence
phase3_consort <- consort(
  data = data.frame(),
  initialN = 3000,
  notEligibleN = 1000,
  randomizedN = 2000,
  arm1Label = "New Treatment",
  arm1N = 1000,
  arm1AnalyzedN = 850,
  arm2Label = "Standard Care",
  arm2N = 1000,
  arm2AnalyzedN = 875
)
```

# Integration with jamovi

The CONSORT function integrates seamlessly with jamovi:

1. **Open jamovi** and load your module
2. **Navigate** to Exploration → ClinicoPath Reports → CONSORT Flowchart
3. **Enter your trial numbers** in the interface:
   - Enrollment numbers
   - Randomization details
   - Arm-specific data
   - Exclusion reasons
4. **Review outputs**:
   - Summary statistics table
   - Visual flow diagram
   - Text summary

## Tips for jamovi Users

- **Start with estimates**: Begin with approximate numbers and refine
- **Use descriptive labels**: Clear arm names improve readability
- **Include detailed reasons**: Specific exclusion text enhances transparency
- **Export for publication**: Use the generated diagrams directly in manuscripts

# Statistical Considerations

## Sample Size and Power
CONSORT diagrams help readers assess whether the study had adequate power:

```{r power-assessment, eval=FALSE}
# Calculate effective sample sizes from our examples
cv_effective_n <- cv_trial$arm1_analyzed + cv_trial$arm2_analyzed
cancer_effective_n <- cancer_trial$arm1_analyzed + cancer_trial$arm2_analyzed

cat("Cardiovascular trial analyzed N:", cv_effective_n, "\n")
cat("Cancer trial analyzed N:", cancer_effective_n, "\n")

# Dropout rates
cv_dropout_rate <- (cv_trial$arm1_lost_followup + cv_trial$arm2_lost_followup) / cv_trial$randomized * 100
cancer_dropout_rate <- (cancer_trial$arm1_lost_followup + cancer_trial$arm2_lost_followup) / cancer_trial$randomized * 100

cat("CV trial dropout rate:", round(cv_dropout_rate, 1), "%\n")
cat("Cancer trial dropout rate:", round(cancer_dropout_rate, 1), "%\n")
```

## Bias Assessment

CONSORT diagrams help identify potential sources of bias:

- **Imbalanced dropout**: Unequal loss between arms
- **High overall dropout**: >20% may compromise validity
- **Differential reasons**: Different dropout reasons between arms
- **Post-randomization exclusions**: Should be minimal and balanced

## Reporting Standards

Different journals may have specific requirements:

- **High-impact journals**: Often require detailed CONSORT compliance
- **Specialty journals**: May have field-specific modifications
- **Regulatory submissions**: FDA/EMA have specific formatting expectations

# Troubleshooting Common Issues

## Inconsistent Numbers

Ensure logical flow:
```{r number-checks, eval=FALSE}
# Check: randomized = arm1 + arm2
randomized_check <- arm1N + arm2N == randomizedN

# Check: analyzed <= received <= allocated
arm1_logic <- arm1AnalyzedN <= arm1ReceivedN & arm1ReceivedN <= arm1N
arm2_logic <- arm2AnalyzedN <= arm2ReceivedN & arm2ReceivedN <= arm2N
```

## Missing Data Handling

Be explicit about:
- Patients with missing primary endpoint data
- Protocol violations
- Consent withdrawals
- Lost to follow-up vs. study discontinuation

## Complex Study Designs

For crossover, cluster randomized, or adaptive designs:
- Modify the standard CONSORT template
- Add additional flowchart elements as needed
- Consult CONSORT extensions for specific designs

# Conclusion

The ClinicoPath CONSORT function provides:

✅ **Easy generation** of publication-ready flow diagrams  
✅ **Automatic calculations** of percentages and summaries  
✅ **Professional formatting** meeting journal standards  
✅ **Comprehensive examples** for different trial types  
✅ **Integration with jamovi** for interactive use  

## Next Steps

1. **Practice** with the provided examples
2. **Adapt** the parameters to your own trial data
3. **Export** diagrams for your publications
4. **Combine** with other ClinicoPath functions for complete trial reporting

## References

- Schulz KF, Altman DG, Moher D, for the CONSORT Group. CONSORT 2010 Statement: updated guidelines for reporting parallel group randomised trials. BMJ 2010;340:c332.
- CONSORT website: http://www.consort-statement.org/
- Clinical trial reporting guidelines: https://www.equator-network.org/

---

*For more information about ClinicoPath, visit our documentation at [ClinicoPath Guide](https://www.serdarbalci.com/ClinicoPathJamoviModule/)*