---
title: "Comprehensive IHC Analysis Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive IHC Analysis Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The `ihcstats` function in ClinicoPath provides comprehensive immunohistochemical (IHC) expression analysis implementing methodologies from landmark research papers:

1. **Matsuoka et al. (2011)** - Ward's hierarchical clustering for colorectal cancer prognosis
2. **Carvalho et al. (2011)** - Iterative marker selection for renal oncocytoma diagnosis  
3. **Olsen et al. (2006)** - Differential diagnosis clustering for sarcomas
4. **Sterlacci et al. (2019)** - TIL signature analysis for NSCLC

## Setup

```{r setup}
library(ClinicoPath)
library(dplyr)
library(ggplot2)
library(knitr)

# Load available dataset
data("ihc_test_data")

# Check available variables
cat("Available variables in ihc_test_data:\n")
print(names(ihc_test_data))
```

## Core Functionality Overview

### Clustering Methods Available

```{r clustering_overview}
cat("IHC Clustering Methods:\n")
cat("1. Hierarchical clustering with multiple distance metrics\n")
cat("2. PAM (Partitioning Around Medoids) clustering\n")
cat("3. K-means clustering\n")
cat("4. Fuzzy clustering\n\n")

cat("Distance Metrics:\n")
cat("- Gower: Mixed data types (categorical + continuous)\n")
cat("- Jaccard: Binary/categorical data\n")
cat("- Euclidean: Continuous data\n")
cat("- Manhattan: Robust to outliers\n\n")
```

## Methodology Implementation

### Matsuoka 2011 Methodology

```{r matsuoka_methodology}
cat("Matsuoka 2011 Features:\n")
cat("- Ward's hierarchical clustering (ward.D2)\n")
cat("- 3-tier scoring system (0, 1, 2)\n")
cat("- Prognostic clustering with survival analysis\n")
cat("- Multi-region tumor analysis (central vs invasive)\n")
cat("- Cox regression for prognostic groups\n\n")

cat("Target: Colorectal cancer prognosis\n")
cat("Key markers: p53, Ki67, VEGF, COX2\n")
```

### Carvalho 2011 Methodology

```{r carvalho_methodology}
cat("Carvalho 2011 Features:\n")
cat("- Iterative marker selection algorithm\n")
cat("- Binary scoring system for specific markers\n")
cat("- Renal tumor differential diagnosis\n")
cat("- Automatic marker importance ranking\n")
cat("- Diagnostic accuracy optimization\n\n")

cat("Target: Renal oncocytoma vs chromophobe RCC\n")
cat("Key markers: CK7, CD117, E-cadherin, Claudin-7\n")
```

### Olsen 2006 Methodology

```{r olsen_methodology}
cat("Olsen 2006 Features:\n")
cat("- Hierarchical clustering for sarcoma classification\n")
cat("- Standard IHC scoring scale integration\n")
cat("- Differential diagnosis clustering\n")
cat("- Subtype-specific marker panels\n")
cat("- Morphology correlation analysis\n\n")

cat("Target: Sarcoma differential diagnosis\n")
cat("Key markers: Vimentin, SMA, Desmin, S100, CD34\n")
```

### Sterlacci 2019 Methodology

```{r sterlacci_methodology}
cat("Sterlacci 2019 Features:\n")
cat("- TIL (Tumor-Infiltrating Lymphocytes) signature analysis\n")
cat("- CD3, CD8, CD20, PD-L1 integration\n")
cat("- Immune microenvironment clustering\n")
cat("- NSCLC histotype correlation\n")
cat("- Prognostic immune signatures\n\n")

cat("Target: NSCLC immune microenvironment\n")
cat("Key markers: CD3, CD8, CD20, PD-L1, FOXP3\n")
```

## Advanced Features

### H-Score Calculation

```{r hscore_features}
cat("H-Score Calculation:\n")
cat("- Automated H-Score computation from IHC data\n")
cat("- Integration with clustering results\n")
cat("- Intensity and percentage scoring\n")
cat("- Quality control and validation metrics\n\n")
```

### Prognostic Analysis

```{r prognostic_features}
cat("Prognostic Analysis Features:\n")
cat("- Survival time and event integration\n")
cat("- Cox proportional hazards modeling\n")
cat("- Kaplan-Meier survival curves\n")
cat("- Log-rank test for cluster comparison\n")
cat("- Hazard ratio calculation with confidence intervals\n\n")
```

### Visualization Options

```{r visualization_features}
cat("Visualization Components:\n")
cat("- Hierarchical clustering dendrograms\n")
cat("- IHC expression heatmaps\n")
cat("- Cluster assignment plots\n")
cat("- Survival curves by cluster\n")
cat("- Marker importance rankings\n")
cat("- Correlation matrices\n\n")
```

## Clinical Applications

### Diagnostic Workflows

```{r diagnostic_workflows}
cat("Diagnostic Applications:\n")
cat("1. Sarcoma subtype classification\n")
cat("2. Renal tumor differential diagnosis\n")
cat("3. Colorectal cancer prognostic grouping\n")
cat("4. NSCLC immune profiling\n")
cat("5. Custom marker panel optimization\n\n")
```

### Research Applications

```{r research_applications}
cat("Research Applications:\n")
cat("1. Biomarker discovery and validation\n")
cat("2. Tumor heterogeneity analysis\n")
cat("3. Immune microenvironment characterization\n")
cat("4. Multi-region tumor analysis\n")
cat("5. Prognostic signature development\n\n")
```

## Data Requirements

### Input Data Format

```{r data_requirements}
cat("Required Data Structure:\n")
cat("- Patient/case identifier column\n")
cat("- IHC marker expression variables\n")
cat("- Optional: Survival time and event variables\n")
cat("- Optional: Clinical and pathological variables\n")
cat("- Optional: Multi-region scoring variables\n\n")

cat("Scoring Systems Supported:\n")
cat("- Binary (0/1, negative/positive)\n")
cat("- 3-tier (0/1/2, negative/weak/strong)\n")
cat("- 4-tier (0/1/2/3)\n")
cat("- H-Score (0-300)\n")
cat("- Percentage (0-100)\n")
```

## Implementation in jamovi

All IHC analysis functions are available in jamovi:

1. **IHC Stats**: `Exploration > Specialized Analyses > IHC Statistical Analysis`
2. **Data Input**: Direct import of IHC scoring spreadsheets
3. **Interactive Parameters**: Point-and-click parameter selection
4. **Visual Output**: Integrated plots and tables in jamovi results

## Best Practices

### Data Preparation

1. **Standardize Scoring**: Use consistent scoring systems across observers
2. **Quality Control**: Implement inter-observer agreement assessment
3. **Missing Data**: Handle missing scores appropriately
4. **Validation**: Cross-validate results with independent datasets

### Analysis Strategy

1. **Method Selection**: Choose methodology based on clinical question
2. **Parameter Optimization**: Test different clustering parameters
3. **Validation**: Validate clusters with clinical outcomes
4. **Interpretation**: Consider biological plausibility of results

### Reporting Guidelines

1. **Methodology**: Report complete analysis methodology
2. **Parameters**: Document all analysis parameters used
3. **Validation**: Include validation metrics and statistics
4. **Clinical Relevance**: Discuss clinical implications of findings

## Further Resources

- **Individual Function Vignettes**: For detailed parameter usage
- **Clinical Workflow Guide**: For integrated analysis pipelines
- **Getting Started Guide**: For basic ClinicoPath functionality
- **Visualization Gallery**: For advanced plot customization

## Conclusion

The `ihcstats` function provides a comprehensive platform for IHC analysis implementing proven methodologies from the literature. The integration with jamovi makes these advanced techniques accessible to clinical researchers and pathologists without programming experience.

For specific use cases and detailed examples, consult the individual function vignettes and clinical workflow guides.
