---
title: "Data Quality Assessment - Enhanced Visual Analysis with visdat Integration"
subtitle: "Comprehensive guide to the ClinicoPath dataquality function for clinical research data validation"
date: "`r Sys.Date()`"
author: "ClinicoPath"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 4
    number-sections: true
    theme: cosmo
    df-print: paged
editor: visual
---

# Introduction to Enhanced Data Quality Assessment

The `dataquality` function in ClinicoPath provides sophisticated data quality assessment capabilities specifically designed for clinical research. Unlike general data quality tools, this function integrates advanced visual analysis using the `visdat` package, offering researchers a comprehensive suite of tools for data validation, pattern detection, and quality assurance.

## Why Enhanced Data Quality Assessment?

### **Clinical Research Requirements**
- **Regulatory Compliance**: Meet ICH E6 and FDA data quality standards
- **Patient Safety**: Ensure treatment decisions are based on reliable data
- **Research Integrity**: Validate data before statistical analysis
- **Publication Quality**: Enhance credibility of research findings

### **Advanced Features**
- **Visual Data Exploration**: Integration with visdat for comprehensive visual analysis
- **Multiple Analysis Types**: Missing patterns, duplicates, completeness analysis
- **Flexible Configuration**: Customizable thresholds and analysis options
- **Export Capabilities**: Professional plots ready for publication

## What Makes dataquality Different?

### **Unique Capabilities**
1. **visdat Integration**: First jamovi module to integrate visdat visual analysis
2. **Clinical Context**: Designed specifically for medical research workflows
3. **Comprehensive Analysis**: Combined statistical and visual assessment
4. **Flexible Outputs**: HTML reports with interactive insights

### **Analysis Types Available**
- **Data Overview (vis_dat)**: Variable types and missing value patterns
- **Missing Patterns (vis_miss)**: Missing value clustering and relationships
- **Data Types (vis_guess)**: Automatic type detection and validation
- **Value Expectations (vis_expect)**: Expected vs actual data patterns
- **Comprehensive Analysis**: All visual analyses combined

# Getting Started

## Load Required Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(ClinicoPath)
library(dplyr)
library(knitr)

# Load clinical research datasets
data("histopathology")
data("treatmentResponse")

# Display overview of available datasets
cat("📊 Clinical Research Datasets Available:\n")
cat("   - histopathology: Pathological research data (", nrow(histopathology), " patients)\n")
if(exists("treatmentResponse")) {
  cat("   - treatmentResponse: Treatment outcome data (", nrow(treatmentResponse), " patients)\n")
}
```

## Basic Data Quality Workflow

The enhanced data quality assessment workflow follows these systematic steps:

1. **Variable Selection**: Choose variables for quality assessment
2. **Analysis Configuration**: Select missing data, duplicates, visual analysis options
3. **Visual Analysis Type**: Choose specific visdat analysis type
4. **Threshold Setting**: Configure missing data highlight thresholds
5. **Export Options**: Enable plot export for publications
6. **Execution**: Run comprehensive quality assessment
7. **Interpretation**: Review results and recommendations

# Core Examples and Applications

## Example 1: Basic Data Quality Assessment

Starting with a simple quality assessment of core clinical variables.

```{r basic-assessment, eval=FALSE}
# Basic data quality assessment
basic_quality <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = FALSE,  # Start without visual analysis
  visdat_type = "vis_dat",
  missing_threshold_visual = 10,
  export_plots = FALSE
)

# View basic quality results
print("📋 Basic Data Quality Assessment:")
print("================================")
basic_quality$text
```

**Expected Results:**
- Missing data analysis for Age, Sex, and Grade variables
- Duplicate value detection within each variable
- Data completeness summary
- Recommendations based on quality findings

## Example 2: Enhanced Visual Data Quality Assessment

Using visdat integration for comprehensive visual analysis.

```{r visual-assessment, eval=FALSE}
# Enhanced visual data quality assessment
visual_quality <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,  # Enable visual analysis
  visdat_type = "vis_dat",  # Data overview visualization
  missing_threshold_visual = 5,  # Highlight variables with >5% missing
  export_plots = FALSE
)

# View visual quality results
print("👁️ Visual Data Quality Assessment:")
print("=================================")
print("Text Summary:")
visual_quality$text

print("\nVisual Analysis:")
visual_quality$plot
```

**Visual Analysis Benefits:**
- **Pattern Recognition**: Identify systematic missing data patterns
- **Type Validation**: Verify variable types match expectations
- **Missing Clustering**: Detect non-random missing data mechanisms
- **Quality Overview**: Comprehensive visual summary of data structure

## Example 3: Missing Data Pattern Analysis

Focused analysis of missing data patterns using vis_miss.

```{r missing-patterns, eval=FALSE}
# Missing data pattern analysis
missing_analysis <- dataquality(
  data = histopathology,
  vars = c("Age", "Grade", "TStage", "MeasurementA", "MeasurementB"),
  check_duplicates = FALSE,  # Focus on missing data
  check_missing = TRUE,
  complete_cases_only = TRUE,  # Analyze complete cases
  visual_analysis = TRUE,
  visdat_type = "vis_miss",  # Missing pattern visualization
  missing_threshold_visual = 10,
  export_plots = FALSE
)

# Examine missing data insights
print("🔍 Missing Data Pattern Analysis:")
print("=================================")
missing_analysis$text
missing_analysis$plot
```

**Missing Pattern Insights:**
- **MCAR Detection**: Identify missing completely at random patterns
- **MAR Assessment**: Detect missing at random relationships
- **MNAR Warning**: Flag potentially missing not at random issues
- **Pattern Clustering**: Visualize missing data relationships

## Example 4: Data Type Validation

Using vis_guess for automatic data type detection and validation.

```{r type-validation, eval=FALSE}
# Data type validation analysis
type_validation <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "OverallTime", "Death"),
  check_duplicates = FALSE,
  check_missing = FALSE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,
  visdat_type = "vis_guess",  # Type detection visualization
  missing_threshold_visual = 15,
  export_plots = TRUE  # Enable export for documentation
)

# Review type validation results
print("🔢 Data Type Validation Analysis:")
print("================================")
type_validation$text
type_validation$plot
```

**Type Validation Applications:**
- **Import Verification**: Confirm data imported with correct types
- **Analysis Readiness**: Ensure variables are analysis-ready
- **Error Detection**: Identify unexpected type conversions
- **Documentation**: Create type validation reports

## Example 5: Comprehensive Quality Assessment

Complete data quality assessment using all analysis types.

```{r comprehensive-assessment, eval=FALSE}
# Comprehensive data quality assessment
comprehensive_quality <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI", "OverallTime", "Death"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = TRUE,
  visual_analysis = TRUE,
  visdat_type = "all_visual",  # All visual analyses
  missing_threshold_visual = 5,
  export_plots = TRUE
)

# Review comprehensive results
print("📊 Comprehensive Data Quality Assessment:")
print("=======================================")
print("Quality Summary:")
comprehensive_quality$text

print("\nVisual Analysis Suite:")
comprehensive_quality$plot
```

**Comprehensive Benefits:**
- **Complete Coverage**: All quality dimensions assessed
- **Integrated Analysis**: Statistical and visual assessment combined
- **Publication Ready**: Professional output for research documentation
- **Regulatory Compliance**: Meets clinical research standards

# Advanced Configuration Options

## Visual Analysis Types

### **Data Overview (vis_dat)**
```{r vis-dat-config, eval=FALSE}
# Data overview configuration
overview_config <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage"),
  visual_analysis = TRUE,
  visdat_type = "vis_dat",
  missing_threshold_visual = 10
)
```

**Use Cases:**
- Initial data exploration
- Variable type verification
- Overall data structure assessment
- Missing data overview

### **Missing Patterns (vis_miss)**
```{r vis-miss-config, eval=FALSE}
# Missing pattern configuration
missing_config <- dataquality(
  data = histopathology,
  vars = c("MeasurementA", "MeasurementB", "OverallTime"),
  visual_analysis = TRUE,
  visdat_type = "vis_miss",
  missing_threshold_visual = 5  # Stricter threshold for missing data
)
```

**Use Cases:**
- Missing data mechanism investigation
- Pattern clustering analysis
- Systematic missingness detection
- Imputation planning

### **Type Detection (vis_guess)**
```{r vis-guess-config, eval=FALSE}
# Type detection configuration
type_config <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "Death"),
  visual_analysis = TRUE,
  visdat_type = "vis_guess",
  missing_threshold_visual = 15
)
```

**Use Cases:**
- Data import validation
- Type conversion verification
- Analysis preparation
- Error detection

### **Value Expectations (vis_expect)**
```{r vis-expect-config, eval=FALSE}
# Value expectation configuration
expect_config <- dataquality(
  data = histopathology,
  vars = c("Age", "OverallTime"),
  visual_analysis = TRUE,
  visdat_type = "vis_expect",
  missing_threshold_visual = 10
)
```

**Use Cases:**
- Outlier pattern detection
- Expected value validation
- Data anomaly identification
- Quality threshold assessment

## Missing Data Threshold Configuration

### **Threshold Selection Guidelines**

```{r threshold-guidelines, eval=FALSE}
# Conservative threshold (highlight variables with >2% missing)
conservative_analysis <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  visual_analysis = TRUE,
  visdat_type = "vis_miss",
  missing_threshold_visual = 2  # Very strict
)

# Standard threshold (highlight variables with >10% missing)
standard_analysis <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  visual_analysis = TRUE,
  visdat_type = "vis_miss",
  missing_threshold_visual = 10  # Recommended
)

# Permissive threshold (highlight variables with >25% missing)
permissive_analysis <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  visual_analysis = TRUE,
  visdat_type = "vis_miss",
  missing_threshold_visual = 25  # Lenient
)
```

**Threshold Recommendations:**
- **Primary Endpoints**: 2-5% (strict quality requirements)
- **Secondary Variables**: 10-15% (standard clinical research)
- **Exploratory Analysis**: 20-25% (hypothesis generation)

## Export and Documentation Options

### **Publication-Ready Plots**
```{r export-config, eval=FALSE}
# Export-ready configuration
export_analysis <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,
  visdat_type = "all_visual",
  missing_threshold_visual = 10,
  export_plots = TRUE  # Enable export features
)

# Export information will be included in results
export_analysis$plot
```

**Export Features:**
- High-resolution plot generation ready for manuscripts
- Code examples for reproducing analyses
- Professional formatting for publications
- Integration with visdat export capabilities

# Clinical Research Applications

## Biomarker Data Quality Assessment

### **Precision Medicine Requirements**
```{r biomarker-quality, eval=FALSE}
# Create biomarker test data (simulated for demonstration)
biomarker_data <- data.frame(
  patient_id = paste0("BM", sprintf("%03d", 1:100)),
  her2_expression = c(
    # HER2 negative patients
    rlnorm(80, meanlog = 1.5, sdlog = 0.8),
    # HER2 positive patients  
    rlnorm(20, meanlog = 3.2, sdlog = 0.6)
  ),
  pdl1_percentage = pmax(0, pmin(100, rnorm(100, 25, 20))),  # Bounded 0-100%
  response_status = factor(sample(c("Complete Response", "Partial Response", "Stable Disease", "Progressive Disease"), 
                                 100, replace = TRUE))
)

# Assess biomarker quality
biomarker_quality <- dataquality(
  data = biomarker_data,
  vars = c("her2_expression", "pdl1_percentage", "response_status"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,
  visdat_type = "vis_dat",
  missing_threshold_visual = 5,  # Strict for biomarkers
  export_plots = TRUE
)

print("🧬 Biomarker Data Quality Assessment:")
biomarker_quality$text
biomarker_quality$plot
```

**Biomarker Quality Standards:**
- **Coefficient of Variation**: <20% for most assays
- **Missing Data**: <5% for regulatory submissions
- **Range Validation**: Biologically meaningful bounds
- **Outlier Investigation**: Essential for clinical decisions

## Clinical Trial Data Monitoring

### **Real-Time Quality Monitoring**
```{r clinical-trial-monitoring, eval=FALSE}
# Simulate ongoing clinical trial data
trial_data <- data.frame(
  patient_id = paste0("CT", sprintf("%03d", 1:75)),
  enrollment_date = seq(as.Date("2024-01-01"), as.Date("2024-03-15"), length.out = 75),
  primary_endpoint = c(
    rnorm(60, 2.5, 1.2),  # Most patients
    rep(NA, 10),          # Recent enrollments, not yet assessed
    c(8.5, -2.1, 15.3, 0.1, 9.8)  # Some outliers
  ),
  safety_score = sample(0:4, 75, replace = TRUE, prob = c(0.4, 0.3, 0.2, 0.08, 0.02)),
  treatment_group = factor(sample(c("Treatment", "Control"), 75, replace = TRUE))
)

# Monitor trial data quality
trial_quality <- dataquality(
  data = trial_data,
  vars = c("primary_endpoint", "safety_score", "treatment_group"),
  check_duplicates = FALSE,  # Not applicable for trial data
  check_missing = TRUE,
  complete_cases_only = TRUE,
  visual_analysis = TRUE,
  visdat_type = "vis_miss",  # Focus on missing patterns
  missing_threshold_visual = 10,
  export_plots = FALSE
)

print("🏥 Clinical Trial Data Quality Monitoring:")
trial_quality$text
trial_quality$plot

# Check if quality meets regulatory standards
missing_pct <- 100 * sum(is.na(trial_data$primary_endpoint)) / nrow(trial_data)
if (missing_pct > 10) {
  cat("⚠️  WARNING: Missing data exceeds 10% - investigate data collection procedures\n")
} else {
  cat("✅ PASS: Missing data within acceptable limits\n")
}
```

## Pharmacovigilance Data Quality

### **Adverse Event Reporting Quality**
```{r pharmacovigilance-quality, eval=FALSE}
# Simulate adverse event data
ae_data <- data.frame(
  patient_id = paste0("AE", sprintf("%03d", 1:150)),
  severity_score = sample(1:5, 150, replace = TRUE, prob = c(0.4, 0.3, 0.2, 0.08, 0.02)),
  onset_days = c(
    sample(1:30, 120, replace = TRUE),  # Normal onset times
    rep(NA, 20),                       # Missing onset times
    c(365, 400, -5, 0, 2000, 1500, 750, 800, 1200, 950)  # Problematic values
  ),
  recovery_status = factor(
    sample(c("Recovered", "Recovering", "Not Recovered", "Unknown"), 
           150, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.1))
  )
)

# Assess adverse event data quality
ae_quality <- dataquality(
  data = ae_data,
  vars = c("severity_score", "onset_days", "recovery_status"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,
  visdat_type = "all_visual",  # Comprehensive analysis for safety
  missing_threshold_visual = 15,
  export_plots = TRUE
)

print("⚠️ Pharmacovigilance Data Quality Assessment:")
ae_quality$text
ae_quality$plot
```

# Quality Assurance Best Practices

## Data Quality Validation Workflow

### **Pre-Analysis Quality Check**
```{r pre-analysis-workflow, eval=FALSE}
# Function for standardized pre-analysis quality assessment
perform_preanalysis_quality_check <- function(data, primary_vars, secondary_vars = NULL) {
  
  cat("📋 PRE-ANALYSIS QUALITY ASSESSMENT\n")
  cat("=================================\n")
  
  # Primary variables quality assessment
  primary_quality <- dataquality(
    data = data,
    vars = primary_vars,
    check_duplicates = TRUE,
    check_missing = TRUE,
    complete_cases_only = TRUE,
    visual_analysis = TRUE,
    visdat_type = "all_visual",
    missing_threshold_visual = 5,  # Strict for primary variables
    export_plots = TRUE
  )
  
  cat("PRIMARY VARIABLES ASSESSMENT:\n")
  cat("----------------------------\n")
  primary_quality$text
  
  # Secondary variables if provided
  if (!is.null(secondary_vars)) {
    secondary_quality <- dataquality(
      data = data,
      vars = secondary_vars,
      check_duplicates = TRUE,
      check_missing = TRUE,
      complete_cases_only = FALSE,
      visual_analysis = TRUE,
      visdat_type = "vis_miss",
      missing_threshold_visual = 15,  # More lenient for secondary
      export_plots = FALSE
    )
    
    cat("\nSECONDARY VARIABLES ASSESSMENT:\n")
    cat("------------------------------\n")
    secondary_quality$text
  }
  
  return(list(primary = primary_quality, secondary = secondary_quality))
}

# Example usage
primary_vars <- c("Age", "Sex", "Grade", "TStage")
secondary_vars <- c("LVI", "PNI", "MeasurementA", "MeasurementB")

quality_results <- perform_preanalysis_quality_check(
  data = histopathology,
  primary_vars = primary_vars,
  secondary_vars = secondary_vars
)
```

### **Quality Grading System**

The dataquality function provides implicit quality grading based on analysis results:

**Grade A: Excellent Quality**
- Missing data: 0-5%
- Minimal duplicates
- Appropriate data types
- No systematic patterns

**Grade B: Good Quality**
- Missing data: 5-15%
- Some duplicate patterns
- Minor type issues
- Limited systematic concerns

**Grade C: Concerning Quality**
- Missing data: 15-30%
- Notable duplicate issues
- Type validation problems
- Systematic patterns detected

**Grade D: Poor Quality**
- Missing data: >30%
- Significant quality issues
- Multiple systematic problems
- Requires intervention

## Integration with jamovi Workflow

### **Menu Navigation**
```
jamovi → Exploration → ClinicoPath Descriptives → Data Quality Assessment
```

### **Parameter Configuration in jamovi**
1. **Variables**: Select variables for quality assessment
2. **Duplicates**: Enable duplicate detection analysis
3. **Missing Values**: Enable missing data analysis
4. **Complete Cases**: Enable complete case analysis
5. **Visual Analysis**: Enable visdat integration
6. **Analysis Type**: Choose vis_dat, vis_miss, vis_guess, vis_expect, or all_visual
7. **Missing Threshold**: Set percentage threshold for missing data highlighting
8. **Export Plots**: Enable plot export functionality

### **Interpreting jamovi Results**
- **Instructions Panel**: Comprehensive guidance for using the function
- **Text Results**: Detailed quality assessment summary
- **Visual Results**: Interactive plots showing data quality patterns
- **Export Information**: Guidelines for saving and sharing results

# Troubleshooting and Common Issues

## visdat Package Dependencies

### **Installation Requirements**
```{r visdat-dependencies, eval=FALSE}
# Check visdat availability
if (!requireNamespace("visdat", quietly = TRUE)) {
  cat("📦 Installing visdat package for enhanced visual analysis...\n")
  install.packages("visdat")
}

# Verify installation
if (requireNamespace("visdat", quietly = TRUE)) {
  cat("✅ visdat package available - full visual analysis enabled\n")
} else {
  cat("⚠️  visdat package not available - visual analysis will show informative messages\n")
  cat("   Install with: install.packages('visdat')\n")
}
```

### **Alternative Analysis Without visdat**
```{r no-visdat-analysis, eval=FALSE}
# Data quality assessment without visual analysis
no_visual_quality <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = FALSE,  # Disable visual analysis
  visdat_type = "vis_dat",
  missing_threshold_visual = 10,
  export_plots = FALSE
)

print("📊 Data Quality Assessment (No Visual Analysis):")
no_visual_quality$text
```

## Performance Optimization

### **Large Dataset Handling**
```{r large-dataset-tips, eval=FALSE}
# Tips for large datasets (>1000 rows, >50 variables)

# 1. Subset variables for initial assessment
key_vars <- c("Age", "Sex", "Grade", "TStage", "Death")
subset_quality <- dataquality(
  data = histopathology,
  vars = key_vars,  # Focus on key variables first
  check_duplicates = FALSE,  # Skip duplicates for speed
  check_missing = TRUE,
  complete_cases_only = FALSE,
  visual_analysis = TRUE,
  visdat_type = "vis_dat",  # Use faster vis_dat over all_visual
  missing_threshold_visual = 10,
  export_plots = FALSE
)

# 2. Use sampling for very large datasets
if (nrow(histopathology) > 1000) {
  sample_data <- histopathology[sample(nrow(histopathology), 500), ]
  sample_quality <- dataquality(
    data = sample_data,
    vars = c("Age", "Sex", "Grade"),
    visual_analysis = TRUE,
    visdat_type = "all_visual",
    missing_threshold_visual = 10
  )
}
```

### **Memory Management**
```{r memory-management, eval=FALSE}
# Monitor memory usage during quality assessment
cat("Memory usage before analysis:", format(object.size(histopathology), units = "MB"), "\n")

# Perform analysis
quality_result <- dataquality(
  data = histopathology,
  vars = c("Age", "Sex", "Grade"),
  visual_analysis = TRUE,
  visdat_type = "vis_dat"
)

# Clean up if needed
# rm(quality_result)
# gc()  # Garbage collection
```

# Advanced Integration Examples

## Integration with Other ClinicoPath Functions

### **Quality Assessment Before crosstable Analysis**
```{r integration-crosstable, eval=FALSE}
# Step 1: Assess data quality
quality_check <- dataquality(
  data = histopathology,
  vars = c("Sex", "Grade", "Group"),
  check_duplicates = TRUE,
  check_missing = TRUE,
  visual_analysis = TRUE,
  visdat_type = "vis_miss"
)

print("Quality assessment before crosstable analysis:")
quality_check$text

# Step 2: Proceed with crosstable if quality is acceptable
if (# quality meets standards - implement quality check logic here
    TRUE) {  # Placeholder condition
  
  crosstable_result <- crosstable(
    data = histopathology,
    vars = c("Sex", "Grade"),
    group = "Group"
  )
  
  cat("✅ Quality check passed - proceeding with crosstable analysis\n")
}
```

### **Quality Assessment Before Survival Analysis**
```{r integration-survival, eval=FALSE}
# Assess survival analysis variables
survival_quality <- dataquality(
  data = histopathology,
  vars = c("OverallTime", "Death", "Grade", "TStage"),
  check_duplicates = FALSE,  # Not applicable for survival data
  check_missing = TRUE,
  complete_cases_only = TRUE,  # Critical for survival analysis
  visual_analysis = TRUE,
  visdat_type = "vis_dat",
  missing_threshold_visual = 5  # Strict for survival endpoints
)

print("Survival analysis data quality:")
survival_quality$text
```

## Automated Quality Reporting

### **Generate Quality Report**
```{r automated-reporting, eval=FALSE}
# Function to generate automated quality report
generate_quality_report <- function(data, variables, report_title = "Data Quality Report") {
  
  cat("📊", report_title, "\n")
  cat(rep("=", nchar(report_title) + 4), "\n")
  cat("Generated on:", Sys.Date(), "\n")
  cat("Dataset:", deparse(substitute(data)), "\n")
  cat("Variables analyzed:", length(variables), "\n")
  cat("Observations:", nrow(data), "\n\n")
  
  # Comprehensive quality assessment
  quality_result <- dataquality(
    data = data,
    vars = variables,
    check_duplicates = TRUE,
    check_missing = TRUE,
    complete_cases_only = TRUE,
    visual_analysis = TRUE,
    visdat_type = "all_visual",
    missing_threshold_visual = 10,
    export_plots = TRUE
  )
  
  cat("QUALITY ASSESSMENT RESULTS:\n")
  cat("-------------------------\n")
  quality_result$text
  
  cat("\nVISUAL ANALYSIS:\n")
  cat("---------------\n")
  quality_result$plot
  
  return(quality_result)
}

# Generate report for histopathology data
quality_report <- generate_quality_report(
  data = histopathology,
  variables = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI"),
  report_title = "Histopathology Data Quality Report"
)
```

# Conclusion

The ClinicoPath `dataquality` function represents a significant advancement in clinical research data validation tools. By integrating sophisticated visual analysis capabilities with traditional statistical quality assessment, it provides researchers with unprecedented insight into data quality issues.

## Key Benefits

1. **Comprehensive Analysis**: Statistical and visual assessment in one tool
2. **Clinical Focus**: Designed specifically for medical research workflows
3. **Advanced Visualization**: First jamovi integration with visdat package
4. **Flexible Configuration**: Customizable analysis options and thresholds
5. **Export Capabilities**: Publication-ready outputs for research documentation
6. **Regulatory Compliance**: Supports ICH E6 and FDA data quality standards

## Success Factors

1. **Regular Assessment**: Integrate quality checks throughout data lifecycle
2. **Visual Interpretation**: Leverage visual patterns for deeper insights
3. **Threshold Customization**: Adjust quality standards for research context
4. **Documentation Standards**: Maintain comprehensive quality audit trails
5. **Continuous Monitoring**: Implement ongoing quality surveillance

## Future Enhancements

- **Machine Learning Integration**: Automated pattern recognition and anomaly detection
- **Real-Time Monitoring**: Continuous quality assessment during data collection
- **Predictive Quality Models**: Early warning systems for quality deterioration
- **Advanced Export Options**: Enhanced publication-ready output formats
- **Integration Expansion**: Deeper connections with other ClinicoPath modules

The `dataquality` function establishes a new standard for clinical research data validation, combining the power of statistical analysis with the insight of visual exploration to ensure the highest quality foundation for medical research and evidence-based healthcare decisions.

---

*This comprehensive guide demonstrates the full capabilities of enhanced data quality assessment in the ClinicoPath dataquality function, providing users with the theoretical foundation, practical skills, and professional standards needed for effective clinical research data management and quality assurance.*