---
title: "Interrater Reliability & Agreement Analysis - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interrater Reliability & Agreement Analysis - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Load the package
devtools::load_all("../")

# Base path for data files
data_path <- "../data-raw/non-rda/"
```

# Interrater Reliability & Agreement Analysis

## Overview

The **Agreement** module (`agreement`) provides a comprehensive suite of interrater
reliability and agreement statistics for clinicopathological research. It supports
categorical, ordinal, and continuous data with 2 or more raters, and includes
advanced features such as hierarchical kappa, mixed-effects comparisons,
bootstrap confidence intervals, and built-in sample size calculators.

All examples below use the bundled test datasets located in `data-raw/non-rda/`.
In jamovi, open the corresponding `.omv` file. From R, read the `.csv` directly.

---

## Datasets Used in This Guide

| Dataset | Type | Raters | Description |
|---------|------|--------|-------------|
| `agreement_pathology` | Categorical | 2 | Surgical pathology diagnoses (Benign/Low-grade/High-grade) |
| `agreement_binary` | Binary | 2 | Positive/Negative classification |
| `agreement_ordinal` | Ordinal | 2 | Tumor grading (Grade 1-3) |
| `agreement_continuous` | Continuous | 2 | Continuous measurements |
| `agreement_threeRater` | Categorical | 3 | Three-rater panel |
| `agreement_multiRater` | Categorical | 5 | Five pathologists of varying experience |
| `agreement_hierarchical` | Categorical | 6 | Multi-center (3 hospitals, 2 raters each) |
| `agreement_testRetest` | Categorical | 6 | Test-retest (3 raters x 2 timepoints) |
| `agreement_missing` | Categorical | 3 | Dataset with missing values |
| `agreement_paired_comparison` | Categorical | 4 | Manual vs AI rater comparison |
| `agreement_mixed_effects` | Continuous | 2 | Pre/Post training condition comparison |
| `agreement_heatmap_test` | Ordinal | 3 | HER2 scoring for heatmap visualization |
| `agreement_perfect` | Categorical | 2 | Perfect agreement (edge case) |
| `agreement_poor` | Categorical | 2 | Poor agreement |
| `agreement_small` | Binary | 2 | Small sample (n=30) |
| `pathology_agreement_main` | Continuous | 3 | Ki-67 scoring (HALO, Aiforia, Manual) |
| `pathology_agreement_multimethod` | Continuous | 4 | Ki-67 multi-platform |
| `pathology_agreement_ai` | Continuous | 2 | AI vs pathologist Ki-67 |
| `pathology_agreement_missing` | Continuous | 3 | Continuous data with NAs |
| `comprehensive_agreement_data` | Categorical | 4 | Multi-specialty with metadata |
| `digital_pathology_validation` | Continuous | 2 | Digital pathology AI validation |

---

## 1. Categorical Agreement (Default Kappa)

The default analysis computes Cohen's kappa (2 raters) or Fleiss' kappa (3+ raters)
with percent agreement.

### Basic two-rater kappa

```{r categorical-basic}
# Dataset: agreement_pathology
# Raters: Pathologist1, Pathologist2
# Diagnoses: Benign, Low-grade malignant, High-grade malignant

agreement_pathology <- read.csv(paste0(data_path, "agreement_pathology.csv"))

agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2)
)
```

### With frequency table and additional statistics

```{r categorical-full}
agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    sft = TRUE,                     # Show contingency (frequency) table
    pabak = TRUE,                   # Prevalence-Adjusted Bias-Adjusted Kappa
    gwet = TRUE,                    # Gwet's AC1/AC2 (stable with prevalence imbalance)
    gwetWeights = "unweighted",     # or "linear", "quadratic"
    specificAgreement = TRUE,       # Category-specific agreement (PSA/NSA)
    specificAllCategories = TRUE,   # Show for all categories
    specificConfidenceIntervals = TRUE,
    raterBias = TRUE,               # McNemar test for systematic rater bias
    bhapkar = TRUE,                 # Bhapkar test (multivariate extension)
    stuartMaxwell = TRUE,           # Stuart-Maxwell marginal homogeneity test
    agreementHeatmap = TRUE,        # Visual heatmap of cross-tabulation
    heatmapColorScheme = "bluered", # Color scheme (bluered/viridis/heat/grayscale)
    heatmapShowPercentages = TRUE,
    heatmapShowCounts = TRUE,
    heatmapAnnotationSize = 3.5
)
```

### With subgroup analysis

```{r categorical-subgroup}
agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    agreementBySubgroup = TRUE,
    subgroupVariable = specimen_type,   # Compare kappa across specimen types
    subgroupForestPlot = TRUE,          # Forest plot of subgroup kappas
    subgroupMinCases = 10               # Minimum cases per subgroup
)
```

---

## 2. Binary Agreement

```{r binary}
# Dataset: agreement_binary
# Raters: PathologistX, PathologistY (Positive/Negative)

agreement_binary <- read.csv(paste0(data_path, "agreement_binary.csv"))

agreement(
    data = agreement_binary,
    conditionBVars = NULL,
    vars = vars(PathologistX, PathologistY),
    specificAgreement = TRUE,
    specificPositiveCategory = "Positive",  # Category for PSA/NSA
    specificConfidenceIntervals = TRUE,
    pabak = TRUE,
    agreementBySubgroup = TRUE,
    subgroupVariable = specimen_quality
)
```

---

## 3. Ordinal / Weighted Kappa

For ordinal data (e.g., tumor grades), weighted kappa accounts for the magnitude
of disagreement. A Grade 1 vs Grade 3 mismatch is penalized more heavily than
Grade 1 vs Grade 2.

```{r ordinal-linear}
# Dataset: agreement_ordinal
# Raters: PathologistA, PathologistB (Grade 1/2/3)

agreement_ordinal <- read.csv(paste0(data_path, "agreement_ordinal.csv"))

# Convert to ordered factors (required for weighted kappa & rank-based methods)
grade_levels <- c("Grade 1", "Grade 2", "Grade 3", "Grade 4")
agreement_ordinal$PathologistA <- factor(agreement_ordinal$PathologistA,
    levels = grade_levels, ordered = TRUE)
agreement_ordinal$PathologistB <- factor(agreement_ordinal$PathologistB,
    levels = grade_levels, ordered = TRUE)

# Linear weights
agreement(
    data = agreement_ordinal,
    conditionBVars = NULL,
    vars = vars(PathologistA, PathologistB),
    wght = "equal",              # Linear weights
    showLevelInfo = TRUE,        # Show category-level details
    kendallW = TRUE,             # Kendall's W (coefficient of concordance)
    agreementBySubgroup = TRUE,
    subgroupVariable = tumor_site
)
```

```{r ordinal-quadratic}
# Quadratic weights (more forgiving of near-misses)
agreement(
    data = agreement_ordinal,
    conditionBVars = NULL,
    vars = vars(PathologistA, PathologistB),
    wght = "squared"
)
```

---

## 4. Three or More Raters (Fleiss' / Light's Kappa)

When 3+ raters are present, the module automatically computes Fleiss' kappa.
Light's kappa is the average of all pairwise Cohen's kappas.

```{r three-raters}
# Dataset: agreement_threeRater
# Raters: Rater1, Rater2, Rater3

agreement_threeRater <- read.csv(paste0(data_path, "agreement_threeRater.csv"))

agreement(
    data = agreement_threeRater,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2, Rater3),
    exct = TRUE,                        # Exact kappa (small samples)
    lightKappa = TRUE,                  # Light's Kappa (mean pairwise)
    finn = TRUE,                        # Finn coefficient
    finnLevels = 3,                     # Number of response categories
    finnModel = "oneway",               # "oneway" or "twoway"
    multiAnnotatorConcordance = TRUE,   # Multi-annotator concordance (F1)
    predictionColumn = 1,               # Which rater is the "prediction"
    agreementBySubgroup = TRUE,
    subgroupVariable = tissue_site
)
```

### Five raters with clustering and profiles

```{r multi-rater}
# Dataset: agreement_multiRater
# Raters: SeniorPath1, SeniorPath2, MidLevelPath, JuniorPath1, JuniorPath2

agreement_multiRater <- read.csv(paste0(data_path, "agreement_multiRater.csv"))

agreement(
    data = agreement_multiRater,
    conditionBVars = NULL,
    vars = vars(SeniorPath1, SeniorPath2, MidLevelPath, JuniorPath1, JuniorPath2),
    lightKappa = TRUE,
    # Rater clustering
    raterClustering = TRUE,
    clusterMethod = "hierarchical",     # or "kmeans"
    clusterDistance = "correlation",     # or "euclidean", "manhattan"
    clusterLinkage = "average",         # or "complete", "ward.D2"
    nClusters = 2,
    showDendrogram = TRUE,
    showClusterHeatmap = TRUE,
    # Case clustering
    caseClustering = TRUE,
    caseClusterMethod = "hierarchical",
    caseClusterDistance = "correlation",
    caseClusterLinkage = "average",
    nCaseClusters = 2,
    showCaseDendrogram = TRUE,
    showCaseClusterHeatmap = TRUE,
    # Rater profiles
    raterProfiles = TRUE,
    raterProfileType = "boxplot",       # "boxplot", "violin", or "barplot"
    raterProfileShowPoints = TRUE,
    # Subgroup
    agreementBySubgroup = TRUE,
    subgroupVariable = difficulty,
    subgroupForestPlot = TRUE,
    subgroupMinCases = 10
)
```

---

## 5. Continuous Data (ICC / Bland-Altman / CCC)

For continuous measurements, kappa is not appropriate. The module automatically
detects continuous data and suggests ICC, Bland-Altman, and CCC.

### Two continuous raters

```{r continuous-basic}
# Dataset: agreement_continuous
# Raters: MeasurementA, MeasurementB

agreement_continuous <- read.csv(paste0(data_path, "agreement_continuous.csv"))

agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    # Intraclass Correlation Coefficient
    icc = TRUE,
    iccType = "icc21",          # ICC(2,1) two-way random, single measures
    # Bland-Altman analysis
    blandAltmanPlot = TRUE,
    baConfidenceLevel = 0.95,
    proportionalBias = TRUE,    # Test for proportional bias
    # Lin's Concordance Correlation Coefficient
    linCCC = TRUE,
    # Total Deviation Index
    tdi = TRUE,
    tdiCoverage = 90,           # Percentage coverage
    tdiLimit = 10,              # Acceptable limit
    # Additional methods
    meanPearson = TRUE,         # Mean Pearson correlation
    meanSpearman = TRUE,        # Mean Spearman correlation
    robinsonA = TRUE,           # Robinson's A
    maxwellRE = TRUE,           # Maxwell's Random Effect model
    iota = TRUE,                # Iota index
    iotaStandardize = TRUE,     # Standardize before computing
    # Subgroup
    agreementBySubgroup = TRUE,
    subgroupVariable = tumor_type
)
```

### Cycling through all ICC types

```{r icc-types}
# ICC(1,1): One-way random, single measures
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc11"
)

# ICC(2,1): Two-way random, single measures (most common)
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc21"
)

# ICC(3,1): Two-way mixed, single measures
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc31"
)

# ICC(1,k): One-way random, average measures
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc1k"
)

# ICC(2,k): Two-way random, average measures
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc2k"
)

# ICC(3,k): Two-way mixed, average measures
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    icc = TRUE, iccType = "icc3k"
)
```

### Three continuous raters (Ki-67)

```{r continuous-three}
# Dataset: pathology_agreement_multimethod (using 3 of 4 raters)
# Raters: Ki67_HALO, Ki67_Aiforia, Ki67_ImageJ

pathology_multi <- read.csv(paste0(data_path, "pathology_agreement_multimethod.csv"))

agreement(
    data = pathology_multi,
    conditionBVars = NULL,
    vars = vars(Ki67_HALO, Ki67_Aiforia, Ki67_ImageJ),
    icc = TRUE,
    meanPearson = TRUE,
    raterProfiles = TRUE,
    kripp = TRUE,
    krippMethod = "interval"    # For continuous data
)
```

### Four continuous raters with bootstrap

```{r continuous-four}
# Dataset: pathology_agreement_multimethod
# Raters: Ki67_HALO, Ki67_Aiforia, Ki67_ImageJ, Ki67_Manual

pathology_multi <- read.csv(paste0(data_path, "pathology_agreement_multimethod.csv"))

agreement(
    data = pathology_multi,
    conditionBVars = NULL,
    vars = vars(Ki67_HALO, Ki67_Aiforia, Ki67_ImageJ, Ki67_Manual),
    icc = TRUE,
    raterClustering = TRUE,
    caseClustering = TRUE,
    bootstrapCI = TRUE,
    nBoot = 200
)
```

---

## 6. Krippendorff's Alpha

Krippendorff's alpha handles missing data natively and supports all measurement
levels. It is the recommended general-purpose agreement statistic.

### Categorical with missing data

```{r kripp-categorical}
# Dataset: agreement_missing
# Raters: Rater1, Rater2, Rater3 (with NAs)

agreement_missing <- read.csv(paste0(data_path, "agreement_missing.csv"))

agreement(
    data = agreement_missing,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2, Rater3),
    kripp = TRUE,
    krippMethod = "nominal",    # "nominal", "ordinal", "interval", "ratio"
    bootstrap = TRUE,           # Bootstrap CI for Krippendorff's alpha
    bootstrapCI = TRUE,
    nBoot = 200
)
```

### Continuous with missing data

```{r kripp-continuous}
# Dataset: pathology_agreement_missing
# Raters: HALO_Score, Aiforia_Score, ImageJ_Score

pathology_missing <- read.csv(paste0(data_path, "pathology_agreement_missing.csv"))

agreement(
    data = pathology_missing,
    conditionBVars = NULL,
    vars = vars(HALO_Score, Aiforia_Score, ImageJ_Score),
    kripp = TRUE,
    krippMethod = "interval",   # Continuous measurement level
    icc = TRUE                  # Compare with ICC
)
```

---

## 7. Gwet's AC and PABAK

Gwet's AC1/AC2 is more stable than kappa when prevalence is skewed. PABAK
adjusts for both prevalence and bias.

```{r gwet-pabak}
# Dataset: agreement_poor (low kappa due to prevalence, not poor raters)

agreement_poor <- read.csv(paste0(data_path, "agreement_poor.csv"))

agreement(
    data = agreement_poor,
    conditionBVars = NULL,
    vars = vars(PathologistA, PathologistB),
    gwet = TRUE,
    gwetWeights = "unweighted",     # "unweighted", "linear", "quadratic"
    pabak = TRUE,
    agreementBySubgroup = TRUE,
    subgroupVariable = difficulty_level
)
```

---

## 8. Hierarchical / Multi-Center Kappa

For multi-center studies where raters are nested within institutions, hierarchical
kappa decomposes agreement into within-center and between-center components.

**Note:** Hierarchical mixed-effects decomposition (variance decomposition,
hierarchical ICC) requires **continuous** numeric ratings. For categorical data,
use standard kappa or cluster-specific kappa analyses.

### Continuous data: Full hierarchical decomposition

```{r hierarchical-continuous}
# Dataset: digital_pathology_validation (continuous Ki-67 scores)
# Raters: Ki67_Manual, Ki67_AI_Assisted
# Cluster variable: Institution (4 centers)

digital_path_hier <- read.csv(paste0(data_path, "digital_pathology_validation.csv"))

agreement(
    data = digital_path_hier,
    conditionBVars = NULL,
    vars = vars(Ki67_Manual, Ki67_AI_Assisted),
    hierarchicalKappa = TRUE,
    clusterVariable = Institution,           # Nesting variable
    clusterSpecificKappa = TRUE,             # Kappa per center
    varianceDecomposition = TRUE,            # Variance components
    testClusterHomogeneity = TRUE,           # Test if agreement differs across centers
    shrinkageEstimates = TRUE,               # Empirical Bayes shrinkage
    clusterRankings = TRUE,                  # Rank centers by agreement
    iccHierarchical = TRUE,                  # Hierarchical ICC
    multipleTestCorrection = "bonferroni"    # "none", "bonferroni", "bh", "holm"
)
```

### Categorical data: Pairwise and cluster-specific kappa

```{r hierarchical-categorical}
# Dataset: agreement_hierarchical (categorical diagnoses)
# Raters: HospitalA_Rater1/2, HospitalB_Rater1/2, HospitalC_Rater1/2

agreement_hierarchical <- read.csv(paste0(data_path, "agreement_hierarchical.csv"))

agreement(
    data = agreement_hierarchical,
    conditionBVars = NULL,
    vars = vars(
        HospitalA_Rater1, HospitalA_Rater2,
        HospitalB_Rater1, HospitalB_Rater2,
        HospitalC_Rater1, HospitalC_Rater2
    ),
    pairwiseKappa = TRUE,
    referenceRater = HospitalA_Rater1,
    rankRaters = TRUE
)
```

### Pairwise kappa with reference rater

```{r pairwise}
# Dataset: comprehensive_agreement_data
# Raters: Rater_1, Rater_2, Rater_3, Rater_4

comprehensive <- read.csv(paste0(data_path, "comprehensive_agreement_data.csv"))

agreement(
    data = comprehensive,
    conditionBVars = NULL,
    vars = vars(Rater_1, Rater_2, Rater_3, Rater_4),
    pairwiseKappa = TRUE,
    referenceRater = Rater_1,       # Compare all raters against this one
    rankRaters = TRUE,              # Rank raters by agreement with reference
    agreementBySubgroup = TRUE,
    subgroupVariable = Difficulty
)
```

---

## 9. Test-Retest / Inter-Intra Rater Reliability

Separates inter-rater from intra-rater reliability using time-point suffixes
in variable names (e.g., `Rater1_T1`, `Rater1_T2`).

```{r test-retest}
# Dataset: agreement_testRetest
# Raters: Rater1_T1, Rater1_T2, Rater2_T1, Rater2_T2, Rater3_T1, Rater3_T2

agreement_testRetest <- read.csv(paste0(data_path, "agreement_testRetest.csv"))

agreement(
    data = agreement_testRetest,
    conditionBVars = NULL,
    vars = vars(Rater1_T1, Rater1_T2, Rater2_T1, Rater2_T2, Rater3_T1, Rater3_T2),
    interIntraRater = TRUE,
    interIntraSeparator = "_",      # Character separating rater from timepoint
    finn = TRUE,
    finnLevels = 3
)
```

---

## 10. Paired Agreement Comparison (Manual vs AI)

Compare agreement under two conditions (e.g., manual vs AI-assisted) using a
bootstrap test of kappa difference.

```{r paired}
# Dataset: agreement_paired_comparison
# Condition A (vars): Manual_Rater1, Manual_Rater2
# Condition B (conditionBVars): AI_Rater1, AI_Rater2

agreement_paired <- read.csv(paste0(data_path, "agreement_paired_comparison.csv"))

agreement(
    data = agreement_paired,
    vars = vars(Manual_Rater1, Manual_Rater2),
    conditionBVars = vars(AI_Rater1, AI_Rater2),
    pairedAgreementTest = TRUE,
    pairedBootN = 200,              # Number of bootstrap replicates
    agreementBySubgroup = TRUE,
    subgroupVariable = tumor_type
)
```

---

## 11. Mixed-Effects Condition Comparison

Compare agreement between conditions (e.g., before vs after training) using
mixed-effects models. Requires a condition variable in the dataset.

```{r mixed-effects}
# Dataset: agreement_mixed_effects
# Raters: Rater1, Rater2
# Condition: condition (Pre_Training / Post_Training)

agreement_mixed <- read.csv(paste0(data_path, "agreement_mixed_effects.csv"))

agreement(
    data = agreement_mixed,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2),
    mixedEffectsComparison = TRUE,
    conditionVariable = condition,
    icc = TRUE,
    multipleTestCorrection = "BH"   # Benjamini-Hochberg
)
```

### Digital pathology validation

```{r mixed-effects-digital}
# Dataset: digital_pathology_validation
# Raters: Ki67_Manual, Ki67_AI_Assisted
# Condition: Pathologist_Experience

digital_path <- read.csv(paste0(data_path, "digital_pathology_validation.csv"))

agreement(
    data = digital_path,
    conditionBVars = NULL,
    vars = vars(Ki67_Manual, Ki67_AI_Assisted),
    mixedEffectsComparison = TRUE,
    conditionVariable = Pathologist_Experience,
    blandAltmanPlot = TRUE,
    linCCC = TRUE,
    agreementBySubgroup = TRUE,
    subgroupVariable = Tumor_Type
)
```

---

## 12. AI Validation

```{r ai-validation}
# Dataset: pathology_agreement_ai
# Raters: Ki67_AI, Ki67_Pathologist

pathology_ai <- read.csv(paste0(data_path, "pathology_agreement_ai.csv"))

agreement(
    data = pathology_ai,
    conditionBVars = NULL,
    vars = vars(Ki67_AI, Ki67_Pathologist),
    blandAltmanPlot = TRUE,
    linCCC = TRUE,
    icc = TRUE,
    agreementBySubgroup = TRUE,
    subgroupVariable = TumorType
)
```

---

## 13. Agreement Heatmap

Dedicated visualization of cross-tabulated agreement patterns.

```{r heatmap}
# Dataset: agreement_heatmap_test
# Raters: Scorer_A, Scorer_B, Scorer_C (HER2 scores: 0, 1+, 2+, 3+)

agreement_heatmap <- read.csv(paste0(data_path, "agreement_heatmap_test.csv"))

# Convert HER2 scores to ordered factors
her2_levels <- c("0", "1+", "2+", "3+")
agreement_heatmap$Scorer_A <- factor(agreement_heatmap$Scorer_A,
    levels = her2_levels, ordered = TRUE)
agreement_heatmap$Scorer_B <- factor(agreement_heatmap$Scorer_B,
    levels = her2_levels, ordered = TRUE)
agreement_heatmap$Scorer_C <- factor(agreement_heatmap$Scorer_C,
    levels = her2_levels, ordered = TRUE)

agreement(
    data = agreement_heatmap,
    conditionBVars = NULL,
    vars = vars(Scorer_A, Scorer_B, Scorer_C),
    wght = "equal",                     # Linear weights for ordinal HER2
    sft = TRUE,                         # Contingency table
    agreementHeatmap = TRUE,
    heatmapColorScheme = "bluered",     # "bluered", "viridis", "heat", "grayscale"
    heatmapShowPercentages = TRUE,
    heatmapShowCounts = TRUE,
    heatmapAnnotationSize = 3.5
)
```

---

## 14. Confusion Matrix

```{r confusion-matrix-none}
# Dataset: agreement_pathology

agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    confusionMatrix = TRUE,
    confusionNormalize = "none"     # "none", "row", "column"
)
```

```{r confusion-matrix-row}
# Row-normalized (sensitivity per category)
agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    confusionMatrix = TRUE,
    confusionNormalize = "row"
)
```

---

## 15. Bootstrap Confidence Intervals

BCa (bias-corrected accelerated) bootstrap confidence intervals for all
agreement statistics.

```{r bootstrap}
# Dataset: agreement_small (n=30)

agreement_small <- read.csv(paste0(data_path, "agreement_small.csv"))

agreement(
    data = agreement_small,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2),
    bootstrapCI = TRUE,
    nBoot = 200,                    # Number of bootstrap replicates (50-5000)
    confLevel = 0.95                # Confidence level
)
```

---

## 16. Multi-Annotator Concordance (F1)

Evaluates whether a prediction (one rater column) matches the consensus of the
other annotators.

```{r concordance-f1}
# Dataset: agreement_threeRater

agreement(
    data = agreement_threeRater,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2, Rater3),
    multiAnnotatorConcordance = TRUE,
    predictionColumn = 1            # Which rater is the "prediction" (1-indexed)
)
```

---

## 17. Specific Agreement (Positive/Negative)

Category-specific agreement rates, important when overall kappa is misleading
due to prevalence effects.

```{r specific-agreement}
# Dataset: agreement_binary

agreement(
    data = agreement_binary,
    conditionBVars = NULL,
    vars = vars(PathologistX, PathologistY),
    specificAgreement = TRUE,
    specificPositiveCategory = "Positive",  # Target category for PSA/NSA
    specificAllCategories = TRUE,           # Show all categories
    specificConfidenceIntervals = TRUE      # Wilson score CIs
)
```

---

## 18. Computed Variables

The module can add new computed columns to the dataset.

### Consensus variable

```{r consensus}
# Dataset: agreement_threeRater

agreement(
    data = agreement_threeRater,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2, Rater3),
    consensusName = "consensus_rating",
    consensusRule = "majority",     # "majority", "supermajority", "unanimous"
    tieBreaker = "exclude"          # "exclude", "first", "lowest", "highest"
)
```

### Level of Agreement variable

```{r loa-variable}
# Dataset: agreement_pathology

agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    loaVariable = TRUE,
    detailLevel = "detailed",       # "simple" or "detailed"
    loaThresholds = "custom",       # "custom", "quartiles", "tertiles"
    loaHighThreshold = 75,
    loaLowThreshold = 56,
    loaVariableName = "agreement_level",
    showLoaTable = TRUE
)
```

---

## 19. Edge Cases

### Perfect agreement

```{r perfect}
agreement_perfect <- read.csv(paste0(data_path, "agreement_perfect.csv"))

agreement(
    data = agreement_perfect,
    conditionBVars = NULL,
    vars = vars(RaterA, RaterB)
)
# Expect: kappa = 1.0, 100% agreement
```

### Poor agreement with Gwet's AC

```{r poor-gwet}
agreement(
    data = agreement_poor,
    conditionBVars = NULL,
    vars = vars(PathologistA, PathologistB),
    gwet = TRUE      # Gwet's AC is stable even when kappa is paradoxically low
)
```

### Small sample with bootstrap

```{r small-bootstrap}
agreement(
    data = agreement_small,
    conditionBVars = NULL,
    vars = vars(Rater1, Rater2),
    bootstrapCI = TRUE,
    nBoot = 200
)
```

---

## 20. Sample Size Calculator

The sample size calculator works independently of data -- it only requires the
analysis options to be set.

```{r sample-size-kappa}
# Kappa-based sample size
agreement(
    data = agreement_pathology,         # Any dataset (not used for calculation)
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    agreementSampleSize = TRUE,
    ssMetric = "kappa",                 # "kappa", "fleiss", "icc"
    ssKappaNull = 0.4,                  # Null hypothesis kappa
    ssKappaAlt = 0.7,                   # Alternative hypothesis kappa
    ssNRaters = 2,                      # Number of raters
    ssNCategories = 4,                  # Number of categories
    ssAlpha = 0.05,                     # Significance level
    ssPower = 0.80                      # Desired power
)
```

```{r sample-size-icc}
# ICC-based sample size
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    agreementSampleSize = TRUE,
    ssMetric = "icc",
    ssKappaNull = 0.5,
    ssKappaAlt = 0.8,
    ssNRaters = 2,
    ssAlpha = 0.05,
    ssPower = 0.80
)
```

---

## 21. Display Options

### Confidence level

```{r conf-level}
# Change from default 95% to 90%
agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    confLevel = 0.90       # Applies to all CI calculations
)
```

---

## 22. Complete Example: All Categorical Options

```{r kitchen-sink-categorical}
agreement(
    data = agreement_pathology,
    conditionBVars = NULL,
    vars = vars(Pathologist1, Pathologist2),
    # Display
    sft = TRUE,

    confLevel = 0.95,
    # Weighted kappa
    wght = "unweighted",
    showLevelInfo = TRUE,
    # Additional statistics
    kripp = TRUE,
    krippMethod = "nominal",
    bootstrap = TRUE,
    gwet = TRUE,
    gwetWeights = "unweighted",
    pabak = TRUE,
    # Specific agreement
    specificAgreement = TRUE,
    specificAllCategories = TRUE,
    specificConfidenceIntervals = TRUE,
    # Bias tests
    raterBias = TRUE,
    bhapkar = TRUE,
    stuartMaxwell = TRUE,
    # Confusion matrix
    confusionMatrix = TRUE,
    confusionNormalize = "none",
    # Visualization
    agreementHeatmap = TRUE,
    # Bootstrap
    bootstrapCI = TRUE,
    nBoot = 200,
    # Subgroup
    agreementBySubgroup = TRUE,
    subgroupVariable = specimen_type,
    subgroupForestPlot = TRUE,
    subgroupMinCases = 10,
    # Computed variables
    consensusName = "consensus",
    consensusRule = "majority",
    tieBreaker = "exclude",
    loaVariable = TRUE,
    detailLevel = "detailed",
    loaThresholds = "custom",
    loaHighThreshold = 75,
    loaLowThreshold = 56,
    loaVariableName = "agreement_level",
    showLoaTable = TRUE,
    # Sample size
    agreementSampleSize = TRUE,
    ssMetric = "kappa",
    ssKappaNull = 0.4,
    ssKappaAlt = 0.7,
    ssNRaters = 2,
    ssNCategories = 3,
    ssAlpha = 0.05,
    ssPower = 0.80
)
```

---

## 24. Complete Example: All Continuous Options

```{r kitchen-sink-continuous}
agreement(
    data = agreement_continuous,
    conditionBVars = NULL,
    vars = vars(MeasurementA, MeasurementB),
    confLevel = 0.95,
    # ICC
    icc = TRUE,
    iccType = "icc21",
    # Concordance
    linCCC = TRUE,
    # Bland-Altman
    blandAltmanPlot = TRUE,
    baConfidenceLevel = 0.95,
    proportionalBias = TRUE,
    # TDI
    tdi = TRUE,
    tdiCoverage = 90,
    tdiLimit = 10,
    # Additional statistics
    meanPearson = TRUE,
    maxwellRE = TRUE,
    iota = TRUE,
    iotaStandardize = TRUE,
    kripp = TRUE,
    krippMethod = "interval",
    # Bootstrap
    bootstrapCI = TRUE,
    nBoot = 200,
    # Rater profiles
    raterProfiles = TRUE,
    raterProfileType = "violin",
    raterProfileShowPoints = TRUE,
    # Subgroup
    agreementBySubgroup = TRUE,
    subgroupVariable = tumor_type,
    subgroupForestPlot = TRUE,
    # Sample size
    agreementSampleSize = TRUE,
    ssMetric = "icc",
    ssKappaNull = 0.5,
    ssKappaAlt = 0.8,
    ssNRaters = 2,
    ssAlpha = 0.05,
    ssPower = 0.80
)
```

---

## Quick Reference: Choosing the Right Method

| Data Type | Raters | Recommended Methods |
|-----------|--------|---------------------|
| Binary/Nominal | 2 | Cohen's kappa, Gwet's AC1, PABAK, specific agreement |
| Binary/Nominal | 3+ | Fleiss' kappa, Light's kappa, Krippendorff's alpha |
| Ordinal | 2+ | Weighted kappa (linear/quadratic), Kendall's W |
| Continuous | 2 | ICC, Bland-Altman, Lin's CCC, TDI |
| Continuous | 3+ | ICC, mean Pearson/Spearman, rater clustering |
| Any with NAs | Any | Krippendorff's alpha |
| Multi-center | Any | Hierarchical kappa, variance decomposition |
| Before/After | Any | Paired agreement test, mixed-effects comparison |
| Test-retest | Any | Inter/intra-rater analysis |

---

## References

- Cohen J (1960). A coefficient of agreement for nominal scales. *Educational and Psychological Measurement*, 20(1), 37-46.
- Fleiss JL (1971). Measuring nominal scale agreement among many raters. *Psychological Bulletin*, 76(5), 378-382.
- Krippendorff K (2011). Computing Krippendorff's alpha-reliability. *Annenberg School for Communication Departmental Papers*, Paper 43. University of Pennsylvania.
- Gwet KL (2008). Computing inter-rater reliability and its variance in the presence of high agreement. *British Journal of Mathematical and Statistical Psychology*, 61(1), 29-48.
- Shrout PE, Fleiss JL (1979). Intraclass correlations: uses in assessing rater reliability. *Psychological Bulletin*, 86(2), 420-428.
- Lin LI (1989). A concordance correlation coefficient to evaluate reproducibility. *Biometrics*, 45(1), 255-268.
- Bland JM, Altman DG (1986). Statistical methods for assessing agreement between two methods of clinical measurement. *The Lancet*, 327(8476), 307-310.
- Donner A, Eliasziw M (1992). A goodness-of-fit approach to inference procedures for the kappa statistic: Confidence interval construction, significance-testing and sample size estimation. *Statistics in Medicine*, 11(11), 1511-1519.
