---
title: "Function Reference: alluvial() - Complete Parameter Guide"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function Reference: alluvial() - Complete Parameter Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

The `alluvial()` function in ClinicoPath creates comprehensive alluvial diagrams for visualizing categorical data flows. This reference document provides complete parameter documentation and usage examples.

## Function Signature

```r
alluvial(
  data,
  vars,
  condensationvar = NULL,
  excl = FALSE,
  marg = FALSE,
  fill = "first_variable",
  bin = "default",
  orient = "vert",
  usetitle = FALSE,
  mytitle = "Alluvial Plot"
)
```

## Parameters

### Required Parameters

#### `data`
- **Type**: Data frame
- **Description**: The dataset containing the variables for analysis
- **Example**: `data = histopathology`

#### `vars`
- **Type**: Vector of variable names
- **Description**: 2-8 categorical variables for the alluvial diagram
- **Example**: `vars = c("Sex", "Grade_Level", "LVI", "PNI")`

### Optional Parameters

#### `condensationvar`
- **Type**: Variable name (string)
- **Default**: `NULL`
- **Description**: Variable used to create a condensation plot showing how this variable relates to the main flow patterns
- **Example**: `condensationvar = "Sex"`
- **Note**: When specified, generates a second plot alongside the main alluvial diagram

#### `excl`
- **Type**: Boolean
- **Default**: `FALSE`
- **Description**: Whether to exclude missing values from the analysis
- **Options**:
  - `FALSE`: Include missing values (default)
  - `TRUE`: Exclude missing values
- **Example**: `excl = TRUE`

#### `marg`
- **Type**: Boolean
- **Default**: `FALSE`
- **Description**: Whether to add marginal histograms showing variable distributions
- **Options**:
  - `FALSE`: No marginal plots (default)
  - `TRUE`: Add marginal histograms
- **Constraints**: 
  - Only available with vertical orientation (`orient = "vert"`)
  - Cannot be used with custom titles (`usetitle = FALSE`)
- **Example**: `marg = TRUE`

#### `fill`
- **Type**: String
- **Default**: `"first_variable"`
- **Description**: Strategy for coloring the flows in the diagram
- **Options**:
  - `"first_variable"`: Color by the first variable in vars (default)
  - `"last_variable"`: Color by the last variable in vars
  - `"all_flows"`: Each unique flow path gets a different color
  - `"values"`: Color by variable values
- **Example**: `fill = "all_flows"`

#### `bin`
- **Type**: String
- **Default**: `"default"`
- **Description**: Method for labeling bins when continuous variables are automatically binned
- **Options**:
  - `"default"`: Uses labels c("LL", "ML", "M", "MH", "HH")
  - `"mean"`: Bin labels based on mean values
  - `"median"`: Bin labels based on median values
  - `"min_max"`: Bin labels showing min-max ranges
  - `"cuts"`: Bin labels based on cut points
- **Example**: `bin = "median"`

#### `orient`
- **Type**: String
- **Default**: `"vert"`
- **Description**: Orientation of the alluvial diagram
- **Options**:
  - `"vert"`: Vertical flow (default)
  - `"horr"`: Horizontal flow
- **Constraints**: Marginal plots only work with vertical orientation
- **Example**: `orient = "horr"`

#### `usetitle`
- **Type**: Boolean
- **Default**: `FALSE`
- **Description**: Whether to use a custom title for the plot
- **Options**:
  - `FALSE`: Use default title (default)
  - `TRUE`: Use custom title specified in `mytitle`
- **Constraints**: Cannot be used with marginal plots (`marg = FALSE`)
- **Example**: `usetitle = TRUE`

#### `mytitle`
- **Type**: String
- **Default**: `"Alluvial Plot"`
- **Description**: Custom title text for the plot
- **Note**: Only used when `usetitle = TRUE`
- **Example**: `mytitle = "Patient Flow Analysis"`

## Parameter Constraints and Interactions

### Mutual Exclusions
1. **Marginal plots + Custom titles**: Cannot use both simultaneously
   - If `marg = TRUE`, then `usetitle` must be `FALSE`
   - If `usetitle = TRUE`, then `marg` must be `FALSE`

2. **Marginal plots + Horizontal orientation**: Cannot use both simultaneously
   - If `marg = TRUE`, then `orient` must be `"vert"`
   - If `orient = "horr"`, then `marg` must be `FALSE`

### Valid Combinations
```r
# Valid: Marginal plots with vertical orientation
alluvial(data = histopathology, vars = c("Sex", "LVI"), marg = TRUE, orient = "vert")

# Valid: Custom title with horizontal orientation  
alluvial(data = histopathology, vars = c("Sex", "LVI"), usetitle = TRUE, orient = "horr")

# Invalid: This combination will cause issues
# alluvial(data = histopathology, vars = c("Sex", "LVI"), marg = TRUE, usetitle = TRUE)
```

## Return Value

The function returns an `alluvialClass` object that inherits from `alluvialBase`. This object contains:
- Plot rendering methods (`.plot` and `.plot2`)
- Parameter settings
- Data processing methods
- jamovi-compatible analysis structure

## Examples

### Basic Usage

```{r basic-examples}
library(ClinicoPath)
data("histopathology")

# Minimal example
alluvial(
  data = histopathology,
  vars = c("Sex", "Grade_Level", "LVI")
)

# With parameter customization
alluvial(
  data = histopathology,
  vars = c("Sex", "Race", "Group", "Grade_Level"),
  fill = "last_variable",
  excl = TRUE,
  orient = "horr"
)
```

### Advanced Usage

```{r advanced-examples}
# Comprehensive parameter usage
alluvial(
  data = histopathology,
  vars = c("Sex", "Race", "Group", "Grade_Level", "LVI"),
  condensationvar = "PNI",
  excl = TRUE,
  marg = FALSE,
  fill = "all_flows",
  bin = "median",
  orient = "vert",
  usetitle = TRUE,
  mytitle = "Comprehensive Clinical Characteristics Flow"
)

# Marginal plots example
alluvial(
  data = histopathology,
  vars = c("Sex", "Grade_Level", "LVI"),
  marg = TRUE,
  orient = "vert",
  fill = "first_variable"
)

# Condensation analysis
alluvial(
  data = histopathology,
  vars = c("Grade_Level", "LVI", "PNI"),
  condensationvar = "Sex",
  fill = "values"
)
```

### Different Datasets

```{r dataset-examples}
# Colon cancer analysis
data("colon")
# Convert numeric variables to factors for alluvial plot
colon$sex <- factor(colon$sex, levels = c(0, 1), labels = c("Female", "Male"))
colon$differ <- factor(colon$differ, levels = c(1, 2, 3), labels = c("Well", "Moderate", "Poor"))
colon$extent <- factor(colon$extent, levels = c(1, 2, 3, 4), labels = c("Submucosa", "Muscle", "Serosa", "Contiguous"))

alluvial(
  data = colon,
  vars = c("sex", "differ", "extent"),
  fill = "first_variable",
  usetitle = TRUE,
  mytitle = "Colon Cancer Characteristics"
)

# Using histopathology dataset for treatment analysis
data("histopathology")
alluvial(
  data = histopathology,
  vars = c("Grade", "LVI", "PNI"),
  fill = "last_variable"
)
```

## Technical Details

### Underlying Implementation
- Uses the `easyalluvial` R package for core functionality
- Supports up to 8 variables (`max_variables = 8` in easyalluvial)
- Integrates with jamovi's R6 class system
- Provides dual plot functionality (main + condensation)

### Data Requirements
- Variables should be categorical (factors or character vectors)
- Continuous variables are automatically binned using the specified `bin` method
- Missing values are handled according to the `excl` parameter
- Minimum 2 variables, maximum 8 variables recommended

### Performance Considerations
- Works best with moderate sample sizes (50-500 cases)
- Large datasets may produce cluttered visualizations
- Complex flows with many categories may be difficult to interpret

## Error Handling

The function includes comprehensive error handling:
- Empty datasets trigger appropriate error messages
- Invalid parameter combinations are prevented through UI constraints
- Missing required parameters show helpful todo messages

## See Also

- [Comprehensive Alluvial Guide](clinicopath-descriptives-05-alluvial.html) - Complete usage examples
- `easyalluvial` package documentation
- Related ClinicoPath visualization functions: `jjalluvial()`, `riverplot()`

## Version Information

This documentation is current as of ClinicoPath version 0.0.3.32.
