---
title: "Continuous Variable Comparisons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Variable Comparisons}
  %\VignetteEngine{rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
library(jjstatsplot)
data(mtcars)
```

This article focuses on functions that compare continuous variables.

## Comparing groups with `jjbetweenstats()`

`jjbetweenstats()` combines box and violin plots to compare a continuous
response across groups. Statistical tests such as ANOVA or their
non-parametric equivalents are performed automatically depending on the
`typestatistics` argument.

```{r}
jjbetweenstats(data = mtcars, dep = mpg, group = cyl)
```

## Repeated measures with `jjwithinstats()`

For paired or repeated observations, `jjwithinstats()` displays the
measurements for each condition alongside the results of a paired test.
Below we compare miles-per-gallon recorded in two hypothetical
conditions represented by `mpg` and `hp` just for demonstration.

```{r}
jjwithinstats(data = mtcars, dep1 = mpg, dep2 = hp)
```

## Histograms with `jjhistostats()`

`jjhistostats()` produces histograms and overlays the results of a test
for normality. The example below shows the distribution of engine
horsepower.

```{r}
jjhistostats(data = mtcars, dep = hp)
```


