---
title: "Summarising Clinicopathological Data"
author: "ClinicoPathDescriptives Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarising Clinicopathological Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette focuses on the functions that generate tables and textual summaries.

## Continuous Variables

`summarydata()` provides descriptive statistics (mean, median, quartiles) for numeric variables.

```{r cont-sum, eval=FALSE}
summarydata(
  data = histopathology,
  vars = vars(Age, TumorSize)
)
```

## Categorical Variables

`reportcat()` summarises categorical variables with counts and percentages.

```{r cat-sum, eval=FALSE}
reportcat(
  data = histopathology,
  vars = vars(Sex, Grade)
)
```

## Cross Tables

Use `crosstable()` to create cross tabulations with statistical tests. The `group` argument defines the column variable.

```{r crosstable-vig, eval=FALSE}
crosstable(
  data = histopathology,
  vars = vars(Sex, Grade),
  group = "PreinvasiveComponent",
  sty = "nejm",
  excl = TRUE
)
```

## Table One for Publications

`tableone()` creates publication-ready tables of baseline characteristics.

```{r tableone-vig, eval=FALSE}
tableone(
  data = histopathology,
  vars = vars(Sex, PreinvasiveComponent, LVI, PNI, Grade, Age),
  sty = "t3",
  excl = TRUE
)
```

These functions output results objects that contain tables which can be converted to data frames for further customisation.

