---
title: "Introduction to Decision Panel Optimization"
author: "meddecide Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Decision Panel Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The Decision Panel Optimization module in the `meddecide` package provides a comprehensive framework for optimizing diagnostic test combinations in medical decision-making. This vignette introduces the basic concepts and demonstrates core functionality.

## Key Concepts

### Testing Strategies

When multiple diagnostic tests are available, they can be combined in different ways:

1. **Single Testing**: Use individual tests independently
2. **Parallel Testing**: Perform multiple tests simultaneously
   - ANY rule (OR): Positive if any test is positive
   - ALL rule (AND): Positive only if all tests are positive
   - MAJORITY rule: Positive if majority of tests are positive
3. **Sequential Testing**: Perform tests in sequence based on previous results
   - Stop on first positive
   - Confirmatory (require multiple positives)
   - Exclusion (require multiple negatives)

### Optimization Criteria

The module can optimize test panels based on various criteria:

- **Accuracy**: Overall correct classification rate
- **Sensitivity**: Ability to detect disease (minimize false negatives)
- **Specificity**: Ability to rule out disease (minimize false positives)
- **Predictive Values**: PPV and NPV
- **Cost-Effectiveness**: Balance performance with resource utilization
- **Utility**: Custom utility functions incorporating costs of errors

## Installation and Loading

```{r installation, eval=FALSE}
# Install meddecide package
install.packages("meddecide")

# Or install from GitHub
devtools::install_github("ClinicoPath/meddecide")
```




```{r loading}
# Load required packages
library(meddecide)
library(dplyr)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(knitr)
library(forcats)
```



```{r include=FALSE}
# Test Datasets for Decision Panel Optimization Module

# Set seed for reproducibility
set.seed(42)

# Create data directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# ============================================================================
# DATASET 1: COVID-19 SCREENING
# ============================================================================

create_covid_data <- function(n = 1000, prevalence = 0.15) {
  # True disease status
  disease <- rbinom(n, 1, prevalence)
  
  # Rapid Antigen Test (RAT)
  # Sensitivity: 65%, Specificity: 98%
  rat_prob <- ifelse(disease == 1, 0.65, 0.02)
  rat_result <- rbinom(n, 1, rat_prob)
  
  # PCR Test
  # Sensitivity: 95%, Specificity: 99%
  pcr_prob <- ifelse(disease == 1, 0.95, 0.01)
  pcr_result <- rbinom(n, 1, pcr_prob)
  
  # Chest CT
  # Sensitivity: 90%, Specificity: 85%
  ct_prob <- ifelse(disease == 1, 0.90, 0.15)
  ct_result <- rbinom(n, 1, ct_prob)
  
  # Clinical symptoms score (0-10)
  # Higher in disease
  symptoms <- ifelse(disease == 1,
                     pmin(10, round(rnorm(sum(disease == 1), 7, 2))),
                     pmax(0, round(rnorm(sum(disease == 0), 3, 2))))
  
  # Create dataset
  covid_data <- data.frame(
    patient_id = 1:n,
    rapid_antigen = factor(rat_result, levels = c(0, 1), 
                          labels = c("Negative", "Positive")),
    pcr = factor(pcr_result, levels = c(0, 1), 
                labels = c("Negative", "Positive")),
    chest_ct = factor(ct_result, levels = c(0, 1), 
                     labels = c("Normal", "Abnormal")),
    symptom_score = symptoms,
    covid_status = factor(disease, levels = c(0, 1), 
                         labels = c("Negative", "Positive")),
    age = round(rnorm(n, 45, 15)),
    risk_group = factor(sample(c("Low", "Medium", "High"), n, 
                              replace = TRUE, prob = c(0.6, 0.3, 0.1)))
  )
  
  # Add some missing values realistically
  # PCR might be missing if rapid test is negative
  missing_pcr <- which(covid_data$rapid_antigen == "Negative" & 
                      runif(n) < 0.3)
  covid_data$pcr[missing_pcr] <- NA
  
  return(covid_data)
}

# Generate COVID dataset
covid_screening_data <- create_covid_data(n = 1000, prevalence = 0.15)

# ============================================================================
# DATASET 2: BREAST CANCER SCREENING
# ============================================================================

create_breast_cancer_data <- function(n = 2000, prevalence = 0.005) {
  # True disease status (low prevalence for screening)
  disease <- rbinom(n, 1, prevalence)
  
  # Clinical Breast Exam (CBE)
  # Sensitivity: 54%, Specificity: 94%
  cbe_prob <- ifelse(disease == 1, 0.54, 0.06)
  cbe_result <- rbinom(n, 1, cbe_prob)
  
  # Mammography
  # Sensitivity: 85%, Specificity: 95%
  mammo_prob <- ifelse(disease == 1, 0.85, 0.05)
  mammo_result <- rbinom(n, 1, mammo_prob)
  
  # Ultrasound
  # Sensitivity: 80%, Specificity: 90%
  us_prob <- ifelse(disease == 1, 0.80, 0.10)
  us_result <- rbinom(n, 1, us_prob)
  
  # MRI (for high-risk patients)
  # Sensitivity: 95%, Specificity: 85%
  mri_prob <- ifelse(disease == 1, 0.95, 0.15)
  mri_result <- rbinom(n, 1, mri_prob)
  
  # Create risk factors
  age <- round(rnorm(n, 55, 10))
  age[age < 40] <- 40
  age[age > 75] <- 75
  
  family_history <- rbinom(n, 1, 0.15)
  brca_status <- rbinom(n, 1, 0.02)
  
  # Create dataset
  breast_cancer_data <- data.frame(
    patient_id = 1:n,
    clinical_exam = factor(cbe_result, levels = c(0, 1), 
                          labels = c("Normal", "Abnormal")),
    mammography = factor(mammo_result, levels = c(0, 1), 
                        labels = c("BIRADS 1-2", "BIRADS 3-5")),
    ultrasound = factor(us_result, levels = c(0, 1), 
                       labels = c("Normal", "Suspicious")),
    mri = factor(mri_result, levels = c(0, 1), 
                labels = c("Normal", "Suspicious")),
    cancer_status = factor(disease, levels = c(0, 1), 
                          labels = c("No Cancer", "Cancer")),
    age = age,
    family_history = factor(family_history, levels = c(0, 1), 
                           labels = c("No", "Yes")),
    brca_mutation = factor(brca_status, levels = c(0, 1), 
                          labels = c("Negative", "Positive")),
    breast_density = factor(sample(c("A", "B", "C", "D"), n, 
                                  replace = TRUE, 
                                  prob = c(0.1, 0.4, 0.4, 0.1)))
  )
  
  # MRI typically only done for high-risk
  low_risk_idx <- which(breast_cancer_data$family_history == "No" & 
                       breast_cancer_data$brca_mutation == "Negative")
  breast_cancer_data$mri[sample(low_risk_idx, 
                               length(low_risk_idx) * 0.9)] <- NA
  
  return(breast_cancer_data)
}

# Generate breast cancer dataset
breast_cancer_data <- create_breast_cancer_data(n = 2000, prevalence = 0.005)

# ============================================================================
# DATASET 3: TUBERCULOSIS DIAGNOSIS
# ============================================================================

create_tb_data <- function(n = 1500, prevalence = 0.20) {
  # True disease status (high prevalence in TB clinic)
  disease <- rbinom(n, 1, prevalence)
  
  # Symptom screening (cough > 2 weeks, fever, weight loss, night sweats)
  # Sensitivity: 80%, Specificity: 60%
  symptom_prob <- ifelse(disease == 1, 0.80, 0.40)
  symptom_result <- rbinom(n, 1, symptom_prob)
  
  # Sputum smear microscopy
  # Sensitivity: 60%, Specificity: 98%
  smear_prob <- ifelse(disease == 1, 0.60, 0.02)
  smear_result <- rbinom(n, 1, smear_prob)
  
  # GeneXpert MTB/RIF
  # Sensitivity: 88%, Specificity: 98%
  genexpert_prob <- ifelse(disease == 1, 0.88, 0.02)
  genexpert_result <- rbinom(n, 1, genexpert_prob)
  
  # Culture (gold standard-ish)
  # Sensitivity: 95%, Specificity: 100%
  culture_prob <- ifelse(disease == 1, 0.95, 0.00)
  culture_result <- rbinom(n, 1, culture_prob)
  
  # Chest X-ray
  # Sensitivity: 85%, Specificity: 75%
  cxr_prob <- ifelse(disease == 1, 0.85, 0.25)
  cxr_result <- rbinom(n, 1, cxr_prob)
  
  # HIV status affects presentation
  hiv_status <- rbinom(n, 1, 0.25)
  
  # Create dataset
  tb_data <- data.frame(
    patient_id = 1:n,
    symptoms = factor(symptom_result, levels = c(0, 1), 
                     labels = c("No", "Yes")),
    sputum_smear = factor(smear_result, levels = c(0, 1), 
                         labels = c("Negative", "Positive")),
    genexpert = factor(genexpert_result, levels = c(0, 1), 
                      labels = c("MTB not detected", "MTB detected")),
    culture = factor(culture_result, levels = c(0, 1), 
                    labels = c("Negative", "Positive")),
    chest_xray = factor(cxr_result, levels = c(0, 1), 
                       labels = c("Normal", "Abnormal")),
    tb_status = factor(disease, levels = c(0, 1), 
                      labels = c("No TB", "TB")),
    hiv_status = factor(hiv_status, levels = c(0, 1), 
                       labels = c("Negative", "Positive")),
    age = round(rnorm(n, 35, 15)),
    contact_history = factor(rbinom(n, 1, 0.30), levels = c(0, 1), 
                           labels = c("No", "Yes"))
  )
  
  # Culture takes time, might not be done for all
  no_culture_idx <- which(tb_data$genexpert == "MTB not detected" & 
                         tb_data$symptoms == "No" & 
                         runif(n) < 0.4)
  tb_data$culture[no_culture_idx] <- NA
  
  return(tb_data)
}

# Generate TB dataset
tb_diagnosis_data <- create_tb_data(n = 1500, prevalence = 0.20)

# ============================================================================
# DATASET 4: MYOCARDIAL INFARCTION RULE-OUT (FIXED)
# ============================================================================

create_mi_data <- function(n = 800, prevalence = 0.10) {
  # True disease status
  disease <- rbinom(n, 1, prevalence)
  
  # ECG changes
  # Sensitivity: 55%, Specificity: 95%
  ecg_prob <- ifelse(disease == 1, 0.55, 0.05)
  ecg_result <- rbinom(n, 1, ecg_prob)
  
  # Initial troponin
  # Sensitivity: 85%, Specificity: 95%
  trop1_prob <- ifelse(disease == 1, 0.85, 0.05)
  trop1_result <- rbinom(n, 1, trop1_prob)
  
  # 3-hour troponin
  # Sensitivity: 98%, Specificity: 95%
  trop3_prob <- ifelse(disease == 1, 0.98, 0.05)
  trop3_result <- rbinom(n, 1, trop3_prob)
  
  # Make sure 3-hour is at least as positive as initial
  trop3_result <- pmax(trop1_result, trop3_result)
  
  # CT Angiography
  # Sensitivity: 95%, Specificity: 90%
  cta_prob <- ifelse(disease == 1, 0.95, 0.10)
  cta_result <- rbinom(n, 1, cta_prob)
  
  # Clinical risk score components
  age <- round(rnorm(n, 60, 15))
  age[age < 30] <- 30
  age[age > 90] <- 90
  
  # FIXED: Generate chest pain type for each individual based on their disease status
  chest_pain_type <- character(n)
  pain_types <- c("Typical", "Atypical", "Non-cardiac")
  
  for (i in 1:n) {
    if (disease[i] == 1) {
      # Disease present - more likely to have typical pain
      chest_pain_type[i] <- sample(pain_types, 1, prob = c(0.6, 0.3, 0.1))
    } else {
      # No disease - more likely to have non-cardiac pain
      chest_pain_type[i] <- sample(pain_types, 1, prob = c(0.1, 0.3, 0.6))
    }
  }
  
  # Create dataset
  mi_data <- data.frame(
    patient_id = 1:n,
    ecg = factor(ecg_result, levels = c(0, 1), 
                labels = c("Normal", "Ischemic changes")),
    troponin_initial = factor(trop1_result, levels = c(0, 1), 
                             labels = c("Normal", "Elevated")),
    troponin_3hr = factor(trop3_result, levels = c(0, 1), 
                         labels = c("Normal", "Elevated")),
    ct_angiography = factor(cta_result, levels = c(0, 1), 
                           labels = c("No stenosis", "Significant stenosis")),
    mi_status = factor(disease, levels = c(0, 1), 
                      labels = c("No MI", "MI")),
    age = age,
    chest_pain = factor(chest_pain_type),
    diabetes = factor(rbinom(n, 1, 0.25), levels = c(0, 1), 
                     labels = c("No", "Yes")),
    smoking = factor(rbinom(n, 1, 0.30), levels = c(0, 1), 
                    labels = c("No", "Yes")),
    prior_cad = factor(rbinom(n, 1, 0.20), levels = c(0, 1), 
                      labels = c("No", "Yes"))
  )
  
  # CTA typically only for intermediate risk
  low_risk_idx <- which(mi_data$chest_pain == "Non-cardiac" & 
                       mi_data$ecg == "Normal" & 
                       mi_data$troponin_initial == "Normal")
  if (length(low_risk_idx) > 0) {
    mi_data$ct_angiography[sample(low_risk_idx, 
                                 min(length(low_risk_idx), 
                                     floor(length(low_risk_idx) * 0.8)))] <- NA
  }
  
  return(mi_data)
}

# Generate MI dataset
mi_ruleout_data <- create_mi_data(n = 800, prevalence = 0.10)

# ============================================================================
# DATASET 5: THYROID NODULE EVALUATION
# ============================================================================

create_thyroid_data <- function(n = 600, prevalence = 0.05) {
  # True disease status (thyroid cancer)
  disease <- rbinom(n, 1, prevalence)
  
  # Ultrasound features (TI-RADS)
  # Sensitivity: 90%, Specificity: 70%
  us_prob <- ifelse(disease == 1, 0.90, 0.30)
  us_result <- rbinom(n, 1, us_prob)
  
  # Fine Needle Aspiration (FNA) cytology
  # Sensitivity: 95%, Specificity: 98%
  fna_prob <- ifelse(disease == 1, 0.95, 0.02)
  fna_result <- rbinom(n, 1, fna_prob)
  
  # Molecular testing (ThyroSeq/Afirma)
  # Sensitivity: 91%, Specificity: 85%
  molecular_prob <- ifelse(disease == 1, 0.91, 0.15)
  molecular_result <- rbinom(n, 1, molecular_prob)
  
  # Thyroglobulin levels
  # Sensitivity: 70%, Specificity: 80%
  tg_prob <- ifelse(disease == 1, 0.70, 0.20)
  tg_result <- rbinom(n, 1, tg_prob)
  
  # Create dataset
  thyroid_data <- data.frame(
    patient_id = 1:n,
    ultrasound = factor(us_result, levels = c(0, 1), 
                       labels = c("TI-RADS 1-3", "TI-RADS 4-5")),
    fna_cytology = factor(fna_result, levels = c(0, 1), 
                         labels = c("Benign/Indeterminate", "Suspicious/Malignant")),
    molecular_test = factor(molecular_result, levels = c(0, 1), 
                           labels = c("Benign", "Suspicious")),
    thyroglobulin = factor(tg_result, levels = c(0, 1), 
                          labels = c("Normal", "Elevated")),
    cancer_status = factor(disease, levels = c(0, 1), 
                          labels = c("Benign", "Malignant")),
    nodule_size = round(rlnorm(n, log(15), 0.5)),
    age = round(rnorm(n, 50, 15)),
    gender = factor(sample(c("Female", "Male"), n, 
                          replace = TRUE, prob = c(0.75, 0.25))),
    radiation_history = factor(rbinom(n, 1, 0.05), levels = c(0, 1), 
                             labels = c("No", "Yes"))
  )
  
  # Molecular testing only for indeterminate FNA
  molecular_not_done <- which(thyroid_data$fna_cytology != "Benign/Indeterminate" |
                             runif(n) < 0.5)
  thyroid_data$molecular_test[molecular_not_done] <- NA
  
  return(thyroid_data)
}

# Generate thyroid dataset
thyroid_nodule_data <- create_thyroid_data(n = 600, prevalence = 0.05)

# ============================================================================
# SAVE ALL DATASETS
# ============================================================================

# Save as RData file
save(covid_screening_data, 
     breast_cancer_data,
     tb_diagnosis_data,
     mi_ruleout_data,
     thyroid_nodule_data,
     file = "data/decision_panel_test_data.RData")

# Also save as CSV files for external use
write.csv(covid_screening_data, "data/covid_screening_data.csv", row.names = FALSE)
write.csv(breast_cancer_data, "data/breast_cancer_data.csv", row.names = FALSE)
write.csv(tb_diagnosis_data, "data/tb_diagnosis_data.csv", row.names = FALSE)
write.csv(mi_ruleout_data, "data/mi_ruleout_data.csv", row.names = FALSE)
write.csv(thyroid_nodule_data, "data/thyroid_nodule_data.csv", row.names = FALSE)

# Print confirmation
cat("✓ All datasets generated and saved successfully!\n")

# ============================================================================
# DATASET SUMMARIES FOR VERIFICATION
# ============================================================================

summarize_dataset <- function(data, disease_col, test_cols) {
  cat("\nDataset Summary:\n")
  cat("Total observations:", nrow(data), "\n")
  
  # Check if disease column exists and has the expected levels
  if (disease_col %in% names(data)) {
    disease_levels <- levels(factor(data[[disease_col]]))
    if (length(disease_levels) >= 2) {
      prevalence <- mean(data[[disease_col]] == disease_levels[2], na.rm = TRUE)
      cat("Disease prevalence:", round(prevalence * 100, 1), "%\n")
    } else {
      cat("Disease column found but insufficient levels\n")
    }
  } else {
    cat("Disease column not found\n")
  }
  
  cat("\nTest performance:\n")
  
  for (test in test_cols) {
    if (test %in% names(data) && disease_col %in% names(data)) {
      # Remove rows with NAs in either variable
      complete_data <- data[!is.na(data[[test]]) & !is.na(data[[disease_col]]), ]
      
      if (nrow(complete_data) > 0) {
        tab <- table(complete_data[[test]], complete_data[[disease_col]])
        
        if (nrow(tab) >= 2 && ncol(tab) >= 2) {
          # Assume positive is the second level (index 2)
          if (nrow(tab) == 2 && ncol(tab) == 2) {
            sens <- tab[2,2] / sum(tab[,2])
            spec <- tab[1,1] / sum(tab[,1])
            cat(sprintf("  %s: Sens=%.1f%%, Spec=%.1f%% (n=%d)\n", 
                        test, sens*100, spec*100, nrow(complete_data)))
          } else {
            cat(sprintf("  %s: Complex table structure (n=%d)\n", 
                        test, nrow(complete_data)))
          }
        } else {
          cat(sprintf("  %s: Insufficient data for analysis (n=%d)\n", 
                      test, nrow(complete_data)))
        }
      } else {
        cat(sprintf("  %s: No complete cases\n", test))
      }
    } else {
      cat(sprintf("  %s: Column not found\n", test))
    }
  }
}

# Print summaries for verification
cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("DATASET SUMMARIES")
cat("\n", paste(rep("=", 60), collapse = ""), "\n")

cat("\n=== COVID-19 SCREENING DATA ===")
summarize_dataset(covid_screening_data, "covid_status", 
                 c("rapid_antigen", "pcr", "chest_ct"))

cat("\n=== BREAST CANCER SCREENING DATA ===")
summarize_dataset(breast_cancer_data, "cancer_status",
                 c("clinical_exam", "mammography", "ultrasound", "mri"))

cat("\n=== TUBERCULOSIS DIAGNOSIS DATA ===")
summarize_dataset(tb_diagnosis_data, "tb_status",
                 c("symptoms", "sputum_smear", "genexpert", "culture", "chest_xray"))

cat("\n=== MYOCARDIAL INFARCTION DATA ===")
summarize_dataset(mi_ruleout_data, "mi_status",
                 c("ecg", "troponin_initial", "troponin_3hr", "ct_angiography"))

cat("\n=== THYROID NODULE DATA ===")
summarize_dataset(thyroid_nodule_data, "cancer_status",
                 c("ultrasound", "fna_cytology", "molecular_test", "thyroglobulin"))

cat("\n", paste(rep("=", 60), collapse = ""), "\n")
```








## Basic Example: COVID-19 Screening

Let's start with a simple example using COVID-19 screening data:

```{r covid_example}
# Examine the data structure
str(covid_screening_data)

# Check disease prevalence
table(covid_screening_data$covid_status)
prop.table(table(covid_screening_data$covid_status))
```

### Running Basic Analysis

```{r basic_analysis, eval=FALSE}
# Basic decision panel analysis
covid_panel <- decisionpanel(
  data = covid_screening_data,
  tests = c("rapid_antigen", "pcr", "chest_ct"),
  testLevels = c("Positive", "Positive", "Abnormal"),
  gold = "covid_status",
  goldPositive = "Positive",
  strategies = "all",
  optimizationCriteria = "accuracy"
)
```

### Interpreting Results

The analysis provides several key outputs:

1. **Individual Test Performance**: How each test performs alone
2. **Optimal Panel**: The best combination of tests
3. **Strategy Comparison**: Performance of different testing approaches
4. **Decision Tree**: Optimal sequence for testing

## Understanding Testing Strategies

### Parallel Testing Example

```{r parallel_example}
# Simulate parallel testing with ANY rule
# Positive if rapid_antigen OR pcr is positive
parallel_any <- with(covid_screening_data, 
  rapid_antigen == "Positive" | pcr == "Positive"
)

# Create confusion matrix
conf_matrix_any <- table(
  Predicted = parallel_any,
  Actual = covid_screening_data$covid_status == "Positive"
)

print(conf_matrix_any)

# Calculate metrics
sensitivity_any <- conf_matrix_any[2,2] / sum(conf_matrix_any[,2])
specificity_any <- conf_matrix_any[1,1] / sum(conf_matrix_any[,1])

cat("Parallel ANY Rule:\n")
cat(sprintf("Sensitivity: %.1f%%\n", sensitivity_any * 100))
cat(sprintf("Specificity: %.1f%%\n", specificity_any * 100))
```

### Sequential Testing Example

```{r sequential_example}
# Simulate sequential testing
# Start with rapid test, only do PCR if rapid is positive
sequential_result <- rep("Negative", nrow(covid_screening_data))

# Those with positive rapid test
rapid_pos_idx <- which(covid_screening_data$rapid_antigen == "Positive")

# Among those, check PCR
sequential_result[rapid_pos_idx] <- 
  ifelse(covid_screening_data$pcr[rapid_pos_idx] == "Positive", 
         "Positive", "Negative")

# Create confusion matrix
conf_matrix_seq <- table(
  Predicted = sequential_result == "Positive",
  Actual = covid_screening_data$covid_status == "Positive"
)

print(conf_matrix_seq)

# Calculate metrics
sensitivity_seq <- conf_matrix_seq[2,2] / sum(conf_matrix_seq[,2])
specificity_seq <- conf_matrix_seq[1,1] / sum(conf_matrix_seq[,1])

cat("\nSequential Testing:\n")
cat(sprintf("Sensitivity: %.1f%%\n", sensitivity_seq * 100))
cat(sprintf("Specificity: %.1f%%\n", specificity_seq * 100))

# Calculate cost savings
pcr_tests_saved <- sum(covid_screening_data$rapid_antigen == "Negative")
cat(sprintf("PCR tests saved: %d (%.1f%%)\n", 
            pcr_tests_saved, 
            pcr_tests_saved/nrow(covid_screening_data) * 100))
```

## Cost-Effectiveness Analysis

When costs are considered, the optimal strategy may change:

```{r cost_analysis, eval=FALSE}
# Analysis with costs
covid_panel_cost <- decisionpanel(
  data = covid_screening_data,
  tests = c("rapid_antigen", "pcr", "chest_ct"),
  testLevels = c("Positive", "Positive", "Abnormal"),
  gold = "covid_status",
  goldPositive = "Positive",
  strategies = "all",
  optimizationCriteria = "utility",
  useCosts = TRUE,
  testCosts = "5,50,200",  # Costs for each test
  fpCost = 500,            # Cost of false positive
  fnCost = 5000            # Cost of false negative
)
```

## Visualization

### Performance Comparison Plot

```{r visualization}
# Create performance comparison data
strategies <- data.frame(
  Strategy = c("Rapid Only", "PCR Only", "Parallel ANY", "Sequential"),
  Sensitivity = c(65, 95, 98, 62),
  Specificity = c(98, 99, 97, 99.9),
  Cost = c(5, 50, 55, 15)
)

# Plot sensitivity vs specificity
ggplot(strategies, aes(x = 100 - Specificity, y = Sensitivity)) +
  geom_point(aes(size = Cost), alpha = 0.6) +
  geom_text(aes(label = Strategy), vjust = -1) +
  scale_size_continuous(range = c(3, 10)) +
  xlim(0, 5) + ylim(60, 100) +
  labs(
    title = "Testing Strategy Comparison",
    x = "False Positive Rate (%)",
    y = "Sensitivity (%)",
    size = "Cost ($)"
  ) +
  theme_minimal()
```

## Decision Trees

Decision trees provide clear algorithms for clinical use:

```{r decision_tree, eval=FALSE}
# Generate decision tree
covid_tree <- decisionpanel(
  data = covid_screening_data,
  tests = c("rapid_antigen", "pcr", "chest_ct", "symptom_score"),
  testLevels = c("Positive", "Positive", "Abnormal", ">5"),
  gold = "covid_status",
  goldPositive = "Positive",
  createTree = TRUE,
  treeMethod = "cart",
  maxDepth = 3
)
```

### Interpreting the Tree

A typical decision tree output might look like:

```
1. Start with Rapid Antigen Test
   ├─ If Positive (2% of patients)
   │  └─ Confirm with PCR
   │     ├─ If Positive → COVID Positive (PPV: 95%)
   │     └─ If Negative → COVID Negative (NPV: 98%)
   └─ If Negative (98% of patients)
      ├─ If Symptoms > 5
      │  └─ Perform Chest CT
      │     ├─ If Abnormal → Perform PCR
      │     └─ If Normal → COVID Negative
      └─ If Symptoms ≤ 5 → COVID Negative
```

## Advanced Features

### Cross-Validation

Validate panel performance using k-fold cross-validation:

```{r cross_validation, eval=FALSE}
# Run with cross-validation
covid_panel_cv <- decisionpanel(
  data = covid_screening_data,
  tests = c("rapid_antigen", "pcr", "chest_ct"),
  testLevels = c("Positive", "Positive", "Abnormal"),
  gold = "covid_status",
  goldPositive = "Positive",
  crossValidate = TRUE,
  nFolds = 5,
  seed = 123
)
```

### Bootstrap Confidence Intervals

Get uncertainty estimates for performance metrics:

```{r bootstrap, eval=FALSE}
# Run with bootstrap
covid_panel_boot <- decisionpanel(
  data = covid_screening_data,
  tests = c("rapid_antigen", "pcr", "chest_ct"),
  testLevels = c("Positive", "Positive", "Abnormal"),
  gold = "covid_status",
  goldPositive = "Positive",
  bootstrap = TRUE,
  bootReps = 1000,
  seed = 123
)
```

## Best Practices

1. **Start Simple**: Begin with individual test performance before combinations
2. **Consider Context**: Screening vs. diagnosis requires different strategies
3. **Validate Results**: Use cross-validation or separate test sets
4. **Include Costs**: Real-world decisions must consider resources
5. **Think Sequentially**: Often more efficient than parallel testing
6. **Set Constraints**: Define minimum acceptable performance
7. **Interpret Clinically**: Statistical optimality isn't everything

## Conclusion

The Decision Panel Optimization module provides a systematic approach to combining diagnostic tests. By considering various strategies, costs, and constraints, it helps identify practical testing algorithms that balance performance with resource utilization.

## Next Steps

- See the "Clinical Applications" vignette for disease-specific examples
- Review "Advanced Optimization" for complex scenarios
- Check "Implementation Guide" for deploying algorithms in practice

## Session Information

```{r session_info}
sessionInfo()
```
