---
title: "Comprehensive Guide: Summary of Continuous Variables"
author: "ClinicoPath jamovi Module Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide: Summary of Continuous Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `summarydata` analysis provides descriptive summaries for numeric variables
in jamovi and R. It reports available and missing counts, mean, sample standard
deviation, median, minimum, and maximum. Optional panels add distribution
diagnostics, IQR-based outlier screening, and draft statistical sentences.

# Example workflow in R

```{r}
library(ClinicoPath)
data("summarydata_test", package = "ClinicoPath")

summarydata(
  data = summarydata_test,
  vars = c("age_normal", "hemoglobin_lab"),
  distr = TRUE,
  outliers = TRUE,
  report_sentences = TRUE,
  decimal_places = 2
)
```

# Interpretation

The analysis uses each variable's available observations; it does not impute
missing data. Review the reported missing counts, especially when variables have
different amounts of missingness.

The Shapiro-Wilk test can detect a departure from normality but cannot establish
normality. Interpret it with the displayed distribution and the sample size.

The outlier option flags observations outside Q1 - 1.5 x IQR or Q3 + 1.5 x IQR.
These are statistical screening fences, not expected values or clinical
reference limits. Review flagged values for data errors, measurement problems,
and plausible biological extremes before considering exclusion.

Draft sentences require editing before use: add measurement units, the study
population, the measurement time point, and the chosen missing-data approach.
This descriptive analysis does not estimate or validate clinical reference
intervals and does not verify the assumptions of a later statistical model.
