---
title: "Comprehensive Guide: Table One"
author: "ClinicoPath jamovi Module Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide: Table One}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Scope

`ClinicoPath::tableone()` summarizes one overall cohort. It does not stratify by
group, compare groups, produce plots, impute values, or compute p-values,
confidence intervals or standardized mean differences. Each row is one case.
Check repeated records before calling the counts patients.

# Example in R

```{r example-data}
library(ClinicoPath)
data("tableone_test", package = "ClinicoPath")
example_data <- tableone_test[1:40, c("Age", "Sex", "TumorStage")]
example_data$Age[1:5] <- NA
```

The default style reports mean (SD) for numeric measurements and N (% recorded)
for categorical variables. For a binary factor it displays the second level,
named in the row. The Missing column reports percentages, not counts.

```{r standard-table}
standard <- ClinicoPath::tableone(
  data = example_data,
  vars = c("Age", "Sex", "TumorStage"),
  sty = "t1",
  showSummary = TRUE,
  showReportSentence = TRUE
)
cat(standard$tablestyle1$content)
```

For a median and quartiles, use `nonnormal = TRUE` with t1. This option does not
change the other engines or run a normality test.

```{r median-table}
median_table <- ClinicoPath::tableone(
  data = example_data, vars = c("Age", "Sex"),
  sty = "t1", nonnormal = TRUE
)
cat(median_table$tablestyle1$content)
```

# Style choices

| `sty` | Output |
| --- | --- |
| `"t1"` | tableone: mean (SD), or median (Q1, Q3); missing percentages |
| `"t2"` | gtsummary: median (Q1, Q3), N (%); Unknown counts |
| `"t3"` | arsenal: mean (SD), range, N (%); N-Miss counts |
| `"t4"` | janitor: categorical frequencies, totals and missing-value rows |

gtsummary can treat numeric variables with fewer than 10 distinct values as
categorical. Numeric category codes should be explicitly converted to factors
in R, or nominal/ordinal in jamovi.

Janitor accepts factors, ordered factors, text and logicals with at most 20
recorded categories. It skips numeric measurements instead of binning them.
Unused factor levels are not displayed. Single-row gtsummary summaries name
the counted level: `flag = TRUE` with `0 (0%)` means no recorded value was TRUE,
not that there were no observations.

```{r frequency-table, results='asis'}
frequencies <- ClinicoPath::tableone(
  data = example_data, vars = c("Sex", "TumorStage"), sty = "t4"
)
knitr::asis_output(frequencies$tablestyle4$content)
```

# Missingness and eligibility

With `excl = FALSE`, variables use their own recorded denominators. In t1–t3,
categorical percentages exclude missing values; missingness is reported
separately. In t4, Percent includes the NA row, while Valid Percent (shown when
needed) uses recorded values only.

With `excl = TRUE`, a case missing any included variable is removed, so retained
variables share one complete-case cohort. This is listwise deletion, not
imputation. Summary and report text still describe source missingness and case loss.

```{r complete-cases, results='asis'}
complete_cases <- ClinicoPath::tableone(
  data = example_data, vars = c("Age", "Sex", "TumorStage"),
  excl = TRUE, showSummary = TRUE
)
knitr::asis_output(complete_cases$summary$content)
```

All-missing variables and unsupported R storage classes (Date, date-time and
difftime, for example) are named and omitted before exclusion. Convert dates or
durations to measurements with explicitly chosen units before selecting them.
Omitted variables cannot change the analyzed cohort or appear in report text.

Matrix, array and list columns supplied through R are rejected before they can
be flattened. Extract the intended scalar measurement explicitly first.
Actual NA factor levels, including those created by `addNA()`, are treated as
missing before exclusion. Literal text categories `"NA"` and `"Unknown"` remain
categories. For example, the following table and report use 30 complete cases:

```{r explicit-na-level, results='asis'}
na_level_data <- data.frame(
  group = factor(c(rep("A", 20), rep("B", 10), rep(NA, 10)), exclude = NULL)
)
na_level_result <- ClinicoPath::tableone(
  na_level_data, "group", sty = "t3", excl = TRUE, showReportSentence = TRUE
)
stopifnot(grepl("30 cases with complete data", na_level_result$reportSentence$content))
knitr::asis_output(na_level_result$reportSentence$content)
```

If frequency weights are supplied, counts represent replicated rows rather
than verified unique patients. This does not provide complex-survey inference.

Data-quality notices precede the table. Small-sample, missingness and case-loss
thresholds are descriptive heuristics, not validated clinical cutoffs. Review
actual counts and missing-data patterns, not only whether a threshold passed.
Copy-ready text is editable and withheld when no table can be produced.

# In jamovi

Open **ExplorationT → ClinicoPath Descriptives → Table One**, select variables,
choose a style, and expand Output Options for the summary, educational guidance
or report sentence. The median option applies to t1 only. An empty variable
selection displays instructions.
