---
title: "Advanced Survival Analysis with jsurvival"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Advanced Survival Analysis with jsurvival}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This vignette covers advanced survival analysis techniques and methodological considerations when using jsurvival. It addresses complex scenarios that researchers commonly encounter in clinical and epidemiological studies.

# Advanced Cox Regression Topics

## Checking Proportional Hazards Assumption

The Cox proportional hazards model assumes that hazard ratios remain constant over time. Violation of this assumption can lead to misleading results.

### Visual Assessment

```{r ph-visual}
# Use survival analysis with extended follow-up time
ph_check <- survival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE,
  endplot = 120,  # Extended follow-up for visual inspection
  sc = TRUE       # Schoenfeld residuals (if available)
)
```

**Visual Signs of PH Violation**:
- Crossing survival curves
- Changing hazard ratios over time
- Non-random patterns in Schoenfeld residuals

### Solutions for PH Violations

#### Option 1: Stratified Cox Model

```{r stratified-cox}
# When PH assumption is violated for a categorical variable
# Use stratification approach in multivariable analysis
stratified_result <- multisurvival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = c(treatment, age, sex),
  # Note: Stratification would need to be implemented
  # This is conceptual - actual implementation may vary
)
```

#### Option 2: Time-Dependent Effects

For variables with time-varying effects, consider:
- Landmark analyses at multiple time points
- Piecewise exponential models
- Flexible parametric models

## Model Selection and Variable Selection

### Forward/Backward Selection

```{r variable-selection}
# Start with univariate screening
candidate_vars <- c("age", "sex", "stage", "grade", "biomarker")
univariate_results <- list()

for(var in candidate_vars) {
  univariate_results[[var]] <- survival(
    data = mydata,
    elapsedtime = time_months,
    outcome = death,
    outcomeLevel = "1",
    explanatory = !!sym(var),
    ph_cox = TRUE
  )
}

# Variables with p < 0.20 in univariate analysis
# proceed to multivariable model
significant_vars <- c("age", "stage", "biomarker")  # Based on results

final_model <- multisurvival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = significant_vars
)
```

### Model Validation

```{r model-validation}
# Internal validation using bootstrap or cross-validation
# External validation in independent dataset

validation_result <- multisurvival(
  data = validation_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = final_model_variables
)
```

## Handling Missing Data

### Complete Case Analysis

```{r complete-case}
# Default approach - uses only complete cases
complete_result <- survival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment
)
```

### Multiple Imputation Approach

```{r multiple-imputation}
# Conceptual approach for multiple imputation
# 1. Create multiple imputed datasets
# 2. Analyze each dataset separately
# 3. Pool results using Rubin's rules

# Example workflow (implementation details may vary):
imputed_results <- list()
for(i in 1:5) {  # 5 imputed datasets
  imputed_results[[i]] <- survival(
    data = imputed_data[[i]],
    elapsedtime = time_months,
    outcome = death,
    outcomeLevel = "1",
    explanatory = treatment
  )
}

# Pool results (conceptual)
pooled_hr <- mean(sapply(imputed_results, function(x) x$hazard_ratio))
pooled_se <- sqrt(mean(sapply(imputed_results, function(x) x$se^2)) + 
                  (1 + 1/5) * var(sapply(imputed_results, function(x) x$hazard_ratio)))
```

# Cut-point Optimization

## Methodological Considerations

### Multiple Cut-point Testing

```{r multiple-cutpoints}
# When testing multiple cut-points, adjust for multiple comparisons
cutpoint_analysis <- survivalcont(
  data = biomarker_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  contexpl = biomarker_value,
  findcut = TRUE,
  # Consider correction for multiple testing
  padjustmethod = "holm"
)
```

### Cross-Validation of Cut-points

```{r cutpoint-validation}
# Split data into training and validation sets
set.seed(123)
train_idx <- sample(nrow(biomarker_data), nrow(biomarker_data) * 0.7)
train_data <- biomarker_data[train_idx, ]
test_data <- biomarker_data[-train_idx, ]

# Find cut-point in training data
train_cutpoint <- survivalcont(
  data = train_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  contexpl = biomarker_value,
  findcut = TRUE
)

# Apply cut-point to test data
test_data$biomarker_high <- ifelse(test_data$biomarker_value >= optimal_cutpoint, 
                                   "High", "Low")

validation_result <- survival(
  data = test_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = biomarker_high
)
```

## Alternative Approaches to Dichotomization

### Tertiles or Quartiles

```{r tertiles}
# Instead of single cut-point, use tertiles
biomarker_data$biomarker_tertile <- cut(biomarker_data$biomarker_value,
                                        breaks = quantile(biomarker_data$biomarker_value,
                                                         c(0, 1/3, 2/3, 1), na.rm = TRUE),
                                        labels = c("Low", "Medium", "High"),
                                        include.lowest = TRUE)

tertile_analysis <- survival(
  data = biomarker_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = biomarker_tertile,
  analysistype = "overall"
)
```

### Spline Analysis

```{r spline-concept}
# For truly continuous relationships
# Consider restricted cubic splines or other flexible approaches
# This would require additional implementation beyond basic jsurvival
```

# Time-Dependent Covariates

## Landmark Analysis Implementation

### Multiple Landmark Points

```{r multiple-landmarks}
# Analyze survival at different landmark times
landmark_times <- c(6, 12, 24)  # months
landmark_results <- list()

for(t in landmark_times) {
  # Create landmark dataset
  landmark_data <- subset(original_data, 
                         time_months > t | (time_months <= t & death == 0))
  
  # Adjust survival times
  landmark_data$time_from_landmark <- pmax(0, landmark_data$time_months - t)
  
  # Analyze
  landmark_results[[paste0("Month_", t)]] <- survival(
    data = landmark_data,
    elapsedtime = time_from_landmark,
    outcome = death,
    outcomeLevel = "1",
    explanatory = response_status,
    uselandmark = TRUE,
    landmark = t
  )
}
```

### Dynamic Prediction

```{r dynamic-prediction}
# Conditional survival probabilities
# Probability of surviving additional 2 years given survival to 1 year

# Patients alive at 12 months
alive_12m <- subset(mydata, time_months > 12 | (time_months <= 12 & death == 0))
alive_12m$time_conditional <- pmax(0, alive_12m$time_months - 12)

conditional_survival <- singlearm(
  data = alive_12m,
  elapsedtime = time_conditional,
  outcome = death,
  outcomeLevel = "1",
  cutp = "24",  # Additional 2 years
  timetypeoutput = "months"
)
```

# Competing Risks Analysis

## Cause-Specific Hazards

```{r cause-specific}
# Separate analyses for different causes of death
# Cancer-specific mortality
cancer_death <- survival(
  data = competing_data,
  elapsedtime = time_months,
  outcome = death_cancer,  # 1 = cancer death, 0 = alive or other death
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE
)

# Other-cause mortality
other_death <- survival(
  data = competing_data,
  elapsedtime = time_months,
  outcome = death_other,   # 1 = other death, 0 = alive or cancer death
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE
)
```

## Subdistribution Hazards (Fine-Gray Model)

```{r fine-gray-concept}
# Fine-Gray model for cumulative incidence
# This approach treats competing events as censoring at the time they occur
# but keeps subjects in the risk set

# Implementation would require specialized competing risks functions
# Currently beyond basic jsurvival functionality
```

# Sample Size and Power Calculations

## Post-hoc Power Analysis

```{r power-analysis}
# Given observed data, calculate achieved power
observed_events <- 120
observed_hr <- 0.75
alpha <- 0.05

# Use standard formulas or specialized software for power calculation
# Example calculation (conceptual):
log_hr <- log(observed_hr)
se_log_hr <- 1.96 / abs(qnorm(alpha/2))  # Approximate from CI
power <- pnorm(abs(log_hr)/se_log_hr - qnorm(1-alpha/2))
```

## Required Sample Size

```{r sample-size}
# For planning future studies
target_hr <- 0.70        # Clinically meaningful difference
alpha <- 0.05            # Type I error
power <- 0.80            # Desired power
accrual_time <- 24       # months
followup_time <- 36      # months
median_survival <- 30    # months in control group

# Use specialized software or formulas for calculation
# Required events ≈ 4 * (Z_α/2 + Z_β)² / (log(HR))²
```

# Meta-Analysis of Survival Data

## Individual Patient Data Meta-Analysis

```{r ipd-meta}
# Combine multiple datasets
combined_data <- rbind(
  transform(study1_data, study = "Study1"),
  transform(study2_data, study = "Study2"),
  transform(study3_data, study = "Study3")
)

# Stratified analysis by study
meta_result <- multisurvival(
  data = combined_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = c(treatment, age, sex, study)
  # Consider stratification by study
)
```

## Fixed Effects vs. Random Effects

```{r meta-approaches}
# Fixed effects: assumes same treatment effect across studies
fixed_effects <- survival(
  data = combined_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment
)

# Random effects: allows heterogeneity between studies
# Would require specialized meta-analysis functions
```

# Quality Control and Validation

## Data Quality Checks

```{r quality-checks}
# Check for data inconsistencies
data_checks <- list(
  # Negative survival times
  negative_times = sum(mydata$time_months < 0, na.rm = TRUE),
  
  # Events after last follow-up
  impossible_events = sum(mydata$death == 1 & mydata$time_months > max_followup),
  
  # Missing key variables
  missing_time = sum(is.na(mydata$time_months)),
  missing_event = sum(is.na(mydata$death)),
  
  # Extreme values
  extreme_times = sum(mydata$time_months > 200, na.rm = TRUE)  # >16 years
)

print(data_checks)
```

## Model Diagnostics

```{r model-diagnostics}
# Concordance index (C-index)
# Measures discriminative ability of the model
# Values >0.7 generally considered good

# Calibration assessment
# Agreement between predicted and observed survival

# Model comparison using AIC/BIC
# Lower values indicate better fit
```

# Reporting and Interpretation

## Effect Size Interpretation

```{r effect-interpretation}
# Hazard Ratio Interpretation:
# HR = 0.50: 50% reduction in hazard (strong effect)
# HR = 0.75: 25% reduction in hazard (moderate effect)
# HR = 0.90: 10% reduction in hazard (small effect)
# HR = 1.00: No effect
# HR = 1.25: 25% increase in hazard (moderate harm)
# HR = 2.00: 100% increase in hazard (strong harm)

# Number Needed to Treat (NNT) calculation
survival_control <- 0.60  # 5-year survival in control
survival_treatment <- 0.70  # 5-year survival in treatment
absolute_benefit <- survival_treatment - survival_control
nnt <- 1 / absolute_benefit  # Number needed to treat
```

## Confidence Interval Interpretation

```{r ci-interpretation}
# HR = 0.75 (95% CI: 0.60-0.95)
# Interpretation:
# - Point estimate suggests 25% reduction in hazard
# - We can be 95% confident the true HR is between 0.60 and 0.95
# - Since CI excludes 1.0, result is statistically significant
# - Minimum plausible benefit is 5% (HR=0.95)
# - Maximum plausible benefit is 40% (HR=0.60)
```

# Conclusion

Advanced survival analysis requires careful consideration of:

1. **Model assumptions** and their validation
2. **Missing data patterns** and appropriate handling
3. **Multiple comparisons** and adjustment strategies
4. **Clinical relevance** beyond statistical significance
5. **Robust validation** in independent datasets

Key principles for advanced analysis:

- **Pre-specify analysis plans** to avoid data dredging
- **Validate findings** in independent cohorts when possible
- **Consider clinical context** in statistical decisions
- **Report limitations** and assumptions clearly
- **Collaborate with statisticians** for complex analyses

For complex analyses beyond the scope of basic jsurvival functions, consider:
- Specialized R packages (survival, survminer, rms, cmprsk)
- Statistical software with advanced survival capabilities
- Consultation with biostatisticians

Additional resources:
- [Survival Analysis Handbook](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html)
- [Clinical Prediction Models](https://www.clinicalpredictionmodels.org/)
- [STROBE Guidelines](https://www.strobe-statement.org/) for observational studies