---
title: "Troubleshooting Guide and FAQ for jjstatsplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting Guide and FAQ for jjstatsplot}
  %\VignetteEngine{rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, eval = FALSE)
```

# Troubleshooting Guide and FAQ

This comprehensive guide addresses common issues, error messages, and frequently asked questions when using jjstatsplot in both jamovi and R.

## Installation Issues

### Q: jjstatsplot doesn't appear in jamovi's module list

**Solutions:**
1. **Check jamovi version**: Requires jamovi ≥ 1.2.19
   ```
   Help → About jamovi (check version number)
   ```

2. **Refresh module library**:
   - Close jamovi completely
   - Restart jamovi
   - Try searching again in jamovi library

3. **Manual installation**:
   - Download `.jmo` file from GitHub releases
   - jamovi → Modules → Install from file

4. **Internet connection**: Ensure stable internet for library access

### Q: "Module failed to install" error

**Solutions:**
1. **Clear module cache**:
   - Close jamovi
   - Delete jamovi modules folder (varies by OS)
   - Restart jamovi and reinstall

2. **Permission issues**: Run jamovi as administrator (Windows) or check write permissions

3. **Dependency conflicts**: Uninstall conflicting modules temporarily

### Q: R installation fails with dependency errors

```{r r-install-issues}
# Common dependency installation issues:

# Issue: Package compilation fails
# Solution: Install binary packages
install.packages("jjstatsplot", type = "binary")

# Issue: ggstatsplot not found
# Solution: Install dependencies first
install.packages(c("ggstatsplot", "jmvcore", "R6"))
devtools::install_github("sbalci/jjstatsplot")

# Issue: devtools not available
# Solution: Install devtools first
install.packages("devtools")
```

## Data-Related Issues

### Q: "No data to plot" error

**Common Causes & Solutions:**

1. **Missing variable selection**:
   - **jamovi**: Ensure variables are dragged to appropriate boxes
   - **R**: Check variable names are correct and exist in data

2. **All missing values**:
   ```{r missing-data}
   # Check for missing data
   summary(your_data)
   
   # Remove rows with missing values for specific variables
   clean_data <- your_data[complete.cases(your_data[c("var1", "var2")]), ]
   ```

3. **Incorrect variable types**:
   ```{r variable-types}
   # Check variable types
   str(your_data)
   
   # Convert variable types if needed
   your_data$categorical_var <- factor(your_data$categorical_var)
   your_data$numeric_var <- as.numeric(your_data$numeric_var)
   ```

### Q: "Insufficient data for analysis" message

**Solutions:**
1. **Check sample sizes**:
   - Minimum 10 observations for most analyses
   - At least 5 observations per group for group comparisons
   - Correlation analyses need at least 10 complete pairs

2. **Group-specific issues**:
   ```{r sample-sizes}
   # Check group sizes
   table(your_data$grouping_variable)
   
   # Remove small groups if necessary
   group_counts <- table(your_data$grouping_variable)
   large_groups <- names(group_counts[group_counts >= 5])
   filtered_data <- your_data[your_data$grouping_variable %in% large_groups, ]
   ```

### Q: Variables show as "character" instead of "factor"

**Solution:**
```{r character-to-factor}
# In R: Convert character to factor
your_data$category <- factor(your_data$category)

# In jamovi: 
# 1. Double-click variable name in data view
# 2. Change "Data Type" from "Text" to "Nominal" or "Ordinal"
# 3. Set appropriate factor levels if needed
```

## Statistical Test Issues

### Q: "Statistical test failed" or "Assumptions violated" warnings

**Common Issues & Solutions:**

1. **Normality violations**:
   ```{r normality-issues}
   # Check normality
   shapiro.test(your_data$variable)
   
   # Solutions:
   # 1. Use non-parametric alternatives (usually automatic in jjstatsplot)
   # 2. Transform data
   log_transformed <- log(your_data$positive_variable)
   sqrt_transformed <- sqrt(your_data$variable)
   
   # 3. Use robust statistical methods (available in options)
   ```

2. **Homogeneity of variance violations**:
   ```{r variance-issues}
   # Check variance homogeneity
   bartlett.test(variable ~ group, data = your_data)
   
   # Solutions:
   # 1. Use Welch's t-test (often default)
   # 2. Use non-parametric tests
   # 3. Transform data to stabilize variance
   ```

3. **Small sample sizes for specific tests**:
   - Fisher's exact test for small contingency tables
   - Bootstrap methods for small samples
   - Non-parametric alternatives

### Q: Correlation analysis produces NaN or infinite values

**Solutions:**
```{r correlation-issues}
# Check for issues
summary(your_data[, numeric_variables])

# Remove constant variables
non_constant <- sapply(your_data[numeric_variables], function(x) length(unique(x)) > 1)
valid_variables <- numeric_variables[non_constant]

# Check for extreme outliers
outlier_check <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  extreme_outliers <- x < (Q1 - 3 * IQR) | x > (Q3 + 3 * IQR)
  return(sum(extreme_outliers, na.rm = TRUE))
}

sapply(your_data[valid_variables], outlier_check)
```

## Plot Display Issues

### Q: Plots appear blank or don't display

**Solutions:**

1. **jamovi-specific**:
   - Refresh analysis (click away and back)
   - Resize plot area
   - Check if plot is in "plot2" section for grouped analyses

2. **R-specific**:
   ```{r plot-display}
   # Check if plot object exists
   result <- jjhistostats(data = mtcars, dep = "mpg", grvar = NULL)
   
   # For single plots
   if (exists("result") && !is.null(result$plot)) {
     print(result$plot)
   }
   
   # For grouped plots
   if (exists("result") && !is.null(result$plot2)) {
     print(result$plot2)
   }
   ```

3. **Graphics device issues**:
   ```{r graphics-device}
   # Reset graphics device
   dev.off()
   
   # For RStudio users
   # Go to Plots pane → Export → Copy to Clipboard
   ```

### Q: Plot text is too small or overlapping

**Solutions:**

1. **In jamovi**: 
   - Increase plot size in results panel
   - Export plot at higher resolution

2. **In R**:
   ```{r plot-sizing}
   # Adjust plot size
   result <- jjhistostats(data = mtcars, dep = "mpg", grvar = NULL)
   
   # Modify plot theme
   improved_plot <- result$plot + 
     ggplot2::theme(
       text = ggplot2::element_text(size = 12),
       axis.text = ggplot2::element_text(size = 10),
       plot.title = ggplot2::element_text(size = 14)
     )
   
   print(improved_plot)
   ```

### Q: Statistical annotations are cut off or overlapping

**Solutions:**
```{r annotation-issues}
# Adjust plot margins
library(ggplot2)

result <- jjscatterstats(data = mtcars, dep = "mpg", group = "hp")

# Increase plot margins
fixed_plot <- result$plot + 
  theme(
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt")
  )

# For very long annotations, consider:
# 1. Shorter variable names
# 2. Custom subtitle/caption
# 3. Separate statistical table
```

## Performance Issues

### Q: Analysis takes very long time or jamovi freezes

**Solutions:**

1. **Large datasets**:
   ```{r large-data}
   # Sample large datasets
   if (nrow(your_data) > 10000) {
     sampled_data <- your_data[sample(nrow(your_data), 5000), ]
   }
   
   # Or focus on specific subsets
   subset_data <- your_data[your_data$important_condition == "focus_group", ]
   ```

2. **Complex grouping**:
   - Reduce number of groups
   - Combine similar categories
   - Use simpler analyses for exploration

3. **Memory issues**:
   ```{r memory-issues}
   # Check memory usage
   object.size(your_data)
   
   # Clean workspace
   rm(list = ls()[!ls() %in% c("essential_objects")])
   gc()  # Garbage collection
   ```

## Common Error Messages

### "Error in eval(expr, envir, enclos): object 'variable_name' not found"

**Cause**: Variable name mismatch or typo

**Solutions:**
```{r variable-not-found}
# Check available variables
names(your_data)

# Check for exact spelling
"variable_name" %in% names(your_data)

# Common issues:
# - Case sensitivity: "Variable" vs "variable"
# - Spaces: "Variable Name" vs "Variable_Name"
# - Special characters: accents, symbols
```

### "Error: Aesthetics must be either length 1 or the same as the data"

**Cause**: Data length mismatch, often from grouping issues

**Solutions:**
```{r aesthetic-error}
# Check data consistency
table(your_data$grouping_variable, useNA = "always")

# Remove missing values in grouping variable
clean_data <- your_data[!is.na(your_data$grouping_variable), ]

# Ensure balanced groups if needed
balanced_data <- your_data %>%
  group_by(grouping_variable) %>%
  filter(n() >= 5) %>%
  ungroup()
```

### "Warning: Computation failed in `stat_*()`: Non-numeric variable"

**Cause**: Numeric variable treated as categorical or vice versa

**Solutions:**
```{r numeric-variable-error}
# Check variable type
class(your_data$problematic_variable)

# Convert to appropriate type
your_data$numeric_var <- as.numeric(as.character(your_data$numeric_var))
your_data$factor_var <- factor(your_data$factor_var)

# Handle factor levels that look numeric
your_data$category <- factor(your_data$category, 
                            levels = c("Low", "Medium", "High"))
```

## Best Practices to Avoid Issues

### Data Preparation

```{r data-prep-best-practices}
# 1. Always check data structure first
str(your_data)
summary(your_data)

# 2. Handle missing values explicitly
your_data_clean <- your_data %>%
  filter(!is.na(dependent_variable) & !is.na(grouping_variable))

# 3. Set appropriate variable types
your_data$categorical <- factor(your_data$categorical)
your_data$ordered_cat <- ordered(your_data$ordered_cat, 
                                levels = c("Low", "Med", "High"))

# 4. Check for outliers
boxplot(your_data$continuous_variable)

# 5. Verify sufficient sample sizes
table(your_data$grouping_variable)
```

### Analysis Workflow

1. **Start simple**: Begin with basic plots before adding grouping
2. **Check assumptions**: Use diagnostic plots and tests
3. **Validate results**: Compare with alternative methods
4. **Document choices**: Keep track of data transformations and exclusions

### Reporting Guidelines

1. **Report sample sizes**: Always include n for each group
2. **Describe missing data**: How were missing values handled?
3. **State assumptions**: Were statistical assumptions met?
4. **Include effect sizes**: Don't rely only on p-values
5. **Show uncertainty**: Include confidence intervals when available

## Getting Additional Help

### jamovi Community Resources
- **jamovi Forum**: [forum.jamovi.org](https://forum.jamovi.org)
- **jamovi Documentation**: [docs.jamovi.org](https://docs.jamovi.org)
- **Video Tutorials**: Search "jamovi tutorials" on YouTube

### ggstatsplot Resources
- **Package Website**: [indrajeetpatil.github.io/ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/)
- **GitHub Issues**: For ggstatsplot-specific problems
- **Vignettes**: Comprehensive function documentation

### jjstatsplot Specific
- **GitHub Issues**: [github.com/sbalci/ClinicoPathJamoviModule/issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues)
- **Email Support**: drserdarbalci@gmail.com
- **Package Documentation**: Use `?function_name` in R

### When Reporting Issues

Include the following information:
1. **jamovi version** (Help → About)
2. **jjstatsplot version**
3. **Operating system**
4. **Sample data** (if possible)
5. **Exact error message**
6. **Steps to reproduce**
7. **Screenshots** (for jamovi issues)

### Creating Minimal Reproducible Examples

```{r reproducible-example}
# For R issues, create minimal example:
library(ClinicoPath)

# Use built-in data
data(mtcars)

# Minimal code that reproduces the problem
result <- jjhistostats(data = mtcars, dep = "mpg", grvar = NULL)
result$plot  # If this fails, report it
```

This troubleshooting guide covers the most common issues encountered with jjstatsplot. For complex statistical questions, consider consulting with a statistician or posting detailed questions on statistical forums like Cross Validated (stats.stackexchange.com).