---
title: "Data Preparation for Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation for Survival Analysis}
  %\VignetteEngine{rmarkdown::html_vignette}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

Proper data preparation is crucial for reliable survival analysis. This vignette provides comprehensive guidance on preparing data for use with jsurvival, covering common data issues and their solutions.

# Essential Data Components

## Required Variables

Every survival analysis requires at minimum:

1. **Time variable**: Time from study entry to event or censoring
2. **Event indicator**: Binary variable indicating whether event occurred

## Optional Variables

- **Grouping factors**: For comparing survival between groups
- **Continuous predictors**: For cut-point analysis or multivariable models
- **Date variables**: For calculating time intervals
- **Stratification variables**: For adjusted analyses

# Time Variable Preparation

## Calculating Time from Dates

### Using Built-in Time Interval Calculator

The jsurvival module includes a time interval calculator:

```{r time-calculator}
# In jamovi: Survival → Data Preparation → Time Interval Calculator
# Assign:
# - Start Date: Study entry date (e.g., diagnosis_date)
# - End Date: Event/censoring date (e.g., death_date or last_followup_date)
# - Output Unit: months or years

# This creates a new variable with calculated time intervals
```

### Manual Calculation in R

```{r manual-time-calc}
library(lubridate)

# Example data with dates
clinical_data <- data.frame(
  patient_id = 1:100,
  diagnosis_date = as.Date("2020-01-01") + sample(0:365, 100, replace = TRUE),
  death_date = as.Date("2020-01-01") + sample(200:1000, 100, replace = TRUE),
  last_followup_date = as.Date("2023-12-31")
)

# Calculate time to death or censoring
clinical_data$death_occurred <- !is.na(clinical_data$death_date)
clinical_data$event_date <- ifelse(clinical_data$death_occurred,
                                   clinical_data$death_date,
                                   clinical_data$last_followup_date)
clinical_data$event_date <- as.Date(clinical_data$event_date, origin = "1970-01-01")

# Time in months
clinical_data$time_months <- as.numeric(
  difftime(clinical_data$event_date, clinical_data$diagnosis_date, units = "days")
) / 30.44  # Average days per month

# Time in years
clinical_data$time_years <- as.numeric(
  difftime(clinical_data$event_date, clinical_data$diagnosis_date, units = "days")
) / 365.25  # Accounting for leap years
```

## Handling Time Zero Issues

### Left Truncation

```{r left-truncation}
# Patients entering study at different times
# Example: analyzing from landmark time (e.g., 3 months post-diagnosis)

# Include only patients who survived to landmark
landmark_time <- 3  # months
eligible_data <- subset(clinical_data, time_months >= landmark_time)

# Adjust time origin to landmark
eligible_data$time_from_landmark <- eligible_data$time_months - landmark_time
```

### Negative Time Values

```{r negative-times}
# Check for and handle negative times
negative_times <- clinical_data$time_months < 0
if(any(negative_times, na.rm = TRUE)) {
  print(paste("Found", sum(negative_times, na.rm = TRUE), "negative time values"))
  
  # Options:
  # 1. Exclude these patients
  clinical_data <- clinical_data[!negative_times | is.na(negative_times), ]
  
  # 2. Set to small positive value
  # clinical_data$time_months[negative_times] <- 0.001
  
  # 3. Investigate data entry errors
  # View(clinical_data[negative_times, ])
}
```

# Event Indicator Preparation

## Creating Binary Event Variables

### Death as Event

```{r death-event}
# From death status variable
clinical_data$death_binary <- ifelse(clinical_data$vital_status == "Dead", 1, 0)

# From death date
clinical_data$death_binary <- ifelse(is.na(clinical_data$death_date), 0, 1)

# Verify coding
table(clinical_data$death_binary, clinical_data$vital_status, useNA = "always")
```

### Other Events

```{r other-events}
# Disease progression
clinical_data$progression <- ifelse(clinical_data$progression_status == "Progressive Disease", 1, 0)

# Recurrence
clinical_data$recurrence <- ifelse(!is.na(clinical_data$recurrence_date), 1, 0)

# Composite endpoint (death or progression, whichever comes first)
clinical_data$death_or_progression <- pmax(clinical_data$death_binary, 
                                          clinical_data$progression, 
                                          na.rm = TRUE)
```

## Multiple Event Types

### Competing Risks Setup

```{r competing-risks-setup}
# Create separate indicators for different event types
clinical_data$death_cancer <- ifelse(clinical_data$death_cause == "Cancer", 1, 0)
clinical_data$death_other <- ifelse(clinical_data$death_cause == "Other" | 
                                   clinical_data$death_cause == "Cardiovascular", 1, 0)

# For patients still alive, both should be 0
alive_patients <- clinical_data$vital_status == "Alive"
clinical_data$death_cancer[alive_patients] <- 0
clinical_data$death_other[alive_patients] <- 0

# Verify: sum should equal total deaths
table(clinical_data$death_cancer + clinical_data$death_other, 
      clinical_data$vital_status)
```

# Data Quality Checks

## Consistency Checks

```{r consistency-checks}
# Function to perform comprehensive data quality checks
check_survival_data <- function(data, time_var, event_var) {
  
  checks <- list()
  
  # 1. Missing values
  checks$missing_time <- sum(is.na(data[[time_var]]))
  checks$missing_event <- sum(is.na(data[[event_var]]))
  
  # 2. Negative or zero times
  checks$negative_times <- sum(data[[time_var]] <= 0, na.rm = TRUE)
  
  # 3. Event indicator values
  event_values <- unique(data[[event_var]])
  checks$valid_events <- all(event_values %in% c(0, 1, NA))
  checks$event_table <- table(data[[event_var]], useNA = "always")
  
  # 4. Extreme values
  checks$extreme_times <- list(
    max_time = max(data[[time_var]], na.rm = TRUE),
    min_time = min(data[[time_var]], na.rm = TRUE),
    outliers = sum(data[[time_var]] > quantile(data[[time_var]], 0.99, na.rm = TRUE), na.rm = TRUE)
  )
  
  # 5. Event rate
  checks$event_rate <- mean(data[[event_var]], na.rm = TRUE)
  
  return(checks)
}

# Run checks
data_quality <- check_survival_data(clinical_data, "time_months", "death_binary")
print(data_quality)
```

## Outlier Detection

```{r outlier-detection}
# Identify potential outliers in survival times
Q1 <- quantile(clinical_data$time_months, 0.25, na.rm = TRUE)
Q3 <- quantile(clinical_data$time_months, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define outliers using 1.5 * IQR rule
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- clinical_data$time_months < lower_bound | clinical_data$time_months > upper_bound
print(paste("Found", sum(outliers, na.rm = TRUE), "potential outliers"))

# Review outliers
if(sum(outliers, na.rm = TRUE) > 0) {
  outlier_data <- clinical_data[outliers & !is.na(outliers), 
                               c("patient_id", "time_months", "death_binary")]
  print(outlier_data)
}
```

# Handling Missing Data

## Missing Data Patterns

```{r missing-patterns}
# Assess missing data patterns
library(VIM)  # Visualization and Imputation of Missing values

# Create missingness indicator
clinical_data$time_missing <- is.na(clinical_data$time_months)
clinical_data$event_missing <- is.na(clinical_data$death_binary)

# Missing data summary
missing_summary <- clinical_data %>%
  summarise(
    n_total = n(),
    time_missing = sum(time_missing),
    event_missing = sum(event_missing),
    both_missing = sum(time_missing & event_missing),
    complete_cases = sum(!time_missing & !event_missing)
  )

print(missing_summary)
```

## Strategies for Missing Data

### Complete Case Analysis

```{r complete-case}
# Use only patients with complete survival data
complete_data <- clinical_data[!is.na(clinical_data$time_months) & 
                              !is.na(clinical_data$death_binary), ]

print(paste("Complete cases:", nrow(complete_data), "of", nrow(clinical_data)))
```

### Administrative Censoring

```{r admin-censoring}
# For missing death dates, censor at last known follow-up
clinical_data$time_months_imputed <- clinical_data$time_months
clinical_data$death_binary_imputed <- clinical_data$death_binary

# If death status missing but last follow-up available, assume alive
missing_death <- is.na(clinical_data$death_binary) & !is.na(clinical_data$last_followup_date)
if(any(missing_death)) {
  clinical_data$death_binary_imputed[missing_death] <- 0
  clinical_data$time_months_imputed[missing_death] <- 
    as.numeric(difftime(clinical_data$last_followup_date[missing_death],
                       clinical_data$diagnosis_date[missing_death], 
                       units = "days")) / 30.44
}
```

# Explanatory Variable Preparation

## Categorical Variables

### Factor Ordering

```{r factor-ordering}
# Ensure proper ordering for categorical variables
clinical_data$tumor_grade <- factor(clinical_data$tumor_grade, 
                                   levels = c("Grade 1", "Grade 2", "Grade 3", "Grade 4"),
                                   ordered = TRUE)

clinical_data$stage <- factor(clinical_data$stage,
                             levels = c("Stage I", "Stage II", "Stage III", "Stage IV"),
                             ordered = TRUE)

# For unordered factors, set reference level
clinical_data$treatment <- factor(clinical_data$treatment)
clinical_data$treatment <- relevel(clinical_data$treatment, ref = "Standard")
```

### Handling Rare Categories

```{r rare-categories}
# Combine rare categories to ensure adequate sample sizes
table(clinical_data$histology)

# Combine categories with <10 patients
rare_histologies <- names(table(clinical_data$histology))[table(clinical_data$histology) < 10]
clinical_data$histology_grouped <- as.character(clinical_data$histology)
clinical_data$histology_grouped[clinical_data$histology_grouped %in% rare_histologies] <- "Other"
clinical_data$histology_grouped <- factor(clinical_data$histology_grouped)

table(clinical_data$histology_grouped)
```

## Continuous Variables

### Age Preparation

```{r age-preparation}
# Check age distribution
summary(clinical_data$age_at_diagnosis)

# Create age groups if needed
clinical_data$age_group <- cut(clinical_data$age_at_diagnosis,
                              breaks = c(0, 50, 65, 80, Inf),
                              labels = c("≤50", "51-65", "66-80", ">80"),
                              include.lowest = TRUE)

# Age as decades
clinical_data$age_decades <- floor(clinical_data$age_at_diagnosis / 10) * 10
```

### Biomarker Preparation

```{r biomarker-prep}
# Log transformation for skewed biomarkers
clinical_data$log_biomarker <- log10(clinical_data$biomarker_value + 1)

# Standardization (z-score)
clinical_data$biomarker_zscore <- scale(clinical_data$biomarker_value)[,1]

# Handle extreme values
biomarker_99th <- quantile(clinical_data$biomarker_value, 0.99, na.rm = TRUE)
clinical_data$biomarker_winsorized <- pmin(clinical_data$biomarker_value, biomarker_99th)
```

# Data Validation

## Cross-Validation with Source Data

```{r data-validation}
# Compare calculated times with expected ranges
validation_checks <- function(data) {
  issues <- list()
  
  # Check if all events occurred before end of study
  study_end <- as.Date("2023-12-31")
  late_events <- data$event_date > study_end & data$death_binary == 1
  issues$late_events <- sum(late_events, na.rm = TRUE)
  
  # Check if follow-up times are reasonable
  unreasonable_followup <- data$time_months > 240  # > 20 years
  issues$long_followup <- sum(unreasonable_followup, na.rm = TRUE)
  
  # Check for impossible age at events
  impossible_age <- data$age_at_diagnosis < 0 | data$age_at_diagnosis > 120
  issues$impossible_age <- sum(impossible_age, na.rm = TRUE)
  
  return(issues)
}

validation_results <- validation_checks(clinical_data)
print(validation_results)
```

## Final Dataset Preparation

```{r final-prep}
# Create final analysis dataset
analysis_data <- clinical_data %>%
  # Include only complete cases for primary analysis
  filter(!is.na(time_months), !is.na(death_binary)) %>%
  # Remove patients with negative or zero follow-up time
  filter(time_months > 0) %>%
  # Remove patients with missing key covariates
  filter(!is.na(age_at_diagnosis), !is.na(stage)) %>%
  # Select relevant variables
  select(
    patient_id,
    time_months,
    death_binary,
    age_at_diagnosis,
    age_group,
    sex,
    stage,
    tumor_grade,
    treatment,
    biomarker_value,
    histology_grouped
  )

# Final data summary
cat("Final analysis dataset:\n")
cat("N =", nrow(analysis_data), "\n")
cat("Events =", sum(analysis_data$death_binary), "\n")
cat("Event rate =", round(mean(analysis_data$death_binary) * 100, 1), "%\n")
cat("Median follow-up =", round(median(analysis_data$time_months), 1), "months\n")
```

# Export for jamovi

## CSV Export

```{r csv-export}
# Export cleaned data for use in jamovi
write.csv(analysis_data, "survival_analysis_data.csv", row.names = FALSE)

# Include data dictionary
data_dictionary <- data.frame(
  variable = names(analysis_data),
  description = c(
    "Patient identifier",
    "Time from diagnosis to death or censoring (months)",
    "Death indicator (1=death, 0=alive/censored)",
    "Age at diagnosis (years)",
    "Age group categories",
    "Patient sex",
    "Cancer stage",
    "Tumor grade",
    "Treatment received",
    "Biomarker value",
    "Histology type (grouped)"
  ),
  type = c(
    "Identifier",
    "Continuous",
    "Binary",
    "Continuous", 
    "Categorical",
    "Categorical",
    "Ordinal",
    "Ordinal",
    "Categorical",
    "Continuous",
    "Categorical"
  )
)

write.csv(data_dictionary, "data_dictionary.csv", row.names = FALSE)
```

# Quality Assurance Checklist

## Pre-Analysis Checklist

Before proceeding with survival analysis, verify:

- [ ] Time variable contains no negative or zero values
- [ ] Event indicator is properly coded (0/1 or No/Yes)
- [ ] Missing data patterns have been assessed and addressed
- [ ] Categorical variables have appropriate reference levels
- [ ] Continuous variables have been checked for outliers
- [ ] Data types are correct (numeric, factor, date)
- [ ] Sample size is adequate (rule of thumb: 10 events per variable)
- [ ] Follow-up time is sufficient for meaningful analysis

## Common Data Preparation Mistakes

1. **Incorrect time calculation**: Using calendar time instead of time from study entry
2. **Wrong event coding**: Mixing up 0 and 1 for events and censoring
3. **Including impossible values**: Negative times, events after last follow-up
4. **Ignoring competing risks**: Treating all deaths as same event type
5. **Improper handling of missing data**: Excluding too many patients without justification
6. **Not checking assumptions**: Ignoring data distribution and outliers

# Conclusion

Proper data preparation is essential for reliable survival analysis results. Key principles include:

1. **Careful time calculation** with appropriate handling of dates and time zero
2. **Consistent event coding** with validation against source data
3. **Comprehensive quality checks** to identify and resolve data issues
4. **Appropriate handling of missing data** based on missingness patterns
5. **Proper preparation of explanatory variables** with attention to clinical meaning

Taking time for thorough data preparation will ensure more reliable and interpretable survival analysis results in jsurvival.