---
title: "Independent vs Dependent Sampling: Statistical Implications for Pathology Adequacy Analysis"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Independent vs Dependent Sampling: Statistical Implications for Pathology Adequacy Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What the analysis estimates

`pathsampling` describes sampling and detection in the supplied cases. Its first-detection
curves use cases in which a lesion was eventually recorded. A missing first-detection position
means that no detection was recorded; it does not establish an independently verified
negative disease status. The recorded positive fraction and the conditional detection curve
therefore do not identify sensitivity in a population with known disease status.

The observation window matters. Cases examined for different numbers of samples may have
different opportunities for detection. Neither a geometric fit to observed detections nor
resampling those detections recovers lesions missed throughout the available window.

# Independence is an assumption about outcomes

Separate anatomical structures are not necessarily statistically independent. Lymph nodes
from the same case can share disease burden, and consecutive tissue samples can share tissue
features. Sampling without replacement also differs from independent Bernoulli sampling.
Choose a model using the sampling design and its assumptions, rather than anatomy alone.

For independent trials with the same positive probability `q`, the probability of at least one
positive result in `n` trials is `1 - (1 - q)^n`. The geometric distribution describes the position
of the first positive result under the same assumptions. R counts failures before the first
success, so its geometric quantile needs one added to obtain a sample position.

```{r independent_example}
q <- 0.2
n <- 5
stats::pbinom(0, size = n, prob = q, lower.tail = FALSE)
stats::qgeom(0.95, prob = q) + 1
```

These give a probability of 0.67232 and a model-based target position of 14. The latter is a
calculation under the assumed `q`, not a validated clinical sampling requirement.

## Dependence can change detection in either direction

Dependence does not imply systematic underestimation by the independence model. Consider two
trials, each with marginal positive probability 0.2. If their outcomes always agree, their
probability of any positive result is 0.2. If positive results are mutually exclusive, it is 0.4.
Independence gives 0.36, between these two possibilities.

```{r dependence_example}
data.frame(
  relationship = c("Always agree", "Independent", "Mutually exclusive positives"),
  probability_any_positive = c(0.2, 1 - (1 - 0.2)^2, 0.2 + 0.2)
)
```

These are probability examples, not assumptions about a particular tissue or disease.

## Selecting observed detections changes the fitted distribution

Suppose the underlying first positive position follows a geometric distribution with `q = 0.1`,
but observation stops after five samples. Fitting `1 / mean(first position)` only to the detected
cases estimates from a truncated distribution. Even with infinitely many cases, that estimate
need not equal the underlying `q`.

```{r observation_window}
positions <- 1:5
mass <- stats::dgeom(positions - 1, prob = 0.1)
probability_detected_by_five <- sum(mass)
mean_position_among_detected <- sum(positions * mass) / sum(mass)
c(
  underlying_q = 0.1,
  detected_fraction = probability_detected_by_five,
  inverse_observed_mean = 1 / mean_position_among_detected
)
```

The detected fraction is about 0.410. The inverse observed mean is about 0.358. This illustrates
why the analysis labels its fitted first-detection curve as an observed-positive working model.
It does not fit a censoring or ascertainment model.

# Methods available in pathsampling

| Output | Inputs and interpretation | Main limitation |
|---|---|---|
| Binomial/geometric prediction | A fitted per-sample `q`; probability of at least one positive sample | Constant-probability independence working model; observed-detection selection can bias the fitted `q` |
| Empirical cumulative curve and bootstrap | First-detection positions among eventually observed positives | Describes this selected cohort; does not correct unequal observation windows or missed lesions |
| Hypergeometric prediction | Known finite population `N` and positive count `K`; sampling without replacement | Requires those counts to be meaningful for the proposed sampling design |
| Beta-binomial prediction | Paired positive and total counts; fitted variation across case probabilities | Exchangeability and distributional assumptions remain; fitted parameters must be identifiable |
| Node-count adequacy summary | Fraction meeting a user-selected node-count threshold, with a Wilson interval | Describes compliance with the selected threshold, not a survival benefit or validated disease-specific requirement |
| Sample-position groups and gap indices | Lists of positive sample positions and a gap threshold | Sampling order is not anatomical distance; these heuristics do not establish lesions, stage, or treatment |

Hypergeometric probabilities are averaged across all eligible cases with equal case weights.
A case with fewer positive states than the requested detection target contributes probability
zero; it is not removed merely because the target is impossible in that case.

The beta-binomial model estimates a distribution of case probabilities from paired counts.
Warnings or fitting failures are reported, and failed fits do not produce predictions.

## Selecting the estimator

With `estimationMethod = "auto"`, the general, lymph-node, and omentum contexts use eligible
positive counts when supplied; otherwise they use the detected-case geometric estimate. Tumor
and margin contexts use the geometric estimate. Explicit `"geometric"` and `"empirical"` choices
are honored. Explicit empirical estimation requires eligible paired counts and does not
silently substitute the geometric estimator.

For count-based estimation, the numerator and denominator use the same eligible detected-case
rows. The analysis screens the coefficient of variation of the case-level positive proportions:
values above 0.3 produce a caution and values above 0.5 withhold the single-`q` predictions.
These cutoffs are heuristic screens, not a validated model-selection test. Subgroup estimates
use the selected estimator and the same screen within each group. Groups without an eligible
estimate retain their recorded counts and positive fraction.

The separate automatic between-group warning is enabled by default. It screens the CV of
inverse mean first-detection positions across sample types, irrespective of the prediction
estimator. It requires at least 10 detected cases overall and two groups with at least three
detected cases each. A CV above 0.3 triggers this descriptive warning, not a significance test.

The approximate chi-square goodness-of-fit calculation is available only for the detected-case
geometric estimator, after pooling sparse expected bins. It is a check of an untruncated working
model, conditional on the supplied detections. It does not validate independence, ascertainment,
or an empirical-count estimate of `q`.

## Bootstrap uncertainty

Bootstrap intervals resample complete observed-positive cases and evaluate the cumulative curve
at each sample position. The two bootstrap displays share their resamples. These intervals
reflect variation within that cohort; they are not assumption-free intervals for clinical
sensitivity. Correlation between cases would also require a resampling design that the current
analysis does not provide.

Use `setSeed = TRUE` and a specified `seedValue` for reproducible resampling. The analysis restores
the calling session's random-number state. The default is 10,000 bootstrap iterations when
bootstrap output is enabled. All reported bootstrap and Wilson intervals use a fixed 95% level;
`targetConfidence` specifies the target detection probability, not the interval confidence level.

# Preparing data and using the options

Use one row per case and integer counts. First-detection positions must lie between one and the
total number of examined samples; leave the position missing when detection was not recorded.
Positive counts must not exceed their paired totals. Invalid rows are excluded from the affected
calculation with an explanation, while row alignment is retained for optional variables.

Provide actual total foci through `totalFoci` for comparisons with the largest count in one
sample (`maxPositiveSingle`). Positive cassettes and foci have different units and are not
interchangeable. Lists of positive sample positions support sequence heuristics only.

Finite-population models, lymph-node summaries, and planning with an assumed `q` can run using
their own inputs without selecting first-detection variables. Advanced displays remain optional.
The model applicability guide explains available methods; it does not automatically establish
which model is clinically valid.

```{r analysis_example, eval = FALSE}
# Illustrative data: blocks examined per case and the position of the first positive block
set.seed(1)
cases <- data.frame(samples_examined = sample(4:10, 80, replace = TRUE))
first <- rgeom(80, prob = 0.35) + 1
cases$first_positive_position <- ifelse(first <= cases$samples_examined, first, NA)

ClinicoPath::pathsampling(
  data = cases,
  totalSamples = "samples_examined",
  firstDetection = "first_positive_position",
  estimationMethod = "geometric",
  showBootstrap = TRUE,
  bootstrapIterations = 10000,
  setSeed = TRUE,
  seedValue = 42
)
```

Interpret the outputs alongside the observation design and independently justified clinical
criteria. Descriptive yield bands, node thresholds, and sample-position groups do not establish
a stopping rule, disease stage, or a treatment recommendation.

# Statistical references

The probability definitions and R conventions are documented in the official R manuals for the
[geometric distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Geometric.html)
and the [hypergeometric distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Hypergeometric.html).
