---
title: "Getting Started with ClinicoPathDescriptives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ClinicoPathDescriptives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

ClinicoPathDescriptives is a comprehensive R package designed for descriptive analysis in clinicopathological research. It provides both programmatic R functions and a graphical user interface through jamovi, making advanced statistical analysis accessible to researchers with varying levels of programming experience.

### Key Features

- **Descriptive Statistics**: Generate comprehensive "Table One" summaries
- **Cross-tabulations**: Create contingency tables with appropriate statistical tests  
- **Visualization Tools**: Age pyramids, alluvial diagrams, Venn diagrams, and more
- **Treatment Response Analysis**: Waterfall and swimmer plots for oncology research
- **Data Quality Assessment**: Benford's law analysis for fraud detection
- **Dual Interface**: Use in R or through jamovi's point-and-click interface

## Installation

### Install from GitHub

```{r eval=FALSE}
# Install development version
if (!require("remotes")) install.packages("remotes")
remotes::install_github("sbalci/ClinicoPathDescriptives")
```

### Install as jamovi Module

Download the `.jmo` file from [Releases](https://github.com/sbalci/ClinicoPathDescriptives/releases) and install in jamovi via: `jamovi menu > Modules > Install from file`

## Loading the Package

```{r setup}
library(ClinicoPath)
library(dplyr)    # For data manipulation
library(ggplot2)  # For additional plotting
```

## Example Dataset

The package includes a comprehensive histopathology dataset with 250 cancer patients and 38 variables including demographics, pathological features, biomarkers, and outcomes.

```{r}
# Load the example dataset
data(histopathology)

# Examine the structure
str(histopathology, give.attr = FALSE)
```

```{r}
# Quick overview of key variables
histopathology %>%
  select(Age, Sex, Grade, TStage, Group, Outcome, OverallTime) %>%
  head(10)
```

## Basic Descriptive Analysis

### Table One

Generate a comprehensive descriptive summary table - the cornerstone of clinical research manuscripts:

```{r}
# Basic Table One
tableone(
  data = histopathology,
  vars = c("Age", "Sex", "Grade", "TStage", "LVI", "PNI"),
  group = "Group"
)
```

### Continuous Variables Summary

Generate detailed summaries for continuous variables:

```{r}
summarydata(
  data = histopathology,
  vars = c("Age", "OverallTime", "MeasurementA"),
  group = "Grade"
)
```

### Categorical Variables Summary

Analyze categorical variables with counts and percentages:

```{r}
reportcat(
  data = histopathology,
  vars = c("Sex", "Grade", "TStage", "LVI", "PNI"),
  group = "Group"
)
```

## Cross-tabulation Analysis

Create contingency tables with appropriate statistical tests:

```{r}
crosstable(
  data = histopathology,
  rows = "Grade",
  cols = "Group",
  total = TRUE,
  expected = TRUE,
  percentages = "row"
)
```

```{r}
# Multiple variables cross-tabulation
crosstable(
  data = histopathology,
  rows = c("Grade", "TStage"),
  cols = "LymphNodeMetastasis",
  total = TRUE,
  percentages = "column"
)
```

## Data Visualization

### Age Pyramid

Visualize age and sex distribution:

```{r}
agepyramid(
  data = histopathology,
  age = "Age",
  sex = "Sex",
  group = "Group"
)
```

### Alluvial Diagrams

Show relationships between categorical variables:

```{r}
alluvial(
  data = histopathology,
  vars = c("Grade", "TStage", "LymphNodeMetastasis", "Outcome")
)
```

### Venn Diagrams

Analyze overlap between sets:

```{r}
# Create binary indicators for Venn diagram
histopathology_venn <- histopathology %>%
  mutate(
    LVI_present = ifelse(LVI == "Present", 1, 0),
    PNI_present = ifelse(PNI == "Present", 1, 0),
    LN_positive = ifelse(LymphNodeMetastasis == "Present", 1, 0)
  )

venn(
  data = histopathology_venn,
  vars = c("LVI_present", "PNI_present", "LN_positive"),
  names = c("LVI", "PNI", "LN Metastasis")
)
```

## Data Quality Assessment

### Benford's Law Analysis

Assess data quality using Benford's law - useful for detecting fabricated data:

```{r}
# Analyze first digits of continuous measurements
benford(
  data = histopathology,
  var = "MeasurementA"
)
```

### Variable Tree Summary

Get a hierarchical overview of your dataset:

```{r}
vartree(
  data = histopathology,
  vars = c("Grade", "TStage", "LVI", "PNI", "LymphNodeMetastasis")
)
```

## Working with Treatment Response Data

Load the treatment response dataset for oncology-specific analyses:

```{r}
# Load treatment response data
data(treatmentResponse)
head(treatmentResponse)
```

### Waterfall Plot

Visualize individual patient treatment responses:

```{r}
waterfall(
  data = treatmentResponse,
  patient_id = "PatientID",
  response = "ResponseValue"
)
```

## Integration with jamovi

All functions in this package are also available through jamovi's graphical interface:

1. **Install the module** in jamovi
2. **Navigate to**: `Exploration > ClinicoPath Descriptives`
3. **Select your analysis** function
4. **Configure options** through the GUI
5. **Generate results** with point-and-click ease

### jamovi Advantages

- No programming required
- Interactive parameter adjustment
- Real-time result updates
- Export capabilities for reports
- Accessible to non-programmers

## Next Steps

This vignette covered the basic functionality of ClinicoPathDescriptives. For more advanced usage, see:

- **Clinical Workflow Vignette**: Complete analysis workflows for clinical research
- **Treatment Response Analysis**: Advanced oncology analysis techniques  
- **Data Quality and Exploration**: Comprehensive data assessment methods
- **Visualization Gallery**: Advanced plotting techniques and customization

## Citation

```{r eval=FALSE}
citation("ClinicoPathDescriptives")
```

## Support

- 📖 [Documentation](https://www.serdarbalci.com/ClinicoPathDescriptives/)
- 🐛 [Report Issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues)
- 🌐 [jamovi Module Hub](https://sbalci.github.io/ClinicoPathJamoviModule/)