---
title: "Comprehensive Guide to IHC Expression Analysis with ihcstats"
subtitle: "Implementing Four Research Methodologies for Immunohistochemical Data Analysis"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide to IHC Expression Analysis with ihcstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 150
)
```

## Introduction

The `ihcstats` function in the ClinicoPath jamovi module provides comprehensive immunohistochemical (IHC) expression analysis by implementing methodologies from four landmark research papers:

1. **Matsuoka et al. (2011)** - Ward's hierarchical clustering for colorectal cancer prognosis
2. **Carvalho et al. (2011)** - Iterative marker selection for renal oncocytoma diagnosis  
3. **Olsen et al. (2006)** - Differential diagnosis clustering for sarcomas
4. **Sterlacci et al. (2019)** - TIL signature analysis for NSCLC

This vignette demonstrates how to use all features of the `ihcstats` function with comprehensive synthetic datasets that cover all implemented methodologies.

## Loading Required Libraries and Data

```{r libraries}
library(ClinicoPath)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cluster)
library(knitr)
library(DT)

# Load the comprehensive dataset
data("ihc_comprehensive_data")

# Load specialized datasets
data("sarcoma_ihc_data")
data("nsclc_til_data") 
data("colorectal_ihc_data")
data("renal_ihc_data")
```

## Dataset Overview

Our comprehensive dataset includes **300 patients** with **60 variables** covering:

- **8 tumor types** for differential diagnosis
- **4 NSCLC histotypes** for TIL analysis
- **34 IHC markers** with different scoring scales
- **Multi-region tumor data** (central vs invasive)
- **Survival outcomes** and clinical variables
- **TIL quantification** and immune signatures

```{r data_overview}
# Dataset dimensions
cat("Dataset dimensions:", dim(ihc_comprehensive_data), "\n")

# Tumor type distribution
tumor_table <- table(ihc_comprehensive_data$Tumor_Type)
kable(data.frame(Tumor_Type = names(tumor_table), Count = as.numeric(tumor_table)),
      caption = "Distribution of Tumor Types")

# IHC scoring scales represented
cat("\nIHC Scoring Scales in Dataset:\n")
cat("- Standard (0, 1+, 2+, 3+): Core sarcoma markers\n")
cat("- Matsuoka (Mild, Moderate, Marked): Colorectal markers\n") 
cat("- Carvalho (0, 1, 2): Renal markers\n")
cat("- Binary (Negative, Positive): Selected markers\n")
```

## Basic IHC Expression Analysis

### Example 1: Standard Hierarchical Clustering

Let's start with a basic analysis using sarcoma markers with the standard IHC scoring scale.

```{r basic_analysis}
# Basic hierarchical clustering with sarcoma markers
basic_result <- ihcstats(
  data = sarcoma_ihc_data,
  markers = c("Vimentin", "SMA", "Desmin", "S100", "CD34"),
  clusterMethod = "hierarchical",
  distanceMetric = "gower",
  linkageMethod = "average",
  nClusters = 4,
  showDendrogram = TRUE,
  showHeatmap = TRUE,
  computeHScore = TRUE
)

# Display H-Score results
kable(basic_result$hscoreTable$asDF, 
      caption = "H-Score Analysis for Sarcoma Markers")
```

### Example 2: Different Distance Metrics and Clustering Methods

```{r clustering_methods}
# Compare different clustering approaches
methods_comparison <- list(
  "Hierarchical + Gower" = ihcstats(
    data = sarcoma_ihc_data,
    markers = c("Vimentin", "SMA", "Desmin", "S100"),
    clusterMethod = "hierarchical",
    distanceMetric = "gower",
    nClusters = 3
  ),
  
  "Hierarchical + Jaccard" = ihcstats(
    data = sarcoma_ihc_data,
    markers = c("Vimentin", "SMA", "Desmin", "S100"),
    clusterMethod = "hierarchical", 
    distanceMetric = "jaccard",
    nClusters = 3
  ),
  
  "PAM Clustering" = ihcstats(
    data = sarcoma_ihc_data,
    markers = c("Vimentin", "SMA", "Desmin", "S100"),
    clusterMethod = "pam",
    nClusters = 3
  ),
  
  "K-means" = ihcstats(
    data = sarcoma_ihc_data,
    markers = c("Vimentin", "SMA", "Desmin", "S100"),
    clusterMethod = "kmeans",
    nClusters = 3
  )
)

# Compare cluster sizes
cluster_comparison <- data.frame(
  Method = names(methods_comparison),
  Cluster_1 = sapply(methods_comparison, function(x) x$clusterSummary$asDF$size[1]),
  Cluster_2 = sapply(methods_comparison, function(x) x$clusterSummary$asDF$size[2]),
  Cluster_3 = sapply(methods_comparison, function(x) x$clusterSummary$asDF$size[3])
)

kable(cluster_comparison, caption = "Comparison of Clustering Methods")
```

## Matsuoka 2011 Methodology: Prognostic Clustering

### Ward's Hierarchical Clustering with Survival Analysis

The Matsuoka approach uses Ward's method for prognostic clustering with survival outcomes.

```{r matsuoka_analysis}
# Matsuoka-style prognostic clustering
matsuoka_result <- ihcstats(
  data = colorectal_ihc_data,
  markers = c("p53", "Ki67", "VEGF", "COX2"),
  clusterMethod = "hierarchical",
  linkageMethod = "ward.D2",  # Ward's method
  scoringScale = "matsuoka",  # 3-tier system
  nClusters = 3,
  prognosticClustering = TRUE,
  survivalTimeVar = "Overall_Survival_Months",
  survivalEventVar = "Death_Event",
  showDendrogram = TRUE,
  showHeatmap = TRUE
)

# Display prognostic clustering results
kable(matsuoka_result$prognosticTable$asDF,
      caption = "Matsuoka Prognostic Clustering Results")

# Display survival analysis
kable(matsuoka_result$survivalTable$asDF,
      caption = "Cox Regression Analysis for Prognostic Groups")
```

### Multi-Region Tumor Analysis

The Matsuoka methodology includes analysis of central vs invasive tumor regions.

```{r multiregion_analysis}
# Multi-region analysis (central vs invasive)
regional_result <- ihcstats(
  data = colorectal_ihc_data,
  markers = c("p53", "Ki67", "VEGF", "COX2"),
  tumorRegionAnalysis = TRUE,
  centralRegionVar = c("p53_Central", "Ki67_Central", "VEGF_Central", "COX2_Central"),
  invasiveRegionVar = c("p53_Invasive", "Ki67_Invasive", "VEGF_Invasive", "COX2_Invasive"),
  scoringScale = "matsuoka"
)

# Display regional analysis results
kable(regional_result$regionalTable$asDF,
      caption = "Multi-Region Analysis: Central vs Invasive Front")
```

## Carvalho 2011 Methodology: Iterative Marker Selection

### Optimal Marker Panel Identification

The Carvalho approach uses iterative selection to identify optimal marker combinations.

```{r carvalho_analysis}
# Carvalho-style iterative marker selection
carvalho_result <- ihcstats(
  data = renal_ihc_data,
  markers = c("CD117", "Vimentin_Carvalho", "E_Cadherin", "CK7"),
  scoringScale = "carvalho",  # 0-2 scale
  iterativeClustering = TRUE,
  clusterMethod = "hierarchical",
  distanceMetric = "gower",
  nClusters = 3,
  showClusterValidation = TRUE
)

# Display optimal markers
kable(carvalho_result$optimalMarkersTable$asDF,
      caption = "Carvalho Optimal Marker Selection Results")
```

### PCA Analysis with K-means

```{r carvalho_pca}
# PCA + K-means approach (alternative Carvalho method)
carvalho_pca <- ihcstats(
  data = renal_ihc_data,
  markers = c("CD117", "Vimentin_Carvalho", "E_Cadherin", "CK7"),
  clusterMethod = "pca_kmeans",
  pcaAnalysis = TRUE,
  showPCAPlot = TRUE,
  standardizeData = TRUE,
  nClusters = 3
)

# Display PCA results
kable(carvalho_pca$pcaTable$asDF,
      caption = "Principal Component Analysis Results")
```

## Olsen 2006 Methodology: Differential Diagnosis

### Sarcoma Differential Diagnosis Clustering

The Olsen approach focuses on differential diagnosis using antibody panels.

```{r olsen_analysis}
# Olsen-style differential diagnosis
olsen_result <- ihcstats(
  data = sarcoma_ihc_data,
  markers = c("Vimentin", "SMA", "Desmin", "S100", "CD34", "Cytokeratin", "MDM2", "CDK4"),
  differentialDiagnosis = TRUE,
  tumorTypeVar = "Tumor_Type",
  clusterMethod = "hierarchical",
  linkageMethod = "complete",  # Complete linkage as in Olsen
  clusterCutHeight = 0.6,
  calculateDiagnosticMetrics = TRUE,
  antibodyOptimization = TRUE,
  olsenVisualization = TRUE
)

# Display differential diagnosis results
kable(head(olsen_result$differentialTable$asDF, 10),
      caption = "Olsen Differential Diagnosis Results (First 10 Cases)")
```

### Antibody Performance Analysis

```{r antibody_performance}
# Display antibody performance metrics
kable(olsen_result$antibodyPerformanceTable$asDF,
      caption = "Individual Antibody Diagnostic Performance")

# Display optimal panels
kable(head(olsen_result$optimalPanelTable$asDF, 5),
      caption = "Optimal Antibody Panel Combinations")
```

## Sterlacci 2019 Methodology: TIL Signature Analysis

### Comprehensive TIL Analysis

The Sterlacci approach focuses on tumor-infiltrating lymphocyte analysis with immune signatures.

```{r sterlacci_analysis}
# Sterlacci TIL signature analysis
sterlacci_result <- ihcstats(
  data = nsclc_til_data,
  markers = c("CD3", "CD4", "CD8", "CD20", "Granzyme_B", "PD1", "PDL1", "FOXP3"),
  sterlacciAnalysis = TRUE,
  immuneSignatureFocus = TRUE,
  tilAnalysisMode = "combined_til",
  clusterMethod = "hierarchical",
  linkageMethod = "complete",  # As in Sterlacci paper
  distanceMetric = "gower",    # Gower distance for TIL data
  nClusters = 3,
  multipleTesting = "bonferroni",
  significanceThreshold = 0.05
)

# Display Sterlacci TIL results
kable(sterlacci_result$sterlacciTable$asDF,
      caption = "Sterlacci TIL Signature Analysis Results")

# Display TIL signature characterization
kable(sterlacci_result$tilSignatureTable$asDF,
      caption = "TIL Signature Characterization by Cluster")
```

### Supervised Clustering by Histotype

```{r supervised_clustering}
# Supervised clustering within histotypes
supervised_result <- ihcstats(
  data = nsclc_til_data,
  markers = c("CD3", "CD4", "CD8", "Granzyme_B", "PD1", "PDL1"),
  supervisedClustering = TRUE,
  groupVariable = "Histotype",
  sterlacciAnalysis = TRUE,
  nClusters = 2  # Within each histotype
)

# Display supervised clustering results
kable(supervised_result$supervisedResultsTable$asDF,
      caption = "Supervised Clustering Results by Histotype")
```

### Reproducibility Testing

```{r reproducibility}
# Reproducibility analysis with Cohen's kappa
reproducibility_result <- ihcstats(
  data = nsclc_til_data,
  markers = c("CD4", "CD8", "Granzyme_B", "PD1"),
  reproductibilityTesting = TRUE,
  sterlacciAnalysis = TRUE,
  nClusters = 3
)

# Display reproducibility results
kable(reproducibility_result$reproductibilityTable$asDF,
      caption = "Cluster Reproducibility Analysis (Cohen's κ)")
```

## Advanced Visualization Options

### Comprehensive Visualization Analysis

```{r visualization_analysis}
# Full visualization suite
viz_result <- ihcstats(
  data = sarcoma_ihc_data,
  markers = c("Vimentin", "SMA", "Desmin", "S100", "CD34", "MDM2"),
  showDendrogram = TRUE,
  showHeatmap = TRUE,
  showPCAPlot = TRUE,
  showScoreDist = TRUE,
  showClusterValidation = TRUE,
  pcaAnalysis = TRUE,
  clusterMethod = "hierarchical",
  nClusters = 4,
  standardizeData = TRUE
)

cat("Visualization analysis completed with all plot types enabled.")
```

## Diagnostic Performance Analysis

### Comprehensive Diagnostic Metrics

```{r diagnostic_performance}
# Diagnostic performance with grouping variable
diagnostic_result <- ihcstats(
  data = sarcoma_ihc_data,
  markers = c("Vimentin", "SMA", "Desmin", "S100", "CD34"),
  showDiagnostics = TRUE,
  groupVariable = "Tumor_Type",
  calculateDiagnosticMetrics = TRUE
)

# Display diagnostic performance
kable(head(diagnostic_result$diagnosticTable$asDF, 15),
      caption = "Diagnostic Performance Analysis")
```

## Multiple Scoring Scales Comparison

### Comparing Different IHC Scoring Systems

```{r scoring_scales}
# Compare different scoring scales
scoring_comparison <- list(
  "Binary" = ihcstats(
    data = transform(sarcoma_ihc_data, 
                    Vimentin_Binary = ifelse(Vimentin %in% c("2+", "3+"), "Positive", "Negative")),
    markers = "Vimentin_Binary",
    scoringScale = "binary",
    nClusters = 2
  ),
  
  "Standard" = ihcstats(
    data = sarcoma_ihc_data,
    markers = "Vimentin",
    scoringScale = "standard",
    nClusters = 3
  ),
  
  "Matsuoka" = ihcstats(
    data = colorectal_ihc_data,
    markers = "p53",
    scoringScale = "matsuoka",
    nClusters = 3
  ),
  
  "Carvalho" = ihcstats(
    data = renal_ihc_data,
    markers = "CD117",
    scoringScale = "carvalho", 
    nClusters = 3
  )
)

cat("Scoring scale comparison completed for Binary, Standard, Matsuoka, and Carvalho systems.")
```

## Optimal K Selection Methods

### Comparing Cluster Number Selection Approaches

```{r optimal_k}
# Compare different optimal K methods
k_methods <- c("elbow", "silhouette", "gap")
k_results <- list()

for (method in k_methods) {
  k_results[[method]] <- ihcstats(
    data = sarcoma_ihc_data,
    markers = c("Vimentin", "SMA", "Desmin", "S100"),
    optimalKMethod = method,
    showClusterValidation = TRUE,
    nClusters = 4  # Will be optimized
  )
}

cat("Optimal K selection analysis completed using elbow, silhouette, and gap statistic methods.")
```

## Clinical Integration Examples

### Complete Clinical Workflow

```{r clinical_workflow}
# Complete clinical analysis workflow
clinical_result <- ihcstats(
  data = ihc_comprehensive_data,
  markers = c("Vimentin", "SMA", "CD34", "S100", "CD3", "CD8", "Ki67", "p53"),
  
  # Basic clustering
  clusterMethod = "hierarchical",
  linkageMethod = "ward.D2",
  distanceMetric = "gower",
  nClusters = 4,
  
  # Multiple analysis types
  prognosticClustering = TRUE,
  differentialDiagnosis = TRUE,
  sterlacciAnalysis = TRUE,
  
  # Variables for analysis
  survivalTimeVar = "Overall_Survival_Months",
  survivalEventVar = "Death_Event",
  tumorTypeVar = "Tumor_Type",
  groupVariable = "Risk_Category",
  
  # Advanced options
  iterativeClustering = TRUE,
  antibodyOptimization = TRUE,
  reproductibilityTesting = TRUE,
  
  # Visualization
  showDendrogram = TRUE,
  showHeatmap = TRUE,
  showPCAPlot = TRUE,
  computeHScore = TRUE,
  
  # Statistical options
  multipleTesting = "bonferroni",
  significanceThreshold = 0.05
)

# Summary of clinical workflow results
cat("Clinical workflow analysis completed with:\n")
cat("- Cluster analysis:", nrow(clinical_result$clusterSummary$asDF), "clusters identified\n")
cat("- H-score analysis:", nrow(clinical_result$hscoreTable$asDF), "markers analyzed\n")
if (nrow(clinical_result$prognosticTable$asDF) > 0) {
  cat("- Prognostic analysis: Completed\n")
}
if (nrow(clinical_result$sterlacciTable$asDF) > 0) {
  cat("- TIL signature analysis: Completed\n")
}
```

## Summary and Recommendations

### Key Features Summary

The `ihcstats` function provides comprehensive IHC analysis with the following key capabilities:

#### 1. **Multiple Research Methodologies**
- **Matsuoka 2011**: Ward's clustering + survival analysis
- **Carvalho 2011**: Iterative marker selection
- **Olsen 2006**: Differential diagnosis clustering  
- **Sterlacci 2019**: TIL signature analysis

#### 2. **Flexible Scoring Systems**
- Binary (Negative/Positive)
- Standard IHC (0, 1+, 2+, 3+)
- Carvalho scale (0, 1, 2)
- Matsuoka 3-tier (Mild/Moderate/Marked)
- H-score calculation (0-300)

#### 3. **Advanced Clustering Options**
- Hierarchical clustering (multiple linkage methods)
- PAM (Partitioning Around Medoids)
- K-means clustering
- PCA + K-means combination

#### 4. **Distance Metrics**
- Gower distance (mixed data types)
- Jaccard distance (categorical data)

#### 5. **Specialized Analyses**
- Multi-region tumor analysis
- Survival analysis integration
- Diagnostic performance metrics
- Reproducibility testing
- Antibody panel optimization

### Clinical Applications

#### **Pathology Practice Integration**

1. **Sarcoma Diagnosis** (Olsen methodology)
   - Use comprehensive antibody panels
   - Optimize marker combinations
   - Calculate diagnostic performance

2. **Lung Cancer TIL Analysis** (Sterlacci methodology)
   - Quantify immune infiltration
   - Characterize TIL signatures
   - Predict immunotherapy response

3. **Colorectal Cancer Prognosis** (Matsuoka methodology)
   - Identify prognostic clusters
   - Perform survival analysis
   - Analyze tumor heterogeneity

4. **Renal Tumor Diagnosis** (Carvalho methodology)
   - Iterative marker selection
   - Optimize diagnostic panels
   - Validate marker performance

### Best Practices

#### **Data Preparation**
```r
# Ensure factors are properly coded
data$marker <- as.factor(data$marker)

# Check for missing values
summary(data)

# Verify survival data format
# Time: numeric (months/years)
# Event: 0/1 or FALSE/TRUE
```

#### **Analysis Workflow**
```r
# 1. Start with basic clustering
result1 <- ihcstats(data, markers, nClusters = 3)

# 2. Add methodology-specific options
result2 <- ihcstats(data, markers, prognosticClustering = TRUE, 
                   survivalTimeVar = "time", survivalEventVar = "event")

# 3. Include visualizations
result3 <- ihcstats(data, markers, showDendrogram = TRUE, 
                   showHeatmap = TRUE, showPCAPlot = TRUE)
```

#### **Result Interpretation**
- **Cluster sizes**: Should be reasonably balanced
- **Silhouette values**: >0.5 indicates good clustering
- **Survival p-values**: <0.05 for significant prognostic value
- **Cohen's κ**: >0.6 indicates good reproducibility

### Troubleshooting

#### **Common Issues**
1. **Small cluster sizes**: Reduce number of clusters
2. **Poor clustering quality**: Try different distance metrics
3. **Missing survival data**: Exclude survival analysis options
4. **Factor level mismatches**: Ensure consistent coding

#### **Performance Optimization**
- Use focused marker panels (5-10 markers)
- Consider data size limitations
- Pre-filter relevant cases

This comprehensive guide demonstrates the full capabilities of the `ihcstats` function for modern pathology practice and research applications.

---

## References

The `ihcstats` function implements methodologies from the following key research papers:

### Primary IHC Clustering Methods

1. **Matsuoka T, Mitomi H, Fukui N, Kanazawa H, Saito T, Hayashi T, Yao T.** (2011). Cluster analysis of claudin-1 and -4, E-cadherin, and β-catenin expression in colorectal cancers. *Journal of Surgical Oncology*, 103(7), 674-686. doi: 10.1002/jso.21854. PMID: 21360533.

2. **Carvalho JC, Wasco MJ, Kunju LP, Thomas DG, Shah RB.** (2011). Cluster analysis of immunohistochemical profiles delineates CK7, vimentin, S100A1 and C-kit (CD117) as an optimal panel in the differential diagnosis of renal oncocytoma from its mimics. *Histopathology*, 58(2), 169-179. doi: 10.1111/j.1365-2559.2011.03753.x. PMID: 21323945.

3. **Olsen SH, Thomas DG, Lucas DR.** (2006). Cluster analysis of immunohistochemical profiles in synovial sarcoma, malignant peripheral nerve sheath tumor, and Ewing sarcoma. *Modern Pathology*, 19(5), 659-668. doi: 10.1038/modpathol.3800569. PMID: 16528378.

4. **Sterlacci W, Fiegl M, Juskevicius D, Tzankov A.** (2020). Cluster Analysis According to Immunohistochemistry is a Robust Tool for Non-Small Cell Lung Cancer and Reveals a Distinct, Immune Signature-defined Subgroup. *Applied Immunohistochemistry & Molecular Morphology*, 28(4), 274-283. doi: 10.1097/PAI.0000000000000751. PMID: 31058655.

### Related High-Risk Classification Methods

5. **Laas E, Ballester M, Cortez A, Graesslin O, Daraï E.** (2019). Unsupervised Clustering of Immunohistochemical Markers to Define High-Risk Endometrial Cancer. *Pathology & Oncology Research*, 25(2), 461-469. doi: 10.1007/s12253-017-0335-y. PMID: 29264761.

### Diagnostic Style Analysis

6. **Usubutun A, Mutter GL, Saglam A, Dolgun A, Ozkan EA, Ince T, Akyol A, Bulbul HD, Calay Z, Eren F, Gumurdulu D, Haberal AN, Ilvan S, Karaveli S, Koyuncuoglu M, Muezzinoglu B, Muftuoglu KH, Ozdemir N, Ozen O, Baykara S, Pestereli E, Ulukus EC, Zekioglu O.** (2012). Reproducibility of endometrial intraepithelial neoplasia diagnosis is good, but influenced by the diagnostic style of pathologists. *Modern Pathology*, 25(6), 877-884. doi: 10.1038/modpathol.2011.220. PMID: 22301705.

---

*For additional support, please refer to the ClinicoPath documentation or contact the development team.*