---
title: "jjcorrmat: Comprehensive Correlation Matrix Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

The `jjcorrmat` function provides a comprehensive interface for creating correlation matrices with advanced statistical analysis capabilities. This function is a wrapper around `ggstatsplot::ggcorrmat` and `ggstatsplot::grouped_ggcorrmat`, offering both single and grouped correlation matrix visualizations.

## Key Features

- **Multiple correlation methods**: Parametric (Pearson), nonparametric (Spearman), robust, and Bayesian
- **Grouped analysis**: Create separate correlation matrices for different groups
- **Performance optimized**: Uses internal caching to eliminate redundant computations
- **Comprehensive visualization**: High-quality correlation plots with statistical annotations

## Installation and Setup

```{r setup}
echo: true
eval: false

# Install ClinicoPath if not already installed
if (!require("ClinicoPath")) {
  devtools::install_github("sbalci/ClinicoPathJamoviModule")
}

library(ClinicoPath)
library(ggplot2)
```

## Quick Start

### Basic Correlation Matrix

```{r basic-example}
echo: true
eval: false

# Load test data
data(jjcorrmat_test_data)

# Basic correlation matrix
result <- jjcorrmat(
  data = jjcorrmat_test_data,
  dep = c("ki67_percent", "p53_score", "her2_intensity", "tumor_size_mm"),
  typestatistics = "parametric"
)

# View the plot
result$plot
```

### Grouped Correlation Matrix

```{r grouped-example}
echo: true
eval: false

# Grouped correlation matrix by tumor grade
result_grouped <- jjcorrmat(
  data = jjcorrmat_test_data,
  dep = c("ki67_percent", "p53_score", "her2_intensity"),
  grvar = "tumor_grade",
  typestatistics = "nonparametric"
)

# View the grouped plot
result_grouped$plot2
```

## Function Parameters

### Core Parameters

- **`data`**: Input data frame containing the variables to analyze
- **`dep`**: List of continuous variables for correlation analysis (must be numeric)
- **`grvar`**: Optional grouping variable for creating separate correlation matrices
- **`typestatistics`**: Type of correlation analysis to perform

### Statistical Methods

The `typestatistics` parameter supports four different approaches:

1. **"parametric"** (default): Pearson product-moment correlation
   - Assumes normal distribution
   - Best for linear relationships
   - Most commonly used

2. **"nonparametric"**: Spearman's rank correlation
   - Distribution-free method
   - Robust to outliers
   - Captures monotonic relationships

3. **"robust"**: Percentage bend correlation
   - Robust to outliers and non-normality
   - Good compromise between parametric and nonparametric
   - Uses WRS2::pbcor()

4. **"bayes"**: Bayesian correlation analysis
   - Provides Bayes factors for evidence assessment
   - Incorporates prior beliefs
   - Gives probabilistic interpretation

## Performance Optimizations

### Version History

The function has been significantly optimized for performance:

**Previous Issues:**
- Redundant data processing in `.plot()` and `.plot2()` methods
- Unused caching infrastructure
- Repeated formula construction and variable processing

**Current Optimizations:**
- **Data Caching**: Uses `.prepareData()` method to cache processed data
- **Options Caching**: Uses `.prepareOptions()` method to cache formula processing
- **Eliminated Redundancy**: Both plot methods now use cached results
- **Better Progress Feedback**: Clear user messaging during processing

### Performance Benefits

```{r performance-comparison}
echo: true
eval: false

# Performance comparison (conceptual)
# Before optimization: 
# - Data processed twice (once for each plot)
# - Formula construction repeated
# - Variable processing duplicated

# After optimization:
# - Data processed once and cached
# - Formula construction done once
# - Significant speedup for large datasets
```

## Advanced Usage Examples

### Multiple Statistical Methods

```{r multiple-methods}
echo: true
eval: false

# Compare different correlation methods
methods <- c("parametric", "nonparametric", "robust", "bayes")
variables <- c("ki67_percent", "p53_score", "her2_intensity")

results <- list()
for (method in methods) {
  results[[method]] <- jjcorrmat(
    data = jjcorrmat_test_data,
    dep = variables,
    typestatistics = method
  )
}

# Each result contains the correlation matrix plot
# results$parametric$plot
# results$nonparametric$plot
# etc.
```

### Clinical Research Example

```{r clinical-example}
echo: true
eval: false

# Analyze biomarker correlations in breast cancer data
biomarkers <- c("ki67_percent", "p53_score", "her2_intensity", "tumor_size_mm", "age_years")

# Overall correlation matrix
overall_corr <- jjcorrmat(
  data = jjcorrmat_test_data,
  dep = biomarkers,
  typestatistics = "parametric"
)

# Stratified by hormone receptor status
stratified_corr <- jjcorrmat(
  data = jjcorrmat_test_data,
  dep = biomarkers,
  grvar = "hormone_status",
  typestatistics = "parametric"
)

# Compare correlations across tumor grades
grade_corr <- jjcorrmat(
  data = jjcorrmat_test_data,
  dep = biomarkers[1:4],  # Exclude age for clarity
  grvar = "tumor_grade",
  typestatistics = "spearman"
)
```

## Data Requirements

### Input Data Structure

The input data should be a data frame with:

- **Continuous variables**: Numeric columns for correlation analysis
- **Grouping variables**: Factor or character columns for stratified analysis
- **Complete cases**: Missing values are automatically excluded

### Example Data Structure

```{r data-structure}
echo: true
eval: false

# Structure of test data
str(jjcorrmat_test_data)

# Key variables:
# - ki67_percent: Numeric (0-100)
# - p53_score: Numeric (0-50)
# - her2_intensity: Numeric (0-30)
# - tumor_size_mm: Numeric (5-50)
# - age_years: Numeric (18-90)
# - tumor_grade: Factor (Grade 1/2/3)
# - hormone_status: Factor (ER+/PR+, etc.)
```

## Best Practices

### Variable Selection

1. **Choose appropriate variables**: Select variables that theoretically should be correlated
2. **Check distributions**: Consider log transformation for skewed variables
3. **Handle missing data**: Decide on complete case analysis vs. imputation
4. **Sample size**: Ensure adequate sample size for stable correlations

### Statistical Method Selection

- **Use parametric** when variables are approximately normal
- **Use nonparametric** for ordinal data or non-normal distributions
- **Use robust** when there are outliers but you want parametric-like interpretation
- **Use Bayesian** when you want to quantify evidence for/against correlations

### Interpretation Guidelines

- **Correlation strength**: 
  - 0.00-0.30: Weak
  - 0.30-0.70: Moderate  
  - 0.70-1.00: Strong
- **Statistical significance**: Consider both p-values and effect sizes
- **Clinical significance**: Strong statistical correlation may not be clinically meaningful

## Troubleshooting

### Common Issues

1. **"Data contains no (complete) rows"**
   - Check for missing values in your variables
   - Ensure at least some complete cases exist

2. **No plot generated**
   - Verify that you have at least 2 continuous variables
   - Check that variables are properly formatted as numeric

3. **Slow performance**
   - The optimized version should be much faster
   - For very large datasets, consider sampling

### Error Messages

```{r error-handling}
echo: true
eval: false

# Example error handling
tryCatch({
  result <- jjcorrmat(
    data = my_data,
    dep = c("var1", "var2"),
    typestatistics = "parametric"
  )
}, error = function(e) {
  message("Error in correlation analysis: ", e$message)
  message("Check your data structure and variable types")
})
```

## Technical Details

### Underlying Functions

The `jjcorrmat` function is built on:

- **ggstatsplot::ggcorrmat**: For single correlation matrices
- **ggstatsplot::grouped_ggcorrmat**: For grouped analyses
- **jmvcore**: For data handling and formula processing

### Caching Implementation

```{r caching-details}
echo: true
eval: false

# Internal caching structure (conceptual)
# private$.processedData: Cached cleaned data
# private$.processedOptions: Cached formula and variable processing
# 
# Benefits:
# - Eliminates redundant jmvcore::naOmit() calls
# - Avoids repeated formula construction
# - Shares processed data between plot methods
```

## Conclusion

The optimized `jjcorrmat` function provides:

- **High performance**: Significant speed improvements through caching
- **Flexibility**: Multiple statistical methods and grouping options
- **Reliability**: Comprehensive error handling and validation
- **Usability**: Clear documentation and examples

The function is well-suited for clinical research, biomarker analysis, and any scenario requiring robust correlation matrix visualization.

## Session Information

```{r session-info}
sessionInfo()
```
