---
title: "Line Charts for Clinical Time Series and Trend Analysis"
author: "ClinicoPath Package"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    self-contained: true
execute:
  warning: false
  message: false
  echo: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  dpi = 300
)

# Load required packages
suppressPackageStartupMessages({
  library(jmvcore)
  library(ggplot2)
  library(dplyr)
  if (requireNamespace("ClinicoPath", quietly = TRUE)) {
    library(ClinicoPath)
  }
})
```

# Introduction to Line Charts in Clinical Research

## What are Line Charts?

Line charts are fundamental visualization tools that display data points connected by straight line segments, making them ideal for showing **trends and changes over time** or ordered categories. In clinical and pathological research, line charts are essential for:

- **Longitudinal studies**: Tracking patient outcomes over time
- **Treatment monitoring**: Visualizing response to interventions
- **Biomarker evolution**: Following laboratory values or disease markers
- **Quality metrics**: Monitoring healthcare performance indicators
- **Dose-response relationships**: Understanding therapeutic relationships

### Key Advantages

1. **Trend Identification**: Easily spot increasing, decreasing, or stable patterns
2. **Comparative Analysis**: Compare multiple groups or treatments simultaneously  
3. **Time Series Visualization**: Perfect for longitudinal and follow-up studies
4. **Statistical Integration**: Combine with trend lines, confidence intervals, and correlations
5. **Clinical Communication**: Intuitive format for presenting to clinicians and patients

## When to Use Line Charts

### Ideal Clinical Scenarios

- **Laboratory Values Over Time**: Hemoglobin, tumor markers, inflammatory indices
- **Treatment Response Monitoring**: Blood pressure, pain scores, functional assessments
- **Disease Progression**: Tumor size, cognitive decline, symptom severity
- **Quality Improvement**: Infection rates, readmission rates, patient satisfaction
- **Dose-Response Studies**: Efficacy or toxicity versus drug concentration

### Data Requirements

- **X-axis**: Time points, visit numbers, or ordered categories
- **Y-axis**: Continuous numeric outcomes (laboratory values, scores, measurements)
- **Optional Grouping**: Treatment arms, patient populations, hospital units
- **Minimum Data**: At least 3 time points for meaningful trend analysis

# Statistical Background

## Trend Analysis Methods

### Linear Trends
Line charts can incorporate **linear regression** to identify overall trends:

$$Y = \beta_0 + \beta_1 \times Time + \epsilon$$

Where:
- $\beta_1$ represents the rate of change per time unit
- Positive $\beta_1$: improving/increasing trend
- Negative $\beta_1$: declining/decreasing trend

### Correlation Analysis
The relationship between time and outcome can be quantified using:

- **Pearson correlation**: For linear relationships
- **Spearman correlation**: For monotonic (rank-based) relationships
- **R-squared**: Proportion of variance explained by time

### Confidence Intervals
Line charts can display **uncertainty** around trends:
- **Standard error bands**: Show precision of the line
- **Confidence intervals**: Typically 95% CI around the trend
- **Prediction intervals**: Bounds for future observations

## Clinical Interpretation Guidelines

### Trend Magnitude
```{r trend-interpretation, echo=FALSE}
trend_guide <- data.frame(
  `Change per Unit Time` = c("< 5%", "5-15%", "15-30%", "> 30%"),
  `Clinical Significance` = c("Minimal", "Small", "Moderate", "Large"),
  `Clinical Action` = c("Monitor", "Consider intervention", "Likely clinically meaningful", "Definite clinical impact"),
  `Examples` = c("Lab variation", "Modest improvement", "Notable response", "Major clinical change")
)
knitr::kable(trend_guide, caption = "Clinical Interpretation of Trend Magnitude")
```

### Correlation Strength
```{r correlation-interpretation, echo=FALSE}
correlation_guide <- data.frame(
  `Correlation (r)` = c("0.0 - 0.1", "0.1 - 0.3", "0.3 - 0.5", "0.5 - 0.7", "0.7 - 1.0"),
  `Strength` = c("Negligible", "Weak", "Moderate", "Strong", "Very Strong"),
  `Clinical Utility` = c("No trend", "Minimal trend", "Noticeable pattern", "Clear trend", "Highly predictable"),
  `R² (Variance Explained)` = c("< 1%", "1-9%", "9-25%", "25-49%", "49-100%")
)
knitr::kable(correlation_guide, caption = "Correlation Strength Interpretation")
```

# Clinical Examples

## Example 1: Hemoglobin Monitoring During Cancer Treatment

This example demonstrates monitoring anemia treatment in cancer patients receiving erythropoietin (EPO) therapy.

```{r hemoglobin-example}
# Create realistic hemoglobin data
set.seed(123)
n_patients <- 90
n_visits <- 6
visit_weeks <- c(0, 2, 4, 8, 12, 16)

# Generate patient data
hemoglobin_data <- expand.grid(
  patient_id = paste0("P", sprintf("%03d", 1:n_patients)),
  visit_week = visit_weeks
) %>%
  left_join(
    data.frame(
      patient_id = paste0("P", sprintf("%03d", 1:n_patients)),
      treatment_group = sample(c("Control", "EPO_Low", "EPO_High"), n_patients, 
                              replace = TRUE, prob = c(0.3, 0.4, 0.3)),
      baseline_hgb = rnorm(n_patients, mean = 9.5, sd = 1.2)
    ),
    by = "patient_id"
  ) %>%
  arrange(patient_id, visit_week)

# Generate hemoglobin trajectories
hemoglobin_data$hemoglobin_g_dl <- NA

for (i in 1:nrow(hemoglobin_data)) {
  week <- hemoglobin_data$visit_week[i]
  treatment <- hemoglobin_data$treatment_group[i]
  baseline <- hemoglobin_data$baseline_hgb[i]
  
  # Treatment-specific trends
  if (treatment == "Control") {
    trend <- -0.03 * week + rnorm(1, 0, 0.3)
  } else if (treatment == "EPO_Low") {
    trend <- 0.08 * week + rnorm(1, 0, 0.25)
  } else {
    trend <- 0.12 * week + rnorm(1, 0, 0.2)
  }
  
  hemoglobin_data$hemoglobin_g_dl[i] <- baseline + trend
}

# Ensure realistic ranges
hemoglobin_data$hemoglobin_g_dl <- pmax(6, pmin(16, hemoglobin_data$hemoglobin_g_dl))
hemoglobin_data$hemoglobin_g_dl <- round(hemoglobin_data$hemoglobin_g_dl, 1)

# Display data structure
str(hemoglobin_data)
head(hemoglobin_data, 12)
```

### Running Line Chart Analysis

```{r hemoglobin-analysis, eval=FALSE}
# In jamovi, you would select:
# Analyses → JJStatsPlot → Line Chart

# Programmatically (if running in R):
hemoglobin_result <- linechart(
  data = hemoglobin_data,
  xvar = "visit_week",
  yvar = "hemoglobin_g_dl",
  groupby = "treatment_group",
  confidence = TRUE,
  trendline = TRUE,
  points = TRUE,
  refline = 12,
  reflineLabel = "Normal Lower Limit",
  xlabel = "Weeks Since Treatment Start",
  ylabel = "Hemoglobin (g/dL)",
  title = "Hemoglobin Response to EPO Treatment",
  colorPalette = "clinical"
)
```

### Expected Clinical Results

1. **Visual Patterns**:
   - Control group: Slight decline or stable low values
   - EPO Low dose: Gradual improvement over time
   - EPO High dose: Steeper improvement, reaching normal range

2. **Statistical Analysis**:
   - Strong positive correlation for EPO groups (r > 0.7)
   - Weak negative correlation for control group
   - Significant trend p-values for treatment groups

3. **Clinical Interpretation**:
   - EPO therapy effectively treats cancer-related anemia
   - Higher doses provide faster hemoglobin recovery
   - Reference line helps identify patients reaching normal range

## Example 2: Blood Pressure Response to Interventions

This example shows hypertension management comparing lifestyle interventions, medications, and combined therapy.

```{r blood-pressure-example}
# Create blood pressure monitoring data
set.seed(456)
n_subjects <- 75
n_months <- 8
months <- 0:(n_months-1)

bp_data <- expand.grid(
  subject_id = paste0("BP", sprintf("%03d", 1:n_subjects)),
  month = months
) %>%
  left_join(
    data.frame(
      subject_id = paste0("BP", sprintf("%03d", 1:n_subjects)),
      intervention = sample(c("Lifestyle", "Medication", "Combined"), n_subjects, 
                           replace = TRUE, prob = c(0.35, 0.35, 0.3)),
      baseline_sbp = rnorm(n_subjects, mean = 155, sd = 15)
    ),
    by = "subject_id"
  ) %>%
  arrange(subject_id, month)

# Generate systolic BP trajectories
bp_data$systolic_bp <- NA

for (i in 1:nrow(bp_data)) {
  month <- bp_data$month[i]
  intervention <- bp_data$intervention[i]
  baseline <- bp_data$baseline_sbp[i]
  
  # Intervention-specific effects
  if (intervention == "Lifestyle") {
    reduction <- 2 * sqrt(month) + rnorm(1, 0, 2)
  } else if (intervention == "Medication") {
    reduction <- ifelse(month <= 3, 8 * (month / 3), 8 + 2 * ((month - 3) / 5)) + rnorm(1, 0, 3)
  } else {
    reduction <- ifelse(month <= 2, 6 * (month / 2), 6 + 8 * ((month - 2) / 6)) + rnorm(1, 0, 2)
  }
  
  bp_data$systolic_bp[i] <- baseline - reduction
}

# Ensure realistic ranges
bp_data$systolic_bp <- pmax(90, pmin(200, bp_data$systolic_bp))
bp_data$systolic_bp <- round(bp_data$systolic_bp)

# Display summary
cat("Blood Pressure Study Summary:\n")
cat("Subjects:", length(unique(bp_data$subject_id)), "\n")
cat("Interventions:", paste(unique(bp_data$intervention), collapse = ", "), "\n")
cat("Baseline SBP range:", min(bp_data[bp_data$month == 0, "systolic_bp"]), "-", 
    max(bp_data[bp_data$month == 0, "systolic_bp"]), "mmHg\n")
cat("Final SBP range:", min(bp_data[bp_data$month == max(bp_data$month), "systolic_bp"]), "-", 
    max(bp_data[bp_data$month == max(bp_data$month), "systolic_bp"]), "mmHg\n")
```

### Blood Pressure Analysis Setup

```{r bp-analysis-setup, eval=FALSE}
bp_result <- linechart(
  data = bp_data,
  xvar = "month",
  yvar = "systolic_bp",
  groupby = "intervention",
  confidence = TRUE,
  trendline = TRUE,
  refline = 140,
  reflineLabel = "Hypertension Threshold",
  xlabel = "Months Since Intervention Start",
  ylabel = "Systolic Blood Pressure (mmHg)",
  title = "Blood Pressure Response to Different Interventions",
  colorPalette = "colorblind",
  theme = "publication"
)
```

### Expected Clinical Findings

1. **Intervention Comparisons**:
   - **Lifestyle**: Gradual, sustained reduction
   - **Medication**: Rapid initial improvement, then plateau
   - **Combined**: Best overall reduction combining both benefits

2. **Clinical Milestones**:
   - Time to reach target BP (<140 mmHg)
   - Magnitude of reduction from baseline
   - Sustainability of improvements

## Example 3: Biomarker Response to Immunotherapy

This example tracks inflammatory biomarkers (IL-6, TNF-α) during cancer immunotherapy to assess treatment response.

```{r biomarker-example}
# Create biomarker immunotherapy data
set.seed(789)
n_patients <- 45
visit_days <- c(0, 21, 42, 84, 126, 168)

biomarker_data <- expand.grid(
  patient_id = paste0("IMM", sprintf("%03d", 1:n_patients)),
  visit_day = visit_days
) %>%
  left_join(
    data.frame(
      patient_id = paste0("IMM", sprintf("%03d", 1:n_patients)),
      response_type = sample(c("Responder", "Non_Responder", "Progressive"), n_patients,
                            replace = TRUE, prob = c(0.4, 0.35, 0.25)),
      baseline_il6 = rlnorm(n_patients, meanlog = 2, sdlog = 0.8)
    ),
    by = "patient_id"
  ) %>%
  arrange(patient_id, visit_day)

# Generate IL-6 trajectories
biomarker_data$il6_pg_ml <- NA

for (i in 1:nrow(biomarker_data)) {
  day <- biomarker_data$visit_day[i]
  response <- biomarker_data$response_type[i]
  baseline <- biomarker_data$baseline_il6[i]
  
  # Response-specific patterns
  if (response == "Responder") {
    factor <- exp(-0.003 * day) * (1 + rnorm(1, 0, 0.2))
  } else if (response == "Non_Responder") {
    factor <- 1 + 0.1 * sin(day / 50) * (1 + rnorm(1, 0, 0.3))
  } else {
    factor <- 1 + 0.002 * day * (1 + rnorm(1, 0, 0.4))
  }
  
  biomarker_data$il6_pg_ml[i] <- baseline * factor
}

# Ensure positive values
biomarker_data$il6_pg_ml <- pmax(0.1, pmin(100, biomarker_data$il6_pg_ml))
biomarker_data$il6_pg_ml <- round(biomarker_data$il6_pg_ml, 2)

# Calculate summary by response type
response_summary <- biomarker_data %>%
  filter(visit_day %in% c(0, 168)) %>%
  select(patient_id, response_type, visit_day, il6_pg_ml) %>%
  tidyr::pivot_wider(names_from = visit_day, values_from = il6_pg_ml, names_prefix = "day_") %>%
  mutate(percent_change = round(100 * (day_168 - day_0) / day_0, 1)) %>%
  group_by(response_type) %>%
  summarise(
    n = n(),
    mean_baseline = round(mean(day_0, na.rm = TRUE), 1),
    mean_final = round(mean(day_168, na.rm = TRUE), 1),
    mean_change = round(mean(percent_change, na.rm = TRUE), 1),
    .groups = 'drop'
  )

print(response_summary)
```

### Biomarker Analysis

```{r biomarker-analysis, eval=FALSE}
biomarker_result <- linechart(
  data = biomarker_data,
  xvar = "visit_day",
  yvar = "il6_pg_ml", 
  groupby = "response_type",
  confidence = TRUE,
  smooth = TRUE,  # Use smoothed lines for biomarker data
  xlabel = "Days Since Treatment Start",
  ylabel = "IL-6 (pg/mL)",
  title = "IL-6 Response Patterns During Immunotherapy",
  colorPalette = "viridis"
)
```

### Clinical Biomarker Interpretation

1. **Response Patterns**:
   - **Responders**: Progressive decline in inflammatory markers
   - **Non-responders**: Stable levels with minimal change  
   - **Progressive disease**: Increasing inflammatory activity

2. **Predictive Value**:
   - Early biomarker changes may predict long-term response
   - Useful for treatment modification decisions
   - Complement to imaging and clinical assessments

# Advanced Line Chart Features

## Confidence Intervals and Uncertainty

### Standard Error vs. Confidence Intervals
```{r confidence-intervals, eval=FALSE}
# Line chart with confidence intervals
result <- linechart(
  data = clinical_data,
  xvar = "time_point",
  yvar = "biomarker_level",
  groupby = "treatment",
  confidence = TRUE,     # Shows 95% CI around lines
  smooth = FALSE,        # Linear vs. smoothed confidence bands
  theme = "publication"
)
```

**Clinical Applications**:
- **Uncertainty quantification**: Show precision of measurements
- **Group comparisons**: Overlapping CI suggests similar trends
- **Sample size effects**: Wider CI with smaller groups
- **Statistical significance**: Non-overlapping CI indicates differences

## Reference Lines and Clinical Thresholds

### Normal Ranges and Target Values
```{r reference-lines, eval=FALSE}
# Add clinical reference lines
result <- linechart(
  data = lab_data,
  xvar = "visit_week", 
  yvar = "hemoglobin_g_dl",
  refline = 12,                    # Normal hemoglobin threshold
  reflineLabel = "Normal Range",   # Clear clinical labeling
  title = "Hemoglobin Monitoring with Clinical Targets"
)
```

**Common Clinical References**:
- **Laboratory normal ranges**: Hemoglobin, liver enzymes, kidney function
- **Blood pressure targets**: <140/90 mmHg for hypertension
- **Pain score thresholds**: ≤3/10 for adequate pain control
- **Quality metrics**: Infection rates, readmission benchmarks

## Statistical Trend Analysis

### Linear vs. Non-linear Trends
```{r trend-analysis, eval=FALSE}
# Compare linear and smoothed trends
linear_result <- linechart(
  data = longitudinal_data,
  xvar = "time",
  yvar = "outcome", 
  trendline = TRUE,    # Linear regression line
  smooth = FALSE       # Straight line connections
)

smooth_result <- linechart(
  data = longitudinal_data,
  xvar = "time",
  yvar = "outcome",
  smooth = TRUE,       # LOESS smoothing
  confidence = TRUE    # Smoothed confidence bands
)
```

**When to Use Each**:
- **Linear trends**: Constant rate of change, dose-response relationships
- **Smoothed trends**: Complex patterns, seasonal effects, non-linear responses

# Clinical Applications by Specialty

## Oncology Applications

### Treatment Response Monitoring
```{r oncology-examples, echo=FALSE}
oncology_apps <- data.frame(
  `Clinical Scenario` = c(
    "Tumor marker trends",
    "Chemotherapy response", 
    "Quality of life during treatment",
    "Biomarker-guided therapy",
    "Survivorship monitoring"
  ),
  `X-axis Variable` = c(
    "Months since diagnosis",
    "Treatment cycles",
    "Weeks since treatment start", 
    "Days since targeted therapy",
    "Years post-treatment"
  ),
  `Y-axis Variable` = c(
    "PSA, CA-125, CEA levels",
    "Tumor size (RECIST)",
    "EORTC QLQ scores",
    "PD-L1, circulating tumor DNA",
    "Late effects scores"
  ),
  `Grouping Variable` = c(
    "Tumor stage/grade",
    "Treatment regimen",
    "Performance status",
    "Biomarker status",
    "Treatment modality"
  )
)
knitr::kable(oncology_apps, caption = "Oncology Line Chart Applications")
```

### Key Clinical Insights
- **Early response prediction**: Biomarker changes predict long-term outcomes
- **Treatment optimization**: Adjust therapy based on trends
- **Quality of life**: Balance efficacy with patient-reported outcomes

## Cardiology Applications

### Cardiovascular Risk Management
```{r cardiology-examples, echo=FALSE}
cardiology_apps <- data.frame(
  `Clinical Scenario` = c(
    "Blood pressure management",
    "Lipid control monitoring",
    "Heart failure progression",
    "Post-MI recovery",
    "Device monitoring"
  ),
  `Outcome Measures` = c(
    "SBP/DBP (mmHg)",
    "LDL cholesterol (mg/dL)",
    "EF%, BNP levels",
    "Exercise capacity, LVEF",
    "ICD/pacemaker parameters"
  ),
  `Clinical Targets` = c(
    "<140/90 mmHg",
    "<70 mg/dL (high risk)",
    "Stable/improved EF",
    "Return to baseline function",
    "Appropriate device function"
  ),
  `Typical Timeline` = c(
    "Months to years",
    "3-6 month intervals", 
    "Quarterly assessments",
    "6-12 months post-MI",
    "Continuous monitoring"
  )
)
knitr::kable(cardiology_apps, caption = "Cardiology Monitoring Applications")
```

## Critical Care Applications

### ICU Monitoring and Outcomes
```{r critical-care-examples, echo=FALSE}
icu_apps <- data.frame(
  `Monitoring Domain` = c(
    "Hemodynamic stability",
    "Respiratory function", 
    "Organ function recovery",
    "Sedation management",
    "Nutritional status"
  ),
  `Key Variables` = c(
    "MAP, cardiac output",
    "P/F ratio, PEEP levels",
    "Creatinine, liver enzymes", 
    "RASS scores, delirium",
    "Albumin, weight changes"
  ),
  `Time Resolution` = c(
    "Hours to days",
    "Hourly to daily",
    "Daily assessments",
    "Shift-based scoring",
    "Daily to weekly"
  ),
  `Clinical Goals` = c(
    "Hemodynamic optimization",
    "Weaning from ventilator",
    "Organ recovery",
    "Optimal sedation",
    "Adequate nutrition"
  )
)
knitr::kable(icu_apps, caption = "Critical Care Line Chart Applications")
```

# Best Practices for Clinical Line Charts

## Data Preparation Guidelines

### 1. Time Variable Standardization
```{r time-standardization, eval=FALSE}
# Standardize time points for multi-site studies
clinical_data <- clinical_data %>%
  mutate(
    # Convert visit dates to study days
    study_day = as.numeric(visit_date - enrollment_date),
    
    # Group irregular visits into windows
    visit_window = case_when(
      study_day <= 7 ~ "Week 1",
      study_day <= 28 ~ "Month 1", 
      study_day <= 84 ~ "Month 3",
      study_day <= 168 ~ "Month 6",
      TRUE ~ "Later"
    ),
    
    # Handle missing values appropriately
    outcome_clean = ifelse(is.na(lab_value), NA, lab_value)
  )
```

### 2. Missing Data Strategies
```{r missing-data, eval=FALSE}
# Handle missing data patterns
missing_analysis <- clinical_data %>%
  group_by(patient_id) %>%
  summarise(
    n_visits = n(),
    n_missing = sum(is.na(primary_outcome)),
    last_visit = max(study_day, na.rm = TRUE),
    dropout_pattern = case_when(
      n_missing == 0 ~ "Complete",
      n_missing < n_visits/2 ~ "Sporadic missing",
      last_visit < 168 ~ "Early dropout",
      TRUE ~ "Late dropout"
    )
  )

# Visualize missing patterns
missing_plot <- linechart(
  data = clinical_data,
  xvar = "study_day",
  yvar = "primary_outcome",
  groupby = "dropout_pattern",
  title = "Outcome by Missing Data Pattern"
)
```

### 3. Outlier Management
```{r outlier-management, eval=FALSE}
# Identify and handle outliers
clinical_data <- clinical_data %>%
  group_by(visit_week) %>%
  mutate(
    # Calculate z-scores within time points
    z_score = abs(scale(lab_value)[,1]),
    
    # Flag extreme outliers
    outlier_flag = z_score > 3,
    
    # Winsorize extreme values
    lab_value_winsorized = case_when(
      z_score > 3 ~ quantile(lab_value, 0.95, na.rm = TRUE),
      z_score < -3 ~ quantile(lab_value, 0.05, na.rm = TRUE),
      TRUE ~ lab_value
    )
  ) %>%
  ungroup()
```

## Visualization Design Principles

### 1. Color and Theme Selection
```{r design-principles, eval=FALSE}
# Clinical publication-ready themes
publication_chart <- linechart(
  data = clinical_data,
  xvar = "time_point",
  yvar = "biomarker",
  groupby = "treatment",
  colorPalette = "colorblind",  # Accessible to colorblind readers
  theme = "publication",        # Clean, professional appearance
  xlabel = "Time (weeks)",
  ylabel = "Biomarker Level (units)",
  title = "Treatment Response Over Time"
)

# Specialty-specific color schemes
cardiology_chart <- linechart(
  data = bp_data,
  colorPalette = "clinical",    # Red/blue for medical contexts
  refline = 140,
  reflineLabel = "Hypertension Threshold"
)
```

### 2. Axis and Label Optimization
```{r axis-optimization, eval=FALSE}
# Optimize for clinical communication
optimized_chart <- linechart(
  data = clinical_data,
  xvar = "visit_month",
  yvar = "hemoglobin_g_dl",
  
  # Clear, clinical labels
  xlabel = "Months Since Treatment Initiation",
  ylabel = "Hemoglobin (g/dL)",
  title = "Hemoglobin Response to Iron Supplementation",
  
  # Reference lines for clinical context
  refline = 12,
  reflineLabel = "Normal Lower Limit (Women)",
  
  # Appropriate size for presentations
  width = 800,
  height = 600
)
```

## Statistical Reporting Standards

### 1. Correlation and Trend Reporting
```{r statistical-reporting, eval=FALSE}
# Extract statistical results for reporting
trend_analysis <- linechart(
  data = longitudinal_data,
  xvar = "time",
  yvar = "outcome",
  trendline = TRUE
)

# Report format for publications:
# "Hemoglobin levels showed a significant positive correlation with time 
# (r = 0.78, 95% CI: 0.65-0.87, p < 0.001), with an average increase 
# of 0.45 g/dL per month (95% CI: 0.32-0.58)."
```

### 2. Group Comparison Reporting  
```{r group-reporting, eval=FALSE}
# Compare trends between groups
group_analysis <- linechart(
  data = treatment_data,
  xvar = "weeks",
  yvar = "pain_score", 
  groupby = "treatment_arm",
  trendline = TRUE
)

# Report format:
# "Pain scores decreased significantly in both treatment groups over 12 weeks. 
# The experimental group showed a steeper decline (slope = -0.8 points/week, 
# r = -0.85, p < 0.001) compared to control (slope = -0.3 points/week, 
# r = -0.62, p = 0.003)."
```

# Troubleshooting Common Issues

## 1. Irregular Time Points

**Problem**: Clinical visits don't occur at regular intervals

**Solutions**:
```{r irregular-timepoints, eval=FALSE}
# Option 1: Use actual visit times
irregular_chart <- linechart(
  data = visit_data,
  xvar = "actual_visit_day",  # Use actual times
  yvar = "lab_result",
  smooth = TRUE              # Smooth over irregular intervals
)

# Option 2: Standardize to visit windows
standardized_data <- visit_data %>%
  mutate(
    visit_period = case_when(
      actual_visit_day <= 10 ~ "Baseline",
      actual_visit_day <= 35 ~ "Month 1",
      actual_visit_day <= 95 ~ "Month 3",
      TRUE ~ "Month 6"
    )
  )

window_chart <- linechart(
  data = standardized_data,
  xvar = "visit_period",
  yvar = "lab_result"
)
```

## 2. Highly Variable Data

**Problem**: Large fluctuations obscure underlying trends

**Solutions**:
```{r variable-data, eval=FALSE}
# Use smoothed lines for noisy data
smooth_chart <- linechart(
  data = noisy_data,
  xvar = "time",
  yvar = "variable_outcome",
  smooth = TRUE,             # LOESS smoothing
  confidence = TRUE,         # Show uncertainty
  points = FALSE             # Hide individual points
)

# Alternative: Moving averages
smoothed_data <- noisy_data %>%
  arrange(patient_id, time_point) %>%
  group_by(patient_id) %>%
  mutate(
    moving_avg = zoo::rollmean(outcome, k = 3, fill = NA, align = "center")
  ) %>%
  ungroup()
```

## 3. Multiple Y-Variables

**Problem**: Want to show multiple related outcomes

**Solutions**:
```{r multiple-outcomes, eval=FALSE}
# Option 1: Standardize to z-scores
standardized_data <- clinical_data %>%
  mutate(
    sbp_z = as.numeric(scale(systolic_bp)),
    dbp_z = as.numeric(scale(diastolic_bp))
  ) %>%
  tidyr::pivot_longer(
    cols = c(sbp_z, dbp_z),
    names_to = "bp_type",
    values_to = "z_score"
  )

multi_outcome_chart <- linechart(
  data = standardized_data,
  xvar = "visit_month",
  yvar = "z_score",
  groupby = "bp_type",
  ylabel = "Standardized Blood Pressure (Z-score)"
)

# Option 2: Separate charts with same scale
sbp_chart <- linechart(data = bp_data, xvar = "month", yvar = "systolic_bp")
dbp_chart <- linechart(data = bp_data, xvar = "month", yvar = "diastolic_bp")
```

## 4. Small Sample Sizes

**Problem**: Few patients or time points

**Solutions**:
```{r small-samples, eval=FALSE}
# Emphasize individual trajectories
small_sample_chart <- linechart(
  data = pilot_data,
  xvar = "time",
  yvar = "outcome", 
  points = TRUE,           # Show all data points
  confidence = FALSE,      # Don't show CI with small n
  smooth = FALSE,          # Use straight lines
  title = "Individual Patient Trajectories (n=8)"
)

# Consider spaghetti plots for individual patients
individual_chart <- linechart(
  data = pilot_data,
  xvar = "time",
  yvar = "outcome",
  groupby = "patient_id",  # Each patient as separate line
  theme = "minimal"
)
```

# Publication and Presentation Guidelines

## Journal Submission Standards

### Figure Preparation Checklist
```{r publication-checklist, eval=FALSE}
# Publication-ready line chart
publication_figure <- linechart(
  data = study_data,
  xvar = "time_weeks",
  yvar = "primary_endpoint",
  groupby = "treatment_arm",
  
  # Professional appearance
  theme = "publication",
  colorPalette = "colorblind",
  
  # Appropriate sizing
  width = 800,   # Suitable for 2-column format
  height = 600,
  
  # Clear labeling
  xlabel = "Time (weeks)",
  ylabel = "Primary Endpoint (units)",
  title = "",    # Usually caption provides title
  
  # Statistical elements
  confidence = TRUE,
  trendline = TRUE
)
```

### Essential Figure Elements
1. **Clear axes labels** with units
2. **Legible font sizes** (minimum 10pt)
3. **Colorblind-friendly palettes**
4. **Statistical information** (p-values, confidence intervals)
5. **Sample sizes** for each group and time point
6. **Reference lines** for clinical thresholds

### Caption Writing Guidelines
```
Figure 1. Treatment response over time by intervention group. 
Line chart showing mean (±95% CI) primary endpoint values over 
24 weeks of follow-up. Control group (n=45, blue), Intervention 
group (n=43, red). Dotted horizontal line indicates clinical 
significance threshold (threshold value). P-values from linear 
mixed-effects models comparing slopes between groups.
```

## Presentation Best Practices

### 1. Audience-Specific Design
```{r presentation-design, eval=FALSE}
# For clinical audiences
clinical_presentation <- linechart(
  data = clinical_data,
  colorPalette = "clinical",
  theme = "classic",
  
  # Larger fonts for projection
  width = 1000,
  height = 700,
  
  # Clear reference lines
  refline = clinical_threshold,
  reflineLabel = "Clinical Target"
)

# For research conferences
research_presentation <- linechart(
  data = research_data,
  theme = "publication",
  confidence = TRUE,        # Show statistical rigor
  trendline = TRUE,
  colorPalette = "viridis"  # Professional color scheme
)
```

### 2. Interactive Elements
For presentations where interaction is possible:

- **Hover information**: Display exact values and sample sizes
- **Group toggle**: Allow showing/hiding specific treatment groups
- **Time range selection**: Focus on specific periods of interest
- **Statistical overlay**: Toggle trend lines and confidence intervals

# Conclusion

Line charts are powerful tools for clinical research visualization that enable:

## Key Clinical Applications
- **Longitudinal Analysis**: Track patient outcomes over time
- **Treatment Monitoring**: Assess intervention effectiveness  
- **Biomarker Evolution**: Follow disease progression markers
- **Quality Improvement**: Monitor healthcare performance metrics
- **Comparative Research**: Compare treatments or populations

## Statistical Capabilities
- **Trend Analysis**: Linear and non-linear pattern detection
- **Correlation Assessment**: Quantify time-outcome relationships
- **Confidence Intervals**: Display uncertainty and precision
- **Group Comparisons**: Statistical comparison of trends

## Best Practices Summary
1. **Design for audience**: Clinical vs. research presentations
2. **Include clinical context**: Reference lines and thresholds  
3. **Handle missing data**: Transparent reporting of dropout patterns
4. **Statistical rigor**: Report correlations, confidence intervals, and p-values
5. **Professional appearance**: Publication-ready formatting

## Future Directions
- **Real-time monitoring**: Integration with electronic health records
- **Predictive modeling**: Combine trends with machine learning
- **Patient-specific trajectories**: Personalized medicine applications
- **Multi-outcome visualization**: Simultaneous monitoring of related endpoints

The `linechart` function in ClinicoPath provides researchers and clinicians with a comprehensive tool for creating publication-quality visualizations that effectively communicate temporal patterns and treatment effects in clinical data.

---

*For more information about ClinicoPath visualization tools and clinical data analysis, visit the package documentation and tutorials.*