---
title: "Social Science Statistical Visualization with jsjplot"
subtitle: "Comprehensive Guide to sjPlot Integration for Research"
date: last-modified
author: "ClinicoPath"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    number-sections: true
    number-depth: 2
    code-fold: show
    code-tools: true
    code-copy: true
    fig-width: 12
    fig-height: 8
    fig-dpi: 300
    df-print: paged
execute:
  warning: false
  message: false
  echo: true
  eval: true
  error: false
  freeze: auto
---

# Introduction

The `jsjplot` function provides comprehensive social science statistical visualization using the powerful `sjPlot` package. This function is designed specifically for researchers in psychology, sociology, political science, economics, and related fields who need publication-ready statistical visualizations and model summaries.

## Key Features

- **Multiple Analysis Types**: Regression tables, coefficient plots, interaction visualizations, marginal effects, frequency tables, correlation matrices, and PCA plots
- **Model Support**: Linear models (lm), generalized linear models (glm), logistic regression, Poisson regression, and mixed effects models
- **Publication Ready**: High-quality plots with extensive customization options
- **Statistical Tables**: HTML-formatted regression tables with confidence intervals and significance tests
- **Performance Optimized**: Intelligent caching system for faster repeated analyses
- **Comprehensive Output**: Model statistics, summaries, and interpretative guidance

## Installation and Setup

```{r setup}
# Load required libraries
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Set options for better output
options(digits = 3)
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
  dpi = 300,
  out.width = "100%"
)

# Check if sjPlot is available
if (!requireNamespace("sjPlot", quietly = TRUE)) {
  message("Note: sjPlot package not available. Install with: install.packages('sjPlot')")
}
```

# Data Preparation

Let's create realistic datasets for different types of social science analyses:

## Basic Research Data

```{r data-prep}
# Create comprehensive research dataset
set.seed(123)
research_data <- data.frame(
  # Continuous outcomes
  life_satisfaction = rnorm(300, 65, 15),
  academic_performance = rnorm(300, 75, 12),
  income = exp(rnorm(300, 10.5, 0.8)), # Log-normal distribution
  
  # Binary outcomes
  employment_status = rbinom(300, 1, 0.7), # 1 = employed
  college_graduate = rbinom(300, 1, 0.4),
  
  # Count outcomes
  social_activities = rpois(300, 3),
  volunteer_hours = rpois(300, 2),
  
  # Continuous predictors
  age = round(rnorm(300, 35, 12)),
  years_education = round(rnorm(300, 14, 3)),
  
  # Categorical predictors
  gender = factor(sample(c("Male", "Female"), 300, replace = TRUE)),
  marital_status = factor(sample(c("Single", "Married", "Divorced"), 
                                300, replace = TRUE, prob = c(0.4, 0.5, 0.1))),
  income_bracket = factor(sample(c("Low", "Medium", "High"), 
                                300, replace = TRUE, prob = c(0.3, 0.4, 0.3))),
  
  # Interaction variables
  treatment = factor(sample(c("Control", "Intervention"), 300, replace = TRUE)),
  baseline_score = rnorm(300, 50, 10)
) %>%
  mutate(
    # Create realistic relationships
    life_satisfaction = life_satisfaction + 
      0.2 * age + 
      2 * years_education + 
      ifelse(marital_status == "Married", 8, 0) + 
      ifelse(gender == "Female", 3, 0) + 
      rnorm(300, 0, 8),
    
    academic_performance = academic_performance + 
      1.5 * years_education + 
      ifelse(income_bracket == "High", 5, 0) + 
      rnorm(300, 0, 6),
    
    employment_status = rbinom(300, 1, plogis(-2 + 
      0.1 * years_education + 
      0.02 * age + 
      ifelse(gender == "Male", 0.3, 0)))
  )

# Display data structure
str(research_data)
```

# Basic Usage

## Coefficient Plots

The most common visualization in social science research - coefficient plots show effect sizes with confidence intervals:

```{r coefficient-plot-basic}
# Basic coefficient plot
result_coef <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender", "marital_status"),
  model_type = "lm"
)

# The result contains multiple components
print(names(result_coef))
```

## Regression Tables

Publication-ready regression tables with multiple customization options:

```{r regression-table}
result_table <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender", "marital_status"),
  model_type = "lm",
  show_p_values = TRUE,
  standardized = FALSE,
  confidence_level = 0.95
)
```

# Analysis Types

## 1. Linear Regression Analysis

### Continuous Outcomes

```{r linear-regression}
# Comprehensive linear model
linear_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket", "marital_status"),
  model_type = "lm",
  show_values = TRUE,
  sort_estimates = TRUE,
  remove_intercept = TRUE,
  theme_style = "minimal"
)
```

### Model Diagnostics and Statistics

```{r linear-diagnostics}
# Same model with full statistics display
linear_stats <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket", "marital_status"),
  model_type = "lm",
  show_statistics = TRUE,
  show_summary = TRUE,
  html_output = TRUE
)
```

## 2. Logistic Regression Analysis

### Binary Outcomes

```{r logistic-regression}
# Logistic regression for binary outcomes
logistic_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "employment_status",
  independent_vars = c("age", "years_education", "gender", "marital_status"),
  model_type = "logistic",
  show_values = TRUE,
  colors = "colorblind",
  title = "Predictors of Employment Status"
)
```

### Odds Ratios Visualization

```{r odds-ratios}
# Logistic regression with transformed coefficients
odds_ratio_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "college_graduate",
  independent_vars = c("age", "years_education", "income_bracket"),
  model_type = "logistic",
  transform_axis = "log",
  show_values = TRUE,
  title = "Odds Ratios for College Graduation"
)
```

## 3. Poisson Regression Analysis

### Count Outcomes

```{r poisson-regression}
# Poisson regression for count data
poisson_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "social_activities",
  independent_vars = c("age", "life_satisfaction", "marital_status"),
  model_type = "poisson",
  show_values = TRUE,
  theme_style = "apa"
)
```

## 4. Generalized Linear Models

### Different GLM Families

```{r glm-families}
# Gaussian GLM (equivalent to linear regression)
gaussian_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education"),
  model_type = "glm",
  family = "gaussian"
)

# Binomial GLM for binary outcomes
binomial_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "employment_status",
  independent_vars = c("age", "years_education"),
  model_type = "glm",
  family = "binomial"
)

# Poisson GLM for count outcomes
poisson_glm_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "volunteer_hours",
  independent_vars = c("age", "life_satisfaction"),
  model_type = "glm",
  family = "poisson"
)
```

# Advanced Features

## Interaction Effects

### Continuous × Categorical Interactions

```{r interactions-cont-cat}
# Create interaction data
interaction_data <- research_data %>%
  mutate(
    # Interaction between age and treatment
    outcome_with_interaction = baseline_score + 
      0.5 * age + 
      ifelse(treatment == "Intervention", 10, 0) + 
      # Interaction: age effect is stronger in intervention group
      ifelse(treatment == "Intervention", 0.3 * age, 0) + 
      rnorm(300, 0, 8)
  )

# Interaction plot
interaction_result <- jsjplot(
  data = interaction_data,
  analysis_type = "interaction_plot",
  dependent_var = "outcome_with_interaction",
  interaction_vars = c("age", "treatment"),
  independent_vars = c("baseline_score"),
  title = "Age × Treatment Interaction Effect"
)
```

### Categorical × Categorical Interactions

```{r interactions-cat-cat}
# Interaction between gender and marital status
categorical_interaction <- jsjplot(
  data = research_data,
  analysis_type = "interaction_plot",
  dependent_var = "life_satisfaction",
  interaction_vars = c("gender", "marital_status"),
  independent_vars = c("age"),
  colors = "viridis"
)
```

## Marginal Effects

### Predicted Values and Effects

```{r marginal-effects}
# Marginal effects visualization
marginal_result <- jsjplot(
  data = research_data,
  analysis_type = "marginal_effects",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "marital_status"),
  model_type = "lm",
  plot_type = "eff"
)
```

### Predicted Values Plot

```{r predicted-values}
# Predicted values plot
predicted_result <- jsjplot(
  data = research_data,
  analysis_type = "marginal_effects",
  dependent_var = "employment_status",
  independent_vars = c("age", "years_education"),
  model_type = "logistic",
  plot_type = "pred"
)
```

# Descriptive Analyses

## Frequency Tables

### Categorical Variable Analysis

```{r frequency-tables}
# Comprehensive frequency analysis
freq_result <- jsjplot(
  data = research_data,
  analysis_type = "frequency_table"
)
```

### Cross-Tabulations

```{r cross-tabs}
# Create dataset with more categorical variables for better demonstration
categorical_data <- research_data %>%
  mutate(
    age_group = cut(age, breaks = c(0, 30, 50, 100), labels = c("Young", "Middle", "Older")),
    satisfaction_level = cut(life_satisfaction, 
                           breaks = c(0, 50, 70, 100), 
                           labels = c("Low", "Medium", "High"))
  )

# Frequency analysis with more categories
freq_categorical <- jsjplot(
  data = categorical_data,
  analysis_type = "frequency_table"
)
```

## Correlation Matrices

### Correlation Visualization

```{r correlation-matrix}
# Select numeric variables for correlation
numeric_vars <- research_data %>%
  select(life_satisfaction, academic_performance, age, years_education, 
         social_activities, volunteer_hours, baseline_score)

# Correlation matrix analysis
corr_result <- jsjplot(
  data = numeric_vars,
  analysis_type = "correlation_matrix"
)
```

## Principal Component Analysis

### Dimensionality Reduction

```{r pca-analysis}
# PCA for dimensionality reduction
pca_result <- jsjplot(
  data = numeric_vars,
  analysis_type = "pca_plot"
)
```

### High-Dimensional Data

```{r pca-highdim}
# Create high-dimensional dataset
set.seed(456)
highdim_data <- data.frame(
  replicate(10, rnorm(200))
) %>%
  setNames(paste0("variable_", 1:10)) %>%
  mutate(
    # Create some factor structure
    factor1 = variable_1 + variable_2 + variable_3 + rnorm(200, 0, 0.5),
    factor2 = variable_4 + variable_5 + variable_6 + rnorm(200, 0, 0.5)
  )

# PCA on high-dimensional data
pca_highdim <- jsjplot(
  data = highdim_data,
  analysis_type = "pca_plot"
)
```

# Customization and Styling

## Visual Themes

### Different Theme Styles

```{r theme-comparison}
themes <- c("sjplot", "minimal", "classic", "apa", "bw")

# Demonstrate different themes (showing code for minimal theme)
theme_minimal <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender"),
  theme_style = "minimal",
  title = "Minimal Theme Example"
)

# APA style for publications
theme_apa <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender"),
  theme_style = "apa",
  title = "APA Style for Publications"
)
```

## Color Schemes

### Color-blind Friendly Options

```{r color-schemes}
# Color-blind safe palette
colorblind_plot <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket"),
  colors = "colorblind",
  title = "Color-blind Friendly Plot"
)

# Viridis color scheme
viridis_plot <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket"),
  colors = "viridis",
  title = "Viridis Color Scheme"
)
```

## Plot Customization

### Size and Spacing Options

```{r plot-customization}
# Customized plot appearance
custom_plot <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender", "marital_status"),
  dot_size = 4,
  line_size = 1.2,
  grid_breaks = 0.5,
  show_values = TRUE,
  title = "Customized Coefficient Plot",
  axis_labels = "Age,Education,Gender,Marital Status"
)
```

### Confidence Intervals

```{r confidence-intervals}
# Different confidence levels
ci_90 <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education"),
  confidence_level = 0.90,
  title = "90% Confidence Intervals"
)

ci_99 <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education"),
  confidence_level = 0.99,
  title = "99% Confidence Intervals"
)
```

# Advanced Statistical Options

## Standardized Coefficients

### Beta Coefficients

```{r standardized-coefficients}
# Standardized coefficients for comparison
standardized_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "baseline_score"),
  standardized = TRUE,
  title = "Standardized Coefficients (Beta)",
  show_values = TRUE
)

# Compare with unstandardized
unstandardized_result <- jsjplot(
  data = research_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "baseline_score"),
  standardized = FALSE,
  title = "Unstandardized Coefficients",
  show_values = TRUE
)
```

## Model Comparison

### Multiple Model Types

```{r model-comparison}
# Linear model
linear_comparison <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket"),
  model_type = "lm",
  title = "Linear Model Results"
)

# GLM with Gaussian family (equivalent to linear)
glm_comparison <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "income_bracket"),
  model_type = "glm",
  family = "gaussian",
  title = "GLM (Gaussian) Results"
)
```

# Real-World Applications

## Psychology Research Example

```{r psychology-example}
# Create psychology research dataset
psychology_data <- data.frame(
  participant_id = 1:250,
  # Psychological measures
  depression_score = rnorm(250, 15, 8),
  anxiety_score = rnorm(250, 12, 6),
  stress_score = rnorm(250, 18, 10),
  # Demographics
  age = round(rnorm(250, 30, 10)),
  gender = factor(sample(c("Male", "Female"), 250, replace = TRUE)),
  therapy_type = factor(sample(c("CBT", "Psychodynamic", "Control"), 250, replace = TRUE)),
  # Outcome
  wellbeing_improvement = rnorm(250, 10, 15)
) %>%
  mutate(
    # Create realistic relationships
    wellbeing_improvement = wellbeing_improvement - 
      0.3 * depression_score - 
      0.2 * anxiety_score - 
      0.1 * stress_score + 
      ifelse(therapy_type == "CBT", 8, 0) + 
      ifelse(therapy_type == "Psychodynamic", 5, 0) + 
      rnorm(250, 0, 8)
  )

# Psychology research analysis
psych_result <- jsjplot(
  data = psychology_data,
  analysis_type = "coefficient_plot",
  dependent_var = "wellbeing_improvement",
  independent_vars = c("depression_score", "anxiety_score", "therapy_type", "age"),
  model_type = "lm",
  theme_style = "apa",
  show_values = TRUE,
  title = "Predictors of Wellbeing Improvement"
)
```

## Economics Research Example

```{r economics-example}
# Create economics dataset
economics_data <- data.frame(
  # Economic indicators
  gdp_growth = rnorm(200, 2.5, 1.5),
  unemployment_rate = rnorm(200, 6, 2),
  inflation_rate = rnorm(200, 2, 1),
  # Policy variables
  fiscal_spending = rnorm(200, 20, 5),
  tax_rate = rnorm(200, 25, 8),
  # Country characteristics
  country_size = factor(sample(c("Small", "Medium", "Large"), 200, replace = TRUE)),
  region = factor(sample(c("Europe", "Asia", "Americas", "Africa"), 200, replace = TRUE)),
  # Outcome
  economic_satisfaction = rnorm(200, 60, 20)
) %>%
  mutate(
    economic_satisfaction = economic_satisfaction + 
      2 * gdp_growth - 
      1.5 * unemployment_rate + 
      0.5 * fiscal_spending + 
      ifelse(country_size == "Large", 10, 0) + 
      rnorm(200, 0, 12)
  )

# Economics analysis
econ_result <- jsjplot(
  data = economics_data,
  analysis_type = "coefficient_plot",
  dependent_var = "economic_satisfaction",
  independent_vars = c("gdp_growth", "unemployment_rate", "fiscal_spending", "country_size"),
  model_type = "lm",
  colors = "metro",
  title = "Economic Determinants of Satisfaction"
)
```

## Clinical Research Example

```{r clinical-example}
# Create clinical trial dataset
clinical_data <- data.frame(
  patient_id = 1:300,
  # Baseline characteristics
  age = round(rnorm(300, 55, 15)),
  gender = factor(sample(c("Male", "Female"), 300, replace = TRUE)),
  baseline_severity = rnorm(300, 50, 15),
  comorbidities = rpois(300, 1.5),
  # Treatment
  treatment_arm = factor(sample(c("Placebo", "Low_Dose", "High_Dose"), 
                               300, replace = TRUE)),
  # Outcomes
  treatment_response = rbinom(300, 1, 0.4),
  side_effects = rpois(300, 0.8),
  quality_of_life = rnorm(300, 70, 20)
) %>%
  mutate(
    treatment_response = rbinom(300, 1, plogis(-1.5 + 
      ifelse(treatment_arm == "Low_Dose", 0.8, 0) + 
      ifelse(treatment_arm == "High_Dose", 1.5, 0) - 
      0.01 * baseline_severity)),
    
    quality_of_life = quality_of_life + 
      ifelse(treatment_response == 1, 15, 0) - 
      0.2 * age - 
      3 * comorbidities + 
      rnorm(300, 0, 12)
  )

# Clinical analysis - treatment response
clinical_response <- jsjplot(
  data = clinical_data,
  analysis_type = "coefficient_plot",
  dependent_var = "treatment_response",
  independent_vars = c("treatment_arm", "age", "baseline_severity", "gender"),
  model_type = "logistic",
  theme_style = "minimal",
  title = "Predictors of Treatment Response"
)

# Quality of life analysis
clinical_qol <- jsjplot(
  data = clinical_data,
  analysis_type = "coefficient_plot",
  dependent_var = "quality_of_life",
  independent_vars = c("treatment_response", "age", "comorbidities"),
  model_type = "lm",
  title = "Factors Affecting Quality of Life"
)
```

# Error Handling and Troubleshooting

## Missing Data

```{r missing-data-handling}
# Create data with missing values
missing_data <- research_data
missing_data$life_satisfaction[sample(1:300, 30)] <- NA
missing_data$age[sample(1:300, 20)] <- NA

# Analysis with missing data
missing_result <- jsjplot(
  data = missing_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender"),
  model_type = "lm"
)
```

## Small Sample Sizes

```{r small-samples}
# Small dataset
small_data <- research_data[1:30, ]

# Analysis with small sample
small_result <- jsjplot(
  data = small_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education"),
  model_type = "lm"
)
```

## Edge Cases

```{r edge-cases}
# Constant variable (no variation)
edge_data <- research_data
edge_data$constant_var <- 5

# Perfect correlation
edge_data$age_copy <- edge_data$age

# Analysis with edge cases
edge_result <- jsjplot(
  data = edge_data,
  analysis_type = "coefficient_plot",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "constant_var"),
  model_type = "lm"
)
```

# Performance and Optimization

## Caching Benefits

```{r performance-demo}
# Large dataset for performance testing
large_data <- do.call(rbind, replicate(3, research_data, simplify = FALSE))
large_data$id <- 1:nrow(large_data)

# First run (will be slower)
system.time({
  perf_result1 <- jsjplot(
    data = large_data,
    analysis_type = "coefficient_plot",
    dependent_var = "life_satisfaction",
    independent_vars = c("age", "years_education", "gender", "marital_status"),
    model_type = "lm"
  )
})

# Second run with same data/options (should be faster due to caching)
system.time({
  perf_result2 <- jsjplot(
    data = large_data,
    analysis_type = "coefficient_plot",
    dependent_var = "life_satisfaction",
    independent_vars = c("age", "years_education", "gender", "marital_status"),
    model_type = "lm"
  )
})
```

# Best Practices

## 1. Model Selection Guidelines

```{r model-selection-guide}
# Create guidance table
model_guide <- data.frame(
  Outcome_Type = c(
    "Continuous (normal)",
    "Binary (0/1)",
    "Count (0,1,2,...)",
    "Ordinal",
    "Time-to-event"
  ),
  Recommended_Model = c(
    "lm (Linear Model)",
    "logistic (Logistic Regression)",
    "poisson (Poisson Regression)",
    "logistic with ordinal package",
    "Survival models (not implemented)"
  ),
  jsjplot_Setting = c(
    "model_type = 'lm'",
    "model_type = 'logistic'",
    "model_type = 'poisson'",
    "model_type = 'logistic'",
    "Not available"
  ),
  Example_Use = c(
    "Life satisfaction, test scores",
    "Employment status, treatment response",
    "Number of events, visits",
    "Likert scales, rankings",
    "Time to treatment response"
  )
)

print(model_guide)
```

## 2. Visualization Guidelines

```{r visualization-guide}
# Best practices for different analysis types
viz_guide <- data.frame(
  Analysis_Type = c(
    "coefficient_plot",
    "regression_table",
    "interaction_plot",
    "marginal_effects",
    "frequency_table",
    "correlation_matrix",
    "pca_plot"
  ),
  Best_For = c(
    "Comparing effect sizes across predictors",
    "Publication-ready regression results",
    "Visualizing interaction effects",
    "Showing predicted values and effects",
    "Descriptive analysis of categories",
    "Exploring variable relationships",
    "Dimensionality reduction visualization"
  ),
  Recommended_Options = c(
    "sort_estimates=TRUE, remove_intercept=TRUE",
    "show_p_values=TRUE, confidence_level=0.95",
    "Clear labeling of interaction variables",
    "plot_type='eff' for effects",
    "Limit to important categorical variables",
    "Use with numeric variables only",
    "Good for 3+ numeric variables"
  )
)

print(viz_guide)
```

## 3. Statistical Interpretation

```{r interpretation-guide}
# Effect size interpretation guidelines
effect_guide <- data.frame(
  Effect_Size = c("Small", "Medium", "Large"),
  Cohen_d = c("0.2", "0.5", "0.8"),
  R_squared = c("0.01", "0.09", "0.25"),
  Interpretation = c(
    "Subtle but potentially important",
    "Moderate practical significance",
    "Large practical significance"
  )
)

print(effect_guide)
```

# Reporting Results

## APA Style Reporting

```{r apa-reporting}
# Example of APA-style analysis
apa_result <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "marital_status"),
  model_type = "lm",
  theme_style = "apa",
  confidence_level = 0.95,
  show_p_values = TRUE,
  standardized = TRUE
)
```

## Publication-Ready Tables

```{r publication-tables}
# High-quality table for publication
pub_table <- jsjplot(
  data = research_data,
  analysis_type = "regression_table",
  dependent_var = "life_satisfaction",
  independent_vars = c("age", "years_education", "gender", "marital_status", "income_bracket"),
  model_type = "lm",
  html_output = TRUE,
  show_p_values = TRUE,
  standardized = FALSE,
  confidence_level = 0.95
)
```

# Summary

The `jsjplot` function provides a comprehensive solution for social science statistical visualization and analysis. Key advantages include:

## Core Capabilities

- **Multiple Analysis Types**: Seven different analysis types covering most social science research needs
- **Model Flexibility**: Support for linear, logistic, Poisson, and generalized linear models
- **Publication Quality**: High-quality plots and tables ready for publication
- **Statistical Rigor**: Confidence intervals, significance tests, and model diagnostics

## Performance Features

- **Intelligent Caching**: Automatic caching of models and results for faster repeated analyses
- **Error Handling**: Robust error handling for missing data and edge cases
- **Memory Efficiency**: Optimized for large datasets with smart data preparation
- **Validation**: Comprehensive input validation and informative error messages

## Customization Options

- **Visual Themes**: Multiple themes including APA style for publications
- **Color Schemes**: Color-blind friendly and aesthetically pleasing palettes
- **Plot Customization**: Extensive control over plot elements and styling
- **Output Formats**: HTML tables and high-resolution plots

## Use Cases

- **Academic Research**: Psychology, sociology, political science, economics
- **Clinical Research**: Treatment effects, patient outcomes, clinical trials
- **Business Analytics**: Customer satisfaction, employee surveys, market research
- **Policy Analysis**: Public policy evaluation, social program assessment

## Next Steps

- Explore the [ClinicoPath package documentation](https://github.com/sbalci/ClinicoPathJamoviModule) for more analysis functions
- Try different analysis types with your own data
- Experiment with customization options for publication-ready outputs
- Consider the sjPlot package documentation for advanced features

```{r session-info}
sessionInfo()
```