---
title: "Predictive Power Score (PPS) Analysis with jpps"
subtitle: "Comprehensive Guide to Modern Relationship Detection"
date: "2025-07-07"
author: "ClinicoPath"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    highlight-style: github
    theme: bootstrap
    embed-resources: true
    standalone: true
    self-contained: true
    code-fold: false
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
  eval: true
editor: visual
---

# Introduction

## What is Predictive Power Score (PPS)?

Predictive Power Score (PPS) is a modern metric that detects **linear and non-linear relationships** between variables using machine learning. Unlike traditional correlation analysis, PPS:

- **Detects non-linear relationships** that correlation misses
- **Works with mixed data types** (numeric, categorical)
- Provides **asymmetric scores** (X→Y may differ from Y→X)
- **Uses machine learning** for robust relationship detection
- Ranges from **0 (no predictive power) to 1 (perfect prediction)**

## Why Use PPS Instead of Correlation?

| Feature | Correlation | PPS |
|---------|-------------|-----|
| **Relationship Types** | Linear only | Linear + Non-linear |
| **Data Types** | Numeric only | Numeric + Categorical |
| **Symmetry** | Symmetric (X↔Y) | Asymmetric (X→Y ≠ Y→X) |
| **Method** | Mathematical formula | Machine learning |
| **Range** | -1 to +1 | 0 to 1 |
| **Outlier Sensitivity** | High | Moderate |

## The jpps Function

The `jpps()` function in ClinicoPath provides a comprehensive interface for PPS analysis with four analysis types:

1. **Single**: One predictor → one target
2. **Predictors**: Multiple predictors → one target  
3. **Matrix**: All variables → all variables
4. **Compare**: PPS vs correlation comparison

# Getting Started

```{r setup}
# Load required packages
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Set random seed for reproducibility
set.seed(20250707)
```

## Quick Example

Let's start with a simple example showing PPS detecting a non-linear relationship:

```{r quick_example}
# Create data with non-linear relationship
data <- data.frame(
  x = seq(-3, 3, length.out = 50),
  y_linear = 2 * seq(-3, 3, length.out = 50) + rnorm(50, 0, 0.5),
  y_quadratic = seq(-3, 3, length.out = 50)^2 + rnorm(50, 0, 0.5)
)

# Analyze with PPS
result <- jpps(
  data = data,
  analysis_type = "predictors",
  target_var = "y_quadratic",
  predictor_vars = c("x"),
  algorithm = "tree"
)

# View results
result
```

This demonstrates how PPS can detect the quadratic relationship between `x` and `y_quadratic`.

# Analysis Types

## Single Predictor Analysis

Use when you want to analyze **one specific predictor-target relationship**:

```{r single_analysis}
# Create test data
single_data <- data.frame(
  sales = c(100, 150, 120, 180, 140, 200, 160, 220, 180, 250),
  advertising = c(5, 8, 6, 10, 7, 12, 9, 13, 10, 15)
)

# Single predictor analysis
single_result <- jpps(
  data = single_data,
  analysis_type = "single",
  target_var = "sales",
  predictor_var = "advertising",
  algorithm = "auto"
)

single_result
```

## Multiple Predictors Analysis

Use to **identify the best predictors** for a target variable:

```{r predictors_analysis}
# Create business data
business_data <- data.frame(
  revenue = rnorm(40, 1000, 200),
  marketing_spend = rnorm(40, 50, 15),
  employee_count = round(rnorm(40, 25, 8)),
  customer_satisfaction = rnorm(40, 4.2, 0.8),
  market_share = rnorm(40, 15, 5)
)

# Multiple predictors analysis
predictors_result <- jpps(
  data = business_data,
  analysis_type = "predictors",
  target_var = "revenue",
  predictor_vars = c("marketing_spend", "employee_count", "customer_satisfaction", "market_share"),
  sort_results = "pps_desc",
  show_summary = TRUE
)

predictors_result
```

## Matrix Analysis

Use for **comprehensive relationship exploration** between all variables:

```{r matrix_analysis}
# Create correlation matrix data
matrix_data <- data.frame(
  var1 = rnorm(30, 50, 10),
  var2 = rnorm(30, 100, 20),
  var3 = rnorm(30, 25, 5),
  var4 = rnorm(30, 75, 15)
) %>%
  mutate(
    # Create some relationships
    var2 = var2 + 0.5 * var1,  # Linear relationship
    var3 = var1^0.5 + rnorm(30, 0, 2),  # Non-linear relationship
    var4 = ifelse(var1 > 50, var4 + 20, var4)  # Threshold relationship
  )

# Matrix analysis with heatmap
matrix_result <- jpps(
  data = matrix_data,
  analysis_type = "matrix",
  matrix_vars = c("var1", "var2", "var3", "var4"),
  show_heatmap = TRUE,
  color_scheme = "viridis",
  show_values_on_plot = TRUE
)

matrix_result
```

## Comparison Analysis

Use to **compare PPS with correlation** and identify where PPS provides additional insights:

```{r comparison_analysis}
# Create data where PPS outperforms correlation
comparison_data <- data.frame(
  x = runif(50, -3, 3)
) %>%
  mutate(
    linear_rel = 0.8 * x + rnorm(50, 0, 0.5),      # High correlation, high PPS
    quadratic_rel = x^2 + rnorm(50, 0, 0.5),       # Low correlation, high PPS
    no_rel = rnorm(50, 0, 1)                       # Low correlation, low PPS
  )

# Comparison analysis
comparison_result <- jpps(
  data = comparison_data,
  analysis_type = "compare",
  matrix_vars = c("x", "linear_rel", "quadratic_rel", "no_rel"),
  show_correlation_comparison = TRUE,
  correlation_method = "pearson"
)

comparison_result
```

# Algorithm Options

PPS supports three algorithms, each with different strengths:

## Decision Tree Algorithm

**Best for**: Interpretable results, categorical variables, threshold relationships

```{r tree_algorithm}
# Data with clear thresholds
threshold_data <- data.frame(
  income = runif(40, 20000, 100000),
  age = round(runif(40, 25, 65))
) %>%
  mutate(
    loan_approved = factor(ifelse(income > 50000 & age > 30, "Yes", "No"))
  )

tree_result <- jpps(
  data = threshold_data,
  analysis_type = "single",
  target_var = "loan_approved",
  predictor_var = "income",
  algorithm = "tree"
)

tree_result
```

## Random Forest Algorithm

**Best for**: Complex relationships, mixed data types, high accuracy

```{r forest_algorithm}
forest_result <- jpps(
  data = threshold_data,
  analysis_type = "single", 
  target_var = "loan_approved",
  predictor_var = "income",
  algorithm = "forest"
)

forest_result
```

## Auto Algorithm

**Best for**: General use, automatic algorithm selection

```{r auto_algorithm}
auto_result <- jpps(
  data = threshold_data,
  analysis_type = "single",
  target_var = "loan_approved", 
  predictor_var = "income",
  algorithm = "auto"
)

auto_result
```

# Working with Different Data Types

## Mixed Data Types

PPS excels with mixed categorical and numerical data:

```{r mixed_data}
# Create mixed data
mixed_data <- data.frame(
  # Categorical variables
  education = factor(sample(c("High School", "Bachelor", "Master", "PhD"), 60, replace = TRUE)),
  department = factor(sample(c("Sales", "Marketing", "Engineering", "HR"), 60, replace = TRUE)),
  
  # Numerical variables
  experience_years = round(runif(60, 0, 20)),
  performance_score = rnorm(60, 75, 15)
) %>%
  mutate(
    # Salary influenced by education and experience
    salary = case_when(
      education == "PhD" ~ 80000 + experience_years * 3000 + rnorm(sum(education == "PhD"), 0, 5000),
      education == "Master" ~ 60000 + experience_years * 2500 + rnorm(sum(education == "Master"), 0, 4000),
      education == "Bachelor" ~ 45000 + experience_years * 2000 + rnorm(sum(education == "Bachelor"), 0, 3000),
      TRUE ~ 30000 + experience_years * 1500 + rnorm(sum(education == "High School"), 0, 2000)
    )
  )

# Analyze mixed data
mixed_result <- jpps(
  data = mixed_data,
  analysis_type = "predictors",
  target_var = "salary",
  predictor_vars = c("education", "department", "experience_years", "performance_score"),
  sort_results = "pps_desc"
)

mixed_result
```

## Time Series Data

PPS can detect temporal patterns:

```{r timeseries_data}
# Create time series data
ts_data <- data.frame(
  month = 1:24,
  seasonal_factor = sin(2 * pi * (1:24) / 12)
) %>%
  mutate(
    sales = 1000 + 50 * month + 200 * seasonal_factor + rnorm(24, 0, 50),
    lagged_sales = lag(sales, 1)
  ) %>%
  filter(!is.na(lagged_sales))

# Analyze temporal relationships
ts_result <- jpps(
  data = ts_data,
  analysis_type = "matrix",
  matrix_vars = c("month", "seasonal_factor", "sales", "lagged_sales"),
  show_heatmap = TRUE
)

ts_result
```

# Advanced Features

## Sample Size Management

For large datasets, use sample size limits for faster analysis:

```{r sample_size}
# Create large dataset
large_data <- data.frame(
  x = rnorm(1000),
  y = rnorm(1000)
) %>%
  mutate(z = 0.6 * x + 0.4 * y + rnorm(1000, 0, 0.3))

# Use sampling for efficiency
sampled_result <- jpps(
  data = large_data,
  analysis_type = "matrix",
  matrix_vars = c("x", "y", "z"),
  sample_size = 200,  # Use only 200 samples
  algorithm = "tree"
)

sampled_result
```

## Cross-Validation Options

Adjust cross-validation for different dataset sizes:

```{r cross_validation}
# Small dataset - use fewer folds
small_result <- jpps(
  data = mixed_data[1:20, ],
  analysis_type = "single",
  target_var = "salary",
  predictor_var = "experience_years",
  cv_folds = 3  # Fewer folds for small data
)

# Larger dataset - use more folds
large_result <- jpps(
  data = mixed_data,
  analysis_type = "single",
  target_var = "salary",
  predictor_var = "experience_years", 
  cv_folds = 10  # More folds for better validation
)
```

## Threshold Filtering

Filter results by minimum PPS score:

```{r threshold_filtering}
# Show only relationships with PPS > 0.1
threshold_result <- jpps(
  data = matrix_data,
  analysis_type = "matrix",
  matrix_vars = c("var1", "var2", "var3", "var4"),
  min_pps_threshold = 0.1,
  sort_results = "pps_desc"
)

threshold_result
```

# Visualization Options

## Heatmap Customization

Customize heatmaps for matrix analysis:

```{r heatmap_custom}
# Custom color scheme
custom_heatmap <- jpps(
  data = matrix_data,
  analysis_type = "matrix",
  matrix_vars = c("var1", "var2", "var3", "var4"),
  show_heatmap = TRUE,
  color_scheme = "custom",
  custom_color_low = "#FFFFFF",
  custom_color_high = "#FF6B35",
  show_values_on_plot = TRUE,
  plot_title = "Custom PPS Heatmap"
)

custom_heatmap
```

## Barplot Options

Customize barplots for predictor analysis:

```{r barplot_custom}
# Custom barplot
barplot_result <- jpps(
  data = business_data,
  analysis_type = "predictors",
  target_var = "revenue",
  predictor_vars = c("marketing_spend", "employee_count", "customer_satisfaction"),
  show_barplot = TRUE,
  show_values_on_plot = TRUE,
  plot_title = "Revenue Predictors",
  sort_results = "pps_desc"
)

barplot_result
```

# Practical Applications

## Marketing Analytics

Identify which marketing channels drive sales:

```{r marketing_example}
# Marketing data
marketing_data <- data.frame(
  # Marketing channels
  tv_spend = runif(50, 1000, 10000),
  digital_spend = runif(50, 500, 5000),
  radio_spend = runif(50, 200, 2000),
  print_spend = runif(50, 100, 1000),
  
  # Customer metrics
  brand_awareness = runif(50, 20, 80),
  website_visits = round(runif(50, 1000, 10000))
) %>%
  mutate(
    # Sales influenced by digital and TV (non-linear)
    sales = 5000 + 
      sqrt(digital_spend) * 2 +  # Non-linear digital effect
      tv_spend * 0.3 +           # Linear TV effect
      ifelse(brand_awareness > 60, 2000, 0) +  # Threshold effect
      rnorm(50, 0, 1000)
  )

# Analyze marketing effectiveness
marketing_result <- jpps(
  data = marketing_data,
  analysis_type = "predictors",
  target_var = "sales",
  predictor_vars = c("tv_spend", "digital_spend", "radio_spend", "print_spend", "brand_awareness"),
  algorithm = "forest",
  show_summary = TRUE
)

marketing_result
```

## Medical Research

Identify predictors of health outcomes:

```{r medical_example}
# Clinical data
clinical_data <- data.frame(
  age = round(runif(80, 25, 75)),
  gender = factor(sample(c("Male", "Female"), 80, replace = TRUE)),
  bmi = rnorm(80, 26, 4),
  smoking = factor(sample(c("Never", "Former", "Current"), 80, replace = TRUE, prob = c(0.5, 0.3, 0.2))),
  exercise_hours = pmax(0, rnorm(80, 3, 2))
) %>%
  mutate(
    # Risk score with complex relationships
    risk_score = case_when(
      smoking == "Current" & age > 50 ~ 70 + rnorm(sum(smoking == "Current" & age > 50), 0, 10),
      smoking == "Current" ~ 50 + rnorm(sum(smoking == "Current" & age <= 50), 0, 8),
      age > 60 ~ 40 + rnorm(sum(smoking != "Current" & age > 60), 0, 12),
      TRUE ~ 20 + rnorm(sum(smoking != "Current" & age <= 60), 0, 8)
    ),
    risk_score = pmax(0, pmin(100, risk_score))
  )

# Analyze health risk factors
health_result <- jpps(
  data = clinical_data,
  analysis_type = "predictors",
  target_var = "risk_score",
  predictor_vars = c("age", "gender", "bmi", "smoking", "exercise_hours"),
  algorithm = "forest",
  show_summary = TRUE
)

health_result
```

## Financial Analysis

Detect relationships in financial data:

```{r financial_example}
# Financial indicators
financial_data <- data.frame(
  # Market indicators
  sp500_return = rnorm(60, 0.08, 0.15),
  volatility_index = abs(rnorm(60, 20, 8)),
  bond_yield = rnorm(60, 3.5, 1.2),
  
  # Economic indicators  
  gdp_growth = rnorm(60, 2.5, 1.0),
  inflation = rnorm(60, 2.8, 0.8),
  unemployment = rnorm(60, 5.2, 1.5)
) %>%
  mutate(
    # Stock performance with complex relationships
    stock_performance = 
      sp500_return * 1.2 +                           # Market beta
      -0.5 * log(volatility_index + 1) +            # Non-linear volatility effect
      ifelse(gdp_growth > 3, 0.05, 0) +             # Threshold GDP effect
      rnorm(60, 0, 0.1)
  )

# Analyze financial relationships
financial_result <- jpps(
  data = financial_data,
  analysis_type = "compare",
  matrix_vars = c("sp500_return", "volatility_index", "bond_yield", "gdp_growth", "stock_performance"),
  show_correlation_comparison = TRUE,
  correlation_method = "pearson"
)

financial_result
```

# Performance and Optimization

## Caching System

The jpps function includes intelligent caching to speed up repeated analyses:

```{r caching_demo, eval=FALSE}
# First run - computes PPS
system.time({
  result1 <- jpps(
    data = large_data,
    analysis_type = "matrix",
    matrix_vars = c("x", "y", "z")
  )
})

# Second run with same parameters - uses cache (much faster)
system.time({
  result2 <- jpps(
    data = large_data,
    analysis_type = "matrix", 
    matrix_vars = c("x", "y", "z")
  )
})
```

## Memory Management

For very large datasets, use strategic sampling:

```{r memory_management, eval=FALSE}
# For datasets > 10,000 rows
huge_data <- data.frame(
  x = rnorm(50000),
  y = rnorm(50000),
  z = rnorm(50000)
)

# Use sample_size to manage memory
efficient_result <- jpps(
  data = huge_data,
  analysis_type = "matrix",
  matrix_vars = c("x", "y", "z"),
  sample_size = 1000,    # Sample for efficiency
  algorithm = "tree"     # Faster algorithm
)
```

# Interpretation Guidelines

## PPS Score Interpretation

| PPS Score | Interpretation | Action |
|-----------|----------------|--------|
| **0.0** | No predictive power | No relationship |
| **0.0 - 0.2** | Weak relationship | Investigate further |
| **0.2 - 0.5** | Moderate relationship | Potentially useful |
| **0.5 - 0.8** | Strong relationship | Very useful predictor |
| **0.8 - 1.0** | Very strong to perfect | Excellent predictor |

## When to Use PPS vs Correlation

### Use PPS when:
- Exploring **unknown relationships**
- Working with **mixed data types**
- Suspecting **non-linear patterns**
- Need **asymmetric relationship detection**
- Want **robust outlier handling**

### Use Correlation when:
- Specifically testing **linear relationships**
- Need **fast computation**
- Working with **continuous variables only**
- Require **exact mathematical interpretation**

## Clinical/Research Interpretation

```{r interpretation_example}
# Example interpretation workflow
interpretation_data <- data.frame(
  treatment_dose = c(0, 5, 10, 15, 20, 25, 30),
  response_rate = c(10, 25, 45, 65, 75, 80, 82)
)

# Analyze dose-response relationship
dose_response <- jpps(
  data = interpretation_data,
  analysis_type = "single",
  target_var = "response_rate",
  predictor_var = "treatment_dose",
  algorithm = "tree"
)

dose_response
```

**Interpretation**: A high PPS score (>0.8) suggests treatment dose is a strong predictor of response rate, likely following a non-linear dose-response curve typical in pharmacology.

# Best Practices

## Data Preparation

1. **Clean missing values** appropriately
2. **Scale/normalize** when comparing vastly different ranges
3. **Consider transformations** for heavily skewed data
4. **Encode categorical variables** consistently

## Analysis Strategy

1. **Start with matrix analysis** for exploration
2. **Use comparison analysis** to validate against correlation
3. **Focus on predictors analysis** for modeling
4. **Apply domain knowledge** to interpretation

## Performance Tips

1. **Use sampling** for datasets >5,000 rows
2. **Choose appropriate algorithms** for data type
3. **Adjust CV folds** based on sample size
4. **Leverage caching** for repeated analyses

# Common Pitfalls and Solutions

## Pitfall 1: Overfitting with Small Samples

**Problem**: High PPS scores with very small datasets

**Solution**: Use more conservative CV folds and interpret cautiously

```{r small_sample_solution}
# Small sample - use conservative settings
small_data <- data.frame(
  x = 1:8,
  y = rnorm(8)
)

conservative_result <- jpps(
  data = small_data,
  analysis_type = "single",
  target_var = "y",
  predictor_var = "x",
  cv_folds = 3,  # Fewer folds for small data
  algorithm = "tree"  # Simpler algorithm
)
```

## Pitfall 2: Ignoring Data Types

**Problem**: Treating categorical variables as numeric

**Solution**: Ensure proper factor encoding

```{r data_types_solution}
# Correct categorical encoding
corrected_data <- data.frame(
  category = factor(c("A", "B", "C", "A", "B", "C")),  # Proper factor
  outcome = c(10, 20, 30, 12, 22, 28)
)

correct_result <- jpps(
  data = corrected_data,
  analysis_type = "single",
  target_var = "outcome",
  predictor_var = "category"
)
```

## Pitfall 3: Misinterpreting Causation

**Problem**: Assuming high PPS implies causation

**Solution**: Remember PPS shows **predictive power**, not **causation**

# Conclusion

The `jpps()` function provides a powerful, modern approach to relationship detection that goes beyond traditional correlation analysis. Key advantages include:

✅ **Detects non-linear relationships**  
✅ **Works with mixed data types**  
✅ **Provides asymmetric insights**  
✅ **Uses robust machine learning**  
✅ **Includes performance optimizations**  
✅ **Offers comprehensive visualization**  

## Next Steps

1. **Explore your data** with matrix analysis
2. **Compare with correlation** to identify missed relationships  
3. **Focus on high-PPS predictors** for modeling
4. **Validate findings** with domain knowledge
5. **Use for feature selection** in machine learning pipelines

Remember: PPS is a **"quick and dirty"** exploration tool. Use it to guide more detailed analysis rather than as a final answer.

---

*For more information about ClinicoPath functions, visit our documentation or explore other vignettes in this series.*