---
title: "Categorical Plot Functions"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
library(ClinicoPath)
data(mtcars)
```

This vignette demonstrates the functions designed for categorical data:
`jjbarstats()`, `jjpiestats()` and `jjdotplotstats()`.

## Bar charts with `jjbarstats()`

`jjbarstats()` creates a bar chart and automatically performs a
chi-squared test to compare the distribution of two categorical
variables. The example below compares the number of cylinders (`cyl`)
across transmission types (`am`).

```{r}
jjbarstats(data = mtcars, dep = cyl, group = am, grvar = NULL)
```

## Pie charts with `jjpiestats()`

`jjpiestats()` is similar to `jjbarstats()` but displays the results as a
pie chart.

```{r}
jjpiestats(data = mtcars, dep = cyl, group = am, grvar = NULL)
```

## Dot charts with `jjdotplotstats()`

`jjdotplotstats()` shows group means using a dot plot. In this example we
plot horsepower (`hp`) by engine configuration (`vs`).

```{r}
jjdotplotstats(data = mtcars, dep = hp, group = vs, grvar = NULL)
```

Each function returns a results object whose `plot` element contains the
`ggplot2` visualisation.
