---
title: "Network Visualization with jjarcdiagram"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Network Visualization with jjarcdiagram}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

## Introduction

The `jjarcdiagram()` function in ClinicoPath provides elegant network visualization using arc diagrams. Arc diagrams are particularly effective for displaying hierarchical relationships, sequential connections, and network structures with minimal visual clutter.

### Key Features

- **Multiple Network Types**: Social networks, organizational hierarchies, gene regulation, supply chains
- **Flexible Layouts**: Horizontal and vertical arrangements
- **Dynamic Sizing**: Nodes sized by degree centrality or fixed values
- **Group Coloring**: Automatic color assignment based on categorical variables
- **Weight Visualization**: Arc thickness proportional to connection strength
- **Network Analytics**: Built-in calculation of network metrics and centrality measures
- **Interactive Customization**: Extensive styling and layout options

This vignette demonstrates all features of the jjarcdiagram function using various real-world network datasets.

## Loading Required Libraries and Data

```{r setup}
library(ClinicoPath)

# Load the original arc diagram dataset
data("arcDiagram")

# Preview the data structure
str(arcDiagram)
head(arcDiagram)
```

## Data Requirements

The jjarcdiagram function works with edge list data representing network connections:

### Required Variables:
- **Source Node**: Starting point of connections (factor or character)
- **Target Node**: Endpoint of connections (factor or character)

### Optional Variables:
- **Edge Weight**: Numeric values representing connection strength
- **Node Groups**: Categorical variable for color coding nodes

### Data Format:
- Each row represents one edge/connection
- Source and target can be any identifiers (names, IDs, etc.)
- Weights should be positive numeric values
- Groups can be any categorical variable

## Basic Arc Diagram

### Simple Network Visualization

```{r basic-arc}
# Basic arc diagram without weights or groups
jjarcdiagram(
  data = arcDiagram,
  source = "source",
  target = "target",
  showNodes = TRUE,
  horizontal = TRUE,
  plotTitle = "Basic Network Structure",
  group = NULL,
  weight = NULL
)
```

### Network with Edge Weights

```{r weighted-arc}
# Arc diagram with edge weights
jjarcdiagram(
  data = arcDiagram,
  source = "source",
  target = "target",
  weight = "weight",
  arcWidth = "weight",
  showNodes = TRUE,
  plotTitle = "Weighted Network Connections",
  group = NULL
)
```

### Network with Group Colors

```{r grouped-arc}
# Arc diagram with group-based coloring
jjarcdiagram(
  data = arcDiagram,
  source = "source",
  target = "target",
  weight = "weight",
  group = "group",
  colorByGroup = TRUE,
  showLegend = TRUE,
  showStats = TRUE,
  plotTitle = "Network with Group-Based Coloring"
)
```

## Advanced Network Datasets

### Social Network Analysis

```{r social-network}
# Load social network test data
data("jjarcdiagram_social_network")

# Preview the social network data
str(social_network_data)
head(social_network_data)

# Create social network visualization
jjarcdiagram(
  data = social_network_data,
  source = "source",
  target = "target", 
  weight = "weight",
  group = "group",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  showStats = TRUE,
  plotTitle = "Social Network Analysis - Relationship Strength"
)
```

### Academic Collaboration Network

```{r academic-network}
# Load academic collaboration data
data("jjarcdiagram_academic_network")

# Preview the academic network data
str(academic_network_data)
head(academic_network_data)

# Create academic collaboration visualization
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "publications",
  group = "department",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  arcWidth = "weight",
  sortNodes = "group",
  showStats = TRUE,
  plotTitle = "Academic Collaboration Network by Department"
)
```

### Organizational Hierarchy

```{r org-hierarchy}
# Load organizational hierarchy data
data("jjarcdiagram_org_hierarchy")

# Preview the organizational data
str(org_hierarchy_data)
head(org_hierarchy_data)

# Create organizational hierarchy visualization
jjarcdiagram(
  data = org_hierarchy_data,
  source = "employee",
  target = "reports_to",
  weight = "relationship_strength",
  group = "department",
  colorByGroup = TRUE,
  showLegend = TRUE,
  directed = TRUE,
  sortNodes = "group",
  nodeSize = "degree",
  horizontal = FALSE,
  showStats = TRUE,
  plotTitle = "Organizational Reporting Structure"
)
```

### Gene Regulatory Network

```{r gene-network}
# Load gene regulatory network data
data("jjarcdiagram_gene_network")

# Preview the gene network data
str(gene_network_data)
head(gene_network_data)

# Create gene regulatory network visualization
jjarcdiagram(
  data = gene_network_data,
  source = "regulator",
  target = "target",
  weight = "regulation_score",
  group = "pathway",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  arcWidth = "weight",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  showStats = TRUE,
  plotTitle = "Gene Regulatory Network by Pathway"
)
```

### Supply Chain Network

```{r supply-chain}
# Load supply chain data
data("jjarcdiagram_supply_chain")

# Preview the supply chain data
str(supply_chain_data)
head(supply_chain_data)

# Create supply chain visualization
jjarcdiagram(
  data = supply_chain_data,
  source = "supplier",
  target = "customer",
  weight = "volume",
  group = "industry",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  arcWidth = "weight",
  sortNodes = "group",
  horizontal = TRUE,
  showStats = TRUE,
  plotTitle = "Supply Chain Network by Industry"
)
```

## Layout and Styling Options

### Horizontal vs Vertical Layouts

```{r layout-comparison}
# Horizontal layout
jjarcdiagram(
  data = simple_weighted_network,
  source = "from_node",
  target = "to_node",
  weight = "connection_weight",
  horizontal = TRUE,
  showNodes = TRUE,
  plotTitle = "Horizontal Arc Diagram"
)

# Vertical layout
jjarcdiagram(
  data = simple_weighted_network,
  source = "from_node",
  target = "to_node",
  weight = "connection_weight",
  horizontal = FALSE,
  showNodes = TRUE,
  plotTitle = "Vertical Arc Diagram"
)
```

### Node Sizing Options

```{r node-sizing}
# Fixed node size
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "publications",
  group = "department",
  nodeSize = "fixed",
  nodeSizeValue = 3,
  colorByGroup = TRUE,
  plotTitle = "Fixed Node Sizes"
)

# Degree-based node sizing
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "publications",
  group = "department",
  nodeSize = "degree",
  colorByGroup = TRUE,
  plotTitle = "Degree-Based Node Sizing"
)
```

### Arc Width Customization

```{r arc-width}
# Fixed arc width
jjarcdiagram(
  data = supply_chain_data,
  source = "supplier",
  target = "customer",
  weight = "volume",
  arcWidth = "fixed",
  arcWidthValue = 2,
  plotTitle = "Fixed Arc Width"
)

# Weight-proportional arc width
jjarcdiagram(
  data = supply_chain_data,
  source = "supplier",
  target = "customer",
  weight = "volume",
  arcWidth = "weight",
  arcWidthValue = 3,
  plotTitle = "Weight-Proportional Arc Width"
)
```

### Arc Transparency Effects

```{r transparency}
# Test different transparency levels
transparency_levels <- c(0.2, 0.5, 0.8)

for(alpha in transparency_levels) {
  cat("\n=== Arc Transparency:", alpha, "===\n")
  result <- jjarcdiagram(
    data = social_network_data,
    source = "source",
    target = "target",
    weight = "weight",
    group = "group",
    arcTransparency = alpha,
    colorByGroup = TRUE,
    plotTitle = paste("Arc Transparency =", alpha)
  )
}
```

## Node Sorting and Ordering

### Sorting by Different Criteria

```{r node-sorting}
# Sort by node names
jjarcdiagram(
  data = gene_network_data,
  source = "regulator",
  target = "target",
  weight = "regulation_score",
  group = "pathway",
  sortNodes = "name",
  colorByGroup = TRUE,
  plotTitle = "Sorted by Node Names"
)

# Sort by degree centrality
jjarcdiagram(
  data = gene_network_data,
  source = "regulator",
  target = "target",
  weight = "regulation_score",
  group = "pathway",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  colorByGroup = TRUE,
  plotTitle = "Sorted by Degree (Descending)"
)

# Sort by groups
jjarcdiagram(
  data = gene_network_data,
  source = "regulator",
  target = "target",
  weight = "regulation_score",
  group = "pathway",
  sortNodes = "group",
  colorByGroup = TRUE,
  showLegend = TRUE,
  plotTitle = "Sorted by Groups"
)
```

## Network Direction Options

### Directed vs Undirected Networks

```{r direction-comparison}
# Directed network (default for hierarchical data)
jjarcdiagram(
  data = org_hierarchy_data,
  source = "employee",
  target = "reports_to",
  weight = "relationship_strength",
  group = "department",
  directed = TRUE,
  colorByGroup = TRUE,
  plotTitle = "Directed Network (Organizational Hierarchy)"
)

# Undirected network (treat connections as bidirectional)
jjarcdiagram(
  data = org_hierarchy_data,
  source = "employee",
  target = "reports_to",
  weight = "relationship_strength",
  group = "department",
  directed = FALSE,
  colorByGroup = TRUE,
  plotTitle = "Undirected Network (Mutual Relationships)"
)
```

## Edge Case Scenarios

### Minimal Networks

```{r minimal-network}
# Load minimal test data
data("jjarcdiagram_minimal_network")

# Visualize minimal network
jjarcdiagram(
  data = minimal_network_data,
  source = "from",
  target = "to",
  weight = "strength",
  group = "category",
  colorByGroup = TRUE,
  showStats = TRUE,
  plotTitle = "Minimal Network (3 nodes)"
)
```

### Networks with Self-loops

```{r selfloop-network}
# Load self-loop test data
data("jjarcdiagram_selfloop_network")

# Visualize network with self-loops
jjarcdiagram(
  data = selfloop_network_data,
  source = "node1",
  target = "node2",
  weight = "weight",
  group = "type",
  colorByGroup = TRUE,
  showNodes = TRUE,
  showStats = TRUE,
  plotTitle = "Network with Self-loops"
)
```

### Large Networks

```{r large-network}
# Load large network test data
data("jjarcdiagram_large_network")

# Preview large network
cat("Large network size:", nrow(large_network_data), "edges\n")
cat("Unique nodes:", length(unique(c(large_network_data$source, large_network_data$target))), "\n")

# Visualize large network
jjarcdiagram(
  data = large_network_data,
  source = "source",
  target = "target",
  weight = "weight",
  group = "cluster",
  colorByGroup = TRUE,
  nodeSize = "degree",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  arcWidth = "weight",
  showStats = TRUE,
  plotTitle = "Large Network (50+ nodes)"
)
```

## Network Analytics and Statistics

### Network Metrics Interpretation

```{r network-stats}
# Generate comprehensive network statistics
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "publications",
  group = "department",
  colorByGroup = TRUE,
  showStats = TRUE,
  nodeSize = "degree",
  sortNodes = "degree",
  plotTitle = "Academic Network with Comprehensive Statistics"
)
```

### Understanding Network Metrics

The network statistics provide valuable insights:

- **Network Density**: Ratio of actual to possible connections (0-1 scale)
- **Connectivity**: Whether all nodes can reach each other
- **Degree Centrality**: Number of connections per node
- **Betweenness Centrality**: Nodes that act as bridges between others

## Customization and Styling

### Custom Colors and Labels

```{r custom-styling}
# Custom plot with enhanced labeling
jjarcdiagram(
  data = social_network_data,
  source = "source",
  target = "target",
  weight = "weight",
  group = "group",
  colorByGroup = TRUE,
  showLegend = TRUE,
  labelSize = 1.2,
  nodeSize = "degree",
  nodeSizeValue = 2.5,
  arcWidth = "weight",
  arcTransparency = 0.7,
  plotTitle = "Customized Social Network Visualization",
  showStats = TRUE
)
```

### Multiple Group Variables

```{r multiple-groups}
# Compare different grouping strategies
# grouping_vars <- c("group", "interaction_type")
 
# for(var in grouping_vars) {
#   cat("\n=== Grouping by", var, "===\n")
#   result <- jjarcdiagram(
#     data = social_network_data,
#     source = "source",
#     target = "target",
#     weight = "weight",
#     group = var,
#     colorByGroup = TRUE,
#     showLegend = TRUE,
#     plotTitle = paste("Grouped by", var)
#   )
# }
```

## Clinical and Research Applications

### Biological Networks

```{r biological-applications}
# Gene regulation network analysis
jjarcdiagram(
  data = gene_network_data,
  source = "regulator",
  target = "target",
  weight = "regulation_score",
  group = "effect_type",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  sortNodes = "degree",
  showStats = TRUE,
  plotTitle = "Gene Regulatory Network by Effect Type"
)
```

### Organizational Analysis

```{r organizational-applications}
# Department-level analysis
jjarcdiagram(
  data = org_hierarchy_data,
  source = "employee",
  target = "reports_to",
  weight = "relationship_strength",
  group = "level",
  colorByGroup = TRUE,
  showLegend = TRUE,
  directed = TRUE,
  sortNodes = "group",
  horizontal = FALSE,
  showStats = TRUE,
  plotTitle = "Organizational Structure by Hierarchy Level"
)
```

### Collaboration Networks

```{r collaboration-applications}
# Research collaboration analysis
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "collaboration_years",
  group = "department",
  colorByGroup = TRUE,
  showLegend = TRUE,
  nodeSize = "degree",
  arcWidth = "weight",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  showStats = TRUE,
  plotTitle = "Long-term Research Collaborations"
)
```

## Performance and Scalability

### Handling Different Network Sizes

```{r performance}
# Compare performance with different network sizes
network_sizes <- list(
  "Small" = minimal_network_data,
  "Medium" = social_network_data,
  "Large" = large_network_data
)

for(size_name in names(network_sizes)) {
  data_set <- network_sizes[[size_name]]
  n_nodes <- length(unique(c(data_set[[1]], data_set[[2]])))
  n_edges <- nrow(data_set)
  
  cat("\n=== Network Size:", size_name, "===\n")
  cat("Nodes:", n_nodes, ", Edges:", n_edges, "\n")
}
```

## Best Practices and Recommendations

### Data Preparation Guidelines

1. **Data Quality**: Ensure clean source and target identifiers
2. **Weight Scaling**: Normalize weights for better visualization
3. **Group Selection**: Choose meaningful categorical variables
4. **Edge Validation**: Remove duplicate or invalid connections

### Visualization Guidelines

1. **Layout Choice**: Use horizontal for wide networks, vertical for tall ones
2. **Node Sizing**: Use degree-based sizing for centrality emphasis
3. **Color Coding**: Limit groups to 8-10 for clarity
4. **Arc Transparency**: Use 0.5-0.7 for overlapping connections

### Analysis Workflow

1. **Explore Data Structure**: Check node distribution and edge patterns
2. **Basic Visualization**: Start with simple layout and default settings
3. **Add Grouping**: Include categorical variables for insights
4. **Customize Appearance**: Adjust sizing, colors, and layout
5. **Analyze Metrics**: Review network statistics for interpretation

## Advanced Features

### Network Comparison

```{r network-comparison}
# Compare networks with different parameters
comparison_data <- academic_network_data

# Network sorted by names
jjarcdiagram(
  data = comparison_data,
  source = "author1",
  target = "author2",
  weight = "publications",
  group = "department",
  sortNodes = "name",
  colorByGroup = TRUE,
  plotTitle = "Academic Network - Alphabetical Order"
)

# Network sorted by centrality
jjarcdiagram(
  data = comparison_data,
  source = "author1",
  target = "author2",
  weight = "publications", 
  group = "department",
  sortNodes = "degree",
  sortDecreasing = TRUE,
  colorByGroup = TRUE,
  plotTitle = "Academic Network - Centrality Order"
)
```

### Dynamic Network Properties

```{r dynamic-properties}
# Analyze temporal aspects using collaboration years
jjarcdiagram(
  data = academic_network_data,
  source = "author1",
  target = "author2",
  weight = "collaboration_years",
  group = "department",
  colorByGroup = TRUE,
  showLegend = TRUE,
  arcWidth = "weight",
  nodeSize = "degree",
  showStats = TRUE,
  plotTitle = "Collaboration Duration Network"
)
```

## Conclusion

The `jjarcdiagram()` function provides a powerful and flexible approach to network visualization that is particularly well-suited for:

### Key Advantages:

- **Clarity**: Minimal visual clutter compared to traditional network plots
- **Scalability**: Handles networks from 3 to 100+ nodes effectively
- **Flexibility**: Supports various network types and data formats
- **Analytics**: Built-in network metrics and centrality measures
- **Customization**: Extensive styling and layout options

### Ideal Use Cases:

- **Hierarchical Networks**: Organizational charts, taxonomy trees
- **Sequential Processes**: Workflow diagrams, pipeline analysis
- **Collaboration Networks**: Research partnerships, social connections
- **Biological Networks**: Gene regulation, protein interactions
- **Supply Chains**: Manufacturing and distribution networks

### Best Applications:

- Networks with clear directional flow
- Hierarchical or semi-hierarchical structures
- Medium-sized networks (10-100 nodes)
- Networks where edge overlap is problematic in traditional layouts

For additional resources and advanced network analysis techniques, see the ClinicoPath documentation and other visualization vignettes in this package.
