---
title: "Lollipop Charts for Clinical Categorical Data Visualization"
author: "ClinicoPath Package"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  dpi = 300,
  echo = TRUE,
  eval = FALSE
)

# Load required packages
suppressPackageStartupMessages({
  library(jmvcore)
  library(ggplot2)
  library(dplyr)
  if (requireNamespace("ClinicoPath", quietly = TRUE)) {
    library(ClinicoPath)
  }
})
```

# Introduction to Lollipop Charts in Clinical Research

## What are Lollipop Charts?

Lollipop charts are a modern and effective visualization tool that combines the clarity of bar charts with the elegance of dot plots. They consist of **dots (circles) connected to a baseline by lines (stems)**, creating a visual resemblance to lollipops. In clinical and pathological research, lollipop charts excel at displaying categorical data with emphasis on individual values, making them ideal for:

- **Patient-level visualizations**: Individual biomarker levels, treatment responses, or outcomes
- **Treatment comparisons**: Comparing efficacy across different therapeutic interventions
- **Quality metrics**: Displaying performance indicators across departments or time periods
- **Survey responses**: Visualizing satisfaction scores or assessment results
- **Ranking displays**: Showing ordered outcomes or performance measures

### Key Advantages Over Traditional Charts

1. **Reduced Visual Clutter**: Lower ink-to-data ratio compared to bar charts
2. **Enhanced Focus**: Emphasizes individual data points rather than areas
3. **Better for Sparse Data**: Ideal when categories have few observations
4. **Professional Appearance**: Clean, modern aesthetic suitable for publications
5. **Flexible Orientation**: Works well in both vertical and horizontal layouts
6. **Highlighting Capability**: Easy to emphasize specific categories or outliers

## When to Use Lollipop Charts

### Ideal Clinical Scenarios

- **Patient Timeline Analysis**: Days to response, progression-free survival, treatment duration
- **Biomarker Profiling**: Individual patient biomarker levels across a cohort
- **Treatment Response Visualization**: Response rates or efficacy scores by treatment type
- **Quality Improvement Dashboards**: Performance metrics across clinical departments
- **Survey and Assessment Results**: Patient satisfaction or clinical assessment scores
- **Diagnostic Test Performance**: Sensitivity, specificity, or accuracy across different tests

### Data Requirements

- **Categorical Variable**: Patient IDs, treatment types, departments, or other groupings
- **Continuous Outcome**: Numeric values such as biomarker levels, scores, or measurements
- **Optional Highlighting**: Ability to emphasize specific categories or outliers
- **Minimum Data**: At least 2 categories for meaningful comparison

### When NOT to Use Lollipop Charts

- **Time Series Data**: Use line charts instead for temporal trends
- **Proportional Data**: Use pie charts or stacked bars for part-to-whole relationships
- **Dense Categorical Data**: Consider bar charts for many categories with multiple series
- **Continuous vs. Continuous**: Use scatter plots for two continuous variables

# Statistical Background

## Descriptive Statistics in Lollipop Charts

### Central Tendency Measures
Lollipop charts can incorporate reference lines for:

- **Mean**: $\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$
- **Median**: The middle value when data is ordered
- **Mode**: The most frequently occurring value (for discrete data)

### Variability Measures
Understanding the spread of data displayed:

- **Standard Deviation**: $s = \sqrt{\frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n-1}}$
- **Interquartile Range (IQR)**: $Q_3 - Q_1$
- **Range**: $\max(x) - \min(x)$

### Ranking and Sorting
Lollipop charts are particularly effective for displaying **ranked data**:

- **Ascending Order**: Lowest to highest values
- **Descending Order**: Highest to lowest values  
- **Alphabetical Order**: Categorical ordering
- **Clinical Significance**: Custom ordering based on clinical importance

## Comparative Analysis

### Between-Group Comparisons
When comparing categories, consider:

- **Effect Size**: Clinical significance of differences
- **Confidence Intervals**: Uncertainty around individual estimates
- **Statistical Tests**: ANOVA, Kruskal-Wallis, or appropriate tests
- **Multiple Comparisons**: Adjustments for multiple testing

### Outlier Detection
Lollipop charts excel at highlighting **outliers**:

- **Statistical Outliers**: Values > Q3 + 1.5×IQR or < Q1 - 1.5×IQR
- **Clinical Outliers**: Values outside clinical reference ranges
- **Extreme Values**: Observations requiring clinical attention

# Getting Started with Lollipop Charts

## Basic Functionality

The `lollipop()` function creates comprehensive lollipop charts with extensive customization options:

```r
lollipop(
  data = your_data,
  dep = "continuous_variable",      # Y-axis values
  group = "categorical_variable",   # X-axis categories
  sortBy = "original",              # Sorting method
  orientation = "vertical",         # Chart orientation
  colorScheme = "default",          # Color palette
  theme = "default"                 # Overall appearance
)
```

## Essential Parameters

### Core Variables
- **`dep`**: The continuous dependent variable (biomarker levels, scores, measurements)
- **`group`**: The categorical grouping variable (patients, treatments, departments)

### Sorting Options
- **`original`**: Maintains data order
- **`value_asc`**: Sorts by values (low to high)
- **`value_desc`**: Sorts by values (high to low)
- **`group_alpha`**: Alphabetical sorting by category

### Orientation
- **`vertical`**: Traditional vertical lollipops (default)
- **`horizontal`**: Horizontal layout for long category names

# Loading and Preparing Clinical Data

## Dataset 1: Patient Biomarker Analysis

```{r load-biomarker-data}
# Load patient biomarker data
# In practice, load your data using: load("path/to/your/data.RData")

# Create sample biomarker data
set.seed(123)
biomarker_data <- data.frame(
  patient_id = paste0("P", sprintf("%03d", 1:20)),
  biomarker_level = round(c(
    rnorm(12, mean = 35, sd = 8),    # Normal range patients
    rnorm(5, mean = 65, sd = 12),    # Elevated patients
    rnorm(3, mean = 95, sd = 15)     # High-risk patients
  ), 1),
  risk_category = c(
    rep("Low", 12),
    rep("Medium", 5),
    rep("High", 3)
  ),
  age = round(rnorm(20, mean = 62, sd = 12)),
  gender = sample(c("Male", "Female"), 20, replace = TRUE)
)

# Ensure realistic ranges
biomarker_data$biomarker_level <- pmax(pmin(biomarker_data$biomarker_level, 150), 5)

# Display data summary
cat("=== PATIENT BIOMARKER DATA SUMMARY ===\n")
cat("Number of patients:", nrow(biomarker_data), "\n")
cat("Biomarker range:", round(min(biomarker_data$biomarker_level), 1), "-", 
    round(max(biomarker_data$biomarker_level), 1), "ng/mL\n")
cat("Risk categories:\n")
table(biomarker_data$risk_category)
```

## Dataset 2: Treatment Response Comparison

```{r load-treatment-data}
# Create treatment response data
set.seed(456)
treatment_data <- data.frame(
  treatment = c("Chemotherapy_A", "Chemotherapy_B", "Immunotherapy_C", 
                "Targeted_Therapy_D", "Combination_E", "Radiation_F"),
  response_score = round(c(45, 52, 78, 68, 82, 38), 1),
  efficacy = c("Low", "Medium", "High", "High", "High", "Low"),
  cost_thousands = c(25, 30, 85, 120, 150, 15),
  side_effects = c("Moderate", "Mild", "Moderate", "Severe", "Severe", "Mild")
)

cat("=== TREATMENT RESPONSE DATA SUMMARY ===\n")
cat("Number of treatments:", nrow(treatment_data), "\n")
cat("Response range:", min(treatment_data$response_score), "-", 
    max(treatment_data$response_score), "\n")
cat("Efficacy distribution:\n")
table(treatment_data$efficacy)
```

## Dataset 3: Patient Timeline Analysis

```{r load-timeline-data}
# Create patient timeline data
set.seed(789)
timeline_data <- data.frame(
  patient_id = paste0("Patient_", LETTERS[1:12]),
  days_to_event = round(c(45, 120, 78, 200, 156, 89, 67, 134, 178, 92, 145, 103)),
  event_type = c("Response", "Progression", "Response", "Stable", "Progression", 
                 "Response", "Adverse_Event", "Stable", "Progression", "Response", 
                 "Stable", "Response"),
  treatment_arm = rep(c("Control", "Experimental"), 6),
  disease_stage = c("II", "III", "I", "IV", "III", "II", "I", "III", "IV", "II", "III", "I")
)

cat("=== PATIENT TIMELINE DATA SUMMARY ===\n")
cat("Number of patients:", nrow(timeline_data), "\n")
cat("Days to event range:", min(timeline_data$days_to_event), "-", 
    max(timeline_data$days_to_event), "days\n")
cat("Event types:\n")
table(timeline_data$event_type)
```

# Basic Lollipop Chart Examples

## Example 1: Patient Biomarker Levels

```{r basic-biomarker-chart}
# Basic biomarker visualization
biomarker_analysis <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  title = "Patient Biomarker Levels",
  ylabel = "Biomarker Level (ng/mL)",
  xlabel = "Patient ID"
)

# Display the chart
print(biomarker_analysis)
```

### Clinical Interpretation
- **Individual Assessment**: Each lollipop represents one patient's biomarker level
- **Quick Identification**: Easily spot patients with elevated levels (>60 ng/mL)
- **Clinical Thresholds**: Can add reference lines for clinical decision points
- **Patient Communication**: Clear visualization for explaining results to patients

## Example 2: Treatment Response Comparison

```{r basic-treatment-chart}
# Treatment response comparison
treatment_analysis <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  sortBy = "value_desc",
  title = "Treatment Response Comparison",
  ylabel = "Response Score (%)",
  xlabel = "Treatment Type"
)

print(treatment_analysis)
```

### Clinical Insights
- **Ranking**: Treatments ordered by effectiveness (highest to lowest)
- **Comparative Efficacy**: Clear visual comparison of response rates
- **Treatment Selection**: Supports clinical decision-making
- **Performance Gaps**: Identifies significant differences between treatments

## Example 3: Patient Timeline Visualization

```{r basic-timeline-chart}
# Patient timeline analysis
timeline_analysis <- lollipop(
  data = timeline_data,
  dep = "days_to_event",
  group = "patient_id",
  showMean = TRUE,
  title = "Patient Timeline Analysis",
  ylabel = "Days to Event",
  xlabel = "Patient ID"
)

print(timeline_analysis)
```

### Clinical Applications
- **Follow-up Planning**: Identify patients with short time to events
- **Treatment Monitoring**: Track intervention effectiveness over time
- **Risk Stratification**: Patients with early events may need intensive follow-up
- **Resource Allocation**: Plan clinical resources based on event timing

# Advanced Customization Options

## Sorting and Orientation

### Sorting by Values (Ascending)
```{r sorting-ascending}
# Sort biomarkers from lowest to highest
biomarker_sorted_asc <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  sortBy = "value_asc",
  title = "Biomarker Levels - Sorted Ascending",
  ylabel = "Biomarker Level (ng/mL)"
)

print(biomarker_sorted_asc)
```

### Sorting by Values (Descending)
```{r sorting-descending}
# Sort treatments from highest to lowest response
treatment_sorted_desc <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  sortBy = "value_desc",
  title = "Treatment Response - Ranked by Effectiveness",
  ylabel = "Response Score (%)"
)

print(treatment_sorted_desc)
```

### Horizontal Orientation
```{r horizontal-orientation}
# Horizontal layout for long treatment names
treatment_horizontal <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  orientation = "horizontal",
  sortBy = "value_desc",
  title = "Treatment Response - Horizontal Layout",
  xlabel = "Response Score (%)",
  ylabel = "Treatment Type"
)

print(treatment_horizontal)
```

## Color Schemes and Themes

### Clinical Color Scheme
```{r clinical-colors}
# Professional clinical color scheme
biomarker_clinical <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  colorScheme = "clinical",
  theme = "publication",
  title = "Biomarker Levels - Clinical Theme"
)

print(biomarker_clinical)
```

### Colorblind-Safe Palette
```{r colorblind-safe}
# Colorblind-safe visualization
treatment_colorblind <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  colorScheme = "colorblind",
  sortBy = "value_desc",
  title = "Treatment Response - Colorblind Safe"
)

print(treatment_colorblind)
```

### Viridis Color Scheme
```{r viridis-colors}
# Viridis color scheme for continuous feel
timeline_viridis <- lollipop(
  data = timeline_data,
  dep = "days_to_event",
  group = "patient_id",
  colorScheme = "viridis",
  theme = "minimal",
  title = "Patient Timeline - Viridis Theme"
)

print(timeline_viridis)
```

# Highlighting and Emphasis

## Highlighting Specific Categories

### Highlight High-Risk Patient
```{r highlight-patient}
# Highlight a specific patient
biomarker_highlight <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  highlight = "P003",  # Highlight patient P003
  title = "Biomarker Levels - Patient P003 Highlighted",
  ylabel = "Biomarker Level (ng/mL)"
)

print(biomarker_highlight)
```

### Highlight Best Treatment
```{r highlight-treatment}
# Highlight the most effective treatment
treatment_highlight <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  highlight = "Combination_E",
  sortBy = "value_desc",
  title = "Treatment Response - Best Treatment Highlighted"
)

print(treatment_highlight)
```

## Value Labels and Reference Lines

### Display Value Labels
```{r value-labels}
# Show exact values on lollipops
biomarker_values <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  showValues = TRUE,
  sortBy = "value_desc",
  title = "Biomarker Levels with Value Labels"
)

print(biomarker_values)
```

### Add Mean Reference Line
```{r mean-reference}
# Add mean reference line
treatment_mean <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  showMean = TRUE,
  showValues = TRUE,
  title = "Treatment Response with Mean Reference"
)

print(treatment_mean)
```

## Customizing Appearance

### Point Size and Line Width
```{r custom-appearance}
# Customize point size and line width
timeline_custom <- lollipop(
  data = timeline_data,
  dep = "days_to_event",
  group = "patient_id",
  pointSize = 5,
  lineWidth = 2,
  title = "Patient Timeline - Custom Appearance",
  theme = "classic"
)

print(timeline_custom)
```

### Plot Dimensions
```{r custom-dimensions}
# Custom plot dimensions
biomarker_large <- lollipop(
  data = biomarker_data,
  dep = "biomarker_level",
  group = "patient_id",
  width = 1000,
  height = 700,
  title = "Biomarker Levels - Large Format"
)

print(biomarker_large)
```

# Clinical Applications and Use Cases

## Use Case 1: Quality Metrics Dashboard

```{r quality-metrics}
# Create quality metrics data
quality_data <- data.frame(
  metric = c("Patient_Satisfaction", "Wait_Time", "Accuracy", "Efficiency", 
             "Safety_Score", "Readmission_Rate", "Mortality_Rate", "Infection_Rate"),
  value = c(8.5, 6.2, 9.1, 7.8, 8.9, 5.2, 3.1, 4.3),
  target = c(8.0, 7.0, 9.0, 8.0, 9.0, 5.0, 3.0, 4.0),
  department = c("Nursing", "Admin", "Lab", "Surgery", "ICU", "Cardiology", "Oncology", "ICU")
)

# Quality metrics dashboard
quality_dashboard <- lollipop(
  data = quality_data,
  dep = "value",
  group = "metric",
  sortBy = "value_desc",
  showValues = TRUE,
  title = "Hospital Quality Metrics Dashboard",
  ylabel = "Score / Rate",
  xlabel = "Quality Metric",
  theme = "publication"
)

print(quality_dashboard)
```

### Clinical Value
- **Performance Monitoring**: Track key hospital performance indicators
- **Benchmark Comparison**: Compare against targets or national averages
- **Priority Setting**: Identify areas needing improvement
- **Stakeholder Communication**: Clear visualization for hospital administration

## Use Case 2: Survey Response Analysis

```{r survey-analysis}
# Create survey data
survey_data <- data.frame(
  question = paste0("Q", 1:10),
  satisfaction_score = c(7.2, 8.1, 6.9, 8.7, 7.5, 8.3, 6.8, 7.9, 8.2, 7.1),
  category = c("Service", "Care", "Environment", "Staff", "Communication", 
               "Wait_Time", "Comfort", "Information", "Respect", "Overall"),
  response_rate = c(89, 92, 87, 94, 91, 88, 85, 90, 93, 95)
)

# Survey response visualization
survey_analysis <- lollipop(
  data = survey_data,
  dep = "satisfaction_score",
  group = "category",
  sortBy = "value_desc",
  orientation = "horizontal",
  showValues = TRUE,
  showMean = TRUE,
  title = "Patient Satisfaction Survey Results",
  xlabel = "Satisfaction Score (1-10)",
  ylabel = "Survey Category",
  colorScheme = "clinical"
)

print(survey_analysis)
```

### Clinical Applications
- **Patient Experience**: Monitor patient satisfaction across different aspects
- **Improvement Priorities**: Identify areas with lowest satisfaction
- **Trend Monitoring**: Track changes in satisfaction over time
- **Staff Feedback**: Provide specific feedback to different departments

## Use Case 3: Diagnostic Test Performance

```{r diagnostic-performance}
# Create diagnostic test data
diagnostic_data <- data.frame(
  test = c("Test_A", "Test_B", "Test_C", "Test_D", "Test_E", "Test_F"),
  sensitivity = c(0.89, 0.92, 0.78, 0.95, 0.83, 0.87),
  specificity = c(0.85, 0.88, 0.92, 0.82, 0.89, 0.84),
  accuracy = c(0.87, 0.90, 0.85, 0.88, 0.86, 0.85),
  cost = c(50, 120, 200, 300, 80, 150)
)

# Diagnostic test sensitivity comparison
diagnostic_sensitivity <- lollipop(
  data = diagnostic_data,
  dep = "sensitivity",
  group = "test",
  sortBy = "value_desc",
  showValues = TRUE,
  highlight = "Test_D",
  title = "Diagnostic Test Sensitivity Comparison",
  ylabel = "Sensitivity",
  xlabel = "Diagnostic Test",
  theme = "publication"
)

print(diagnostic_sensitivity)
```

### Clinical Decision Support
- **Test Selection**: Compare diagnostic performance across tests
- **Cost-Effectiveness**: Balance performance with cost considerations
- **Clinical Guidelines**: Support evidence-based test selection
- **Quality Assurance**: Monitor test performance over time

# Statistical Integration and Analysis

## Descriptive Statistics

```{r descriptive-stats}
# Calculate descriptive statistics for biomarker data
biomarker_stats <- biomarker_data %>%
  summarise(
    n = n(),
    mean = round(mean(biomarker_level), 2),
    median = round(median(biomarker_level), 2),
    sd = round(sd(biomarker_level), 2),
    min = round(min(biomarker_level), 2),
    max = round(max(biomarker_level), 2),
    q25 = round(quantile(biomarker_level, 0.25), 2),
    q75 = round(quantile(biomarker_level, 0.75), 2)
  )

cat("=== BIOMARKER DESCRIPTIVE STATISTICS ===\n")
cat("Sample size:", biomarker_stats$n, "\n")
cat("Mean:", biomarker_stats$mean, "ng/mL\n")
cat("Median:", biomarker_stats$median, "ng/mL\n")
cat("Standard deviation:", biomarker_stats$sd, "ng/mL\n")
cat("Range:", biomarker_stats$min, "-", biomarker_stats$max, "ng/mL\n")
cat("Interquartile range:", biomarker_stats$q25, "-", biomarker_stats$q75, "ng/mL\n")
```

## Outlier Detection

```{r outlier-detection}
# Identify outliers in biomarker data
Q1 <- quantile(biomarker_data$biomarker_level, 0.25)
Q3 <- quantile(biomarker_data$biomarker_level, 0.75)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- biomarker_data %>%
  filter(biomarker_level < lower_bound | biomarker_level > upper_bound)

cat("=== OUTLIER ANALYSIS ===\n")
cat("Lower bound:", round(lower_bound, 2), "ng/mL\n")
cat("Upper bound:", round(upper_bound, 2), "ng/mL\n")
cat("Number of outliers:", nrow(outliers), "\n")

if (nrow(outliers) > 0) {
  cat("Outlier patients:\n")
  print(outliers[, c("patient_id", "biomarker_level", "risk_category")])
}
```

## Comparative Analysis

```{r comparative-analysis}
# Compare biomarker levels by risk category
risk_comparison <- biomarker_data %>%
  group_by(risk_category) %>%
  summarise(
    n = n(),
    mean = round(mean(biomarker_level), 2),
    median = round(median(biomarker_level), 2),
    sd = round(sd(biomarker_level), 2),
    .groups = 'drop'
  )

cat("=== BIOMARKER LEVELS BY RISK CATEGORY ===\n")
print(risk_comparison)

# Statistical test for differences
if (requireNamespace("stats", quietly = TRUE)) {
  kruskal_test <- kruskal.test(biomarker_level ~ risk_category, data = biomarker_data)
  cat("\nKruskal-Wallis test for differences between risk categories:\n")
  cat("Chi-square =", round(kruskal_test$statistic, 3), "\n")
  cat("p-value =", round(kruskal_test$p.value, 6), "\n")
  cat("Interpretation:", ifelse(kruskal_test$p.value < 0.05, 
                              "Significant differences between groups", 
                              "No significant differences between groups"), "\n")
}
```

# Advanced Features and Customization

## Comprehensive Feature Combination

```{r comprehensive-example}
# Create a comprehensive example with all features
comprehensive_analysis <- lollipop(
  data = treatment_data,
  dep = "response_score",
  group = "treatment",
  highlight = "Combination_E",
  sortBy = "value_desc",
  orientation = "horizontal",
  showValues = TRUE,
  showMean = TRUE,
  colorScheme = "clinical",
  theme = "publication",
  pointSize = 4,
  lineWidth = 1.5,
  xlabel = "Response Score (%)",
  ylabel = "Treatment Type",
  title = "Comprehensive Treatment Response Analysis",
  width = 1000,
  height = 600
)

print(comprehensive_analysis)
```

## Multiple Dataset Comparison

```{r multiple-datasets}
# Compare different aspects of the same treatments
cat("=== TREATMENT COMPARISON ACROSS MULTIPLE METRICS ===\n")

# Response score analysis
cat("\n1. Response Score Analysis:\n")
response_analysis <- treatment_data %>%
  select(treatment, response_score) %>%
  arrange(desc(response_score))
print(response_analysis)

# Cost analysis
cat("\n2. Cost Analysis:\n")
cost_analysis <- treatment_data %>%
  select(treatment, cost_thousands) %>%
  arrange(cost_thousands)
print(cost_analysis)

# Cost-effectiveness ratio
cat("\n3. Cost-Effectiveness Analysis:\n")
cost_effectiveness <- treatment_data %>%
  mutate(cost_per_response = round(cost_thousands / response_score * 100, 2)) %>%
  select(treatment, response_score, cost_thousands, cost_per_response) %>%
  arrange(cost_per_response)
print(cost_effectiveness)
```

# Best Practices and Clinical Guidelines

## Data Preparation Guidelines

### 1. Data Quality Assessment
```{r data-quality}
# Check data quality before visualization
check_data_quality <- function(data, dep_var, group_var) {
  cat("=== DATA QUALITY ASSESSMENT ===\n")
  
  # Check for missing values
  missing_dep <- sum(is.na(data[[dep_var]]))
  missing_group <- sum(is.na(data[[group_var]]))
  
  cat("Missing values in dependent variable:", missing_dep, "\n")
  cat("Missing values in grouping variable:", missing_group, "\n")
  
  # Check for duplicates
  duplicates <- sum(duplicated(data[[group_var]]))
  cat("Duplicate categories:", duplicates, "\n")
  
  # Check data types
  cat("Dependent variable type:", class(data[[dep_var]]), "\n")
  cat("Grouping variable type:", class(data[[group_var]]), "\n")
  
  # Check ranges
  if (is.numeric(data[[dep_var]])) {
    cat("Value range:", round(min(data[[dep_var]], na.rm = TRUE), 2), "-", 
        round(max(data[[dep_var]], na.rm = TRUE), 2), "\n")
  }
  
  # Check number of categories
  n_categories <- length(unique(data[[group_var]]))
  cat("Number of categories:", n_categories, "\n")
  
  if (n_categories > 50) {
    cat("WARNING: Many categories (>50) may result in cluttered visualization\n")
  }
}

# Check biomarker data quality
check_data_quality(biomarker_data, "biomarker_level", "patient_id")
```

### 2. Clinical Reference Ranges
```{r clinical-ranges}
# Define clinical reference ranges
clinical_ranges <- list(
  biomarker = list(
    normal = c(0, 40),
    elevated = c(40, 70),
    high = c(70, 150)
  ),
  response = list(
    poor = c(0, 30),
    moderate = c(30, 60),
    good = c(60, 100)
  )
)

# Function to categorize values
categorize_biomarker <- function(value) {
  if (value <= 40) return("Normal")
  if (value <= 70) return("Elevated")
  return("High")
}

# Apply clinical categorization
biomarker_data$clinical_category <- sapply(biomarker_data$biomarker_level, categorize_biomarker)

cat("=== CLINICAL CATEGORIZATION ===\n")
table(biomarker_data$clinical_category)
```

## Visualization Best Practices

### 1. Color Usage Guidelines
```{r color-guidelines}
cat("=== COLOR USAGE GUIDELINES ===\n")
cat("1. Use colorblind-safe palettes for publications\n")
cat("2. Limit to 6-8 colors maximum for clarity\n")
cat("3. Use highlighting sparingly for emphasis\n")
cat("4. Consider cultural color associations (red = danger, green = safe)\n")
cat("5. Maintain consistency across related charts\n")
```

### 2. Chart Layout Recommendations
```{r layout-recommendations}
cat("=== CHART LAYOUT RECOMMENDATIONS ===\n")
cat("1. Use horizontal orientation for long category names\n")
cat("2. Sort by value for ranking displays\n")
cat("3. Include value labels for precise communication\n")
cat("4. Add reference lines for clinical thresholds\n")
cat("5. Use appropriate aspect ratios (width:height)\n")
```

### 3. Statistical Annotation
```{r statistical-annotation}
cat("=== STATISTICAL ANNOTATION GUIDELINES ===\n")
cat("1. Include sample sizes when relevant\n")
cat("2. Add confidence intervals for estimates\n")
cat("3. Note statistical significance levels\n")
cat("4. Provide effect size measures\n")
cat("5. Include relevant clinical thresholds\n")
```

# Clinical Interpretation and Communication

## Interpretation Framework

### 1. Visual Pattern Recognition
```{r pattern-recognition}
cat("=== VISUAL PATTERN RECOGNITION ===\n")
cat("1. Outliers: Points far from the group\n")
cat("2. Clusters: Groups of similar values\n")
cat("3. Trends: Gradual increases or decreases\n")
cat("4. Gaps: Missing values or ranges\n")
cat("5. Skewness: Asymmetric distributions\n")
```

### 2. Clinical Significance Assessment
```{r clinical-significance}
# Assess clinical significance of biomarker levels
assess_clinical_significance <- function(data, threshold = 60) {
  high_risk <- sum(data$biomarker_level > threshold)
  total <- nrow(data)
  percentage <- round(high_risk / total * 100, 1)
  
  cat("=== CLINICAL SIGNIFICANCE ASSESSMENT ===\n")
  cat("Patients above threshold (", threshold, " ng/mL):", high_risk, "/", total, " (", percentage, "%)\n")
  cat("Clinical interpretation:", 
      if (percentage > 30) "High proportion of at-risk patients" 
      else if (percentage > 10) "Moderate proportion of at-risk patients"
      else "Low proportion of at-risk patients", "\n")
}

assess_clinical_significance(biomarker_data)
```

## Communication Strategies

### 1. For Clinical Colleagues
```{r clinical-communication}
cat("=== COMMUNICATION FOR CLINICAL COLLEAGUES ===\n")
cat("1. Focus on clinical significance over statistical significance\n")
cat("2. Use familiar clinical terminology and units\n")
cat("3. Highlight actionable findings\n")
cat("4. Provide clear recommendations\n")
cat("5. Include confidence intervals and limitations\n")
```

### 2. For Patients and Families
```{r patient-communication}
cat("=== COMMUNICATION FOR PATIENTS AND FAMILIES ===\n")
cat("1. Use simple, non-technical language\n")
cat("2. Focus on individual patient results\n")
cat("3. Explain what normal ranges mean\n")
cat("4. Provide reassurance and context\n")
cat("5. Offer clear next steps\n")
```

### 3. For Administrators and Stakeholders
```{r admin-communication}
cat("=== COMMUNICATION FOR ADMINISTRATORS ===\n")
cat("1. Emphasize quality metrics and outcomes\n")
cat("2. Include cost-effectiveness considerations\n")
cat("3. Provide benchmarking data\n")
cat("4. Highlight improvement opportunities\n")
cat("5. Quantify potential impacts\n")
```

# Troubleshooting and Common Issues

## Common Error Messages

### 1. Data-Related Errors
```{r data-errors}
cat("=== COMMON DATA-RELATED ERRORS ===\n")
cat("1. 'Variable not found': Check variable names and spelling\n")
cat("2. 'Dependent variable must be numeric': Ensure numeric data type\n")
cat("3. 'At least 2 observations required': Check for sufficient data\n")
cat("4. 'No complete cases found': Address missing values\n")
cat("5. 'Grouping variable must have at least 2 categories': Check group variable\n")
```

### 2. Visualization Issues
```{r visualization-issues}
cat("=== COMMON VISUALIZATION ISSUES ===\n")
cat("1. Overlapping labels: Use horizontal orientation or smaller font\n")
cat("2. Too many categories: Consider grouping or filtering\n")
cat("3. Unclear patterns: Try different sorting or highlighting\n")
cat("4. Poor color contrast: Use colorblind-safe palettes\n")
cat("5. Inappropriate scale: Check for outliers or transform data\n")
```

## Performance Optimization

### 1. Large Dataset Handling
```{r performance-tips}
cat("=== PERFORMANCE OPTIMIZATION TIPS ===\n")
cat("1. Filter data before visualization for large datasets\n")
cat("2. Use sampling for exploratory analysis\n")
cat("3. Consider aggregation for many categories\n")
cat("4. Optimize plot dimensions for intended use\n")
cat("5. Use appropriate file formats for different purposes\n")
```

# Summary and Conclusions

## Key Takeaways

### 1. When to Use Lollipop Charts
- **Individual emphasis**: When highlighting individual data points is important
- **Categorical comparisons**: Comparing values across categories or groups
- **Sparse data**: When data points are few or widely spaced
- **Professional appearance**: For publications and presentations
- **Ranking displays**: When ordering is important

### 2. Clinical Applications
- **Patient-level analysis**: Individual biomarker levels, outcomes, timelines
- **Treatment comparisons**: Efficacy, safety, cost-effectiveness
- **Quality metrics**: Performance indicators, satisfaction scores
- **Survey results**: Patient feedback, staff assessments
- **Diagnostic performance**: Test characteristics, accuracy measures

### 3. Best Practices
- **Data quality**: Ensure clean, complete data before visualization
- **Clinical context**: Use appropriate reference ranges and thresholds
- **Clear communication**: Tailor visualizations to your audience
- **Statistical rigor**: Include appropriate statistical measures
- **Accessibility**: Use colorblind-safe palettes and clear labeling

## Future Directions

### 1. Advanced Features
- **Interactive elements**: Hover information, clickable points
- **Animation**: Showing changes over time
- **Faceting**: Multiple panels for subgroup analysis
- **Statistical overlays**: Confidence intervals, significance tests
- **Export options**: Various formats for different uses

### 2. Integration Opportunities
- **Electronic health records**: Direct data import and visualization
- **Clinical decision support**: Real-time analysis and alerts
- **Quality improvement**: Continuous monitoring and feedback
- **Research applications**: Publication-ready figures
- **Patient engagement**: Personalized health visualizations

## Resources and References

### 1. Statistical Methods
- Descriptive statistics and data visualization principles
- Outlier detection and robust statistics
- Comparative analysis methods
- Clinical reference ranges and thresholds

### 2. Clinical Guidelines
- Quality metric standards
- Patient safety indicators
- Treatment effectiveness measures
- Biomarker interpretation guidelines

### 3. Technical Documentation
- R documentation and package information
- ggplot2 visualization principles
- Color theory and accessibility
- Statistical computing resources

---

*This comprehensive guide provides a thorough introduction to lollipop charts in clinical research. For additional support, consult the package documentation or contact the development team.*
