---
title: "Forest Plot Visualization with jforester"
subtitle: "Publication-Ready Forest Plots for Meta-Analyses and Clinical Trials"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Forest Plot Visualization with jforester}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(ClinicoPath)
library(ggplot2)
library(dplyr)
```

# Introduction

The `jforester` function provides comprehensive forest plot visualization capabilities for meta-analyses, systematic reviews, and clinical trial results. It creates publication-ready forest plots that display point estimates with confidence intervals in a professional format suitable for academic publications.

Forest plots are essential tools in evidence-based medicine for visualizing the results of multiple studies or subgroup analyses. They allow researchers to:

- Display effect sizes with confidence intervals across multiple studies
- Visualize heterogeneity between studies
- Present summary statistics and meta-analysis results
- Compare subgroups or treatment effects
- Communicate statistical findings effectively

# Basic Forest Plot Creation

## Simple Meta-Analysis Forest Plot

Let's start with a basic forest plot using meta-analysis data with odds ratios:

```{r}
# Load meta-analysis data
data(jforester_meta_analysis_data, package = "ClinicoPath")

# Display the first few rows of our dataset
head(jforester_meta_analysis_data)
```

```{r}
# Create a basic forest plot
basic_forest <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  plot_title = "Meta-Analysis of Treatment Effectiveness",
  x_axis_label = "Odds Ratio (95% CI)"
)

# The plot will be displayed in jamovi interface
```

## Understanding the Components

The forest plot displays several key components:

- **Study labels**: Names or identifiers for each study (left side)
- **Point estimates**: Central points representing effect sizes
- **Confidence intervals**: Horizontal lines showing uncertainty
- **Reference line**: Vertical line indicating "no effect" (typically 1 for ratios)
- **Scale**: X-axis showing the range of effect sizes

# Different Effect Types

## Odds Ratios

Odds ratios are commonly used in case-control studies and logistic regression:

```{r}
# Meta-analysis with odds ratios
or_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  plot_title = "Treatment Effect: Odds Ratios",
  x_axis_label = "Odds Ratio",
  reference_line = 1.0
)
```

## Risk Ratios

Risk ratios are used in cohort studies and clinical trials:

```{r}
# Load cardiovascular risk ratio data
data(jforester_cardio_rr_data, package = "ClinicoPath")

# Create risk ratio forest plot
rr_plot <- jforester(
  data = jforester_cardio_rr_data,
  study_labels = "trial_name",
  estimates = "risk_ratio",
  ci_lower = "rr_lower",
  ci_upper = "rr_upper",
  effect_type = "rr",
  plot_title = "Cardiovascular Risk Reduction",
  x_axis_label = "Risk Ratio (95% CI)",
  reference_line = 1.0
)
```

## Hazard Ratios

Hazard ratios are used in survival analysis:

```{r}
# Load survival data
data(jforester_survival_hr_data, package = "ClinicoPath")

# Create hazard ratio forest plot
hr_plot <- jforester(
  data = jforester_survival_hr_data,
  study_labels = "study_id",
  estimates = "hazard_ratio",
  ci_lower = "hr_lower",
  ci_upper = "hr_upper",
  effect_type = "hr",
  plot_title = "Survival Analysis: Treatment Effect",
  x_axis_label = "Hazard Ratio (95% CI)",
  reference_line = 1.0
)
```

## Mean Differences

Mean differences are used for continuous outcomes:

```{r}
# Load mean difference data
data(jforester_mean_diff_data, package = "ClinicoPath")

# Create mean difference forest plot
md_plot <- jforester(
  data = jforester_mean_diff_data,
  study_labels = "study_ref",
  estimates = "mean_difference",
  ci_lower = "md_lower",
  ci_upper = "md_upper",
  effect_type = "md",
  plot_title = "Cognitive Score Improvement",
  x_axis_label = "Mean Difference in Cognitive Score",
  reference_line = 0.0,
  log_scale = FALSE
)
```

## Standardized Mean Differences

Standardized mean differences (Cohen's d) are used for effect size comparisons:

```{r}
# Load standardized mean difference data
data(jforester_smd_data, package = "ClinicoPath")

# Create standardized mean difference forest plot
smd_plot <- jforester(
  data = jforester_smd_data,
  study_labels = "paper_citation",
  estimates = "cohens_d",
  ci_lower = "smd_lower",
  ci_upper = "smd_upper",
  effect_type = "smd",
  plot_title = "Effect Size Analysis",
  x_axis_label = "Standardized Mean Difference (Cohen's d)",
  reference_line = 0.0,
  log_scale = FALSE
)
```

# Advanced Customization

## Color Schemes

The function provides several built-in color schemes:

```{r}
# Medical color scheme (red)
medical_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  color_scheme = "medical",
  plot_title = "Medical Color Scheme"
)

# Forest green color scheme
forest_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  color_scheme = "forest",
  plot_title = "Forest Green Color Scheme"
)

# Custom colors
custom_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  color_scheme = "custom",
  custom_point_color = "#FF5722",
  custom_ci_color = "#9E9E9E",
  plot_title = "Custom Color Scheme"
)
```

## Point Sizes and Sample Size Weighting

Point sizes can reflect sample sizes to provide visual weight indication:

```{r}
# Forest plot with sample size weighting
weighted_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  sample_sizes = "sample_size",
  effect_type = "or",
  point_size_range = "large",
  plot_title = "Sample Size Weighted Forest Plot"
)
```

## Confidence Levels

Different confidence levels can be specified:

```{r}
# 99% confidence intervals
conf_99_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  confidence_level = "99",
  plot_title = "99% Confidence Intervals"
)

# 90% confidence intervals
conf_90_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  confidence_level = "90",
  plot_title = "90% Confidence Intervals"
)
```

# Summary Statistics and Meta-Analysis

## Summary Effect Display

You can display overall summary statistics:

```{r}
# Forest plot with summary effect
summary_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  show_summary = TRUE,
  summary_estimate = 1.45,
  summary_ci_lower = 1.23,
  summary_ci_upper = 1.67,
  plot_title = "Meta-Analysis with Summary Effect"
)
```

## Heterogeneity Statistics

Display heterogeneity information:

```{r}
# Forest plot with heterogeneity statistics
heterogeneity_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  show_heterogeneity = TRUE,
  plot_title = "Forest Plot with Heterogeneity Assessment"
)
```

## Study Weights

Include study weights in the analysis:

```{r}
# Forest plot with study weights
weights_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  sample_sizes = "sample_size",
  effect_type = "or",
  include_weights = TRUE,
  show_table = TRUE,
  plot_title = "Forest Plot with Study Weights"
)
```

# Data Tables and Additional Information

## Comprehensive Data Display

Show detailed data tables alongside the forest plot:

```{r}
# Forest plot with comprehensive data table
comprehensive_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  sample_sizes = "sample_size",
  events = "events",
  effect_type = "or",
  show_table = TRUE,
  include_weights = TRUE,
  plot_title = "Comprehensive Forest Plot Display"
)
```

## Arrow Labels for Interpretation

Add directional arrows to aid interpretation:

```{r}
# Forest plot with arrow labels
arrow_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  arrow_labels = TRUE,
  left_arrow_label = "Favors Control",
  right_arrow_label = "Favors Treatment",
  plot_title = "Forest Plot with Directional Labels"
)
```

# Scale and Reference Lines

## Logarithmic vs Linear Scales

Choose appropriate scales for your data:

```{r}
# Logarithmic scale (default for ratios)
log_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  log_scale = TRUE,
  plot_title = "Logarithmic Scale Forest Plot"
)

# Linear scale (for mean differences)
linear_plot <- jforester(
  data = jforester_mean_diff_data,
  study_labels = "study_ref",
  estimates = "mean_difference",
  ci_lower = "md_lower",
  ci_upper = "md_upper",
  effect_type = "md",
  log_scale = FALSE,
  plot_title = "Linear Scale Forest Plot"
)
```

## Custom Reference Lines

Set custom reference lines for specific contexts:

```{r}
# Custom reference line
custom_ref_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  reference_line = 1.5,
  plot_title = "Custom Reference Line (OR = 1.5)"
)
```

# Complex Meta-Analysis Example

Using our complex meta-analysis dataset with multiple variables:

```{r}
# Load complex meta-analysis data
data(jforester_complex_meta_data, package = "ClinicoPath")

# Display dataset structure
str(jforester_complex_meta_data)
```

```{r}
# Comprehensive forest plot with all features
complex_plot <- jforester(
  data = jforester_complex_meta_data,
  study_labels = "study_label",
  estimates = "effect_estimate",
  ci_lower = "lower_ci",
  ci_upper = "upper_ci",
  sample_sizes = "total_participants",
  effect_type = "or",
  show_summary = TRUE,
  summary_estimate = 1.25,
  summary_ci_lower = 1.12,
  summary_ci_upper = 1.39,
  show_heterogeneity = TRUE,
  include_weights = TRUE,
  show_table = TRUE,
  color_scheme = "medical",
  plot_title = "Comprehensive Meta-Analysis Forest Plot",
  x_axis_label = "Effect Size (Odds Ratio)",
  arrow_labels = TRUE,
  left_arrow_label = "Favors Control",
  right_arrow_label = "Favors Treatment"
)
```

# Publication and Export Options

## Plot Dimensions and Quality

Customize plot dimensions for publication:

```{r}
# High-quality plot for publication
publication_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  plot_width = 12,
  plot_height = 10,
  dpi = 300,
  font_family = "Times",
  export_format = "pdf",
  plot_title = "Publication-Ready Forest Plot"
)
```

## Font Customization

Different font families for various publication requirements:

```{r}
# Arial font (default)
arial_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  font_family = "Arial",
  plot_title = "Arial Font Forest Plot"
)

# Times New Roman font
times_plot <- jforester(
  data = jforester_meta_analysis_data,
  study_labels = "study_name",
  estimates = "odds_ratio",
  ci_lower = "or_lower_ci",
  ci_upper = "or_upper_ci",
  effect_type = "or",
  font_family = "Times",
  plot_title = "Times New Roman Font Forest Plot"
)
```

# Interpretation Guidelines

## Understanding Effect Measures

### Odds Ratios (OR)
- **OR = 1**: No difference between groups
- **OR > 1**: Increased odds in treatment group
- **OR < 1**: Decreased odds in treatment group
- **Clinical significance**: Consider both statistical significance and clinical relevance

### Risk Ratios (RR)
- **RR = 1**: No difference in risk
- **RR > 1**: Increased risk in treatment group
- **RR < 1**: Decreased risk in treatment group
- **Interpretation**: More intuitive than odds ratios for absolute risk

### Hazard Ratios (HR)
- **HR = 1**: No difference in hazard
- **HR > 1**: Increased hazard (worse survival)
- **HR < 1**: Decreased hazard (better survival)
- **Time-to-event**: Represents instantaneous risk over time

### Mean Differences (MD)
- **MD = 0**: No difference between groups
- **MD > 0**: Higher values in treatment group
- **MD < 0**: Lower values in treatment group
- **Units**: Same units as original measurement

### Standardized Mean Differences (SMD)
- **Small effect**: |SMD| = 0.2
- **Medium effect**: |SMD| = 0.5
- **Large effect**: |SMD| = 0.8
- **Unitless**: Allows comparison across different scales

## Assessing Heterogeneity

### Visual Assessment
- **Overlapping CIs**: Suggests consistency
- **Non-overlapping CIs**: Suggests heterogeneity
- **Outliers**: Studies with unusual results

### Statistical Measures
- **I² statistic**: Percentage of variation due to heterogeneity
  - 0-25%: Low heterogeneity
  - 25-50%: Moderate heterogeneity
  - 50-75%: Substantial heterogeneity
  - 75-100%: Considerable heterogeneity

## Clinical Considerations

### Statistical vs Clinical Significance
- **Statistical significance**: p < 0.05 (CI doesn't cross null)
- **Clinical significance**: Meaningful difference in practice
- **Minimum clinically important difference**: Pre-specified threshold

### Confidence Intervals
- **Wide CIs**: Greater uncertainty, smaller studies
- **Narrow CIs**: Less uncertainty, larger studies
- **Crossing null**: Non-significant result

# Best Practices

## Data Quality
1. **Complete case analysis**: Remove studies with missing data
2. **Data validation**: Check for logical inconsistencies
3. **Outlier detection**: Identify and investigate extreme values

## Visualization
1. **Appropriate scale**: Log scale for ratios, linear for differences
2. **Clear labeling**: Descriptive titles and axis labels
3. **Consistent formatting**: Use standard conventions

## Reporting
1. **Study characteristics**: Include sample sizes, study quality
2. **Statistical methods**: Describe analysis approach
3. **Heterogeneity assessment**: Report I² and Q-test results
4. **Sensitivity analysis**: Test robustness of findings

# Troubleshooting Common Issues

## Data Format Issues
```r
# Ensure numeric variables are properly formatted
data$odds_ratio <- as.numeric(data$odds_ratio)
data$ci_lower <- as.numeric(data$ci_lower)
data$ci_upper <- as.numeric(data$ci_upper)
```

## Missing Data Handling
```r
# Check for missing values
summary(data[c("odds_ratio", "ci_lower", "ci_upper")])

# Remove incomplete cases
complete_data <- data[complete.cases(data[c("odds_ratio", "ci_lower", "ci_upper")]), ]
```

## Scale Issues
```r
# For ratios, use log scale
log_scale = TRUE

# For differences, use linear scale
log_scale = FALSE
```

# Conclusion

The `jforester` function provides a comprehensive solution for creating publication-ready forest plots in R. Key features include:

- **Multiple effect types**: Support for OR, RR, HR, MD, and SMD
- **Flexible customization**: Colors, fonts, sizes, and layouts
- **Statistical integration**: Summary statistics and heterogeneity assessment
- **Publication ready**: High-quality output for journals and presentations

Forest plots are essential tools for visualizing meta-analysis results and communicating statistical findings effectively. The `jforester` function makes it easy to create professional-quality forest plots that meet publication standards and enhance the presentation of your research findings.

For more advanced meta-analysis techniques and statistical methods, consider consulting specialized meta-analysis packages and statistical guidance from your institution's biostatistician or epidemiologist.

# References

1. Cochrane Handbook for Systematic Reviews of Interventions
2. PRISMA Statement for Reporting Systematic Reviews
3. Higgins JP, Thompson SG. Quantifying heterogeneity in a meta-analysis. Stat Med. 2002;21(11):1539-58.
4. Borenstein M, Hedges LV, Higgins JP, Rothstein HR. Introduction to Meta-Analysis. Wiley, 2009.
5. Riley RD, Higgins JP, Deeks JJ. Interpretation of random effects meta-analyses. BMJ. 2011;342:d549.