---
title: "Advanced Raincloud Plots - Enhanced Distribution Visualization"
description: "Comprehensive guide to advanced raincloud plots with longitudinal connections, Likert scale support, and flexible positioning using ggrain"
author: "ClinicoPath Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Advanced Raincloud Plots - Enhanced Distribution Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  dpi = 300
)

library(ClinicoPath)
library(ggplot2)
library(dplyr)
```

## Overview

The `advancedraincloud` module provides **enhanced raincloud visualization** using the modern `ggrain` package, offering advanced features beyond standard raincloud plots. This module is specifically designed for complex research scenarios requiring longitudinal connections, Likert scale support, and flexible positioning.

### Key Advantages Over Standard Raincloud

| Feature | Standard Raincloud (ggdist) | Advanced Raincloud (ggrain) |
|---------|---------------------------|----------------------------|
| **Package** | ggdist | ggrain |
| **Longitudinal Connections** | ❌ | ✅ Connect repeated measures |
| **Likert Scale Support** | ❌ | ✅ Y-axis jittering for ordinal data |
| **Positioning Options** | Fixed | ✅ Left/Right/Flanking |
| **Covariate Mapping** | ❌ | ✅ Point color remapping |
| **Clinical Applications** | Basic distribution | ✅ Complex research designs |

### When to Use Advanced Raincloud

- **Longitudinal Studies**: Before/after or repeated measures data
- **Survey Analysis**: Likert scales and ordinal responses  
- **Clinical Trials**: Treatment response with individual tracking
- **Complex Comparisons**: Multi-group analysis with covariates

## Dataset Overview

We'll demonstrate using the comprehensive `histopathology` dataset:

```{r data-overview}
# Load the histopathology dataset
data("histopathology")

# Dataset structure for raincloud plots
str(histopathology[c("Group", "Age", "OverallTime", "Sex", "Grade_Level", "ID")])

# Key variables for advanced raincloud plots
cat("Variables suitable for advanced raincloud visualization:\n")
cat("Y-axis (continuous):", paste(c("Age", "OverallTime", "MeasurementA", "MeasurementB"), collapse = ", "), "\n")
cat("X-axis (grouping):", paste(c("Group", "Sex", "Grade_Level"), collapse = ", "), "\n")
cat("Fill variable:", paste(c("Sex", "Grade_Level"), collapse = ", "), "\n")
cat("ID variable:", "ID (for longitudinal connections)\n")

# Check for repeated measures (simulated)
set.seed(123)
histopathology$ID <- rep(1:125, 2)  # Simulate paired data
histopathology$Time <- rep(c("Baseline", "Follow-up"), each = 125)
cat("Simulated longitudinal data: 125 patients with baseline and follow-up measurements\n")
```

## Basic Advanced Raincloud Plot

Let's start with a basic advanced raincloud plot:

```{r basic-example, eval=FALSE}
# Basic advanced raincloud plot
advancedraincloud(
  data = histopathology,
  y_var = "Age",
  x_var = "Group",
  plot_title = "Age Distribution by Treatment Group",
  show_statistics = TRUE,
  show_interpretation = TRUE
)
```

### R Code Example for Basic Plot

```{r basic-code, eval=TRUE}
# Reproducible R code using ggrain
library(ggrain)
library(ggplot2)

# Basic advanced raincloud with ggrain
basic_plot <- ggplot(histopathology, aes(x = Group, y = Age, fill = Group)) +
  geom_rain(rain.side = "l") +
  scale_fill_manual(values = c("#2E86AB", "#A23B72")) +
  theme_minimal() +
  labs(
    title = "Advanced Raincloud Plot - Basic Example",
    x = "Treatment Group",
    y = "Age (years)"
  ) +
  theme(legend.position = "none")

print(basic_plot)
```

## Longitudinal Connections

The most powerful feature of advanced raincloud plots is connecting repeated observations:

```{r longitudinal-example, eval=FALSE}
# Advanced raincloud with longitudinal connections
advancedraincloud(
  data = histopathology,
  y_var = "OverallTime", 
  x_var = "Time",
  id_var = "ID",
  show_longitudinal = TRUE,
  plot_title = "Overall Survival Time: Baseline vs Follow-up",
  x_label = "Time Point",
  y_label = "Overall Time (months)"
)
```

### R Code Example for Longitudinal Connections

```{r longitudinal-code, eval=TRUE}
# Create simulated longitudinal data
longitudinal_data <- histopathology %>%
  filter(!is.na(OverallTime)) %>%
  slice_head(n = 50) %>%  # Smaller sample for clearer visualization
  mutate(
    Baseline = OverallTime,
    Follow_up = OverallTime + rnorm(n(), 2, 1),  # Simulated follow-up
    ID = row_number()
  ) %>%
  tidyr::pivot_longer(
    cols = c(Baseline, Follow_up),
    names_to = "Time",
    values_to = "Measurement"
  )

# Advanced raincloud with longitudinal connections
longitudinal_plot <- ggplot(longitudinal_data, aes(x = Time, y = Measurement, fill = Time)) +
  geom_rain(
    id.long.var = "ID",  # Connect observations by ID
    rain.side = "f"      # Flanking rainclouds
  ) +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e")) +
  theme_minimal() +
  labs(
    title = "Advanced Raincloud with Longitudinal Connections",
    x = "Time Point",
    y = "Measurement Value",
    caption = "Lines connect repeated observations from the same subjects"
  )

print(longitudinal_plot)
```

## Raincloud Positioning Options

Advanced raincloud plots offer flexible positioning:

```{r positioning-examples, eval=FALSE}
# Left-side raincloud (default)
advancedraincloud(
  data = histopathology,
  y_var = "Age",
  x_var = "Sex", 
  rain_side = "l",
  plot_title = "Left-side Raincloud"
)

# Right-side raincloud
advancedraincloud(
  data = histopathology,
  y_var = "Age",
  x_var = "Sex",
  rain_side = "r", 
  plot_title = "Right-side Raincloud"
)

# Flanking rainclouds (both sides)
advancedraincloud(
  data = histopathology,
  y_var = "Age",
  x_var = "Sex",
  rain_side = "f",
  plot_title = "Flanking Rainclouds"
)
```

### R Code Examples for Different Positions

```{r positioning-code, eval=TRUE}
# Create comparison plots
library(gridExtra)

# Left-side raincloud
left_plot <- ggplot(histopathology, aes(x = Sex, y = Age, fill = Sex)) +
  geom_rain(rain.side = "l") +
  scale_fill_manual(values = c("#e74c3c", "#3498db")) +
  theme_minimal() +
  labs(title = "Left-side Raincloud", x = "Sex", y = "Age") +
  theme(legend.position = "none")

# Right-side raincloud  
right_plot <- ggplot(histopathology, aes(x = Sex, y = Age, fill = Sex)) +
  geom_rain(rain.side = "r") +
  scale_fill_manual(values = c("#e74c3c", "#3498db")) +
  theme_minimal() +
  labs(title = "Right-side Raincloud", x = "Sex", y = "Age") +
  theme(legend.position = "none")

# Flanking rainclouds
flanking_plot <- ggplot(histopathology, aes(x = Sex, y = Age, fill = Sex)) +
  geom_rain(rain.side = "f") +
  scale_fill_manual(values = c("#e74c3c", "#3498db")) +
  theme_minimal() +
  labs(title = "Flanking Rainclouds", x = "Sex", y = "Age") +
  theme(legend.position = "none")

# Display plots
grid.arrange(left_plot, right_plot, flanking_plot, ncol = 3)
```

## Likert Scale Support

For ordinal survey data, advanced raincloud plots offer specialized support:

```{r likert-example, eval=FALSE}
# Simulate Likert scale data
histopathology$Satisfaction <- sample(1:5, nrow(histopathology), replace = TRUE,
                                     prob = c(0.1, 0.2, 0.4, 0.2, 0.1))

# Advanced raincloud for Likert data
advancedraincloud(
  data = histopathology,
  y_var = "Satisfaction",
  x_var = "Group",
  likert_mode = TRUE,  # Enable Likert mode
  plot_title = "Patient Satisfaction by Treatment Group",
  y_label = "Satisfaction Score (1-5)"
)
```

### R Code Example for Likert Data

```{r likert-code, eval=TRUE}
# Create Likert scale data
likert_data <- histopathology %>%
  mutate(
    Satisfaction = sample(1:5, n(), replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
    Satisfaction_Factor = factor(Satisfaction, levels = 1:5, 
                                labels = c("Very Poor", "Poor", "Average", "Good", "Excellent"))
  )

# Advanced raincloud for Likert data with jittering
likert_plot <- ggplot(likert_data, aes(x = Group, y = Satisfaction, fill = Group)) +
  geom_rain(
    likert = TRUE,  # Enable Likert mode for Y-axis jittering
    rain.side = "l"
  ) +
  scale_fill_manual(values = c("#2E86AB", "#A23B72")) +
  scale_y_continuous(breaks = 1:5, labels = c("Very Poor", "Poor", "Average", "Good", "Excellent")) +
  theme_minimal() +
  labs(
    title = "Patient Satisfaction by Treatment Group - Likert Scale",
    x = "Treatment Group",
    y = "Satisfaction Score",
    caption = "Y-axis jittering applied for discrete ordinal responses"
  ) +
  theme(legend.position = "none")

print(likert_plot)
```

## Covariate Mapping

Advanced raincloud plots support point color remapping based on additional variables:

```{r covariate-example, eval=FALSE}
# Advanced raincloud with covariate mapping
advancedraincloud(
  data = histopathology,
  y_var = "OverallTime",
  x_var = "Group",
  cov_var = "Age",  # Map point colors to age
  plot_title = "Overall Time by Group with Age Mapping",
  show_statistics = TRUE
)
```

### R Code Example for Covariate Mapping

```{r covariate-code, eval=TRUE}
# Advanced raincloud with covariate mapping
covariate_plot <- ggplot(histopathology, aes(x = Group, y = OverallTime, fill = Group)) +
  geom_rain(
    rain.side = "l",
    cov = "Age"  # Map point colors to age covariate
  ) +
  scale_fill_manual(values = c("#2E86AB", "#A23B72")) +
  scale_color_viridis_c(name = "Age") +  # Continuous color scale for age
  theme_minimal() +
  labs(
    title = "Overall Time by Group with Age Covariate Mapping",
    x = "Treatment Group", 
    y = "Overall Time (months)",
    caption = "Point colors represent patient age"
  )

print(covariate_plot)
```

## Color Palettes

The module includes 6 professional color palettes:

```{r color-palettes, eval=FALSE}
# Clinical palette (default)
advancedraincloud(data = histopathology, y_var = "Age", x_var = "Grade_Level", 
                 color_palette = "clinical")

# Viridis palette
advancedraincloud(data = histopathology, y_var = "Age", x_var = "Grade_Level",
                 color_palette = "viridis")

# Pastel palette
advancedraincloud(data = histopathology, y_var = "Age", x_var = "Grade_Level",
                 color_palette = "pastel")
```

### R Code Examples for Color Palettes

```{r palette-code, eval=TRUE}
# Create palette comparison
grade_data <- histopathology %>% filter(!is.na(Grade_Level))

# Clinical palette
clinical_colors <- c("#2E86AB", "#A23B72", "#F18F01")
clinical_plot <- ggplot(grade_data, aes(x = Grade_Level, y = Age, fill = Grade_Level)) +
  geom_rain() +
  scale_fill_manual(values = clinical_colors) +
  theme_minimal() +
  labs(title = "Clinical Palette") +
  theme(legend.position = "none")

# Viridis palette
viridis_plot <- ggplot(grade_data, aes(x = Grade_Level, y = Age, fill = Grade_Level)) +
  geom_rain() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Viridis Palette") +
  theme(legend.position = "none")

# Pastel palette  
pastel_colors <- c("#FFB3BA", "#BAFFC9", "#BAE1FF")
pastel_plot <- ggplot(grade_data, aes(x = Grade_Level, y = Age, fill = Grade_Level)) +
  geom_rain() +
  scale_fill_manual(values = pastel_colors) +
  theme_minimal() +
  labs(title = "Pastel Palette") +
  theme(legend.position = "none")

# Display palette comparison
grid.arrange(clinical_plot, viridis_plot, pastel_plot, ncol = 3)
```

## Statistical Integration

Advanced raincloud plots include comprehensive statistical analysis:

```{r statistics-example, eval=FALSE}
# With summary statistics and group comparisons
advancedraincloud(
  data = histopathology,
  y_var = "Age", 
  x_var = "Group",
  show_statistics = TRUE,      # Summary statistics table
  show_comparisons = TRUE,     # Statistical tests
  show_interpretation = TRUE   # Feature guide
)
```

### Statistical Test Details

The module automatically selects appropriate tests:

- **2 Groups**: Wilcoxon rank-sum test (robust, non-parametric)
- **>2 Groups**: Kruskal-Wallis test (robust, non-parametric)

```{r statistics-demo, eval=TRUE}
# Demonstrate statistical testing
group_data <- histopathology %>% filter(Group %in% c("Control", "Treatment"))

# Wilcoxon test for two groups
wilcox_result <- wilcox.test(Age ~ Group, data = group_data)

cat("Statistical Test Results:\n")
cat("Wilcoxon rank-sum test\n")
cat("W =", wilcox_result$statistic, "\n")
cat("p-value =", format.pval(wilcox_result$p.value, digits = 3), "\n")

# Summary statistics by group
summary_stats <- group_data %>%
  group_by(Group) %>%
  summarise(
    n = n(),
    mean = round(mean(Age, na.rm = TRUE), 1),
    median = round(median(Age, na.rm = TRUE), 1),
    sd = round(sd(Age, na.rm = TRUE), 1),
    .groups = 'drop'
  )

print(summary_stats)
```

## Customization Options

Advanced raincloud plots offer extensive customization:

```{r customization-example, eval=FALSE}
# Highly customized advanced raincloud
advancedraincloud(
  data = histopathology,
  y_var = "OverallTime",
  x_var = "Group", 
  fill_var = "Sex",
  point_size = 2.0,          # Larger points
  point_alpha = 0.8,         # Point transparency
  violin_alpha = 0.6,        # Violin transparency
  boxplot_width = 0.15,      # Boxplot width
  plot_title = "Customized Advanced Raincloud Plot",
  x_label = "Treatment Group",
  y_label = "Overall Survival Time (months)"
)
```

### R Code Example for Customization

```{r customization-code, eval=TRUE}
# Highly customized plot
custom_plot <- ggplot(histopathology, aes(x = Group, y = OverallTime, fill = Sex)) +
  geom_rain(
    rain.side = "l",
    point.args = list(size = 2.0, alpha = 0.8),      # Custom points
    violin.args = list(alpha = 0.6),                  # Custom violin
    boxplot.args = list(width = 0.15)                 # Custom boxplot
  ) +
  scale_fill_manual(values = c("#e74c3c", "#3498db")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  ) +
  labs(
    title = "Customized Advanced Raincloud Plot",
    x = "Treatment Group",
    y = "Overall Survival Time (months)",
    fill = "Sex"
  )

print(custom_plot)
```

## Clinical Research Applications

### 1. Before/After Treatment Analysis

```{r clinical-before-after, eval=TRUE}
# Simulate before/after treatment data
treatment_data <- data.frame(
  Patient_ID = rep(1:30, 2),
  Time = rep(c("Before", "After"), each = 30),
  Biomarker = c(
    rnorm(30, 100, 15),                    # Before treatment
    rnorm(30, 85, 12)                      # After treatment (improvement)
  ),
  Treatment = rep(c("Drug A", "Drug B"), each = 15, times = 2)
)

# Advanced raincloud with connections
before_after_plot <- ggplot(treatment_data, aes(x = Time, y = Biomarker, fill = Treatment)) +
  geom_rain(
    id.long.var = "Patient_ID",  # Connect before/after for same patients
    rain.side = "f"              # Flanking for comparison
  ) +
  scale_fill_manual(values = c("#2E86AB", "#A23B72")) +
  theme_minimal() +
  labs(
    title = "Before/After Treatment Analysis with Patient Connections",
    x = "Time Point",
    y = "Biomarker Level",
    fill = "Treatment",
    caption = "Lines connect measurements from the same patients"
  )

print(before_after_plot)
```

### 2. Multi-Group Clinical Trial

```{r clinical-trial, eval=TRUE}
# Simulate clinical trial data
trial_data <- histopathology %>%
  mutate(
    Treatment_Arm = case_when(
      Group == "Control" ~ "Placebo",
      Group == "Treatment" ~ sample(c("Low Dose", "High Dose"), n(), replace = TRUE)
    ),
    Response_Score = case_when(
      Treatment_Arm == "Placebo" ~ rnorm(n(), 3, 1),
      Treatment_Arm == "Low Dose" ~ rnorm(n(), 5, 1.2),
      Treatment_Arm == "High Dose" ~ rnorm(n(), 7, 1.5)
    )
  ) %>%
  filter(!is.na(Treatment_Arm))

# Multi-group comparison
trial_plot <- ggplot(trial_data, aes(x = Treatment_Arm, y = Response_Score, fill = Treatment_Arm)) +
  geom_rain(rain.side = "l") +
  scale_fill_manual(values = c("#95a5a6", "#3498db", "#e74c3c")) +
  theme_minimal() +
  labs(
    title = "Clinical Trial: Treatment Response by Dose",
    x = "Treatment Arm",
    y = "Response Score",
    caption = "Advanced raincloud plot showing distribution differences"
  ) +
  theme(legend.position = "none")

print(trial_plot)
```

## Best Practices

### 1. Choosing the Right Approach

- **Standard Raincloud**: Simple group comparisons, basic distribution analysis
- **Advanced Raincloud**: Repeated measures, longitudinal data, complex designs

### 2. Data Preparation

```{r data-preparation}
# Always check your data structure
cat("Data preparation checklist:\n")
cat("✓ Y-variable: continuous/numeric\n")
cat("✓ X-variable: categorical/factor\n")
cat("✓ ID variable: unique identifier for each subject\n")
cat("✓ No excessive missing values\n")

# Example data check
data_check <- histopathology %>%
  summarise(
    n_total = n(),
    n_complete = sum(complete.cases(Group, Age)),
    missing_pct = round((n_total - n_complete) / n_total * 100, 1)
  )

cat("\nData completeness:", data_check$n_complete, "of", data_check$n_total, 
    "cases (", data_check$missing_pct, "% missing)\n")
```

### 3. Interpretation Guidelines

```{r interpretation-guidelines}
cat("Advanced Raincloud Plot Interpretation:\n")
cat("📊 Violin: Shows probability density distribution\n")
cat("📦 Box: Median, quartiles, and outliers\n") 
cat("🔵 Points: Individual observations\n")
cat("📈 Connections: Longitudinal relationships (when enabled)\n")
cat("🎨 Colors: Group or covariate distinctions\n")
```

## Package Dependencies

The advanced raincloud module requires:

```{r dependencies, eval=FALSE}
# Core dependencies
library(ggrain)      # Enhanced raincloud plots
library(ggplot2)     # Graphics
library(dplyr)       # Data manipulation
library(jmvcore)     # jamovi framework

# Optional enhancements
library(viridis)     # Color palettes
library(RColorBrewer) # Additional palettes
```

## Conclusion

Advanced raincloud plots represent a significant enhancement over standard raincloud visualization, providing:

### Key Benefits

1. **Longitudinal Connections**: Track individual subjects across time/conditions
2. **Likert Scale Support**: Specialized handling for ordinal survey data
3. **Flexible Positioning**: Optimal layout for different research designs
4. **Covariate Integration**: Multi-dimensional data visualization
5. **Statistical Integration**: Robust non-parametric testing
6. **Clinical Focus**: Designed for medical and clinical research

### When to Use

- **Repeated Measures Studies**: Before/after, pre/post designs
- **Survey Research**: Likert scales, ordinal responses
- **Clinical Trials**: Treatment comparisons with individual tracking
- **Longitudinal Analysis**: Multi-timepoint studies
- **Complex Visualizations**: Multi-group, multi-variable analyses

### Next Steps

- Explore the interactive features in jamovi
- Experiment with different positioning options
- Try longitudinal connections with your repeated measures data
- Use covariate mapping for multi-dimensional analysis

For additional statistical plotting modules in ClinicoPath:
- `raincloud` for standard raincloud plots
- `jjscatterstats` for correlation visualization
- `jjbarstats` for categorical data analysis
- `jjhistostats` for distribution analysis