---
title: "Introduction to jjstatsplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jjstatsplot}
  %\VignetteEngine{rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

`jjstatsplot` provides easy wrappers around functions from the
[`ggstatsplot`](https://indrajeetpatil.github.io/ggstatsplot/) package.
These wrappers are primarily used by the *jamovi* interface but can also
be called directly from R. The package includes functions for creating
bar charts, box--violin plots, scatter plots, histograms, and more while
reporting relevant statistical tests automatically.

The main functions exported are:

- `jjbarstats()` -- bar charts for categorical variables
- `jjbetweenstats()` -- compare continuous variables between groups
- `jjcorrmat()` -- correlation matrix visualisations
- `jjdotplotstats()` -- dot charts with summary statistics
- `jjhistostats()` -- histograms with statistical annotations
- `jjpiestats()` -- pie charts for categorical data
- `jjscatterstats()` -- scatter plots with regression details
- `jjwithinstats()` -- paired comparisons of repeated measures

All functions return a jamovi results object, but the primary component
of interest is usually the generated **ggplot2** object stored in the
`plot` element. In this vignette we give a quick overview of how to call
the wrappers from R.

We will use the built-in `mtcars` data set for the examples below.

```{r}
library(jjstatsplot)
data(mtcars)
head(mtcars)
```

The following sections demonstrate some of the available plots.


