---
title: "jjhistostats: Comprehensive Histogram Analysis"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

The `jjhistostats` function provides a comprehensive interface for creating statistical histograms that visualize the distribution of continuous variables with normality testing and statistical summaries. This function is a wrapper around `ggstatsplot::gghistostats` and `ggstatsplot::grouped_gghistostats`, offering both single and grouped histogram visualizations with advanced statistical analysis.

## Key Features

- **Multiple distribution assessment methods**: Parametric (Shapiro-Wilk), nonparametric (Anderson-Darling), robust, and Bayesian approaches
- **Flexible bin width control**: Automatic or manual bin width specification
- **Normal curve overlay**: Compare data distribution with theoretical normal distribution
- **Centrality indicators**: Display mean, median, or robust centrality measures
- **Multi-variable support**: Create multiple histograms in a single analysis
- **Grouped analysis**: Separate histograms for different subgroups
- **Performance optimized**: Uses internal caching to eliminate redundant computations

## Installation and Setup

```{r setup}
echo: true
eval: false

# Install ClinicoPath if not already installed
if (!require("ClinicoPath")) {
  devtools::install_github("sbalci/ClinicoPathJamoviModule")
}

library(ClinicoPath)
library(ggplot2)
```

## Quick Start

### Basic Histogram

```{r basic-example}
echo: true
eval: false

# Load test data
data(jjhistostats_test_data)

# Basic histogram with normality testing
result <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "age_years",
  typestatistics = "parametric",
  normalcurve = TRUE,
  centralityline = TRUE
)

# View the plot
result$plot
```

### Multiple Variables Histogram

```{r multiple-example}
echo: true
eval: false

# Multiple histograms in one analysis
result_multiple <- jjhistostats(
  data = jjhistostats_test_data,
  dep = c("age_years", "tumor_size_mm", "bmi"),
  typestatistics = "parametric",
  resultssubtitle = TRUE,
  centralityline = TRUE
)

# View the combined plot
result_multiple$plot
```

### Grouped Histogram

```{r grouped-example}
echo: true
eval: false

# Grouped histogram by disease stage
result_grouped <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "psa_level",
  grvar = "disease_stage",
  typestatistics = "nonparametric",
  normalcurve = TRUE,
  changebinwidth = TRUE,
  binwidth = 2.0
)

# View the grouped plot
result_grouped$plot2
```

## Function Parameters

### Core Parameters

- **`data`**: Input data frame containing the variables to analyze
- **`dep`**: One or more continuous numeric variables for histogram creation
- **`grvar`**: Optional grouping variable for stratified analysis
- **`typestatistics`**: Type of normality test to perform

### Statistical Methods

The `typestatistics` parameter supports four different approaches:

1. **"parametric"** (default): Shapiro-Wilk normality test
   - Best for sample sizes 3 < n < 5000
   - Tests null hypothesis of normality
   - Most commonly used in clinical research

2. **"nonparametric"**: Anderson-Darling test
   - More powerful than Shapiro-Wilk for detecting departures from normality
   - Better for detecting deviations in the tails
   - Good for larger sample sizes

3. **"robust"**: Robust normality tests
   - Less sensitive to outliers
   - Uses robust statistical measures
   - Good when data may have contamination

4. **"bayes"**: Bayesian normality assessment
   - Provides Bayes factors for normality
   - Quantifies evidence for/against normality
   - Incorporates prior beliefs

### Display Options

#### Binning Control

- **`changebinwidth`**: Enable manual bin width control
- **`binwidth`**: Specify exact bin width (when changebinwidth = TRUE)

Automatic binning uses the formula: `(max(x) - min(x)) / sqrt(n)`

#### Visual Elements

- **`centralityline`**: Display vertical line for central tendency
  - Mean for parametric analysis
  - Median for nonparametric analysis
  - Trimmed mean for robust analysis

- **`normalcurve`**: Overlay theoretical normal distribution
  - Uses sample mean and standard deviation
  - Helps visualize departure from normality

- **`resultssubtitle`**: Show statistical test results as subtitle

## Performance Optimizations

### Version History

The function has been significantly optimized for performance:

**Previous Issues:**
- Redundant data processing in `.plot()` and `.plot2()` methods
- No caching infrastructure
- Repeated variable conversion and formula construction
- Inefficient multi-variable handling

**Current Optimizations:**
- **Data Caching**: Uses `.prepareData()` method to cache processed data
- **Options Caching**: Uses `.prepareOptions()` method to cache option processing
- **Eliminated Redundancy**: Both plot methods now use cached results
- **Efficient Multi-Variable Processing**: Shared data preparation for multiple histograms
- **Better Progress Feedback**: Clear user messaging during processing

### Performance Benefits

```{r performance-comparison}
echo: true
eval: false

# Performance comparison (conceptual)
# Before optimization: 
# - Data processed separately for each plot method
# - Variable conversion repeated for each histogram
# - Options processing duplicated

# After optimization:
# - Data processed once and cached
# - Variable conversion done once for all variables
# - Significant speedup for multiple variables and grouped analysis
```

## Advanced Usage Examples

### Distribution Shape Analysis

```{r distribution-shapes}
echo: true
eval: false

# Analyze different distribution shapes in the test data

# Normal distribution (age)
age_hist <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "age_years",
  typestatistics = "parametric",
  normalcurve = TRUE,
  centralityline = TRUE,
  resultssubtitle = TRUE
)

# Right-skewed distribution (tumor size)
tumor_hist <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "tumor_size_mm",
  typestatistics = "nonparametric",
  normalcurve = TRUE,
  changebinwidth = TRUE,
  binwidth = 5
)

# Bimodal distribution (PSA levels)
psa_hist <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "psa_level",
  typestatistics = "robust",
  normalcurve = TRUE,
  changebinwidth = TRUE,
  binwidth = 3
)
```

### Clinical Biomarker Analysis

```{r clinical-example}
echo: true
eval: false

# Comprehensive biomarker distribution analysis
biomarkers <- c("hemoglobin_level", "creatinine_level", "crp_level")

# Analyze all biomarkers simultaneously
biomarker_analysis <- jjhistostats(
  data = jjhistostats_test_data,
  dep = biomarkers,
  typestatistics = "parametric",
  centralityline = TRUE,
  resultssubtitle = TRUE,
  normalcurve = TRUE
)

# Stratified analysis by treatment group
biomarker_by_treatment <- jjhistostats(
  data = jjhistostats_test_data,
  dep = "crp_level",
  grvar = "treatment_group",
  typestatistics = "nonparametric",
  normalcurve = TRUE,
  changebinwidth = TRUE,
  binwidth = 5
)
```

### Normality Assessment Workflow

```{r normality-assessment}
echo: true
eval: false

# Systematic normality testing across multiple methods
variable_of_interest <- "ejection_fraction"

# Parametric approach (Shapiro-Wilk)
norm_parametric <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable_of_interest,
  typestatistics = "parametric",
  normalcurve = TRUE,
  centralityline = TRUE
)

# Nonparametric approach (Anderson-Darling)
norm_nonparametric <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable_of_interest,
  typestatistics = "nonparametric",
  normalcurve = TRUE,
  centralityline = TRUE
)

# Robust approach
norm_robust <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable_of_interest,
  typestatistics = "robust",
  normalcurve = TRUE,
  centralityline = TRUE
)

# Bayesian approach
norm_bayes <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable_of_interest,
  typestatistics = "bayes",
  normalcurve = TRUE,
  centralityline = TRUE
)
```

### Custom Binning Strategies

```{r binning-strategies}
echo: true
eval: false

# Different binning approaches for the same variable
variable <- "lab_score"

# Automatic binning
auto_bins <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable,
  changebinwidth = FALSE
)

# Fine binning (many narrow bins)
fine_bins <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable,
  changebinwidth = TRUE,
  binwidth = 2
)

# Coarse binning (few wide bins)
coarse_bins <- jjhistostats(
  data = jjhistostats_test_data,
  dep = variable,
  changebinwidth = TRUE,
  binwidth = 10
)
```

## Data Requirements

### Input Data Structure

The input data should be a data frame with:

- **Continuous variables**: Numeric columns for histogram analysis
- **Grouping variables**: Factor or character columns for stratified analysis
- **Complete cases**: Missing values are automatically excluded
- **Reasonable sample size**: At least 10-15 observations per group for meaningful analysis

### Example Data Structure

```{r data-structure}
echo: true
eval: false

# Structure of test data
str(jjhistostats_test_data)

# Key variables:
# - age_years: Normal distribution (20-95 years)
# - tumor_size_mm: Right-skewed distribution (3-150 mm)
# - ejection_fraction: Left-skewed distribution (15-80%)
# - psa_level: Bimodal distribution (0.1-100 ng/mL)
# - bmi: Nearly normal distribution (15-45 kg/m²)
# - disease_stage: Ordered factor (Stage I-IV)
# - treatment_group: Factor (Control, Treatment A, Treatment B)
```

## Best Practices

### Variable Selection and Preparation

1. **Choose appropriate variables**: 
   - Ensure variables are truly continuous
   - Check for reasonable ranges and outliers
   
2. **Sample size considerations**:
   - Minimum 10-15 observations for basic histogram
   - At least 30 observations for reliable normality testing
   - Consider power for detecting departures from normality

3. **Handle missing data**: 
   - Understand missing data patterns
   - Document exclusions due to missing values
   
4. **Outlier assessment**:
   - Identify potential outliers before analysis
   - Consider robust methods if outliers are present

### Statistical Method Selection

- **Use parametric** for well-behaved, moderately-sized samples (n=30-5000)
- **Use nonparametric** for larger samples or when concerned about tail behavior
- **Use robust** when outliers are present or data quality is uncertain
- **Use Bayesian** when you want to quantify evidence for normality

### Binning Strategy Guidelines

```{r binning-guidelines}
echo: true
eval: false

# General binning guidelines:

# Automatic binning (recommended starting point)
# - Good for initial exploration
# - Uses sqrt(n) rule or Sturges' formula

# Manual binning considerations:
# Small datasets (n < 50): 5-10 bins
# Medium datasets (n = 50-200): 10-20 bins  
# Large datasets (n > 200): 20+ bins

# Rule of thumb for bin width:
# binwidth ≈ (max - min) / desired_number_of_bins
```

### Interpretation Guidelines

#### Histogram Shape Assessment

- **Normal**: Bell-shaped, symmetric around mean
- **Right-skewed**: Long tail extending to the right
- **Left-skewed**: Long tail extending to the left
- **Bimodal**: Two distinct peaks
- **Uniform**: Relatively flat across the range

#### Normality Test Interpretation

- **p > 0.05**: Fail to reject normality (data may be normal)
- **p ≤ 0.05**: Reject normality (data significantly non-normal)
- **Visual inspection**: Always combine with histogram visual assessment

## Troubleshooting

### Common Issues

1. **"Data contains no (complete) rows"**
   - Check for missing values in selected variables
   - Ensure at least some complete cases exist
   - Consider imputation strategies

2. **Binning issues**
   - If automatic binning creates too few/many bins, use manual binwidth
   - For integer data, consider binwidth = 1
   - For continuous data, experiment with different bin widths

3. **Performance issues**
   - The optimized version should be much faster
   - For very large datasets (n > 10,000), consider sampling
   - Multiple variables create multiple plots - expect longer processing

### Error Handling

```{r error-handling}
echo: true
eval: false

# Example error handling
tryCatch({
  result <- jjhistostats(
    data = my_data,
    dep = "continuous_var",
    typestatistics = "parametric"
  )
}, error = function(e) {
  message("Error in histogram analysis: ", e$message)
  message("Check your data structure and variable types")
  
  # Diagnostic information
  cat("Data structure:\n")
  str(my_data)
  cat("\nVariable summary:\n")
  summary(my_data$continuous_var)
})
```

## Technical Details

### Underlying Functions

The `jjhistostats` function is built on:

- **ggstatsplot::gghistostats**: For single histograms
- **ggstatsplot::grouped_gghistostats**: For grouped analyses
- **jmvcore**: For data handling and option processing

### Caching Implementation

```{r caching-details}
echo: true
eval: false

# Internal caching structure (conceptual)
# private$.processedData: Cached cleaned data with converted variables
# private$.processedOptions: Cached option processing including binwidth
# 
# Benefits:
# - Eliminates redundant jmvcore::naOmit() calls
# - Avoids repeated variable conversion
# - Shares processed data between plot methods
# - Optimizes multi-variable processing
```

## Clinical Applications

### Distribution Analysis in Research

Histograms are essential for:

- **Data exploration**: Understanding variable distributions before analysis
- **Assumption checking**: Verifying normality for parametric tests
- **Outlier detection**: Identifying unusual values
- **Data quality assessment**: Detecting measurement errors or data entry issues

### Research Scenarios

1. **Clinical Trials**: Assess baseline characteristics and endpoint distributions
2. **Epidemiology**: Examine exposure and outcome variable distributions
3. **Laboratory Medicine**: Establish reference ranges and quality control
4. **Biomarker Studies**: Understand biomarker distributions across populations

### Normality Assessment Workflow

```{r clinical-workflow}
echo: true
eval: false

# Typical clinical research workflow

# Step 1: Initial exploration
initial_assessment <- jjhistostats(
  data = clinical_data,
  dep = c("primary_endpoint", "key_biomarker"),
  normalcurve = TRUE,
  centralityline = TRUE
)

# Step 2: Formal normality testing  
normality_test <- jjhistostats(
  data = clinical_data,
  dep = "primary_endpoint",
  typestatistics = "parametric",
  resultssubtitle = TRUE
)

# Step 3: Stratified analysis
stratified_analysis <- jjhistostats(
  data = clinical_data,
  dep = "primary_endpoint",
  grvar = "treatment_group",
  typestatistics = "parametric"
)

# Step 4: Decision for downstream analysis
# Based on normality results, choose appropriate statistical tests
```

## Conclusion

The optimized `jjhistostats` function provides:

- **High performance**: Significant speed improvements through caching
- **Statistical rigor**: Multiple normality testing approaches
- **Flexibility**: Comprehensive customization options for binning and display
- **Clinical relevance**: Designed for biomedical research applications
- **Usability**: Clear documentation and comprehensive examples

The function is well-suited for clinical research, biomarker analysis, data exploration, and any scenario requiring robust distribution visualization and normality assessment.

## Session Information

```{r session-info}
sessionInfo()
```
