---
title: "Forest Plots from Regression Models with jforestmodel"
subtitle: "Professional Forest Plot Visualization for Linear, Logistic, and Cox Regression Models"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Forest Plots from Regression Models with jforestmodel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(ClinicoPath)
library(ggplot2)
library(dplyr)
```

# Introduction

The `jforestmodel` function provides comprehensive forest plot visualization capabilities for regression models including linear regression, logistic regression, and Cox proportional hazards models. Unlike traditional forest plots that display pre-calculated effect sizes, this function fits regression models directly from raw data and creates publication-ready forest plots of the model coefficients.

Forest plots from regression models are essential tools in medical research for:

- Visualizing the effect of multiple predictors simultaneously
- Comparing the relative importance of different risk factors
- Presenting odds ratios, hazard ratios, or linear coefficients with confidence intervals
- Creating publication-ready figures for medical journals
- Communicating statistical findings to clinical audiences

# Key Features

The `jforestmodel` function supports:

- **Multiple Model Types**: Linear regression (lm), logistic regression (glm), and Cox proportional hazards (coxph)
- **Automatic Model Fitting**: Fits models directly from your data
- **Professional Visualization**: Uses the forestmodel package for high-quality plots
- **Flexible Customization**: Colors, fonts, sizes, and layout options
- **Clinical Interpretation**: Automatic interpretation guidance
- **Export Options**: Multiple formats for publication

# Basic Linear Regression

## Simple Linear Model

Let's start with a basic linear regression using the histopathology dataset:

```{r}
# Load histopathology data
data(histopathology, package = "ClinicoPath")

# Display the first few rows
head(histopathology[c("Age", "Grade", "TStage", "LVI", "PNI")])
```

```{r}
# Create a basic linear regression forest plot
linear_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Age",
  predictor_vars = c("Grade", "TStage", "LVI", "PNI"),
  model_type = "lm",
  plot_title = "Linear Regression: Predictors of Age",
  x_axis_label = "Beta Coefficient (Years)",
  show_summary = TRUE,
  show_interpretation = TRUE
)

# The plot will be displayed in jamovi interface
```

## Understanding Linear Regression Forest Plots

In linear regression forest plots:

- **Beta coefficients** represent the change in the dependent variable for a one-unit increase in each predictor
- **Confidence intervals** that don't cross zero indicate statistically significant associations
- **Point estimates** show the direction and magnitude of the effect
- **Reference line** at zero represents no effect

# Logistic Regression

## Binary Outcome Models

Logistic regression is commonly used for binary outcomes. Let's predict lymphovascular invasion (LVI):

```{r}
# Logistic regression with odds ratios
logistic_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  plot_title = "Logistic Regression: Predictors of Lymphovascular Invasion",
  x_axis_label = "Odds Ratio (95% CI)",
  show_summary = TRUE,
  confidence_level = 0.95
)
```

## Working with the Wisconsin Breast Cancer Dataset

Let's use the classic Wisconsin Breast Cancer dataset for cancer diagnosis prediction:

```{r}
# Load breast cancer data
data(BreastCancer, package = "ClinicoPath")

# Display dataset structure
str(BreastCancer[c("Class", "Cl.thickness", "Cell.size", "Cell.shape", "Marg.adhesion")])
```

```{r}
# Comprehensive breast cancer prediction model
breast_cancer_model <- jforestmodel(
  data = BreastCancer,
  dependent_var = "Class",
  predictor_vars = c("Cl.thickness", "Cell.size", "Cell.shape", 
                    "Marg.adhesion", "Epith.c.size", "Bare.nuclei"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  plot_title = "Breast Cancer Diagnosis Prediction Model",
  x_axis_label = "Odds Ratio (95% CI)",
  color_scheme = "blue",
  factor_separate_line = TRUE,
  show_p_values = TRUE,
  show_confidence_intervals = TRUE
)
```

## Different GLM Families

The function supports various GLM families:

```{r}
# Poisson regression example
poisson_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Age",  # Treating age as count for demonstration
  predictor_vars = c("Grade", "TStage", "LVI"),
  model_type = "glm",
  family = "poisson",
  exponentiate = TRUE,
  plot_title = "Poisson Regression Model",
  x_axis_label = "Rate Ratio"
)

# Gaussian GLM (equivalent to linear regression)
gaussian_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Age",
  predictor_vars = c("Grade", "TStage", "LVI"),
  model_type = "glm",
  family = "gaussian",
  exponentiate = FALSE,
  plot_title = "Gaussian GLM Model",
  x_axis_label = "Coefficient"
)
```

# Cox Proportional Hazards Models

## Survival Analysis with Colon Cancer Data

Cox regression is used for time-to-event analysis. Let's use the colon cancer dataset:

```{r}
# Load colon cancer survival data
data(colon, package = "ClinicoPath")

# Display relevant variables
head(colon[c("time", "status", "age", "sex", "obstruct", "perfor", "adhere")])
```

```{r}
# Cox proportional hazards model
cox_model <- jforestmodel(
  data = colon,
  dependent_var = "status",  # Required by interface but not used directly
  predictor_vars = c("age", "sex", "obstruct", "perfor", "adhere"),
  model_type = "coxph",
  time_var = "time",
  event_var = "status",
  exponentiate = TRUE,
  plot_title = "Cox Regression: Colon Cancer Survival",
  x_axis_label = "Hazard Ratio (95% CI)",
  confidence_level = 0.95,
  show_summary = TRUE
)
```

## Melanoma Survival Analysis

Let's also examine melanoma survival data:

```{r}
# Load melanoma data
data(melanoma, package = "ClinicoPath")

# Display melanoma variables
head(melanoma[c("time", "status", "age", "sex", "thickness", "ulcer")])
```

```{r}
# Melanoma survival model
melanoma_model <- jforestmodel(
  data = melanoma,
  dependent_var = "status",
  predictor_vars = c("age", "sex", "thickness", "ulcer"),
  model_type = "coxph",
  time_var = "time",
  event_var = "status",
  exponentiate = TRUE,
  plot_title = "Melanoma Survival Analysis",
  x_axis_label = "Hazard Ratio (95% CI)",
  color_scheme = "red",
  show_interpretation = TRUE
)
```

# Advanced Customization

## Color Schemes and Visual Options

The function provides multiple color schemes and visual customization:

```{r}
# Custom color scheme
custom_color_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  color_scheme = "custom",
  custom_color = "#FF5722",
  point_size = 3.0,
  line_size = 1.2,
  plot_title = "Custom Styled Forest Plot"
)

# Professional medical color scheme
medical_style_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  color_scheme = "green",
  plot_title = "Medical Style Forest Plot",
  show_reference_line = TRUE,
  reference_value = 1
)
```

## Confidence Levels and Statistical Options

```{r}
# 99% confidence intervals
high_confidence_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  confidence_level = 0.99,
  plot_title = "High Confidence Model (99% CI)"
)

# 90% confidence intervals
low_confidence_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  confidence_level = 0.90,
  plot_title = "Conservative Model (90% CI)"
)
```

## Factor Variable Handling

```{r}
# Factor variables on separate lines (default)
factor_separate_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Death",
  predictor_vars = c("Age", "Grade", "TStage", "LVI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  factor_separate_line = TRUE,
  plot_title = "Factor Levels on Separate Lines"
)

# Factor variables on same line
factor_together_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Death",
  predictor_vars = c("Age", "Grade", "TStage", "LVI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  factor_separate_line = FALSE,
  plot_title = "Factor Levels Together"
)
```

# Variable Sorting and Selection

## Sorting Options

The function provides several sorting options for variables:

```{r}
# Sort by coefficient size
coefficient_sorted_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  sort_variables = "coefficient",
  plot_title = "Sorted by Coefficient Size"
)

# Sort by p-value
pvalue_sorted_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  sort_variables = "pvalue",
  plot_title = "Sorted by P-value"
)

# Alphabetical sorting
alphabetical_sorted_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  sort_variables = "alphabetical",
  plot_title = "Alphabetically Sorted"
)
```

## Covariate Selection

You can specify which covariates to display in the plot:

```{r}
# Display only specific covariates
selected_covariates_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  covariates = c("Age", "Grade"),  # Only show these in plot
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  plot_title = "Selected Covariates Only"
)
```

# Model Diagnostics and Interpretation

## Model Summary Statistics

```{r}
# Comprehensive model with full reporting
comprehensive_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  show_summary = TRUE,
  show_interpretation = TRUE,
  plot_title = "Comprehensive Model Analysis"
)
```

## Linear Model Diagnostics

For linear models, additional diagnostic information is provided:

```{r}
# Linear model with diagnostics
linear_diagnostics_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Age",
  predictor_vars = c("Grade", "TStage", "LVI", "PNI"),
  model_type = "lm",
  show_summary = TRUE,
  plot_title = "Linear Model with Diagnostics"
)
```

# Publication-Ready Examples

## High-Quality Clinical Example

```{r}
# Publication-ready breast cancer model
publication_model <- jforestmodel(
  data = BreastCancer,
  dependent_var = "Class",
  predictor_vars = c("Cl.thickness", "Cell.size", "Cell.shape", 
                    "Marg.adhesion", "Epith.c.size"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  confidence_level = 0.95,
  plot_title = "Predictors of Breast Cancer Malignancy",
  x_axis_label = "Odds Ratio (95% Confidence Interval)",
  color_scheme = "blue",
  point_size = 2.5,
  line_size = 0.8,
  factor_separate_line = TRUE,
  show_p_values = TRUE,
  show_confidence_intervals = TRUE,
  show_reference_line = TRUE,
  reference_value = 1,
  show_summary = TRUE,
  show_interpretation = TRUE
)
```

## Multi-Model Comparison

You can create multiple models for comparison:

```{r}
# Model 1: Basic predictors
basic_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Death",
  predictor_vars = c("Age", "Grade"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  plot_title = "Basic Model: Age and Grade Only"
)

# Model 2: Extended predictors
extended_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Death",
  predictor_vars = c("Age", "Grade", "TStage", "LVI", "PNI"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  plot_title = "Extended Model: Multiple Predictors"
)
```

# Layout and Panel Customization

## Panel Width Ratios

```{r}
# Custom panel width ratios
custom_layout_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  panel_width_ratio = "2:3:1",
  plot_title = "Custom Panel Layout"
)
```

## Reference Lines and Values

```{r}
# Custom reference line for linear model
linear_reference_model <- jforestmodel(
  data = histopathology,
  dependent_var = "Age",
  predictor_vars = c("Grade", "TStage", "LVI"),
  model_type = "lm",
  show_reference_line = TRUE,
  reference_value = 0,
  plot_title = "Linear Model with Zero Reference"
)

# Custom reference line for logistic model
logistic_reference_model <- jforestmodel(
  data = histopathology,
  dependent_var = "LVI",
  predictor_vars = c("Age", "Grade", "TStage"),
  model_type = "glm",
  family = "binomial",
  exponentiate = TRUE,
  show_reference_line = TRUE,
  reference_value = 1,
  plot_title = "Logistic Model with OR=1 Reference"
)
```

# Interpretation Guidelines

## Understanding Different Effect Measures

### Linear Regression Coefficients
- **Interpretation**: Change in dependent variable per unit increase in predictor
- **Significance**: Confidence intervals not including zero
- **Clinical relevance**: Consider magnitude along with statistical significance

### Odds Ratios (Logistic Regression)
- **OR = 1**: No association
- **OR > 1**: Increased odds of outcome
- **OR < 1**: Decreased odds of outcome
- **95% CI**: Must not include 1 for statistical significance

### Hazard Ratios (Cox Regression)
- **HR = 1**: No effect on hazard
- **HR > 1**: Increased hazard (worse survival)
- **HR < 1**: Decreased hazard (better survival)
- **Interpretation**: Instantaneous risk at any given time

## Model Assessment Guidelines

### Model Fit Assessment
1. **R-squared** (linear): Proportion of variance explained
2. **AIC** (all models): Lower values indicate better fit
3. **Residual analysis** (linear/GLM): Check model assumptions
4. **Proportional hazards assumption** (Cox): Verify with diagnostics

### Clinical Interpretation
1. **Statistical vs Clinical Significance**: Large effects may not be clinically meaningful
2. **Confidence Intervals**: Wide intervals suggest uncertainty
3. **Multiple Comparisons**: Consider adjustment for multiple testing
4. **Model Limitations**: Remember unmeasured confounders

# Best Practices

## Model Building
1. **Variable Selection**: Use clinical knowledge and statistical criteria
2. **Sample Size**: Ensure adequate power for the number of predictors
3. **Missing Data**: Handle appropriately before modeling
4. **Assumption Checking**: Verify model assumptions

## Visualization
1. **Appropriate Scale**: Use exponentiation for ratios, raw coefficients for linear
2. **Clear Labeling**: Descriptive titles and axis labels
3. **Consistent Formatting**: Use standard conventions
4. **Color Choices**: Consider accessibility and journal requirements

## Reporting
1. **Model Details**: Report model type, sample size, and selection criteria
2. **Assumption Testing**: Document assumption verification
3. **Effect Sizes**: Include both point estimates and confidence intervals
4. **Clinical Context**: Interpret findings in clinical context

# Troubleshooting Common Issues

## Data Preparation
```r
# Ensure variables are properly formatted
data$outcome <- as.factor(data$outcome)  # For binary outcomes
data$time <- as.numeric(data$time)       # For survival time
data$event <- as.numeric(data$event)     # For event indicator
```

## Model Convergence Issues
```r
# For convergence problems, try:
# 1. Check for perfect separation in logistic regression
# 2. Scale continuous variables
# 3. Check for multicollinearity
# 4. Reduce number of predictors
```

## Missing Data Handling
```r
# Remove incomplete cases or use multiple imputation
complete_data <- data[complete.cases(data[c("outcome", "pred1", "pred2")]), ]
```

# Advanced Features

## Custom Model Objects

The function also supports custom model objects (future enhancement):

```{r, eval=FALSE}
# Fit your own model first
my_model <- glm(LVI ~ Age + Grade + TStage, 
               data = histopathology, 
               family = binomial())

# Then create forest plot
custom_model_plot <- jforestmodel(
  model_object = my_model,  # Future feature
  model_type = "custom",
  exponentiate = TRUE,
  plot_title = "Custom Model Forest Plot"
)
```

## Integration with Other Packages

The forest plots can be integrated with other R packages for enhanced analysis:

- **broom**: For model tidying and extraction
- **performance**: For model performance metrics
- **see**: For additional visualization options
- **report**: For automated interpretation

# Conclusion

The `jforestmodel` function provides a comprehensive solution for creating professional forest plots directly from regression models. Key benefits include:

- **Direct Model Fitting**: No need to pre-calculate effect sizes
- **Multiple Model Types**: Support for linear, logistic, and Cox regression
- **Professional Quality**: Publication-ready outputs
- **Extensive Customization**: Colors, layouts, and formatting options
- **Clinical Integration**: Appropriate for medical research applications

Forest plots from regression models are powerful tools for communicating statistical findings in medical research. The `jforestmodel` function makes it easy to create high-quality visualizations that meet publication standards and effectively communicate your research findings.

For advanced statistical modeling considerations, consult with a biostatistician and refer to specialized statistical texts and guidelines from your research domain.

# References

1. Lewis S, Clarke M. Forest plots: trying to see the wood and the trees. BMJ. 2001;322(7300):1479-80.
2. Wickham H. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.
3. Therneau T. A Package for Survival Analysis in R. 2023.
4. Harrell FE. Regression Modeling Strategies. Springer, 2015.
5. Steyerberg EW. Clinical Prediction Models. Springer, 2019.