---
title: "Advanced Bar Charts - 7 Professional Approaches"
description: "Comprehensive guide to creating professional bar charts with multiple visualization approaches, BBC styling, and GraphPad Prism integration."
author: "ClinicoPath Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Advanced Bar Charts - 7 Professional Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  dpi = 300
)

library(ClinicoPath)
library(ggplot2)
library(dplyr)
```

## Overview

The `advancedbarplot` module provides **7 distinct professional approaches** to bar chart visualization, each optimized for different use cases in clinical research, data presentation, and publication. This comprehensive guide demonstrates all approaches with the histopathology dataset.

### Key Features

- **7 Chart Approaches**: From basic exploration to publication-ready designs
- **Professional Styling**: BBC News and GraphPad Prism authentic themes
- **Statistical Integration**: Built-in statistical tests and annotations
- **26+ Color Palettes**: Including BBC colors and Prism palettes
- **Interactive Options**: Plotly integration for web-based exploration
- **Export Optimization**: Publication and presentation-ready outputs

## Dataset Overview

We'll use the comprehensive `histopathology` dataset throughout this guide:

```{r data-overview}
# Load the histopathology dataset
data("histopathology")

# Dataset structure
str(histopathology)

# Key variables for bar charts
cat("Categorical variables suitable for X-axis:\n")
cat("- Sex:", unique(histopathology$Sex), "\n")
cat("- Race:", unique(histopathology$Race), "\n") 
cat("- Group:", unique(histopathology$Group), "\n")
cat("- Grade_Level:", unique(histopathology$Grade_Level), "\n")

cat("\nNumeric variables suitable for Y-axis:\n")
cat("- Age: range", range(histopathology$Age, na.rm = TRUE), "\n")
cat("- Grade: range", range(histopathology$Grade, na.rm = TRUE), "\n")
cat("- OverallTime: range", range(histopathology$OverallTime, na.rm = TRUE), "\n")
```

## Approach 1: Basic ggplot2

**Use Case**: Data exploration, quick analysis, educational purposes

**Strengths**: Fast rendering, straightforward implementation, highly customizable

```{r approach-basic, eval=FALSE}
# Basic ggplot2 approach
advancedbarplot(
  data = histopathology,
  x_var = "Group",
  y_var = "Age", 
  chart_approach = "basic",
  stat_type = "mean",
  show_values = TRUE,
  plot_title = "Mean Age by Treatment Group",
  x_title = "Treatment Group",
  y_title = "Mean Age (years)"
)
```

### R Code Example for Basic Approach

```{r basic-code, eval=TRUE}
# Reproducible R code for basic approach
library(ggplot2)
library(dplyr)

# Summarize data
summary_data <- histopathology %>%
  group_by(Group) %>%
  summarise(
    mean_age = mean(Age, na.rm = TRUE),
    se = sd(Age, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Create basic plot
basic_plot <- ggplot(summary_data, aes(x = Group, y = mean_age)) +
  geom_col(width = 0.8, alpha = 0.9) +
  geom_text(aes(label = round(mean_age, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(
    title = "Mean Age by Treatment Group - Basic Approach",
    x = "Treatment Group",
    y = "Mean Age (years)"
  )

print(basic_plot)
```

## Approach 2: Polished Presentation

**Use Case**: Business presentations, stakeholder reports, dashboard displays

**Strengths**: Professional appearance, enhanced readability, presentation-ready

```{r approach-polished, eval=FALSE}
# Polished presentation approach
advancedbarplot(
  data = histopathology,
  x_var = "Grade_Level", 
  y_var = "OverallTime",
  fill_var = "Sex",
  chart_approach = "polished",
  stat_type = "mean",
  color_palette = "clinical",
  bar_position = "dodge",
  show_values = TRUE,
  value_format = "decimal1",
  plot_title = "Overall Survival Time by Grade and Sex",
  legend_position = "top"
)
```

### R Code Example for Polished Approach

```{r polished-code, eval=TRUE}
# Polished presentation approach
summary_data_grouped <- histopathology %>%
  group_by(Grade_Level, Sex) %>%
  summarise(
    mean_time = mean(OverallTime, na.rm = TRUE),
    se = sd(OverallTime, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Clinical color palette
clinical_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728")

polished_plot <- ggplot(summary_data_grouped, aes(x = Grade_Level, y = mean_time, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.8, alpha = 0.9) +
  geom_text(
    aes(label = sprintf("%.1f", mean_time)),
    position = position_dodge(width = 0.8),
    vjust = -0.5,
    size = 3.5
  ) +
  scale_fill_manual(values = clinical_colors) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "top"
  ) +
  labs(
    title = "Overall Survival Time by Grade and Sex - Polished Approach",
    x = "Grade Level",
    y = "Mean Overall Time (months)",
    fill = "Sex"
  )

print(polished_plot)
```

## Approach 3: Statistical Annotations

**Use Case**: Research publications, clinical studies, comparative analysis

**Strengths**: Built-in statistics, automated annotations, hypothesis testing

```{r approach-statistical, eval=FALSE}
# Statistical annotations approach
advancedbarplot(
  data = histopathology,
  x_var = "Group",
  y_var = "Grade",
  chart_approach = "statistical", 
  stat_type = "mean",
  add_statistics = TRUE,
  stat_method = "ttest",
  error_bars = "se",
  show_values = TRUE,
  plot_title = "Mean Grade by Treatment Group with Statistical Test",
  color_palette = "nature"
)
```

### R Code Example for Statistical Approach

```{r statistical-code, eval=TRUE}
# Statistical approach with error bars
summary_stats <- histopathology %>%
  group_by(Group) %>%
  summarise(
    mean_grade = mean(Grade, na.rm = TRUE),
    se = sd(Grade, na.rm = TRUE) / sqrt(n()),
    sd = sd(Grade, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Perform t-test
control_data <- histopathology$Grade[histopathology$Group == "Control"]
treatment_data <- histopathology$Grade[histopathology$Group == "Treatment"]
t_result <- t.test(control_data, treatment_data)

statistical_plot <- ggplot(summary_stats, aes(x = Group, y = mean_grade)) +
  geom_col(width = 0.8, alpha = 0.9, fill = "steelblue") +
  geom_errorbar(
    aes(ymin = mean_grade - se, ymax = mean_grade + se),
    width = 0.2,
    position = position_dodge(width = 0.8)
  ) +
  geom_text(aes(label = sprintf("%.2f", mean_grade)), vjust = -1.5) +
  theme_minimal() +
  labs(
    title = paste0("Mean Grade by Treatment Group - Statistical Approach\n",
                   "t(", round(t_result$parameter, 1), ") = ", round(t_result$statistic, 3),
                   ", p = ", format.pval(t_result$p.value, digits = 3)),
    x = "Treatment Group", 
    y = "Mean Grade",
    caption = paste0("Error bars represent standard error\n",
                     "Mean difference: ", round(diff(t_result$estimate), 3))
  )

print(statistical_plot)
```

## Approach 4: Interactive Plotly

**Use Case**: Web applications, interactive reports, data exploration tools

**Strengths**: Interactive exploration, hover information, zoom and pan capabilities

```{r approach-interactive, eval=FALSE}
# Interactive plotly approach
advancedbarplot(
  data = histopathology,
  x_var = "Race",
  y_var = "MeasurementA", 
  chart_approach = "interactive",
  stat_type = "median",
  color_palette = "viridis",
  plot_title = "Median Measurement A by Race - Interactive",
  show_values = TRUE
)
```

### R Code Example for Interactive Approach

```{r interactive-code, eval=TRUE}
library(plotly)

# Prepare data for interactive plot
race_summary <- histopathology %>%
  group_by(Race) %>%
  summarise(
    median_measurement = median(MeasurementA, na.rm = TRUE),
    mad = mad(MeasurementA, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Create ggplot
interactive_base <- ggplot(race_summary, aes(x = Race, y = median_measurement, fill = Race)) +
  geom_col(width = 0.8, alpha = 0.9) +
  geom_text(aes(label = sprintf("%.3f", median_measurement)), vjust = -0.5) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Median Measurement A by Race - Interactive Approach",
    x = "Race",
    y = "Median Measurement A"
  )

# Convert to interactive plot
interactive_plot <- ggplotly(interactive_base, tooltip = c("x", "y"))

print(interactive_plot)
```

## Approach 5: Publication Ready

**Use Case**: Academic papers, journal submissions, conference presentations

**Strengths**: Journal formatting standards, high-resolution export, citation ready

```{r approach-publication, eval=FALSE}
# Publication ready approach
advancedbarplot(
  data = histopathology,
  x_var = "Sex",
  y_var = "MeasurementB",
  chart_approach = "publication",
  stat_type = "mean",
  error_bars = "ci95", 
  show_values = TRUE,
  value_format = "decimal2",
  plot_title = "Mean Measurement B by Sex",
  theme_style = "publication",
  export_options = TRUE
)
```

### R Code Example for Publication Approach

```{r publication-code, eval=TRUE}
# Publication-ready approach
sex_summary <- histopathology %>%
  group_by(Sex) %>%
  summarise(
    mean_measurement = mean(MeasurementB, na.rm = TRUE),
    se = sd(MeasurementB, na.rm = TRUE) / sqrt(n()),
    ci95 = 1.96 * se,
    n = n(),
    .groups = "drop"
  )

publication_plot <- ggplot(sex_summary, aes(x = Sex, y = mean_measurement)) +
  geom_col(
    width = 0.8,
    alpha = 0.7,
    color = "black",
    size = 0.3,
    fill = "steelblue"
  ) +
  geom_errorbar(
    aes(ymin = mean_measurement - ci95, ymax = mean_measurement + ci95),
    width = 0.2
  ) +
  geom_text(
    aes(label = sprintf("%.2f", mean_measurement)),
    vjust = -1.5,
    size = 3
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 11, family = "serif"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 11, face = "bold"),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.5)
  ) +
  labs(
    title = "Mean Measurement B by Sex",
    x = "Sex",
    y = "Mean Measurement B",
    caption = "Error bars represent 95% confidence intervals"
  )

print(publication_plot)
```

## Approach 6: BBC News Style

**Use Case**: News reports, journalism, public communications, professional presentations

**Strengths**: BBC design standards, professional typography, brand consistency

```{r approach-bbc, eval=FALSE}
# BBC News style approach
advancedbarplot(
  data = histopathology,
  x_var = "Grade_Level",
  y_var = "Age",
  chart_approach = "bbc_style",
  stat_type = "mean",
  color_palette = "bbc_multi",
  show_values = TRUE,
  plot_title = "Age Distribution by Grade Level",
  subtitle_text = "BBC Visual and Data Journalism Style",
  source_text = "ClinicoPath Histopathology Dataset"
)
```

### R Code Example for BBC Style

```{r bbc-code, eval=TRUE}
# BBC News style approach
grade_age_summary <- histopathology %>%
  group_by(Grade_Level) %>%
  summarise(
    mean_age = mean(Age, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# BBC color palette
bbc_colors <- c("#1380A1", "#FAAB18", "#007f7f", "#333333")

bbc_plot <- ggplot(grade_age_summary, aes(x = Grade_Level, y = mean_age, fill = Grade_Level)) +
  geom_col(width = 0.8, alpha = 1.0) +
  scale_fill_manual(values = bbc_colors) +
  theme(
    # Text elements
    text = element_text(family = "Arial", size = 18, color = "#222222"),
    plot.title = element_text(family = "Arial", size = 28, face = "bold", color = "#222222"),
    plot.subtitle = element_text(family = "Arial", size = 22, margin = margin(9, 0, 9, 0)),
    plot.caption = element_text(size = 14, color = "#666666"),
    
    # Legend
    legend.position = "none",
    
    # Axis
    axis.title = element_blank(),
    axis.text = element_text(family = "Arial", size = 18, color = "#222222"),
    axis.text.x = element_text(margin = margin(5, b = 10)),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    
    # Grid lines (BBC style: horizontal only)
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "#cbcbcb"),
    panel.grid.major.x = element_blank(),
    
    # Background
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title = "Age Distribution by Grade Level",
    subtitle = "BBC Visual and Data Journalism Style",
    caption = "Source: ClinicoPath Histopathology Dataset"
  )

print(bbc_plot)
```

## Approach 7: GraphPad Prism Style

**Use Case**: Scientific software integration, GraphPad users, statistical software consistency

**Strengths**: Authentic Prism styling, scientific software aesthetics, statistical focus

```{r approach-prism, eval=FALSE}
# GraphPad Prism style approach  
advancedbarplot(
  data = histopathology,
  x_var = "Group",
  y_var = "OverallTime",
  fill_var = "Grade_Level",
  chart_approach = "prism_style",
  stat_type = "mean",
  color_palette = "prism_floral",
  bar_position = "dodge",
  theme_style = "prism_default",
  show_values = TRUE,
  plot_title = "Overall Survival Time by Treatment and Grade",
  subtitle_text = "GraphPad Prism Style Visualization"
)
```

### R Code Example for Prism Style

```{r prism-code, eval=TRUE}
# GraphPad Prism style approach
prism_summary <- histopathology %>%
  group_by(Group, Grade_Level) %>%
  summarise(
    mean_time = mean(OverallTime, na.rm = TRUE),
    se = sd(OverallTime, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Prism-style colors (approximation)
prism_colors <- c("#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4")

prism_plot <- ggplot(prism_summary, aes(x = Group, y = mean_time, fill = Grade_Level)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.8,
    alpha = 0.9
  ) +
  scale_fill_manual(values = prism_colors) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "black", fill = NA),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "top"
  ) +
  labs(
    title = "Overall Survival Time by Treatment and Grade",
    subtitle = "GraphPad Prism Style Visualization", 
    x = "Treatment Group",
    y = "Mean Overall Time (months)",
    fill = "Grade Level"
  )

print(prism_plot)
```

## Statistical Testing Integration

All approaches support integrated statistical testing:

```{r statistical-tests}
# Example statistical comparisons
cat("Available statistical tests in advancedbarplot:\n")
cat("- ANOVA: For comparing multiple groups\n")
cat("- t-test: For comparing two groups\n") 
cat("- Wilcoxon: Non-parametric alternative to t-test\n")
cat("- Chi-square: For categorical data associations\n")
cat("- Kruskal-Wallis: Non-parametric alternative to ANOVA\n")

# Example ANOVA test
anova_result <- aov(Age ~ Group, data = histopathology)
cat("\nExample ANOVA result (Age by Group):\n")
summary(anova_result)
```

## Color Palette Showcase

The module includes 26+ professional color palettes:

```{r color-palettes, eval=FALSE}
# BBC Color Palettes
- bbc_blue: "#1380A1" 
- bbc_orange: "#FAAB18"
- bbc_multi: Multiple BBC brand colors

# GraphPad Prism Palettes  
- prism_floral: 12 colors
- prism_candy_bright: 9 colors
- prism_office: 9 colors
- prism_colorblind_safe: 6 colors

# Scientific Journal Palettes
- nature: Nature journal colors
- science: Science journal colors
- clinical: Clinical research optimized

# Standard Palettes
- viridis: Colorblind-friendly
- set1, dark2: RColorBrewer palettes
```

## Export and Reproducibility

### High-Quality Export Options

```{r export-tips, eval=FALSE}
# For publication export
advancedbarplot(
  # ... your parameters ...
  plot_width = 10,      # Width in inches
  plot_height = 6,      # Height in inches
  export_options = TRUE # Optimize for export
)
```

### Reproducible R Code

Each approach generates reproducible R code that you can copy and modify:

```{r reproducible-example, eval=FALSE}
# Example generated code
ggplot(data, aes(x = Group, y = Age)) +
  geom_col(alpha = 0.8) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(legend.position = "none")
```

## Best Practices

### 1. Choosing the Right Approach

- **Basic**: Quick exploration, learning ggplot2
- **Polished**: Business presentations, dashboards  
- **Statistical**: Research papers, hypothesis testing
- **Interactive**: Web apps, exploration tools
- **Publication**: Journal submissions, formal reports
- **BBC Style**: News graphics, public communication
- **Prism Style**: Scientific software integration

### 2. Data Preparation

```{r data-prep-tips}
# Always check your data first
summary(histopathology[c("Group", "Age", "Grade_Level")])

# Handle missing values
cat("Missing values check:\n")
sapply(histopathology[c("Group", "Age", "Grade_Level")], function(x) sum(is.na(x)))

# Ensure proper factor levels
histopathology$Group <- factor(histopathology$Group)
histopathology$Grade_Level <- factor(histopathology$Grade_Level, levels = c("low", "high"))
```

### 3. Statistical Considerations

```{r statistical-considerations}
# Check assumptions before statistical tests
cat("Sample sizes by group:\n")
table(histopathology$Group)

cat("\nNormality check (Shapiro-Wilk test p-values):\n")
by(histopathology$Age, histopathology$Group, function(x) shapiro.test(x)$p.value)

cat("\nEqual variances test (F-test p-value):\n")
var.test(Age ~ Group, data = histopathology)$p.value
```

## Conclusion

The `advancedbarplot` module provides a comprehensive toolkit for professional bar chart visualization. Whether you need quick exploration plots or publication-ready graphics, the 7 approaches cover all major use cases in clinical research and data presentation.

### Key Takeaways

1. **7 distinct approaches** for different use cases
2. **Professional styling** with BBC and Prism authenticity
3. **Statistical integration** with automated testing
4. **Extensive customization** with 26+ color palettes
5. **Export optimization** for publication and presentation

### Next Steps

- Explore the interactive features in jamovi
- Try different color palettes for your specific use case
- Experiment with statistical testing options
- Use the generated R code for further customization

For more advanced statistical plotting, see the related modules:
- `jjhistogram` for distribution analysis
- `jjscatter` for correlation visualization  
- `survival` for time-to-event analysis
