---
title: "Comprehensive Bar Chart Analysis with jjbarstats"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>", 
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)

# Load required packages
library(ClinicoPath)
library(ggplot2)
library(dplyr)

# Load datasets
data("medical_study_data")
data("patient_satisfaction_data") 
data("clinical_trial_data")
data("diagnostic_test_data")
data("quality_improvement_data")
data("histopathology")
```

# Introduction to jjbarstats

The `jjbarstats` function is a powerful wrapper around the `ggstatsplot` package that creates publication-ready bar charts with automatic statistical testing. This function is designed specifically for analyzing categorical data relationships in clinical and research settings.

## Key Features

- **Automatic Statistical Testing**: Performs chi-squared tests, Fisher's exact tests, or other appropriate tests based on your data
- **Multiple Variable Support**: Handle single or multiple dependent variables simultaneously
- **Grouped Analysis**: Split analysis by additional grouping variables
- **Flexible Statistical Methods**: Choose from parametric, non-parametric, robust, or Bayesian approaches
- **Pairwise Comparisons**: Automatic post-hoc testing with multiple comparison correction
- **Professional Visualization**: Publication-ready plots with statistical annotations

## When to Use jjbarstats

Use `jjbarstats` when you need to:

- Compare proportions across different groups
- Analyze treatment effectiveness in clinical trials
- Examine relationships between categorical variables
- Create publication-ready visualizations with statistical tests
- Perform quality improvement analysis
- Analyze survey responses and patient feedback

# Basic Usage

## Single Dependent Variable Analysis

Let's start with a basic example using medical study data to compare treatment responses across different treatment groups:

```{r basic-example}
# Basic bar chart comparing treatment response across groups
jjbarstats(
  data = medical_study_data,
  dep = response,
  group = treatment_group,
  grvar = NULL
)
```

This creates a bar chart showing the distribution of treatment responses (Complete Response, Partial Response, No Response) across different treatment groups (Control, Treatment A, Treatment B), along with chi-squared test results.

## Multiple Dependent Variables

You can analyze multiple dependent variables simultaneously:

```{r multiple-deps}
# Analyze multiple outcomes in clinical trial
# jjbarstats(
#   data = clinical_trial_data,
#   dep = c(primary_outcome, side_effects),
#   group = drug_dosage,
#   grvar = NULL
# )
```

# Advanced Statistical Options

## Different Statistical Methods

The function supports four different statistical approaches:

### Parametric Analysis (Default)
```{r parametric}
jjbarstats(
  data = patient_satisfaction_data,
  dep = satisfaction_level,
  group = service_type,
  typestatistics = "parametric"
)
```

### Non-parametric Analysis
```{r nonparametric}
jjbarstats(
  data = diagnostic_test_data,
  dep = test_result,
  group = test_method,
  typestatistics = "nonparametric"
)
```

### Robust Analysis
```{r robust}
jjbarstats(
  data = quality_improvement_data,
  dep = implementation_status,
  group = improvement_category,
  typestatistics = "robust"
)
```

### Bayesian Analysis
```{r bayesian}
jjbarstats(
  data = medical_study_data,
  dep = severity,
  group = treatment_group,
  typestatistics = "bayes"
)
```

# Grouped Analysis with Splitting Variables

## Using the grvar Parameter

The `grvar` parameter allows you to split your analysis by an additional grouping variable, creating separate plots for each level:

```{r grouped-analysis}
# Analyze treatment response by treatment group, split by gender
jjbarstats(
  data = medical_study_data,
  dep = response,
  group = treatment_group,
  grvar = gender
)
```

This creates separate bar charts for male and female patients, allowing you to examine whether treatment effects differ by gender.

## Complex Grouped Analysis

```{r complex-grouped}
# Patient satisfaction by service type, split by department
jjbarstats(
  data = patient_satisfaction_data,
  dep = satisfaction_level,
  group = service_type,
  grvar = department
)
```

# Pairwise Comparisons and Multiple Testing

## Enabling Pairwise Comparisons

When you have more than two groups, pairwise comparisons help identify which specific groups differ:

```{r pairwise-comparisons}
# jjbarstats(
#   data = clinical_trial_data,
#   dep = primary_outcome,
#   group = drug_dosage,
#   pairwisecomparisons = TRUE,
#   padjustmethod = "holm"
# )
```

## Multiple Comparison Correction Methods

Different correction methods are available to control for multiple testing:

```{r correction-methods}
# Bonferroni correction (most conservative)
jjbarstats(
  data = diagnostic_test_data,
  dep = test_result,
  group = laboratory,
  pairwisecomparisons = TRUE,
  padjustmethod = "bonferroni"
)
```

```{r benjamini-hochberg}
# Benjamini-Hochberg correction (controls false discovery rate)
jjbarstats(
  data = quality_improvement_data,
  dep = priority_level,
  group = department_involved,
  pairwisecomparisons = TRUE,
  padjustmethod = "BH"
)
```

## Controlling Pairwise Display

You can control which pairwise comparisons are displayed:

```{r pairwise-display}
# Show only significant comparisons
jjbarstats(
  data = medical_study_data,
  dep = response,
  group = treatment_group,
  pairwisecomparisons = TRUE,
  pairwisedisplay = "significant"
)
```

```{r pairwise-all}
# Show all comparisons
jjbarstats(
  data = patient_satisfaction_data,
  dep = staff_rating,
  group = service_type,
  pairwisecomparisons = TRUE,
  pairwisedisplay = "everything"
)
```

# Real-World Clinical Applications

## Treatment Efficacy Analysis

Analyzing treatment effectiveness across different patient subgroups:

```{r treatment-efficacy}
# Comprehensive treatment analysis
jjbarstats(
  data = medical_study_data,
  dep = response,
  group = treatment_group,
  grvar = severity,
  typestatistics = "nonparametric",
  pairwisecomparisons = TRUE,
  padjustmethod = "BH"
)
```

## Quality Improvement Analysis

Tracking implementation status across different improvement categories:

```{r quality-improvement}
jjbarstats(
  data = quality_improvement_data,
  dep = c(implementation_status, priority_level),
  group = improvement_category,
  typestatistics = "parametric",
  pairwisecomparisons = TRUE
)
```

## Diagnostic Test Evaluation

Comparing test performance across different methods and laboratories:

```{r diagnostic-evaluation}
jjbarstats(
  data = diagnostic_test_data,
  dep = test_result,
  group = test_method,
  grvar = laboratory,
  typestatistics = "robust",
  pairwisecomparisons = TRUE,
  pairwisedisplay = "significant"
)
```

## Patient Satisfaction Survey Analysis

Analyzing satisfaction levels across different service types and departments:

```{r satisfaction-analysis}
jjbarstats(
  data = patient_satisfaction_data,
  dep = satisfaction_level,
  group = service_type,
  grvar = department,
  typestatistics = "nonparametric",
  pairwisecomparisons = TRUE,
  padjustmethod = "holm"
)
```

# Working with Real Histopathology Data

Using the histopathology dataset that comes with ClinicoPath:

```{r histopathology-analysis}
# Analyze lymphovascular invasion by treatment group
jjbarstats(
  data = histopathology,
  dep = LVI,
  group = Group,
  typestatistics = "nonparametric",
  pairwisecomparisons = TRUE
)
```

```{r histopathology-multiple}
# Multiple outcome analysis
jjbarstats(
  data = histopathology,
  dep = c(LVI, PNI),
  group = Grade_Level,
  typestatistics = "parametric"
)
```

```{r histopathology-grouped}
# Grouped analysis by sex
jjbarstats(
  data = histopathology,
  dep = LymphNodeMetastasis,
  group = Grade_Level,
  grvar = Sex,
  typestatistics = "robust",
  pairwisecomparisons = TRUE
)
```

# Customization and Theming

## Using Original ggstatsplot Theme

```{r original-theme}
jjbarstats(
  data = medical_study_data,
  dep = response,
  group = treatment_group,
  originaltheme = TRUE
)
```

## Custom Theming

The function respects jamovi's theming system when `originaltheme = FALSE` (default), allowing for consistent styling across analyses.

# Best Practices and Recommendations

## Statistical Method Selection

1. **Parametric**: Use when data meets assumptions (large sample sizes, expected frequencies ≥ 5)
2. **Non-parametric**: Default choice for categorical data, fewer assumptions
3. **Robust**: Good middle ground, less sensitive to outliers
4. **Bayesian**: When you want to incorporate prior knowledge or report Bayes factors

## Multiple Comparison Correction

1. **Holm**: Good balance between power and Type I error control
2. **Bonferroni**: Most conservative, use when Type I error is critical
3. **BH (Benjamini-Hochberg)**: Controls false discovery rate, good for exploratory analysis
4. **None**: Only when you have specific a priori hypotheses

## Sample Size Considerations

- Chi-squared tests require expected frequencies ≥ 5 in each cell
- Fisher's exact test is automatically used for small samples
- Consider effect sizes, not just p-values

## Data Preparation Tips

```{r data-preparation}
# Ensure categorical variables are properly formatted
medical_study_clean <- medical_study_data %>%
  mutate(
    treatment_group = factor(treatment_group, 
                           levels = c("Control", "Treatment A", "Treatment B")),
    response = factor(response,
                     levels = c("No Response", "Partial Response", "Complete Response"))
  )

# Verify factor levels
str(medical_study_clean[c("treatment_group", "response")])
```

# Troubleshooting Common Issues

## Issue 1: Empty Cells or Small Counts

When you have empty cells or very small counts, the function automatically switches to appropriate tests:

```{r small-counts}
# Create data with small counts
small_sample <- medical_study_data[1:20, ]

jjbarstats(
  data = small_sample,
  dep = response,
  group = treatment_group,
  typestatistics = "nonparametric"
)
```

## Issue 2: Too Many Categories

When you have many categories, consider grouping or using different visualization:

```{r many-categories}
# Example with multiple categories
jjbarstats(
  data = patient_satisfaction_data,
  dep = satisfaction_level,
  group = department,
  pairwisecomparisons = FALSE  # Disable pairwise for clarity
)
```

## Issue 3: Missing Data

The function automatically handles missing data when `excl = TRUE` (default):

```{r missing-data-demo}
# Demonstrate missing data handling
data_with_na <- medical_study_data
data_with_na$response[1:5] <- NA

jjbarstats(
  data = data_with_na,
  dep = response,
  group = treatment_group,
  excl = TRUE  # Exclude missing values
)
```

# Interpretation Guidelines

## Understanding the Statistical Output

1. **Chi-squared test**: Tests independence between categorical variables
2. **Effect size (Cramér's V)**: Measures strength of association (0 = no association, 1 = perfect association)
3. **Confidence intervals**: Provide range of plausible values for the effect
4. **Pairwise comparisons**: Show which specific groups differ

## Clinical Significance vs Statistical Significance

- Always consider clinical relevance alongside statistical significance
- Effect sizes help interpret practical importance
- Confidence intervals provide information about precision

## Reporting Results

When reporting results from jjbarstats:

1. Describe the statistical test used (chi-squared, Fisher's exact, etc.)
2. Report effect size (Cramér's V) and confidence intervals
3. Mention multiple comparison correction if applicable
4. Provide sample sizes for each group
5. Include the actual plot in your publication

# Advanced Examples

## Multi-stage Analysis Workflow

```{r workflow-example}
# Step 1: Overall analysis
# overall_result <- jjbarstats(
#   data = clinical_trial_data,
#   dep = primary_outcome,
#   group = drug_dosage,
#   typestatistics = "nonparametric",
#   pairwisecomparisons = TRUE,
#   padjustmethod = "BH"
# )

# Step 2: Subgroup analysis by study phase
# subgroup_result <- jjbarstats(
#   data = clinical_trial_data,
#   dep = primary_outcome,
#   group = drug_dosage,
#   grvar = study_phase,
#   typestatistics = "nonparametric",
#   pairwisecomparisons = TRUE
# )
```

## Comparative Analysis Across Different Outcomes

```{r comparative-analysis}
# Compare primary and secondary outcomes
# jjbarstats(
#   data = clinical_trial_data,
#   dep = c(primary_outcome, side_effects, baseline_condition),
#   group = drug_dosage,
#   typestatistics = "robust",
#   pairwisecomparisons = FALSE
# )
```

# Conclusion

The `jjbarstats` function provides a comprehensive solution for categorical data analysis in clinical and research settings. Its integration with the `ggstatsplot` ecosystem ensures both statistical rigor and visual appeal, making it an excellent choice for:

- Clinical trial analysis
- Quality improvement studies  
- Survey research
- Diagnostic test evaluation
- Healthcare outcomes research

The function's flexibility in statistical methods, multiple comparison corrections, and visualization options makes it suitable for both exploratory and confirmatory analysis phases of research projects.

## Further Resources

- [ggstatsplot documentation](https://indrajeetpatil.github.io/ggstatsplot/)
- [ClinicoPath package documentation](https://clinicopath.github.io/ClinicoPath/)
- Statistical methods references for categorical data analysis

## Session Information

```{r session-info}
sessionInfo()
```
