---
title: "Professional Waffle Charts with jwaffle"
subtitle: "Proportional Data Visualization Using Grid-Based Charts"
date: last-modified
author: "ClinicoPath"
output: html_document
---

# Introduction

The `jwaffle` function provides professional waffle chart visualizations for proportional data display. Waffle charts (also known as square pie charts) use a grid of squares to represent data proportions, making them particularly effective for displaying percentages, survey responses, and categorical distributions.

## What are Waffle Charts?

Waffle charts are a powerful alternative to traditional pie charts that offer several advantages:

- **Easy to read**: Grid-based layout is more intuitive than pie slices
- **Precise comparison**: Individual squares make exact proportions clear
- **Flexible scaling**: Can represent different scales (each square = 1%, 10 cases, etc.)
- **Professional appearance**: Clean, modern aesthetic suitable for presentations
- **Accessible design**: Color patterns are easier to distinguish than pie segments

## Key Features of jwaffle

- **Flexible Data Input**: Works with counts, frequencies, or case-level data
- **Advanced Customization**: Multiple color palettes, themes, and styling options
- **Faceting Support**: Create multi-panel charts for complex data comparison
- **Performance Optimized**: Intelligent caching for large datasets and repeated analyses
- **Professional Styling**: Publication-ready output with customizable themes

## Installation and Setup

```{r setup}
# Load required libraries
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Set options for better output
options(digits = 3)
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
  dpi = 300,
  out.width = "100%"
)

# Check if required packages are available
if (!requireNamespace("waffle", quietly = TRUE)) {
  message("Note: waffle package required for waffle chart creation")
}

if (!requireNamespace("scales", quietly = TRUE)) {
  message("Note: scales package recommended for enhanced formatting")
}
```

# Data Preparation

Let's create diverse datasets to demonstrate different waffle chart applications:

## Market Research Data

```{r market-data}
# Create market research dataset
set.seed(123)
market_data <- data.frame(
  # Core product preferences
  preferred_brand = factor(sample(c("Brand_A", "Brand_B", "Brand_C", "Brand_D", "Brand_E"), 
                                 500, replace = TRUE, prob = c(0.35, 0.25, 0.20, 0.15, 0.05))),
  
  purchase_frequency = sample(1:12, 500, replace = TRUE, 
                             prob = c(0.3, 0.2, 0.15, 0.1, 0.08, 0.07, rep(0.02, 6))),
  
  # Demographic segmentation
  age_group = factor(sample(c("18-25", "26-35", "36-45", "46-55", "55+"), 
                           500, replace = TRUE, prob = c(0.15, 0.25, 0.25, 0.20, 0.15))),
  
  region = factor(sample(c("North", "South", "East", "West"), 
                        500, replace = TRUE)),
  
  income_level = factor(sample(c("Low", "Middle", "High"), 
                              500, replace = TRUE, prob = c(0.25, 0.50, 0.25))),
  
  # Customer behavior
  loyalty_program = factor(sample(c("Member", "Non_Member"), 
                                 500, replace = TRUE, prob = c(0.4, 0.6))),
  
  purchase_channel = factor(sample(c("Online", "In_Store", "Mobile", "Phone"), 
                                  500, replace = TRUE, prob = c(0.45, 0.35, 0.15, 0.05)))
)

# Display structure
str(market_data)
head(market_data)
```

## Survey Response Data

```{r survey-data}
# Create employee satisfaction survey data
survey_data <- data.frame(
  # Satisfaction ratings (Likert scale)
  job_satisfaction = factor(sample(c("Very_Satisfied", "Satisfied", "Neutral", "Dissatisfied", "Very_Dissatisfied"), 
                                  400, replace = TRUE, prob = c(0.20, 0.35, 0.25, 0.15, 0.05)),
                           levels = c("Very_Dissatisfied", "Dissatisfied", "Neutral", "Satisfied", "Very_Satisfied")),
  
  work_life_balance = factor(sample(c("Excellent", "Good", "Fair", "Poor"), 
                                   400, replace = TRUE, prob = c(0.15, 0.45, 0.30, 0.10)),
                            levels = c("Poor", "Fair", "Good", "Excellent")),
  
  # Employee demographics
  department = factor(sample(c("Engineering", "Sales", "Marketing", "HR", "Finance", "Operations"), 
                            400, replace = TRUE, prob = c(0.25, 0.20, 0.15, 0.10, 0.15, 0.15))),
  
  tenure_years = factor(sample(c("Less_than_1", "1-2", "3-5", "5-10", "More_than_10"), 
                              400, replace = TRUE, prob = c(0.15, 0.25, 0.30, 0.20, 0.10))),
  
  employment_type = factor(sample(c("Full_Time", "Part_Time", "Contract"), 
                                 400, replace = TRUE, prob = c(0.80, 0.15, 0.05))),
  
  remote_work = factor(sample(c("Fully_Remote", "Hybrid", "On_Site"), 
                             400, replace = TRUE, prob = c(0.30, 0.45, 0.25)))
)

str(survey_data)
```

## Healthcare Data

```{r healthcare-data}
# Create healthcare outcome data
healthcare_data <- data.frame(
  # Treatment outcomes
  treatment_response = factor(sample(c("Complete_Response", "Partial_Response", "Stable_Disease", "Progressive_Disease"), 
                                    300, replace = TRUE, prob = c(0.25, 0.35, 0.25, 0.15))),
  
  # Patient characteristics
  age_category = factor(sample(c("Under_40", "40-60", "60-75", "Over_75"), 
                              300, replace = TRUE, prob = c(0.15, 0.35, 0.35, 0.15))),
  
  disease_stage = factor(sample(c("Early", "Intermediate", "Advanced"), 
                               300, replace = TRUE, prob = c(0.30, 0.45, 0.25))),
  
  treatment_type = factor(sample(c("Surgery", "Chemotherapy", "Radiation", "Immunotherapy", "Combination"), 
                                300, replace = TRUE, prob = c(0.20, 0.25, 0.20, 0.15, 0.20))),
  
  hospital_type = factor(sample(c("Academic", "Community", "Cancer_Center"), 
                               300, replace = TRUE, prob = c(0.35, 0.40, 0.25))),
  
  insurance_type = factor(sample(c("Private", "Medicare", "Medicaid"), 
                                300, replace = TRUE, prob = c(0.60, 0.30, 0.10)))
)

str(healthcare_data)
```

# Basic Usage

## Simple Waffle Chart

The most basic waffle chart shows the distribution of a categorical variable:

```{r basic-waffle}
# Basic waffle chart
basic_chart <- jwaffle(
  data = market_data,
  groups = "preferred_brand"
)

# Display the result
print("Basic waffle chart created successfully")
```

This creates a 10x10 grid (100 squares) where each square represents approximately 1% of the data.

## Understanding Waffle Chart Components

A waffle chart consists of:

- **Squares**: Individual units that make up the grid
- **Colors**: Different colors represent different categories
- **Legend**: Explains what each color represents
- **Caption**: Shows what each square represents (e.g., "Each square represents 5 cases")

## Customizing Rows and Layout

```{r custom-layout}
# Different grid sizes
waffle_5x20 <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  rows = 5  # Creates 5x20 grid (still 100 squares)
)

waffle_8x12 <- jwaffle(
  data = market_data,
  groups = "preferred_brand", 
  rows = 8  # Creates 8x12 grid (approximately 100 squares)
)

# Flipped orientation
waffle_flipped <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  rows = 10,
  flip = TRUE
)

print("Custom layout waffle charts created")
```

## Using Count Variables

When you have aggregated data with counts:

```{r count-variables}
# Aggregate data first
brand_summary <- market_data %>%
  group_by(preferred_brand) %>%
  summarise(customer_count = n(), .groups = "drop")

print(brand_summary)

# Create waffle chart with counts
count_waffle <- jwaffle(
  data = brand_summary,
  groups = "preferred_brand",
  counts = "customer_count"
)

print("Count-based waffle chart created")
```

# Advanced Features

## Faceting for Multi-Panel Charts

Create separate waffle charts for different subgroups:

```{r faceting}
# Faceted waffle chart
faceted_chart <- jwaffle(
  data = survey_data,
  groups = "job_satisfaction",
  facet = "department"
)

print("Faceted waffle chart created")

# Faceting with geographic data
regional_chart <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  facet = "region"
)

print("Regional comparison waffle chart created")
```

Faceting is particularly useful for:
- Comparing patterns across different groups
- Showing changes over time periods
- Analyzing subpopulations separately

## Color Palettes for Professional Presentation

```{r color-palettes}
# Default palette
default_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "default"
)

# Colorblind-friendly palette
colorblind_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "colorblind"
)

# Professional business palette
professional_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "professional"
)

# Presentation palette (high contrast)
presentation_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "presentation"
)

# Journal publication palette
journal_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "journal"
)

# Pastel colors
pastel_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "pastel"
)

# Dark theme colors
dark_palette <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  color_palette = "dark"
)

print("Multiple color palette waffle charts created")
```

### Color Palette Guidelines

- **Default**: General purpose, good balance of visibility and aesthetics
- **Colorblind**: Optimized for accessibility, distinguishable by all users
- **Professional**: Business-appropriate colors for corporate presentations
- **Presentation**: High contrast colors for projector displays
- **Journal**: Grayscale-friendly for academic publications
- **Pastel**: Soft colors for informal or creative presentations
- **Dark**: Bold colors that work well on dark backgrounds

## Titles and Labels

```{r titles-labels}
# Comprehensive labeling
labeled_chart <- jwaffle(
  data = survey_data,
  groups = "work_life_balance",
  facet = "employment_type",
  mytitle = "Work-Life Balance by Employment Type",
  legendtitle = "Balance Rating",
  color_palette = "professional",
  show_legend = TRUE
)

print("Labeled waffle chart created")

# Minimal labeling (no legend)
minimal_chart <- jwaffle(
  data = market_data,
  groups = "purchase_channel",
  mytitle = "Customer Purchase Channel Preferences", 
  show_legend = FALSE,
  color_palette = "presentation"
)

print("Minimal waffle chart created")
```

# Real-World Applications

## Market Research Analysis

### Brand Preference Study

```{r market-research}
# Overall brand preference
brand_preference <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  mytitle = "Overall Brand Preference Distribution",
  legendtitle = "Brand",
  color_palette = "professional",
  show_legend = TRUE
)

# Brand preference by age group
age_brand_analysis <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  facet = "age_group",
  mytitle = "Brand Preference by Age Group",
  legendtitle = "Brand",
  color_palette = "presentation",
  rows = 8
)

# Purchase channel analysis
channel_analysis <- jwaffle(
  data = market_data,
  groups = "purchase_channel",
  facet = "income_level",
  mytitle = "Purchase Channel Preference by Income Level",
  legendtitle = "Channel",
  color_palette = "colorblind"
)

print("Market research waffle charts created")
```

### Customer Segmentation

```{r customer-segmentation}
# Loyalty program membership
loyalty_chart <- jwaffle(
  data = market_data,
  groups = "loyalty_program",
  facet = "region",
  mytitle = "Loyalty Program Membership by Region",
  legendtitle = "Membership",
  color_palette = "professional",
  rows = 6
)

print("Customer segmentation chart created")
```

## Employee Satisfaction Survey

### Job Satisfaction Analysis

```{r job-satisfaction}
# Overall job satisfaction
overall_satisfaction <- jwaffle(
  data = survey_data,
  groups = "job_satisfaction",
  mytitle = "Employee Job Satisfaction Distribution",
  legendtitle = "Satisfaction Level",
  color_palette = "journal",
  rows = 10
)

# Satisfaction by department
dept_satisfaction <- jwaffle(
  data = survey_data,
  groups = "job_satisfaction",
  facet = "department",
  mytitle = "Job Satisfaction by Department",
  legendtitle = "Satisfaction",
  color_palette = "professional"
)

# Work-life balance analysis
balance_analysis <- jwaffle(
  data = survey_data,
  groups = "work_life_balance",
  facet = "remote_work",
  mytitle = "Work-Life Balance by Work Arrangement",
  legendtitle = "Balance Rating",
  color_palette = "colorblind",
  rows = 8
)

print("Employee satisfaction charts created")
```

### Tenure and Employment Analysis

```{r tenure-analysis}
# Employment type distribution
employment_dist <- jwaffle(
  data = survey_data,
  groups = "employment_type",
  mytitle = "Employment Type Distribution",
  legendtitle = "Employment Type",
  color_palette = "presentation",
  rows = 6
)

# Tenure analysis
tenure_analysis <- jwaffle(
  data = survey_data,
  groups = "tenure_years",
  facet = "department",
  mytitle = "Employee Tenure by Department",
  legendtitle = "Tenure",
  color_palette = "pastel"
)

print("Tenure analysis charts created")
```

## Healthcare Outcomes Research

### Treatment Response Analysis

```{r treatment-response}
# Overall treatment response
treatment_overall <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  mytitle = "Treatment Response Distribution",
  legendtitle = "Response Type",
  color_palette = "professional",
  rows = 10
)

# Response by treatment type
response_by_treatment <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  facet = "treatment_type",
  mytitle = "Treatment Response by Treatment Type", 
  legendtitle = "Response",
  color_palette = "journal"
)

# Response by disease stage
response_by_stage <- jwaffle(
  data = healthcare_data,
  groups = "treatment_response",
  facet = "disease_stage",
  mytitle = "Treatment Response by Disease Stage",
  legendtitle = "Response",
  color_palette = "colorblind",
  rows = 8
)

print("Healthcare outcome charts created")
```

### Healthcare System Analysis

```{r healthcare-system}
# Hospital type distribution
hospital_dist <- jwaffle(
  data = healthcare_data,
  groups = "hospital_type",
  mytitle = "Patient Distribution by Hospital Type",
  legendtitle = "Hospital Type",
  color_palette = "presentation"
)

# Insurance analysis
insurance_analysis <- jwaffle(
  data = healthcare_data,
  groups = "insurance_type",
  facet = "age_category",
  mytitle = "Insurance Type by Age Category",
  legendtitle = "Insurance",
  color_palette = "professional",
  rows = 6
)

print("Healthcare system charts created")
```

# Advanced Customization

## Complex Multi-Variable Analysis

```{r complex-analysis}
# Create complex segmented data
complex_data <- market_data %>%
  mutate(
    customer_segment = case_when(
      loyalty_program == "Member" & income_level == "High" ~ "Premium_Customer",
      loyalty_program == "Member" & income_level == "Middle" ~ "Standard_Member", 
      loyalty_program == "Member" & income_level == "Low" ~ "Budget_Member",
      loyalty_program == "Non_Member" & income_level == "High" ~ "High_Value_Prospect",
      TRUE ~ "Standard_Prospect"
    )
  )

# Complex segmentation analysis
segment_analysis <- jwaffle(
  data = complex_data,
  groups = "customer_segment",
  facet = "purchase_channel",
  mytitle = "Customer Segmentation by Purchase Channel",
  legendtitle = "Customer Segment",
  color_palette = "professional",
  rows = 8
)

print("Complex segmentation analysis created")
```

## Weighted Analysis with Purchase Frequency

```{r weighted-analysis}
# Use purchase frequency as weights
weighted_analysis <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  counts = "purchase_frequency",
  facet = "region",
  mytitle = "Brand Preference Weighted by Purchase Frequency",
  legendtitle = "Brand",
  color_palette = "presentation"
)

print("Weighted analysis chart created")
```

## Time-Based Analysis

```{r time-analysis}
# Create quarterly data
quarterly_data <- data.frame(
  quarter = rep(c("Q1", "Q2", "Q3", "Q4"), each = 100),
  satisfaction = factor(sample(c("Satisfied", "Neutral", "Dissatisfied"), 
                              400, replace = TRUE, prob = c(0.6, 0.25, 0.15))),
  product_line = factor(sample(c("Product_A", "Product_B", "Product_C"), 
                              400, replace = TRUE))
)

# Quarterly satisfaction trends
quarterly_satisfaction <- jwaffle(
  data = quarterly_data,
  groups = "satisfaction",
  facet = "quarter",
  mytitle = "Customer Satisfaction Trends by Quarter",
  legendtitle = "Satisfaction",
  color_palette = "journal",
  rows = 10
)

print("Quarterly trend analysis created")
```

# Performance and Optimization

## Large Dataset Handling

```{r large-dataset}
# Create large dataset for performance testing
large_data <- do.call(rbind, replicate(5, market_data, simplify = FALSE))
large_data$customer_id <- 1:nrow(large_data)

# Performance timing
system.time({
  large_chart1 <- jwaffle(
    data = large_data,
    groups = "preferred_brand",
    color_palette = "professional"
  )
})

# Second run should be faster due to caching
system.time({
  large_chart2 <- jwaffle(
    data = large_data,
    groups = "preferred_brand", 
    color_palette = "professional"
  )
})

print("Large dataset analysis completed")
```

## Caching Performance Benefits

The jwaffle function includes intelligent caching that provides several performance benefits:

- **Data Caching**: Prepared data is cached to avoid reprocessing
- **Plot Caching**: Generated plots are cached for identical parameters
- **Palette Caching**: Color palettes are cached based on options and group count
- **Hash-Based Invalidation**: Cache is automatically invalidated when data or options change

```{r cache-demo}
# Demonstrate caching with repeated calls
for (i in 1:3) {
  start_time <- Sys.time()
  
  result <- jwaffle(
    data = survey_data,
    groups = "job_satisfaction",
    color_palette = "professional"
  )
  
  end_time <- Sys.time()
  cat("Run", i, "completed in", round(as.numeric(end_time - start_time), 3), "seconds\n")
}
```

# Best Practices

## When to Use Waffle Charts

Waffle charts are ideal for:

- **Proportional data**: When showing percentages or parts of a whole
- **Survey responses**: Likert scales and categorical responses
- **Market share**: Brand preference and competitive analysis
- **Demographics**: Population breakdowns and characteristics
- **Binary outcomes**: Yes/no, success/failure, member/non-member

```{r best-practices-table}
# Create best practices guide
best_practices <- data.frame(
  Scenario = c(
    "Survey Results",
    "Market Share",
    "Demographics", 
    "Binary Outcomes",
    "Progress Tracking",
    "Budget Allocation"
  ),
  Waffle_Advantage = c(
    "Easy to see exact proportions",
    "Clear competitive positioning",
    "Intuitive population representation",
    "Simple visual comparison",
    "Progress visualization", 
    "Resource distribution clarity"
  ),
  Key_Settings = c(
    "Use colorblind palette, show legend",
    "Professional palette, custom title",
    "Facet by subgroups, clean layout",
    "High contrast colors, minimal legend",
    "Sequential colors, progress labels",
    "Professional theme, clear categories"
  )
)

knitr::kable(best_practices, caption = "Waffle Chart Best Practices by Use Case")
```

## Design Guidelines

### Color Selection

```{r color-guidelines}
# Create color guideline examples
color_guide <- data.frame(
  Palette = c("Default", "Colorblind", "Professional", "Presentation", "Journal", "Pastel", "Dark"),
  Best_For = c(
    "General purpose presentations",
    "Accessible design, inclusive audiences", 
    "Business meetings, corporate reports",
    "Projector displays, conference presentations",
    "Academic papers, grayscale printing",
    "Creative projects, informal presentations",
    "Dark themes, modern designs"
  ),
  Avoid_When = c(
    "Accessibility is critical",
    "Creative expression is primary goal",
    "Informal or creative contexts",
    "Fine detail distinction needed",
    "Color emphasis is important",
    "Professional formality required",
    "Light backgrounds are used"
  )
)

knitr::kable(color_guide, caption = "Color Palette Selection Guide")
```

### Layout Recommendations

```{r layout-recommendations}
# Layout guidelines
layout_guide <- data.frame(
  Grid_Size = c("5 rows", "8 rows", "10 rows", "12+ rows"),
  Best_For = c(
    "Simple data (2-4 categories)",
    "Moderate complexity (4-6 categories)",
    "Standard presentations (any categories)",
    "Detailed analysis (many categories)"
  ),
  Considerations = c(
    "Large squares, easy to count",
    "Good balance of detail and clarity",
    "Standard 10x10 grid, familiar format",
    "Smaller squares, harder to count individually"
  )
)

knitr::kable(layout_guide, caption = "Grid Size Selection Guide")
```

## Accessibility Considerations

```{r accessibility}
# Accessible waffle chart example
accessible_chart <- jwaffle(
  data = survey_data,
  groups = "job_satisfaction",
  color_palette = "colorblind",  # Colorblind-friendly
  mytitle = "Employee Job Satisfaction Survey Results (N=400)",  # Descriptive title
  legendtitle = "Satisfaction Level",  # Clear legend
  show_legend = TRUE,  # Always show legend for accessibility
  rows = 10  # Standard grid for easy counting
)

print("Accessible waffle chart created")
```

Accessibility best practices:
- Use colorblind-friendly palettes
- Include descriptive titles and legends
- Provide sample size information
- Use standard grid sizes when possible
- Consider alternative text descriptions for digital formats

# Integration with Research Workflows

## Statistical Analysis Integration

```{r statistical-integration}
# Combine waffle charts with statistical tests
library(dplyr)

# Chi-square test for independence
if (requireNamespace("stats", quietly = TRUE)) {
  # Test brand preference vs age group
  contingency_table <- table(market_data$preferred_brand, market_data$age_group)
  chi_test <- chisq.test(contingency_table)
  
  cat("Chi-square test results:\n")
  cat("Chi-square =", round(chi_test$statistic, 3), "\n")
  cat("p-value =", round(chi_test$p.value, 4), "\n")
  cat("Significant association:", ifelse(chi_test$p.value < 0.05, "Yes", "No"), "\n")
}

# Combine with waffle visualization
statistical_chart <- jwaffle(
  data = market_data,
  groups = "preferred_brand",
  facet = "age_group",
  mytitle = paste("Brand Preference by Age Group",
                  ifelse(chi_test$p.value < 0.05, "(p < 0.05)", "(n.s.)")),
  legendtitle = "Brand",
  color_palette = "journal"
)

print("Statistical analysis with waffle chart created")
```

## Report Generation

```{r report-generation}
# Function to generate waffle chart reports
generate_waffle_report <- function(data, group_var, facet_var = NULL, title_prefix = "") {
  
  # Basic summary statistics
  summary_stats <- data %>%
    group_by(.data[[group_var]]) %>%
    summarise(
      count = n(),
      percentage = round(n() / nrow(data) * 100, 1),
      .groups = "drop"
    )
  
  # Create waffle chart
  chart <- jwaffle(
    data = data,
    groups = group_var,
    facet = facet_var,
    mytitle = paste(title_prefix, "Distribution of", gsub("_", " ", group_var)),
    legendtitle = gsub("_", " ", group_var),
    color_palette = "professional"
  )
  
  # Return summary and chart
  return(list(
    summary = summary_stats,
    chart = chart,
    total_n = nrow(data)
  ))
}

# Generate reports
brand_report <- generate_waffle_report(
  data = market_data,
  group_var = "preferred_brand",
  title_prefix = "Market Research:"
)

satisfaction_report <- generate_waffle_report(
  data = survey_data,
  group_var = "job_satisfaction",
  facet_var = "department",
  title_prefix = "Employee Survey:"
)

# Display summaries
cat("Brand Preference Summary:\n")
print(brand_report$summary)
cat("\nTotal respondents:", brand_report$total_n, "\n")

cat("\nJob Satisfaction Summary:\n")
print(satisfaction_report$summary)
cat("\nTotal employees:", satisfaction_report$total_n, "\n")
```

# Troubleshooting

## Common Issues and Solutions

### Data Validation

```{r data-validation}
# Function to validate waffle chart data
validate_waffle_data <- function(data, groups, counts = NULL, facet = NULL) {
  issues <- c()
  
  # Check if data is not empty
  if (nrow(data) == 0) {
    issues <- c(issues, "Data frame is empty")
  }
  
  # Check if group variable exists
  if (!groups %in% names(data)) {
    issues <- c(issues, paste("Group variable", groups, "not found in data"))
  }
  
  # Check if count variable exists (if specified)
  if (!is.null(counts) && !counts %in% names(data)) {
    issues <- c(issues, paste("Count variable", counts, "not found in data"))
  }
  
  # Check if facet variable exists (if specified)
  if (!is.null(facet) && !facet %in% names(data)) {
    issues <- c(issues, paste("Facet variable", facet, "not found in data"))
  }
  
  # Check for sufficient data in groups
  if (groups %in% names(data)) {
    group_counts <- table(data[[groups]])
    if (any(group_counts == 0)) {
      issues <- c(issues, "Some groups have zero observations")
    }
  }
  
  # Check for missing values
  if (groups %in% names(data)) {
    missing_groups <- sum(is.na(data[[groups]]))
    if (missing_groups > 0) {
      issues <- c(issues, paste(missing_groups, "missing values in group variable"))
    }
  }
  
  return(issues)
}

# Test validation
validation_result <- validate_waffle_data(
  data = market_data,
  groups = "preferred_brand",
  facet = "region"
)

if (length(validation_result) == 0) {
  cat("✓ Data validation passed\n")
} else {
  cat("⚠ Data validation issues:\n")
  for (issue in validation_result) {
    cat("-", issue, "\n")
  }
}
```

### Handling Missing Data

```{r missing-data}
# Create data with missing values
missing_data <- market_data
missing_data$preferred_brand[sample(1:500, 25)] <- NA
missing_data$region[sample(1:500, 15)] <- NA

# Check missing data impact
cat("Original data:", nrow(market_data), "rows\n")
cat("Data with missing values:", nrow(missing_data), "rows\n")
cat("Complete cases:", sum(complete.cases(missing_data)), "rows\n")

# Create waffle chart (automatically handles missing data)
missing_chart <- jwaffle(
  data = missing_data,
  groups = "preferred_brand",
  facet = "region",
  mytitle = "Brand Preference with Missing Data Handling",
  legendtitle = "Brand",
  color_palette = "colorblind"
)

print("Missing data chart created (missing values automatically excluded)")
```

### Performance Optimization Tips

```{r performance-tips}
# Performance optimization guidelines
perf_tips <- data.frame(
  Scenario = c(
    "Large datasets (>1000 rows)",
    "Many categories (>8 groups)",
    "Frequent updates",
    "Multiple facets",
    "Presentation mode"
  ),
  Recommendation = c(
    "Use caching, avoid unnecessary recalculation",
    "Consider grouping rare categories",
    "Minimize data changes between calls",
    "Limit number of facet levels",
    "Use high-contrast palettes"
  ),
  Implementation = c(
    "Keep same data object, change options only",
    "Create 'Other' category for small groups",
    "Batch process multiple charts",
    "Filter data to key comparisons",
    "Use 'presentation' color palette"
  )
)

knitr::kable(perf_tips, caption = "Performance Optimization Guide")
```

# Summary

The `jwaffle` function provides a comprehensive solution for creating professional waffle charts. Key advantages include:

## Core Visualization Capabilities

- **Intuitive Design**: Grid-based layout that's easier to interpret than pie charts
- **Flexible Input**: Works with raw data, counts, or aggregated summaries
- **Professional Output**: Publication and presentation-ready visualizations
- **Accessibility**: Colorblind-friendly palettes and clear visual hierarchy

## Advanced Features

- **Faceting Support**: Multi-panel comparisons across subgroups
- **Custom Styling**: Multiple color palettes and theme options
- **Performance Optimization**: Intelligent caching for large datasets
- **Error Handling**: Graceful handling of edge cases and missing data

## Real-World Applications

- **Market Research**: Brand preference, customer segmentation, purchase behavior
- **Survey Analysis**: Employee satisfaction, customer feedback, opinion polls
- **Healthcare Research**: Treatment outcomes, patient demographics, quality metrics
- **Business Intelligence**: Resource allocation, performance metrics, KPI tracking

## Integration Benefits

- **Statistical Workflow**: Seamless integration with statistical testing
- **Report Generation**: Automated chart creation for regular reporting
- **Data Validation**: Built-in checks for data quality and completeness
- **Accessibility**: Support for inclusive design and universal access

## Next Steps

- Explore different applications for your specific data visualization needs
- Experiment with color palettes and styling for your organization's brand
- Try the performance optimization features with your datasets
- Consider jwaffle for standardizing proportional data visualization

```{r session-info}
sessionInfo()
```
