---
title: "Firth's Penalized Likelihood Regression - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Firth's Penalized Likelihood Regression - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
data_path <- "data-raw/non-rda/"
data_path2 <- "data/"
```

> **Note:** The `firthregression()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference.

# Firth's Penalized Likelihood Regression

## Overview

Firth's penalized likelihood regression adds a Jeffreys-prior penalty to the likelihood function, producing finite and bias-corrected coefficient estimates even when standard maximum likelihood fails. This is essential for clinical studies with small samples, rare events, or complete/quasi-complete separation -- situations where standard logistic or Cox regression produces infinite or severely biased estimates.

Unlike LASSO (which selects variables by shrinking some to zero), Firth regression keeps all variables but corrects first-order bias. It provides profile likelihood confidence intervals (more accurate than Wald for small samples), penalized likelihood ratio test p-values, and automatic separation detection.

The module supports two modes: **Firth logistic regression** (binary outcomes via the `logistf` package) and **Firth Cox regression** (survival outcomes via the `coxphf` package). Both are reference implementations by Georg Heinze.

---

## Datasets Used in This Guide

| Dataset | N | Events | Primary Use |
|---------|---|--------|-------------|
| `firth_standard` | 120 | 58 (logistic) / 31 (Cox) | Balanced clinical data, both modes |
| `firth_separation` | 80 | 11 (~14%) | Rare events with complete separation |
| `firth_smallcox` | 50 | 23 | Small survival cohort |

---

## 1. Firth Logistic Regression

### Basic logistic analysis

```{r logistic-basic}
standard <- read.csv(paste0(data_path, "firth_standard.csv"))
str(standard[, c("mortality", "age", "grade", "tumor_size", "lvi", "marker")])

firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size", "lvi", "marker"),
  suitabilityCheck = TRUE,
  separationCheck = TRUE,
  compareStandard = TRUE,
  showModelFit = TRUE
)
```

Look for: coefficients with OR, profile likelihood CIs, p-values from penalized LR tests, bias reduction %, model fit statistics, separation diagnostics.

---

## 2. Firth Cox Regression

### Survival analysis mode

```{r cox-basic}
firthregression(
  data = standard,
  analysisType = "cox",
  time = "follow_up_time",
  outcome = "status",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size", "lvi", "marker"),
  suitabilityCheck = TRUE,
  compareStandard = TRUE
)
```

Look for: coefficients with HR, profile CIs (always used for Cox via coxphf), comparison with standard Cox model.

---

## 3. Confidence Intervals

### Profile vs Wald CIs

Profile likelihood CIs are recommended (default) as they are more accurate for small samples. Wald CIs are faster but less reliable near separation.

```{r ci-profile}
firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size"),
  ciLevel = 0.95,
  ciMethod = "profile",
  suitabilityCheck = FALSE
)
```

```{r ci-wald}
firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size"),
  ciLevel = 0.99,
  ciMethod = "wald",
  suitabilityCheck = FALSE
)
```

Compare the CI widths: profile CIs are typically asymmetric (reflecting the actual likelihood shape), while Wald CIs are symmetric.

---

## 4. Separation Detection

### Data with complete separation

The `firth_separation` dataset includes a variable (`margin_positive`) that perfectly predicts the outcome -- all positive-margin patients had recurrence. Standard logistic regression would produce infinite coefficients.

```{r separation}
separation <- read.csv(paste0(data_path, "firth_separation.csv"))
table(separation$outcome, separation$margin_positive)

firthregression(
  data = separation,
  analysisType = "logistic",
  outcome = "outcome",
  outcomeLevel = "Recurrence",
  predictors = c("age", "bmi", "grade", "margin_positive"),
  separationCheck = TRUE,
  compareStandard = TRUE,
  suitabilityCheck = TRUE
)
```

Look for: separation diagnostics table flagging `margin_positive`, comparison table showing how standard GLM produces extreme estimates while Firth gives finite values, bias reduction %.

---

## 5. Bias Reduction Comparison

### Firth vs Standard model side-by-side

```{r comparison}
firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size", "lvi", "marker"),
  compareStandard = TRUE,
  showModelFit = TRUE,
  suitabilityCheck = FALSE
)
```

Look for: comparison table with Firth coefficient, Standard coefficient, Firth OR, Standard OR, and p-values from both models. The bias_reduction column in the coefficients table shows % change.

---

## 6. Plots

### Forest plot

```{r forest-plot}
firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size", "lvi", "marker"),
  forestPlot = TRUE,
  suitabilityCheck = FALSE
)
```

### Separation diagnostic plot

```{r separation-plot}
firthregression(
  data = separation,
  analysisType = "logistic",
  outcome = "outcome",
  outcomeLevel = "Recurrence",
  predictors = c("age", "grade", "margin_positive"),
  separationPlot = TRUE,
  suitabilityCheck = FALSE
)
```

---

## 7. Clinical Output

### Results summary and explanations

```{r clinical-output}
firthregression(
  data = standard,
  analysisType = "logistic",
  outcome = "mortality",
  outcomeLevel = "Dead",
  predictors = c("age", "grade", "tumor_size"),
  showSummary = TRUE,
  showExplanations = TRUE,
  suitabilityCheck = FALSE
)
```

Look for: natural-language results summary suitable for reports, and methodological explanations about when and why to use Firth's method.

---

## 8. Small Cox Cohort

```{r small-cox}
smallcox <- read.csv(paste0(data_path, "firth_smallcox.csv"))
table(smallcox$status)

firthregression(
  data = smallcox,
  analysisType = "cox",
  time = "time",
  outcome = "status",
  outcomeLevel = "Dead",
  predictors = c("age", "treatment", "biomarker"),
  suitabilityCheck = TRUE,
  compareStandard = TRUE,
  showSummary = TRUE
)
```

---

## References

- Firth D (1993). "Bias reduction of maximum likelihood estimates." *Biometrika*, 80(1), 27-38.
- Heinze G, Schemper M (2001). "A solution to the problem of separation in logistic regression." *Statistics in Medicine*, 20(2), 169-177.
- Heinze G, Ploner M (2023). *logistf: Firth's Bias-Reduced Logistic Regression*. R package.
- Heinze G, Ploner M, Beyea J (2023). *coxphf: Cox Regression with Firth's Penalized Likelihood*. R package.
