---
title: "Conditional Survival Estimation - Comprehensive Guide"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Conditional Survival Estimation - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** `conditionalsurvival()` is designed primarily for the jamovi GUI. The R
> syntax shown here is for reference and advanced scripting.

# Conditional Survival Estimation

## Overview

In clinical oncology, a patient's prognosis improves with each additional year of
disease-free survival. Standard Kaplan-Meier estimates describe the probability of
surviving beyond time *t* from the moment of diagnosis or treatment. **Conditional
survival** answers a different and often more useful question: given that a patient
has already survived to time *s*, what is their probability of surviving to time *t*?

Formally: CS(t | s) = P(T > t | T > s) = S(t) / S(s)

This is implemented via the `conditionalsurvival()` function in the jsurvival module
(menu: SurvivalD > ClinicoPath Survival). Four estimation methods are available:

| Method | Key | Description |
|--------|-----|-------------|
| Kaplan-Meier Weights | `km` | Weighted KM estimation via `condSURV::KMW()` (falls back to S(t)/S(s) ratio when condSURV is not available) |
| Landmark Approach | `landmark` | Subsets cohort to patients surviving past the conditioning time, then refits KM |
| Inverse Probability Weighting | `ipw` | IPCW-based estimation (stub; currently falls back to manual KM ratio) |
| Presmoothed Kaplan-Meier | `pkm` | Kernel-smoothed KM (stub; currently falls back to manual KM ratio) |

## Datasets

This vignette uses the bundled `conditionalsurvival_test` dataset:

| Dataset | N | Scenario | Key Variables |
|---------|---|----------|---------------|
| `conditionalsurvival_test` | 150 | Colorectal cancer, Weibull survival, stage-dependent hazard | `OverallTime` (months), `Event` (0/1), `EventFactor` (Alive/Dead), `Treatment`, `Stage`, `Grade`, `Age`, `Sex`, `CEA` |
| `histopathology` | varies | Built-in ClinicoPath dataset | General clinicopathological variables |

```{r load-data}
data("conditionalsurvival_test")
str(conditionalsurvival_test)
summary(conditionalsurvival_test[, c("OverallTime", "Event", "Treatment", "Stage")])
```

---

# Basic Usage

## Unstratified Analysis (Overall Cohort)

The simplest use case: estimate conditional survival for the entire cohort with
default settings. When `conditionTime = 0`, the function automatically uses the
median follow-up time.

```{r basic-unstratified}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 0,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = TRUE
)
```

## Unstratified with Fixed Conditioning Time

Set `conditionTime = 12` to ask: "Given a patient survived 12 months, what is their
probability of surviving to each subsequent time point?"

```{r basic-fixed-condtime}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = FALSE
)
```

---

# Stratified Analysis

## By Treatment

Compare conditional survival between Surgery and Surgery+Chemo groups. The
`conditionVar` option activates stratified analysis, producing separate
conditional survival estimates and curves per group.

```{r stratified-treatment}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = TRUE
)
```

## By Stage

Stage has 4 levels (I, II, III, IV). Groups with fewer than 3 events are
silently skipped. Early-stage patients with few events at late conditioning
times may be excluded automatically.

```{r stratified-stage}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Stage,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = FALSE
)
```

## By Grade

Three tumor differentiation levels: Well, Moderate, Poor.

```{r stratified-grade}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Grade,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

---

# Estimation Methods

## Kaplan-Meier Weights (Default)

The default method. When the `condSURV` package is installed, uses KMW weights
for estimation. Otherwise falls back to the manual S(t)/S(s) ratio.

```{r method-km}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = TRUE
)
```

## Landmark Approach

The landmark method subsets the data to only patients who survived past the
conditioning time, then estimates survival from that point forward. This is the
simplest and most transparent approach but discards information from patients
censored before the landmark.

```{r method-landmark}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "landmark",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = TRUE
)
```

## Inverse Probability Weighting

The IPW method is intended to use inverse probability of censoring weights for
unbiased estimation. Currently implemented as a stub that falls back to the
manual KM ratio method.

```{r method-ipw}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "ipw",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = TRUE
)
```

## Presmoothed Kaplan-Meier

The PKM method applies kernel smoothing to the KM estimator. Currently a stub
that falls back to the manual KM ratio. The `bandwidth` option is reserved for
controlling the smoothing bandwidth when this method is fully implemented.

```{r method-pkm}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "pkm",
  bandwidth    = 0,
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = TRUE
)
```

---

# Custom Time Points

By default, the function generates ~5 evenly spaced time points from the
conditioning time to the maximum follow-up. Use the `timePoints` option to
specify exact time points of clinical interest (comma-separated string).

## Clinically Meaningful Time Points

For example, 1-year, 2-year, 3-year, and 5-year conditional survival given
survival to 12 months:

```{r custom-timepoints}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 12,
  timePoints   = "24,36,48,60",
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Time Points Including Values Before Conditioning Time

When a requested time point is at or before the conditioning time, the
conditional survival is exactly 1.0 (the patient has already survived past
that point).

```{r custom-timepoints-before}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 24,
  timePoints   = "12,18,24,36,48,60",
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Sparse Time Points

Only two time points of interest:

```{r custom-timepoints-sparse}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 6,
  timePoints   = "24,60",
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

---

# Factor Outcome Variable

The `outcomeVar` accepts both numeric (0/1) and factor (2-level) variables.
Factor variables are converted internally: the first level becomes 0 (censored),
the second level becomes 1 (event).

## Using EventFactor (Alive/Dead)

```{r factor-outcome}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = EventFactor,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = FALSE
)
```

---

# Confidence Level Variations

The `confInt` option controls the width of confidence intervals around
conditional survival estimates. The default is 0.95 (95% CI).

## 90% Confidence Intervals

```{r ci-90}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 12,
  confInt      = 0.90,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## 99% Confidence Intervals

```{r ci-99}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 12,
  confInt      = 0.99,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## 80% Confidence Intervals

```{r ci-80}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  confInt      = 0.80,
  method       = "landmark",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

---

# Edge Cases

## Conditioning Time at Median (Auto-Selection)

When `conditionTime` is 0, NULL, or NA, the function uses the median follow-up
time. This is the safest default for exploratory analyses.

```{r edge-auto-condtime}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 0,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Late Conditioning Time

A late conditioning time reduces the risk set. Estimates become less stable
with wider confidence intervals.

```{r edge-late-condtime}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 60,
  method       = "km",
  showTable    = TRUE,
  showPlot     = TRUE,
  showExplanations = FALSE
)
```

## Conditioning Time Beyond Maximum Follow-Up (Expect Error)

When the conditioning time is at or beyond the maximum observed follow-up,
the function rejects the analysis with an informative error.

```{r edge-condtime-beyond-max}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 200,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Too Few Events (Expect Error)

A dataset with fewer than 5 events is rejected.

```{r edge-few-events}
few_events_data <- data.frame(
  time  = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50,
            55, 60, 65, 70, 75, 80, 85, 90, 95, 100),
  event = c(1, 1, 0, 0, 0, 0, 0, 0, 0, 0,
            0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  group = rep(c("A", "B"), 10)
)

conditionalsurvival(
  data         = few_events_data,
  timeVar      = time,
  outcomeVar   = event,
  conditionTime = 10,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Three-Level Factor Outcome (Expect Error)

The outcome variable must have exactly 2 levels. A 3-level factor is rejected.

```{r edge-3level-factor}
bad_data <- conditionalsurvival_test
bad_data$BadOutcome <- factor(
  sample(c("Alive", "Dead", "Unknown"), nrow(bad_data), replace = TRUE)
)

conditionalsurvival(
  data         = bad_data,
  timeVar      = OverallTime,
  outcomeVar   = BadOutcome,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Single-Level Conditioning Variable (Expect Error)

A grouping variable with only one level cannot be used for stratified analysis.

```{r edge-single-level}
single_data <- conditionalsurvival_test
single_data$SingleGroup <- "Everyone"

conditionalsurvival(
  data         = single_data,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = SingleGroup,
  conditionTime = 12,
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Display Options: All Off

The `showTable`, `showPlot`, and `showExplanations` toggles control visibility
of their respective panels. The `todo`, `reportSentence`, and `assumptions`
panels are always visible.

```{r edge-display-all-off}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionVar = Treatment,
  conditionTime = 12,
  method       = "km",
  showTable    = FALSE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

## Plot Type Option (Currently Curves Only)

The `plotType` option accepts `"curves"`, `"probability"`, or `"both"`, but
the plot rendering currently always draws the curves style regardless of
the selected value.

```{r edge-plottype}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 12,
  plotType     = "probability",
  method       = "km",
  showTable    = FALSE,
  showPlot     = TRUE,
  showExplanations = FALSE
)
```

---

# Clinical Interpretation

Conditional survival is most valuable in clinical settings where patients
return for follow-up visits and want updated prognostic information.

**Key interpretation points:**

1. **CS(t | s) is always >= S(t)**: A patient who has survived to time *s*
   always has a survival probability at least as high as the original
   unconditional estimate at time *t*.

2. **The conditioning effect is strongest early**: The biggest prognostic
   "improvement" from conditional survival typically occurs in the first
   1--3 years after treatment, when the hazard is highest.

3. **Plateau behavior**: If the hazard decreases over time (as in many
   cancers after successful treatment), conditional survival probabilities
   approach 1.0 as the conditioning time increases.

4. **Group comparisons**: Stratified conditional survival can reveal whether
   prognostic differences between treatment arms or stage groups narrow
   over time (a common pattern in oncology).

5. **Clinical communication**: The report sentence generated by this function
   is designed to be copied directly into manuscript results sections.

## Example: Prognostic Update at Different Conditioning Times

```{r clinical-update-6mo}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 6,
  timePoints   = "12,24,36,60",
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

```{r clinical-update-24mo}
conditionalsurvival(
  data         = conditionalsurvival_test,
  timeVar      = OverallTime,
  outcomeVar   = Event,
  conditionTime = 24,
  timePoints   = "36,48,60,72",
  method       = "km",
  showTable    = TRUE,
  showPlot     = FALSE,
  showExplanations = FALSE
)
```

---

# References

1. Hieke S, Kleber M, Konig C, Engelhardt M, Schumacher M. Conditional
   survival: a useful concept to provide information on how prognosis
   evolves over time. *Clin Cancer Res*. 2015;21(7):1530-1536.

2. Zamboni BA, Yothers G, Choi M, et al. Conditional survival and the
   choice of conditioning set for patients with colon cancer: an analysis
   of NSABP trials C-03 through C-07. *J Clin Oncol*. 2010;28(15):2544-2548.

3. Skuladottir H, Olsen JH. Conditional survival of patients with the four
   main types of lung cancer in Denmark, 1978-2001. *Cancer*. 2003;97(8):
   2014-2019.

4. Beran R. Nonparametric regression with randomly censored survival data.
   Technical Report, University of California, Berkeley. 1981.

5. Meira-Machado L, de Una-Alvarez J, Cadarso-Suarez C. Nonparametric
   estimation of transition probabilities in a non-Markov illness-death
   model. *Lifetime Data Anal*. 2006;12(3):325-344.
