---
title: "Comprehensive Survival Analysis with ClinicoPath"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Comprehensive Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## 1. Introduction

The `survival` function in ClinicoPath provides a comprehensive suite of
survival analysis tools designed for clinicopathological research. It combines
Kaplan-Meier estimation, Cox proportional hazards regression, and a wide range
of advanced methods into a single, unified analysis.

**What it produces by default (no optional flags needed):**

- Median survival time with 95% confidence intervals for each group
- Cox proportional hazards regression with hazard ratios
- Survival probability table at user-specified time points (default: 12, 36, 60 months)

**What you can add with a single checkbox:**

- Kaplan-Meier plots, cumulative events/hazard plots, KMunicate-style plots
- Pairwise group comparisons with multiple testing correction
- Proportional hazards assumption testing
- Restricted Mean Survival Time (RMST)
- Person-time analysis with incidence rates
- Weighted log-rank tests (Fleming-Harrington family)
- Bootstrap internal validation (optimism-corrected C-index)
- Calibration curves for Cox model assessment
- Non-linearity assessment via restricted cubic splines
- Age-adjusted, age-stratified, and age-as-time-scale models
- Competing risks and cause-specific survival
- REMARK reporting checklist
- Natural language summaries and clinical interpretation

**Clinical safety features:**

- Analysis is blocked when fewer than 10 events are detected
- Warnings are issued for low event counts (10-19, 20-49)
- Competing risk mode automatically skips analyses that are not valid in that context

## 2. Loading the Data

The `survival_test` dataset bundled with ClinicoPath contains 200 simulated
patients with survival time, binary outcome, treatment group, stage, grade,
age, sex, performance status, and a continuous biomarker value. It is designed
to demonstrate all features of the survival function.

```{r load-data}
library(ClinicoPath)
data(survival_test, package = "ClinicoPath")
str(survival_test)
```

The key columns are:

| Column | Description | Type |
|--------|-------------|------|
| `elapsedtime` | Survival time in months | numeric |
| `outcome` | Event indicator (1 = event, 0 = censored) | numeric |
| `treatment` | Treatment group (Control, Treatment A, Treatment B) | factor |
| `stage` | Cancer stage (I, II, III, IV) | factor |
| `grade` | Tumor grade (1, 2, 3) | factor |
| `age` | Patient age at diagnosis | numeric |
| `sex` | Patient sex (Female, Male) | factor |
| `biomarker_value` | Continuous biomarker measurement | numeric |

## 3. Basic Survival Analysis

A minimal call requires three variables: `elapsedtime`, `outcome`, and
`explanatory`. The function automatically computes median survival,
Cox regression, and survival probability tables. Note that `dod`, `dooc`,
`awd`, and `awod` must be supplied (as empty strings) even when not used.

```{r basic-analysis}
result <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = ""
)
```

This produces:

1. **Median Survival Table** -- median survival time, restricted mean, number
   of events, and 95% CI for each level of the explanatory variable.
2. **Cox Regression Table** -- hazard ratios (univariable) with 95% CI and
   p-values for each level compared to the reference.
3. **Survival Probability Table** -- survival probabilities at 12, 36, and 60
   months (customizable via `cutp`).

## 4. Survival Plots

ClinicoPath provides five distinct plot types. Enable each with a single
boolean option.

### 4.1 Kaplan-Meier Survival Curve

```{r km-plot}
result_plots <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  sc = TRUE,
  ci95 = TRUE,
  risktable = TRUE,
  censored = TRUE,
  pplot = TRUE,
  medianline = "hv",
  endplot = 60,
  byplot = 12,
  ybegin_plot = 0,
  yend_plot = 1
)
```

**Plot appearance options:**

- `ci95` -- shaded 95% confidence band around each curve
- `risktable` -- number-at-risk table below the x-axis
- `censored` -- tick marks where censoring occurred
- `pplot` -- log-rank p-value annotation
- `medianline` -- horizontal ("h"), vertical ("v"), or both ("hv") median line
- `endplot`, `byplot` -- control x-axis range and tick spacing
- `ybegin_plot`, `yend_plot` -- control y-axis range

### 4.2 Cumulative Events

Shows the cumulative proportion of events over time (1 minus survival).

```{r ce-plot}
result_ce <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  ce = TRUE,
  endplot = 60,
  byplot = 12
)
```

### 4.3 Cumulative Hazard

Displays the Nelson-Aalen cumulative hazard estimate.

```{r ch-plot}
result_ch <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  ch = TRUE,
  endplot = 60,
  byplot = 12
)
```

### 4.4 KMunicate-Style Plot

A publication-ready survival plot following the KMunicate recommendations for
improved readability of Kaplan-Meier curves.

```{r kmunicate-plot}
result_kmunicate <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  kmunicate = TRUE,
  endplot = 60,
  byplot = 12
)
```

### 4.5 Log-Log Plot

Used for visual assessment of the proportional hazards assumption. If the
log-log curves are approximately parallel, the proportional hazards assumption
is reasonable.

```{r loglog-plot}
result_loglog <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  loglog = TRUE
)
```

## 5. Survival Probability Tables

The survival probability table reports estimates at user-specified time points.
Use the `cutp` argument to define these cutpoints as a comma-separated string.

```{r surv-table}
result_table <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  cutp = "6, 12, 24, 36, 48, 60"
)
```

The output table includes, for each stratum and time point:

- Number at risk
- Number of events
- Survival probability (with 95% CI)

## 6. Pairwise Comparisons

When the explanatory variable has more than two levels, pairwise log-rank tests
identify which specific groups differ significantly. Multiple testing correction
is applied via the `padjustmethod` option.

```{r pairwise}
result_pw <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  pw = TRUE,
  padjustmethod = "bonferroni"
)
```

Available adjustment methods: `holm` (default), `hochberg`, `hommel`,
`bonferroni`, `BH`, `BY`, `fdr`, `none`.

## 7. Proportional Hazards Assumption Testing

The Cox model assumes proportional hazards -- that the hazard ratio between
groups is constant over time. This can be tested formally using Schoenfeld
residuals and visually via the Schoenfeld residual plot.

```{r ph-test}
result_ph <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  ph_cox = TRUE
)
```

The output includes:

- **Formal test** -- `cox.zph()` test with chi-square statistic and p-value.
  A significant p-value (< 0.05) suggests violation.
- **Interpretation** -- automated plain-language assessment with recommendations
- **Schoenfeld residual plot** -- scaled residuals vs. time; a flat trend
  supports proportional hazards

## 8. RMST Analysis

Restricted Mean Survival Time represents the area under the survival curve up
to a specified time horizon (tau). It is particularly useful when:

- The survival curves cross (non-proportional hazards)
- Median survival cannot be estimated (too few events)
- A clinically meaningful time horizon exists (e.g., 5-year RMST)

```{r rmst}
result_rmst <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  rmst_analysis = TRUE,
  rmst_tau = 48
)
```

If `rmst_tau` is set to 0 (default), the function automatically uses the 75th
percentile of observed follow-up time. A table note indicates the chosen tau.

The RMST table reports, for each group:

- RMST estimate (average survival time up to tau)
- Standard error
- 95% confidence interval
- The tau value used

## 9. Person-Time Analysis

Person-time analysis calculates incidence rates that properly account for
varying follow-up durations. This is essential for epidemiological studies
where participants have unequal observation periods.

```{r person-time}
result_pt <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  person_time = TRUE,
  time_intervals = "12, 24, 36, 48, 60",
  rate_multiplier = 100
)
```

The person-time table shows, for each time interval:

- Number of events
- Total person-time accumulated
- Incidence rate (events per `rate_multiplier` person-time units)
- 95% confidence interval for the rate

The `rate_multiplier` controls the denominator: 100 gives rates per 100
person-months, 1000 gives rates per 1000 person-months.

## 10. Weighted Log-Rank Tests

The standard log-rank test weights all time points equally. Weighted variants
from the Fleming-Harrington family emphasize different parts of the survival
curve, which is important when hazards are non-proportional.

```{r weighted-logrank}
result_wlr <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  weightedLogRank = TRUE,
  survivalTestType = "fleming_harrington"
)
```

When enabled, the function runs all four weighted tests simultaneously:

| Test | rho | Emphasizes |
|------|-----|------------|
| Log-Rank (standard) | 0 | All time points equally |
| Gehan-Breslow-Wilcoxon | -- | Early differences |
| Tarone-Ware | -- | Moderate early weighting |
| Peto-Peto | -- | Early-to-mid differences |

The `survivalTestType` option also controls which test is used for pairwise
comparisons (via `survminer::pairwise_survdiff(rho=...)`).

## 11. Bootstrap Internal Validation

Bootstrap validation provides optimism-corrected estimates of model
discriminative ability. It resamples the data, fits the Cox model on each
bootstrap sample, and evaluates on both the bootstrap and original data to
estimate the optimism bias.

```{r bootstrap}
result_boot <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  bootstrapValidation = TRUE,
  bootstrapValN = 200
)
```

The validation table reports:

| Metric | Apparent | Optimism | Corrected |
|--------|----------|----------|-----------|
| C-index | Training performance | Overfitting bias | Honest estimate |
| Somers' Dxy | 2 * (C-index - 0.5) | ... | ... |
| Calibration slope | Should be near 1 | ... | ... |

**Important notes:**

- `survival::concordance()` treats higher linear predictor as *better*
  prognosis by default. The function uses `reverse = TRUE` because the Cox
  linear predictor has higher values = *worse* prognosis.
- Minimum 50 resamples, maximum 1000. Default 200 provides a reasonable
  bias-variance trade-off.
- This analysis can be computationally intensive for large datasets.

## 12. Calibration Curves

Calibration assesses whether predicted survival probabilities match observed
outcomes. The function groups patients into risk quantiles, computes
Kaplan-Meier survival within each group, and compares it to the model's
predicted survival.

```{r calibration}
result_cal <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  calibration_curves = TRUE,
  calibration_timepoint = 36,
  calibration_ngroups = 5
)
```

The output includes:

1. **Calibration metrics table** -- calibration slope, calibration-in-the-large,
   with ideal values and interpretation
2. **Calibration by risk group table** -- predicted vs. observed survival for
   each risk quantile
3. **Calibration plot** -- predicted vs. observed with 45-degree perfect
   calibration reference line

If `calibration_timepoint` is 0, the median observed follow-up time is used.
Each risk group should have at least 20-30 patients for reliable estimates.

## 13. Non-Linearity Assessment (Restricted Cubic Splines)

For continuous predictors like age, tumor size, or biomarker levels, the
relationship with survival may not be linear on the log-hazard scale. RCS
analysis tests for non-linearity and visualizes the dose-response curve.

```{r rcs}
result_rcs <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  rcs_analysis = TRUE,
  rcs_variable = "biomarker_value",
  rcs_knots = 4
)
```

The output includes:

1. **Non-linearity test table** -- likelihood ratio test comparing the spline
   model (non-linear) against a linear model. A significant p-value indicates
   the relationship is non-linear.
2. **Hazard ratio curve plot** -- HR as a function of the continuous predictor,
   with 95% confidence band. The reference value is typically the median.

Knot recommendations: 3 knots (simplest, 2 df), 4 knots (recommended default,
3 df), 5 knots (more flexible, 4 df). Knots are placed at Harrell-recommended
percentiles.

## 14. Age-Adjusted Analysis

Age is the most important confounder in survival studies. ClinicoPath provides
multiple approaches to handle age.

### 14.1 Age-Adjusted Cox Regression

Includes age as a covariate in the Cox model, producing both unadjusted and
age-adjusted hazard ratios side by side.

```{r age-adjusted}
result_age <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  age_adjustment = TRUE,
  age_variable = "age"
)
```

### 14.2 Age x Group Interaction

Tests whether the treatment effect varies by age.

```{r age-interaction}
result_age_int <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  age_adjustment = TRUE,
  age_variable = "age",
  age_interaction = TRUE
)
```

### 14.3 Age-Stratified KM Plots

Displays Kaplan-Meier curves within age groups defined by cutpoints.

```{r age-stratified-km}
result_age_km <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  age_adjustment = TRUE,
  age_variable = "age",
  age_stratified_km = TRUE,
  age_group_cutpoints = "50, 65, 75"
)
```

### 14.4 Age as Time Scale

Uses biological age rather than follow-up time as the time axis. This is the
most rigorous approach for cancer epidemiology where age is the primary driver
of risk.

```{r age-time-scale}
result_age_ts <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  age_adjustment = TRUE,
  age_variable = "age",
  age_time_scale = TRUE
)
```

## 15. Competing Risks Analysis

When multiple causes of death exist, standard survival analysis can
overestimate the probability of the event of interest. Competing risks
analysis properly accounts for events that preclude observation of the
primary outcome.

```{r competing-risks, eval=FALSE}
# Requires a dataset with multi-level outcome variable
# Example using histopathology dataset:
data(histopathology, package = "ClinicoPath")

result_compete <- survival(
  data = histopathology,
  elapsedtime = "OverallTime",
  outcome = "Outcome2",
  outcomeLevel = "",
  explanatory = "Grade_Level",
  multievent = TRUE,
  dod = "DOD",
  dooc = "DOOC",
  awd = "AWD",
  awod = "AWOD",
  analysistype = "compete",
  sc = TRUE
)
```

**Analysis types:**

- `overall` -- standard overall survival (death from any cause)
- `cause` -- cause-specific survival (censors deaths from other causes)
- `compete` -- competing risks (cumulative incidence functions)

**Important:** When `analysistype = "compete"`, the following analyses are
automatically skipped because they are not valid for competing risks: Cox
regression, pairwise comparisons, RMST, calibration, RCS, bootstrap
validation, and person-time analysis.

## 16. Landmark Analysis

Landmark analysis evaluates survival conditional on surviving past a
specified time point. This is useful for assessing the effect of a
time-varying treatment or exposure.

```{r landmark}
result_landmark <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  uselandmark = TRUE,
  landmark = 12
)
```

All patients who experienced an event or were censored before the landmark
time are excluded. The analysis then proceeds on the remaining cohort.

## 17. Residual Diagnostics

Cox model residuals help identify influential observations and assess model
fit.

```{r residuals}
result_resid <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  residual_diagnostics = TRUE
)
```

The output includes:

- **Residuals table** -- Martingale, deviance, score, and Schoenfeld residuals
- **Diagnostic plot** -- visual assessment of residual patterns

## 18. Clinical Interpretations and Explanations

Two complementary features help users understand their results.

### 18.1 Analysis Explanations

Provides detailed statistical methodology explanations for each output
section (median survival, Cox regression, survival tables, plots, etc.).

```{r explanations}
result_explain <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  showExplanations = TRUE
)
```

### 18.2 Natural Language Summaries

Provides plain-language interpretations of the statistical results, including
clinical interpretation, a glossary of terms, and copy-ready sentences for
reports.

```{r summaries}
result_summary <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  showSummaries = TRUE
)
```

These features produce:

- **Clinical interpretation** -- context-aware summary of findings
- **Clinical glossary** -- definitions of survival analysis terminology
- **Copy-ready sentences** -- publication-ready text for results sections

## 19. REMARK Reporting Checklist

The REMARK (REporting recommendations for tumor MARKer prognostic studies)
checklist helps ensure comprehensive reporting of biomarker-based survival
studies.

```{r remark}
result_remark <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  remark_checklist = TRUE
)
```

The checklist shows which REMARK items are addressed by the current analysis
configuration and which require additional reporting.

## 20. Data Export

Survival estimates can be exported for use in external analyses or
visualization tools.

```{r export}
result_export <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",
  export_survival_data = TRUE
)
```

## 21. Complete Analysis Example

Here is a comprehensive example that enables most features simultaneously.

```{r comprehensive}
result_full <- survival(
  data = survival_test,
  elapsedtime = "elapsedtime",
  outcome = "outcome",
  outcomeLevel = "1",
  explanatory = "treatment",
  dod = "",
  dooc = "",
  awd = "",
  awod = "",

  # Survival probability cutpoints

  cutp = "6, 12, 24, 36, 48, 60",

  # Plots
  sc = TRUE,
  ce = TRUE,
  ch = TRUE,
  kmunicate = TRUE,
  loglog = TRUE,
  ci95 = TRUE,
  risktable = TRUE,
  censored = TRUE,
  pplot = TRUE,
  medianline = "hv",
  endplot = 60,
  byplot = 12,

  # Statistical tests
  pw = TRUE,
  padjustmethod = "holm",
  ph_cox = TRUE,

  # Advanced analysis
  rmst_analysis = TRUE,
  rmst_tau = 48,
  weightedLogRank = TRUE,
  person_time = TRUE,
  time_intervals = "12, 24, 36, 48, 60",
  rate_multiplier = 100,

  # Model validation
  bootstrapValidation = TRUE,
  bootstrapValN = 200,
  calibration_curves = TRUE,
  calibration_timepoint = 36,
  calibration_ngroups = 5,

  # Residual diagnostics
  residual_diagnostics = TRUE,

  # Interpretation aids
  showExplanations = TRUE,
  showSummaries = TRUE,
  remark_checklist = TRUE
)
```

## 22. Summary of All Options

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `elapsedtime` | Variable | -- | Survival time variable |
| `outcome` | Variable | -- | Event indicator |
| `outcomeLevel` | Level | -- | Which level = event |
| `explanatory` | Variable | -- | Grouping variable |
| `cutp` | String | "12, 36, 60" | Cutpoints for survival table |
| `tint` | Bool | FALSE | Calculate time from dates |
| `dxdate` | Variable | -- | Diagnosis date (if tint) |
| `fudate` | Variable | -- | Follow-up date (if tint) |
| `timetypedata` | List | ymd | Date format in data |
| `timetypeoutput` | List | months | Output time unit |
| `uselandmark` | Bool | FALSE | Landmark analysis |
| `landmark` | Integer | 3 | Landmark time |
| `multievent` | Bool | FALSE | Multiple event levels |
| `dod` | Level | -- | Dead of Disease level |
| `dooc` | Level | -- | Dead of Other Causes level |
| `awd` | Level | -- | Alive with Disease level |
| `awod` | Level | -- | Alive without Disease level |
| `analysistype` | List | overall | overall/cause/compete |
| `sc` | Bool | FALSE | Survival curve plot |
| `kmunicate` | Bool | FALSE | KMunicate-style plot |
| `ce` | Bool | FALSE | Cumulative events plot |
| `ch` | Bool | FALSE | Cumulative hazard plot |
| `loglog` | Bool | FALSE | Log-log plot |
| `ci95` | Bool | FALSE | 95% CI on plots |
| `risktable` | Bool | FALSE | Risk table on plots |
| `censored` | Bool | FALSE | Censoring marks |
| `pplot` | Bool | FALSE | P-value on plots |
| `medianline` | List | none | Median line style |
| `endplot` | Integer | 60 | Plot end time |
| `byplot` | Integer | 12 | Time interval for axis |
| `ybegin_plot` | Number | 0.0 | Y-axis start |
| `yend_plot` | Number | 1.0 | Y-axis end |
| `pw` | Bool | FALSE | Pairwise comparisons |
| `padjustmethod` | List | holm | P-value adjustment |
| `ph_cox` | Bool | FALSE | PH assumption test |
| `stratified_cox` | Bool | FALSE | Stratified Cox model |
| `strata_variable` | Variable | -- | Stratification variable |
| `rmst_analysis` | Bool | FALSE | RMST analysis |
| `rmst_tau` | Number | 0 | RMST time horizon |
| `person_time` | Bool | FALSE | Person-time analysis |
| `time_intervals` | String | "12, 36, 60" | Person-time intervals |
| `rate_multiplier` | Integer | 100 | Rate per N person-time |
| `weightedLogRank` | Bool | FALSE | Weighted log-rank tests |
| `survivalTestType` | List | logrank | Test family selection |
| `bootstrapValidation` | Bool | FALSE | Bootstrap validation |
| `bootstrapValN` | Integer | 200 | Number of resamples |
| `calibration_curves` | Bool | FALSE | Calibration assessment |
| `calibration_timepoint` | Number | 0 | Timepoint (0=median) |
| `calibration_ngroups` | Integer | 5 | Risk group count |
| `rcs_analysis` | Bool | FALSE | Non-linearity test |
| `rcs_variable` | Variable | -- | Continuous predictor |
| `rcs_knots` | Integer | 4 | Number of knots |
| `residual_diagnostics` | Bool | FALSE | Model diagnostics |
| `age_adjustment` | Bool | FALSE | Age-adjusted Cox |
| `age_variable` | Variable | -- | Age variable |
| `age_interaction` | Bool | FALSE | Age x group test |
| `age_stratified_cox` | Bool | FALSE | Stratify Cox by age |
| `age_group_cutpoints` | String | "50, 65, 75" | Age group boundaries |
| `age_time_scale` | Bool | FALSE | Age as time axis |
| `age_standardization` | Bool | FALSE | SMR computation |
| `age_standardization_method` | List | indirect | SMR method |
| `age_stratified_km` | Bool | FALSE | Age-stratified KM |
| `adjusted_curves` | Bool | FALSE | Age-adjusted curves |
| `showExplanations` | Bool | FALSE | Method explanations |
| `showSummaries` | Bool | FALSE | Plain-language summaries |
| `remark_checklist` | Bool | FALSE | REMARK checklist |
| `export_survival_data` | Output | -- | Export estimates |

## References

- Klein JP, Moeschberger ML (2003). *Survival Analysis: Techniques for Censored and Truncated Data.* Springer.
- Therneau TM, Grambsch PM (2000). *Modeling Survival Data: Extending the Cox Model.* Springer.
- Royston P, Parmar MK (2013). Restricted mean survival time: an alternative to the hazard ratio. *BMC Medical Research Methodology* 13:152.
- Morris TP et al. (2019). Proposals on Kaplan-Meier plots in medical research and a survey of stakeholder views: KMunicate. *BMJ Open* 9:e030215.
- McShane LM et al. (2005). REporting recommendations for tumour MARKer prognostic studies (REMARK). *British Journal of Cancer* 93:387-391.
