---
title: "Use jsurvival in R"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
    echo = TRUE,
  eval = FALSE
)
```

```{r setup}
library(ClinicoPath)
```


```{r data histopathology}
data("histopathology")
```

```{r}
#| glimpse histopathology

dplyr::glimpse(histopathology)
```


```{r}
ClinicoPath::survival(
    data = histopathology,
    elapsedtime = "OverallTime",
    tint = TRUE,
    dxdate = SurgeryDate,
    fudate = LastFollowUpDate,
    explanatory = LVI,
    outcome = Outcome,
    outcomeLevel = NULL,
    dod = NULL,
    dooc = NULL,
    awd = NULL,
    awod = NULL,
    timetypedata = "ymdhms",
    sc = TRUE,
    ce = TRUE,
    ch = TRUE
    )
```




```{r}
overall_survival <-
ClinicoPath::survival(
    data = histopathology,
    elapsedtime = "OverallTime",
    tint = TRUE,
    dxdate = SurgeryDate,
    fudate = LastFollowUpDate,
    explanatory = LVI,
    outcome = Outcome,
    outcomeLevel = NULL,
    dod = NULL,
    dooc = NULL,
    awd = NULL,
    awod = NULL,
    timetypedata = "ymdhms",
    sc = TRUE,
    ce = TRUE,
    ch = TRUE
    )
```


```{r}
ls(overall_survival)
```



```{r}
overall_survival$tCoxtext2
```


```{r}
overall_survival$survTableSummary
```


```{r}
overall_survival$survTable
```



```{r}
overall_survival$subtitle
```



```{r}
overall_survival$plot3
```



```{r}
overall_survival$plot2
```



```{r}
overall_survival$plot
```



```{r}
overall_survival$medianTable
```


```{r}
overall_survival$medianSummary
```



```{r}
overall_survival$coxTable
```



```{r}
overall_survival$coxTable$asDF
```




```{r}
overall_survival$coxSummary
```




```{r}
overall_survival$plot

```


```{r}
# overall_survival$plot$saveAs("overall_survival.png")
# 
# overall_survival <- png::readPNG("overall_survival.png")
# 
# tiff::writeTIFF(what = overall_survival, where = "overall_survival.tiff", compression = "none")
```
 
 



```{r}
# htmlTable::htmlTable(
#   kableExtra::kable(
#     overall_survival$coxTable$asDF
#     ))


```




All functions

histopathology
	

histopathology

jsurvival-package _PACKAGE jsurvival
	

jsurvival: Analysis for jsurvivalological Research

multisurvival()
	

Multivariable Survival Analysis

multisurvivalClass
	

Multivariable Survival Analysis

oddsratio()
	

Odds Ratio Table and Plot

oddsratioClass
	

Odds Ratio Table and Plot

singlearm()
	

Single Arm Survival

singlearmClass
	

Single Arm Survival

survival()
	

Survival Analysis

survivalClass
	

Survival Analysis

survivalcont()
	

Survival Analysis for Continuous Variable

survivalcontClass
	

Survival Analysis for Continuous Explanatory Variable
Contents

    Reference


