---
title: "SCAD/MCP Cox Regression for Variable Selection in Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCAD/MCP Cox Regression for Variable Selection in Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `ncvregcox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `ncvregcoxClass$new(options = ncvregcoxOptions$new(...), data = mydata)`.

# SCAD/MCP Cox Regression

## Overview

The **SCAD/MCP Cox Regression** module (`ncvregcox`) performs variable selection using
non-convex penalties -- **SCAD** (Smoothly Clipped Absolute Deviation) and **MCP**
(Minimax Concave Penalty) -- for Cox proportional hazards models. These penalties
solve a fundamental problem with LASSO: **over-penalization of large coefficients**.

Key features:

- **Two non-convex penalty types**: SCAD and MCP
- **Oracle properties**: Consistent selection + unbiased estimation for large effects
- **Data suitability assessment** (traffic-light system with 6 checks)
- **Cross-validated lambda selection** (min or 1-SE rule)
- **Regularization path** visualization
- **Variable importance** analysis

---

## Why Non-Convex Penalties?

LASSO applies a constant penalty rate regardless of coefficient size, which means:

- Small (noise) coefficients: shrunk to zero (good)
- Large (true) coefficients: also over-shrunk (bad -- introduces bias)

**SCAD** and **MCP** use penalties that flatten out for large coefficients, solving
this problem. The key insight is that once a coefficient is clearly non-zero, there
is no reason to keep penalizing it.

### Penalty Comparison

| Penalty | Available Here? | Behavior | Bias for Large Effects | Selection |
|---------|:--------------:|----------|----------------------|-----------|
| **SCAD** | Yes | Penalty flattens at $\gamma\lambda$ | Nearly unbiased | Oracle property |
| **MCP** | Yes | Penalty flattens sooner | Nearly unbiased | Oracle property |
| LASSO | No (see `lassocox`) | Constant penalty rate | High bias | Consistent under conditions |
| Ridge | No | Quadratic penalty | Moderate bias | No selection |

> **Note:** LASSO is *not* available in this module. For LASSO-penalized Cox regression,
> use the dedicated **LASSO Cox** (`lassocox`) function, which includes its own suitability
> assessment and is optimized for the L1 penalty. This module focuses exclusively on
> non-convex penalties that offer oracle properties LASSO cannot provide.

### Mathematical Formulation

**SCAD penalty** (Fan & Li, 2001):
$$P(\beta) = \begin{cases}
\lambda |\beta| & \text{if } |\beta| \leq \lambda \\
\frac{-(\beta^2 - 2\gamma\lambda|\beta| + \lambda^2)}{2(\gamma-1)} & \text{if } \lambda < |\beta| \leq \gamma\lambda \\
\frac{(\gamma+1)\lambda^2}{2} & \text{if } |\beta| > \gamma\lambda
\end{cases}$$

**MCP** (Zhang, 2010):
$$P(\beta) = \begin{cases}
\lambda |\beta| - \frac{\beta^2}{2\gamma} & \text{if } |\beta| \leq \gamma\lambda \\
\frac{\gamma\lambda^2}{2} & \text{if } |\beta| > \gamma\lambda
\end{cases}$$

The default $\gamma = 3.7$ for SCAD was recommended by Fan & Li (2001) based on
Bayesian arguments and minimization of a Bayesian risk. For MCP, the recommended
default is $\gamma = 3.0$ (Zhang, 2010).

Both penalties behave like LASSO near zero (good for eliminating noise) but stop
penalizing coefficients once they exceed the threshold $\gamma\lambda$. This
two-regime behavior is what gives them the **oracle property**: under regularity
conditions, they select the correct model and estimate non-zero coefficients as
if the true model were known in advance.

---

## When to Use SCAD/MCP

| Scenario | Recommendation |
|----------|---------------|
| Suspect strong true effects that LASSO would bias | **SCAD or MCP** |
| Building a model for publication (unbiased coefficients) | **SCAD recommended** |
| Want oracle property without two-stage procedure | **SCAD or MCP** |
| Moderate number of predictors | **MCP** (more aggressive selection) |
| Very high dimensional (p >> n) | Consider `lassocox` -- LASSO may be more stable |
| Need guaranteed convexity for optimization | Use `lassocox` |

---

## Datasets Used in This Guide

All datasets ship with the package and can be loaded directly:

| Dataset | N | Covariates | Description |
|---------|---|------------|-------------|
| `ncvregcox_clinical` | 200 | 14 mixed | Clinical study with known effects |
| `ncvregcox_sparse` | 100 | 30 continuous | Many noise variables, few true signals |
| `ncvregcox_small` | 25 | 5 | Small-sample edge case for EPV warnings |
| `ncvregcox_collinear` | 150 | 10 | Extreme multicollinearity (r > 0.9) |

---

## 1. Loading Test Data

```{r load-clinical-data}
# Load the clinical dataset (n=200, 14 covariates)
data(ncvregcox_clinical, package = "ClinicoPath")
str(ncvregcox_clinical)
cat("N:", nrow(ncvregcox_clinical),
    "Events:", sum(ncvregcox_clinical$event == "1", na.rm = TRUE),
    "Event rate:", round(mean(ncvregcox_clinical$event == "1", na.rm = TRUE), 2), "\n")
```

**True effects in `ncvregcox_clinical`:** age (weak), tumor_diameter (strong),
ldh_level (moderate), cea_level (weak-moderate), t_stage T3/T4 (strong),
n_stage N2 (strong). All other variables are noise.

```{r load-sparse-data}
# Load the sparse/high-dimensional dataset (n=100, 30 covariates)
data(ncvregcox_sparse, package = "ClinicoPath")
cat("N:", nrow(ncvregcox_sparse),
    "Events:", sum(ncvregcox_sparse$event == "1", na.rm = TRUE),
    "Covariates:", ncol(ncvregcox_sparse) - 2, "\n")
```

**True effects in `ncvregcox_sparse`:** x1 (+0.8), x5 (-0.6), x12 (+0.7),
x20 (-0.4). The remaining 26 variables are noise with block-correlated structure.

---

## 2. Data Suitability Assessment

Before running the penalized regression, the module performs a comprehensive
suitability check with a traffic-light system (green/yellow/red) across 6 dimensions:

1. **Events-Per-Variable (EPV)** -- are there enough events relative to predictors?
2. **Regularization Need** -- is the dimensionality high enough to warrant penalization?
3. **Sample Size** -- is n adequate for stable cross-validation?
4. **Event Rate** -- is the event/censoring balance reasonable?
5. **Multicollinearity** -- are predictors highly correlated?
6. **Data Quality** -- how much missing data exists?

```{r suitability-clinical}
# Run with suitability check enabled (default)
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin", "wbc_count",
                 "neutrophil_ratio", "platelet_count", "gender",
                 "n_stage", "histology"),
  penalty = "SCAD",
  suitabilityCheck = TRUE,
  cv_folds = 10,
  lambda_type = "min",
  plot_path = FALSE,
  plot_cv = FALSE,
  variable_importance = FALSE
)
```

The suitability report appears as an HTML table with colored dots indicating the
status of each check. For the clinical dataset (n=200, 14 covariates), you should
see mostly green indicators.

---

## 3. SCAD Penalty (Default)

SCAD is the recommended starting point -- it provides nearly unbiased estimates
for truly important variables while setting noise variables to zero.

```{r scad-clinical}
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin", "wbc_count",
                 "neutrophil_ratio", "platelet_count", "gender",
                 "n_stage", "histology"),
  penalty = "SCAD",
  gamma = 3.7,
  alpha = 1.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Expected:** SCAD should select age, tumor_diameter, ldh_level, cea_level, and
t_stage/n_stage. Noise variables should be eliminated with nearly unbiased
coefficient estimates.

---

## 4. MCP Penalty

MCP is more aggressive than SCAD -- it transitions to zero penalty more quickly,
leading to sparser models.

```{r mcp-clinical}
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin", "wbc_count",
                 "neutrophil_ratio", "platelet_count", "gender",
                 "n_stage", "histology"),
  penalty = "MCP",
  gamma = 3.0,
  alpha = 1.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Compare with SCAD:** MCP typically selects fewer variables. If both methods select
the same core set of predictors, this provides stronger evidence that those variables
are truly associated with the outcome.

---

## 5. Gamma Parameter

The gamma parameter controls how quickly the penalty flattens:

- **SCAD**: gamma must be > 2. Default 3.7 (Fan & Li recommendation)
- **MCP**: gamma must be > 1. Default 3.0. Lower gamma = more aggressive selection

```{r gamma-scad-low}
# SCAD with lower gamma (more aggressive -- penalty flattens sooner)
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  gamma = 2.5,
  cv_folds = 10,
  lambda_type = "min",
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r gamma-scad-high}
# SCAD with higher gamma (less aggressive -- closer to LASSO-like behavior)
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  gamma = 6.0,
  cv_folds = 10,
  lambda_type = "min",
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Interpretation:** As gamma increases, SCAD behaves more like LASSO (constant
penalty rate). As gamma decreases toward 2, SCAD becomes more aggressive at
removing the penalty from large coefficients. The default 3.7 is a well-studied
compromise.

---

## 6. Lambda Selection: min vs 1se

The cross-validation procedure identifies the optimal penalty strength (lambda).
Two strategies are available:

- **min**: Lambda that minimizes CV error -- selects more variables
- **1se**: Lambda within 1 SE of minimum -- more parsimonious (Breiman's rule)

```{r lambda-min}
# Minimum CV error -- selects more variables
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  lambda_type = "min",
  cv_folds = 10,
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r lambda-1se}
# One SE rule -- more parsimonious
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  lambda_type = "1se",
  cv_folds = 10,
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Practical tip:** If the 1-SE model drops a variable you consider clinically
important, inspect the CV error plot. If the two lambda values are close,
the difference is likely within noise. If they are far apart, the 1-SE model
is genuinely simpler and may generalize better.

---

## 7. Sparse Scenario (Many Noise Variables)

The `ncvregcox_sparse` dataset has 30 covariates but only 4 true signals.
This is where non-convex penalties shine -- SCAD/MCP should recover the true
model more accurately than LASSO would.

```{r sparse-scad}
data(ncvregcox_sparse, package = "ClinicoPath")

# SCAD should recover x1, x5, x12, x20
ncvregcox(
  data = ncvregcox_sparse,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = paste0("x", 1:30),
  penalty = "SCAD",
  gamma = 3.7,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r sparse-mcp}
# MCP comparison -- often sparser than SCAD
ncvregcox(
  data = ncvregcox_sparse,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = paste0("x", 1:30),
  penalty = "MCP",
  gamma = 3.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = FALSE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**What to look for:** Both SCAD and MCP should identify x1, x5, x12, and x20 as
the important variables. Any noise variables that sneak in are false positives.
The sparse dataset also has correlated blocks (x1-x5 with rho=0.4), so some
leakage from x1 to x2-x4 is possible.

---

## 8. Elastic Net Mixing

Combine SCAD/MCP with an L2 (ridge-like) penalty by setting alpha < 1. This
can stabilize selection when predictors are correlated.

```{r elastic-net}
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "bmi", "tumor_diameter", "ldh_level", "cea_level",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  alpha = 0.7,  # Mix with L2 penalty
  cv_folds = 10,
  lambda_type = "min",
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

> **Note:** `ncvreg` requires alpha > 0. Pure ridge (alpha = 0) is not supported.
> Values between 0.5 and 1.0 provide a useful range of sparsity/stability tradeoffs.

---

## 9. Standardization

When covariates are on different scales (e.g., age in years vs. LDH in U/L),
standardization ensures the penalty is applied fairly across all variables.

```{r no-standardize}
# Without standardization (not recommended for mixed-scale variables)
ncvregcox(
  data = ncvregcox_clinical,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "tumor_diameter", "ldh_level", "cea_level"),
  penalty = "SCAD",
  standardize = FALSE,
  cv_folds = 10,
  lambda_type = "min",
  suitabilityCheck = FALSE,
  variable_importance = TRUE
)
```

**Caution:** Without standardization, variables with larger numeric ranges (like
LDH ~ 80-600) will be penalized differently from variables with smaller ranges
(like tumor diameter ~ 0.3-20). The default `standardize = TRUE` is almost
always the correct choice.

---

## 10. Edge Cases

### Small Sample (n=25)

The `ncvregcox_small` dataset tests how the module handles very small samples.
The suitability assessment should flag concerns about sample size and EPV.

```{r edge-small}
data(ncvregcox_small, package = "ClinicoPath")

cat("N:", nrow(ncvregcox_small),
    "Events:", sum(ncvregcox_small$event == "1", na.rm = TRUE),
    "Covariates:", ncol(ncvregcox_small) - 2, "\n")

ncvregcox(
  data = ncvregcox_small,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = c("age", "marker1", "marker2", "marker3", "grade"),
  penalty = "SCAD",
  gamma = 3.7,
  cv_folds = 5,        # Reduced folds for small sample
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Expected suitability flags:**

- **Sample Size**: Yellow or red (n=25 is small)
- **EPV**: Likely yellow (few events relative to 5 covariates)
- **Regularization Need**: Yellow (low dimensionality -- standard Cox may suffice)

With only 25 observations, cross-validation folds should be reduced (e.g., 5
instead of 10) to ensure each fold has enough events for reliable estimation.

### Extreme Multicollinearity

The `ncvregcox_collinear` dataset has pairs of variables with correlations
exceeding 0.9 (x1/x2 have r ~ 0.995). This is a known challenge for
penalized methods.

```{r edge-collinear}
data(ncvregcox_collinear, package = "ClinicoPath")

cat("N:", nrow(ncvregcox_collinear),
    "Events:", sum(ncvregcox_collinear$event == "1", na.rm = TRUE), "\n")

# Check the correlation structure
num_cols <- ncvregcox_collinear[, paste0("x", 1:10)]
cor_mat <- cor(num_cols, use = "pairwise.complete.obs")
cat("Max pairwise |r|:", round(max(abs(cor_mat[upper.tri(cor_mat)])), 3), "\n")
```

```{r edge-collinear-scad}
# SCAD under extreme collinearity
ncvregcox(
  data = ncvregcox_collinear,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = paste0("x", 1:10),
  penalty = "SCAD",
  gamma = 3.7,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r edge-collinear-mcp}
# MCP under extreme collinearity
ncvregcox(
  data = ncvregcox_collinear,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = paste0("x", 1:10),
  penalty = "MCP",
  gamma = 3.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r edge-collinear-elastic}
# Elastic net mixing can stabilize selection under collinearity
ncvregcox(
  data = ncvregcox_collinear,
  time = "time",
  event = "event",
  outcomeLevel = "1",
  censorLevel = "0",
  covariates = paste0("x", 1:10),
  penalty = "SCAD",
  alpha = 0.5,     # Strong L2 component for stability
  gamma = 3.7,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  suitabilityCheck = FALSE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Expected suitability flags:**

- **Multicollinearity**: Red (max |r| > 0.9)
- The suitability report will recommend considering Elastic Net (alpha < 1)

**Key observation:** Under extreme collinearity, SCAD and MCP may arbitrarily
swap which of the correlated variables (x1 vs x2 vs x3) gets selected. The
*group* is correctly identified as important, but the specific member chosen
is unstable. This is a fundamental limitation of any variable selection method
when predictors are nearly identical. Elastic net mixing (alpha < 1) can help
by distributing the coefficient across correlated predictors rather than
forcing a single winner.

---

## Interpreting Results

### Selected Variables Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name |
| Coefficient | Log hazard ratio (nearly unbiased with SCAD/MCP) |
| Hazard Ratio | exp(coefficient) |

### Key Advantage Over LASSO

For a true coefficient of 0.8:

- LASSO might estimate 0.5 (biased downward)
- SCAD/MCP might estimate 0.78 (nearly unbiased)

This matters for hazard ratio interpretation in clinical settings. If tumor
diameter truly doubles the hazard (HR = 2.0, beta = 0.69), LASSO might report
HR = 1.5 while SCAD reports HR = 1.95. The SCAD estimate is more trustworthy
for clinical decision-making.

### Variable Importance

Variables are ranked by absolute coefficient value. With SCAD/MCP, the ranking
is more reliable than LASSO because coefficients are less biased.

### Suitability Assessment

The traffic-light table provides a quick pre-analysis diagnostic:

| Color | Meaning |
|-------|---------|
| Green | No concerns -- proceed with confidence |
| Yellow | Usable but review the flagged item |
| Red | Potential issue -- results may be unreliable |

---

## SCAD vs MCP: Which to Choose?

| Criterion | SCAD | MCP |
|-----------|------|-----|
| Bias for large effects | Very low | Very low |
| Selection aggressiveness | Moderate | More aggressive |
| Stability | Good | Slightly less stable |
| Recommended gamma | 3.7 (default) | 3.0 (default) |
| Publication standard | More established | Gaining adoption |
| Convergence | Generally reliable | Can be faster |

**Practical recommendation**: Start with SCAD (gamma=3.7). Switch to MCP if
you want sparser selection. If both methods agree on the selected variables,
you have strong evidence of a robust signal.

---

## Common Pitfalls

1. **Setting gamma too low for SCAD**: gamma < 2.5 makes optimization unstable.
   Stick with the default 3.7 unless you have a specific reason to change it.

2. **Comparing SCAD/MCP coefficients directly to LASSO**: SCAD/MCP coefficients
   are less biased. A SCAD coefficient of 0.8 is not "bigger" than a LASSO
   coefficient of 0.5 for the same variable -- the LASSO version is just more shrunk.

3. **Not reporting the penalty type**: Always specify whether you used SCAD or MCP,
   along with the gamma value and lambda selection method.

4. **Using SCAD/MCP for very high dimensional data (p >> 10n)**: In extremely
   high-dimensional settings, LASSO or Elastic Net may be more stable.
   SCAD/MCP work best when p is moderately large relative to n.

5. **Ignoring the suitability assessment**: The traffic-light report is there for
   a reason. A red flag on multicollinearity or sample size means the selected
   model may not be trustworthy. Consider the suggested remedies (reduce
   covariates, use elastic net mixing, increase sample size).

6. **Post-selection inference**: Coefficients from SCAD/MCP are *selected*
   coefficients. Standard confidence intervals and p-values from a refit Cox
   model on the selected variables are **not valid** for inference because the
   selection step was not accounted for. If formal inference is needed, consider
   methods such as selective inference (Lee et al., 2016) or sample splitting.
   The coefficients reported here are appropriate for prediction and variable
   ranking, but should not be interpreted as confirmed causal effects without
   further validation.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** (`lassocox`) | Standard LASSO with suitability assessment |
| **Adaptive LASSO** (`adaptivelasso`) | Two-stage adaptive weights approach |
| **High-Dimensional Cox** (`highdimcox`) | Multiple regularization methods unified |
| **PLS Cox** (`plscox`) | Dimensionality reduction for very high p |
| **Multivariable Survival** | Standard Cox, no regularization needed |

---

## References

- Fan J, Li R. Variable selection via nonconcave penalized likelihood and its oracle properties. *J Am Stat Assoc*. 2001;96(456):1348-1360.
- Zhang CH. Nearly unbiased variable selection under minimax concave penalty. *Ann Stat*. 2010;38(2):894-942.
- Breheny P, Huang J. Coordinate descent algorithms for nonconvex penalized regression, with applications to biological feature selection. *Ann Appl Stat*. 2011;5(1):232-253.
- Fan J, Li R. Variable selection for Cox's proportional hazards model and frailty model. *Ann Stat*. 2002;30(1):74-99.
- Lee JD, Sun DL, Sun Y, Taylor JE. Exact post-selection inference, with application to the lasso. *Ann Stat*. 2016;44(3):907-927.
