---
title: "Adaptive LASSO Cox Regression for Variable Selection in Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive LASSO Cox Regression for Variable Selection in Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `adaptivelasso()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `adaptivelassoClass$new(options = adaptivelassoOptions$new(...), data = mydata)`.

# Adaptive LASSO Cox Regression

## Overview

The **Adaptive LASSO Cox Regression** module (`adaptivelasso`) performs penalized
variable selection for Cox proportional hazards models using data-driven adaptive
weights. Unlike standard LASSO, the adaptive LASSO assigns differential penalties
to each coefficient based on an initial estimate of their importance. This achieves
the **oracle property** -- meaning that asymptotically, it selects the true model
as if the true sparse structure were known in advance.

This matters in practice because standard LASSO can over-shrink large coefficients
and under-shrink small ones, producing biased estimates. The adaptive variant
corrects this by penalizing unimportant variables more aggressively while leaving
truly important predictors relatively unpenalized, yielding sparser models with
less bias.

Key capabilities:

- **Five weight methods** (ridge, univariate Cox, full Cox, marginal correlation, equal)
- **Elastic net mixing** for handling correlated predictors
- **Cross-validated lambda selection** with deviance or C-index criteria
- **Stability selection** via bootstrap for robust variable identification
- **Proportional hazards testing** and influence diagnostics
- **Risk group stratification** with survival predictions
- **Six visualization types** (regularization path, CV curve, stability, survival curves, baseline hazard, diagnostics)

---

## What is Adaptive LASSO?

### Mathematical Formulation

The Cox proportional hazards model is:

$$h(t) = h_0(t) \exp(\beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p)$$

**Standard LASSO** penalizes all coefficients equally:

$$\text{Maximize:} \quad \ell(\beta) - \lambda \sum_{j=1}^{p} |\beta_j|$$

**Adaptive LASSO** uses data-driven weights $\hat{w}_j$:

$$\text{Maximize:} \quad \ell(\beta) - \lambda \sum_{j=1}^{p} \hat{w}_j |\beta_j|$$

where $\hat{w}_j = 1 / |\hat{\beta}_j^{init}|^{\gamma}$ and $\hat{\beta}_j^{init}$
is an initial consistent estimator of $\beta_j$. The power parameter $\gamma > 0$
controls how aggressively the weights differentiate between important and unimportant
predictors.

### Why Adaptive Weights Matter

The key insight is that variables with large initial coefficient estimates receive
*smaller* penalties (low $\hat{w}_j$), while variables with near-zero initial
estimates receive *larger* penalties (high $\hat{w}_j$). This asymmetry is what
grants the oracle property.

### Weight Methods Comparison

| Method | Initial Estimator | When to Use |
|--------|-------------------|-------------|
| **Ridge** (default) | Ridge-penalized Cox coefficients | High-dimensional data (p > n), correlated predictors |
| **Univariate Cox** | Individual Cox regressions per variable | Moderate dimensions, uncorrelated predictors |
| **Full Cox** | Standard multivariable Cox model | Low dimensions (p << n), well-conditioned design |
| **Marginal Correlation** | Correlation with survival outcome | Quick screening, very large p |
| **Equal** (Standard LASSO) | All weights = 1 (no adaptation) | Baseline comparison, when initial estimates are unreliable |

### LASSO vs Adaptive LASSO

| Property | Standard LASSO | Adaptive LASSO |
|----------|---------------|----------------|
| Variable selection consistency | No (in general) | Yes (oracle property) |
| Coefficient bias | Systematically biased | Reduced bias for true signals |
| Penalty structure | Uniform across all variables | Differential, data-driven |
| Computational cost | Lower (single-stage) | Higher (two-stage: initial + penalized) |
| Best for | Quick screening, exploratory | Confirmatory analysis, final models |

---

## Datasets Used in This Guide

The package includes three test datasets designed for adaptive LASSO analysis:

| Dataset | N | Events | Event Rate | Predictors | Strata | Description |
|---------|---|--------|------------|------------|--------|-------------|
| `adaptivelasso_test_data` | 180 | 122 | 68% | 12 (6 numeric + 6 factor) | `center` (3 levels) | Standard clinical scenario |
| `adaptivelasso_small_data` | 40 | 11 | 28% | 6 (2 numeric + 4 factor) | `center` (2 levels) | Small sample / edge case |
| `adaptivelasso_highdim_data` | 80 | 61 | 76% | 17 (9 numeric + 8 factor) | -- | High-dimensional with genomic features |

---

## 1. Basic Usage (Default Settings)

### 1.1 Minimal Example

A minimal adaptive LASSO analysis requires only the time variable, event indicator,
event level, and predictor variables. All other options use sensible defaults
(ridge weights, alpha=1.0, 10-fold CV, deviance criterion).

```{r basic-minimal}
data("adaptivelasso_test_data", package = "ClinicoPath")

adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  suitabilityCheck = TRUE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

The suitability assessment (traffic-light system) evaluates:

- **Events per variable (EPV)**: Are there enough events relative to the number of predictors?
- **Sample size adequacy**: Is n large enough for stable estimation?
- **Event rate**: Are events too rare or too common?
- **Missing data**: Are there problematic amounts of missingness?
- **Multicollinearity**: Are predictors highly correlated?
- **Variable types**: Are there enough numeric predictors for penalized regression?

### 1.2 With Stratification

When there is a known grouping variable (e.g., treatment center, study site) that
should not be penalized but accounted for as a stratum, use the `strata` option.
This fits separate baseline hazards for each stratum while sharing the regression
coefficients.

```{r basic-stratified}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  strata = "center",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

Stratification is particularly useful when the center effect is a nuisance variable
you want to adjust for without estimating its coefficient.

---

## 2. Weight Methods Comparison

The choice of weight method determines the initial coefficient estimates used to
calculate adaptive penalties. This section compares all five methods on the same
dataset. Look at which variables are selected and the magnitude of adaptive weights
to understand how each method behaves.

### 2.1 Ridge Weights (Default)

Ridge regression always produces non-zero coefficients, making it a stable source
of initial weights even in high-dimensional or correlated settings.

```{r weight-ridge}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  weight_method = "ridge",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 2.2 Univariate Cox Weights

Each predictor is fitted in a separate univariate Cox model. Variables with strong
marginal associations get small weights (less penalized), while marginally weak
predictors get large weights. This does not account for confounding between variables.

```{r weight-univariate}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  weight_method = "univariate",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 2.3 Full Cox Weights

Uses the full multivariable Cox model for initial estimates. This accounts for
confounding but can be unstable when p is close to n or when there is
multicollinearity. Best reserved for low-dimensional problems.

```{r weight-cox}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  weight_method = "cox",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 2.4 Marginal Correlation Weights

Uses the correlation between each predictor and the survival outcome (via
concordance or similar measure). Fast to compute, suitable as a rough screen in
very high-dimensional settings.

```{r weight-correlation}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  weight_method = "correlation",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 2.5 Equal Weights (Standard LASSO)

Setting weights to equal recovers the standard LASSO -- no adaptation. This is
useful as a baseline comparison to see how much the adaptive weighting changes
the variable selection.

```{r weight-equal}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  weight_method = "equal",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**What to compare across methods:** The "Adaptive Weight" column in the coefficients
table shows how aggressively each variable is penalized. Variables with small weights
are protected from shrinkage; variables with large weights are strongly penalized.
Different weight methods may select different variables -- this is expected and
informative about which predictors are robust across estimation strategies.

---

## 3. Elastic Net and Adaptive Penalty

### 3.1 Alpha Sensitivity

The `alpha` parameter controls the mix between L1 (LASSO) and L2 (ridge) penalties.
Values between 0 and 1 give elastic net behavior, which is helpful when predictors
are correlated and you want grouped selection.

**Alpha = 1.0 (Pure LASSO):** Maximum sparsity. Variables are either in or out.

```{r alpha-1}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  alpha = 1.0,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Alpha = 0.5 (Elastic Net):** Compromise between LASSO and ridge. Correlated
predictors tend to be selected or excluded together.

```{r alpha-05}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  alpha = 0.5,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Alpha = 0.0 (Pure Ridge):** No variable selection -- all coefficients are
shrunk but none set exactly to zero. This is not typically the goal of adaptive
LASSO, but it is useful as a comparison.

```{r alpha-0}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  alpha = 0.0,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 3.2 Gamma Sensitivity

The `gamma` parameter controls how sharply the adaptive weights differentiate
between strong and weak predictors. Higher gamma gives more aggressive selection.

**Gamma = 0.5 (Mild adaptation):** Weights are less extreme. More variables tend
to survive selection.

```{r gamma-05}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  gamma = 0.5,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Gamma = 1.0 (Default):** Standard adaptive LASSO.

```{r gamma-1}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  gamma = 1.0,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Gamma = 2.0 (Strong adaptation):** Weights are very extreme. Only the strongest
predictors survive; marginal predictors are aggressively excluded.

```{r gamma-2}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  gamma = 2.0,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Guidance:** Use gamma=0.5 for exploratory analysis (retain more variables),
gamma=1.0 as the default, and gamma=2.0 when you need a very sparse final model.
In high-dimensional settings, larger gamma helps control false selections.

---

## 4. Cross-Validation Options

### 4.1 Different CV Measures

The cross-validation criterion determines what the model optimizes during lambda
selection. The two available measures are partial likelihood deviance and
Harrell's C-index.

**Deviance (default):** Optimizes the partial likelihood. This is the standard
choice and is computationally efficient.

```{r cv-deviance}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  cv_folds = 10,
  cv_measure = "deviance",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = TRUE
)
```

**C-index (Harrell's concordance):** Optimizes discrimination ability -- the
probability that for a random pair of patients, the one with the higher risk score
has the earlier event. This focuses on ranking rather than calibration.

```{r cv-cindex}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  cv_folds = 10,
  cv_measure = "C",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = TRUE
)
```

**Choosing between deviance and C-index:**

- Use **deviance** (default) for general-purpose model building and when calibration matters.
- Use **C-index** when discrimination (ranking patients correctly) is the primary goal, such as for prognostic risk scores.

### 4.2 Custom Lambda Sequence

Instead of the automatic lambda sequence, you can specify a custom range. This is
useful when you want to explore a specific penalty region or when the automatic
range does not cover the optimal region well.

```{r cv-custom-lambda}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  lambda_sequence = "custom",
  lambda_custom_max = 0.5,
  lambda_custom_min = 0.0001,
  n_lambda = 50,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_selection_path = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

### 4.3 Single Lambda

Fix the penalty parameter to a single value. This bypasses cross-validation and
fits the model at exactly the specified lambda. Useful for comparing models at
specific penalty levels, or when you have a pre-determined lambda from prior
analysis.

```{r cv-single-lambda}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  lambda_sequence = "single",
  lambda_single = 0.01,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 4.4 Varying the Number of CV Folds

More folds give better bias-variance trade-off in the CV estimate but increase
computation time. Fewer folds (e.g., 5) are faster; more folds (e.g., 15-20)
are more stable.

```{r cv-5fold}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  cv_folds = 5,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 4.5 Lambda Path Resolution

The `n_lambda` parameter controls how many lambda values are evaluated along the
regularization path. More values give a smoother path but take longer. The
`lambda_min_ratio` parameter controls how far down the path extends.

```{r cv-fine-path}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  n_lambda = 200,
  lambda_min_ratio = 0.0001,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = TRUE,
  plot_cv_curve = FALSE
)
```

---

## 5. Stability Selection

Stability selection runs the adaptive LASSO on many bootstrap subsamples and
tracks how frequently each variable is selected. Variables selected consistently
across subsamples are considered "stably selected" and are more likely to be truly
important. This provides stronger evidence than a single-run selection.

### 5.1 Basic Stability Selection

```{r stability-basic}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  stability_selection = TRUE,
  stability_threshold = 0.6,
  bootstrap_samples = 100,
  subsampling_ratio = 0.8,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_stability = TRUE
)
```

The stability results table shows:

- **Selection Frequency**: How often each variable was selected across bootstrap samples (0 to 1).
- **Stability Score**: A normalized score reflecting selection reliability.
- **Stable Selection**: "Yes" if the selection frequency exceeds the threshold, "No" otherwise.
- **Error Bound**: An upper bound on the expected number of false selections (Meinshausen and Buhlmann, 2010).

### 5.2 Varying Threshold

A higher threshold is more conservative -- fewer variables are declared "stably
selected" but you have more confidence in those that are. A lower threshold is
more liberal.

**Conservative threshold (0.9):**

```{r stability-conservative}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  stability_selection = TRUE,
  stability_threshold = 0.9,
  bootstrap_samples = 100,
  subsampling_ratio = 0.8,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_stability = TRUE
)
```

**Liberal threshold (0.5):**

```{r stability-liberal}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  stability_selection = TRUE,
  stability_threshold = 0.5,
  bootstrap_samples = 100,
  subsampling_ratio = 0.8,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_stability = TRUE
)
```

**Practical guidance:** A threshold of 0.6 is a reasonable default. For
confirmatory analysis or when false selections are costly, use 0.8--0.9. For
exploratory screening, 0.5 is acceptable.

---

## 6. Model Diagnostics

### 6.1 Proportional Hazards Test

The Cox model assumes that hazard ratios are constant over time. The proportional
hazards (PH) test evaluates this assumption for each selected variable using
scaled Schoenfeld residuals.

```{r diag-ph}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  proportional_hazards = TRUE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = TRUE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

A significant p-value (< 0.05) in the PH test suggests that the effect of that
variable changes over time. If the PH assumption is violated, consider
stratification by that variable or a time-varying coefficient approach.

### 6.2 Influence Diagnostics

Influence diagnostics identify observations that have an outsized effect on the
model estimates. This includes dfbeta values (change in each coefficient when an
observation is removed) and leverage measures.

```{r diag-influence}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  influence_diagnostics = TRUE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = TRUE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_diagnostics = TRUE
)
```

### 6.3 Goodness of Fit

The goodness-of-fit assessment reports overall model performance including
concordance statistics (C-index), model deviance, and calibration measures. This
tells you how well the model discriminates between patients and how well-calibrated
the predictions are.

```{r diag-gof}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  goodness_of_fit = TRUE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = TRUE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 6.4 All Diagnostics Combined

For a comprehensive diagnostic assessment, enable all three diagnostic options
together.

```{r diag-all}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  proportional_hazards = TRUE,
  influence_diagnostics = TRUE,
  goodness_of_fit = TRUE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = TRUE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_diagnostics = TRUE
)
```

---

## 7. Risk Groups and Predictions

### 7.1 Risk Stratification

The adaptive LASSO model computes a linear predictor (risk score) for each patient.
Patients are then divided into risk groups based on quantiles of this score. The
risk group table shows the number of subjects, events, median survival, and hazard
ratios relative to the lowest-risk group.

**Three risk groups (default):**

```{r risk-3groups}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  risk_groups = 3,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_survival_curves = TRUE
)
```

**Two risk groups (high vs low):**

```{r risk-2groups}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  risk_groups = 2,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_survival_curves = TRUE
)
```

**Five risk groups (fine-grained stratification):**

```{r risk-5groups}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  risk_groups = 5,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_survival_curves = TRUE
)
```

### 7.2 Survival Predictions at Time Points

The `time_points` option specifies at which time points to estimate the baseline
survival probability. These predictions are useful for communicating prognosis in
clinical terms ("the 5-year survival probability is...").

```{r predictions-timepoints}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  baseline_survival = TRUE,
  time_points = "5, 10, 15, 20, 25",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE,
  plot_baseline_hazard = TRUE
)
```

The predictions table shows the baseline survival estimate with 95% confidence
intervals at each requested time point. Individual patient survival can be derived
by combining the baseline survival with their risk score:

$$S(t | x) = S_0(t)^{\exp(\hat{\beta}^T x)}$$

---

## 8. Visualization Options

### 8.1 All Plots Enabled

This example enables all six plot types simultaneously. In practice you would
typically enable only the plots relevant to your analysis stage.

```{r plots-all}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade"),
  stability_selection = TRUE,
  bootstrap_samples = 50,
  proportional_hazards = TRUE,
  influence_diagnostics = TRUE,
  goodness_of_fit = TRUE,
  baseline_survival = TRUE,
  time_points = "5, 10, 20",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_selection_path = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE,
  plot_stability = TRUE,
  plot_survival_curves = TRUE,
  plot_baseline_hazard = TRUE,
  plot_diagnostics = TRUE
)
```

**What each plot shows:**

| Plot | Content |
|------|---------|
| **Regularization Path** | Coefficient trajectories vs. log(lambda). Shows which variables enter/leave the model as penalty changes. |
| **CV Curve** | Cross-validation error vs. log(lambda) with confidence bands. Vertical lines mark lambda.min and lambda.1se. |
| **Stability Selection** | Bar chart of selection frequencies across bootstrap samples. Variables above the threshold line are stably selected. |
| **Risk Group Survival** | Kaplan-Meier curves stratified by adaptive LASSO risk groups. Well-separated curves indicate good discrimination. |
| **Baseline Hazard** | Estimated baseline hazard and cumulative hazard from the Breslow estimator. Helps assess the temporal risk pattern. |
| **Diagnostics** | Residual plots, influential observations, and PH assessment. Identify model violations and outliers. |

---

## 9. Advanced Options

### 9.1 Tied Times Methods

Tied survival times (multiple events at the same time) require approximation in
the partial likelihood. Two methods are available.

**Breslow (default):** Faster, adequate when ties are not excessive.

```{r tie-breslow}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  tie_method = "breslow",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Efron:** Better approximation to the exact partial likelihood, especially when
there are many tied event times. Slightly more computationally intensive.

```{r tie-efron}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  tie_method = "efron",
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**When to use Efron:** If your data has many tied event times (e.g., survival
measured in months rather than days), Efron provides a better approximation. For
continuous time measurements with few ties, both methods give nearly identical
results.

### 9.2 Standardization

Standardizing predictors before fitting ensures that the penalty is applied
fairly across variables measured on different scales. This is strongly recommended
when variables have different units (e.g., age in years vs. tumor size in cm vs.
Ki-67 percentage).

**With standardization (default, recommended):**

```{r standardize-on}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  standardize = TRUE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**Without standardization (use with caution):**

```{r standardize-off}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  standardize = FALSE,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

Without standardization, variables on larger scales (e.g., age 30--90) will be
penalized more heavily than variables on smaller scales (e.g., albumin 2--5).
Only disable standardization if all predictors are already on the same scale.

### 9.3 Convergence Settings

The coordinate descent algorithm used internally has convergence and iteration
parameters. The defaults are adequate for most problems. Adjust only when you
receive convergence warnings.

**Tighter convergence (more precise but slower):**

```{r convergence-tight}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  convergence_threshold = 1e-10,
  max_iterations = 50000,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

**More relaxed convergence (faster, for exploration):**

```{r convergence-relaxed}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  convergence_threshold = 1e-5,
  max_iterations = 5000,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = FALSE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

### 9.4 Reproducibility (Random Seed)

The `random_seed` option ensures that cross-validation folds and bootstrap samples
are identical across runs. Always report the seed used in publications.

```{r seed-42}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  random_seed = 42,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

```{r seed-999}
# Different seed may produce slightly different fold assignments and results
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin"),
  random_seed = 999,
  suitabilityCheck = FALSE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_selection_path = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

If the same variables are selected regardless of the seed, that is strong evidence
of genuine signal. If selection changes substantially across seeds, consider using
stability selection for more robust results.

---

## 10. Edge Cases

### 10.1 Small Sample

The small dataset (n=40, 11 events) tests the function's behavior under
challenging conditions: low EPV, few events, and limited statistical power.

```{r edge-small}
data("adaptivelasso_small_data", package = "ClinicoPath")

adaptivelasso(
  data = adaptivelasso_small_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "biomarker_a", "biomarker_b",
                 "gender", "stage"),
  strata = "center",
  suitabilityCheck = TRUE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

Expected suitability flags:

- **EPV**: 11 events / 6 predictors = 1.8 (likely red -- very low)
- **Sample size**: n=40 (likely yellow or red)
- **Event rate**: 28% (acceptable but low event count)

With such small samples, results should be interpreted with extreme caution. The
adaptive LASSO may select zero variables (empty model) as the most honest answer,
or it may select one or two variables with wide confidence intervals.

### 10.2 High-Dimensional Data

The high-dimensional dataset (n=80, 17 predictors including genomic features)
tests the scenario where the number of candidate predictors is large relative to
the sample size.

```{r edge-highdim}
data("adaptivelasso_highdim_data", package = "ClinicoPath")

adaptivelasso(
  data = adaptivelasso_highdim_data,
  time = "time",
  event = "event",
  event_level = "Dead",
  predictors = c("gene_a", "gene_b", "gene_c", "gene_d",
                 "crp", "il6", "tnf", "fibrinogen",
                 "molecular_subtype", "grade", "lymph_node",
                 "margin_status", "receptor_status",
                 "bmi", "hemoglobin", "platelet", "ldh"),
  weight_method = "ridge",
  gamma = 1.0,
  suitabilityCheck = TRUE,
  stability_selection = TRUE,
  bootstrap_samples = 50,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = FALSE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE,
  plot_stability = TRUE
)
```

In high-dimensional settings, ridge weights are recommended because they always
produce non-zero initial estimates. Univariate or full Cox weights may be unstable.
Stability selection adds an extra layer of robustness for identifying truly
important features.

### 10.3 Data Suitability Assessment

The suitability check runs automated assessments and produces a traffic-light
report (green/yellow/red). This example demonstrates the suitability check with
many predictors.

```{r edge-suitability}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size", "ki67_index", "hemoglobin",
                 "crp_level", "albumin", "gender", "stage", "grade",
                 "treatment", "smoking_status"),
  suitabilityCheck = TRUE,
  show_coefficients = FALSE,
  show_selection_path = FALSE,
  show_cv_results = FALSE,
  show_diagnostics = FALSE,
  plot_selection_path = FALSE,
  plot_cv_curve = FALSE
)
```

The suitability assessment never blocks the analysis -- it is advisory. Even with
red flags, the model will still run, but results should be interpreted accordingly.

### 10.4 Minimal Predictors

When only two or three predictors are available, the adaptive LASSO may not add
much over standard Cox regression. This tests the boundary case.

```{r edge-minimal-predictors}
adaptivelasso(
  data = adaptivelasso_test_data,
  time = "time",
  event = "event",
  event_level = "Event",
  predictors = c("age", "tumor_size"),
  suitabilityCheck = TRUE,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  show_diagnostics = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

---

## Interpreting Results

### Coefficients Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name (with dummy level for factors) |
| Coefficient | Log hazard ratio (positive = increased risk, negative = protective) |
| Hazard Ratio | exp(coefficient) -- multiplicative effect on hazard per unit change |
| SE | Standard error from post-selection unpenalized Cox refit |
| Lower CI / Upper CI | 95% confidence interval for the hazard ratio |
| Adaptive Weight | The penalty weight applied to this variable (smaller = less penalized) |

### Cross-Validation Results

| Column | Meaning |
|--------|---------|
| Lambda Min | Lambda with minimum CV error (retains more variables) |
| Lambda 1SE | Most parsimonious lambda within 1 SE of minimum (recommended default) |
| Min CV Error | CV error at lambda.min |
| 1SE CV Error | CV error at lambda.1se |
| Variables (Min) | Number of selected variables at lambda.min |
| Variables (1SE) | Number of selected variables at lambda.1se |

### C-index Interpretation

| C-index | Discrimination | Clinical Utility |
|---------|---------------|-----------------|
| 0.50--0.60 | Poor | Limited utility |
| 0.60--0.70 | Fair | May inform decisions |
| 0.70--0.80 | Good | Useful for stratification |
| 0.80--0.90 | Excellent | Strong clinical utility |
| > 0.90 | Outstanding | May indicate overfitting |

---

## Common Pitfalls and Best Practices

### Pitfalls to Avoid

1. **Choosing weight method without considering data dimensions.** Full Cox weights
   (`cox`) fail or are unstable when p is close to or exceeds n. Use ridge weights
   as the safe default.

2. **Ignoring the oracle property's asymptotic nature.** The oracle property is an
   asymptotic result. In small samples, adaptive LASSO may not outperform standard
   LASSO. Use stability selection to assess robustness.

3. **Over-interpreting post-selection p-values.** Confidence intervals and p-values
   from the refitted unpenalized Cox model are conditional on the selected variables
   and tend to be anti-conservative. They are informative but should not be treated
   as formal hypothesis tests.

4. **Setting gamma too high with noisy initial estimates.** If the initial estimates
   are unreliable (small sample, many predictors), a high gamma amplifies this noise.
   Start with gamma=1.0 and only increase after verifying that initial weights are
   sensible.

5. **Forgetting to standardize.** Without standardization, the penalty is applied
   unevenly across variables measured on different scales. Always keep
   `standardize = TRUE` unless you have a specific reason not to.

6. **Not checking the proportional hazards assumption.** A penalized Cox model still
   assumes proportional hazards. Violations can invalidate the selected model.

7. **Feature selection on the full dataset (data leakage).** If you select features
   on all patients, then validate on the same data, performance estimates are biased.
   Feature selection should be done within the training set only. Cross-validation
   within the adaptive LASSO partially addresses this, but external validation is
   still needed for unbiased assessment.

### Best Practices

1. **Use ridge weights as the default** -- they are stable across data dimensions.
2. **Always run suitability check** before interpreting results.
3. **Use stability selection** for confirmatory analysis to strengthen variable
   selection evidence.
4. **Compare weight methods** -- if the same variables are selected across methods,
   you have stronger evidence.
5. **Report the full specification:** weight method, alpha, gamma, lambda selection
   rule, CV folds, seed.
6. **Validate externally** when possible. Internal CV performance is optimistic.
7. **Follow TRIPOD guidelines** for transparent reporting of prediction models.
8. **Combine with clinical knowledge** -- adaptive LASSO selects statistically
   important variables, but clinical relevance requires domain expertise.

---

## When to Use Adaptive LASSO vs Other Methods

| Scenario | Recommended Method |
|----------|-------------------|
| Many predictors, want sparse model | **Adaptive LASSO** |
| Standard LASSO but want less bias | **Adaptive LASSO** |
| Correlated predictors, grouped selection | **Adaptive LASSO (alpha < 1)** or Elastic Net |
| p >> n, genomic/radiomic features | **Adaptive LASSO with ridge weights** |
| Few predictors, traditional inference | Standard Cox regression |
| Quick exploratory screen | Standard LASSO (`weight_method = "equal"`) |
| Baseline comparison | Standard LASSO, then compare with adaptive |
| Grouped variables (e.g., gene pathways) | Group LASSO |

---

## Related ClinicoPath Functions

| Function | Module | Use When |
|----------|--------|----------|
| **LASSO Cox** (`lassocox`) | jsurvival | Standard LASSO without adaptive weights |
| **Adaptive LASSO** (`adaptivelasso`) | jsurvival | Oracle-property variable selection (this function) |
| **Group LASSO** (`grouplasso`) | jsurvival | Select/exclude grouped variables together |
| **NCV Reg Cox** (`ncvregcox`) | jsurvival | Non-convex penalties (SCAD, MCP) |
| **High-Dim Cox** (`highdimcox`) | jsurvival | Very high-dimensional data (p >> n) |
| **PLS Cox** (`plscox`) | jsurvival | Dimension reduction via partial least squares |
| **Survival Analysis** | jsurvival | Standard Kaplan-Meier with log-rank test |
| **Multivariable Survival** | jsurvival | Standard Cox with no regularization |

---

## References

- Zou H. The adaptive lasso and its oracle properties. *J Am Stat Assoc*. 2006;101(476):1418-1429.
- Zhang HH, Lu W. Adaptive Lasso for Cox's proportional hazards model. *Biometrika*. 2007;94(3):691-703.
- Tibshirani R. The lasso method for variable selection in the Cox model. *Stat Med*. 1997;16(4):385-395.
- Simon N, et al. Regularization paths for Cox's proportional hazards model via coordinate descent. *J Stat Softw*. 2011;39(5):1-13.
- Meinshausen N, Buhlmann P. Stability selection. *J R Stat Soc B*. 2010;72(4):417-473.
- Collins GS, et al. Transparent reporting of a multivariable prediction model for individual prognosis or diagnosis (TRIPOD). *BMJ*. 2015;350:g7594.
