---
title: "Comprehensive Survival Analysis with ClinicoPath"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Comprehensive Survival Analysis with ClinicoPath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = TRUE
)
```

## Introduction

The `survival` function in ClinicoPath provides a comprehensive suite of survival analysis tools designed for clinicopathological research. This vignette demonstrates all major features using the `survival_comprehensive` dataset, which simulates a cohort of 200 cancer patients with 15 clinical and pathological variables.

## Setup

```{r load-package}
devtools::load_all(".")
```

## Dataset Overview

```{r load-data}
data("survival_comprehensive")

cat("Dimensions:", nrow(survival_comprehensive), "patients x",
    ncol(survival_comprehensive), "variables\n\n")

cat("Variables:\n")
str(survival_comprehensive, max.level = 1)
```

```{r dataset-summary}
cat("\nOutcome distribution:\n")
print(table(survival_comprehensive$Status))

cat("\nExplanatory variable (TumorStage):\n")
print(table(survival_comprehensive$TumorStage))

cat("\nFollow-up time summary (months):\n")
print(summary(survival_comprehensive$FollowUpMonths))
```

## 1. Basic Kaplan-Meier Analysis

The simplest use case: compare survival between tumor stage groups using overall survival (event = Dead of Disease).

```{r basic-km}
result_basic <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease"
)
```

This produces:

- Median survival table for each stage group
- Cox regression hazard ratios (univariable)
- 1, 3, 5-year survival probability table (at months 12, 36, 60)

## 2. Survival Plots

### 2.1 Standard Kaplan-Meier Curve

```{r km-plot}
result_plots <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  sc = TRUE,
  ci95 = TRUE,
  risktable = TRUE,
  censored = TRUE,
  pplot = TRUE,
  medianline = "hv",
  endplot = 80,
  byplot = 12,
  ybegin_plot = 0.0,
  yend_plot = 1.0
)
```

### 2.2 KMunicate-Style Plot

The KMunicate format provides an extended risk table with cleaner presentation, following the recommendations of Morris et al. (2019).

```{r kmunicate-plot}
result_kmunicate <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  kmunicate = TRUE,
  endplot = 80,
  byplot = 12
)
```

### 2.3 Cumulative Events and Cumulative Hazard

```{r cumulative-plots}
result_cumulative <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  ce = TRUE,
  ch = TRUE,
  endplot = 80,
  byplot = 12
)
```

### 2.4 Log-Log Plot for Proportional Hazards Assessment

The log-log plot displays log(-log(S(t))) vs log(t). Parallel lines suggest the proportional hazards assumption holds.

```{r loglog-plot}
result_loglog <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  loglog = TRUE,
  endplot = 80
)
```

## 3. Cox Regression and Diagnostics

### 3.1 Proportional Hazards Assumption Test

```{r cox-ph}
result_cox <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  ph_cox = TRUE
)
```

### 3.2 Residual Diagnostics

```{r cox-residuals}
result_residuals <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  residual_diagnostics = TRUE
)
```

### 3.3 Stratified Cox Regression

When the PH assumption is violated for a covariate, stratified Cox regression allows different baseline hazards for each stratum.

```{r stratified-cox}
result_stratified <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  stratified_cox = TRUE,
  strata_variable = "MarginStatus"
)
```

## 4. Age Adjustment

Age is one of the most important confounders in cancer research. The survival function provides multiple approaches to handle age.

### 4.1 Age-Adjusted Cox Regression

```{r age-adjusted}
result_age <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  age_adjustment = TRUE,
  age_variable = "Age"
)
```

### 4.2 Age x Group Interaction Test

Test whether the effect of tumor stage on survival differs by patient age.

```{r age-interaction}
result_age_int <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  age_adjustment = TRUE,
  age_variable = "Age",
  age_interaction = TRUE
)
```

### 4.3 Age as Time Scale

In cancer epidemiology, using biological age as the time axis (rather than follow-up time) is often considered more rigorous, as it directly models the age-driven risk.

```{r age-time-scale}
result_age_time <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  age_adjustment = TRUE,
  age_variable = "Age",
  age_time_scale = TRUE
)
```

### 4.4 Age Standardization (SMR)

Standardized Mortality Ratio compares observed deaths to expected deaths based on age-specific rates.

```{r age-standardization}
result_smr <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  age_adjustment = TRUE,
  age_variable = "Age",
  age_standardization = TRUE,
  age_standardization_method = "indirect",
  age_group_cutpoints = "50, 65, 75"
)
```

### 4.5 Age-Stratified KM Plots and Adjusted Curves

```{r age-stratified-km}
result_age_km <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  age_adjustment = TRUE,
  age_variable = "Age",
  age_stratified_km = TRUE,
  adjusted_curves = TRUE,
  age_group_cutpoints = "50, 65, 75",
  sc = TRUE,
  endplot = 80,
  byplot = 12
)
```

## 5. Competing Risks Analysis

When patients can experience multiple types of events (e.g., death from disease vs death from other causes), competing risks analysis using cumulative incidence functions is more appropriate than standard KM.

```{r competing-risks}
result_compete <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  multievent = TRUE,
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  analysistype = "compete",
  explanatory = "TumorStage"
)
```

### 5.1 Cause-Specific Survival

Cause-specific survival censors deaths from other causes, focusing only on the event of interest.

```{r cause-specific}
result_cause <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  multievent = TRUE,
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  analysistype = "cause",
  explanatory = "TumorStage",
  sc = TRUE,
  endplot = 80
)
```

## 6. RMST (Restricted Mean Survival Time)

RMST represents the average survival time up to a specified time horizon. It is particularly useful when:

- Survival curves cross (non-proportional hazards)
- Median survival cannot be estimated (curve does not reach 50%)
- A clinically meaningful time horizon is of interest

```{r rmst-analysis}
result_rmst <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  rmst_analysis = TRUE,
  rmst_tau = 60
)
```

### 6.1 RMST with Automatic Tau Selection

When `rmst_tau = 0`, the function automatically selects the 75th percentile of follow-up time.

```{r rmst-auto-tau}
result_rmst_auto <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  rmst_analysis = TRUE,
  rmst_tau = 0
)
```

## 7. Person-Time Analysis

Person-time analysis calculates incidence rates that account for varying follow-up durations. This is essential for studies with differential loss to follow-up.

```{r person-time}
result_pt <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  person_time = TRUE,
  time_intervals = "12, 36, 60",
  rate_multiplier = 1000
)
```

## 8. Statistical Tests

### 8.1 Pairwise Comparisons

Compare survival between all pairs of groups with multiple testing correction.

```{r pairwise}
result_pw <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  pw = TRUE,
  padjustmethod = "bonferroni"
)
```

### 8.2 Weighted Log-Rank Tests

Different weighting schemes emphasize different parts of the survival curve. The standard log-rank test weights all time points equally, while Gehan-Breslow emphasizes early differences.

```{r weighted-logrank}
result_wlr <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  weightedLogRank = TRUE,
  survivalTestType = "fleming_harrington"
)
```

## 9. Model Validation

### 9.1 Calibration Assessment

Calibration compares predicted survival probabilities from the Cox model against observed Kaplan-Meier estimates, grouped by predicted risk quintiles.

```{r calibration}
result_cal <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  calibration_curves = TRUE,
  calibration_timepoint = 60,
  calibration_ngroups = 5
)
```

### 9.2 Bootstrap Internal Validation

Harrell's bootstrap optimism-corrected C-index quantifies how much the apparent model performance is inflated by overfitting.

```{r bootstrap-validation}
result_boot <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  bootstrapValidation = TRUE,
  bootstrapValN = 100
)
```

### 9.3 Non-Linearity Assessment (RCS)

Restricted cubic splines (RCS) test whether a continuous predictor has a non-linear relationship with the hazard. This is important for variables like age, where the risk may not increase linearly.

```{r rcs-analysis}
result_rcs <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  rcs_analysis = TRUE,
  rcs_variable = "Age",
  rcs_knots = 4
)
```

## 10. Landmark Analysis

Landmark analysis assesses conditional survival for patients who have survived beyond a specified time point. This is useful for updating prognosis after treatment milestones.

```{r landmark}
result_landmark <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  uselandmark = TRUE,
  landmark = 12,
  sc = TRUE,
  endplot = 70,
  byplot = 12
)
```

## 11. Date-Based Time Calculation

When survival time is not pre-calculated, you can provide diagnosis and follow-up dates directly.

```{r date-based}
result_dates <- survival(
  data = survival_comprehensive,
  tint = TRUE,
  dxdate = "DiagnosisDate",
  fudate = "LastFollowUpDate",
  timetypedata = "ymd",
  timetypeoutput = "months",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease"
)
```

## 12. Explanations and Clinical Reporting

### 12.1 Analysis Explanations

The `showExplanations` option provides detailed methodological explanations for each analysis component, helpful for learning and for auditing the analysis.

```{r explanations}
result_explain <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  showExplanations = TRUE
)
```

### 12.2 Natural Language Summaries and REMARK Checklist

```{r summaries-remark}
result_summary <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  showSummaries = TRUE,
  remark_checklist = TRUE
)
```

## 13. Comprehensive Example (All Features)

This example enables all major features simultaneously to demonstrate the full breadth of the analysis.

```{r comprehensive, eval=FALSE}
result_full <- survival(
  data = survival_comprehensive,
  # Core variables
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorStage",
  # Multi-event levels
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  # Survival table cutpoints
  cutp = "12, 36, 60",
  # Plots

  sc = TRUE,
  kmunicate = TRUE,
  ce = TRUE,
  ch = TRUE,
  loglog = TRUE,
  ci95 = TRUE,
  risktable = TRUE,
  censored = TRUE,
  pplot = TRUE,
  medianline = "hv",
  endplot = 80,
  byplot = 12,
  ybegin_plot = 0.0,
  yend_plot = 1.0,
  # Cox diagnostics
  ph_cox = TRUE,
  residual_diagnostics = TRUE,
  stratified_cox = TRUE,
  strata_variable = "MarginStatus",
  # Age adjustment
  age_adjustment = TRUE,
  age_variable = "Age",
  age_interaction = TRUE,
  age_stratified_cox = TRUE,
  age_group_cutpoints = "50, 65, 75",
  age_time_scale = TRUE,
  age_standardization = TRUE,
  age_standardization_method = "indirect",
  age_stratified_km = TRUE,
  adjusted_curves = TRUE,
  # Validation
  calibration_curves = TRUE,
  calibration_timepoint = 60,
  calibration_ngroups = 5,
  bootstrapValidation = TRUE,
  bootstrapValN = 100,
  rcs_analysis = TRUE,
  rcs_variable = "Age",
  rcs_knots = 4,
  # Statistical tests
  pw = TRUE,
  padjustmethod = "bonferroni",
  weightedLogRank = TRUE,
  survivalTestType = "logrank",
  # RMST
  rmst_analysis = TRUE,
  rmst_tau = 60,
  # Person-time
  person_time = TRUE,
  time_intervals = "12, 36, 60",
  rate_multiplier = 1000,
  # Reporting
  showExplanations = TRUE,
  showSummaries = TRUE,
  remark_checklist = TRUE
)
```

## 14. Using Different Explanatory Variables

### 14.1 Binary Variable (Sex)

```{r binary-explanatory}
result_sex <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "Sex",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  sc = TRUE,
  endplot = 80
)
```

### 14.2 Three-Level Variable (TumorGrade)

```{r grade-explanatory}
result_grade <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "TumorGrade",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  sc = TRUE,
  pw = TRUE,
  padjustmethod = "holm",
  endplot = 80
)
```

### 14.3 Clinical Factor (LymphovascularInvasion)

```{r lvi-explanatory}
result_lvi <- survival(
  data = survival_comprehensive,
  elapsedtime = "FollowUpMonths",
  outcome = "Status",
  outcomeLevel = "Dead of Disease",
  explanatory = "LymphovascularInvasion",
  dod = "Dead of Disease",
  dooc = "Dead of Other Causes",
  awd = "Alive with Disease",
  awod = "Alive without Disease",
  sc = TRUE,
  endplot = 80,
  rmst_analysis = TRUE,
  rmst_tau = 60
)
```

## Session Info

```{r session-info}
sessionInfo()
```
