---
title: "LASSO Cox Regression for Variable Selection in Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LASSO Cox Regression for Variable Selection in Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Not yet released.** The `lassocox` analysis is on a development menu route, so it does
> not appear in the jamovi menus of ClinicoPath or of any of its submodules. It is
> documented here ahead of a future release, and its options, defaults and output may
> still change. The R function is exported, so the examples below run from an R
> console; what is not yet available is the jamovi analysis itself.


> **Note:** The examples use the exported R wrapper. Both `outcomeLevel` and
> `censorLevel` are required arguments because they define the event coding
> explicitly. During package development, run `devtools::load_all()` first.

# LASSO Cox Regression

## Overview

The **Lasso-Cox Regression** module (`lassocox`) performs automatic variable selection
in survival analysis using L1-penalized Cox proportional hazards regression.
It estimates a sparse set of retained predictor columns while shrinking
coefficients. Regularization can reduce overfitting, but it does not remove the
need to validate the complete modeling process.

Key features:

- **Data suitability assessment** (traffic-light system)
- **Cross-validated lambda selection** (lambda.min vs lambda.1se)
- **Variable selection** with coefficient shrinkage
- **Apparent risk-score calculation** and exploratory development-sample grouping
- **Comprehensive visualizations** (CV plot, coefficient plot, survival curves)
- **Clinical interpretation guidance**

---

## What is LASSO Cox Regression?

LASSO (Least Absolute Shrinkage and Selection Operator) Cox regression combines the
Cox proportional hazards model with L1 regularization for automatic variable selection.

**Standard Cox model:**

$$h(t) = h_0(t) \exp(\beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p)$$

**LASSO Cox adds a penalty:**

$$\text{Maximize:} \quad \ell(\beta) - \lambda \sum_{j=1}^{p} |\beta_j|$$

where $\ell(\beta)$ is the partial log-likelihood and $\lambda$ controls penalty strength.

The effect: coefficients shrink toward zero and some become exactly zero. A
nonzero coefficient means that a predictor column was retained in this fitted
development model; it does not establish clinical importance or causality.

### LASSO vs Ridge vs Elastic Net

| Method | Penalty | Variable Selection | Best For |
|--------|---------|-------------------|----------|
| **Ridge** | $\lambda \sum \beta_j^2$ (L2) | No - shrinks but keeps all | Correlated predictors, all must stay |
| **LASSO** | $\lambda \sum |\beta_j|$ (L1) | Yes - sets some to zero | Sparse models, feature selection |
| **Elastic Net** | $\alpha \lambda \sum |\beta_j| + (1-\alpha) \lambda \sum \beta_j^2$ | Yes - can retain correlated columns together more readily | Correlated candidate predictors |

In ClinicoPath: `lassocox` uses alpha=1 (pure LASSO). For Elastic Net or Ridge,
use the **Penalized Cox Regression** module.

---

## When to Use LASSO Cox Regression

| Scenario | Recommendation |
|----------|---------------|
| Many predictors relative to sample | Ordinary Cox fitting may be unstable; consider penalization within a fully validated modeling strategy |
| Radiomics / genomics feature reduction | LASSO is one candidate method; assess feature stability and validate the full pipeline |
| Developing a prognostic risk score | LASSO may be considered with appropriate validation and calibration |
| Few predictors (p < 5) with ample events | Standard Cox may suffice |
| Correlated predictors that should all be retained | Consider Elastic Net |
| Inference / p-values are the primary goal | Use standard Cox regression |

---

## Datasets Used in This Guide

| Dataset | N | Events | Predictors | Description |
|---------|---|--------|------------|-------------|
| Synthetic "standard" | 200 | 80 | 3 numeric | Basic advisory-assessment scenario |
| Synthetic "small" | 60 | 25 | 5 numeric | Limited development-sample scenario |
| Synthetic "collinear" | 100 | 45 | 4 (2 correlated) | Multicollinearity detection |
| Synthetic "categorical" | 100 | 40 | 3 factors | All categorical predictors |
| `histopathology` | 250 | ~100 | 8 mixed | Real clinical pathology data |
| `lassocox_genomic` | 80 | - | 50 gene features | High-dimensional genomic scenario |
| `lassocox_multicollinear` | 180 | - | 12 correlated | Multicollinearity stress test |
| `lassocox_breast_cancer` | 250 | - | 20 clinical | Breast cancer clinicopathological |

---

## 1. Basic Analysis

A standard scenario with adequate sample size, events, and predictors.

```{r basic-green}
# Create synthetic data: 200 patients, 80 events, 3 numeric predictors
set.seed(42)
n <- 200
data_green <- data.frame(
  time = abs(rnorm(n, 20, 10)) + 0.1,
  status = factor(c(rep("event", 80), rep("censored", 120)),
                  levels = c("censored", "event")),
  var1 = rnorm(n),
  var2 = rnorm(n),
  var3 = rnorm(n)
)

lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = TRUE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

The suitability report is advisory. For this example it should describe:

- **EPV**: 80/3 = 26.7 (excellent)
- **Sample size**: n=200 (adequate)
- **Event rate**: 40% (balanced)
- **Multicollinearity**: uncorrelated predictors

---

## 2. Small Sample

When sample size is limited, the suitability assessment flags potential issues.

```{r small-sample}
set.seed(42)
data_small <- data.frame(
  time = abs(rnorm(60, 15, 8)) + 0.1,
  status = factor(c(rep("event", 25), rep("censored", 35)),
                  levels = c("censored", "event")),
  var1 = rnorm(60), var2 = rnorm(60), var3 = rnorm(60),
  var4 = rnorm(60), var5 = rnorm(60)
)

lassocox(
  data = data_small,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3", "var4", "var5"),
  suitabilityCheck = TRUE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

Expected suitability information:

- **EPV**: 25 events divided by the encoded-column count
- **Sample size**: n=60, above the hard minimum but still limited for prediction modeling
- **Regularization**: an advisory comparison with an ordinary Cox model

---

## 3. Multicollinearity Detection

When predictors are highly correlated, the suitability check identifies the pairs.

```{r collinearity}
set.seed(42)
base_var <- rnorm(100)
data_collinear <- data.frame(
  time = abs(rnorm(100, 20, 10)) + 0.1,
  status = factor(c(rep("event", 45), rep("censored", 55)),
                  levels = c("censored", "event")),
  var1 = base_var,
  var2 = base_var + rnorm(100, 0, 0.1),  # r > 0.95 with var1
  var3 = rnorm(100),
  var4 = rnorm(100)
)

lassocox(
  data = data_collinear,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3", "var4"),
  suitabilityCheck = TRUE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

The suitability report should flag:

- **Multicollinearity**: a warning describing the high correlation
- **Top correlated pair**: var1 & var2
- **Recommendation**: Consider Elastic Net (Penalized Cox in ClinicoPath)

---

## 4. Categorical Predictors

LASSO handles factor variables via dummy coding. The collinearity check
correctly excludes within-factor dummy correlations.

```{r categorical}
set.seed(42)
data_cat <- data.frame(
  time = abs(rnorm(100, 20, 10)) + 0.1,
  status = factor(c(rep("event", 40), rep("censored", 60)),
                  levels = c("censored", "event")),
  grade = factor(sample(c("Low", "Med", "High"), 100, replace = TRUE)),
  stage = factor(sample(c("I", "II", "III"), 100, replace = TRUE)),
  treatment = factor(sample(c("A", "B"), 100, replace = TRUE))
)

lassocox(
  data = data_cat,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("grade", "stage", "treatment"),
  suitabilityCheck = TRUE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

---

## 5. Lambda Selection: lambda.min vs lambda.1se

The choice of lambda controls the trade-off between model complexity and parsimony.

```{r lambda-min}
# lambda.min: lambda with minimum cross-validated error estimate
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  lambda = "lambda.min",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

```{r lambda-1se}
# lambda.1se: most regularized value within 1 SE, often retaining fewer columns
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

**Interpretation:**

- **lambda.min**: Minimizes the cross-validated partial-likelihood error estimate.
- **lambda.1se** (default): Uses the more penalized value within one standard error of the minimum and often retains fewer columns. Neither rule guarantees better generalization.

---

## 6. Suitability Check Toggle

The suitability assessment is advisory - it never blocks the analysis.

```{r toggle-off}
# Disable suitability assessment
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,  # No suitability report shown
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE
)
```

---

## 7. Explanatory Output Options

### Method explanations

```{r explanations}
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE,
  showSummary = TRUE,              # Natural-language results paragraph for manuscripts
  showExplanations = TRUE,         # LASSO methodology + plot explanations
  showMethodologyNotes = TRUE,     # Technical details
  includeClinicalGuidance = TRUE   # Clinical interpretation guide
)
```

### Variable importance and model comparison

```{r importance-comparison}
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  lambda = "lambda.min",  # More likely to select variables
  nfolds = 5,
  standardize = TRUE,
  cv_plot = FALSE,
  coef_plot = FALSE,
  survival_plot = FALSE,
  showVariableImportance = TRUE,   # Magnitudes + path inclusion proportions
  showModelComparison = TRUE       # LASSO vs Standard Cox comparison
)
```

---

## 8. Risk Score Output

The computed risk score can be saved as a new variable in the dataset.

```{r risk-score}
result <- lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  lambda = "lambda.min",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = FALSE,
  coef_plot = FALSE,
  survival_plot = FALSE
)
```

The risk score is computed as:

$$\text{Risk Score} = \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k$$

where only LASSO-selected variables have non-zero coefficients.

---

## Interpreting Results

### Model Summary Table

| Statistic | Meaning |
|-----------|---------|
| Total Variables | Number of candidate predictors (after dummy coding) |
| Selected Variables | Variables with non-zero coefficients |
| Selection Proportion | Percentage of variables retained |
| Optimal Lambda | The regularization strength used |
| Sample Size | Complete cases used |
| Number of Events | Events observed |
| Censoring Rate | Percentage of censored observations |

### Selected Variables Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name |
| Penalized Coefficient | Original-scale coefficient from the penalized fit; positive and negative values indicate higher and lower fitted hazard |
| Penalized Hazard Ratio | exp(coefficient), interpreted for the encoded column and its units |
| Scale-Adjusted Magnitude | Absolute coefficient multiplied by the encoded column's complete-case SD; descriptive only |

### Performance Table

| Metric | Meaning |
|--------|---------|
| Apparent C-index | Development-sample discrimination calculated in the same data used for preprocessing, tuning, and fitting |

The apparent C-index is a descriptive development-sample estimate. No universal
C-index threshold establishes calibration, transportability, or clinical utility.

---

## Common Pitfalls and Best Practices

### Pitfalls to Avoid

1. **Using LASSO with too few events**: Penalization does not create information. A low event count relative to the candidate model makes tuning, retained columns, and performance estimates unstable; no fixed event-per-variable threshold guarantees reliability.

2. **Not standardizing variables**: LASSO is sensitive to scale. A variable measured in 0 - 20,000 (e.g., tumor area) will dominate one measured in 0 - 100 (e.g., age) unless standardized. Use `standardize = TRUE` (the default).

3. **Interpreting coefficients as p-values**: LASSO performs *selection*, not *testing*. Prefer wording such as "the Ki67 encoded column was retained at the selected lambda." Do not call it statistically significant or clinically important from this output.

4. **Feature selection on full data (data leakage)**: A very common mistake. If you perform LASSO feature selection on all patients, then build and test the model on the same data, information leaks from the test set into the selection step. Feature selection must be done *within the training set only*.

5. **No validation of the full pipeline**: The displayed cross-validation tunes
   lambda; it is not an independent estimate of final-model performance. Use
   bootstrap optimism correction or nested cross-validation that repeats all
   preprocessing and tuning, followed by external validation where feasible.

6. **Ignoring feature stability**: Radiomic or WSI features may change with scanner, staining protocol, segmentation method, or resolution. Before LASSO, features should be tested for robustness using intraclass correlation coefficient (ICC) or test - retest analysis.

7. **Reporting only selected variables**: Good reporting requires: selected features, lambda value used, cross-validation method, coefficient values, and performance metrics. Follow the TRIPOD statement for transparent reporting.

8. **Ignoring established clinical predictors**: Assess whether candidate models
   add value beyond an appropriate prespecified clinical model. Added value is an
   empirical question and must be validated.

9. **Not assessing model performance properly**: Survival models should report C-index *and* calibration curves. High discrimination (C-index) alone is insufficient - the model must also be well-calibrated. Consider decision curve analysis for clinical utility.

10. **No calibration analysis**: Discrimination alone is insufficient. Evaluate
    calibration-in-the-large, calibration slope, and time-specific calibration
    using methods appropriate for censored survival outcomes.

### Best Practices

1. **Check suitability first**: Use the built-in assessment before interpreting results
2. **Choose and report a lambda rule**: `lambda.1se` is usually more penalized; it is not guaranteed to generalize better than `lambda.min`
3. **Report transparently**: Lambda, CV method, all selected variables, coefficients
4. **Validate externally**: Test in independent cohorts when possible
5. **Assess established clinical predictors**: Prespecify and validate comparisons with appropriate clinical models
6. **Follow TRIPOD guidelines**: For prognostic model reporting
7. **Test feature stability**: Use ICC before entering features into LASSO
8. **Perform selection within training set**: Never select features on the full dataset

---

## Radiomics and Digital Pathology Pipeline

LASSO Cox regression is central to many radiomics and digital pathology prognostic studies.
A robust analysis pipeline follows these steps:

```
WSI / Imaging Data
       ↓
Feature Extraction (e.g., texture, morphology, deep features)
       ↓
Feature Stability Testing (ICC, test - retest)
       ↓
Feature Reduction (remove unstable/redundant features)
       ↓
LASSO Cox Selection (cross-validated, training set only)
       ↓
Model Construction (risk score from selected features)
       ↓
Internal Validation (bootstrap or cross-validation)
       ↓
External Validation (independent cohort)
       ↓
Calibration + Decision Curve Analysis
```

### Application Areas

| Domain | Example Use |
|--------|------------|
| **WSI radiomics** | Texture and morphology features from whole-slide images |
| **Deep feature selection** | CNN-extracted features reduced via LASSO for interpretability |
| **Tumor microenvironment signatures** | Spatial features (immune cell density, stroma ratio) |
| **Gene expression survival models** | Selecting prognostic genes from expression panels |

### Typical Wording in Papers

When reporting LASSO Cox results, consider these example phrasings:

> "Feature selection was performed using LASSO penalized Cox regression with
> 10-fold cross-validation. The selected features were used to construct a
> radiomic risk score."

> "Variables were selected using LASSO regression with 10-fold cross-validation,
> and selected predictors were entered into the multivariable Cox model."

---

## 9. Using Real Test Datasets

The package includes several realistic test datasets for LASSO Cox analysis.

### Breast cancer clinical data

```{r real-breast-cancer}
data(lassocox_breast_cancer)

lassocox(
  data = lassocox_breast_cancer,
  elapsedtime = "survival_months",
  outcome = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  explanatory = c("age", "tumor_size_cm", "grade", "lymph_nodes_positive",
                   "ki67_percent", "er_status", "her2_status", "lvi"),
  suitabilityCheck = FALSE,
  lambda = "lambda.1se",
  nfolds = 10,
  standardize = TRUE,
  cv_plot = TRUE,
  coef_plot = TRUE,
  survival_plot = TRUE,
  showSummary = TRUE
)
```

The `showSummary = TRUE` option produces a natural-language draft. Verify the
event coding, exclusions, estimator labels, and validation caveats before using
the text in a manuscript or report.

---

## 10. Reproducibility with Random Seeds

The `random_seed` option ensures reproducible cross-validation fold assignments.
Using the same seed produces identical results every time.

```{r seed-demo}
# Fixed seed for reproducible results
lassocox(
  data = data_green,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "event",
  censorLevel = "censored",
  explanatory = c("var1", "var2", "var3"),
  suitabilityCheck = FALSE,
  random_seed = 42,
  lambda = "lambda.1se",
  nfolds = 5,
  standardize = TRUE,
  cv_plot = FALSE,
  coef_plot = FALSE,
  survival_plot = FALSE
)
```

Different seeds may produce different retained columns, especially when
predictors provide similar fitted signal. For publications, report the seed and
the complete resampling procedure.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **Survival Analysis** | Standard Kaplan-Meier with log-rank test (few predictors) |
| **Multivariable Survival** | Standard Cox with multiple predictors (no regularization needed) |
| **Penalized Cox Regression** | Elastic Net or Ridge (retain correlated predictors) |
| **Adaptive LASSO** | Two-stage LASSO with oracle properties |
| **Group LASSO** | Select/exclude grouped variables together |

---

## References

- Tibshirani R. The lasso method for variable selection in the Cox model. *Stat Med*. 1997;16(4):385-395.
- Simon N, et al. Regularization paths for Cox's proportional hazards model via coordinate descent. *J Stat Softw*. 2011;39(5):1-13.
- Collins GS, et al. Transparent reporting of a multivariable prediction model for individual prognosis or diagnosis (TRIPOD). *BMJ*. 2015;350:g7594.
