---
title: "PCA Cox Regression - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA Cox Regression - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
data_path <- "data-raw/non-rda/"
data_path2 <- "data/"
```

> **Note:** The `pcacox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference.

# PCA Cox Regression

## Overview

PCA Cox regression addresses the challenge of high-dimensional survival data where the number of predictors approaches or exceeds the sample size. By reducing the predictor space to a smaller set of orthogonal principal components, it enables Cox proportional hazards modeling even when standard regression would fail due to multicollinearity or overfitting.

The module supports four PCA methods: supervised PCA (using survival information to guide component extraction via the superpc package), standard PCA (unsupervised via prcomp), sparse PCA (with interpretable sparse loadings via sparsepca), and kernel PCA (for nonlinear relationships via kernlab). Each method falls back gracefully to standard PCA if the required package is unavailable.

This analysis is particularly valuable for genomic survival studies (gene expression signatures), proteomics/metabolomics panels, imaging feature sets, and any clinical scenario with many correlated predictors.

---

## Datasets Used in This Guide

| Dataset | N | Predictors | Events | Primary Use |
|---------|---|-----------|--------|-------------|
| `pcacox_clinical` | 60 | 10 (mixed clinical) | 30 | Standard clinical scenarios |
| `pcacox_genomic` | 150 | 30 (6 correlated gene blocks) | 72 | High-dimensional genomics |

---

## 1. Standard PCA with Clinical Data

### Basic analysis

```{r standard-pca}
clinical <- read.csv(paste0(data_path, "pcacox_clinical.csv"))
str(clinical[, 1:6])

pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  pca_method = "standard",
  n_components = 3,
  component_selection = "fixed",
  suitabilityCheck = TRUE
)
```

Look for: PCA summary (eigenvalues, variance explained, selection status), Cox model table (coefficients, HRs, p-values), model performance (C-index, R-squared, AIC).

---

## 2. PCA Methods

### Supervised PCA (survival-weighted)

Supervised PCA uses survival information to select features before extracting components, focusing on survival-relevant variation.

```{r supervised-pca}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  pca_method = "supervised",
  survival_weighting = TRUE,
  n_components = 3,
  suitabilityCheck = FALSE
)
```

### Sparse PCA

Sparse PCA produces components with few non-zero loadings, making interpretation easier.

```{r sparse-pca}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  pca_method = "sparse",
  sparse_parameter = 0.1,
  n_components = 3,
  suitabilityCheck = FALSE
)
```

---

## 3. Component Selection

### Cross-validation

CV-based selection evaluates 1-K components and picks the number maximizing out-of-fold C-index.

```{r cv-selection}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  component_selection = "cv",
  cv_folds = 5,
  suitabilityCheck = FALSE
)
```

### Variance threshold

Select the minimum number of components explaining at least the specified proportion of variance.

```{r variance-selection}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  component_selection = "variance",
  variance_threshold = 0.8,
  suitabilityCheck = FALSE
)
```

---

## 4. Validation

### Bootstrap optimism correction

```{r bootstrap-validation}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh"),
  clinical_vars = NULL,
  n_components = 2,
  bootstrap_validation = TRUE,
  n_bootstrap = 50,
  permutation_test = TRUE,
  n_permutations = 50,
  suitabilityCheck = FALSE
)
```

Look for: optimism-corrected C-index (should be lower than apparent), calibration slope, per-component permutation p-values.

---

## 5. Additional Analyses

### Model comparison + feature clusters

```{r additional}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  n_components = 3,
  show_model_comparison = TRUE,
  pathway_analysis = TRUE,
  risk_score = TRUE,
  feature_importance = TRUE,
  suitabilityCheck = FALSE
)
```

Look for: sequential model comparison (1-5 PCs with AIC/BIC/C-index), feature cluster analysis by dominant component, risk group KM curves.

---

## 6. High-Dimensional Genomic Data

```{r genomic}
genomic <- read.csv(paste0(data_path, "pcacox_genomic.csv"))
gene_vars <- names(genomic)[!names(genomic) %in% c("time", "status")]

pcacox(
  data = genomic,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = gene_vars,
  clinical_vars = NULL,
  pca_method = "standard",
  n_components = 5,
  component_selection = "variance",
  variance_threshold = 0.8,
  feature_importance = TRUE,
  pathway_analysis = TRUE,
  suitabilityCheck = TRUE
)
```

With 30 predictors and 150 observations, PCA dimensionality reduction is essential. The feature cluster analysis groups genes by their dominant component.

---

## 7. All Plots

```{r all-plots}
pcacox(
  data = clinical,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi", "albumin", "crp", "ldh",
                 "hemoglobin", "wbc", "platelets", "tumor_size", "ki67"),
  clinical_vars = NULL,
  n_components = 3,
  plot_scree = TRUE,
  plot_loadings = TRUE,
  plot_biplot = TRUE,
  plot_survival = TRUE,
  risk_score = TRUE,
  suitabilityCheck = FALSE
)
```

---

## 8. Edge Case: Small Sample

```{r small-sample}
small <- clinical[1:25, ]

pcacox(
  data = small,
  time = "time",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "albumin", "crp", "ldh"),
  clinical_vars = NULL,
  n_components = 2,
  suitabilityCheck = TRUE,
  plot_scree = FALSE,
  plot_loadings = FALSE,
  plot_biplot = FALSE,
  plot_survival = FALSE,
  bootstrap_validation = FALSE,
  permutation_test = FALSE
)
```

---

## References

- Bair E, Tibshirani R (2004). "Semi-supervised methods to predict patient survival from gene expression data." *PLoS Biology*, 2(4), e108.
- Therneau TM (2026). *survival: Survival Analysis*. R package version 3.8-6.
- Erichson NB, Zheng P, Manohar K, Brunton S, Kutz JN, Aravkin AY (2020). "Sparse Principal Component Analysis via Variable Projection." *SIAM Journal on Applied Mathematics*.
