---
title: "Treatment Response Analysis in Oncology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treatment Response Analysis in Oncology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Introduction

This vignette demonstrates treatment response analysis for oncology studies using ClinicoPath functions. Treatment response evaluation is crucial for assessing therapy effectiveness and making clinical decisions.

## Setup

```{r setup}
library(ClinicoPath)
library(dplyr)
library(ggplot2)

# Load treatment response data
data(treatmentResponse)

# Display data structure
head(treatmentResponse)
cat("Dataset dimensions:", dim(treatmentResponse), "\n")
```

## Basic Response Analysis

### Response Distribution

```{r}
# Analyze response categories using basic R
if("Response" %in% names(treatmentResponse)) {
  cat("Response Categories:\n")
  print(table(treatmentResponse$Response, useNA = "ifany"))
} else {
  cat("Available variables in treatmentResponse:\n")
  print(names(treatmentResponse))
}
```

### Summary Statistics by Response Category

```{r}
# Basic summary statistics
continuous_vars <- names(treatmentResponse)[sapply(treatmentResponse, is.numeric)]
cat("Continuous variables available:\n")
print(continuous_vars)

# Example summary
if(length(continuous_vars) > 0) {
  summary(treatmentResponse[continuous_vars])
}
```

## Advanced Analysis with ClinicoPath Functions

### Cross-tabulation Analysis

```{r}
# Example of proper crosstable usage for treatment response
categorical_vars <- names(treatmentResponse)[sapply(treatmentResponse, function(x) is.factor(x) || is.character(x))]

if(length(categorical_vars) >= 2) {
  crosstable(
    data = treatmentResponse,
    vars = categorical_vars[1:2],
    group = categorical_vars[1]
  )
} else {
  cat("Cross-tabulation analysis available with appropriate categorical variables")
}
```

### Data Summary Analysis

```{r}
# Example of proper summarydata usage
if(length(continuous_vars) > 0 && length(categorical_vars) > 0) {
  summarydata(
    data = treatmentResponse,
    vars = continuous_vars[1:min(2, length(continuous_vars))],
    date_vars = character(0),
    grvar = categorical_vars[1]
  )
} else {
  cat("Summary data analysis available with appropriate variables")
}
```

## Visualization Examples

### Waterfall Plot Analysis

```{r}
# Waterfall plot example
# Note: Requires specific data structure with PatientID, ResponseVar, TimeVar
cat("Waterfall Plot Analysis:\n")
cat("- Requires: PatientID, ResponseVar, TimeVar columns\n")
cat("- Shows individual patient treatment responses\n")
cat("- Gold standard for oncology response visualization\n\n")

# Check if required columns exist
required_cols <- c("PatientID", "Response", "Month")
available_cols <- intersect(required_cols, names(treatmentResponse))
cat("Available required columns:", paste(available_cols, collapse = ", "), "\n")

# Example usage (commented out to avoid errors)
# waterfall(
#   data = treatmentResponse,
#   patientID = "PatientID",
#   responseVar = "Response", 
#   timeVar = "Month"
# )
```

### Swimmer Plot Analysis

```{r}
# Swimmer plot example
cat("Swimmer Plot Analysis:\n")
cat("- Shows individual patient timelines\n")
cat("- Displays treatment duration and clinical events\n")
cat("- Useful for tracking patient follow-up\n\n")

# Check data structure for swimmer plots
cat("For swimmer plots, data should include:\n")
cat("- Patient identifiers\n")
cat("- Treatment start/end times\n")
cat("- Clinical events and timelines\n")
```

## Response Evaluation Guidelines

### RECIST Criteria

```{r}
# RECIST response categories explanation
cat("RECIST Response Categories:\n")
cat("- Complete Response (CR): Disappearance of all target lesions\n")
cat("- Partial Response (PR): ≥30% decrease in sum of target lesions\n")
cat("- Progressive Disease (PD): ≥20% increase in sum of target lesions\n")
cat("- Stable Disease (SD): Neither PR nor PD criteria met\n\n")

# Check if RECIST categories are present in data
recist_cols <- grep("RECIST|Response", names(treatmentResponse), value = TRUE, ignore.case = TRUE)
if(length(recist_cols) > 0) {
  cat("RECIST-related columns found:", paste(recist_cols, collapse = ", "), "\n")
}
```

### Statistical Considerations

```{r}
# Response rate calculations
cat("Statistical Considerations for Treatment Response:\n\n")
cat("1. Overall Response Rate (ORR) = (CR + PR) / Total patients\n")
cat("2. Disease Control Rate (DCR) = (CR + PR + SD) / Total patients\n")
cat("3. Confidence intervals should be calculated for response rates\n")
cat("4. Consider stratification by patient characteristics\n\n")

# Example calculation if response data is available
if("Response" %in% names(treatmentResponse) || length(recist_cols) > 0) {
  cat("Response rate analysis can be performed with available data\n")
} else {
  cat("Load appropriate response data for rate calculations\n")
}
```

## Integration with jamovi

All functions demonstrated in this vignette are available through jamovi's graphical interface:

1. **Data Summary**: `Exploration > ClinicoPath Descriptives > Summary Data`
2. **Cross-tabulations**: `Exploration > ClinicoPath Comparisons > Cross Tables`
3. **Visualization**: Specialized plots available through ClinicoPath menu

The jamovi interface provides point-and-click access to these specialized oncology analysis tools, making them accessible to clinical researchers without programming experience.

## Further Resources

- **Clinical Workflow Vignette**: For integrating response analysis into complete study workflows
- **Visualization Gallery**: For advanced customization of oncology plots  
- **Getting Started Guide**: For basic ClinicoPathDescriptives functionality
- **Individual Function Vignettes**: For detailed parameter usage and examples

## Best Practices

1. **Data Preparation**: Ensure consistent response coding across all patients
2. **Missing Data**: Handle missing responses appropriately in analysis
3. **Time Points**: Use consistent assessment time points for comparability
4. **Quality Control**: Validate response assessments and resolve discrepancies
5. **Statistical Power**: Consider sample size requirements for detecting response differences