---
title: "SCAD/MCP Cox Regression for Variable Selection in Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCAD/MCP Cox Regression for Variable Selection in Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `ncvregcox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `ncvregcoxClass$new(options = ncvregcoxOptions$new(...), data = mydata)`.

# SCAD/MCP Cox Regression

## Overview

The **SCAD Cox Regression** module (`ncvregcox`) performs variable selection using
non-convex penalties — **SCAD** (Smoothly Clipped Absolute Deviation) and **MCP**
(Minimax Concave Penalty) — for Cox proportional hazards models. These penalties
solve a fundamental problem with LASSO: **over-penalization of large coefficients**.

Key features:

- **Three penalty types**: SCAD, MCP, and LASSO (for comparison)
- **Oracle properties**: Consistent selection + unbiased estimation for large effects
- **Cross-validated lambda selection** (min or 1-SE rule)
- **Regularization path** visualization
- **Variable importance** analysis

---

## Why Non-Convex Penalties?

LASSO applies a constant penalty rate regardless of coefficient size, which means:
- Small (noise) coefficients: shrunk to zero (good)
- Large (true) coefficients: also over-shrunk (bad — introduces bias)

**SCAD** and **MCP** use penalties that flatten out for large coefficients:

### Penalty Comparison

| Penalty | Behavior | Bias for Large Effects | Selection |
|---------|----------|----------------------|-----------|
| **LASSO** | Constant penalty rate | High bias | Consistent under conditions |
| **SCAD** | Penalty flattens at $\gamma\lambda$ | Nearly unbiased | Oracle property |
| **MCP** | Penalty flattens sooner | Nearly unbiased | Oracle property |
| **Ridge** | Quadratic penalty | Moderate bias | No selection |

### Mathematical Formulation

**LASSO penalty**: $P(\beta) = \lambda |\beta|$

**SCAD penalty** (Fan & Li, 2001):
$$P(\beta) = \begin{cases}
\lambda |\beta| & \text{if } |\beta| \leq \lambda \\
\frac{-(\beta^2 - 2\gamma\lambda|\beta| + \lambda^2)}{2(\gamma-1)} & \text{if } \lambda < |\beta| \leq \gamma\lambda \\
\frac{(\gamma+1)\lambda^2}{2} & \text{if } |\beta| > \gamma\lambda
\end{cases}$$

**MCP** (Zhang, 2010):
$$P(\beta) = \begin{cases}
\lambda |\beta| - \frac{\beta^2}{2\gamma} & \text{if } |\beta| \leq \gamma\lambda \\
\frac{\gamma\lambda^2}{2} & \text{if } |\beta| > \gamma\lambda
\end{cases}$$

The default $\gamma = 3.7$ for SCAD was recommended by Fan & Li (2001) based on
Bayesian arguments and minimization of a Bayesian risk.

---

## When to Use SCAD/MCP

| Scenario | Recommendation |
|----------|---------------|
| Suspect strong true effects that LASSO would bias | **SCAD or MCP** |
| Building a model for publication (unbiased coefficients) | **SCAD recommended** |
| Very high dimensional (p >> n) | LASSO may be more stable |
| Want oracle property without two-stage procedure | **SCAD or MCP** |
| Moderate number of predictors | **MCP** (more aggressive selection) |
| Need guaranteed convexity for optimization | Use LASSO |

---

## Datasets Used in This Guide

| Dataset | N | Events | Covariates | Description |
|---------|---|--------|------------|-------------|
| Synthetic "clinical" | 200 | ~110 | 15 mixed | Clinical study with known effects |
| Synthetic "sparse" | 100 | ~50 | 30 | Many noise variables, few true signals |

---

## 1. SCAD Penalty (Default)

SCAD is the recommended starting point — it provides nearly unbiased estimates
for truly important variables while setting noise variables to zero.

```{r clinical-data}
# Simulate clinical survival data
set.seed(42)
n <- 200

# Predictors with true effects
age <- rnorm(n, 62, 11)
bmi <- rnorm(n, 27, 5)
tumor_diameter <- rlnorm(n, log(3.5), 0.5)
ldh <- rlnorm(n, log(200), 0.4)
cea <- rlnorm(n, log(5), 0.8)
t_stage_num <- sample(1:4, n, replace = TRUE, prob = c(0.15, 0.3, 0.35, 0.2))
t_stage <- factor(paste0("T", t_stage_num))

# Noise predictors
crp <- rlnorm(n, log(3), 0.7)
albumin <- rnorm(n, 4.0, 0.5)
wbc <- rnorm(n, 8, 2.5)
neutrophil_ratio <- rbeta(n, 5, 3)
platelet <- rnorm(n, 250, 60)
gender <- factor(sample(c("M", "F"), n, replace = TRUE))
n_stage <- factor(sample(c("N0", "N1", "N2"), n, replace = TRUE,
                         prob = c(0.45, 0.35, 0.2)))
histology <- factor(sample(c("Adenocarcinoma", "Squamous", "Other"), n,
                           replace = TRUE, prob = c(0.5, 0.35, 0.15)))

# True survival model
lp <- 0.03 * (age - 62) +
      0.15 * tumor_diameter +
      0.003 * (ldh - 200) +
      0.05 * cea +
      0.3 * (t_stage_num - 1)

surv_time <- rweibull(n, shape = 1.2, scale = 36 * exp(-lp * 0.15))
censor_time <- runif(n, 8, 48)
time <- pmax(pmin(surv_time, censor_time), 0.1)
# ncvregcox expects factor for event
event <- factor(ifelse(surv_time <= censor_time, "1", "0"))

data_clinical <- data.frame(
  time = time, event = event,
  age = age, bmi = bmi, tumor_diameter = tumor_diameter,
  ldh = ldh, cea = cea, t_stage = t_stage,
  crp = crp, albumin = albumin, wbc = wbc,
  neutrophil_ratio = neutrophil_ratio, platelet = platelet,
  gender = gender, n_stage = n_stage, histology = histology
)

cat("N:", n, "Events:", sum(event == "1"),
    "Event rate:", round(mean(event == "1"), 2), "\n")
```

```{r scad-run}
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin", "wbc",
                 "neutrophil_ratio", "platelet", "gender",
                 "n_stage", "histology"),
  penalty = "SCAD",
  gamma = 3.7,
  alpha = 1.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

Expected: SCAD should select age, tumor_diameter, ldh, cea, and t_stage.
Noise variables should be eliminated with nearly unbiased coefficient estimates.

---

## 2. MCP Penalty

MCP is more aggressive than SCAD — it transitions to zero penalty more quickly,
leading to sparser models.

```{r mcp}
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin", "wbc",
                 "neutrophil_ratio", "platelet", "gender",
                 "n_stage", "histology"),
  penalty = "MCP",
  gamma = 3.0,
  alpha = 1.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

---

## 3. LASSO Penalty (Comparison)

Include LASSO for comparison — it typically selects more variables but with
higher coefficient bias.

```{r lasso-comparison}
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin", "wbc",
                 "neutrophil_ratio", "platelet", "gender",
                 "n_stage", "histology"),
  penalty = "lasso",
  alpha = 1.0,
  cv_folds = 10,
  lambda_type = "min",
  standardize = TRUE,
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

**Compare**: SCAD and MCP should have less biased coefficients for age, tumor_diameter,
ldh, cea, and t_stage than LASSO, while still setting noise variables to zero.

---

## 4. Gamma Parameter

The gamma parameter controls how quickly the penalty flattens:
- **SCAD**: gamma must be > 2. Default 3.7 (Fan & Li recommendation)
- **MCP**: gamma must be > 1. Default 3.0. Lower gamma = more aggressive selection

```{r gamma-scad-low}
# SCAD with lower gamma (more aggressive)
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  gamma = 2.5,
  cv_folds = 10,
  lambda_type = "min",
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r gamma-scad-high}
# SCAD with higher gamma (closer to LASSO behavior)
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  gamma = 6.0,
  cv_folds = 10,
  lambda_type = "min",
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

---

## 5. Lambda Selection: min vs 1se

```{r lambda-min}
# Minimum CV error — selects more variables
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  lambda_type = "min",
  cv_folds = 10,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r lambda-1se}
# One SE rule — more parsimonious
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  lambda_type = "1se",
  cv_folds = 10,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

---

## 6. Sparse Scenario (Many Noise Variables)

```{r sparse-data}
# 100 patients, 30 variables, only 4 truly important
set.seed(123)
n3 <- 100
p3 <- 30

X <- matrix(rnorm(n3 * p3), nrow = n3)
colnames(X) <- paste0("X", 1:p3)

# Only 4 variables have true effects
beta_true <- rep(0, p3)
beta_true[c(1, 8, 15, 23)] <- c(1.0, -0.8, 0.6, -0.7)

lp3 <- X %*% beta_true
surv3 <- rweibull(n3, 1.2, 24 * exp(-as.numeric(lp3) * 0.2))
cens3 <- runif(n3, 6, 36)
time3 <- pmax(pmin(surv3, cens3), 0.1)
event3 <- factor(ifelse(surv3 <= cens3, "1", "0"))

data_sparse <- data.frame(time = time3, event = event3, as.data.frame(X))

cat("N:", n3, "Events:", sum(event3 == "1"),
    "True signals: X1, X8, X15, X23\n")
```

```{r sparse-scad}
# SCAD should recover the 4 true variables
ncvregcox(
  data = data_sparse,
  time = "time",
  event = "event",
  covariates = paste0("X", 1:30),
  penalty = "SCAD",
  gamma = 3.7,
  cv_folds = 10,
  lambda_type = "min",
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

```{r sparse-mcp}
# MCP comparison
ncvregcox(
  data = data_sparse,
  time = "time",
  event = "event",
  covariates = paste0("X", 1:30),
  penalty = "MCP",
  gamma = 3.0,
  cv_folds = 10,
  lambda_type = "min",
  plot_path = TRUE,
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

---

## 7. Elastic Net Mixing

Combine SCAD/MCP with L2 penalty (alpha < 1) for correlated predictors.

```{r elastic}
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "bmi", "tumor_diameter", "ldh", "cea",
                 "t_stage", "crp", "albumin"),
  penalty = "SCAD",
  alpha = 0.7,  # Mix with L2 penalty
  cv_folds = 10,
  lambda_type = "min",
  plot_cv = TRUE,
  variable_importance = TRUE
)
```

---

## 8. Standardization

```{r no-standardize}
# Without standardization (not recommended for mixed-scale variables)
ncvregcox(
  data = data_clinical,
  time = "time",
  event = "event",
  covariates = c("age", "tumor_diameter", "ldh", "cea"),
  penalty = "SCAD",
  standardize = FALSE,
  cv_folds = 10,
  lambda_type = "min",
  variable_importance = TRUE
)
```

---

## Interpreting Results

### Selected Variables Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name |
| Coefficient | Log hazard ratio (nearly unbiased with SCAD/MCP) |
| Hazard Ratio | exp(coefficient) |
| Selected | Non-zero coefficient indicator |

### Key Advantage Over LASSO

For a true coefficient of 0.8:
- LASSO might estimate 0.5 (biased downward)
- SCAD/MCP might estimate 0.78 (nearly unbiased)

This matters for hazard ratio interpretation in clinical settings.

### Variable Importance

Variables ranked by absolute coefficient value. With SCAD/MCP, the ranking
is more reliable than LASSO because coefficients are less biased.

---

## SCAD vs MCP: Which to Choose?

| Criterion | SCAD | MCP |
|-----------|------|-----|
| Bias for large effects | Very low | Very low |
| Selection aggressiveness | Moderate | More aggressive |
| Stability | Good | Slightly less stable |
| Recommended gamma | 3.7 (default) | 3.0 (default) |
| Publication standard | More established | Gaining adoption |

**Practical recommendation**: Start with SCAD (gamma=3.7). Switch to MCP if
you want sparser selection. Use LASSO as comparison baseline.

---

## Common Pitfalls

1. **Setting gamma too low for SCAD**: gamma < 2.5 makes optimization unstable.
   Stick with the default 3.7 unless you have a specific reason to change it.

2. **Comparing SCAD/MCP coefficients directly to LASSO**: SCAD/MCP coefficients
   are less biased. A SCAD coefficient of 0.8 is not "bigger" than a LASSO
   coefficient of 0.5 for the same variable — the LASSO version is just more shrunk.

3. **Not reporting the penalty type**: Always specify whether you used SCAD, MCP,
   or LASSO, along with the gamma value and lambda selection method.

4. **Using SCAD/MCP for very high dimensional data (p >> 10n)**: In extremely
   high-dimensional settings, LASSO or Elastic Net may be more stable.
   SCAD/MCP work best when p is moderately large relative to n.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** | Standard LASSO with suitability assessment |
| **Adaptive LASSO** | Two-stage adaptive weights approach |
| **High-Dimensional Cox** | Multiple regularization methods unified |
| **PLS Cox** | Dimensionality reduction for very high p |
| **Multivariable Survival** | Standard Cox, no regularization needed |

---

## References

- Fan J, Li R. Variable selection via nonconcave penalized likelihood and its oracle properties. *J Am Stat Assoc*. 2001;96(456):1348-1360.
- Zhang CH. Nearly unbiased variable selection under minimax concave penalty. *Ann Stat*. 2010;38(2):894-942.
- Breheny P, Huang J. Coordinate descent algorithms for nonconvex penalized regression, with applications to biological feature selection. *Ann Appl Stat*. 2011;5(1):232-253.
- Fan J, Li R. Variable selection for Cox's proportional hazards model and frailty model. *Ann Stat*. 2002;30(1):74-99.
