---
title: "Relative Survival Analysis - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Relative Survival Analysis - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `relativesurvival()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the wrapper function directly.

# Relative Survival Analysis

## Overview

Relative survival analysis compares observed survival in a patient cohort to the
survival expected in a matched general population. Rather than requiring (often
inaccurate) cause-of-death data, it estimates disease-specific mortality indirectly
through the ratio of observed to expected survival. This approach is the standard
in population-based cancer epidemiology and is used by EUROCARE, CONCORD, and
national cancer registries worldwide.

The `relativesurvival` module provides:

- **Four estimation methods**: Pohar-Perme (recommended), Ederer I, Ederer II, Hakulinen
- **15 population rate tables**: 4 from the `relsurv` package (US, Minnesota, France, Slovenia) and 10 WHO-based tables (Turkey, Germany, UK, Italy, Japan, Spain, Brazil, South Korea, China, India), plus a custom option
- **Net survival** and **excess mortality** estimation
- **Crude probability of death** decomposition (disease vs. other causes)
- **ICSS age standardization** for international comparisons
- **Period analysis** to track survival trends over diagnosis years
- **Regression models**: additive excess hazard, multiplicative, and flexible parametric (via `rstpm2`)
- **Four publication-ready plots**: observed, expected, relative survival, and excess mortality

---

## Datasets

| Dataset | N | Events | Key Features |
|---------|---|--------|-------------|
| `relativesurvival_test` | 200 | ~116 deaths | 4 cancer sites (Colon, Breast, Lung, Prostate), stages I-IV, ages 30-90, diagnosis years 2000-2015, covariates (comorbidity, tumor_size) |

```{r data-overview}
data(relativesurvival_test)
str(relativesurvival_test)
```

```{r data-summary}
summary(relativesurvival_test[, c("followup_years", "vital_status",
                                   "age_at_diagnosis", "sex", "diagnosis_year")])
```

---

## 1. Basic Analysis

The minimum call requires five variables: follow-up time, vital status, age at
diagnosis, sex, and calendar year of diagnosis. By default, the module uses the
**Pohar-Perme** estimator and the **US population** rate table.

This produces the main survival table (observed, expected, relative survival at
1, 3, 5, and 10 years), net survival table, excess mortality rates, crude
probability of death, and all four plots.

```{r basic}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year"
)
```

### What the outputs mean

- **Survival Estimates by Time**: Observed (Kaplan-Meier), expected (population), and relative (net) survival at each requested timepoint.
- **Net Survival Estimates**: The probability of surviving the disease if other causes of death were removed. Values near 1.0 mean the disease contributes little to mortality.
- **Excess Mortality Rates**: The additional hazard attributable to the disease in each year interval. Higher values indicate more disease-specific mortality.
- **Crude Probability of Death**: Decomposes total mortality into disease-related and other-cause components, accounting for competing risks.
- **Clinical Interpretation**: Automated 5-year prognosis grading (Excellent >90%, Good >70%, Fair >50%, Poor <50%).

---

## 2. Estimation Methods

The module supports four estimation methods, each differing in how expected
survival is calculated. The **Pohar-Perme** method (default) is the only
unbiased estimator of net survival and is recommended by international guidelines.

### 2a. Pohar-Perme (default, recommended)

Pohar-Perme uses inverse-probability-of-censoring weighting (IPCW) to produce
unbiased net survival estimates. This is the EUROCARE/CONCORD standard.

```{r method-poharperme}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  method = "poharperme"
)
```

### 2b. Ederer II

Ederer II updates the expected survival at each event time. It was the traditional
standard but is now known to be biased for net survival.

```{r method-ederer2}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  method = "ederer2"
)
```

### 2c. Ederer I

Ederer I calculates expected survival based on the full cohort demographics at
baseline, without updating. It tends to overestimate expected survival at later times.

```{r method-ederer1}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  method = "ederer1"
)
```

### 2d. Hakulinen

Hakulinen weights expected survival by the censoring distribution, making it a
compromise between Ederer I and Ederer II.

```{r method-hakulinen}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  method = "hakulinen"
)
```

---

## 3. Rate Tables

Population rate tables are the source of expected survival. The choice of rate
table should match the country and calendar period of the patient cohort.

### 3a. US Population (default)

The `survexp.us` table from the `survival` package, repackaged by `relsurv`.

```{r ratetable-us}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  ratetable = "us"
)
```

### 3b. Minnesota Population

A state-level rate table useful for regional cancer registries.

```{r ratetable-mn}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  ratetable = "mn"
)
```

### 3c. French Population

```{r ratetable-fr}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  ratetable = "fr"
)
```

### 3d. Slovenian Population

```{r ratetable-slovenia}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  ratetable = "slovenia"
)
```

### 3e. WHO-Based Tables (Turkey example)

WHO-based tables use Global Health Observatory life table data bundled with this
module. They cover Turkey, Germany, UK, Italy, Japan, Spain, Brazil, South Korea,
China, and India. If the table is not available, the module falls back to the US
table with a notice.

```{r ratetable-turkey}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  ratetable = "turkey"
)
```

---

## 4. Net Survival & Excess Mortality

By default, both **net survival** (`net_survival = TRUE`) and **excess mortality**
(`excess_mortality = TRUE`) are calculated. You can toggle each independently.

### 4a. Net Survival Only

Net survival is the probability of surviving the disease in a hypothetical world
where the disease is the only possible cause of death.

```{r net-survival-only}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  net_survival = TRUE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_excess = FALSE
)
```

### 4b. Excess Mortality Only

Excess mortality is the additional hazard due to the disease, computed as
`-log(S_net(t) / S_net(t-1))` for each yearly interval.

```{r excess-mortality-only}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  net_survival = FALSE,
  excess_mortality = TRUE,
  crude_probability = FALSE
)
```

---

## 5. Crude Probabilities

Crude probabilities decompose total mortality into disease-specific and other-cause
components, using the method of Cronin and Feuer (implemented in `relsurv::cmp.rel`).
This accounts for competing risks.

```{r crude-probability}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = TRUE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 6. Age Standardization

Age-standardized relative survival uses **ICSS (International Cancer Survival
Standard)** weights to make survival estimates comparable across populations
with different age structures.

ICSS weight groups:
- 0-44 years: 0.07
- 45-54 years: 0.12
- 55-64 years: 0.23
- 65-74 years: 0.29
- 75+ years: 0.29

The module calculates Pohar-Perme net survival within each age group, then
produces a weighted average.

```{r age-standardized}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  age_standardized = TRUE,
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 7. Period Analysis

Period analysis estimates the most recent survival experience by restricting
follow-up to a specific calendar window. Unlike cohort analysis (which follows
patients from diagnosis), period analysis captures the latest treatment effects.

### 7a. Default Period Analysis

Without a cohort definition, the module creates 5-year diagnosis periods
automatically from the data range and computes 5-year relative survival for each
period.

```{r period-analysis}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  period_analysis = TRUE,
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 7b. With Cohort Definition

The `cohort_year` option restricts the analysis to patients diagnosed within a
specific year range. The format is "start-end" (e.g., "2005-2015").

```{r period-analysis-cohort}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  period_analysis = TRUE,
  cohort_year = "2005-2015",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 8. Regression Models

When additional covariates are provided, the module can fit three types of
regression models to assess the effect of prognostic factors on excess mortality.

### 8a. Additive Excess Hazard Model

The additive model (`relsurv::rsadd`) assumes that excess hazard is additive
to the expected hazard. This is the classic Esteve model.

```{r regression-additive}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  covariates = "cancer_site",
  regression_model = "additive",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 8b. Multiplicative Model

The multiplicative model (`relsurv::rsmul`) assumes that excess hazard is
proportional (multiplicative) to baseline excess hazard.

```{r regression-multiplicative}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  covariates = "cancer_site",
  regression_model = "multiplicative",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 8c. Flexible Parametric Model

The flexible parametric model (`rstpm2::stpm2`) uses restricted cubic splines
to model the baseline excess hazard. The `spline_df` option controls the
degrees of freedom (complexity) of the spline.

```{r regression-flexible}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  covariates = "cancer_site",
  regression_model = "flexible",
  spline_df = 3,
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 8d. Multiple Covariates

You can include multiple covariates (continuous and/or categorical).

```{r regression-multiple-covariates}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  covariates = c("cancer_site", "stage"),
  regression_model = "additive",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 9. Time Scales

The module accepts follow-up time in years (default), months, or days. The
`time_scale` option tells the module how to interpret the time variable; it is
internally converted to years (and then to days for the `relsurv` rate table).

### 9a. Time in Months

```{r time-scale-months}
# Create a copy with follow-up in months
test_months <- relativesurvival_test
test_months$followup_months <- test_months$followup_years * 12

relativesurvival(
  data = test_months,
  time = "followup_months",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  time_scale = "months",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 9b. Time in Days

```{r time-scale-days}
# Create a copy with follow-up in days
test_days <- relativesurvival_test
test_days$followup_days <- test_days$followup_years * 365.25

relativesurvival(
  data = test_days,
  time = "followup_days",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  time_scale = "days",
  net_survival = FALSE,
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 10. Custom Timepoints & Confidence Levels

### 10a. Custom Timepoints

The `timepoints` option accepts a comma-separated string of time values (in
years, regardless of the `time_scale` setting). These determine at which
follow-up times the survival estimates are reported.

```{r custom-timepoints}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  timepoints = "0.5, 1, 2, 3, 5, 7",
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 10b. Custom Confidence Level

The `confidence_level` option controls the width of all confidence intervals
in the analysis (default 0.95). Here we use 90%.

```{r custom-confidence}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  confidence_level = 0.90,
  timepoints = "1, 3, 5",
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

---

## 11. Edge Cases

### 11a. Small Sample Warning

The module requires at least 30 complete cases and at least 10 events. With
10-19 events, a strong warning is displayed; with 20-49, a moderate warning.

```{r edge-small-sample}
# Subset to a small sample with few events
small_data <- relativesurvival_test[1:40, ]
relativesurvival(
  data = small_data,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  excess_mortality = FALSE,
  crude_probability = FALSE,
  plot_observed = FALSE,
  plot_expected = FALSE,
  plot_relative = FALSE,
  plot_excess = FALSE
)
```

### 11b. Full Analysis with All Options Enabled

This demonstrates the maximum output: all analysis options, all plots, period
analysis with cohort definition, age standardization, and an additive regression
model.

```{r edge-full-analysis}
relativesurvival(
  data = relativesurvival_test,
  time = "followup_years",
  status = "vital_status",
  age = "age_at_diagnosis",
  sex = "sex",
  year = "diagnosis_year",
  covariates = c("cancer_site", "stage"),
  ratetable = "us",
  method = "poharperme",
  time_scale = "years",
  net_survival = TRUE,
  excess_mortality = TRUE,
  crude_probability = TRUE,
  age_standardized = TRUE,
  period_analysis = TRUE,
  cohort_year = "2000-2015",
  regression_model = "additive",
  plot_observed = TRUE,
  plot_expected = TRUE,
  plot_relative = TRUE,
  plot_excess = TRUE,
  confidence_level = 0.95,
  timepoints = "1, 3, 5, 10"
)
```

---

## Complete Option Reference

| # | Option | Type | Default | Range/Choices | Section Demonstrated |
|---|--------|------|---------|---------------|---------------------|
| 1 | `time` | Variable (numeric) | -- | continuous | 1. Basic Analysis |
| 2 | `status` | Variable (factor/numeric) | -- | binary 0/1 | 1. Basic Analysis |
| 3 | `age` | Variable (numeric) | -- | 0-120 | 1. Basic Analysis |
| 4 | `sex` | Variable (factor) | -- | male/female mappings | 1. Basic Analysis |
| 5 | `year` | Variable (numeric) | -- | 1900-2100 | 1. Basic Analysis |
| 6 | `covariates` | Variables (numeric/factor) | -- | any | 8. Regression Models |
| 7 | `ratetable` | List | `us` | us/mn/fr/slovenia/turkey/germany/uk/italy/japan/spain/brazil/south_korea/china/india/custom | 3. Rate Tables |
| 8 | `method` | List | `poharperme` | poharperme/ederer1/ederer2/hakulinen | 2. Estimation Methods |
| 9 | `time_scale` | List | `years` | years/months/days | 9. Time Scales |
| 10 | `net_survival` | Bool | `true` | -- | 4. Net Survival |
| 11 | `excess_mortality` | Bool | `true` | -- | 4. Excess Mortality |
| 12 | `crude_probability` | Bool | `true` | -- | 5. Crude Probabilities |
| 13 | `age_standardized` | Bool | `false` | -- | 6. Age Standardization |
| 14 | `period_analysis` | Bool | `false` | -- | 7. Period Analysis |
| 15 | `cohort_year` | String | `""` | e.g., "2010-2015" | 7b. Cohort Definition |
| 16 | `regression_model` | List | `none` | none/additive/multiplicative/flexible | 8. Regression Models |
| 17 | `spline_df` | Integer | `4` | 1-10 | 8c. Flexible Parametric |
| 18 | `plot_observed` | Bool | `true` | -- | 1. Basic Analysis |
| 19 | `plot_expected` | Bool | `true` | -- | 1. Basic Analysis |
| 20 | `plot_relative` | Bool | `true` | -- | 1. Basic Analysis |
| 21 | `plot_excess` | Bool | `true` | -- | 4b. Excess Mortality |
| 22 | `confidence_level` | Number | `0.95` | 0.50-0.99 | 10b. Custom Confidence |

All 22 `.a.yaml` options are covered in the examples above.

---

## References

1. Pohar Perme M, Stare J, Esteve J. On Estimation in Relative Survival. *Biometrics*, 2012;68:113-120.
2. Ederer F, Axtell LM, Cutler SJ. The Relative Survival Rate: A Statistical Methodology. *NCI Monograph*, 1961;6:101-121.
3. Hakulinen T. Cancer Survival Corrected for Heterogeneity in Patient Withdrawal. *Biometrics*, 1982;38:933-942.
4. Corazziari I, Quinn M, Capocaccia R. Standard cancer patient population for age standardising survival ratios. *Eur J Cancer*, 2004;40:2307-2316.
5. Dickman PW, Coviello E. Estimating and Modeling Relative Survival. *The Stata Journal*, 2015;15(1):186-215.
6. Lambert PC, Royston P. Further Development of Flexible Parametric Models for Survival Analysis. *The Stata Journal*, 2009;9(2):265-290.
7. Cronin KA, Feuer EJ. Cumulative Cause-Specific Mortality for Cancer Patients in the Presence of Other Causes: A Crude Analogue of Relative Survival. *Statistics in Medicine*, 2000;19:1729-1740.

---

*This vignette is part of the [jsurvival](https://www.serdarbalci.com/jsurvival/) module of the ClinicoPath jamovi package.*
