---
title: "Adaptive LASSO Cox Regression for Variable Selection in Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive LASSO Cox Regression for Variable Selection in Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `adaptivelasso()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `adaptivelassoClass$new(options = adaptivelassoOptions$new(...), data = mydata)`.

# Adaptive LASSO Cox Regression

## Overview

The **Adaptive LASSO Cox Regression** module (`adaptivelasso`) performs variable selection
in survival analysis using adaptively weighted L1-penalized Cox proportional hazards
regression. Unlike standard LASSO, adaptive LASSO uses data-driven penalty weights
to achieve the **oracle property** — consistent variable selection and asymptotically
unbiased coefficient estimates for truly non-zero variables.

Key features:

- **Multiple weight methods** (Ridge, Univariate Cox, Full Cox, Correlation, Equal)
- **Cross-validated lambda selection** with multiple performance measures
- **Stability selection** for robust variable identification
- **Model diagnostics** (proportional hazards test, goodness of fit)
- **Risk group stratification** with Kaplan-Meier curves
- **Comprehensive visualizations** (regularization path, CV curve, stability plot)

---

## What is Adaptive LASSO?

Standard LASSO applies the same penalty to all coefficients, which can lead to:
- **Inconsistent selection**: Some noise variables may be selected
- **Biased estimates**: True non-zero coefficients are over-shrunk

Adaptive LASSO fixes this by using **variable-specific weights**:

$$\text{Maximize:} \quad \ell(\beta) - \lambda \sum_{j=1}^{p} w_j |\beta_j|$$

where $w_j = 1/|\hat{\beta}_j^{(0)}|^\gamma$ are weights based on initial estimates
$\hat{\beta}_j^{(0)}$.

**Effect**: Variables with large initial coefficients get small penalties (kept in model),
while variables with small initial estimates get large penalties (shrunk to zero).

### Weight Methods Comparison

| Method | Initial Estimate | Best For |
|--------|-----------------|----------|
| **Ridge** (default) | Ridge regression coefficients | Stable weights, handles collinearity |
| **Univariate** | Individual Cox regressions | Fast, works with p > n |
| **Full Cox** | Standard Cox model | Accurate when p < n |
| **Correlation** | Marginal survival correlations | Very fast screening |
| **Equal** | All weights = 1 (standard LASSO) | Comparison baseline |

### LASSO vs Adaptive LASSO

| Property | Standard LASSO | Adaptive LASSO |
|----------|---------------|----------------|
| Oracle property | No | Yes |
| Coefficient bias | Higher | Lower for true signals |
| Selection consistency | Not guaranteed | Consistent under mild conditions |
| Computation | One-stage | Two-stage (initial + penalized) |
| Best for | Quick screening | Publishable models |

---

## When to Use Adaptive LASSO

| Scenario | Recommendation |
|----------|---------------|
| Building a prognostic model for publication | **Adaptive LASSO recommended** |
| Many predictors (p >= n/3) | **Use Ridge weights** |
| Correlated predictors | **Use Ridge weights + alpha < 1** |
| Quick variable screening | Standard LASSO may suffice |
| Very few predictors (p < 5) | Standard Cox regression |
| Need robust selection | Enable **stability selection** |

---

## Datasets Used in This Guide

| Dataset | N | Events | Predictors | Description |
|---------|---|--------|------------|-------------|
| Synthetic "standard" | 180 | ~100 | 12 mixed | Clinical study with known effects |
| Synthetic "small" | 60 | ~25 | 8 | Small sample edge case |
| Synthetic "collinear" | 120 | ~55 | 6 (2 correlated pairs) | Collinearity testing |

---

## 1. Basic Analysis with Ridge Weights

The default configuration uses Ridge regression weights for adaptive penalties.

```{r basic-ridge}
# Create synthetic clinical data
set.seed(42)
n <- 180

# Clinical variables
age <- rnorm(n, 60, 12)
tumor_size <- rlnorm(n, log(3), 0.6)
ki67 <- pmin(pmax(rnorm(n, 25, 15), 1), 95)
stage_num <- sample(1:4, n, replace = TRUE, prob = c(0.2, 0.35, 0.3, 0.15))
stage <- factor(c("I", "II", "III", "IV")[stage_num])
grade <- factor(sample(c("Low", "Moderate", "High"), n, replace = TRUE,
                       prob = c(0.25, 0.45, 0.30)))
gender <- factor(sample(c("Male", "Female"), n, replace = TRUE))
treatment <- factor(sample(c("Surgery", "ChemoRT", "Combined"), n, replace = TRUE,
                          prob = c(0.4, 0.3, 0.3)))

# True survival model: age, stage, ki67, treatment affect survival
linear_pred <- 0.03 * (age - 60) +
               0.4 * (stage_num - 1) +
               0.02 * (ki67 - 25) +
               ifelse(treatment == "Combined", -0.5, 0)

# Generate survival times (Weibull)
baseline_time <- 36
time_raw <- rweibull(n, shape = 1.2, scale = baseline_time * exp(-linear_pred * 0.3))
censor_time <- runif(n, 12, 60)
time <- pmin(time_raw, censor_time)
time <- pmax(time, 0.1)
event <- factor(ifelse(time_raw <= censor_time, "Event", "Censored"),
                levels = c("Censored", "Event"))

# Noise variables
hemoglobin <- rnorm(n, 13, 2)
crp <- rlnorm(n, log(5), 0.8)
albumin <- rnorm(n, 4.0, 0.5)
center <- factor(sample(c("CenterA", "CenterB", "CenterC"), n, replace = TRUE))

data_clinical <- data.frame(
  follow_up = time, event = event,
  age = age, tumor_size = tumor_size, ki67 = ki67,
  stage = stage, grade = grade, gender = gender,
  treatment = treatment, hemoglobin = hemoglobin,
  crp = crp, albumin = albumin, center = center
)

cat("Event rate:", mean(event == "Event"), "\n")
cat("N:", n, "Events:", sum(event == "Event"), "\n")
```

```{r run-basic}
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin", "crp", "albumin"),
  weight_method = "ridge",
  alpha = 1.0,
  gamma = 1.0,
  cv_folds = 10,
  cv_measure = "deviance",
  show_coefficients = TRUE,
  show_selection_path = TRUE,
  show_cv_results = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE,
  plot_survival_curves = TRUE
)
```

Expected: Adaptive LASSO should select age, stage, ki67, and treatment as important
predictors. Noise variables (hemoglobin, albumin) should be eliminated.

---

## 2. Comparing Weight Methods

Different weight calculation methods suit different data scenarios.

```{r weight-univariate}
# Univariate Cox weights — good for p > n scenarios
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin", "crp", "albumin"),
  weight_method = "univariate",
  alpha = 1.0,
  gamma = 1.0,
  cv_folds = 10,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  plot_selection_path = TRUE,
  plot_cv_curve = TRUE
)
```

```{r weight-equal}
# Equal weights = standard LASSO (for comparison)
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin", "crp", "albumin"),
  weight_method = "equal",
  alpha = 1.0,
  cv_folds = 10,
  show_coefficients = TRUE,
  show_cv_results = TRUE,
  plot_cv_curve = TRUE
)
```

**Interpretation**: Compare the selected variables between Ridge-weighted, Univariate-weighted,
and Equal-weighted (standard LASSO) models. Ridge typically provides the most stable selection.

---

## 3. Stability Selection

Stability selection identifies variables that are robustly selected across many
bootstrap subsamples — more reliable than single-model selection.

```{r stability}
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin", "crp", "albumin"),
  weight_method = "ridge",
  alpha = 1.0,
  cv_folds = 10,
  stability_selection = TRUE,
  stability_threshold = 0.6,
  bootstrap_samples = 100,
  subsampling_ratio = 0.8,
  show_coefficients = TRUE,
  plot_stability = TRUE,
  plot_cv_curve = TRUE
)
```

Variables selected in >60% of bootstrap samples are considered stable.
This is the recommended approach for publishable prognostic models.

---

## 4. Elastic Net Mixing (Correlated Predictors)

When predictors are correlated, pure LASSO (alpha=1) arbitrarily selects one
from a correlated group. Setting alpha < 1 adds an L2 penalty to encourage
group selection.

```{r elastic-net}
# Create data with correlated predictors
set.seed(42)
n2 <- 120
base1 <- rnorm(n2)
base2 <- rnorm(n2)

data_corr <- data.frame(
  time = pmax(rweibull(n2, 1.2, 24) * exp(-0.3 * base1), 0.1),
  event = factor(sample(c("Event", "Censored"), n2, replace = TRUE,
                        prob = c(0.55, 0.45)),
                 levels = c("Censored", "Event")),
  var1 = base1,
  var2 = base1 + rnorm(n2, 0, 0.2),  # r ~ 0.98 with var1
  var3 = base2,
  var4 = base2 + rnorm(n2, 0, 0.3),  # r ~ 0.95 with var3
  var5 = rnorm(n2),
  var6 = rnorm(n2)
)

# Pure LASSO (alpha = 1) — may drop one from each correlated pair
adaptivelasso(
  strata = NULL,
  data = data_corr,
  time = "time",
  event = "event",
  predictors = c("var1", "var2", "var3", "var4", "var5", "var6"),
  weight_method = "ridge",
  alpha = 1.0,
  cv_folds = 5,
  show_coefficients = TRUE,
  plot_cv_curve = TRUE
)
```

```{r elastic-net-05}
# Elastic Net (alpha = 0.5) — tends to keep correlated groups together
adaptivelasso(
  strata = NULL,
  data = data_corr,
  time = "time",
  event = "event",
  predictors = c("var1", "var2", "var3", "var4", "var5", "var6"),
  weight_method = "ridge",
  alpha = 0.5,
  cv_folds = 5,
  show_coefficients = TRUE,
  plot_cv_curve = TRUE
)
```

---

## 5. Gamma Parameter Effect

The gamma parameter controls the strength of adaptive penalization. Higher gamma
values create larger differences between penalty weights.

```{r gamma-comparison}
# gamma = 0.5 (mild adaptation)
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin"),
  weight_method = "ridge",
  gamma = 0.5,
  cv_folds = 10,
  show_coefficients = TRUE,
  plot_cv_curve = TRUE
)
```

```{r gamma-high}
# gamma = 2.0 (strong adaptation — more aggressive selection)
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "gender", "treatment", "hemoglobin"),
  weight_method = "ridge",
  gamma = 2.0,
  cv_folds = 10,
  show_coefficients = TRUE,
  plot_cv_curve = TRUE
)
```

**Guidelines**: gamma = 1 is the standard choice. Use gamma = 2 for more aggressive
selection when you suspect many noise variables. Use gamma = 0.5 when you want to
retain more variables for further investigation.

---

## 6. Model Diagnostics

Test proportional hazards assumption and assess model goodness of fit.

```{r diagnostics}
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade",
                 "treatment"),
  weight_method = "ridge",
  cv_folds = 10,
  proportional_hazards = TRUE,
  goodness_of_fit = TRUE,
  influence_diagnostics = TRUE,
  show_coefficients = TRUE,
  show_diagnostics = TRUE,
  plot_diagnostics = TRUE
)
```

---

## 7. Risk Group Stratification

Classify patients into risk groups based on the adaptive LASSO linear predictor.

```{r risk-groups}
# 3 risk groups (default)
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade", "treatment"),
  weight_method = "ridge",
  cv_folds = 10,
  risk_groups = 3,
  time_points = "12, 24, 60",
  baseline_survival = TRUE,
  show_coefficients = TRUE,
  plot_survival_curves = TRUE,
  plot_baseline_hazard = TRUE
)
```

---

## 8. Stratified Analysis

Use stratification when the proportional hazards assumption is violated for
a variable (e.g., center effects).

```{r stratified}
adaptivelasso(
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage", "grade", "treatment"),
  strata = "center",
  weight_method = "ridge",
  cv_folds = 10,
  show_coefficients = TRUE,
  plot_cv_curve = TRUE
)
```

---

## 9. Advanced Options

### Tie handling method

```{r ties}
# Efron method (more accurate with many ties)
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage"),
  weight_method = "ridge",
  tie_method = "efron",
  cv_folds = 10,
  show_coefficients = TRUE
)
```

### Convergence and iterations

```{r convergence}
adaptivelasso(
  strata = NULL,
  data = data_clinical,
  time = "follow_up",
  event = "event",
  predictors = c("age", "tumor_size", "ki67", "stage"),
  weight_method = "ridge",
  convergence_threshold = 1e-8,
  max_iterations = 50000,
  random_seed = 456,
  cv_folds = 10,
  show_coefficients = TRUE
)
```

---

## Interpreting Results

### Coefficient Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name |
| Coefficient | Log hazard ratio (positive = risk, negative = protective) |
| Hazard Ratio | exp(coefficient) — multiplicative effect on hazard |
| Adaptive Weight | Penalty weight used (lower = less penalized) |
| Selected | Whether variable survived regularization |

### Stability Selection Results

| Metric | Meaning |
|--------|---------|
| Selection Frequency | Proportion of bootstrap samples selecting the variable |
| Stable | Whether frequency exceeds the stability threshold |
| Rank | Variables ranked by selection frequency |

---

## Common Pitfalls

1. **Using Full Cox weights when p > n**: Full Cox model fails when predictors exceed
   observations. Use Ridge or Univariate weights instead.

2. **Ignoring weight method choice**: Ridge is the safest default. Univariate is faster
   but ignores variable interactions in the initial estimates.

3. **Setting gamma too high**: gamma > 3 can over-penalize moderate effects, leading to
   missed true signals.

4. **Not using stability selection for publication**: Single-run variable selection is
   unstable. Stability selection provides more robust and reproducible results.

5. **Ignoring proportional hazards**: The adaptive LASSO Cox model assumes proportional
   hazards. Check this assumption with the built-in diagnostics.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** | Standard LASSO without adaptive weights (simpler, faster) |
| **SCAD Cox** | Non-convex penalties (SCAD/MCP) for unbiased estimation |
| **High-Dimensional Cox** | Multiple regularization methods in one interface |
| **PLS Cox** | Dimensionality reduction approach (PCA/PLS before Cox) |
| **Multivariable Survival** | Standard Cox with no regularization needed |

---

## References

- Zou H. The adaptive lasso and its oracle properties. *J Am Stat Assoc*. 2006;101(476):1418-1429.
- Zhang HH, Lu W. Adaptive lasso for Cox's proportional hazards model. *Biometrika*. 2007;94(3):691-703.
- Meinshausen N, Buhlmann P. Stability selection. *J R Stat Soc Ser B*. 2010;72(4):417-473.
- Tibshirani R. The lasso method for variable selection in the Cox model. *Stat Med*. 1997;16(4):385-395.
