---
title: "Overall, Cause Specific, and Competing Survival - Comprehensive Guide"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Overall, Cause Specific, and Competing Survival - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** `competingsurvival()` is designed primarily for the jamovi GUI. The R syntax
> shown here is for reference and advanced scripting.

# Competing Risks Survival Analysis

## Overview

In clinical research, patients are often at risk of dying from multiple causes.
Standard Kaplan-Meier analysis treats competing events (e.g., death from other
causes) as censored observations, which **overestimates** the actual probability
of the event of interest. Competing risks methods address this bias.

**Key concepts:**

- **Kaplan-Meier / Overall Survival**: All deaths are events. Simple and
  familiar, but does not distinguish cause.
- **Cause-Specific Survival**: Only deaths from the disease of interest are
  events; competing deaths are censored. Estimates the cause-specific hazard
  but can overestimate cumulative incidence.
- **Cumulative Incidence Function (CIF)**: The proper probability of
  experiencing the event of interest in the presence of competing risks.
  Always <= 1 minus KM estimate.
- **Gray's Test**: The competing-risks analog of the log-rank test. Compares
  CIF curves between groups.
- **Fine-Gray Model**: Subdistribution hazard regression. Unlike cause-specific
  Cox, it directly models the CIF, making hazard ratios interpretable in terms
  of cumulative incidence.

## Datasets

This vignette uses two bundled datasets:

| Dataset | N | Outcome Levels | Key Grouping Variables |
|---------|---|----------------|------------------------|
| `competing_survival_data` | 300 | Dead_of_Disease (115), Dead_of_Other (41), Alive_w_Disease (57), Alive_wo_Disease (87) | Treatment (Experimental_Therapy / Standard_Care), Tumor_Stage (I-IV), Sex |
| `survival_competing` | 180 | Dead of Disease (53), Dead of Other (37), Alive w Disease (35), Alive w/o Disease (55) | treatment (Experimental / Standard), stage (I-IV), molecular_subtype (Type A/B/C) |

```{r load-data}
data("competing_survival_data")
data("survival_competing")

cat("--- competing_survival_data ---\n")
cat("Dimensions:", nrow(competing_survival_data), "x", ncol(competing_survival_data), "\n")
cat("Outcome distribution:\n")
print(table(competing_survival_data$Outcome))

cat("\n--- survival_competing ---\n")
cat("Dimensions:", nrow(survival_competing), "x", ncol(survival_competing), "\n")
cat("Outcome distribution:\n")
print(table(survival_competing$outcome))
```

## 1. Overall Survival

Overall survival treats **all deaths** (disease + other causes) as events.
Alive patients are censored. This is the broadest survival measure.

```{r overall-survival}
result_os <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "overall",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

With 156 total deaths out of 300 patients, this gives a straightforward hazard
ratio for Treatment comparing all-cause mortality.

### Overall Survival Without a Group Variable

When no explanatory variable is provided, the function reports median survival
for the entire cohort.

```{r overall-no-group}
result_os_nogroup <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  analysistype = "overall",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

## 2. Cause-Specific Survival

Cause-specific survival focuses on **disease-related deaths only**. Deaths from
other causes are treated as censored. This estimates the cause-specific hazard
but can overestimate the probability of disease death when competing risks are
substantial.

```{r cause-specific}
result_css <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "cause",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

Only 115 disease deaths are counted as events here. The 41 deaths from other
causes are censored, which means cause-specific survival is always >= overall
survival.

## 3. Competing Risks (CIF)

The cumulative incidence function properly accounts for competing events. Unlike
1-KM, CIF never overestimates the probability of disease death.

```{r competing-risks}
result_cr <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

The `cuminc` table shows the cumulative incidence of disease death and competing
death at the default timepoints (12, 24, 36, 60 months). The `comprisksPlot`
displays the CIF curves.

### Competing Risks with the Second Dataset

The `survival_competing` dataset uses different naming conventions for outcome
levels, demonstrating that the function handles various label formats.

```{r competing-risks-dataset2}
result_cr2 <- competingsurvival(
  data = survival_competing,
  overalltime = "elapsedtime",
  outcome = "outcome",
  explanatory = "treatment",
  analysistype = "compete",
  dod = "Dead of Disease",
  dooc = "Dead of Other",
  awd = "Alive w Disease",
  awod = "Alive w/o Disease"
)
```

## 4. Gray's Test

Gray's test compares cumulative incidence functions between groups. It is the
competing-risks analog of the log-rank test. Unlike the log-rank test, it tests
equality of the subdistribution hazard (i.e., the CIF) rather than the
cause-specific hazard.

```{r grays-test}
result_gray <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  graystest = TRUE
)
```

The summary output includes the Gray's test chi-squared statistic, degrees of
freedom, and p-value for each event type.

### Gray's Test with a Multi-Level Group Variable

Using Tumor_Stage (4 levels) to test whether CIF differs across stages:

```{r grays-test-multilevel}
result_gray_stage <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Tumor_Stage",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  graystest = TRUE
)
```

## 5. Fine-Gray Subdistribution Hazard

The Fine-Gray model estimates the subdistribution hazard ratio (SHR). Unlike
cause-specific Cox regression, the SHR directly relates to the CIF: a higher
SHR means a higher cumulative incidence of the event over time.

```{r fine-gray}
result_fg <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  subdistribution = TRUE
)
```

The `fineGrayTable` shows the coefficient, hazard ratio, confidence interval,
standard error, z-statistic, and p-value. The `survivalTable` also gains a
row for "Subdistribution HR (Fine-Gray)" alongside the standard competing
risks result.

### Fine-Gray with the Second Dataset

```{r fine-gray-dataset2}
result_fg2 <- competingsurvival(
  data = survival_competing,
  overalltime = "elapsedtime",
  outcome = "outcome",
  explanatory = "treatment",
  analysistype = "compete",
  dod = "Dead of Disease",
  dooc = "Dead of Other",
  awd = "Alive w Disease",
  awod = "Alive w/o Disease",
  subdistribution = TRUE
)
```

## 6. Stacked CIF Plot

The stacked probability plot shows how the total probability is partitioned at
each time point into three components: CIF for disease death, CIF for competing
death, and survival probability. They must sum to 100%.

```{r stacked-plot}
result_stacked <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  showStackedPlot = TRUE,
  cifColors = "colorblind"
)
```

At time 0, survival is 100% and both CIFs are 0%. As time progresses, survival
decreases while the cumulative incidence of disease death and competing death
increase, always summing to 100%.

## 7. KM vs CIF Comparison

This plot directly demonstrates why competing risks methods are necessary. The
naive 1-KM estimate (which treats competing deaths as censored) overestimates
the cumulative incidence of disease death compared to the proper CIF.

```{r km-vs-cif}
result_kmcif <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  showKMvsCIF = TRUE
)
```

The dashed line (1-KM) always lies above the solid line (CIF). The gap between
them represents the bias introduced by ignoring competing events. This
difference is clinically meaningful in older populations or when competing
causes of death are frequent.

## 8. Custom Timepoints and Confidence Level

### Custom CIF Timepoints

By default, cumulative incidence is estimated at 12, 24, 36, and 60 months.
Custom timepoints can be specified as a comma-separated string.

```{r custom-timepoints}
result_tp <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  timepoints = "6,12,18,24,30,36,48,60"
)
```

### Adjusting Confidence Level

The default confidence level is 0.95 (95%). You can change this to, for
example, 0.90 for narrower intervals.

```{r confidence-level}
result_cl90 <- competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease",
  subdistribution = TRUE,
  confidencelevel = 0.90
)
```

The Fine-Gray table and timepoint confidence intervals will reflect the 90%
confidence level.

## 9. Assumptions and Interpretation

### Key Assumptions

1. **Independent censoring**: Censoring must be independent of the event
   process. If patients are censored because they are sicker (or healthier),
   all estimates will be biased.

2. **Correct event classification**: Each patient must be correctly classified.
   Misclassifying a disease death as death from other causes (or vice versa)
   directly biases CIF estimates.

3. **Complete case analysis**: Observations with missing time, outcome, or
   grouping variables are excluded. Report the proportion of missing data.

4. **Adequate follow-up**: CIF estimates beyond the maximum observed time are
   unreliable. Late timepoints should be interpreted cautiously.

5. **Fine-Gray model interpretation**: The subdistribution HR has a different
   meaning than the cause-specific HR. It reflects both the direct effect on
   the event and the indirect effect through competing events. An SHR > 1
   means a group has a higher cumulative incidence over time, which is often
   the most clinically relevant quantity.

6. **Gray's test**: Tests equality of CIF curves, not cause-specific hazards.
   It is appropriate when the research question concerns cumulative incidence
   (probability of event), not the instantaneous rate (hazard).

### When to Use Which Approach

| Research Question | Method | Analysis Type |
|-------------------|--------|---------------|
| What is the all-cause mortality? | Overall Survival | `overall` |
| What is the direct effect of treatment on disease death rate? | Cause-Specific Cox | `cause` |
| What is the probability of dying from disease accounting for other deaths? | CIF / Fine-Gray | `compete` |
| Do CIF curves differ between groups? | Gray's Test | `compete` + `graystest = TRUE` |
| What is the treatment effect on cumulative incidence? | Subdistribution HR | `compete` + `subdistribution = TRUE` |

## 10. Edge Cases

### Minimum Event Requirement

The function requires at least 5 events of each type for competing risks
analysis. This protects against unstable CIF and Fine-Gray estimates.

```{r too-few-events}
# Create a small subset with few events
set.seed(42)
small_data <- competing_survival_data[sample(1:300, 15), ]

# This may error if disease or competing events < 5
competingsurvival(
  data = small_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  dooc = "Dead_of_Other",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

### Competing Risks Requires Both Death Types

If only `dod` is defined without `dooc`, competing risks analysis will produce
an informative error.

```{r missing-dooc}
competingsurvival(
  data = competing_survival_data,
  overalltime = "Overall_Time",
  outcome = "Outcome",
  explanatory = "Treatment",
  analysistype = "compete",
  dod = "Dead_of_Disease",
  awd = "Alive_w_Disease",
  awod = "Alive_wo_Disease"
)
```

### All Three Analysis Types in Sequence

A common workflow compares results across all three approaches:

```{r all-three}
# Overall
res_all_os <- competingsurvival(
  data = survival_competing,
  overalltime = "elapsedtime",
  outcome = "outcome",
  explanatory = "treatment",
  analysistype = "overall",
  dod = "Dead of Disease",
  dooc = "Dead of Other",
  awd = "Alive w Disease",
  awod = "Alive w/o Disease"
)

# Cause-specific
res_all_css <- competingsurvival(
  data = survival_competing,
  overalltime = "elapsedtime",
  outcome = "outcome",
  explanatory = "treatment",
  analysistype = "cause",
  dod = "Dead of Disease",
  dooc = "Dead of Other",
  awd = "Alive w Disease",
  awod = "Alive w/o Disease"
)

# Competing risks with full features
res_all_cr <- competingsurvival(
  data = survival_competing,
  overalltime = "elapsedtime",
  outcome = "outcome",
  explanatory = "treatment",
  analysistype = "compete",
  dod = "Dead of Disease",
  dooc = "Dead of Other",
  awd = "Alive w Disease",
  awod = "Alive w/o Disease",
  graystest = TRUE,
  subdistribution = TRUE,
  showStackedPlot = TRUE,
  showKMvsCIF = TRUE,
  timepoints = "12,24,36,60",
  confidencelevel = 0.95
)
```

Comparing the hazard ratios across these three analyses reveals how competing
risks affect the estimated treatment effect. The cause-specific HR captures the
direct treatment effect on disease death rate. The subdistribution HR captures
the net treatment effect on cumulative incidence, which is often the quantity
patients and clinicians care about.

## Complete Option Reference

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `data` | Data | (required) | The data as a data frame |
| `overalltime` | Variable (numeric) | (required) | Follow-up time in months |
| `outcome` | Variable (factor) | (required) | Multi-level outcome factor |
| `explanatory` | Variable (factor) | NULL | Grouping variable for comparison |
| `dod` | Level | NULL | Level of `outcome` indicating death from disease |
| `dooc` | Level | NULL | Level of `outcome` indicating death from other causes |
| `awd` | Level | NULL | Level of `outcome` indicating alive with disease |
| `awod` | Level | NULL | Level of `outcome` indicating alive without disease |
| `analysistype` | List | "overall" | Analysis type: "overall", "cause", or "compete" |
| `graystest` | Bool | FALSE | Perform Gray's test comparing CIF curves between groups |
| `subdistribution` | Bool | FALSE | Fit Fine-Gray subdistribution hazard model |
| `timepoints` | String | "12,24,36,60" | Comma-separated time points (months) for CIF estimates |
| `confidencelevel` | Number | 0.95 | Confidence level for intervals (0.50 to 0.99) |
| `showrisksets` | Bool | FALSE | Show number at risk summary below CIF plot |
| `showStackedPlot` | Bool | FALSE | Display stacked CIF + survival probability plot |
| `showKMvsCIF` | Bool | FALSE | Display 1-KM vs CIF comparison plot |
| `cifColors` | List | "default" | Color palette: "default" (Red/Blue), "colorblind", or "grayscale" |

### Output Elements

| Output | Type | Visibility |
|--------|------|------------|
| `todo` | Html | When no variables selected |
| `summary` | Html | Always |
| `survivalTable` | Table | Always |
| `cuminc` | Table | When `analysistype = "compete"` |
| `comprisksPlot` | Image | When `analysistype = "compete"` |
| `stackedPlot` | Image | When `showStackedPlot = TRUE` and `analysistype = "compete"` |
| `kmvscifPlot` | Image | When `showKMvsCIF = TRUE` and `analysistype = "compete"` |
| `interpretation` | Html | Always |
| `assumptions` | Html | Always |
| `fineGrayTable` | Table | When `subdistribution = TRUE` |

## References

- Fine JP, Gray RJ (1999). A proportional hazards model for the subdistribution
  of a competing risk. *Journal of the American Statistical Association*,
  94(446):496-509.
- Gray RJ (1988). A class of K-sample tests for comparing the cumulative
  incidence of a competing risk. *Annals of Statistics*, 16(3):1141-1154.
- Putter H, Fiocco M, Geskus RB (2007). Tutorial in biostatistics: competing
  risks and multi-state models. *Statistics in Medicine*, 26(11):2389-2430.
- Austin PC, Lee DS, Fine JP (2016). Introduction to the analysis of survival
  data in the presence of competing risks. *Circulation*, 133(6):601-609.
- Dignam JJ, Zhang Q, Kocherginsky M (2012). The use and interpretation of
  competing risks regression models. *Clinical Cancer Research*,
  18(8):2301-2308.
