---
title: "Interpreting the interaction output in Multivariable Survival Analysis"
subtitle: "Prognostic vs. predictive, the two interaction tables, and how the within-subgroup hazard ratios are estimated"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting the interaction output in Multivariable Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeepMd{true}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

This guide is a practical companion to *Interaction terms in Multivariable Survival
Analysis*. That vignette shows **how to build** an interaction; this one shows **how to read
the results** — what the two interaction tables mean, how the within-subgroup hazard ratios
are computed, and why they differ slightly from running separate Cox models in each subgroup.

Throughout we use a concrete breast-cancer example: a set of candidate markers (AR and
others) tested for whether their effect on overall survival **depends on ER status**.

## Prognostic vs. predictive

Two different clinical claims hide behind the word "biomarker":

- A **prognostic** marker predicts outcome *overall* — higher AR is associated with worse
  (or better) survival, regardless of anything else.
- A **predictive** marker's effect *depends on another factor* — AR matters in ER-positive
  tumours but not in ER-negative ones (or vice versa).

A single Cox model fitted per marker cannot separate these. The **interaction test** can: it
asks whether the marker's hazard ratio is the *same* in both ER groups, or *different*. A
different effect across groups is **effect modification**, and it is the statistical
signature of a predictive biomarker.

## The model

When you cross `ER` with a marker in the **Interaction Terms** builder, the module fits a
single Cox model with the crossed term:

```{r}
survival::coxph(survival::Surv(OS_days, OS_event) ~ ER * Marker, data = df)
# ER * Marker expands to:  ER + Marker + ER:Marker
```

The `ER:Marker` coefficient **is** the interaction. Exponentiated, it is a *ratio of hazard
ratios*: how much the marker's HR in ER-positive patients differs from its HR in
ER-negative patients. If it equals 1, the marker behaves identically in both groups (not
predictive). If it departs from 1, the marker's effect is modified by ER.

## Table 1 — Interaction (Effect-Modification) Test

This table has one row per crossed term and reports the formal test of *whether* the effect
differs.

| Interaction | HR | 95% CI | p |
|---|---:|---:|---:|
| `AR : ER` | 0.42 | 0.22 – 0.80 | 0.008 |

- **HR** is the *ratio* of AR's hazard ratio between the ER groups. `0.42` means AR is far
  more protective in one ER stratum than the other.
- **HR = 1** would mean AR does the same thing regardless of ER — no modification, not
  predictive.
- **p** is the *p-for-interaction*. Here `p = 0.008`, so AR's effect on survival genuinely
  differs by ER status.

Read this table as the answer to the predictive-biomarker question: *does the marker's
effect depend on ER?* It tells you **that** the effect differs; the next table tells you
**how**.

## Table 2 — Within-Subgroup Hazard Ratios

This table gives the marker's actual effect size *inside* each ER subgroup — the clinically
usable numbers.

| Interaction | Subgroup | Focal | HR | 95% CI | p |
|---|---|---|---:|---:|---:|
| `AR * ER` | ER0 (ER-negative) | AR | 1.02 | 0.71 – 1.47 | 0.90 |
| `AR * ER` | ER1 (ER-positive) | AR | 0.43 | 0.26 – 0.72 | <0.001 |

AR is **inert in ER-negative** tumours (HR 1.02, p = 0.90) but **strongly protective in
ER-positive** tumours (HR 0.43, p < 0.001). AR is therefore a **predictive marker within
ER-positive disease** — exactly what the significant interaction in Table 1 promised.
Reporting only a pooled AR HR would have averaged these two very different realities into a
misleading middle.

## How the within-subgroup HRs are computed

The subgroup HRs are read straight out of the **one interaction model** — *not* from separate
Cox models fitted in each ER group. The module uses **relevel-and-refit**:

1. Fit the full model `Surv ~ AR * ER` (plus any covariates you added) on all patients.
2. Relevel `ER` so a chosen level (ER0, then ER1) is the reference. This changes only the
   parameterisation, not the fit.
3. In the releveled model, the `AR` main-effect coefficient now *is* AR's effect **within
   that ER level**. Its HR = `exp(coef)`, with the CI and p from the model's own variance.

Because releveling does not alter the fit, this is mathematically identical to reading
`β(AR)` (within ER0) and `β(AR) + β(AR:ER)` (within ER1) off the single model — releveling is
just the clean way to obtain each one *together with its correct standard error*. These are
the model's **implied conditional effects**.

## Why they differ slightly from separate subgroup Cox models

If you fit AR in each ER subgroup *independently*, the point estimates (HRs) will be *almost,
but not exactly*, the same, and the confidence intervals and p-values will differ a little.
This is expected, and it follows from what each method assumes:

| | Interaction model (what the module reports) | Separate Cox per ER group |
|---|---|---|
| Baseline hazard | **common** across ER (ER is a covariate) | each subgroup its **own** |
| Other covariates | effects assumed **shared**, estimated once | **re-estimated** within each group |
| Standard errors | **borrow strength** across the whole sample | use **only** that subgroup's data |
| Consequence | **narrower CIs, more power** | wider CIs, less power |

- **HRs nearly identical:** with only `ER * Marker`, they differ only because the interaction
  model uses a common baseline hazard across ER while a subgroup-only fit uses an ER-specific
  baseline. Add other covariates and the small gap also reflects the interaction model
  assuming common covariate effects.
- **CIs and p differ more:** the interaction model pools variance across both ER groups, so
  its intervals are tighter — the small difference you may notice against a manual
  subgroup check.

Both are valid, but they answer subtly different questions. The interaction-model
(conditional) subgroup effect is the **preferred, more efficient estimand for
predictive-biomarker subgroup analysis**, and it is what the module reports (noted in the
table caption). Separate subgroup fits are a fine sanity check: expect matching point
estimates and slightly wider intervals.

## Getting the subgroup table to fill — focal and moderator

The builder reads a crossed term as **focal : moderator**. The **moderator** is the variable
that defines the subgroups, and it must be **categorical**. The **focal** is the effect
measured within those subgroups.

The module **auto-detects** the moderator: for a term `A:B` it uses whichever of `A` / `B` is
categorical as the moderator (preferring the second variable when both are categorical). So:

- `AR : ER` (ER categorical) → subgroups by ER, AR effect shown within each — the marker-by-ER
  question. Fills the table.
- `Marker : ER` with a continuous marker → still fills: ER is the moderator, and the marker's
  per-unit HR is shown within each ER subgroup.
- If **both** variables in the term are continuous, there are no discrete subgroups to form,
  so the Within-Subgroup table is intentionally left empty (an on-screen note explains this).
  The interaction-test table (Table 1) fills either way.

To be sure a categorical variable is used as the moderator, simply include it in the crossed
term; the module will assign the roles so the subgroups are formed by the categorical
variable.

## A reproducible check in R

To confirm the interaction HR for one marker against the module (matching the module's
complete-case handling for that marker):

```{r}
library(survival); library(broom)

fit <- coxph(Surv(OS_days, OS_event) ~ ER * AR, data = na.omit(df[, c("OS_days","OS_event","ER","AR")]))

# interaction HR / CI / p (matches Table 1)
broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE)[grepl(":", broom::tidy(fit)$term), ]

# within-subgroup AR HR, ER-positive as reference (matches Table 2, ER1 row)
df$ER <- relevel(factor(df$ER), ref = "ER1")
coxph(Surv(OS_days, OS_event) ~ ER * AR, data = df)   # read the AR coefficient
```

## Cautions before concluding "predictive"

- **Power.** Interaction tests need substantially more events than main-effect tests. A
  non-significant interaction in a modest sample is *weak evidence of no modification*, not
  proof of its absence.
- **Multiplicity.** Scanning many markers inflates false positives. Adjust the
  p-for-interaction (for example, with an FDR correction) before calling any single marker
  predictive.
- **Complete cases.** If you compare the module to a per-marker R script, note that the
  module uses complete cases across *all* selected variables, whereas a per-marker script
  drops rows missing *that* marker only. Match the row set when comparing a single marker, or
  N — and therefore the CI and p — can differ for that reason alone.

## Summary

Table 1 tells you **whether** a marker's effect on survival depends on ER (the
predictive-biomarker test). Table 2 tells you **how big** the effect is inside each ER group,
estimated from the same interaction model as its implied conditional effects. Include a
categorical variable in the crossed term and the module forms the subgroups by it
automatically.
