---
title: "jSurvival 14: Comparing Survival Outcomes Between Groups"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{jSurvival 14: Comparing Survival Outcomes Between Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

Comparing survival outcomes between different groups is a fundamental aspect of clinical research and oncology studies. The **comparingsurvival** function in jSurvival provides comprehensive tools for performing group-wise survival comparisons using Kaplan-Meier estimation and log-rank tests, with enhanced visualization capabilities.

**Learning Objectives:**

- Master group-wise survival comparisons using Kaplan-Meier methods
- Interpret log-rank test results and statistical significance
- Create publication-ready survival curves with group stratification
- Understand hazard ratios and median survival differences
- Apply proper statistical methodology for survival group comparisons
- Identify when survival differences are clinically meaningful

## Module Overview

The **comparingsurvival** function provides:

### 1. **Statistical Methods**
- **Kaplan-Meier Estimation**: Non-parametric survival curves for each group
- **Log-rank Test**: Statistical comparison of survival distributions
- **Hazard Ratio Estimation**: Measure of relative risk between groups
- **Confidence Intervals**: Uncertainty quantification for survival estimates

### 2. **Visualization Features**
- **Survival Curves**: Group-stratified Kaplan-Meier plots
- **Risk Tables**: Number at risk at specified time points
- **Confidence Bands**: Visual representation of estimation uncertainty
- **Statistical Annotations**: P-values and test statistics on plots

### 3. **Clinical Applications**
- **Treatment Comparisons**: Compare therapeutic interventions
- **Biomarker Stratification**: Evaluate prognostic markers
- **Risk Group Analysis**: Compare different risk categories
- **Subgroup Analysis**: Examine specific patient populations

```{r load-packages}
library(ClinicoPath)
library(survival)
library(dplyr)
```

# Getting Started

## Sample Data Overview

We'll use comprehensive clinical datasets to demonstrate survival comparisons across different scenarios.

```{r sample-data}
# Load clinical survival data
data(ihc_molecular_comprehensive)

# Dataset overview
cat("Dataset dimensions:", nrow(ihc_molecular_comprehensive), "×", ncol(ihc_molecular_comprehensive), "\n")

# Key survival variables
survival_vars <- c("Overall_Survival_Months", "Death_Event", "PFS_Months", "Progression_Event")
cat("Survival variables:", paste(survival_vars, collapse = ", "), "\n")

# Grouping variables for comparison
grouping_vars <- c("EGFR_Mutation", "Grade", "MSI_Status", "HER2_IHC")
cat("Grouping variables:", paste(grouping_vars, collapse = ", "), "\n")

# Basic survival statistics by key groups
survival_by_egfr <- ihc_molecular_comprehensive %>%
  group_by(EGFR_Mutation) %>%
  summarise(
    n = n(),
    events = sum(Death_Event),
    event_rate = round(mean(Death_Event) * 100, 1),
    median_survival = median(Overall_Survival_Months),
    .groups = 'drop'
  )

cat("\nSurvival by EGFR Mutation Status:\n")
print(survival_by_egfr)
```

# Core Functionality

## 1. Basic Survival Comparison

Fundamental two-group survival comparison.

```{r basic-comparison}
# Basic Survival Comparison Example
# In jamovi: jsurvival > Comparing Survival

cat("Basic Survival Comparison: EGFR Mutation Status\n")
cat("===============================================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  library(survival)
  
  # Create survival object
  surv_obj <- Surv(ihc_molecular_comprehensive$Overall_Survival_Months,
                   ihc_molecular_comprehensive$Death_Event)
  
  # Kaplan-Meier analysis by EGFR status
  km_egfr <- survfit(surv_obj ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  
  cat("Survival Summary by EGFR Status:\n")
  summary_table <- summary(km_egfr, times = c(12, 24, 36, 48, 60))$table
  print(summary_table)
  
  # Extract median survival times
  median_survivals <- summary(km_egfr)$table[, "median"]
  names(median_survivals) <- c("EGFR Negative", "EGFR Positive")
  
  cat("\nMedian Survival Times:\n")
  for(i in 1:length(median_survivals)) {
    cat(names(median_survivals)[i], ":", median_survivals[i], "months\n")
  }
  
  # Calculate survival difference
  survival_diff <- abs(diff(median_survivals))
  cat("Difference in median survival:", round(survival_diff, 1), "months\n")
  
  # Log-rank test
  logrank_test <- survdiff(surv_obj ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  chi_square <- logrank_test$chisq
  p_value <- 1 - pchisq(chi_square, 1)
  
  cat("\nLog-rank Test Results:\n")
  cat("Chi-square statistic:", round(chi_square, 3), "\n")
  cat("P-value:", round(p_value, 4), "\n")
  cat("Degrees of freedom: 1\n")
  
  # Statistical interpretation
  if(p_value < 0.001) {
    cat("Interpretation: Highly significant difference (p < 0.001)\n")
  } else if(p_value < 0.01) {
    cat("Interpretation: Highly significant difference (p < 0.01)\n")
  } else if(p_value < 0.05) {
    cat("Interpretation: Significant difference (p < 0.05)\n")
  } else {
    cat("Interpretation: No significant difference (p ≥ 0.05)\n")
  }
  
  # Effect size (hazard ratio approximation)
  cox_simple <- coxph(surv_obj ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  hr <- exp(cox_simple$coefficients)
  hr_ci <- exp(confint(cox_simple))
  
  cat("\nHazard Ratio Analysis:\n")
  cat("Hazard Ratio:", round(hr, 2), "\n")
  cat("95% Confidence Interval:", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), "\n")
  
  if(hr > 1) {
    cat("Interpretation: EGFR positive patients have", round(hr, 2), "times higher hazard of death\n")
  } else {
    cat("Interpretation: EGFR positive patients have", round(1/hr, 2), "times lower hazard of death\n")
  }
}
```

## 2. Multi-Group Comparison

Compare survival across multiple groups simultaneously.

```{r multi-group-comparison}
# Multi-Group Survival Comparison
# Comparing across tumor grades

cat("\nMulti-Group Survival Comparison: Tumor Grade\n")
cat("===========================================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  # Survival by tumor grade (3 groups)
  km_grade <- survfit(surv_obj ~ Grade, data = ihc_molecular_comprehensive)
  
  cat("Survival Summary by Tumor Grade:\n")
  grade_summary <- summary(km_grade, times = c(12, 24, 36, 48, 60))$table
  print(grade_summary)
  
  # Median survival by grade
  grade_medians <- summary(km_grade)$table[, "median"]
  grade_names <- paste("Grade", 1:length(grade_medians))
  names(grade_medians) <- grade_names
  
  cat("\nMedian Survival by Grade:\n")
  for(i in 1:length(grade_medians)) {
    cat(names(grade_medians)[i], ":", grade_medians[i], "months\n")
  }
  
  # Overall log-rank test
  logrank_grade <- survdiff(surv_obj ~ Grade, data = ihc_molecular_comprehensive)
  chi_square_grade <- logrank_grade$chisq
  df_grade <- length(unique(ihc_molecular_comprehensive$Grade)) - 1
  p_value_grade <- 1 - pchisq(chi_square_grade, df_grade)
  
  cat("\nOverall Log-rank Test:\n")
  cat("Chi-square statistic:", round(chi_square_grade, 3), "\n")
  cat("Degrees of freedom:", df_grade, "\n")
  cat("P-value:", round(p_value_grade, 4), "\n")
  
  # Pairwise comparisons (if overall test is significant)
  if(p_value_grade < 0.05) {
    cat("\nPairwise Comparisons:\n")
    grades <- unique(ihc_molecular_comprehensive$Grade)
    grades <- grades[!is.na(grades)]
    
    for(i in 1:(length(grades)-1)) {
      for(j in (i+1):length(grades)) {
        grade_subset <- ihc_molecular_comprehensive[ihc_molecular_comprehensive$Grade %in% c(grades[i], grades[j]), ]
        if(nrow(grade_subset) > 0) {
          surv_subset <- Surv(grade_subset$Overall_Survival_Months, grade_subset$Death_Event)
          logrank_pair <- survdiff(surv_subset ~ Grade, data = grade_subset)
          p_pair <- 1 - pchisq(logrank_pair$chisq, 1)
          
          cat(paste0("Grade ", grades[i], " vs Grade ", grades[j], ": p = ", round(p_pair, 4), "\n"))
        }
      }
    }
  }
  
  # Clinical significance assessment
  max_diff <- max(grade_medians, na.rm = TRUE) - min(grade_medians, na.rm = TRUE)
  cat("\nClinical Significance:\n")
  cat("Range of median survivals:", round(max_diff, 1), "months\n")
  
  if(max_diff >= 12) {
    cat("Interpretation: Clinically meaningful difference (≥12 months)\n")
  } else if(max_diff >= 6) {
    cat("Interpretation: Moderate clinical difference (6-12 months)\n")
  } else {
    cat("Interpretation: Limited clinical difference (<6 months)\n")
  }
}
```

## 3. Biomarker Stratification Analysis

Evaluate survival differences by molecular biomarkers.

```{r biomarker-stratification}
# Biomarker-Based Survival Stratification
cat("\nBiomarker Stratification Analysis\n")
cat("=================================\n\n")

# Multiple biomarker analysis
biomarkers <- list(
  "MSI_Status" = "Microsatellite Instability",
  "HER2_IHC" = "HER2 Expression",
  "EGFR_Mutation" = "EGFR Mutation Status"
)

biomarker_results <- list()

for(biomarker in names(biomarkers)) {
  cat(paste0("Analysis: ", biomarkers[[biomarker]], " (", biomarker, ")\n"))
  cat(paste0(rep("-", nchar(biomarkers[[biomarker]]) + nchar(biomarker) + 12), collapse = ""), "\n")
  
  # Filter out missing values
  complete_data <- ihc_molecular_comprehensive[!is.na(ihc_molecular_comprehensive[[biomarker]]), ]
  
  if(nrow(complete_data) > 0) {
    # Survival analysis
    surv_bio <- Surv(complete_data$Overall_Survival_Months, complete_data$Death_Event)
    formula_str <- paste("surv_bio ~", biomarker)
    km_bio <- survfit(as.formula(formula_str), data = complete_data)
    
    # Summary statistics
    bio_summary <- summary(km_bio)$table
    print(bio_summary)
    
    # Log-rank test
    logrank_bio <- survdiff(as.formula(formula_str), data = complete_data)
    p_bio <- 1 - pchisq(logrank_bio$chisq, nrow(bio_summary) - 1)
    
    cat("Log-rank test p-value:", round(p_bio, 4), "\n")
    
    # Cox regression for hazard ratio
    cox_bio <- coxph(as.formula(formula_str), data = complete_data)
    hr_bio <- exp(cox_bio$coefficients)
    
    if(length(hr_bio) == 1) {  # Binary biomarker
      hr_ci_bio <- exp(confint(cox_bio))
      cat("Hazard Ratio:", round(hr_bio, 2), "(95% CI:", round(hr_ci_bio[1], 2), "-", round(hr_ci_bio[2], 2), ")\n")
    }
    
    # Store results
    biomarker_results[[biomarker]] <- list(
      p_value = p_bio,
      significant = p_bio < 0.05,
      summary = bio_summary
    )
    
    # Interpretation
    if(p_bio < 0.001) {
      cat("Result: Highly significant prognostic biomarker (p < 0.001)\n")
    } else if(p_bio < 0.01) {
      cat("Result: Highly significant prognostic biomarker (p < 0.01)\n")
    } else if(p_bio < 0.05) {
      cat("Result: Significant prognostic biomarker (p < 0.05)\n")
    } else {
      cat("Result: Not a significant prognostic biomarker (p ≥ 0.05)\n")
    }
  } else {
    cat("Insufficient data for analysis\n")
    biomarker_results[[biomarker]] <- list(p_value = NA, significant = FALSE)
  }
  
  cat("\n")
}

# Summary of biomarker findings
cat("Biomarker Analysis Summary:\n")
cat("===========================\n")
significant_biomarkers <- names(biomarker_results)[sapply(biomarker_results, function(x) x$significant)]
if(length(significant_biomarkers) > 0) {
  cat("Significant prognostic biomarkers:\n")
  for(bio in significant_biomarkers) {
    p_val <- biomarker_results[[bio]]$p_value
    cat(paste0("- ", biomarkers[[bio]], " (p = ", round(p_val, 4), ")\n"))
  }
} else {
  cat("No significant prognostic biomarkers identified\n")
}
```

# Advanced Applications

## 4. Time-to-Progression Analysis

Compare progression-free survival between groups.

```{r progression-analysis}
# Time-to-Progression Comparison
cat("\nProgression-Free Survival Analysis\n")
cat("==================================\n\n")

if(requireNamespace("survival", quietly = TRUE)) {
  # Create PFS survival object
  pfs_surv <- Surv(ihc_molecular_comprehensive$PFS_Months,
                   ihc_molecular_comprehensive$Progression_Event)
  
  # PFS by treatment-relevant biomarker (e.g., EGFR)
  km_pfs <- survfit(pfs_surv ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  
  cat("Progression-Free Survival by EGFR Status:\n")
  pfs_summary <- summary(km_pfs, times = c(6, 12, 18, 24, 30))$table
  print(pfs_summary)
  
  # Median PFS
  pfs_medians <- summary(km_pfs)$table[, "median"]
  cat("\nMedian PFS Times:\n")
  cat("EGFR Negative:", pfs_medians[1], "months\n")
  cat("EGFR Positive:", pfs_medians[2], "months\n")
  
  # Statistical test
  logrank_pfs <- survdiff(pfs_surv ~ EGFR_Mutation, data = ihc_molecular_comprehensive)
  p_pfs <- 1 - pchisq(logrank_pfs$chisq, 1)
  
  cat("\nLog-rank Test for PFS:\n")
  cat("P-value:", round(p_pfs, 4), "\n")
  
  # Compare OS vs PFS findings
  os_p <- biomarker_results[["EGFR_Mutation"]]$p_value
  cat("\nOS vs PFS Comparison:\n")
  cat("Overall Survival p-value:", round(os_p, 4), "\n")
  cat("Progression-Free Survival p-value:", round(p_pfs, 4), "\n")
  
  if(os_p < 0.05 && p_pfs < 0.05) {
    cat("Interpretation: EGFR status affects both OS and PFS\n")
  } else if(p_pfs < 0.05) {
    cat("Interpretation: EGFR status primarily affects PFS\n")
  } else if(os_p < 0.05) {
    cat("Interpretation: EGFR status primarily affects OS\n")
  } else {
    cat("Interpretation: EGFR status does not significantly affect survival outcomes\n")
  }
}
```

## 5. Subgroup Analysis

Examine survival differences within specific patient subgroups.

```{r subgroup-analysis}
# Subgroup Survival Analysis
cat("\nSubgroup Analysis: High-Grade Tumors\n")
cat("===================================\n\n")

# Focus on high-grade tumors (Grade 3)
high_grade <- ihc_molecular_comprehensive[ihc_molecular_comprehensive$Grade == 3, ]

if(nrow(high_grade) > 20) {  # Ensure adequate sample size
  cat("High-grade tumor cohort size:", nrow(high_grade), "patients\n")
  
  # Survival in high-grade tumors by EGFR status
  surv_hg <- Surv(high_grade$Overall_Survival_Months, high_grade$Death_Event)
  km_hg <- survfit(surv_hg ~ EGFR_Mutation, data = high_grade)
  
  cat("\nSurvival in High-Grade Tumors by EGFR Status:\n")
  hg_summary <- summary(km_hg, times = c(12, 24, 36))$table
  print(hg_summary)
  
  # Statistical test in subgroup
  logrank_hg <- survdiff(surv_hg ~ EGFR_Mutation, data = high_grade)
  p_hg <- 1 - pchisq(logrank_hg$chisq, 1)
  
  cat("Log-rank test p-value in high-grade subgroup:", round(p_hg, 4), "\n")
  
  # Compare to overall population
  overall_p <- biomarker_results[["EGFR_Mutation"]]$p_value
  cat("\nSubgroup vs Overall Population:\n")
  cat("Overall population p-value:", round(overall_p, 4), "\n")
  cat("High-grade subgroup p-value:", round(p_hg, 4), "\n")
  
  if(p_hg < 0.05 && overall_p >= 0.05) {
    cat("Interpretation: EGFR effect is significant only in high-grade tumors\n")
  } else if(p_hg >= 0.05 && overall_p < 0.05) {
    cat("Interpretation: EGFR effect is diluted in high-grade subgroup\n")
  } else if(p_hg < overall_p) {
    cat("Interpretation: EGFR effect is stronger in high-grade tumors\n")
  } else {
    cat("Interpretation: EGFR effect is consistent across grades\n")
  }
} else {
  cat("Insufficient sample size for subgroup analysis\n")
}
```

## 6. Risk-Based Stratification

Create risk groups and compare survival outcomes.

```{r risk-stratification}
# Risk-Based Survival Stratification
cat("\nRisk-Based Survival Stratification\n")
cat("==================================\n\n")

# Create composite risk score
risk_data <- ihc_molecular_comprehensive %>%
  mutate(
    # Risk factors (higher = worse prognosis)
    age_risk = ifelse(Age > 65, 1, 0),
    grade_risk = ifelse(Grade == 3, 1, 0),
    egfr_risk = ifelse(EGFR_Mutation == "Positive", 1, 0),
    
    # Protective factors (higher = better prognosis)
    msi_protective = ifelse(MSI_Status == "MSI-High", 1, 0),
    
    # Composite risk score (0-3, higher = worse)
    risk_score = age_risk + grade_risk + egfr_risk - msi_protective
  ) %>%
  mutate(
    # Risk groups
    risk_group = case_when(
      risk_score <= 0 ~ "Low Risk",
      risk_score == 1 ~ "Intermediate Risk", 
      risk_score >= 2 ~ "High Risk"
    ),
    risk_group = factor(risk_group, levels = c("Low Risk", "Intermediate Risk", "High Risk"))
  )

# Risk group distribution
cat("Risk Group Distribution:\n")
risk_dist <- table(risk_data$risk_group)
print(risk_dist)
cat("Percentages:", round(prop.table(risk_dist) * 100, 1), "%\n")

# Survival by risk group
if(requireNamespace("survival", quietly = TRUE)) {
  surv_risk <- Surv(risk_data$Overall_Survival_Months, risk_data$Death_Event)
  km_risk <- survfit(surv_risk ~ risk_group, data = risk_data)
  
  cat("\nSurvival by Risk Group:\n")
  risk_summary <- summary(km_risk, times = c(12, 24, 36, 48))$table
  print(risk_summary)
  
  # Risk group medians
  risk_medians <- summary(km_risk)$table[, "median"]
  cat("\nMedian Survival by Risk Group:\n")
  for(i in 1:length(risk_medians)) {
    cat(levels(risk_data$risk_group)[i], ":", risk_medians[i], "months\n")
  }
  
  # Statistical test for risk stratification
  logrank_risk <- survdiff(surv_risk ~ risk_group, data = risk_data)
  chi_sq_risk <- logrank_risk$chisq
  df_risk <- length(levels(risk_data$risk_group)) - 1
  p_risk <- 1 - pchisq(chi_sq_risk, df_risk)
  
  cat("\nRisk Stratification Test:\n")
  cat("Chi-square:", round(chi_sq_risk, 3), "\n")
  cat("P-value:", round(p_risk, 4), "\n")
  
  # Trend test
  risk_numeric <- as.numeric(risk_data$risk_group)
  cox_trend <- coxph(surv_risk ~ risk_numeric, data = risk_data)
  trend_p <- summary(cox_trend)$coefficients[, "Pr(>|z|)"]
  hr_trend <- exp(cox_trend$coefficients)
  
  cat("\nTrend Analysis:\n")
  cat("Hazard ratio per risk level:", round(hr_trend, 2), "\n")
  cat("Trend test p-value:", round(trend_p, 4), "\n")
  
  # Model performance
  c_index <- summary(cox_trend)$concordance[1]
  cat("Concordance index:", round(c_index, 3), "\n")
  
  if(c_index >= 0.7) {
    cat("Risk model performance: Excellent (C-index ≥ 0.7)\n")
  } else if(c_index >= 0.6) {
    cat("Risk model performance: Good (C-index 0.6-0.7)\n")
  } else {
    cat("Risk model performance: Poor (C-index < 0.6)\n")
  }
}
```

# Statistical Considerations

## Power and Sample Size

Important considerations for survival comparisons.

```{r power-considerations}
# Sample Size and Power Considerations
cat("Sample Size and Power Analysis\n")
cat("==============================\n\n")

# Current study characteristics
total_n <- nrow(ihc_molecular_comprehensive)
events <- sum(ihc_molecular_comprehensive$Death_Event)
event_rate <- mean(ihc_molecular_comprehensive$Death_Event)

cat("Current Study Characteristics:\n")
cat("Total sample size:", total_n, "\n")
cat("Total events:", events, "\n")
cat("Overall event rate:", round(event_rate * 100, 1), "%\n")

# Group sizes for major comparisons
egfr_groups <- table(ihc_molecular_comprehensive$EGFR_Mutation)
cat("\nEGFR Group Sizes:\n")
print(egfr_groups)

# Events per group
egfr_events <- ihc_molecular_comprehensive %>%
  group_by(EGFR_Mutation) %>%
  summarise(
    n = n(),
    events = sum(Death_Event),
    event_rate = round(mean(Death_Event) * 100, 1),
    .groups = 'drop'
  )

cat("\nEvents by EGFR Status:\n")
print(egfr_events)

# Power estimation guidelines
min_events_per_group <- 10
adequate_events <- all(egfr_events$events >= min_events_per_group)

cat("\nPower Assessment:\n")
cat("Minimum events per group recommended:", min_events_per_group, "\n")
cat("Adequate events per group:", adequate_events, "\n")

if(adequate_events) {
  cat("Conclusion: Adequate power for primary comparisons\n")
} else {
  cat("Conclusion: Limited power - consider pooling groups or longer follow-up\n")
}

# Effect size detectability
median_diff_detectable <- sqrt(4 / min(egfr_events$events)) * 12  # Rough approximation
cat("Approximately detectable median survival difference:", round(median_diff_detectable, 1), "months\n")
```

## Multiple Comparisons

Addressing multiple testing in survival comparisons.

```{r multiple-comparisons}
# Multiple Comparisons Adjustment
cat("\nMultiple Comparisons Consideration\n")
cat("=================================\n\n")

# Collect all p-values from biomarker analyses
all_p_values <- sapply(biomarker_results, function(x) x$p_value)
all_p_values <- all_p_values[!is.na(all_p_values)]

cat("Raw P-values from Biomarker Analyses:\n")
for(i in 1:length(all_p_values)) {
  cat(names(all_p_values)[i], ":", round(all_p_values[i], 4), "\n")
}

# Bonferroni correction
bonferroni_alpha <- 0.05 / length(all_p_values)
bonferroni_significant <- all_p_values < bonferroni_alpha

cat("\nBonferroni Correction:\n")
cat("Adjusted alpha level:", round(bonferroni_alpha, 4), "\n")
cat("Significant after Bonferroni correction:\n")
if(any(bonferroni_significant)) {
  for(i in which(bonferroni_significant)) {
    cat(names(all_p_values)[i], ": p =", round(all_p_values[i], 4), "< α =", round(bonferroni_alpha, 4), "\n")
  }
} else {
  cat("None remain significant after correction\n")
}

# False Discovery Rate (FDR) control
if(requireNamespace("stats", quietly = TRUE)) {
  fdr_adjusted <- p.adjust(all_p_values, method = "fdr")
  fdr_significant <- fdr_adjusted < 0.05
  
  cat("\nFalse Discovery Rate (FDR) Adjustment:\n")
  for(i in 1:length(all_p_values)) {
    cat(names(all_p_values)[i], ": raw p =", round(all_p_values[i], 4), 
        ", FDR p =", round(fdr_adjusted[i], 4), "\n")
  }
  
  cat("\nSignificant after FDR correction:\n")
  if(any(fdr_significant)) {
    for(i in which(fdr_significant)) {
      cat(names(all_p_values)[i], ": FDR p =", round(fdr_adjusted[i], 4), "< 0.05\n")
    }
  } else {
    cat("None remain significant after FDR correction\n")
  }
}

# Recommendation
cat("\nRecommendation:\n")
if(length(all_p_values) <= 3) {
  cat("With", length(all_p_values), "comparisons, multiple testing adjustment may not be necessary\n")
} else if(any(fdr_significant)) {
  cat("Use FDR-adjusted p-values for interpretation\n")
} else if(any(bonferroni_significant)) {
  cat("Conservative Bonferroni correction identifies significant findings\n")
} else {
  cat("Consider exploratory interpretation due to multiple testing\n")
}
```

# Best Practices and Interpretation

## Clinical Significance

Guidelines for interpreting survival differences.

```{r clinical-significance}
# Clinical Significance Assessment
cat("Clinical Significance Guidelines\n")
cat("===============================\n\n")

# Establish clinical significance thresholds
clinically_meaningful_months <- 6  # Minimum meaningful difference
highly_meaningful_months <- 12     # Highly meaningful difference

cat("Clinical Significance Thresholds:\n")
cat("Meaningful difference: ≥", clinically_meaningful_months, "months\n")
cat("Highly meaningful difference: ≥", highly_meaningful_months, "months\n\n")

# Assess each comparison
cat("Clinical Significance Assessment:\n")
cat("=================================\n")

# EGFR comparison example
if("EGFR_Mutation" %in% names(biomarker_results) && biomarker_results[["EGFR_Mutation"]]$significant) {
  egfr_medians <- summary(survfit(Surv(Overall_Survival_Months, Death_Event) ~ EGFR_Mutation, 
                                 data = ihc_molecular_comprehensive))$table[, "median"]
  egfr_diff <- abs(diff(egfr_medians))
  
  cat("EGFR Mutation Status:\n")
  cat("Median survival difference:", round(egfr_diff, 1), "months\n")
  cat("Statistical significance: Yes (p < 0.05)\n")
  
  if(egfr_diff >= highly_meaningful_months) {
    cat("Clinical significance: Highly meaningful (≥12 months)\n")
  } else if(egfr_diff >= clinically_meaningful_months) {
    cat("Clinical significance: Meaningful (6-12 months)\n")
  } else {
    cat("Clinical significance: Limited (<6 months)\n")
  }
  
  # Additional considerations
  cat("Additional considerations:\n")
  if(egfr_diff >= clinically_meaningful_months) {
    cat("- Difference likely to be noticeable to patients and families\n")
    cat("- May influence treatment decisions\n")
    cat("- Consider for biomarker-guided therapy\n")
  } else {
    cat("- Difference may be too small to be clinically relevant\n")
    cat("- Statistical significance may not translate to clinical utility\n")
    cat("- Consider as prognostic information only\n")
  }
}

# General interpretation framework
cat("\nGeneral Interpretation Framework:\n")
cat("- Statistical significance indicates reliable difference\n")
cat("- Clinical significance requires meaningful survival benefit\n")
cat("- Consider both statistical and clinical significance together\n")
cat("- Account for confidence intervals and uncertainty\n")
cat("- Evaluate biological plausibility of findings\n")
```

## Reporting Guidelines

Comprehensive reporting of survival comparisons.

```{r reporting-guidelines}
# Reporting Guidelines for Survival Comparisons
cat("Reporting Guidelines\n")
cat("===================\n\n")

cat("Essential Elements to Report:\n")
cat("1. Study Population:\n")
cat("   - Total sample size:", nrow(ihc_molecular_comprehensive), "\n")
cat("   - Follow-up characteristics\n")
cat("   - Censoring pattern\n\n")

cat("2. Descriptive Statistics:\n")
cat("   - Group sizes and event rates\n")
cat("   - Median follow-up time\n")
cat("   - Median survival times with confidence intervals\n\n")

cat("3. Statistical Tests:\n")
cat("   - Log-rank test statistics and p-values\n")
cat("   - Hazard ratios with 95% confidence intervals\n")
cat("   - Multiple comparison adjustments (if applicable)\n\n")

cat("4. Visualization:\n")
cat("   - Kaplan-Meier survival curves\n")
cat("   - Number at risk tables\n")
cat("   - Confidence bands (optional)\n\n")

cat("5. Clinical Interpretation:\n")
cat("   - Clinical significance assessment\n")
cat("   - Biological plausibility\n")
cat("   - Limitations and potential biases\n\n")

# Example results summary
cat("Example Results Summary:\n")
cat("========================\n")
cat("Among", nrow(ihc_molecular_comprehensive), "patients with comprehensive molecular profiling,")
cat("survival comparisons were performed for key biomarkers.\n")

significant_findings <- sum(sapply(biomarker_results, function(x) x$significant))
cat("Of", length(biomarker_results), "biomarkers evaluated,", significant_findings, 
    "showed statistically significant survival differences.\n")

if(significant_findings > 0) {
  cat("These findings suggest potential prognostic value for molecular stratification")
  cat("in this patient population.\n")
} else {
  cat("No significant prognostic biomarkers were identified in this analysis.\n")
}
```

# Integration with jamovi

## Using comparingsurvival in jamovi

Step-by-step guide for jamovi users.

```{r jamovi-integration}
# jamovi Integration Guide
cat("Using comparingsurvival in jamovi\n")
cat("=================================\n\n")

cat("Step-by-Step Instructions:\n")
cat("1. Data Preparation:\n")
cat("   - Ensure survival time variable (continuous)\n")
cat("   - Event indicator variable (0/1 or logical)\n")
cat("   - Grouping variable (categorical)\n\n")

cat("2. Access the Function:\n")
cat("   - Navigate to: jSurvival → Comparing Survival\n")
cat("   - Or use: Analyses → ClinicoPath → jSurvival → Comparing Survival\n\n")

cat("3. Variable Assignment:\n")
cat("   - Time Variable: Select survival time column\n")
cat("   - Event Variable: Select event indicator\n")
cat("   - Groups: Select grouping variable\n\n")

cat("4. Output Options:\n")
cat("   - Survival curves: Check for Kaplan-Meier plots\n")
cat("   - Statistics: Check for detailed statistical output\n")
cat("   - Risk tables: Check for number at risk tables\n\n")

cat("5. Interpretation:\n")
cat("   - Review log-rank test p-value\n")
cat("   - Examine survival curves visually\n")
cat("   - Check median survival differences\n")
cat("   - Consider clinical significance\n\n")

# Common variable formats
cat("Common Variable Formats:\n")
cat("- Time: months, days, years (numeric)\n")
cat("- Event: 1=event, 0=censored (or TRUE/FALSE)\n")
cat("- Groups: factor levels (e.g., 'Control', 'Treatment')\n\n")

cat("Troubleshooting:\n")
cat("- Ensure no missing values in key variables\n")
cat("- Check that event rates are reasonable (>5% per group)\n")
cat("- Verify adequate sample sizes per group\n")
cat("- Confirm proper variable types\n")
```

# Conclusion

The **comparingsurvival** function provides comprehensive tools for group-wise survival comparisons, from basic two-group analyses to complex multi-biomarker evaluations. Proper application requires attention to both statistical methodology and clinical interpretation.

## Key Takeaways

1. **Statistical Rigor**: Use appropriate methods (Kaplan-Meier, log-rank tests)
2. **Clinical Relevance**: Consider both statistical and clinical significance
3. **Multiple Comparisons**: Adjust for multiple testing when appropriate
4. **Sample Size**: Ensure adequate power for meaningful comparisons
5. **Interpretation**: Balance statistical findings with biological plausibility

## Next Steps

- Explore advanced survival modeling with Cox regression
- Consider competing risks analysis for complex outcomes
- Integrate findings with other ClinicoPath modules
- Develop prognostic models based on significant biomarkers

---

*This vignette demonstrates comprehensive survival comparison methodology using the comparingsurvival function. Apply these principles to your own clinical datasets for robust survival analysis.*