---
title: "Using jsurvival for Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using jsurvival for Survival Analysis}
  %\VignetteEngine{rmarkdown::html_vignette}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The `jsurvival` package provides functions to help researchers carry out
survival analyses and produce publication ready tables and plots.
It supports several workflows ranging from simple survival summaries to
multi‑arm comparisons.

This vignette gives a brief tour of the main functions with example code
using the built in `histopathology` data set.

# Example data

The package ships with a small demonstration data frame `histopathology`.
Some of the available variables are:

- `Sex`
- `Race`
- `Group`
- `Grade`
- `SurgeryDate`
- `LastFollowUpDate`
- `Death`
- `DeathTime`

```{r load-data, eval=FALSE}
# load the included data
library(ClinicoPath)
data(histopathology)
head(histopathology)
```

# Overall survival

The core function of the package is `survival()`, which calculates
Kaplan–Meier estimates and related summaries.  The minimal usage requires
a data frame with the follow up time and an event indicator.

```{r basic-survival, eval=FALSE}
res <- survival(
  data = histopathology,
  elapsedtime = DeathTime,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = Group,
  ci95 = TRUE,
  risktable = TRUE
)

# results object contains tables and plots
res$plot
```

# Continuous predictors

For analyses with a continuous explanatory variable the
`survivalcont()` function provides an interface that can optionally find a
cut‑point for dichotomisation.

```{r continuous, eval=FALSE}
res <- survivalcont(
  data = histopathology,
  elapsedtime = DeathTime,
  outcome = Death,
  outcomeLevel = "1",
  contexpl = `Anti-X-intensity`,
  findcut = TRUE
)
```

# Odds ratios

`oddsratio()` computes odds ratios for binary outcomes with optional
adjustment for explanatory factors.

```{r oddsratio, eval=FALSE}
res <- oddsratio(
  data = histopathology,
  outcome = Mortality5yr,
  explanatory = Group
)
```

# Single arm analyses

In situations where only a single treatment group is available,
`singlearm()` summarises survival and provides median estimates.

```{r singlearm, eval=FALSE}
res <- singlearm(
  data = histopathology,
  elapsedtime = DeathTime,
  outcome = Death,
  outcomeLevel = "1"
)
```

# Multiple groups

The `multisurvival()` function extends the framework to multiple
explanatory variables for a comprehensive analysis across several groups.

```{r multisurvival, eval=FALSE}
res <- multisurvival(
  data = histopathology,
  elapsedtime = DeathTime,
  outcome = Death,
  outcomeLevel = "1",
  explanatory = c(Group, Grade, Race)
)
```

# Further reading

The documentation of each function contains many more options.
See `help(package = "jsurvival")` for details and consult the
package website for worked examples.
