---
title: "Enhanced ROC Analysis for Clinical Biomarker Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enhanced ROC Analysis for Clinical Biomarker Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `timeroc()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `timerocClass$new(options = timerocOptions$new(...), data = mydata)`.

# Enhanced ROC Analysis

## 1. Overview

The **Enhanced ROC Analysis** module (`timeroc`) evaluates how well continuous
biomarkers predict clinical outcomes. It provides a dual-mode architecture:

- **Time-Dependent ROC** (via `timeROC` package): Evaluates marker performance
  at specific timepoints using cumulative/dynamic ROC curves with Inverse
  Probability of Censoring Weights (IPCW).
- **General Binary ROC** (via `pROC` package): Standard diagnostic ROC for
  binary classification without a time component.

Key capabilities:

- IPCW weighting methods: Kaplan-Meier (marginal), Cox model, Aalen additive
- Asymptotic confidence intervals via influence functions
- Optimal cutoff calculation using Youden index
- Multi-marker comparison with DeLong, Bootstrap, or Venkatraman tests
- Clinical interpretation with severity-graded notices

### Test Datasets

| Dataset | Description | N | Variables | Primary Use |
|---------|------------|---|-----------|-------------|
| `timeroc_test` | Breast cancer recurrence | 200 | PatientID, FollowUpMonths, Recurrence (0/1), RecurrenceFactor, Age, Stage, Ki67, GeneScore, NoiseMarker | Primary testing |
| `timeroc_cancer_biomarker` | General cancer biomarkers | varies | Multiple continuous markers | Extended testing |
| `timeroc_multi_biomarker` | Multi-marker scenarios | varies | Multiple markers for comparison | ROC comparison |
| `timeroc_edge_cases` | Edge case scenarios | varies | Designed extremes | Edge case testing |

---

## 2. Basic Time-Dependent ROC

The default mode evaluates how a marker discriminates events from non-events
at specific follow-up timepoints. The timeROC package computes
cumulative/dynamic AUC with IPCW correction for censoring.

```{r basic-timedep}
data("timeroc_test")

timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  method = "marginal",
  bootstrapCI = FALSE,
  plotROC = TRUE,
  plotAUC = TRUE,
  showOptimalCutoff = TRUE,
  showMarkerStats = TRUE,
  timetypeoutput = "months"
)
```

**What this produces:**

- AUC table with values at months 12, 36, and 60
- ROC curves for each timepoint (one curve per timepoint)
- AUC-over-time plot showing marker performance trajectory
- Marker descriptive statistics (N, mean, median, SD, IQR, range, events, event rate)
- Optimal cutoff table with Youden index for each timepoint
- Clinical interpretation text

---

## 3. IPCW Weighting Methods

The `method` option controls how censoring is handled via Inverse Probability
of Censoring Weights. Three approaches are available:

### 3a. Kaplan-Meier (Marginal) Weighting

The default. Assumes censoring is independent of the marker.
Uses the Kaplan-Meier estimator for censoring distribution.

```{r method-marginal}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  method = "marginal",
  timepoints = "12, 36, 60",
  timetypeoutput = "months"
)
```

### 3b. Cox Model Weighting

Uses a Cox proportional hazards model for censoring. Appropriate when
censoring may depend on covariates (informative censoring).

```{r method-cox}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  method = "cox",
  timepoints = "12, 36, 60",
  timetypeoutput = "months"
)
```

### 3c. Aalen Additive Model Weighting

Uses the Aalen additive hazards model. Provides flexible non-proportional
adjustment for censoring.

```{r method-aalen}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  method = "aalen",
  timepoints = "12, 36, 60",
  timetypeoutput = "months"
)
```

---

## 4. Confidence Intervals

Enable asymptotic confidence intervals using influence function-based variance
estimation (not bootstrap, despite the option name). This computes the
variance of the AUC estimator analytically and provides 95% CIs.

```{r confidence-intervals}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  method = "marginal",
  bootstrapCI = TRUE,
  timetypeoutput = "months",
  plotAUC = TRUE
)
```

**With CI enabled:**

- AUC table includes SE, 95% CI Lower, and 95% CI Upper columns
- AUC-over-time plot shows confidence ribbon
- Interpretation text includes p-values testing AUC against 0.5
- Model comparison (`compareBaseline`) becomes fully functional

---

## 5. Binary ROC Analysis

Switch to general binary ROC (no time component needed) for standard
diagnostic evaluation. Uses the pROC package with `direction = "auto"`
to auto-detect whether higher or lower marker values indicate cases.

```{r binary-roc}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "GeneScore",
  analysisType = "binary",
  plotROC = TRUE,
  youdenIndex = TRUE
)
```

**What this produces:**

- Binary ROC table: AUC, SE, 95% CI, optimal cutoff, sensitivity, specificity
- Binary ROC curve plot with AUC in legend
- Diagnostic performance summary with clinical interpretation
- AUC classification: Excellent (>=0.9), Good (0.8-0.9), Fair (0.7-0.8), Poor (0.6-0.7), No discrimination (<0.6)

---

## 6. Factor Outcome with Level Selection

When the outcome is a factor (e.g., "Recurrence" / "No Recurrence"),
select the event level using `outcomeLevel`. The function converts
the factor to binary 0/1 internally.

```{r factor-outcome}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "RecurrenceFactor",
  outcomeLevel = "Recurrence",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  timetypeoutput = "months"
)
```

This also works in binary mode:

```{r factor-outcome-binary}
timeroc(
  data = timeroc_test,
  outcome = "RecurrenceFactor",
  outcomeLevel = "Recurrence",
  marker = "Ki67",
  analysisType = "binary",
  youdenIndex = TRUE
)
```

---

## 7. ROC Comparison (DeLong Test)

Compare multiple markers against the primary marker using the DeLong test
(default), bootstrap test, or Venkatraman test. This is available in
binary ROC mode only.

### 7a. DeLong Test (Default)

```{r roc-comparison-delong}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  markers = c("GeneScore", "NoiseMarker"),
  analysisType = "binary",
  compareROCs = TRUE,
  rocComparison = "delong",
  plotROC = TRUE
)
```

**What this produces:**

- ROC comparison table: pairwise comparisons with test statistic, p-value, interpretation
- Multi-marker ROC plot: all curves overlaid with AUC in legend
- Individual entries in binary ROC table for the primary marker

### 7b. Bootstrap Test

```{r roc-comparison-bootstrap}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  markers = c("GeneScore"),
  analysisType = "binary",
  compareROCs = TRUE,
  rocComparison = "bootstrap"
)
```

### 7c. Venkatraman Test

```{r roc-comparison-venkatraman}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  markers = c("GeneScore"),
  analysisType = "binary",
  compareROCs = TRUE,
  rocComparison = "venkatraman"
)
```

---

## 8. Optimal Cutoffs and Youden Index

### Time-Dependent Mode

Optimal cutoffs are computed at each timepoint by maximizing the Youden
index (sensitivity + specificity - 1) from the time-dependent ROC curve.

```{r cutoffs-timedep}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 24, 36, 48, 60",
  showOptimalCutoff = TRUE,
  timetypeoutput = "months"
)
```

The cutoff table shows: timepoint, optimal cutoff value, sensitivity at
cutoff, specificity at cutoff, and Youden index.

### Binary Mode

In binary mode, the Youden index is calculated from the standard ROC curve.
Disable with `youdenIndex = FALSE` to suppress the calculation.

```{r cutoffs-binary}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "binary",
  youdenIndex = TRUE
)
```

```{r cutoffs-binary-no-youden}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "binary",
  youdenIndex = FALSE
)
```

---

## 9. Display Options

### 9a. Marker Statistics

Toggle descriptive statistics for the marker variable (time-dependent mode).

```{r marker-stats-off}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  showMarkerStats = FALSE,
  timepoints = "12, 36, 60",
  timetypeoutput = "months"
)
```

### 9b. Baseline Model Comparison

Compare marker performance against AUC = 0.5 (random). Requires confidence
intervals to be enabled for p-value computation.

```{r compare-baseline}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  method = "marginal",
  bootstrapCI = TRUE,
  compareBaseline = TRUE,
  timetypeoutput = "months"
)
```

### 9c. Smooth AUC Curve

Apply loess smoothing to the AUC-over-time plot. Requires more than
2 timepoints to function; otherwise falls back to a straight line.

```{r smooth-auc}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "6, 12, 18, 24, 36, 48, 60",
  plotAUC = TRUE,
  smoothAUC = TRUE,
  timetypeoutput = "months"
)
```

### 9d. Time Unit Labels

Change the time unit for display in plots and interpretation text.
Does not transform the numeric values.

```{r time-units-years}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  timetypeoutput = "years"
)
```

### 9e. Suppress Plots

Turn off ROC curve plot and/or AUC-over-time plot for table-only output.

```{r no-plots}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  plotROC = FALSE,
  plotAUC = FALSE,
  timetypeoutput = "months"
)
```

---

## 10. Edge Cases

### 10a. Timepoints Beyond Follow-Up

When all specified timepoints exceed the maximum follow-up time,
the function automatically substitutes quartiles of observed follow-up.

```{r edge-beyond-followup}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "120, 240, 360",
  timetypeoutput = "months"
)
```

### 10b. Invalid Timepoints String

Non-parseable timepoint strings fall back to the defaults (12, 36, 60).

```{r edge-invalid-timepoints}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "abc, xyz",
  timetypeoutput = "months"
)
```

### 10c. Noise Marker (AUC Near 0.5)

A random noise marker should produce AUC close to 0.5, triggering
a "Limited Clinical Utility" warning.

```{r edge-noise-marker}
timeroc(
  data = timeroc_test,
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "NoiseMarker",
  analysisType = "binary",
  youdenIndex = TRUE
)
```

### 10d. Single Timepoint

Analysis with only one evaluation timepoint.

```{r edge-single-timepoint}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "Ki67",
  analysisType = "timedep",
  timepoints = "24",
  timetypeoutput = "months"
)
```

### 10e. Many Timepoints with Smoothing

Dense timepoints with loess smoothing for a smooth AUC trajectory.

```{r edge-many-timepoints}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "GeneScore",
  analysisType = "timedep",
  timepoints = "3, 6, 9, 12, 18, 24, 30, 36, 42, 48, 54, 60",
  smoothAUC = TRUE,
  plotAUC = TRUE,
  timetypeoutput = "months"
)
```

---

## 11. Clinical Interpretation

The module provides automated clinical interpretation at multiple levels:

### Time-Dependent Mode

- **Per-timepoint interpretation:** AUC classification (excellent/good/fair/poor/failed)
  with significance testing against AUC = 0.5 (when CI enabled)
- **Overall assessment:** Mean AUC, performance trend (improving/declining/stable),
  best-performing timepoint
- **Method context:** Explanation of the IPCW weighting method used
- **Notices:** Severity-graded feedback (ERROR, STRONG_WARNING, WARNING, INFO)
  for data quality issues, reversed markers, and limited clinical utility

### Binary Mode

- **AUC classification:** Excellent (>=0.9), Good (0.8-0.9), Fair (0.7-0.8),
  Poor (0.6-0.7), No discrimination (<0.6)
- **Diagnostic performance summary:** AUC with 95% CI, interpretation of
  discriminative ability
- **Optimal threshold:** Youden index-based cutoff with sensitivity and specificity

```{r clinical-interpretation}
timeroc(
  data = timeroc_test,
  elapsedtime = "FollowUpMonths",
  outcome = "Recurrence",
  outcomeLevel = "1",
  marker = "GeneScore",
  analysisType = "timedep",
  timepoints = "12, 36, 60",
  method = "marginal",
  bootstrapCI = TRUE,
  showOptimalCutoff = TRUE,
  showMarkerStats = TRUE,
  compareBaseline = TRUE,
  timetypeoutput = "months"
)
```

---

## 12. References

### Statistical Methods

- Blanche P, Dartigues JF, Jacqmin-Gadda H (2013). Estimating and comparing
  time-dependent areas under receiver operating characteristic curves for
  censored event times with competing risks. *Statistics in Medicine*, 32(30),
  5381-5397.
- Hung H, Chiang CT (2010). Estimation methods for time-dependent AUC models
  with survival data. *Canadian Journal of Statistics*, 38(1), 8-26.
- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas under
  two or more correlated receiver operating characteristic curves: a
  nonparametric approach. *Biometrics*, 44(3), 837-845.
- Robin X, Turck N, Hainard A, et al. (2011). pROC: an open-source package
  for R and S+ to analyze and compare ROC curves. *BMC Bioinformatics*, 12, 77.
- Youden WJ (1950). Index for rating diagnostic tests. *Cancer*, 3(1), 32-35.

### R Packages

- **timeROC**: Time-dependent ROC curve estimation
- **pROC**: Binary ROC curve analysis, DeLong test, bootstrap comparison
- **ggplot2**: AUC-over-time plot visualization
- **glue**: HTML template rendering
- **scales**: Axis label formatting

### AUC Interpretation Guidelines

| AUC Range | Interpretation | Clinical Guidance |
|-----------|---------------|-------------------|
| 0.90-1.00 | Excellent | Strong discriminative ability; suitable for clinical decision-making |
| 0.80-0.89 | Good | Useful for clinical purposes with appropriate validation |
| 0.70-0.79 | Fair | Limited but potentially useful in combination with other markers |
| 0.60-0.69 | Poor | Insufficient for standalone clinical use |
| 0.50-0.59 | Fail | No better than random chance |
| < 0.50 | Reversed | Marker direction likely inverted; consider negating values |
