---
title: "PLS Cox Regression for High-Dimensional Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLS Cox Regression for High-Dimensional Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `plscox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `plscoxClass$new(options = plscoxOptions$new(...), data = mydata)`.

# PLS Cox Regression

## Overview

The **PLS Cox Regression** module (`plscox`) combines Partial Least Squares (PLS)
dimensionality reduction with Cox proportional hazards modeling for high-dimensional
survival data. Instead of selecting individual variables like LASSO, PLS creates
**latent components** -- weighted combinations of all predictors that maximally
explain covariance with the survival outcome.

Key features:

- **Component selection** via cross-validation (log-likelihood, C-index), information criteria (BIC, AIC), or manual
- **Multiple scaling methods** (standardization, unit variance, min-max, none)
- **Advanced PLS settings** (sparse PLS, Q-squared stopping, p-value variable selection)
- **Bootstrap validation** (Harrell optimism-corrected C-index)
- **Permutation testing** for overall model significance
- **Risk group stratification** with Kaplan-Meier survival curves
- **Variable importance** (Cox-weighted PLS loadings)
- **Data suitability assessment** (traffic-light system with 6 checks)

---

## PLS vs LASSO: Different Philosophies

| Approach | How It Works | Output |
|----------|-------------|--------|
| **LASSO** | Selects individual variables, drops others | Sparse coefficient vector |
| **PLS** | Creates weighted combinations of ALL variables | Component scores + loadings |

**Use PLS when:**
- Variables are highly correlated (metabolomics, gene expression)
- You want to retain information from all variables
- The underlying signal comes from latent biological processes
- p >> n and LASSO produces unstable selections

**Use LASSO when:**
- You want to identify individual important predictors
- A sparse, interpretable model is needed
- Variables are relatively independent

---

## How PLS Cox Works

1. **Extract PLS components**: Find linear combinations of X that maximize
   covariance with the survival response
2. **Select components**: Choose the optimal number via CV or information criteria
3. **Fit Cox model**: Use selected components as predictors in Cox regression
4. **Interpret**: Back-project component loadings to understand original variable contributions

---

## Datasets Used in This Guide

```{r datasets-table, echo=FALSE, results='asis'}
cat("| Dataset | N | Events | Predictors | Time Var | Status Var | Description |
|---------|---|--------|------------|----------|------------|-------------|
| `plscox_metabolomics` | 120 | ~50% | 80 metabolites + 3 clinical | `survival_months` | `death` | Metabolomic survival study with 3 latent pathways |
| `plscox_small` | 50 | ~60% | 25 biomarkers | `time_months` | `status` | Small sample edge case |
| `plscox_genomic` | 60 | ~55% | 200 genes | `os_time` | `os_event` (numeric 0/1) | True p>>n with missing data |
")
```

```{r load-datasets}
data_path2 <- "data/"

# Load metabolomics dataset (n=120, p=80)
load(paste0(data_path2, "plscox_metabolomics.rda"))
cat("Metabolomics: N =", nrow(plscox_metabolomics),
    "| Events =", sum(plscox_metabolomics$death == "Dead"),
    "| Predictors =", sum(grepl("^METAB_", names(plscox_metabolomics))), "+ 3 clinical\n")

# Load small dataset (n=50, p=25)
load(paste0(data_path2, "plscox_small.rda"))
cat("Small: N =", nrow(plscox_small),
    "| Events =", sum(plscox_small$status == "Dead"),
    "| Predictors =", sum(grepl("^MARKER_", names(plscox_small))), "\n")

# Load genomic dataset (n=60, p=200)
load(paste0(data_path2, "plscox_genomic.rda"))
cat("Genomic: N =", nrow(plscox_genomic),
    "| Events =", sum(plscox_genomic$os_event == 1),
    "| Predictors =", sum(grepl("^GENE_", names(plscox_genomic))),
    "| Missing values =", sum(is.na(plscox_genomic)), "\n")
```

---

## 1. Basic PLS Cox Analysis (Default Settings)

This demonstrates all default settings: 5 components, 10-fold CV, CV log-likelihood
selection, standardization scaling, and all outputs/plots enabled.

```{r basic-pls-defaults}
metab_predictors <- c("age", "gender", "bmi",
                       paste0("METAB_", sprintf("%03d", 1:80)))

plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  cross_validation = "k10",
  component_selection = "cv_loglik",
  scaling_method = "standardize",
  suitabilityCheck = TRUE,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_validation = TRUE,
  plot_survival = TRUE,
  risk_groups = 3,
  confidence_intervals = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

Expected: PLS should identify 2-3 important components reflecting the underlying
pathway structure. Metabolites in blocks 1-15, 25-40, and 55-70 should have
high loadings on the first few components.

---

## 2. Component Selection Methods

### Cross-Validated Log-Likelihood (default)

```{r cv-loglik-selection}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  plot_validation = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

### Cross-Validated C-Index

```{r cv-cindex-selection}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  component_selection = "cv_cindex",
  cross_validation = "k10",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### BIC (no cross-validation needed)

```{r bic-selection}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  component_selection = "bic",
  cross_validation = "none",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### AIC

```{r aic-selection}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  component_selection = "aic",
  cross_validation = "none",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### Manual (fixed number of components)

```{r manual-selection}
# Use exactly 3 components (based on domain knowledge)
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 3,
  component_selection = "manual",
  cross_validation = "none",
  plot_components = TRUE,
  plot_loadings = TRUE,
  feature_importance = TRUE
)
```

---

## 3. Cross-Validation Methods

### 10-Fold CV (default)

```{r cv-k10}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  cross_validation = "k10",
  component_selection = "cv_loglik",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### 5-Fold CV (faster, slightly more bias)

```{r cv-k5}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  cross_validation = "k5",
  component_selection = "cv_loglik",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### Leave-One-Out CV (small samples only)

```{r cv-loo}
small_predictors <- paste0("MARKER_", sprintf("%02d", 1:25))

plscox(
  data = plscox_small,
  time = "time_months",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = small_predictors,
  pls_components = 5,
  cross_validation = "loo",
  component_selection = "cv_loglik",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 4. Scaling Methods

Variable scaling is critical for PLS since it operates on covariances.

### Standardization (Z-scores, default)

```{r scaling-standardize}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  scaling_method = "standardize",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

### Unit Variance Scaling (no centering)

```{r scaling-unit-variance}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  scaling_method = "unit_variance",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

### Min-Max Scaling (range [0,1])

```{r scaling-minmax}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  scaling_method = "minmax",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

### No Scaling

```{r scaling-none}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  scaling_method = "none",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

---

## 5. Advanced PLS Settings

### Sparse PLS (automatic variable selection within components)

```{r sparse-pls}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  sparse_pls = TRUE,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

### Convergence Tolerance

```{r tolerance-strict}
# Strict tolerance for higher precision
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  tolerance = 1e-10,
  component_selection = "manual",
  feature_importance = TRUE
)
```

### Q-Squared Limit (PLS stopping criterion)

```{r limq2set}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  limQ2set = 0.5,
  component_selection = "manual",
  feature_importance = TRUE
)
```

### P-Value Based Variable Selection

```{r pvals-expli}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  pvals_expli = TRUE,
  alpha_pvals_expli = 0.05,
  component_selection = "manual",
  feature_importance = TRUE
)
```

### Tie Handling Method

```{r tie-method-breslow}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  tie_method = "breslow",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

---

## 6. Bootstrap Validation

Bootstrap validation assesses model overfitting using Harrell's optimism-corrected
C-index. Each bootstrap iteration: (1) fit model on bootstrap sample, (2) assess
on bootstrap and original data, (3) compute optimism.

```{r bootstrap-validation}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  bootstrap_validation = TRUE,
  n_bootstrap = 100,
  prediction_accuracy = TRUE
)
```

---

## 7. Permutation Testing

Test whether the PLS components capture real survival signal or random patterns.
The p-value is the proportion of permuted C-indices that exceed the original.

```{r permutation-testing}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  permutation_test = TRUE,
  n_permutations = 50,  # Use 100+ for publication
  prediction_accuracy = TRUE
)
```

---

## 8. Risk Stratification

Patients are stratified into risk groups based on quantiles of the PLS-derived
linear predictor from the Cox model.

### Binary Risk Groups

```{r risk-groups-2}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 3,
  component_selection = "manual",
  risk_groups = 2,
  plot_survival = TRUE,
  confidence_intervals = TRUE
)
```

### Quartile Risk Groups

```{r risk-groups-4}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 3,
  component_selection = "manual",
  risk_groups = 4,
  plot_survival = TRUE,
  confidence_intervals = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 9. Data Suitability Assessment

The traffic-light assessment checks 6 criteria: events-per-variable, reduction need,
sample size, event rate, multicollinearity, and data quality.

```{r suitability-metabolomics}
# Metabolomics data: expected mostly green
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 3,
  component_selection = "manual",
  suitabilityCheck = TRUE,
  plot_components = FALSE,
  plot_loadings = FALSE,
  plot_scores = FALSE,
  plot_validation = FALSE,
  plot_survival = FALSE
)
```

```{r suitability-genomic}
# Genomic p>>n data: expected yellow/red for EPV and data quality
gene_predictors <- paste0("GENE_", sprintf("%03d", 1:200))

plscox(
  data = plscox_genomic,
  time = "os_time",
  status = "os_event",
  outcomeLevel = "1",
  censorLevel = "0",
  predictors = gene_predictors,
  pls_components = 5,
  component_selection = "manual",
  suitabilityCheck = TRUE,
  plot_components = FALSE,
  plot_loadings = FALSE,
  plot_scores = FALSE,
  plot_validation = FALSE,
  plot_survival = FALSE
)
```

---

## 10. Small Sample Analysis

```{r small-sample-analysis}
plscox(
  data = plscox_small,
  time = "time_months",
  status = "status",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = paste0("MARKER_", sprintf("%02d", 1:25)),
  pls_components = 3,
  cross_validation = "k5",
  component_selection = "cv_loglik",
  scaling_method = "standardize",
  risk_groups = 2,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_survival = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE,
  suitabilityCheck = TRUE
)
```

---

## 11. High-Dimensional Genomic Analysis (p >> n)

This is the core PLS use case: more genes than patients.

```{r genomic-pls}
plscox(
  data = plscox_genomic,
  time = "os_time",
  status = "os_event",
  outcomeLevel = "1",
  censorLevel = "0",
  predictors = gene_predictors,
  pls_components = 5,
  component_selection = "bic",
  cross_validation = "none",
  scaling_method = "standardize",
  risk_groups = 3,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_survival = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE,
  suitabilityCheck = TRUE
)
```

---

## 12. All Plots Demonstration

```{r all-plots-demo}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  risk_groups = 3,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_validation = TRUE,
  plot_survival = TRUE,
  feature_importance = TRUE,
  confidence_intervals = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 13. Full Validation Pipeline

Combine bootstrap validation and permutation testing for publication-quality results.

```{r full-validation}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 10,
  component_selection = "cv_cindex",
  cross_validation = "k5",
  scaling_method = "standardize",
  bootstrap_validation = TRUE,
  n_bootstrap = 100,
  permutation_test = TRUE,
  n_permutations = 50,
  risk_groups = 4,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_validation = TRUE,
  plot_survival = TRUE,
  confidence_intervals = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE,
  suitabilityCheck = TRUE
)
```

---

## 14. Edge Cases

### Single Component Model

```{r edge-single-component}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 1,
  component_selection = "manual",
  plot_loadings = TRUE,
  feature_importance = TRUE
)
```

### Few Predictors

```{r edge-few-predictors}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = c("age", "bmi"),
  pls_components = 1,
  component_selection = "manual",
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

### Minimal Output (no plots, no optional tables)

```{r edge-minimal-output}
plscox(
  data = plscox_metabolomics,
  time = "survival_months",
  status = "death",
  outcomeLevel = "Dead",
  censorLevel = "Alive",
  predictors = metab_predictors,
  pls_components = 3,
  component_selection = "manual",
  suitabilityCheck = FALSE,
  plot_components = FALSE,
  plot_loadings = FALSE,
  plot_scores = FALSE,
  plot_validation = FALSE,
  plot_survival = FALSE,
  confidence_intervals = FALSE,
  feature_importance = FALSE,
  prediction_accuracy = FALSE,
  bootstrap_validation = FALSE,
  permutation_test = FALSE
)
```

---

## Interpreting Results

### Model Summary

The Model Summary HTML shows: sample size, events, predictors, number of PLS
components used, selection method, scaling, CV method, tie handling, tolerance,
and training C-index with likelihood ratio and Wald tests.

### Component Selection Table

| Column | Meaning |
|--------|---------|
| Components | Number of PLS components considered |
| CV Score | Cross-validation score (log-likelihood, C-index, AIC, or BIC) |
| SE | Standard error of CV score (when available) |
| C-Index | Concordance index for that number of components (AIC/BIC only) |
| Selected | "Yes" marks the optimal number of components |

### Model Coefficients Table

| Column | Meaning |
|--------|---------|
| PLS Component | Component identifier (PLS_1, PLS_2, ...) |
| Coefficient | Cox regression coefficient |
| Hazard Ratio | exp(coefficient) |
| HR Lower CI / HR Upper CI | 95% confidence interval for HR |
| Standard Error | SE of the coefficient |
| Z-value | Wald test statistic |
| p-value | Significance of component in Cox model |

### Variable Loadings Table

Variables are sorted by Cox-weighted importance: `|loadings| * |cox_coefficients|`.
This accounts for both how much a variable contributes to each component AND
how much each component matters for survival.

### Risk Stratification Table

Patients are divided into quantile-based risk groups. Risk Group 1 is always the
reference (HR = 1.0). Higher-numbered groups should have progressively higher HRs.

---

## PLS vs PCA for Survival Analysis

| Method | Supervision | Components Maximize |
|--------|------------|-------------------|
| **PCA** | Unsupervised | Variance in X only |
| **PLS** | Supervised | Covariance between X and Y (survival) |

PLS components are specifically constructed to predict survival, while PCA
components may capture variance unrelated to the outcome.

---

## Common Pitfalls

1. **Too many components**: More components can mean overfitting. Always use
   cross-validation to select the optimal number.

2. **Ignoring variable scaling**: Metabolites on different scales dominate PLS
   if not standardized. Always use `scaling_method = "standardize"` (the default).

3. **Not validating**: Use bootstrap validation and/or permutation testing to
   assess reliability. The training C-index overestimates true performance.

4. **Interpreting loadings as independent effects**: A high loading means the
   variable contributes to a component, not that it has an independent causal
   effect. Groups of correlated variables share loading magnitude.

5. **Using LOO CV for large datasets**: Leave-one-out is computationally
   expensive and can be unstable. Use 5- or 10-fold CV for n > 50.

6. **Sparse PLS with too few components**: Sparse PLS may return NULL component
   scores for some configurations. If this happens, try increasing components
   or disabling sparse mode.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** (`lassocox`) | Want sparse individual variable selection |
| **Adaptive LASSO** (`adaptivelasso`) | Oracle property variable selection |
| **NCV Reg Cox** (`ncvregcox`) | SCAD/MCP non-convex penalties |
| **High-Dimensional Cox** (`highdimcox`) | Multiple regularization methods |
| **PCA Cox** (`pcacox`) | Unsupervised dimensionality reduction |
| **Multivariable Survival** | Standard Cox with few predictors |

---

## References

- Bastien P, Esposito Vinzi V, Tenenhaus M. PLS generalised linear regression. *Comput Stat Data Anal*. 2005;48(1):17-46.
- Boulesteix AL, Strimmer K. Partial least squares: a versatile tool for the analysis of high-dimensional genomic data. *Brief Bioinform*. 2007;8(1):32-44.
- Li H, Gui J. Partial Cox regression analysis for high-dimensional microarray gene expression data. *Bioinformatics*. 2004;20(Suppl 1):i208-i215.
- Mevik BH, Wehrens R. The pls package: principal component and partial least squares regression in R. *J Stat Softw*. 2007;18(2):1-23.
