---
title: "PLS Cox Regression for High-Dimensional Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PLS Cox Regression for High-Dimensional Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `plscox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `plscoxClass$new(options = plscoxOptions$new(...), data = mydata)`.

# PLS Cox Regression

## Overview

The **PLS Cox Regression** module (`plscox`) combines Partial Least Squares (PLS)
dimensionality reduction with Cox proportional hazards modeling for high-dimensional
survival data. Instead of selecting individual variables like LASSO, PLS creates
**latent components** — weighted combinations of all predictors that maximally
explain survival variation.

Key features:

- **Multiple PLS algorithms** (NIPALS, Kernel PLS, Wide Kernel PLS)
- **Component selection** via cross-validation, information criteria, or manual
- **Multiple scaling methods** (standardization, normalization, min-max)
- **Bootstrap validation** and **permutation testing**
- **Risk group stratification** with survival curves
- **Feature importance** (variable contributions to components)

---

## PLS vs LASSO: Different Philosophies

| Approach | How It Works | Output |
|----------|-------------|--------|
| **LASSO** | Selects individual variables, drops others | Sparse coefficient vector |
| **PLS** | Creates weighted combinations of ALL variables | Component scores + loadings |

**Use PLS when:**
- Variables are highly correlated (metabolomics, gene expression)
- You want to retain information from all variables
- The underlying signal comes from latent biological processes
- p >> n and LASSO produces unstable selections

**Use LASSO when:**
- You want to identify individual important predictors
- A sparse, interpretable model is needed
- Variables are relatively independent

---

## How PLS Cox Works

1. **Extract PLS components**: Find linear combinations of X that maximize
   covariance with survival response
2. **Select components**: Choose the optimal number via CV or information criteria
3. **Fit Cox model**: Use selected components as predictors in Cox regression
4. **Interpret**: Back-project component loadings to understand original variable contributions

$$T_k = X w_k \quad \text{(component scores)}$$
$$h(t) = h_0(t) \exp(\alpha_1 T_1 + \alpha_2 T_2 + \cdots + \alpha_K T_K)$$

where $w_k$ are PLS weight vectors and $T_k$ are component scores.

---

## PLS Algorithms Compared

| Algorithm | Speed | Memory | Best For |
|-----------|-------|--------|----------|
| **NIPALS** (default) | Fast | Low | Standard use, moderate p |
| **Kernel PLS** | Fast | Higher | Large p, when NIPALS is slow |
| **Wide Kernel PLS** | Fast | Moderate | Very wide data (p >> n) |

---

## Datasets Used in This Guide

| Dataset | N | Events | Predictors | Description |
|---------|---|--------|------------|-------------|
| Synthetic "metabolomics" | 120 | ~60 | 80 metabolites + 3 clinical | Metabolomic survival study |
| Synthetic "small" | 50 | ~30 | 25 biomarkers | Small sample edge case |

---

## 1. Basic PLS Cox Analysis

```{r metabolomics-data}
# Simulate metabolomics survival data
# Signal comes from 2-3 latent pathways affecting clusters of metabolites
set.seed(42)
n <- 120
p_metab <- 80

# Latent biological factors
pathway1 <- rnorm(n)  # Affects metabolites 1-15
pathway2 <- rnorm(n)  # Affects metabolites 20-35
pathway3 <- rnorm(n)  # Affects metabolites 50-65

# Generate metabolite measurements from latent factors + noise
metab_matrix <- matrix(rnorm(n * p_metab, sd = 0.5), nrow = n)
colnames(metab_matrix) <- paste0("METAB_", sprintf("%03d", 1:p_metab))

# Pathway 1 loads on metabolites 1-15
for (j in 1:15) {
  loading <- runif(1, 0.3, 0.8) * sample(c(-1, 1), 1)
  metab_matrix[, j] <- metab_matrix[, j] + loading * pathway1
}
# Pathway 2 loads on metabolites 20-35
for (j in 20:35) {
  loading <- runif(1, 0.3, 0.7) * sample(c(-1, 1), 1)
  metab_matrix[, j] <- metab_matrix[, j] + loading * pathway2
}
# Pathway 3 loads on metabolites 50-65
for (j in 50:65) {
  loading <- runif(1, 0.2, 0.6) * sample(c(-1, 1), 1)
  metab_matrix[, j] <- metab_matrix[, j] + loading * pathway3
}

# Clinical variables
age <- rnorm(n, 58, 12)
gender <- factor(sample(c("Male", "Female"), n, replace = TRUE))
bmi <- rnorm(n, 26, 4)

# Survival depends on pathway1 and pathway2 + age
lp <- 0.5 * pathway1 - 0.4 * pathway2 + 0.02 * (age - 58)
surv_time <- rweibull(n, shape = 1.3, scale = 30 * exp(-lp * 0.3))
censor_time <- runif(n, 6, 48)
time <- pmax(pmin(surv_time, censor_time), 0.1)
status <- factor(ifelse(surv_time <= censor_time, "Dead", "Alive"),
                 levels = c("Alive", "Dead"))

data_metab <- data.frame(
  follow_up = time, vital_status = status,
  age = age, gender = gender, bmi = bmi,
  as.data.frame(metab_matrix)
)

cat("N:", n, "Events:", sum(status == "Dead"),
    "Event rate:", round(mean(status == "Dead"), 2), "\n")
cat("Metabolites:", p_metab, "Clinical:", 3, "\n")
```

```{r basic-pls}
pred_all <- c("age", "gender", "bmi",
              paste0("METAB_", sprintf("%03d", 1:p_metab)))

plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  pls_algorithm = "nipals",
  cross_validation = "k10",
  component_selection = "cv_loglik",
  scaling_method = "standardize",
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_validation = TRUE,
  plot_survival = TRUE,
  risk_groups = 3,
  confidence_intervals = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

Expected: PLS should identify 2-3 important components reflecting the underlying
pathway structure. Metabolites 1-15 and 20-35 should have high loadings on the
first two components.

---

## 2. Component Selection Methods

### Cross-validated log-likelihood (default)

```{r cv-loglik}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 10,  # Try up to 10
  component_selection = "cv_loglik",
  cross_validation = "k10",
  plot_validation = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

### Cross-validated C-index

```{r cv-cindex}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 10,
  component_selection = "cv_cindex",
  cross_validation = "k10",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### Information criteria (BIC, AIC)

```{r bic}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 10,
  component_selection = "bic",
  cross_validation = "none",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

```{r aic}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 10,
  component_selection = "aic",
  cross_validation = "none",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### Manual (fixed number)

```{r manual}
# Use exactly 3 components (based on domain knowledge)
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 3,
  component_selection = "manual",
  cross_validation = "none",
  plot_components = TRUE,
  plot_loadings = TRUE,
  feature_importance = TRUE
)
```

---

## 3. PLS Algorithms

### Kernel PLS (faster for large p)

```{r kernel}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  pls_algorithm = "kernel",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  plot_validation = TRUE,
  feature_importance = TRUE
)
```

### Wide Kernel PLS (optimized for p >> n)

```{r widekernelpls}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  pls_algorithm = "widekernelpls",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  plot_validation = TRUE,
  feature_importance = TRUE
)
```

---

## 4. Cross-Validation Methods

### Leave-One-Out CV (most thorough, slowest)

```{r loo}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  cross_validation = "loo",
  component_selection = "cv_loglik",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

### 5-Fold CV (faster, less variance than LOO)

```{r k5}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  cross_validation = "k5",
  component_selection = "cv_loglik",
  plot_validation = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 5. Scaling Methods

Variable scaling matters greatly for PLS since it's based on covariances.

```{r scaling-comparison}
# Unit variance scaling
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  scaling_method = "unit_variance",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

```{r minmax}
# Min-max scaling
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  scaling_method = "minmax",
  component_selection = "cv_loglik",
  cross_validation = "k10",
  prediction_accuracy = TRUE
)
```

---

## 6. Bootstrap Validation

Bootstrap validation assesses the stability and reliability of component selection.

```{r bootstrap}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  bootstrap_validation = TRUE,
  n_bootstrap = 100,
  prediction_accuracy = TRUE
)
```

---

## 7. Permutation Testing

Test whether the PLS components capture real survival signal or random patterns.

```{r permutation}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  permutation_test = TRUE,
  n_permutations = 50,  # Use 100+ for publication
  prediction_accuracy = TRUE
)
```

---

## 8. Risk Stratification

```{r risk-groups}
# Stratify into 2, 3, or 4 risk groups
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 3,
  component_selection = "manual",
  risk_groups = 4,
  plot_survival = TRUE,
  confidence_intervals = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 9. Small Sample Analysis

```{r small-data}
# 50 patients, 25 biomarkers
set.seed(99)
n2 <- 50
p2 <- 25

# Simple latent structure
latent <- rnorm(n2)
X2 <- matrix(rnorm(n2 * p2, sd = 0.5), nrow = n2)
for (j in 1:8) X2[, j] <- X2[, j] + runif(1, 0.3, 0.6) * latent
colnames(X2) <- paste0("BIO_", sprintf("%02d", 1:p2))

lp2 <- 0.4 * latent
surv2 <- rweibull(n2, 1.2, 24 * exp(-lp2 * 0.3))
cens2 <- runif(n2, 6, 36)
time2 <- pmax(pmin(surv2, cens2), 0.1)
status2 <- factor(ifelse(surv2 <= cens2, "Dead", "Alive"),
                  levels = c("Alive", "Dead"))

data_small <- data.frame(
  time = time2, status = status2,
  as.data.frame(X2)
)

cat("N:", n2, "Events:", sum(status2 == "Dead"), "\n")
```

```{r small-run}
plscox(
  data = data_small,
  time = "time",
  status = "status",
  predictors = paste0("BIO_", sprintf("%02d", 1:25)),
  pls_components = 3,
  cross_validation = "k5",
  component_selection = "cv_loglik",
  scaling_method = "standardize",
  risk_groups = 2,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_survival = TRUE,
  feature_importance = TRUE,
  prediction_accuracy = TRUE
)
```

---

## 10. All Plots

```{r all-plots}
plscox(
  data = data_metab,
  time = "follow_up",
  status = "vital_status",
  predictors = c("age", "gender", "bmi", paste0("METAB_", sprintf("%03d", 1:80))),
  pls_components = 5,
  component_selection = "cv_loglik",
  cross_validation = "k10",
  risk_groups = 3,
  plot_components = TRUE,
  plot_loadings = TRUE,
  plot_scores = TRUE,
  plot_validation = TRUE,
  plot_survival = TRUE,
  feature_importance = TRUE,
  confidence_intervals = TRUE,
  prediction_accuracy = TRUE
)
```

---

## Interpreting Results

### Component Summary

| Column | Meaning |
|--------|---------|
| Component | PLS component number |
| Variance Explained | Proportion of X variance captured |
| Cumulative Variance | Running total of variance explained |
| Cox Coefficient | Association with survival (from Cox model) |
| P-value | Significance of component in Cox model |

### Variable Loadings

Loadings indicate how much each original variable contributes to each component.
Variables with high absolute loadings on survival-significant components are
the most important predictors.

### Feature Importance

Variables are ranked by their weighted contribution across all selected components.
This provides a single importance score per variable, making it easy to identify
the most survival-relevant metabolites/biomarkers.

---

## PLS vs PCA for Survival Analysis

| Method | Supervision | Components Maximize |
|--------|------------|-------------------|
| **PCA** | Unsupervised | Variance in X only |
| **PLS** | Supervised | Covariance between X and Y (survival) |

PLS components are specifically constructed to predict survival, while PCA
components may capture variance unrelated to the outcome. **PLS is preferred**
for survival prediction.

---

## Common Pitfalls

1. **Too many components**: More components = overfitting. Always use cross-validation
   to select the optimal number. The CV curve should show a clear minimum or plateau.

2. **Ignoring variable scaling**: Metabolites measured on different scales (e.g.,
   nmol/L vs mg/dL) dominate PLS components if not standardized. Always use
   `scaling_method = "standardize"` (the default).

3. **Not validating**: PLS models can overfit, especially with many components.
   Use bootstrap validation and/or permutation testing to assess reliability.

4. **Interpreting loadings as individual variable effects**: A high loading
   means the variable contributes to a component, not that it has an independent
   effect on survival. Groups of correlated variables share loading magnitude.

5. **Using LOO CV for large datasets**: Leave-one-out CV is computationally
   expensive and can be unstable. Use 5- or 10-fold CV for n > 50.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** | Want sparse individual variable selection |
| **Adaptive LASSO** | Oracle property variable selection |
| **SCAD Cox** | Non-convex penalties for unbiased estimation |
| **High-Dimensional Cox** | Multiple regularization methods in one interface |
| **Multivariable Survival** | Standard Cox with few predictors |

---

## References

- Bastien P, Esposito Vinzi V, Tenenhaus M. PLS generalised linear regression. *Comput Stat Data Anal*. 2005;48(1):17-46.
- Boulesteix AL, Strimmer K. Partial least squares: a versatile tool for the analysis of high-dimensional genomic data. *Brief Bioinform*. 2007;8(1):32-44.
- Li H, Gui J. Partial Cox regression analysis for high-dimensional microarray gene expression data. *Bioinformatics*. 2004;20(Suppl 1):i208-i215.
- Mevik BH, Wehrens R. The pls package: principal component and partial least squares regression in R. *J Stat Softw*. 2007;18(2):1-23.
