---
title: "High-Dimensional Cox Regression for Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-Dimensional Cox Regression for Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `highdimcox()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `highdimcoxClass$new(options = highdimcoxOptions$new(...), data = mydata)`.

# High-Dimensional Cox Regression

## Overview

The **High-Dimensional Cox Regression** module (`highdimcox`) provides a unified
interface for regularized Cox regression when the number of predictors is large
relative to sample size. It supports **LASSO**, **Ridge**, **Elastic Net**, and
**Adaptive LASSO** regularization, with optional stability selection for robust
variable identification.

This module is designed for:

- **Genomic survival studies** (gene expression panels, mutation data)
- **Proteomic/metabolomic data** (high-dimensional biomarker panels)
- **Radiomics studies** (texture and morphology features from imaging)
- **Any p >> n survival scenario**

Key features:

- **Multiple regularization methods** in one interface
- **Cross-validation** with configurable lambda selection (min, 1se)
- **Stability selection** via bootstrap for robust variable identification
- **Variable importance** and coefficient visualization
- **Model diagnostics** and performance metrics

---

## Regularization Methods Explained

When you have more predictors than observations (or close to it), standard Cox
regression fails. Regularization adds a penalty to prevent overfitting.

| Method | Penalty | Alpha | Selects Variables? | Best For |
|--------|---------|-------|-------------------|----------|
| **LASSO** | $\lambda \sum \|\beta_j\|$ | 1.0 | Yes (sets to zero) | Sparse models |
| **Ridge** | $\lambda \sum \beta_j^2$ | 0.0 | No (shrinks all) | Prediction, collinearity |
| **Elastic Net** | Mixed L1+L2 | 0.5 | Yes (grouped) | Correlated groups |
| **Adaptive LASSO** | Weighted L1 | 1.0 | Yes (oracle) | Publication models |

The **Elastic Net** (default, alpha=0.5) is the recommended starting point —
it combines the variable selection of LASSO with the stability of Ridge.

---

## Datasets Used in This Guide

| Dataset | N | Events | Predictors | Description |
|---------|---|--------|------------|-------------|
| Synthetic "genomic" | 150 | ~75 | 100 genes + 5 clinical | Gene expression survival study |
| Synthetic "proteomic" | 80 | ~50 | 50 proteins + 3 clinical | Smaller proteomic study |

---

## 1. LASSO Regularization

Pure LASSO selects the sparsest model — ideal when you expect few true predictors.

```{r create-genomic-data}
# Simulate a genomic survival study (150 patients, 100 genes)
set.seed(42)
n <- 150
p_genes <- 100

# Gene expression matrix
gene_matrix <- matrix(rnorm(n * p_genes), nrow = n)
colnames(gene_matrix) <- paste0("GENE_", sprintf("%03d", 1:p_genes))

# True effects: 6 genes affect survival
true_effects <- rep(0, p_genes)
true_effects[c(5, 12, 27, 43, 68, 91)] <- c(0.8, -0.6, 0.5, -0.4, 0.7, -0.5)

# Clinical variables
age <- rnorm(n, 62, 10)
gender <- factor(sample(c("Male", "Female"), n, replace = TRUE))
stage <- factor(sample(c("I", "II", "III"), n, replace = TRUE,
                       prob = c(0.3, 0.4, 0.3)))

# Survival times
lp <- gene_matrix %*% true_effects + 0.02 * (age - 62) + 0.3 * (as.numeric(stage) - 1)
surv_time <- rweibull(n, shape = 1.3, scale = 30 * exp(-as.numeric(lp) * 0.2))
censor_time <- runif(n, 6, 48)
time <- pmax(pmin(surv_time, censor_time), 0.1)
event <- factor(ifelse(surv_time <= censor_time, "Dead", "Alive"),
                levels = c("Alive", "Dead"))

data_genomic <- data.frame(
  survival_months = time,
  vital_status = event,
  age = age, gender = gender, stage = stage,
  as.data.frame(gene_matrix)
)

cat("N:", n, "Events:", sum(event == "Dead"),
    "Event rate:", round(mean(event == "Dead"), 2), "\n")
cat("Predictors:", ncol(data_genomic) - 2, "\n")
```

```{r lasso-run}
# All gene + clinical predictors
pred_vars <- c("age", "gender", "stage",
               paste0("GENE_", sprintf("%03d", 1:p_genes)))

highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "lasso",
  cv_method = "cv_1se",
  cv_folds = 10,
  show_regularization_path = TRUE,
  show_cv_plot = TRUE,
  show_variable_importance = TRUE,
  show_coefficients_table = TRUE,
  show_model_diagnostics = TRUE
)
```

---

## 2. Ridge Regularization

Ridge keeps all variables but shrinks coefficients — better for prediction
when you believe many variables contribute small effects.

```{r ridge}
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "ridge",
  cv_method = "cv_min",
  cv_folds = 10,
  show_cv_plot = TRUE,
  show_coefficients_table = TRUE,
  show_model_diagnostics = TRUE
)
```

---

## 3. Elastic Net (Default)

The recommended starting point — combines LASSO selection with Ridge stability.

```{r elastic-net}
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "elastic_net",
  alpha_value = 0.5,
  cv_method = "cv_1se",
  cv_folds = 10,
  show_regularization_path = TRUE,
  show_cv_plot = TRUE,
  show_variable_importance = TRUE,
  show_coefficients_table = TRUE
)
```

### Alpha Tuning

```{r alpha-tuning}
# More LASSO-like (alpha = 0.8)
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "elastic_net",
  alpha_value = 0.8,
  cv_folds = 10,
  show_coefficients_table = TRUE,
  show_cv_plot = TRUE
)
```

---

## 4. Adaptive LASSO

Two-stage approach with oracle properties — best for publishable models.

```{r adaptive}
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "adaptive_lasso",
  cv_method = "cv_1se",
  cv_folds = 10,
  show_regularization_path = TRUE,
  show_cv_plot = TRUE,
  show_variable_importance = TRUE,
  show_coefficients_table = TRUE
)
```

---

## 5. Stability Selection

Bootstrap-based stability selection identifies variables that are consistently
selected across many random subsamples.

```{r stability}
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "elastic_net",
  alpha_value = 0.5,
  cv_folds = 10,
  stability_selection = TRUE,
  bootstrap_iterations = 100,
  stability_threshold = 0.8,
  show_variable_importance = TRUE,
  show_coefficients_table = TRUE
)
```

Variables selected in >80% of bootstrap iterations are considered robustly important.

---

## 6. CV Method Comparison

The choice of CV lambda selection affects model complexity.

```{r cv-min}
# Minimum CV error — more variables, better fit but risk of overfitting
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "lasso",
  cv_method = "cv_min",
  cv_folds = 10,
  show_coefficients_table = TRUE,
  show_cv_plot = TRUE
)
```

```{r cv-1se}
# 1-SE rule — more parsimonious, better generalization
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars,
  regularization_method = "lasso",
  cv_method = "cv_1se",
  cv_folds = 10,
  show_coefficients_table = TRUE,
  show_cv_plot = TRUE
)
```

---

## 7. Smaller Dataset (Proteomic)

```{r proteomic-data}
# 80 patients, 50 protein markers
set.seed(123)
n2 <- 80
p_prot <- 50

prot_matrix <- matrix(rnorm(n2 * p_prot, mean = 10, sd = 3), nrow = n2)
colnames(prot_matrix) <- paste0("PROT_", sprintf("%02d", 1:p_prot))

# 4 proteins with true effects
true_prot <- rep(0, p_prot)
true_prot[c(3, 17, 28, 42)] <- c(0.15, -0.12, 0.18, -0.10)

lp2 <- prot_matrix %*% true_prot
surv_time2 <- rweibull(n2, shape = 1.1, scale = 24 * exp(-as.numeric(lp2) * 0.5))
censor_time2 <- runif(n2, 6, 36)
time2 <- pmax(pmin(surv_time2, censor_time2), 0.1)
event2 <- factor(ifelse(surv_time2 <= censor_time2, "Dead", "Alive"),
                 levels = c("Alive", "Dead"))

data_proteomic <- data.frame(
  time = time2, status = event2,
  age = rnorm(n2, 55, 12),
  gender = factor(sample(c("M", "F"), n2, replace = TRUE)),
  as.data.frame(prot_matrix)
)

cat("N:", n2, "Events:", sum(event2 == "Dead"),
    "Event rate:", round(mean(event2 == "Dead"), 2), "\n")
```

```{r proteomic-run}
prot_vars <- c("age", "gender", paste0("PROT_", sprintf("%02d", 1:p_prot)))

highdimcox(
  data = data_proteomic,
  elapsedtime = "time",
  outcome = "status",
  outcomeLevel = "Dead",
  predictors = prot_vars,
  regularization_method = "elastic_net",
  alpha_value = 0.5,
  cv_method = "cv_1se",
  cv_folds = 5,
  show_regularization_path = TRUE,
  show_cv_plot = TRUE,
  show_variable_importance = TRUE,
  show_coefficients_table = TRUE,
  showExplanations = TRUE
)
```

---

## 8. Explanatory Output

```{r explanatory}
highdimcox(
  data = data_genomic,
  elapsedtime = "survival_months",
  outcome = "vital_status",
  outcomeLevel = "Dead",
  predictors = pred_vars[1:20],  # Subset for readability
  regularization_method = "elastic_net",
  cv_folds = 10,
  show_coefficients_table = TRUE,
  showSummaries = TRUE,
  showExplanations = TRUE
)
```

---

## Interpreting Results

### Coefficient Table

| Column | Meaning |
|--------|---------|
| Variable | Predictor name |
| Coefficient | Regularized log hazard ratio |
| Hazard Ratio | exp(coefficient) |
| Importance | Absolute coefficient value |

### Variable Importance Plot

Variables are ranked by the absolute value of their regularized coefficients.
The plot shows which predictors have the strongest association with survival
after accounting for regularization.

### Regularization Path

The coefficient path plot shows how each variable's coefficient changes as the
penalty strength (lambda) increases. Variables that persist at higher lambda values
are more robust predictors.

---

## Method Selection Guide

| Your Data | Recommended Method | Alpha |
|-----------|--------------------|-------|
| p >> n, expect few true predictors | LASSO | 1.0 |
| p >> n, many small effects | Ridge | 0.0 |
| p >> n, correlated groups | Elastic Net | 0.3-0.7 |
| p > n, want publishable model | Adaptive LASSO | 1.0 |
| Moderate p, clinical study | Elastic Net | 0.5 |
| Prediction is primary goal | Ridge or Elastic Net | 0.0-0.5 |
| Interpretation is primary goal | LASSO or Adaptive LASSO | 0.8-1.0 |

---

## Common Pitfalls

1. **Using Ridge when you need variable selection**: Ridge never sets coefficients
   to exactly zero. For interpretable models, use LASSO or Elastic Net.

2. **Not adjusting CV folds for small samples**: With n < 100, use cv_folds = 5
   instead of 10 to ensure adequate events per fold.

3. **Ignoring stability selection**: A single regularized model may give unstable
   variable selections. Use stability selection (bootstrap_iterations >= 200)
   for robust results.

4. **Comparing models with different alpha values**: Models with different alpha
   values optimize different objectives. Compare using the same alpha or use
   nested cross-validation.

5. **Not reporting the regularization method**: Always report which method
   (LASSO/Ridge/Elastic Net), alpha value, CV method, and number of selected variables.

---

## Related ClinicoPath Functions

| Function | Use When |
|----------|----------|
| **LASSO Cox** | Pure LASSO with suitability assessment and clinical guidance |
| **Adaptive LASSO** | Oracle property weights for improved selection |
| **SCAD Cox** | Non-convex penalties avoiding LASSO bias |
| **PLS Cox** | Dimensionality reduction approach for very high p |
| **Multivariable Survival** | Standard Cox when p is small |

---

## References

- Zou H, Hastie T. Regularization and variable selection via the elastic net. *J R Stat Soc Ser B*. 2005;67(2):301-320.
- Simon N, et al. Regularization paths for Cox's proportional hazards model via coordinate descent. *J Stat Softw*. 2011;39(5):1-13.
- Meinshausen N, Buhlmann P. Stability selection. *J R Stat Soc Ser B*. 2010;72(4):417-473.
- Tibshirani R. Regression shrinkage and selection via the lasso. *J R Stat Soc Ser B*. 1996;58(1):267-288.
