---
title: "Leave-One-Center-Out Cross-Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leave-One-Center-Out Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    eval = FALSE,
    echo = TRUE,
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

The `leaveonecenterout` function performs **leave-one-center-out cross-validation (LOOCV)** for internal-external validation of multi-institutional prediction models. It iteratively holds out each center as a test set, trains on the remaining centers, and evaluates predictive performance.

This implements the validation framework recommended by **TRIPOD guidelines** (Debray et al., BMJ 2015).

**Key features:**

- Three model types: **logistic**, **Cox**, and **linear** regression
- Optional **LASSO regularization** with lambda selection (minimum CV error or 1SE rule)
- Per-center and pooled performance metrics (AUC, C-index, R-squared)
- Forest plot visualization of center-specific performance
- Clinical notices for data quality, EPV, discrimination, and heterogeneity

## Test Data

Two bundled datasets are available:

```{r data}
library(ClinicoPath)
data(loocv_multicenter)  # N=200, 5 centers
data(loocv_small)        # N=45, 3 centers
```

## 1. Logistic Regression (Binary Outcome)

```{r logistic}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "treatment_response",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score", "tumor_size_mm", "grade"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "logistic"
)
```

**Outputs:** Study design summary, per-center AUC with 95% CI, pooled AUC, forest plot, interpretation with heterogeneity assessment.

## 2. Cox Regression (Survival Outcome)

```{r cox}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "os_status",
    outcomeLevel = "Dead",
    predictors = c("age", "ki67_score", "grade"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "cox"
)
```

**Metric:** C-index (concordance) with 95% CI per center, clamped to [0, 1].

## 3. Linear Regression (Continuous Outcome)

```{r linear}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "tumor_shrinkage",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score", "tumor_size_mm"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "linear"
)
```

**Metric:** Out-of-sample R-squared per center (can be negative when model predicts worse than the mean). RMSE also computed.

## 4. LASSO Regularization

LASSO (L1 penalty) performs automatic variable selection within each CV fold. Recommended when the number of predictors is large relative to the number of events.

### Lambda selection: 1SE rule (more parsimonious)

```{r lasso_1se}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "treatment_response",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score", "tumor_size_mm", "ctdna_level",
                   "grade", "stage", "gender"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "logistic",
    useLasso = TRUE,
    lambdaMethod = "lambda.1se"
)
```

### Lambda selection: Minimum CV error (less shrinkage)

```{r lasso_min}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "treatment_response",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score", "tumor_size_mm", "ctdna_level",
                   "grade", "stage", "gender"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "logistic",
    useLasso = TRUE,
    lambdaMethod = "lambda.min"
)
```

### LASSO Cox

```{r lasso_cox}
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "os_status",
    outcomeLevel = "Dead",
    predictors = c("age", "ki67_score", "tumor_size_mm", "grade"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    modelType = "cox",
    useLasso = TRUE
)
```

> **Note:** LASSO is not available for linear regression. If `useLasso=TRUE` with `modelType="linear"`, a warning notice is displayed and standard OLS is used.

## 5. Output Options

### Toggle pooled performance and forest plot

```{r options}
# Minimal output: per-center results only
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "treatment_response",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    pooledPerformance = FALSE,
    forestPlot = FALSE
)

# With calibration (Brier score)
leaveonecenterout(
    data = loocv_multicenter,
    outcome = "treatment_response",
    outcomeLevel = "Responder",
    predictors = c("age", "ki67_score", "grade"),
    centerVariable = "institution",
    elapsedtime = "os_time",
    calibrationCheck = TRUE
)
```

## 6. Small Dataset (Edge Cases)

```{r small}
data(loocv_small)

leaveonecenterout(
    data = loocv_small,
    outcome = "diagnosis",
    outcomeLevel = "Positive",
    predictors = c("age", "marker1", "marker2"),
    centerVariable = "center",
    elapsedtime = "surv_time",
    modelType = "logistic"
)
```

With 3 centers and small sample sizes, expect:

- Warning notice for centers with < 5 cases
- Wider confidence intervals
- Some centers may be skipped if no event variation

## Interpretation Guide

### Discrimination Assessment

| Metric Range | Logistic/Cox | Linear |
|---|---|---|
| >= 0.80 (AUC/C-index) / >= 0.50 (R-squared) | Good | Good |
| 0.70-0.80 / 0.20-0.50 | Acceptable | Moderate |
| < 0.70 / < 0.20 | Poor | Poor |
| -- / < 0.00 | -- | Very Poor |

### Heterogeneity

| SD Range | Interpretation |
|---|---|
| <= 0.05 | Low heterogeneity -- consistent performance |
| 0.05-0.10 | Moderate -- some variation across centers |
| > 0.10 | High -- consider center-level covariates or stratification |

### Notice Types

- **Warning** (blue): Data quality concerns (small centers, LASSO+linear)
- **Strong Warning** (yellow): Results may be unreliable (low EPV, poor discrimination)
- **Info** (green): Confirmations (event level, completion summary)

## References

- Debray TPA et al. A new framework to enhance the interpretation of external validation studies of clinical prediction models. *J Clin Epidemiol* 2015;68:279-289.
- Riley RD et al. Minimum sample size for developing a multivariable prediction model. *Stat Med* 2019;38:1262-1275.
- Collins GS et al. Transparent Reporting of a multivariable prediction model for Individual Prognosis Or Diagnosis (TRIPOD). *Ann Intern Med* 2015;162:55-63.
