---
title: "Clinical ROC Analysis (enhancedROC) - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical ROC Analysis (enhancedROC) - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Load the package
devtools::load_all("../")

# Base path for data files
data_path <- "../data-raw/non-rda/"
data_path2 <- "../data/"
```

# Clinical ROC Analysis (enhancedROC)

## Overview

The **enhancedROC** module provides comprehensive Receiver Operating Characteristic
(ROC) curve analysis for clinical diagnostic performance evaluation. It includes
ROC curve analysis, Youden Index optimization, sensitivity/specificity analysis,
optimal cutoff determination, comparative ROC analysis, calibration assessment,
multi-class ROC, clinical impact metrics, and class imbalance detection.

All examples below use the bundled test datasets. In jamovi, open the corresponding
`.omv` file. From R, read the `.csv` directly.

---

## Datasets Used in This Guide

| Dataset | N | Outcome | Prevalence | Key Features | Primary Use |
|---------|---|---------|------------|--------------|-------------|
| `enhancedroc_biomarker` | 300 | Disease/Healthy | 30% | 3 biomarkers + clinical risk score | Single biomarker validation |
| `enhancedroc_comparative` | 400 | Cancer/No Cancer | 25% | 5 markers (varying AUCs) | Compare diagnostic tests |
| `enhancedroc_imbalanced` | 500 | Positive/Negative | 5% | Rare disease, screening/confirmatory markers | Imbalanced data handling |
| `enhancedroc_multiclass` | 350 | 4 severity levels | Mixed | 3 biomarkers | Multi-class severity grading |
| `enhancedroc_calibration` | 300 | Event/No Event | 35% | Predicted probs + risk score | Calibration assessment |
| `enhancedroc_screening` | 600 | Screen Pos/Neg | 8% | High-sensitivity markers | Screening context |
| `enhancedroc_confirmatory` | 250 | Confirmed/Not | 45% | High-specificity markers | Confirmatory testing |
| `enhancedroc_small` | 60 | Positive/Negative | 30% | Single simple marker | Small sample testing |
| `enhancedroc_validation` | 350 | Positive/Negative | 30% | Probability predictors, biomarker | CROC / convex hull |
| `enhancedroc_tiedscores` | 200 | Disease/No Disease | 35% | Ordinal scores (many ties) | Tied score handling |

---

## 1. Single Biomarker Analysis (Default)

The default analysis computes a single ROC curve with AUC, optimal cutoff via
Youden Index, and diagnostic performance metrics.

### Basic single ROC

```{r single-basic}
biomarker_data <- read.csv(paste0(data_path, "enhancedroc_biomarker.csv"))

enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  customCutoffs = NULL
)
```

### With cutoff table and clinical metrics

```{r single-cutoff-clinical}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  cutoffTable = TRUE,
  clinicalMetrics = TRUE,
  prevalence = 0.30,
  useObservedPrevalence = TRUE,
  customCutoffs = NULL
)
```

### Direction control and custom cutoffs

```{r single-direction-custom}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  direction = "higher",
  customCutoffs = "5.0,7.5,10.0"
)
```

---

## 2. Comparative ROC Analysis (Multiple Predictors)

Compare diagnostic performance across multiple biomarkers using DeLong's test.

### Five-marker comparison with DeLong

```{r comparative-delong}
comparative_data <- read.csv(paste0(data_path, "enhancedroc_comparative.csv"))

enhancedROC(
  data = comparative_data,
  outcome = "cancer_status",
  positiveClass = "Cancer",
  predictors = vars(established_marker, novel_marker1, novel_marker2,
                    imaging_score, genetic_risk_score),
  analysisType = "comparative",
  pairwiseComparisons = TRUE,
  comparisonMethod = "delong",
  customCutoffs = NULL
)
```

### Bootstrap comparison with metrics differences

```{r comparative-bootstrap}
enhancedROC(
  data = comparative_data,
  outcome = "cancer_status",
  positiveClass = "Cancer",
  predictors = vars(established_marker, novel_marker1),
  analysisType = "comparative",
  comparisonMethod = "bootstrap",
  bootstrapSamples = 200,
  showMetricsDiff = TRUE,
  statisticalComparison = TRUE,
  customCutoffs = NULL
)
```

### Comprehensive analysis with clinical interpretation

```{r comparative-comprehensive}
enhancedROC(
  data = comparative_data,
  outcome = "cancer_status",
  positiveClass = "Cancer",
  predictors = vars(established_marker, novel_marker1, novel_marker2,
                    imaging_score, genetic_risk_score),
  analysisType = "comprehensive",
  comprehensive_output = TRUE,
  clinical_interpretation = TRUE,
  customCutoffs = NULL
)
```

---

## 3. Bootstrap & Confidence Intervals

### BCa bootstrap confidence intervals

```{r bootstrap-bca}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  useBootstrap = TRUE,
  bootstrapSamples = 200,
  bootstrapMethod = "bca",
  customCutoffs = NULL
)
```

### Percentile bootstrap with stratification

```{r bootstrap-percentile-stratified}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  useBootstrap = TRUE,
  bootstrapSamples = 200,
  bootstrapMethod = "percentile",
  stratifiedBootstrap = TRUE,
  customCutoffs = NULL
)
```

### 90% CI with bootstrap CIs for cutoff and partial AUC

```{r bootstrap-90ci}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  confidenceLevel = 90,
  useBootstrap = TRUE,
  bootstrapSamples = 200,
  bootstrapCutoffCI = TRUE,
  bootstrapPartialAUC = TRUE,
  partialAuc = TRUE,
  customCutoffs = NULL
)
```

---

## 4. ROC Curve Plot Options

### Clinical theme with cutoff points and confidence bands

```{r plot-clinical}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  rocCurve = TRUE,
  showCutoffPoints = TRUE,
  showConfidenceBands = TRUE,
  plotTheme = "clinical",
  customCutoffs = NULL
)
```

### Classic theme with custom dimensions

```{r plot-classic}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  rocCurve = TRUE,
  plotTheme = "classic",
  plotWidth = 800,
  plotHeight = 800,
  customCutoffs = NULL
)
```

### Comparative ROC overlay plot

```{r plot-comparative}
enhancedROC(
  data = comparative_data,
  outcome = "cancer_status",
  positiveClass = "Cancer",
  predictors = vars(established_marker, novel_marker1),
  analysisType = "comparative",
  rocCurve = TRUE,
  customCutoffs = NULL
)
```

---

## 5. Youden Index & Cutoff Analysis

### Youden Index plot

```{r youden-plot}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  youdenOptimization = TRUE,
  customCutoffs = NULL
)
```

### Cutoff analysis plot

```{r cutoff-plot}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  cutoffTable = TRUE,
  customCutoffs = NULL
)
```

### Clinical decision plot

```{r clinical-decision-plot}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  clinicalMetrics = TRUE,
  customCutoffs = NULL
)
```

---

## 6. Advanced ROC Methods

### Binormal smoothing

```{r smooth-binormal}
validation_data <- read.csv(paste0(data_path2, "enhancedroc_validation.csv"))

enhancedROC(
  data = validation_data,
  outcome = "outcome",
  positiveClass = "Positive",
  predictors = "biomarker",
  smoothMethod = "binormal",
  customCutoffs = NULL
)
```

### Partial AUC (high specificity region)

```{r partial-auc-spec}
enhancedROC(
  data = validation_data,
  outcome = "outcome",
  positiveClass = "Positive",
  predictors = "biomarker",
  partialAuc = TRUE,
  partialAucType = "specificity",
  partialRange = "0.8,1.0",
  customCutoffs = NULL
)
```

### Partial AUC (high sensitivity region)

```{r partial-auc-sens}
enhancedROC(
  data = validation_data,
  outcome = "outcome",
  positiveClass = "Positive",
  predictors = "biomarker",
  partialAuc = TRUE,
  partialAucType = "sensitivity",
  partialRange = "0.9,1.0",
  customCutoffs = NULL
)
```

### CROC analysis (Concentrated ROC)

```{r croc-analysis}
enhancedROC(
  data = validation_data,
  outcome = "outcome",
  positiveClass = "Positive",
  predictors = "biomarker",
  crocAnalysis = TRUE,
  crocAlpha = 7.0,
  customCutoffs = NULL
)
```

### Convex hull analysis

```{r convex-hull}
enhancedROC(
  data = validation_data,
  outcome = "outcome",
  positiveClass = "Positive",
  predictors = "biomarker",
  convexHull = TRUE,
  customCutoffs = NULL
)
```

### Tied score handling

```{r tied-scores}
tied_data <- read.csv(paste0(data_path2, "enhancedroc_tiedscores.csv"))

enhancedROC(
  data = tied_data,
  outcome = "disease",
  positiveClass = "Disease",
  predictors = vars(ordinal_score, rounded_lab, composite_score),
  tiedScoreHandling = "average",
  customCutoffs = NULL
)
```

---

## 7. Sensitivity & Specificity Thresholds

### Screening context: high sensitivity threshold

```{r threshold-screening}
screening_data <- read.csv(paste0(data_path, "enhancedroc_screening.csv"))

enhancedROC(
  data = screening_data,
  outcome = "screening_indication",
  positiveClass = "Screen Positive",
  predictors = "sensitive_marker",
  sensitivityThreshold = 0.95,
  specificityThreshold = 0.5,
  customCutoffs = NULL
)
```

### Confirmatory context: high specificity threshold

```{r threshold-confirmatory}
confirmatory_data <- read.csv(paste0(data_path, "enhancedroc_confirmatory.csv"))

enhancedROC(
  data = confirmatory_data,
  outcome = "confirmed_diagnosis",
  positiveClass = "Confirmed",
  predictors = "specific_marker",
  sensitivityThreshold = 0.5,
  specificityThreshold = 0.95,
  customCutoffs = NULL
)
```

---

## 8. Class Imbalance Detection

### Imbalance detection with PRC recommendation

```{r imbalance-detect}
imbalanced_data <- read.csv(paste0(data_path, "enhancedroc_imbalanced.csv"))

enhancedROC(
  data = imbalanced_data,
  outcome = "rare_disease",
  positiveClass = "Positive",
  predictors = vars(screening_marker, confirmatory_marker),
  detectImbalance = TRUE,
  imbalanceThreshold = 3.0,
  showImbalanceWarning = TRUE,
  recommendPRC = TRUE,
  customCutoffs = NULL
)
```

### Lower imbalance threshold (more sensitive detection)

```{r imbalance-sensitive}
enhancedROC(
  data = imbalanced_data,
  outcome = "rare_disease",
  positiveClass = "Positive",
  predictors = "screening_marker",
  detectImbalance = TRUE,
  imbalanceThreshold = 1.5,
  customCutoffs = NULL
)
```

---

## 9. Clinical Context & Presets

### Biomarker screening preset

```{r preset-screening}
enhancedROC(
  data = screening_data,
  outcome = "screening_indication",
  positiveClass = "Screen Positive",
  predictors = "sensitive_marker",
  clinicalContext = "screening",
  clinicalPresets = "biomarker_screening",
  customCutoffs = NULL
)
```

### Confirmatory testing preset

```{r preset-confirmatory}
enhancedROC(
  data = confirmatory_data,
  outcome = "confirmed_diagnosis",
  positiveClass = "Confirmed",
  predictors = "specific_marker",
  clinicalContext = "diagnosis",
  clinicalPresets = "confirmatory_testing",
  customCutoffs = NULL
)
```

### Diagnostic validation preset (balanced)

```{r preset-diagnostic}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  clinicalContext = "prognosis",
  clinicalPresets = "diagnostic_validation",
  customCutoffs = NULL
)
```

### Research comprehensive preset

```{r preset-research}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = vars(biomarker1, biomarker2, clinical_risk_score),
  clinicalPresets = "research_comprehensive",
  customCutoffs = NULL
)
```

### Monitoring context with custom settings

```{r preset-monitoring}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  clinicalContext = "monitoring",
  clinicalPresets = "custom",
  customCutoffs = NULL
)
```

---

## 10. Calibration Analysis

### Full calibration assessment

```{r calibration-full}
calibration_data <- read.csv(paste0(data_path, "enhancedroc_calibration.csv"))

enhancedROC(
  data = calibration_data,
  outcome = "outcome",
  positiveClass = "Event",
  predictors = vars(predicted_prob, risk_score),
  calibrationAnalysis = TRUE,
  calibrationPlot = TRUE,
  brierScore = TRUE,
  calibrationMetrics = TRUE,
  customCutoffs = NULL
)
```

### Hosmer-Lemeshow test

```{r calibration-hl}
enhancedROC(
  data = calibration_data,
  outcome = "outcome",
  positiveClass = "Event",
  predictors = "predicted_prob",
  hosmerLemeshow = TRUE,
  hlGroups = 10,
  customCutoffs = NULL
)
```

### Hosmer-Lemeshow with fewer groups

```{r calibration-hl-5}
enhancedROC(
  data = calibration_data,
  outcome = "outcome",
  positiveClass = "Event",
  predictors = "predicted_prob",
  hosmerLemeshow = TRUE,
  hlGroups = 5,
  customCutoffs = NULL
)
```

---

## 11. Multi-Class ROC

### One-vs-Rest with macro averaging

```{r multiclass-ovr}
multiclass_data <- read.csv(paste0(data_path, "enhancedroc_multiclass.csv"))

enhancedROC(
  data = multiclass_data,
  outcome = "disease_severity",
  positiveClass = "Severe",
  predictors = vars(biomarker_A, biomarker_B, imaging_severity_score),
  multiClassROC = TRUE,
  multiClassStrategy = "ovr",
  multiClassAveraging = "macro",
  customCutoffs = NULL
)
```

### One-vs-One with weighted averaging

```{r multiclass-ovo}
enhancedROC(
  data = multiclass_data,
  outcome = "disease_severity",
  positiveClass = "Severe",
  predictors = "biomarker_A",
  multiClassROC = TRUE,
  multiClassStrategy = "ovo",
  multiClassAveraging = "weighted",
  customCutoffs = NULL
)
```

---

## 12. Clinical Impact & Utility

### Clinical impact with NNT calculation

```{r clinical-impact}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  clinicalImpact = TRUE,
  nntCalculation = TRUE,
  decisionImpactTable = TRUE,
  customCutoffs = NULL
)
```

### Clinical utility curve

```{r clinical-utility}
enhancedROC(
  data = biomarker_data,
  outcome = "disease_status",
  positiveClass = "Disease",
  predictors = "biomarker1",
  clinicalUtilityCurve = TRUE,
  customCutoffs = NULL
)
```

---

## 13. Edge Cases

### Small sample (n=60)

```{r edge-small}
small_data <- read.csv(paste0(data_path, "enhancedroc_small.csv"))

enhancedROC(
  data = small_data,
  outcome = "disease",
  positiveClass = "Positive",
  predictors = "marker",
  useBootstrap = TRUE,
  bootstrapSamples = 200,
  customCutoffs = NULL
)
```

### Low prevalence screening data

```{r edge-low-prevalence}
enhancedROC(
  data = screening_data,
  outcome = "screening_indication",
  positiveClass = "Screen Positive",
  predictors = "sensitive_marker",
  customCutoffs = NULL
)
```

### Tied scores

```{r edge-tied}
enhancedROC(
  data = tied_data,
  outcome = "disease",
  positiveClass = "Disease",
  predictors = "ordinal_score",
  customCutoffs = NULL
)
```

---

## References

- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas under
  two or more correlated receiver operating characteristic curves: a nonparametric
  approach. *Biometrics*, 44(3), 837-845.
- Youden WJ (1950). Index for rating diagnostic tests. *Cancer*, 3(1), 32-35.
- Hanley JA, McNeil BJ (1982). The meaning and use of the area under a receiver
  operating characteristic (ROC) curve. *Radiology*, 143(1), 29-36.
- Robin X, Turck N, Hainard A, et al. (2011). pROC: an open-source package for R
  and S+ to analyze and compare ROC curves. *BMC Bioinformatics*, 12, 77.
- Hosmer DW, Lemeshow S (2000). *Applied Logistic Regression*. 2nd ed. Wiley.
- Brier GW (1950). Verification of forecasts expressed in terms of probability.
  *Monthly Weather Review*, 78(1), 1-3.
