---
title: "Use meddecide in R"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Use meddecide in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```


agreement()
	

Interrater Reliability

agreementClass
	

Interrater Reliability Analysis

decision()
	

Medical Decision

decisionClass
	

Medical Decision Making

decisioncalculator()
	

Medical Decision Calculator

decisioncalculatorClass
	

Decision Calculator

histopathology
	

histopathology

kappaSizeCI()
	

Confidence Interval Approach for the Number of Subjects Required

kappaSizeCIClass
	

Confidence Interval Approach for the Number of Subjects Required

kappaSizeFixedN()
	

Lowest Expected Value for a fixed sample size

kappaSizeFixedNClass
	

Lowest Expected Value for a fixed sample size

kappaSizePower()
	

Power Approach for the Number of Subjects Required

kappaSizePowerClass 

