---
title: "Decision Tree and Markov Chain Analysis Examples"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
    echo = TRUE,
  eval = FALSE
)

library(dplyr)
library(ggplot2)
```

# Overview

This vignette provides comprehensive examples explaining both decision tree and Markov chain analysis types implemented in the ClinicoPath jamovi module.

# Decision Tree Analysis Example

## Scenario: Acute Appendicitis Treatment Decision

**Clinical Question:** Should a patient with suspected appendicitis receive immediate surgery or conservative treatment?

### Creating Example Data

```{r decision-tree-data}
# Create decision tree example data
set.seed(123)
n_patients <- 100

appendicitis_decision_data <- data.frame(
  patient_id = 1:n_patients,

  # DECISION NODES (Square boxes)
  treatment_choice = sample(c("Immediate Surgery", "Conservative Treatment"),
                           n_patients, replace = TRUE),

  # PROBABILITY VARIABLES (for chance nodes - circles)
  prob_appendicitis_confirmed = runif(n_patients, 0.7, 0.9),     # Probability patient actually has appendicitis
  prob_surgery_success = runif(n_patients, 0.95, 0.99),          # Surgery success rate
  prob_conservative_success = runif(n_patients, 0.6, 0.8),       # Conservative treatment success
  prob_complications_surgery = runif(n_patients, 0.02, 0.08),    # Surgery complications
  prob_complications_conservative = runif(n_patients, 0.15, 0.25), # Conservative complications

  # COST VARIABLES (outcomes)
  cost_surgery = rnorm(n_patients, 12000, 2000),                 # Surgery costs
  cost_conservative = rnorm(n_patients, 3000, 500),              # Conservative treatment costs
  cost_complications = rnorm(n_patients, 8000, 1500),            # Complication management costs
  cost_failed_conservative = rnorm(n_patients, 15000, 2500),     # Emergency surgery after failed conservative

  # UTILITY VARIABLES (quality of life outcomes)
  utility_success = runif(n_patients, 0.95, 1.0),               # Full recovery
  utility_minor_complications = runif(n_patients, 0.8, 0.9),    # Recovery with minor issues
  utility_major_complications = runif(n_patients, 0.6, 0.8),    # Recovery with major issues

  # OUTCOME VARIABLES (terminal nodes - triangles)
  clinical_outcome = sample(c("Complete Recovery", "Minor Complications", "Major Complications"),
                           n_patients, replace = TRUE, prob = c(0.8, 0.15, 0.05))
)

# Display first few rows
head(appendicitis_decision_data, 5)
```

### Decision Tree Structure

The decision tree has the following structure:

1. **DECISION NODE (Square):** Treatment Choice
   - Option A: Immediate Surgery
   - Option B: Conservative Treatment

2. **CHANCE NODES (Circles):** Probabilistic Outcomes
   - For Surgery:
     - Success (95-99%): Low cost, high utility
     - Complications (2-8%): Higher cost, lower utility
   - For Conservative:
     - Success (60-80%): Low cost, high utility
     - Failure (20-40%): Requires emergency surgery

3. **TERMINAL NODES (Triangles):** Final Outcomes
   - Each path ends with:
     - Cost: $3,000 - $20,000
     - Utility: 0.6 - 1.0 QALYs

### Expected Value Calculations

```{r decision-tree-calculations}
# Calculate expected values for decision tree
surgery_expected_cost <- mean(appendicitis_decision_data$cost_surgery +
                             appendicitis_decision_data$prob_complications_surgery *
                             appendicitis_decision_data$cost_complications)

conservative_expected_cost <- mean(appendicitis_decision_data$cost_conservative +
                                  (1 - appendicitis_decision_data$prob_conservative_success) *
                                  appendicitis_decision_data$cost_failed_conservative)

surgery_expected_utility <- mean(appendicitis_decision_data$utility_success *
                                appendicitis_decision_data$prob_surgery_success +
                                appendicitis_decision_data$utility_minor_complications *
                                appendicitis_decision_data$prob_complications_surgery)

conservative_expected_utility <- mean(appendicitis_decision_data$utility_success *
                                     appendicitis_decision_data$prob_conservative_success +
                                     appendicitis_decision_data$utility_major_complications *
                                     (1 - appendicitis_decision_data$prob_conservative_success))

# Calculate ICER
incremental_cost <- surgery_expected_cost - conservative_expected_cost
incremental_utility <- surgery_expected_utility - conservative_expected_utility
icer <- incremental_cost / incremental_utility

# Display results
cat("DECISION TREE ANALYSIS RESULTS:\n")
cat("===============================\n")
cat("Surgery Strategy:\n")
cat("  Expected Cost: $", round(surgery_expected_cost, 0), "\n")
cat("  Expected Utility:", round(surgery_expected_utility, 3), "QALYs\n")
cat("\n")
cat("Conservative Strategy:\n")
cat("  Expected Cost: $", round(conservative_expected_cost, 0), "\n")
cat("  Expected Utility:", round(conservative_expected_utility, 3), "QALYs\n")
cat("\n")
cat("Cost-Effectiveness Analysis:\n")
cat("  Incremental Cost: $", round(incremental_cost, 0), "\n")
cat("  Incremental Utility:", round(incremental_utility, 3), "QALYs\n")
cat("  ICER: $", round(icer, 0), "per QALY\n")

if (icer < 50000) {
  cat("  ✓ Surgery is cost-effective (ICER < $50,000/QALY)\n")
} else {
  cat("  ⚠ Surgery may not be cost-effective (ICER > $50,000/QALY)\n")
}
```

### Decision Tree Interpretation

Decision trees are ideal for:

- One-time decisions with immediate outcomes
- Comparing 2-3 distinct treatment strategies
- Situations where timing is not critical
- When outcomes occur relatively quickly (days to months)
- Point-in-time cost-effectiveness analysis

# Markov Chain Analysis Example

## Scenario: Chronic Heart Disease Management

**Clinical Question:** What is the long-term cost-effectiveness of different heart disease management strategies?

### Creating Markov Data

```{r markov-data}
# Create Markov chain example data
set.seed(456)
n_strategies <- 150

# First create basic structure
heart_disease_markov_data <- data.frame(
  strategy_id = 1:n_strategies,

  # DECISION VARIABLES
  management_strategy = sample(c("Standard Care", "Intensive Monitoring", "Preventive Surgery"),
                              n_strategies, replace = TRUE),
  patient_risk_category = sample(c("Low Risk", "Moderate Risk", "High Risk"),
                                n_strategies, replace = TRUE, prob = c(0.4, 0.4, 0.2))
)

# Add transition probabilities
heart_disease_markov_data$prob_asymp_to_symp <- case_when(
  heart_disease_markov_data$management_strategy == "Standard Care" ~ runif(n_strategies, 0.08, 0.12),
  heart_disease_markov_data$management_strategy == "Intensive Monitoring" ~ runif(n_strategies, 0.05, 0.08),
  heart_disease_markov_data$management_strategy == "Preventive Surgery" ~ runif(n_strategies, 0.02, 0.05)
)

heart_disease_markov_data$prob_asymp_to_death <- runif(n_strategies, 0.01, 0.02)

# From Symptomatic state
heart_disease_markov_data$prob_symp_to_hf <- case_when(
  heart_disease_markov_data$management_strategy == "Standard Care" ~ runif(n_strategies, 0.15, 0.25),
  heart_disease_markov_data$management_strategy == "Intensive Monitoring" ~ runif(n_strategies, 0.10, 0.18),
  heart_disease_markov_data$management_strategy == "Preventive Surgery" ~ runif(n_strategies, 0.05, 0.12)
)

heart_disease_markov_data$prob_symp_to_death <- runif(n_strategies, 0.02, 0.04)

# From Heart Failure state
heart_disease_markov_data$prob_hf_to_death <- runif(n_strategies, 0.12, 0.20)

# STATE-SPECIFIC ANNUAL COSTS
heart_disease_markov_data$cost_asymptomatic <- case_when(
  heart_disease_markov_data$management_strategy == "Standard Care" ~ rnorm(n_strategies, 2000, 300),
  heart_disease_markov_data$management_strategy == "Intensive Monitoring" ~ rnorm(n_strategies, 4000, 500),
  heart_disease_markov_data$management_strategy == "Preventive Surgery" ~ rnorm(n_strategies, 8000, 1000)
)

heart_disease_markov_data$cost_symptomatic <- rnorm(n_strategies, 12000, 2000)
heart_disease_markov_data$cost_heart_failure <- rnorm(n_strategies, 35000, 5000)
heart_disease_markov_data$cost_death <- rep(0, n_strategies)  # No ongoing costs after death

# STATE-SPECIFIC ANNUAL UTILITIES (Quality of Life)
heart_disease_markov_data$utility_asymptomatic <- runif(n_strategies, 0.90, 0.95)
heart_disease_markov_data$utility_symptomatic <- runif(n_strategies, 0.70, 0.80)
heart_disease_markov_data$utility_heart_failure <- runif(n_strategies, 0.45, 0.60)
heart_disease_markov_data$utility_death <- rep(0, n_strategies)  # No utility after death

# Display first few rows
head(heart_disease_markov_data, 5)
```

### Markov Chain Structure

**HEALTH STATES (Markov States):**

1. Asymptomatic Heart Disease
2. Symptomatic Heart Disease  
3. Heart Failure
4. Death (Absorbing State)

**TRANSITION PATHWAYS:**

```
Asymptomatic → Symptomatic → Heart Failure → Death
             ↘ Death        ↘ Death
```

### Transition Matrix Example

```{r transition-matrix}
# Create example transition matrix for Standard Care
states <- c("Asymptomatic", "Symptomatic", "Heart Failure", "Death")
trans_matrix_standard <- matrix(0, nrow = 4, ncol = 4)
rownames(trans_matrix_standard) <- states
colnames(trans_matrix_standard) <- states

# Fill transition matrix with average probabilities for Standard Care
standard_data <- heart_disease_markov_data[heart_disease_markov_data$management_strategy == "Standard Care", ]

trans_matrix_standard[1, 1] <- 1 - mean(standard_data$prob_asymp_to_symp) - mean(standard_data$prob_asymp_to_death)  # Stay asymptomatic
trans_matrix_standard[1, 2] <- mean(standard_data$prob_asymp_to_symp)  # Asymp to symptomatic
trans_matrix_standard[1, 4] <- mean(standard_data$prob_asymp_to_death)  # Asymp to death

trans_matrix_standard[2, 2] <- 1 - mean(standard_data$prob_symp_to_hf) - mean(standard_data$prob_symp_to_death)  # Stay symptomatic
trans_matrix_standard[2, 3] <- mean(standard_data$prob_symp_to_hf)  # Symp to heart failure
trans_matrix_standard[2, 4] <- mean(standard_data$prob_symp_to_death)  # Symp to death

trans_matrix_standard[3, 3] <- 1 - mean(standard_data$prob_hf_to_death)  # Stay in heart failure
trans_matrix_standard[3, 4] <- mean(standard_data$prob_hf_to_death)  # HF to death

trans_matrix_standard[4, 4] <- 1.0  # Death is absorbing

cat("EXAMPLE TRANSITION MATRIX (Standard Care):\n")
cat("=========================================\n")
print(round(trans_matrix_standard, 3))
cat("\nRow sums (should equal 1.0):", round(rowSums(trans_matrix_standard), 3), "\n")
```

### Markov Cohort Simulation

```{r markov-simulation}
# Run Markov cohort simulation
num_cycles <- 20
cohort_trace <- matrix(0, nrow = num_cycles + 1, ncol = 4)
colnames(cohort_trace) <- states

# Initial distribution: Everyone starts asymptomatic
cohort_trace[1, 1] <- 1.0

# Run Markov simulation
for (cycle in 2:(num_cycles + 1)) {
  cohort_trace[cycle, ] <- cohort_trace[cycle - 1, ] %*% trans_matrix_standard
}

# Create summary table
trace_df <- data.frame(
  Year = 0:num_cycles,
  Asymptomatic = round(cohort_trace[, 1] * 100, 1),
  Symptomatic = round(cohort_trace[, 2] * 100, 1),
  Heart_Failure = round(cohort_trace[, 3] * 100, 1),
  Death = round(cohort_trace[, 4] * 100, 1)
)

# Show key time points
key_years <- c(1, 6, 11, 16, 21)  # 0, 5, 10, 15, 20 years
cat("Population distribution over time:\n")
print(trace_df[key_years, ])
```

### Cost-Effectiveness Calculation

```{r cost-effectiveness}
# Calculate costs and utilities for Standard Care
state_costs <- c(mean(standard_data$cost_asymptomatic),
                mean(standard_data$cost_symptomatic),
                mean(standard_data$cost_heart_failure),
                0)  # Death

state_utilities <- c(mean(standard_data$utility_asymptomatic),
                    mean(standard_data$utility_symptomatic),
                    mean(standard_data$utility_heart_failure),
                    0)  # Death

names(state_costs) <- states
names(state_utilities) <- states

cat("Annual costs per state:\n")
print(round(state_costs, 0))
cat("\nAnnual utilities per state:\n")
print(round(state_utilities, 3))

# Calculate cumulative discounted costs and utilities
discount_rate <- 0.03
cumulative_costs <- rep(0, num_cycles + 1)
cumulative_utilities <- rep(0, num_cycles + 1)

for (cycle in 2:(num_cycles + 1)) {
  # Calculate cycle costs and utilities
  cycle_cost <- sum(cohort_trace[cycle, ] * state_costs)
  cycle_utility <- sum(cohort_trace[cycle, ] * state_utilities)

  # Apply discounting
  discount_factor <- (1 + discount_rate)^(-(cycle - 1))

  cumulative_costs[cycle] <- cumulative_costs[cycle - 1] + cycle_cost * discount_factor
  cumulative_utilities[cycle] <- cumulative_utilities[cycle - 1] + cycle_utility * discount_factor
}

# Final results
final_cost <- cumulative_costs[num_cycles + 1]
final_qalys <- cumulative_utilities[num_cycles + 1]

cat("\nFINAL 20-YEAR RESULTS (Standard Care):\n")
cat("Total Lifetime Cost: $", round(final_cost, 0), "\n")
cat("Total Lifetime QALYs:", round(final_qalys, 2), "\n")
cat("Cost per QALY: $", round(final_cost / final_qalys, 0), "\n")
```

### Markov Chain Interpretation

Markov chains are ideal for:

- Chronic diseases with multiple stages
- Long-term cost-effectiveness analysis (years to lifetime)
- Disease progression modeling
- Comparing interventions with different timing effects
- Policy decisions affecting population health
- When disease states change over time
- Recurring decisions or ongoing treatments

# When to Use Each Method

## Comparison Table

```{r comparison-table}
comparison_table <- data.frame(
  Aspect = c("Time Horizon", "Disease Type", "Decision Complexity", "Outcomes",
             "Costs", "Best For", "Data Requirements", "Computational Needs"),
  Decision_Tree = c("Short-term (days-months)", "Acute conditions", "Simple (2-3 options)",
                   "One-time outcomes", "One-time costs", "Treatment selection",
                   "Probabilities, costs, utilities", "Low"),
  Markov_Chain = c("Long-term (years-lifetime)", "Chronic conditions", "Complex strategies",
                  "Recurring outcomes", "Ongoing costs", "Disease management",
                  "Transition probabilities, state costs", "Higher")
)

print(comparison_table)
```

## Example Applications

**DECISION TREES are best for:**

- Emergency treatment decisions (appendicitis, trauma)
- Surgical vs non-surgical interventions
- Diagnostic test decisions
- Vaccination decisions
- One-time screening decisions

**MARKOV CHAINS are best for:**

- Chronic disease management (diabetes, heart disease)
- Cancer progression and treatment
- Addiction treatment programs
- Preventive intervention policies
- Healthcare resource planning
- Long-term pharmaceutical studies

## Combined Approaches

Some complex problems benefit from both:

1. Decision tree for initial treatment choice
2. Markov chain for long-term disease progression

**Example:** Cancer treatment selection (tree) + survival modeling (Markov)

# Summary

This vignette demonstrates practical applications of both decision tree and Markov chain methods for medical decision analysis and cost-effectiveness research. The generated datasets can be used with the ClinicoPath jamovi module to perform these analyses interactively.

## Generated Datasets

The example code creates two key datasets:

- `appendicitis_decision_data` - Decision tree example
- `heart_disease_markov_data` - Markov chain example

These datasets demonstrate realistic medical scenarios and can be used to practice both analysis types in jamovi.

## Next Steps

- Load these datasets into jamovi
- Use the ClinicoPath decision analysis modules
- Practice interpreting cost-effectiveness results
- Apply these methods to your own research questions
