---
title: "Advanced ROC Analysis (psychopdaROC) - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced ROC Analysis (psychopdaROC) - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Load the package
devtools::load_all("../")

# Base path for data files
data_path <- "../data-raw/non-rda/"
data_path2 <- "../data/"
```

# Advanced ROC Analysis (psychopdaROC)

## Overview

The **psychopdaROC** module provides advanced Receiver Operating Characteristic
(ROC) curve analysis with optimal cutpoint determination using the `cutpointr`
framework. It supports 12 cutpoint optimization methods, 16 optimization metrics,
subgroup analysis, DeLong's test for comparing ROC curves, IDI/NRI reclassification
indices, effect size and power analysis, bootstrap ROC with prior weighting,
clinical utility analysis, meta-analysis of AUC values, and fixed
sensitivity/specificity analysis.

All examples below use the bundled test datasets. In jamovi, open the corresponding
`.omv` file. From R, read the `.csv` directly.

---

## Datasets Used in This Guide

| Dataset | N | Class Levels | Prevalence | Key Features | Primary Use |
|---------|---|-------------|------------|--------------|-------------|
| `psychopdaROC_test` | 200 | Healthy/Disease | ~50% | 1 biomarker, age, sex | Basic single-marker testing |
| `psychopdaROC_screening` | 250 | No Cancer/Cancer | ~25% | PSA, CA125 | Cancer screening |
| `psychopdaROC_cardiac` | 180 | No MI/MI | ~30% | Troponin, creatinine, BNP | Cardiac biomarkers |
| `psychopdaROC_multibiomarker` | 220 | Negative/Positive | ~30% | 3 markers + combined score | Multi-marker comparison |
| `psychopdaROC_subgroup` | 200 | No Disease/Disease | ~30% | test_score, age_group, sex | Subgroup analysis |
| `psychopdaROC_perfect` | 100 | Negative/Positive | 50% | Perfect separation | AUC=1.0 edge case |
| `psychopdaROC_poor` | 150 | Negative/Positive | ~50% | No discrimination | AUC~0.5 edge case |
| `psychopdaROC_overlap` | 190 | Negative/Positive | ~50% | Moderate overlap | AUC~0.70 |
| `psychopdaROC_rare` | 300 | No Disease/Disease | 5% | Single biomarker | Rare disease |
| `psychopdaROC_costbenefit` | 160 | No Event/Event | ~30% | Risk score, cost columns | Cost-benefit optimization |
| `psychopdaROC_spectrum` | 170 | 3 severity + binary | varies | Continuous marker | Disease spectrum |
| `psychopdaROC_timedep` | 140 | No Event/Event | ~30% | Baseline + follow-up | Longitudinal comparison |
| `psychopdaROC_small` | 30 | Negative/Positive | ~50% | Single marker | Small sample edge case |
| `psychopdaROC_imbalanced` | 200 | No Event/Event | 2% | Single predictor | Extreme imbalance |
| `psychopdaROC_missing` | 150 | Negative/Positive | ~30% | 2 tests + covariate, NAs | Missing data handling |
| `psychopdaROC_constant` | 80 | No Event/Event | ~50% | Constant marker (all=50) | Zero variance edge case |
| `psychopdaROC_large` | 500 | No Disease/Disease | ~30% | 2 biomarkers, site | Large sample / subgroups |
| `psychopdaROC_advanced` | 250 | Negative/Positive | ~30% | 5 markers, site | Meta-analysis / advanced |

---

## 1. Basic ROC Analysis (Default Cutpoint)

The default analysis uses Youden Index optimization (`maximize_metric` with `youden`)
to find the optimal cutpoint.

### Default single-marker ROC

```{r basic-default}
test_data <- read.csv(paste0(data_path, "psychopdaROC_test.csv"))

psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  refVar = NULL
)
```

### With confusion matrix and threshold table

```{r basic-tables}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  sensSpecTable = TRUE,
  showThresholdTable = TRUE,
  maxThresholds = 20,
  refVar = NULL
)
```

### Lower values indicate positive class

```{r basic-direction}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  direction = "<=",
  refVar = NULL
)
```

---

## 2. Cutpoint Optimization Methods

### Key optimization metrics

```{r cut-youden}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_metric",
  metric = "youden",
  refVar = NULL
)
```

```{r cut-accuracy}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_metric",
  metric = "accuracy",
  refVar = NULL
)
```

```{r cut-f1}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_metric",
  metric = "F1_score",
  refVar = NULL
)
```

```{r cut-kappa}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_metric",
  metric = "cohens_kappa",
  refVar = NULL
)
```

### Minimize misclassification cost

```{r cut-misclass}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "minimize_metric",
  metric = "misclassification_cost",
  refVar = NULL
)
```

### LOESS-smoothed Youden

```{r cut-loess}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_loess_metric",
  metric = "youden",
  refVar = NULL
)
```

### Bootstrap-optimized cutpoint

```{r cut-boot}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "maximize_boot_metric",
  boot_runs = 200,
  seed = 123,
  refVar = NULL
)
```

### Kernel-smoothed Youden

```{r cut-kernel}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "oc_youden_kernel",
  refVar = NULL
)
```

### Parametric normal Youden

```{r cut-normal}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "oc_youden_normal",
  refVar = NULL
)
```

### Manual cutpoint

```{r cut-manual}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "oc_manual",
  specifyCutScore = "7.5",
  refVar = NULL
)
```

### Cost-benefit optimized cutpoint

```{r cut-cost}
costbenefit_data <- read.csv(paste0(data_path, "psychopdaROC_costbenefit.csv"))

psychopdaROC(
  data = costbenefit_data,
  classVar = "outcome",
  positiveClass = "Event",
  dependentVars = "risk_score",
  method = "oc_cost_ratio",
  costratioFP = 0.1,
  refVar = NULL
)
```

### Equal sensitivity and specificity

```{r cut-equal}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "oc_equal_sens_spec",
  refVar = NULL
)
```

### Closest to perfect classifier (0,1)

```{r cut-closest01}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  method = "oc_closest_01",
  refVar = NULL
)
```

---

## 3. Cutpoint Fine-Tuning

### Metric tolerance

```{r finetune-tol}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  tol_metric = 0.01,
  refVar = NULL
)
```

### Tie-breaking methods

```{r finetune-ties-mean}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  break_ties = "mean",
  refVar = NULL
)
```

```{r finetune-ties-median}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  break_ties = "median",
  refVar = NULL
)
```

### All observed cutpoints

```{r finetune-all-observed}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  allObserved = TRUE,
  refVar = NULL
)
```

---

## 4. Multiple Test Variables (Comparative)

### Multi-marker ROC overlay

```{r multi-overlay}
multi_data <- read.csv(paste0(data_path, "psychopdaROC_multibiomarker.csv"))

psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3, combined_score),
  plotROC = TRUE,
  combinePlots = TRUE,
  refVar = NULL
)
```

### DeLong test for AUC comparison

```{r multi-delong}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  delongTest = TRUE,
  refVar = NULL
)
```

### Cardiac biomarker comparison

```{r multi-cardiac}
cardiac_data <- read.csv(paste0(data_path, "psychopdaROC_cardiac.csv"))

psychopdaROC(
  data = cardiac_data,
  classVar = "mi_status",
  positiveClass = "MI",
  dependentVars = vars(troponin, creatinine, bnp),
  combinePlots = TRUE,
  delongTest = TRUE,
  refVar = NULL
)
```

---

## 5. ROC Visualization Options

### Optimal point and confidence bands

```{r viz-bands}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  plotROC = TRUE,
  showOptimalPoint = TRUE,
  showConfidenceBands = TRUE,
  refVar = NULL
)
```

### Smoothed ROC with standard error

```{r viz-smooth}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  smoothing = TRUE,
  displaySE = TRUE,
  refVar = NULL
)
```

### Publication-ready clean plot

```{r viz-clean}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  cleanPlot = TRUE,
  refVar = NULL
)
```

### Legend positions

```{r viz-legend}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  combinePlots = TRUE,
  legendPosition = "bottom",
  refVar = NULL
)
```

### Direct curve labels

```{r viz-direct-label}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  combinePlots = TRUE,
  directLabel = TRUE,
  refVar = NULL
)
```

---

## 6. Diagnostic Plots

### Sensitivity/Specificity vs Threshold

```{r diag-criterion}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  showCriterionPlot = TRUE,
  refVar = NULL
)
```

### Predictive values vs Prevalence

```{r diag-prevalence}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  showPrevalencePlot = TRUE,
  refVar = NULL
)
```

### Test value distribution by class

```{r diag-dot}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  showDotPlot = TRUE,
  refVar = NULL
)
```

### Precision-Recall curve

```{r diag-prc}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  precisionRecallCurve = TRUE,
  refVar = NULL
)
```

---

## 7. Prevalence & Prior Settings

### Prior prevalence adjustment for rare disease

```{r prev-prior}
rare_data <- read.csv(paste0(data_path, "psychopdaROC_rare.csv"))

psychopdaROC(
  data = rare_data,
  classVar = "rare_disease",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  usePriorPrev = TRUE,
  priorPrev = 0.05,
  refVar = NULL
)
```

### Sample prevalence (default)

```{r prev-sample}
psychopdaROC(
  data = rare_data,
  classVar = "rare_disease",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  usePriorPrev = FALSE,
  refVar = NULL
)
```

---

## 8. Subgroup Analysis

### By age group

```{r subgroup-age}
subgroup_data <- read.csv(paste0(data_path, "psychopdaROC_subgroup.csv"))

psychopdaROC(
  data = subgroup_data,
  classVar = "disease",
  positiveClass = "Disease",
  dependentVars = "test_score",
  subGroup = "age_group",
  refVar = NULL
)
```

### By sex

```{r subgroup-sex}
psychopdaROC(
  data = subgroup_data,
  classVar = "disease",
  positiveClass = "Disease",
  dependentVars = "test_score",
  subGroup = "sex",
  refVar = NULL
)
```

### Large dataset with site subgroups

```{r subgroup-site}
large_data <- read.csv(paste0(data_path, "psychopdaROC_large.csv"))

psychopdaROC(
  data = large_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = vars(biomarker1, biomarker2),
  subGroup = "site",
  refVar = NULL
)
```

---

## 9. Advanced ROC Analysis

### Partial AUC (high specificity region)

```{r adv-partial-auc}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  partialAUC = TRUE,
  partialAUCfrom = 0.8,
  partialAUCto = 1.0,
  refVar = NULL
)
```

### Partial AUC (very high specificity)

```{r adv-partial-auc-high}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  partialAUC = TRUE,
  partialAUCfrom = 0.9,
  partialAUCto = 1.0,
  refVar = NULL
)
```

### Binormal smoothing

```{r adv-smooth-binormal}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  rocSmoothingMethod = "binormal",
  refVar = NULL
)
```

### Bootstrap confidence intervals

```{r adv-bootstrap-ci}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  bootstrapCI = TRUE,
  bootstrapReps = 200,
  refVar = NULL
)
```

### Quantile-based CIs

```{r adv-quantile-ci}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  quantileCIs = TRUE,
  quantiles = "0.1,0.25,0.5,0.75,0.9",
  refVar = NULL
)
```

---

## 10. Fixed Sensitivity / Specificity Analysis

### Fixed sensitivity for screening (95%)

```{r fixed-sens}
screening_data <- read.csv(paste0(data_path, "psychopdaROC_screening.csv"))

psychopdaROC(
  data = screening_data,
  classVar = "cancer",
  positiveClass = "Cancer",
  dependentVars = "psa_level",
  fixedSensSpecAnalysis = TRUE,
  fixedAnalysisType = "sensitivity",
  fixedSensitivityValue = 0.95,
  refVar = NULL
)
```

### Fixed specificity for confirmation (90%)

```{r fixed-spec}
psychopdaROC(
  data = screening_data,
  classVar = "cancer",
  positiveClass = "Cancer",
  dependentVars = "psa_level",
  fixedSensSpecAnalysis = TRUE,
  fixedAnalysisType = "specificity",
  fixedSpecificityValue = 0.90,
  refVar = NULL
)
```

### Interpolation methods

```{r fixed-interp-linear}
psychopdaROC(
  data = screening_data,
  classVar = "cancer",
  positiveClass = "Cancer",
  dependentVars = "psa_level",
  fixedSensSpecAnalysis = TRUE,
  fixedAnalysisType = "sensitivity",
  fixedSensitivityValue = 0.95,
  fixedInterpolation = "linear",
  refVar = NULL
)
```

### Fixed ROC plot with explanation

```{r fixed-roc-plot}
psychopdaROC(
  data = screening_data,
  classVar = "cancer",
  positiveClass = "Cancer",
  dependentVars = "psa_level",
  fixedSensSpecAnalysis = TRUE,
  fixedAnalysisType = "sensitivity",
  fixedSensitivityValue = 0.95,
  showFixedROC = TRUE,
  showFixedExplanation = TRUE,
  refVar = NULL
)
```

---

## 11. Model Comparison (IDI / NRI)

### Integrated Discrimination Improvement (IDI)

```{r idi}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  calculateIDI = TRUE,
  refVar = NULL,
  idiNriBootRuns = 200
)
```

### Category-based Net Reclassification Index (NRI)

```{r nri-category}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  calculateNRI = TRUE,
  refVar = NULL,
  nriThresholds = "0.3,0.7"
)
```

### Continuous NRI (no thresholds)

```{r nri-continuous}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  calculateNRI = TRUE,
  refVar = NULL,
  nriThresholds = ""
)
```

### Comprehensive classifier comparison

```{r compare-classifiers}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3),
  compareClassifiers = TRUE,
  refVar = NULL
)
```

---

## 12. Effect Size & Power Analysis

### Effect size analysis

```{r effect-size}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  effectSizeAnalysis = TRUE,
  refVar = NULL
)
```

### Post-hoc power analysis

```{r power-posthoc}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  powerAnalysis = TRUE,
  powerAnalysisType = "post_hoc",
  significanceLevel = 0.05,
  targetPower = 0.80,
  refVar = NULL
)
```

### Prospective power analysis

```{r power-prospective}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  powerAnalysis = TRUE,
  powerAnalysisType = "prospective",
  expectedAUCDifference = 0.10,
  refVar = NULL
)
```

### Sample size estimation

```{r power-sample-size}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  powerAnalysis = TRUE,
  powerAnalysisType = "sample_size",
  correlationROCs = 0.5,
  refVar = NULL
)
```

---

## 13. Bootstrap ROC with Prior Weighting

### Default prior (AUC=0.7, precision=10)

```{r bayesian-default}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  bayesianAnalysis = TRUE,
  priorAUC = 0.7,
  priorPrecision = 10,
  refVar = NULL
)
```

### Strong prior belief (AUC=0.9, high precision)

```{r bayesian-strong}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  bayesianAnalysis = TRUE,
  priorAUC = 0.9,
  priorPrecision = 50,
  refVar = NULL
)
```

---

## 14. Clinical Utility Analysis

### Decision curve analysis

```{r clinical-utility}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalUtilityAnalysis = TRUE,
  treatmentThreshold = "0.05,0.5,0.05",
  harmBenefitRatio = 0.25,
  refVar = NULL
)
```

### With intervention cost analysis

```{r clinical-utility-cost}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalUtilityAnalysis = TRUE,
  interventionCost = TRUE,
  harmBenefitRatio = 1.0,
  refVar = NULL
)
```

---

## 15. Meta-Analysis (3+ Variables Required)

### Fixed and random effects meta-analysis

```{r meta-both}
advanced_data <- read.csv(paste0(data_path2, "psychopdaROC_advanced.csv"))

psychopdaROC(
  data = advanced_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker_excellent, marker_good, marker_moderate,
                       marker_fair, marker_poor),
  metaAnalysis = TRUE,
  metaAnalysisMethod = "both",
  heterogeneityTest = TRUE,
  overrideMetaAnalysisWarning = TRUE,
  refVar = NULL
)
```

### Forest plot

```{r meta-forest}
psychopdaROC(
  data = advanced_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker_excellent, marker_good, marker_moderate,
                       marker_fair, marker_poor),
  metaAnalysis = TRUE,
  metaAnalysisMethod = "both",
  forestPlot = TRUE,
  overrideMetaAnalysisWarning = TRUE,
  refVar = NULL
)
```

### Fixed effects only (3 markers)

```{r meta-fixed}
psychopdaROC(
  data = advanced_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker_excellent, marker_good, marker_moderate),
  metaAnalysis = TRUE,
  metaAnalysisMethod = "fixed",
  overrideMetaAnalysisWarning = TRUE,
  refVar = NULL
)
```

---

## 16. Clinical Mode Presets

### Basic mode (minimal clinical output)

```{r mode-basic}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalMode = "basic",
  refVar = NULL
)
```

### Advanced mode

```{r mode-advanced}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalMode = "advanced",
  refVar = NULL
)
```

### Comprehensive mode (full research-grade output)

```{r mode-comprehensive}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalMode = "comprehensive",
  refVar = NULL
)
```

### Screening preset

```{r preset-screening}
psychopdaROC(
  data = screening_data,
  classVar = "cancer",
  positiveClass = "Cancer",
  dependentVars = "psa_level",
  clinicalPreset = "screening",
  refVar = NULL
)
```

### Confirmation preset

```{r preset-confirmation}
psychopdaROC(
  data = cardiac_data,
  classVar = "mi_status",
  positiveClass = "MI",
  dependentVars = "troponin",
  clinicalPreset = "confirmation",
  refVar = NULL
)
```

### Balanced and research presets

```{r preset-balanced}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalPreset = "balanced",
  refVar = NULL
)
```

```{r preset-research}
psychopdaROC(
  data = test_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  clinicalPreset = "research",
  refVar = NULL
)
```

---

## 17. Edge Cases

### Perfect separation (AUC=1.0)

```{r edge-perfect}
perfect_data <- read.csv(paste0(data_path, "psychopdaROC_perfect.csv"))

psychopdaROC(
  data = perfect_data,
  classVar = "condition",
  positiveClass = "Positive",
  dependentVars = "perfect_test",
  refVar = NULL
)
```

### No discrimination (AUC~0.5)

```{r edge-poor}
poor_data <- read.csv(paste0(data_path, "psychopdaROC_poor.csv"))

psychopdaROC(
  data = poor_data,
  classVar = "status",
  positiveClass = "Case",
  dependentVars = "poor_marker",
  refVar = NULL
)
```

### Moderate overlap (AUC~0.70)

```{r edge-overlap}
overlap_data <- read.csv(paste0(data_path, "psychopdaROC_overlap.csv"))

psychopdaROC(
  data = overlap_data,
  classVar = "diagnosis",
  positiveClass = "Diseased",
  dependentVars = "test_value",
  refVar = NULL
)
```

### Small sample (n=30)

```{r edge-small}
small_data <- read.csv(paste0(data_path, "psychopdaROC_small.csv"))

psychopdaROC(
  data = small_data,
  classVar = "class",
  positiveClass = "Positive",
  dependentVars = "marker",
  refVar = NULL
)
```

### Extreme imbalance (2% event rate)

```{r edge-imbalanced}
imbalanced_data <- read.csv(paste0(data_path, "psychopdaROC_imbalanced.csv"))

psychopdaROC(
  data = imbalanced_data,
  classVar = "rare_outcome",
  positiveClass = "Event",
  dependentVars = "predictor",
  refVar = NULL
)
```

### Constant predictor (zero variance)

```{r edge-constant}
constant_data <- read.csv(paste0(data_path, "psychopdaROC_constant.csv"))

psychopdaROC(
  data = constant_data,
  classVar = "outcome",
  positiveClass = "Positive",
  dependentVars = "constant_marker",
  refVar = NULL
)
```

### Missing data

```{r edge-missing}
missing_data <- read.csv(paste0(data_path, "psychopdaROC_missing.csv"))

psychopdaROC(
  data = missing_data,
  classVar = "diagnosis",
  positiveClass = "Disease",
  dependentVars = vars(test_a, test_b),
  refVar = NULL
)
```

### Rare disease (5% prevalence)

```{r edge-rare}
psychopdaROC(
  data = rare_data,
  classVar = "rare_disease",
  positiveClass = "Disease",
  dependentVars = "biomarker",
  refVar = NULL
)
```

### Large dataset (n=500)

```{r edge-large}
psychopdaROC(
  data = large_data,
  classVar = "disease_status",
  positiveClass = "Disease",
  dependentVars = vars(biomarker1, biomarker2),
  refVar = NULL
)
```

---

## 18. Time-Dependent & Spectrum Analysis

### Compare baseline vs follow-up markers

```{r timedep}
timedep_data <- read.csv(paste0(data_path, "psychopdaROC_timedep.csv"))

psychopdaROC(
  data = timedep_data,
  classVar = "outcome",
  positiveClass = "Event",
  dependentVars = vars(baseline_marker, followup_marker),
  delongTest = TRUE,
  combinePlots = TRUE,
  refVar = NULL
)
```

### Disease spectrum data

```{r spectrum}
spectrum_data <- read.csv(paste0(data_path, "psychopdaROC_spectrum.csv"))

psychopdaROC(
  data = spectrum_data,
  classVar = "binary_status",
  positiveClass = "Positive",
  dependentVars = "continuous_marker",
  refVar = NULL
)
```

---

## 19. Comprehensive Publication-Ready Scenario

Full research-grade analysis combining multiple features.

```{r comprehensive-pub}
psychopdaROC(
  data = multi_data,
  classVar = "diagnosis",
  positiveClass = "Positive",
  dependentVars = vars(marker1, marker2, marker3, combined_score),
  clinicalMode = "comprehensive",
  method = "maximize_metric",
  metric = "youden",
  delongTest = TRUE,
  calculateIDI = TRUE,
  calculateNRI = TRUE,
  refVar = NULL,
  bootstrapCI = TRUE,
  bootstrapReps = 200,
  clinicalUtilityAnalysis = TRUE,
  plotROC = TRUE,
  combinePlots = TRUE,
  cleanPlot = TRUE,
  sensSpecTable = TRUE
)
```

---

## References

- Youden WJ (1950). Index for rating diagnostic tests. *Cancer*, 3(1), 32-35.
- DeLong ER, DeLong DM, Clarke-Pearson DL (1988). Comparing the areas under
  two or more correlated receiver operating characteristic curves: a nonparametric
  approach. *Biometrics*, 44(3), 837-845.
- Pencina MJ, D'Agostino RB, D'Agostino RB Jr, Vasan RS (2008). Evaluating the
  added predictive ability of a new marker: from area under the ROC curve to
  reclassification and beyond. *Statistics in Medicine*, 27(2), 157-172.
- Thiele C, Hirschfeld G (2021). cutpointr: Improved estimation and validation of
  optimal cutpoints in R. *Journal of Statistical Software*, 98(11), 1-27.
- Robin X, Turck N, Hainard A, et al. (2011). pROC: an open-source package for R
  and S+ to analyze and compare ROC curves. *BMC Bioinformatics*, 12, 77.
- Vickers AJ, Elkin EB (2006). Decision curve analysis: a novel method for evaluating
  prediction models. *Medical Decision Making*, 26(6), 565-574.
