---
title: "Interobserver reliability beyond a single kappa"
subtitle: "All-pairs Cohen's kappa and per-category bottlenecks (YSRB-style cytology)"
author: "ClinicoPath"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Interobserver reliability beyond a single kappa}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  eval = TRUE
)
```

## When to use this

A reliability study almost never ends with one number. You report an overall
Fleiss' kappa, a reviewer asks *"which two readers actually disagreed?"* and
*"which diagnostic category is dragging the agreement down?"*, and suddenly a
single summary statistic is not enough. Interobserver studies of structured
reporting systems - the Yokohama System for Reporting Breast cytopathology
(YSRB), the Milan system, the Bethesda system, ISUP grading - routinely pair the
overall kappa with two supplementary tables:

* **Table I - every rater pair.** Cohen's kappa for each of the $\binom{k}{2}$
  reader pairs, with a confidence interval and a significance test. This is how
  you spot an outlier reader whose pairwise kappas are systematically lower than
  everyone else's.
* **Table II - every category.** A per-category agreement rate that answers
  *"when readers land on category c, how often do they all agree?"* This is how
  you find the diagnostic bottleneck - typically the "atypical / suspicious"
  middle category.

The `agreement` analysis produces both, alongside the overall kappa, from the
same set of rater columns.

## A synthetic YSRB-like dataset

The real study data are not redistributable, so we simulate a dataset with the
same shape: 99 cases, four readers, and five ordinal YSRB categories
(I = insufficient, II = benign, III = atypical, IV = suspicious, V = malignant)
with the characteristic heavily-benign marginal distribution.

```{r data}
library(ClinicoPath)

set.seed(2024)
n     <- 99
cats  <- c("I", "II", "III", "IV", "V")
prob  <- c(0.05, 0.75, 0.05, 0.04, 0.11)   # mostly benign, as in real cytology
truth <- sample(cats, n, replace = TRUE, prob = prob)

# Each reader agrees with the latent truth most of the time, otherwise slips to
# an adjacent category. Reader 4 is a little noisier than the rest.
slip <- function(x) {
  i <- match(x, cats)
  cats[pmin(length(cats), pmax(1, i + sample(c(-1, 1), 1)))]
}
reader <- function(p_agree) {
  vapply(truth, function(t) if (runif(1) < p_agree) t else slip(t), character(1))
}

ysrb <- data.frame(
  Reader1 = reader(0.92),
  Reader2 = reader(0.90),
  Reader3 = reader(0.91),
  Reader4 = reader(0.84),
  stringsAsFactors = FALSE
)
ysrb[] <- lapply(ysrb, factor, levels = cats, ordered = TRUE)

knitr::kable(head(ysrb), caption = "First few cases (4 readers, YSRB category)")
```

## Run the analysis

We turn on the all-pairs table, the per-category item-modal table, and a
Bonferroni correction for the $\binom{4}{2} = 6$ pairwise tests.

```{r run}
res <- agreement(
  data                   = ysrb,
  vars                   = c("Reader1", "Reader2", "Reader3", "Reader4"),
  allPairsKappa          = TRUE,
  itemModalCategoryAgreement = TRUE,
  multipleTestCorrection = "bonferroni"
)
```

### Overall agreement

```{r overall}
knitr::kable(res$irrtable$asDF, digits = 3,
             caption = "Overall agreement (Fleiss' kappa across all four readers)")
```

The overall kappa tells you the readers agree appreciably better than chance,
but it hides *where* the agreement comes from and *where* it breaks down.

## Table I - All-pairs Cohen's kappa

```{r table1}
allpairs <- res$allPairsKappaTable$asDF
knitr::kable(
  allpairs[, c("rater_a", "rater_b", "n", "peragree", "kappa",
               "ci_lower", "ci_upper", "p_adj")],
  digits = 3,
  col.names = c("Reader A", "Reader B", "n", "Obs. agree", "Kappa",
                "CI lower", "CI upper", "p (Bonf.)"),
  caption = "Cohen's kappa for every reader pair, with 95% CI."
)
```

Two points worth stressing about this table:

* **The confidence intervals are computed from the non-null asymptotic standard
  error** (via `vcd::Kappa`), so they agree with `psych::cohen.kappa()` and are
  *wider* - i.e. honest - compared with intervals built from the kappa/z test
  statistic, which uses the standard error under the null hypothesis of no
  agreement. A too-narrow CI would overstate precision.
* **Pairs involving the noisier Reader 4** sit at the bottom of the kappa
  ranking. In a real study this is exactly the signal that prompts targeted
  re-training rather than a blanket "agreement was moderate" conclusion.

## Table II - Agreement by modal category

For each case we take the modal (most common) reading across the four readers,
then, within each category, average the case-level agreement rate. A 4/4 case
scores 1.00, a 3/1 split scores 0.75, and a 2/2 split has no unique mode and is
excluded.

```{r table2}
modal <- res$itemModalAgreementTable$asDF
knitr::kable(
  modal,
  digits = 3,
  col.names = c("Modal category", "Cases", "Mean agreement",
                "CI lower", "CI upper"),
  caption = "Within-case agreement, by the case's modal category."
)
```

The benign category (II) carries most of the cases and shows high, tightly
estimated agreement. The rarer categories (I, III, IV) are represented by only a
handful of modal cases, so their agreement estimates are unstable and their
confidence intervals are wide - a reminder that a per-category mean is only as
trustworthy as its cell count. In a real, larger YSRB or Milan series this same
table is where the atypical/suspicious middle categories typically reveal
themselves as the agreement bottleneck; here it mainly illustrates that you
should read sparse-category rows with their CIs, not their point estimates.

## Two cautions the analysis flags for you

**The kappa paradox.** When a category is rare, kappa can be paradoxically low
even when observed agreement is high. If any category is sparse, the analysis
attaches a note to the overall table suggesting the prevalence-robust
alternatives it also offers - Gwet's AC1/AC2 and PABAK - as sensitivity
analyses. Enable those (and `bootstrapCI`) when your marginal distribution is as
lopsided as a typical benign-dominated cytology series.

**Multiplicity.** With $k$ readers you run $\binom{k}{2}$ pairwise tests - 6 for
four readers, 15 for six. The `Multiple Testing Correction` option (Bonferroni,
Benjamini-Hochberg, or Holm) adds an adjusted-p column to Table I so you do not
read six raw p-values as if they were one.

## Reproducibility

```{r session}
sessionInfo()
```
