---
title: "Diagnostic Test Evaluation"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",  echo = TRUE,
  eval = FALSE)
```

The **meddecide** package provides tools to evaluate medical diagnostic tests.
This vignette demonstrates the core functions for decision analysis.

## Example Data

Small example datasets are included with the package. You can load them
using `system.file()` and `read.csv()`.

```{r, eval=FALSE}
df_dec <- read.csv(system.file("extdata", "decision_example.csv", package = "meddecide"))
head(df_dec)
```

## Calculating Diagnostic Metrics

The `decision()` function computes sensitivity, specificity and related
metrics from raw test results.

```{r, eval=FALSE}
res <- decision(data = df_dec,
                gold = df_dec$gold,
                goldPositive = 1,
                newtest = df_dec$newtest,
                testPositive = 1,
                ci = TRUE)
res$ratioTable
```

When you only have the four counts (true positives, false positives,
true negatives and false negatives) you can use
`decisioncalculator()` directly.

```{r, eval=FALSE}
calc <- decisioncalculator(TP = 90, FN = 10, TN = 80, FP = 20,
                           ci = TRUE, fagan = TRUE)
calc$ratioTable
```

The option `fagan = TRUE` adds a Fagan nomogram to illustrate how the
pre-test probability is updated by the diagnostic result.

```{r, eval=FALSE}
calc$fagan
```

These tools help summarise diagnostic performance and can be combined
with other functions in **meddecide** for more advanced analysis.
