---
title: "Module Development for Jamovi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module Development for Jamovi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```




# ‘A beginners guide to jamovi modules’ or ‘jamovi modules for dummies’

# R




# jamovi









# General




Serdar Balci Nov 27th, 2019 at 7:00 PM
is it possible to use github-only packages as module dependencies?




2 replies
Also sent to the channel

Serdar Balci  3 months ago
I have installed Version 1.1.8.0 and now I can get github packages. The picture shows module using easystats/report package
Ekran Resmi 2019-11-28 20.14.55.png 
Ekran Resmi 2019-11-28 20.14.55.png



jonathon:whale2:  3 months ago
:+1:



---



Serdar Balci  3 months ago
Only n, nevent and rmean are displayed in jamovi environment but others are only visible in RStudio. Similar to macOS it is not only output, the generated tables have different column names and content. (edited) 
2 files 
Ekran Resmi 2019-12-05 15.48.47.png


Ekran Resmi 2019-12-05 15.48.32.png



Serdar Balci  3 months ago
I found out that survminer package also forms these tables, but its dependencies require never R versions (edited) 

jonathon:whale2:  3 months ago
 But similarly the generated tables by summary function to survfit objects are different.
might be worth chasing up the author of the package you’re using, to try and find why this is. there’s certainly nothing jamovi is doing which would lead to these differences.
yup, i’m working on updating the R version at the moment.
:+1:
1


Serdar Balci  3 months ago
Thank you. I will try to contact them.

Serdar Balci  2 months ago
This summary function issue is no longer a problem with R 3.6 and windows. Now I can use many packages and get results via jamovi module.

jonathon:whale2:  2 months ago
oh great


---


Serdar Balci  2:17 PM
Hello,
Using “preformatted” result element I get a markdown table output. Is there a way to somehow render/convert this output to html version. Or should I go with https://dev.jamovi.org/api_table.html table api? (edited) 
Ekran Resmi 2020-01-09 13.21.46.png 
Ekran Resmi 2020-01-09 13.21.46.png








jonathon:whale2:  11:53 PM
so you’re best to make use of the table api … the table API has a lot more features than an md table. (edited) 
:+1:
1


---




Serdar Balci  2 months ago
Thank you it works
Ekran Resmi 2020-01-19 18.16.14.png 
Ekran Resmi 2020-01-19 18.16.14.png



Damian Dropmann:piggy:  2 months ago
Awesome! That space around the 'new test positive' is a bit strange.

Damian Dropmann:piggy:  2 months ago
Maybe it's wrapped in another container or something.

Serdar Balci  2 months ago
Thank you. I have also recognised that. I thought it was due to spacing. Below is the yaml.
title: Medical Decision
name: decision
jus: '3.0'
stage: 0
compilerMode: tame
children:
  - type: VariableSupplier
    persistentItems: false
    stretchFactor: 1
    children:
      - type: TargetLayoutBox
        label: Gold Standart
        children:
          - type: VariablesListBox
            name: gold
            maxItemCount: 1
            isTarget: true
          - type: LevelSelector
            name: goldPositive
      - type: TargetLayoutBox
        label: New Test
        children:
          - type: VariablesListBox
            name: newtest
            maxItemCount: 1
            isTarget: true
          - type: LevelSelector
            name: testPositive (edited) 


---

     - name: p
       title: "p"
       type: number
       format: zto,pvalue
       
       
---



so wrt width/height, you can set that in the .r.yaml like so:
https://github.com/kylehamilton/MAJOR/blob/master/jamovi/bayesmetacorr.r.yaml#L46-L49
it’s possible to do it programmatically, with … image$setSize()


---







```
Serdar Balci  4:48 PM
 I think I am getting familiar with the codes :)
QuickTime Movie 
JamoviModule.mov
4 MB QuickTime Movie— Click to download
```





```
Serdar Balci Nov 29th, 2019 at 12:39 PM
Module names now have R version and OS in them. Does it mean that this will not work in windows Installing ClinicoPath_0.0.1-macos-R3.3.0.jmo

4 replies

jonathon:whale2:  3 months ago
It depends on whether there are any native R packages in your modules dependencies. Most modules do, but some don't. (You'll notice there's a "uses native" property there now too ... my intention is to use that to determine if a module can be used cross platform or not)

jonathon:whale2:  3 months ago
If there's native dependencies, then the module needs to be built separately for each os.

jonathon:whale2:  3 months ago
But I can take care of building it for different oses

Serdar Balci  3 months ago
Oh, I see. Thank you :slightly_smiling_face:
```



