---
title: "Module Development for Jamovi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module Development for Jamovi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```



# [useR!2017: *jamovi*: a spreadsheet for R](https://channel9.msdn.com/Events/useR-international-R-User-conferences/useR-International-R-User-2017-Conference/jamovi-a-spreadsheet-for-R)

<iframe src="https://channel9.msdn.com/Events/useR-international-R-User-conferences/useR-International-R-User-2017-Conference/jamovi-a-spreadsheet-for-R/player" width="960" height="540" allowFullScreen frameBorder="0" title="*jamovi*: a spreadsheet for R - Microsoft Channel 9 Video"></iframe>




# ‘A beginners guide to jamovi modules’ or ‘jamovi modules for dummies’

Official documentation is here: https://dev.jamovi.org/


## preparing development tools

jmvtools should be installed from the jamovi repo

https://dev.jamovi.org/tuts0101-getting-started.html


```r
install.packages('jmvtools', repos=c('https://repo.jamovi.org', 'https://cran.r-project.org'))
```


```r
jmvtools::check()
```


```r
jmvtools::install()
```

You can use `devtools::install()` to use your codes as a usual R package, submit to github or CRAN. `devtools::check()` does not like some jamovi folders so be sure to add them under `.Rbuildignore`


## add analysis

```
jmvtools::create('function')
```


# DESCRIPTION

`Imports`, `Depends`, `Suggests`, and `Remotes` have practically no difference in building jamovi modules. The `jmvtools::install()` copies libraries under *build* folder.

Under `Imports` `jmvcore` and `R6` are defaults.

With Remotes one can install github packages as well. But with each `jmvtools::install()` command it tries to check the updates, and if you are online throws an error. An `upgrade = FALSE, quick = TRUE` argument like in devtools::install() is not available, yet. One workaround is temporarily deleting Remotes from DESCRIPTION. The package folders continue to remain under build folder.

One can also directly copy package folders from system R package folder (find via `.libPaths()`) as well.


# NAMESPACE

No need to change.


# R

R folder is where the codes are present. There are two files.

## function.h.R

No need to change. Auto-updated and overwritten.

## function.b.R




jmvcore::toNumeric()
https://dev.jamovi.org/tuts0202-handling-data.html

can I just send whole data to plot function?
you usually don't want to, but sometimes it's appropriate. normally you just provide a summary of the data to the plot function ... just enough data for it to do it's job. but if you need the whole data set for the plot function, then you can specify requiresData: true on the image object. that means the plot function can access self$data.
i do it in the correlation matrix for example. there's no summary i could send ... the plot function needs all the data:
https://github.com/jamovi/jmv/blob/master/jamovi/corrmatrix.r.yaml#L143
jamovi/corrmatrix.r.yaml:143
      requiresData: true



# jamovi


## function.a.yaml


## function.r.yaml


### preformatted

Using “preformatted” result element I get a markdown table output. Is there a way to somehow render/convert this output to html version. Or should I go with https://dev.jamovi.org/api_table.html table api?

so you’re best to make use of the table api … the table API has a lot more features than an md table.


### p-value

```
     - name: p
       title: "p"
       type: number
       format: zto,pvalue
```




## function.u.yaml



## 0000.yaml


## 00refs.yaml

prepare a 00refs.yaml like this:
https://github.com/jamovi/jmv/blob/master/jamovi/00refs.yaml

attach references to objects in the .r.yaml file like this:

https://github.com/jamovi/jmv/blob/master/jamovi/ancova.r.yaml#L174


# Tables

I want a long table. I tried to use following but got error.

Below is my current .r.yaml
   - name:  irrtable
     title: Interrater Reliability
      type: Table
      rows:  1
      columns:
        - name: method
          title: 'Method'
          type: text
        - name: subjects
          title: 'Subjects'
          type: integer
        - name: raters
          title: 'Raters'
          type: integer
        - name: peragree
          title: 'Agreement %'
          type: number
        - name: kappa
          title: 'Kappa'
          type: number
        - name: z
          title: 'z'
          type: number
        - name: p
          title: 'p-value'
          type: number
          format: zto,pvalue


- try setting swapRowsColumns to true.
- alternatively, you can name your columns like this method[a], method[b], this will cause the row to be ‘folded’ where the value in method[b] appears below the value in method[a] . an example of this is the t-test:
https://github.com/jamovi/jmv/blob/master/jamovi/ttestis.r.yaml#L20



# Plots


# `build` folder

## `js` folder

### event.js



## R3.6.3-macos



# Installing

# `jmvtools::install`

# `devtools::install`


```r
devtools::install(upgrade = FALSE, quick = TRUE)
```



# General

## YAML


<iframe width="560" height="315" src="https://www.youtube.com/embed/cdLNKUoMc6c" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>



## RStudio options

## .gitignore

add following to .gitignore file

```
# jamovi
/build/
/build-*/
*.jmo
```

## R version

Try to use compatible packages with the jamovi's R version.

Use: **R 3.6.3** https://cran.r-project.org/bin/macosx/R-3.6.3.pkg


Use packages from mran:

```r
options(repos = c('https://repo.jamovi.org',
                  "https://cran.microsoft.com/snapshot/2020-05-01/"
                  )
)
```



## Electron

jamovi is electron based. See R, shiny, and electron based application development here: [Deploying a Shiny app as a desktop application with Electron](https://github.com/sbalci/clinicopathological#deploying-a-shiny-app-as-a-desktop-application-with-electron)




project structure

https://dev.jamovi.org/info_project-structure.html





# Other Codes












## Datasets

```
make a data folder (same as with an R package), and then you put entries in your 0000.yaml file:
https://github.com/gamlj/gamlj/blob/master/jamovi/0000.yaml#L47-L108
jamovi/0000.yaml:47-108
datasets:
  - name: qsport
    path: qsport.csv
    description: Training hours
    tags:
```

.omv and .csv allowed.
excel is also allowed but user does not see if it is csv or excel file.






# Here















## Error messages
data <- data.frame(outcome=c(1,0,0,1,NA,1))
data <- na.omit(data)
if ( ! is.numeric(data$outcome) || any(data$outcome != 0 & data$outcome != 1))
  stop('Outcome variable must only contains 1s and 0s')







it’s good to test lots of different data sets that a user may have … include missing values, really large values, etc. etc. and make sure your analyses always handle them, and provide useful error messages for why an analysis doesn’t work. you don’t want to leave the user uncertain why something isn’t working … otherwise they just give up.

part of our philosophy is that people shouldn't have to set their data up if they can't be bothered ... because with large data sets it can take a lot of time. so i'd encourage you to treat whatever the user provides you with as continuous, by converting it with toNumeric() ... more on our data philosophy here: https://dev.jamovi.org/tuts0202-handling-data.html


https://youtu.be/oWZrrWc6e74


in the options, you’ve got Survival Curve, and in the results, it’s Survival Plot … i’d encourage you to make these consistent.
also, if the Survival Curve is unchecked, i’d hide the Surival plot, rather than leaving all that vacant space there.

visible: (optionName)
https://github.com/jamovi/jmv/blob/master/jamovi/ttestis.r.yaml#L408-L416
jamovi/ttestis.r.yaml:408-416
            - name: qq
              type: Image
              description: Q-Q plot
              width: 350
              height: 300












is there a variable type for dates in jamovi? Can I force a user to add only a date to a VariablesListBox?
I tried to get info from a self$options$var  via lubridate::is.Date  and is.na.POSIXlt but it did not work
hi, we don’t have a date data type at this time … only integer, numeric, and character … you could have people enter dates as character, and parse them yourself, but i appreciate that’s a bit of a hack

Thank you. Dates are always a problem in my routine practice. I work with many international colleagues and always date column is a mess, and  people calculate survival time very differently. I want to have raw dates so that I can calculate survival time. I will try somehow going around.



learn YAML syntax

it’s a pretty straightforward syntax … you’ve basically got ‘objects’ where each of the elements have names, and you’ve got arrays, where each of the objects have an index. and that’s more-or-less all there is to it.
you can take a look at jmv for examples:
https://github.com/jamovi/jmv/tree/master/jamovi

I don't think we've got a list of allowed parameters anywhere. Probably your best bet is to browse through the .yaml files in jmv. I think you'll find there's not that many parameter names.







as a work-around, once it’s installed the package from the Remotes, you can remove it from the DESCRIPTION and it won’t keep installing it over and over







Hi, there are scarce sources for pairwise chi-square tests. I have found  rmngb::pairwise.chisq.test() and rmngb::pairwise.fisher.test()
but that package has been removed from CRAN. Would you consider implementing this feature?
I also thought to add these functions in a module, but I want to ask your policy about removed packages as well.
4 replies

jonathon:whale2:  18 days ago
provided the module can be built with an entry in REMOTES, we don’t care if it’s not on CRAN

jonathon:whale2:  18 days ago
… but you’re obviously taking a risk using something which isn’t maintained

Serdar Balci  18 days ago
Thanks. Maybe just copying that function with appropriate reference may solve maintenance issue. I will think about it.

jonathon:whale2:  18 days ago
oh yup


---

I have a question. I want to user to enter cut points in a box and then evaluate it as a vector.
the function is this:
summary(km_fit, times = c(12,36,60)
I want user to define times vector.
I have tried the following:
utimes <- jmvcore::decomposeTerms(self$options$cutp)
utimes <- as.vector(utimes)
summary(km_fit, times = utimes
a.yaml is as follows:
    - name: cutp
      title: Define at least two cutpoints (in months) for survival table
      type: String
      default: '12, 36, 60'
Would you please guide me to convert input into a vector. (edited)
3 replies

Serdar Balci  13 hours ago
I think this seems to work:
utimes <- self$options$cutp
utimes <- strsplit(utimes, ",")
utimes <- purrr::reduce(utimes, as.vector)
utimes <- as.numeric(utimes)
(edited)

jonathon:whale2:  5 hours ago
yup, this will do it too:
as.numeric(strsplit(utimes, ',')[[1]])
(it's better if you can avoid using purrr, because it's not really necessary, and you're better off reducing the amount of dependencies you use)

Serdar Balci  5 hours ago
thank you. :+1:

---


so wrt width/height, you can set that in the .r.yaml like so:
https://github.com/kylehamilton/MAJOR/blob/master/jamovi/bayesmetacorr.r.yaml#L46-L49
it’s possible to do it programmatically, with … image$setSize()


---







```
Serdar Balci  4:48 PM
 I think I am getting familiar with the codes :)
QuickTime Movie
JamoviModule.mov
4 MB QuickTime Movie— Click to download
```





```
Serdar Balci Nov 29th, 2019 at 12:39 PM
Module names now have R version and OS in them. Does it mean that this will not work in windows Installing ClinicoPath_0.0.1-macos-R3.3.0.jmo

4 replies

jonathon:whale2:  3 months ago
It depends on whether there are any native R packages in your modules dependencies. Most modules do, but some don't. (You'll notice there's a "uses native" property there now too ... my intention is to use that to determine if a module can be used cross platform or not)

jonathon:whale2:  3 months ago
If there's native dependencies, then the module needs to be built separately for each os.

jonathon:whale2:  3 months ago
But I can take care of building it for different oses

Serdar Balci  3 months ago
Oh, I see. Thank you :slightly_smiling_face:
```









---

# Develop



```{r library, eval=FALSE, include=FALSE}
# install.packages('jmvtools', repos=c('https://repo.jamovi.org', 'https://cran.r-project.org'))

# jmvtools::check("C://Program Files//jamovi//bin")

# jmvtools::install(home = "C://Program Files//jamovi//bin")
#
# devtools::build(path = "C:\\ClinicoPathOutput")

# .libPaths(new = "C:\\ClinicoPathLibrary")

# devtools::build(path = "C:\\ClinicoPathOutput", binary = TRUE, args = c('--preclean'))

Sys.setenv(TZ="Europe/Istanbul")

library("jmvtools")
```


```{r check, eval=FALSE, include=FALSE}
jmvtools::check()

# rhub::check_on_macos()

# rhub::check_for_cran()

# codemetar::write_codemeta()


devtools::check()

```


```{r pkgdown build, eval=FALSE, include=FALSE}
rmarkdown::render('/Users/serdarbalciold/histopathRprojects/ClinicoPath/README.Rmd',  encoding = 'UTF-8', knit_root_dir = '~/histopathRprojects/ClinicoPath', quiet = TRUE)

devtools::document()

pkgdown::build_site()
```


```{r git force push, eval=FALSE, include=FALSE}
# gitUpdateCommitPush
CommitMessage <- paste("updated on ", Sys.time(), sep = "")
wd <- getwd()
gitCommand <- paste("cd ", wd, " \n git add . \n git commit --message '", CommitMessage, "' \n git push origin master \n", sep = "")
# gitCommand <- paste("cd ", wd, " \n git add . \n git commit --no-verify --message '", CommitMessage, "' \n git push origin master \n", sep = "")
system(command = gitCommand, intern = TRUE)

```



```{r add analysis, eval=FALSE, include=FALSE}

# jmvtools::install()
#
# jmvtools::create('SuperAwesome')
#
# jmvtools::addAnalysis(name='ttest', title='Independent Samples T-Test')
#
# jmvtools::addAnalysis(name='survival', title='survival')
#
# jmvtools::addAnalysis(name='correlation', title='correlation')
#
# jmvtools::addAnalysis(name='tableone', title='TableOne')
#
# jmvtools::addAnalysis(name='crosstable', title='CrossTable')
#
#
# jmvtools::addAnalysis(name='writesummary', title='WriteSummary')

# jmvtools::addAnalysis(name='finalfit', title='FinalFit')

# jmvtools::addAnalysis(name='multisurvival', title='FinalFit Multivariate Survival')

# jmvtools::addAnalysis(name='report', title='Report General Features')

# jmvtools::addAnalysis(name='frequencies', title='Frequencies')

# jmvtools::addAnalysis(name='statsplot', title='GGStatsPlot')

# jmvtools::addAnalysis(name='statsplot2', title='GGStatsPlot2')

# jmvtools::addAnalysis(name='scat2', title='scat2')

# jmvtools::addAnalysis(name='decisioncalculator', title='Decision Calculator')

# jmvtools::addAnalysis(name='agreement', title='Interrater Intrarater Reliability')

# jmvtools::addAnalysis(name='cluster', title='Cluster Analysis')

# jmvtools::addAnalysis(name='tree', title='Decision Tree')

```






```{r devtools install, eval=FALSE, include=FALSE}
devtools::install()
```


```{r jmvtools install, eval=FALSE, include=FALSE}
# jmvtools::check()
jmvtools::install()
```



```{r construct, eval=FALSE, include=FALSE}
formula <- jmvcore::constructFormula(terms = c("A", "B", "C"), dep = "D")

jmvcore::constructFormula(terms = list("A", "B", c("C", "D")), dep = "E")


jmvcore::constructFormula(terms = list("A", "B", "C"))

vars <- jmvcore::decomposeFormula(formula = formula)

unlist(vars)

cformula <- jmvcore::composeTerm(components = formula)

jmvcore::composeTerm("A")

jmvcore::composeTerm(components = c("A", "B", "C"))

jmvcore::decomposeTerm(term = c("A", "B", "C"))

jmvcore::decomposeTerm(term = formula)

jmvcore::decomposeTerm(term = cformula)



composeTerm <- jmvcore::composeTerm(components = c("A", "B", "C"))

jmvcore::decomposeTerm(term = composeTerm)


```









---

# Example


```{r read data, eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
```


```{r writesummary, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

# library("ClinicoPath")

deneme$Age <- as.numeric(as.character(deneme$Age))

ClinicoPath::writesummary(data = deneme, vars = Age)

ggstatsplot::normality_message(deneme$Age, "Age")


ClinicoPath::writesummary(
    data = deneme,
    vars = Age)


```



```{r finalfit, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
library(dplyr)
library(survival)
library(finalfit)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ClinicoPath::finalfit(data = deneme,
                      explanatory = Sex,
                      outcome = Outcome,
                      overalltime = OverallTime)
```



```{r decision, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

ClinicoPath::decision(
    data = deneme,
    gold = Outcome,
    goldPositive = "1",
    newtest = Smoker,
    testPositive = "TRUE")

ClinicoPath::decision(
    data = deneme,
    gold = LVI,
    goldPositive = "Present",
    newtest = PNI,
    testPositive = "Present")
```

```{r eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ggstatsplot::ggbetweenstats(data = deneme,
                            x = LVI,
                            y = Age)

```





```{r statsplot, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ClinicoPath::statsplot(
    data = deneme,
    dep = Age,
    group = Smoker)
```




```{r decision 2, eval=FALSE, include=FALSE}
mytable <- table(deneme$Outcome, deneme$Smoker)

caret::confusionMatrix(mytable)
confusionMatrix(pred, truth)
confusionMatrix(xtab, prevalence = 0.25)

levels(deneme$Outcome)

mytable[1,2]

d <- "0"

mytable[d, "FALSE"]

mytable[[0]]

```











```{r construct formula, eval=FALSE, include=FALSE}
formula <- jmvcore::constructFormula(terms = c("A", "B", "C"))

jmvcore::constructFormula(terms = list("A", "B", "C"))

vars <- jmvcore::decomposeFormula(formula = formula)

vars <- jmvcore::decomposeTerms(vars)


vars <- unlist(vars)

formula <- as.formula(formula)


my_group <- "lvi"
my_dep <- "age"

formula <- paste0('x = ', group, 'y = ', dep)
myformula <- as.formula(formula)

myformula <- glueformula::gf({my_group}, {my_dep})

myformula <- glue::glue( 'x = ' , {my_group}, ', y = ' , {my_dep})

myformula <- jmvcore::composeTerm(myformula)


```


```{r, eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

library(survival)
km_fit <- survfit(Surv(OverallTime, Outcome) ~ LVI, data = deneme)

library(dplyr)
km_fit_median_df <- summary(km_fit)
km_fit_median_df <- as.data.frame(km_fit_median_df$table) %>%
                janitor::clean_names(dat = ., case = "snake") %>%
                tibble::rownames_to_column(.data = ., var = "LVI")

```



```{r construct formula 2, eval=FALSE, include=FALSE}
library(dplyr)
library(survival)

deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

myoveralltime <- deneme$OverallTime
myoutcome <- deneme$Outcome
myexplanatory <- deneme$LVI

class(myoveralltime)
class(myoutcome)
typeof(myexplanatory)

is.ordered(myexplanatory)

formula2 <- jmvcore::constructFormula(terms = "myexplanatory")
# formula2 <- jmvcore::decomposeFormula(formula = formula2)
# formula2 <- paste("", formula2)
# formula2 <- as.formula(formula2)
formula2 <- jmvcore::composeTerm(formula2)


formulaL <- jmvcore::constructFormula(terms = "myoveralltime")
# formulaL <- jmvcore::decomposeFormula(formula = formulaL)

formulaR <- jmvcore::constructFormula(terms = "myoutcome")
# formulaR <- jmvcore::decomposeFormula(formula = formulaR)

formula <- paste("Surv(", formulaL, ",", formulaR, ")")
# formula <- jmvcore::composeTerm(formula)
# formula <- as.formula(formula)
# jmvcore::constructFormula(terms = c(formula, formula2))

deneme %>%
  finalfit::finalfit(formula, formula2) -> tUni

tUni
```


```{r eval=FALSE, include=FALSE}
library(dplyr)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

results <- deneme %>%
                ggstatsplot::ggbetweenstats(LVI, Age)
results

mydep <- deneme$Age
mygroup <- deneme$LVI


mygroup <- jmvcore::constructFormula(terms = "mygroup")
mygroup <- jmvcore::composeTerm(mygroup)

mydep <- jmvcore::constructFormula(terms = "mydep")
mydep <- jmvcore::composeTerm(mydep)


# not working
# eval(mygroup)
# rlang::eval_tidy(mygroup)
# !!mygroup
# {{mygroup}}
# sym(mygroup)
# quote(mygroup)
# enexpr(mygroup)

mygroup <- jmvcore::constructFormula(terms = "mygroup")
mydep <- jmvcore::constructFormula(terms = "mydep")

formula1 <- paste(mydep)
formula1 <- jmvcore::composeTerm(formula1)


mygroup <- paste(mygroup)
mygroup <- jmvcore::composeTerm(mygroup)

mydep <- deneme$Age
mygroup <- deneme$LVI

mydep <- jmvcore::resolveQuo(jmvcore::enquo(mydep))
mygroup <- jmvcore::resolveQuo(jmvcore::enquo(mygroup))

mydata2 <- data.frame(mygroup=mygroup, mydep=mydep)

results <- mydata2 %>%
                ggstatsplot::ggbetweenstats(
x = mygroup, y = mydep  )

results



myformula <- glue::glue('x = ', {mygroup}, ', y = ' , {mydep})

myformula <- jmvcore::composeTerm(myformula)

myformula <- as.formula(myformula)
















mydep2 <- quote(mydep)
mygroup2 <- quote(mygroup)


results <- deneme %>%
                ggstatsplot::ggbetweenstats(!!mygroup2, !!mydep2)
results
```








```{r construct formula 3, eval=FALSE, include=FALSE}
formula <- jmvcore::constructFormula(terms = c("myoveralltime", "myoutcome"))

vars <- jmvcore::decomposeFormula(formula = formula)


explanatory <- jmvcore::constructFormula(terms = c("explanatory"))

explanatory <- jmvcore::decomposeFormula(formula = explanatory)

explanatory <- unlist(explanatory)

myformula <- paste("Surv(", vars[1], ", ", vars[2], ")")



deneme %>%
finalfit::finalfit(myformula, explanatory) -> tUni

```


```{r table tangram, eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))


table3 <-
  tangram::html5(
    tangram::tangram(
      "Death ~ LVI + PNI + Age", deneme),
    fragment=TRUE,
    inline="nejm.css",
    caption = "HTML5 Table NEJM Style",
    id="tbl3")

table3

mydep <- deneme$Age
mygroup <- deneme$Death


formulaR <- jmvcore::constructFormula(terms =  c("LVI", "PNI", "Age"))

formulaL <- jmvcore::constructFormula(terms = "Death")

formula <- paste(formulaL, '~', formulaR)

formula <- as.formula(formula)


table <- tangram::html5(
    tangram::tangram(formula, deneme
                     ))

table


```






---




## arsenal

```{r arsenal, results='asis', eval=FALSE, include=FALSE}

tab1 <- arsenal::tableby(~ Age + Sex, data = deneme)

results <- summary(tab1)


# results$object
# results$control
# results$totals
# results$hasStrata
# results$text
# results$pfootnote
# results$term.name
#
# tab1$Call
#
# tab1$control

tab1$tables # this is where results lie


```


## define survival time

```{r define survival time, eval=FALSE, include=FALSE}
mydata$int <- lubridate::interval(
  lubridate::ymd(mydata$SurgeryDate),
  lubridate::ymd(mydata$LastFollowUpDate)
  )
mydata$OverallTime <- lubridate::time_length(mydata$int, "month")
mydata$OverallTime <- round(mydata$OverallTime, digits = 1)
```




## Multivariate Analysis Survival


```{r Multivariate Analysis, eval=FALSE, include=FALSE}
library(finalfit)
library(survival)
explanatoryMultivariate <- explanatoryKM
dependentMultivariate <- dependentKM

mydata %>%
  finalfit(dependentMultivariate, explanatoryMultivariate) -> tMultivariate

knitr::kable(tMultivariate, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```


---

```{r, eval=FALSE, include=FALSE}
# Find arguments in yaml

list_of_yaml <- c(
    list.files(path = "~/histopathRprojects/ClinicoPath-Jamovi--prep/jmv",
               pattern = "\\.yaml$",
               full.names = TRUE,
               all.files = TRUE,
               include.dirs = TRUE,
               recursive = TRUE
    )
)


text_of_yaml_yml <- purrr::map(
    .x = list_of_yaml,
    .f = readLines
)

text_of_yaml_yml <- as.vector(unlist(text_of_yaml_yml))

arglist <-
    stringr::str_extract(
        string = text_of_yaml_yml,
        pattern =
            "([[:alnum:]]*):"
    )

arglist <- arglist[!is.na(arglist)]
arglist <- unique(arglist)
arglist <- gsub(pattern = ":", # remove some characters
                    replacement = "",
                    x = arglist)
arglist <- trimws(arglist) # remove whitespace

cat(arglist, sep = "\n")

```






<!-- # -->
<!-- #                 # tUni_df_descr <- paste0("When ", -->
<!-- #                 #                         tUni_df$dependent_surv_overall_time_outcome[1], -->
<!-- #                 #                         " is ", -->
<!-- #                 #                         tUni_df$x[2], -->
<!-- #                 #                         ", there is ", -->
<!-- #                 #                         tUni_df$hr_univariable[2], -->
<!-- #                 #                         " times risk than ", -->
<!-- #                 #                         "when ", -->
<!-- #                 #                         tUni_df$dependent_surv_overall_time_outcome[1], -->
<!-- #                 #                         " is ", -->
<!-- #                 #                         tUni_df$x[1], -->
<!-- #                 #                         "." -->
<!-- #                 # ) -->
<!-- # -->
<!-- #                 # results5 <- tUni_df_descr -->


```{r, eval=FALSE, include=FALSE}
boot::melanoma
rio::export(x = boot::melanoma, file = "data/melanoma.csv")

survival::colon
rio::export(x = survival::colon, file = "data/colon.csv")

# BreastCancerData <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data"
#
# BreastCancerNames <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names"
#
# BreastCancerData <- read.csv(file = BreastCancerData, header = FALSE,
#                 col.names = c("id","CT", "UCSize", "UCShape", "MA", "SECS", "BN", "BC", "NN","M", "diagnosis") )

library(mlbench)

data("BreastCancer")
BreastCancer

rio::export(x = BreastCancer, file = "data/BreastCancer.csv")



```




```{r pairwise, eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
# names(deneme)

mypairwise <-  survminer::pairwise_survdiff(
                            formula = survival::Surv(OverallTime, Outcome) ~ TStage,
                            data = deneme,
                            p.adjust.method = "BH"
                            )


mypairwise2 <- as.data.frame(mypairwise[["p.value"]]) %>%
  tibble::rownames_to_column()

mypairwise2 %>%
  tidyr::pivot_longer(cols = -rowname) %>%
  dplyr::filter(complete.cases(.)) %>%
  dplyr::mutate(description =
                            glue::glue(
                                "The comparison between {rowname} and {name} has a p-value of {round(value, 2)}."
                            )
                    ) %>%
                    dplyr::select(description) %>%
                    dplyr::pull() -> mypairwisedescription

mypairwisedescription <- unlist(mypairwisedescription)

mypairwisedescription <- c(
"In the pairwise comparison of",
mypairwisedescription)




```
