---
title: "Motivation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```



















https://github.com/sbalci/ClinicoPathJamoviModule


This repository is a template for the codes I use in [my and my friends' histopathology research](https://sbalci.github.io/cv/SerdarBalciMDPathologist.html).


---


**See examples**

- See some of the codes here: https://sbalci.github.io/histopathology-template/Report.html


- You may also install [histopathR](https://sbalci.github.io/histopathR/) package to get all the codes as R Markdown Template.


---


``` r
devtools::install_github("sbalci/histopathR")
rmarkdown::draft(file = "myfile",
                 template = "clinicopathological",
                 package = "histopathR",
                 create_dir = TRUE,
                 edit = TRUE)
```


- I have tried to collect common codes in [My R Codes For Data Analysis](https://sbalci.github.io/MyRCodesForDataAnalysis/). I plan to collect them under the package [histopathR](https://sbalci.github.io/histopathR/) gradually.


---


**How to use**

- *In RStudio:*

   - Fork this repository. Then use this repository as template as described [here](https://help.github.com/en/articles/creating-a-repository-from-a-template).

   - Use RStudio project as described [here](https://happygitwithr.com/existing-github-first.html#new-rstudio-project-via-git-clone-1).

   - Do not forget to change chunk functions to `eval = TRUE`.

   - This template contains packages using [renv](https://rstudio.github.io/renv/articles/renv.html).

- *Via Docker:*

https://hub.docker.com/r/serdarbalci/clinicopathological

```bash
docker pull serdarbalci/clinicopathological
```


https://twitter.com/serdarbalci/status/1197849208302252032


- *As a standalone app (for mac only):*



https://sourceforge.net/projects/histopathr/

https://twitter.com/serdarbalci/status/1198348640294711296


https://serdarbalci.gitbook.io/parapathology/computers/computers#deploying-a-shiny-app-as-a-desktop-application-with-electron

- *As a Shiny app:*

Launch here:
https://serdarbalci.shinyapps.io/histopathR/

Run in your own computer R session:

```r
library(shiny)
runUrl("https://github.com/sbalci/histopathRshiny/raw/master/histopathRshiny.tar.gz")
```



Code here:
https://github.com/sbalci/histopathRshiny


- *As a jamovi module:*

Download module here:

https://github.com/sbalci/ClinicoPathJamoviModule/raw/master/ClinicoPath.jmo

https://github.com/sbalci/ClinicoPathJamoviModule

https://sourceforge.net/projects/clinicopathjamovimodule/


---

**Recommended Links for Reproducible and ClinicoPathological Research**

- [Reproducible Templates for Analysis and Dissemination](https://www.coursera.org/learn/reproducible-templates-analysis/home/info)

- [Happy Git and GitHub for the useR](https://happygitwithr.com/)

- [R-Giris Dersi Sunum](https://sbalci.github.io/MyRCodesForDataAnalysis/R-Giris.html)

- [R-Giris Dersi](https://sbalci.github.io/MyRCodesForDataAnalysis/R-Giris.nb.html)

- [R-Markdown Dersi Sunum](https://sbalci.github.io/MyRCodesForDataAnalysis/R-Markdown.nb.html)

- [R-Markdown Dersi](https://sbalci.github.io/MyRCodesForDataAnalysis/R-Markdown.html)

- [How to Prepare Data for Histopathology Research pdf](https://sbalci.github.io/MyRCodesForDataAnalysis/How-to-Prepare-Data-for-Histopathology-Research.pdf)

- [How to Prepare Data for Histopathology Research](https://sbalci.github.io/MyRCodesForDataAnalysis/How-to-Prepare-Data-for-Histopathology-Research.nb.html)

- [How to Prepare Data for Histopathology Research Presentation](https://sbalci.github.io/MyRCodesForDataAnalysis/How-to-Prepare-Data-for-Histopathology-Research.html)


- ["The Quartz guide to bad data"nın Türkçe çevirisi: Kötü Veri Kılavuzu ](https://sbalci.github.io/Kotu-Veri-Kilavuzu/)


- [Create A MS Word Template for R Markdown](https://vimeo.com/110804387)


---


**Please Give Me FeedBack**

https://goo.gl/forms/YjGZ5DHgtPlR1RnB3


