---
title: "Advanced TNM Stage Migration Analysis - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced TNM Stage Migration Analysis - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `stagemigration()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `stagemigrationClass$new(options = stagemigrationOptions$new(...), data = mydata)`.

# Advanced TNM Stage Migration Analysis

## Overview

When a new edition of the AJCC/TNM staging manual is released, pathologists and
oncologists face a critical question: *Does the revised staging system actually
provide better prognostic discrimination than the one it replaces?* This is not a
trivial question. Patients who were Stage II under the old criteria may become
Stage III under the new criteria -- or vice versa. If the reclassification
genuinely separates patients with different prognoses, the new system is an
improvement. If it simply reshuffles patients without prognostic benefit, or worse,
introduces the **Will Rogers phenomenon** (where survival appears to improve in
every stage simply because of patient reclassification), then the new system may be
misleading.

The `stagemigration` module provides a state-of-the-art statistical toolkit for
answering this question. It goes far beyond a simple cross-tabulation of old vs.
new stages. The analysis includes formal discrimination metrics (Harrell's C-index),
reclassification indices (NRI, IDI), decision curve analysis for clinical utility,
bootstrap validation for internal validity, and multiple visualization tools for
presentations and publications. Cancer-type-specific thresholds and interpretation
guidelines are built in for lung, breast, colorectal, prostate, head and neck, and
melanoma.

Whether you are preparing a manuscript evaluating AJCC 8th vs. 7th edition staging,
validating an institutional modification to the standard staging system, or
assessing a biomarker-enhanced staging proposal, this module provides the full
analytical framework recommended by current staging validation literature.

---

## Datasets

The examples in this vignette use synthetic datasets that mimic realistic TNM
staging migration patterns. Because the bundled `.rda` files may require
regeneration, we create the data inline to guarantee reproducibility.

```{r generate-data}
set.seed(12345)

# --- Helper functions ---
generate_survival_times <- function(stage, hazard_base = 0.02,
                                    stage_multipliers = c(1, 1.5, 2.5, 4)) {
  stage_numeric <- as.numeric(stage)
  hazard <- hazard_base * stage_multipliers[stage_numeric]
  times <- rexp(length(stage), rate = hazard) * 12 + rnorm(length(stage), 0, 2)
  pmax(times, 0.1)
}

generate_censoring <- function(survival_times, censoring_rate = 0.3) {
  prob <- pmin(censoring_rate + (survival_times - median(survival_times)) / 100, 0.8)
  prob <- pmax(prob, 0.1)
  rbinom(length(survival_times), 1, 1 - prob)
}

create_stage_migration <- function(old_stage, migration_prob = 0.25) {
  new_stage <- as.numeric(old_stage)
  n <- length(old_stage)
  migrate <- sample(seq_len(n), size = round(n * migration_prob))
  for (i in migrate) {
    cs <- new_stage[i]
    if (cs == 1) new_stage[i] <- sample(c(1, 2, 3), 1, prob = c(0.4, 0.4, 0.2))
    else if (cs == 2) new_stage[i] <- sample(1:4, 1, prob = c(0.2, 0.3, 0.3, 0.2))
    else if (cs == 3) new_stage[i] <- sample(2:4, 1, prob = c(0.2, 0.4, 0.4))
    else new_stage[i] <- sample(3:4, 1, prob = c(0.1, 0.9))
  }
  factor(new_stage, levels = 1:4,
         labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))
}

# --- Combined dataset (breast + lung + colorectal, N = 2100) ---
make_cohort <- function(n, cancer, hazard_base, mig_prob) {
  age <- pmin(pmax(round(rnorm(n, 64, 12)), 30), 90)
  sex <- factor(sample(c("Male", "Female"), n, replace = TRUE))
  old_num <- sample(1:4, n, replace = TRUE, prob = c(0.28, 0.30, 0.24, 0.18))
  old_stage <- factor(old_num, 1:4,
                      labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))
  new_stage <- create_stage_migration(old_stage, migration_prob = mig_prob)
  st <- generate_survival_times(new_stage, hazard_base = hazard_base,
                                stage_multipliers = c(0.7, 1.2, 2.0, 3.5))
  ev <- generate_censoring(st, censoring_rate = 0.35)
  data.frame(age = age, sex = sex, old_stage = old_stage, new_stage = new_stage,
             survival_time = round(st, 1), event = ev, cancer_type = cancer,
             stringsAsFactors = FALSE)
}

lung_df    <- make_cohort(700, "Lung",       0.015, 0.30)
breast_df  <- make_cohort(700, "Breast",     0.008, 0.25)
crc_df     <- make_cohort(700, "Colorectal", 0.012, 0.28)

combined_data <- rbind(lung_df, breast_df, crc_df)
combined_data$patient_id <- seq_len(nrow(combined_data))

# --- Small sample dataset (N = 50, edge-case testing) ---
small_data <- make_cohort(50, "Mixed", 0.015, 0.20)

cat("Combined data:", nrow(combined_data), "patients,",
    sum(combined_data$event), "events\n")
cat("Small data:", nrow(small_data), "patients,",
    sum(small_data$event), "events\n")
```

| Dataset | N | Key Features |
|---------|---|-------------|
| `combined_data` | 2100 | Combined breast/lung/colorectal, 4-stage system |
| `lung_df` | 700 | Lung cancer specific, higher hazard |
| `breast_df` | 700 | Breast cancer specific, lower hazard |
| `small_data` | 50 | Small sample for edge-case testing |

---

## 1. Basic Migration Analysis

The simplest use case: compare two staging systems and display the migration matrix
plus an overview table. This is the starting point for any staging validation.

```{r basic}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "basic",
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE
)
```

The **migration overview** tells you how many patients changed stage (and in which
direction), while the **migration matrix** shows the exact cross-tabulation. Diagonal
cells are patients who stayed in the same stage; off-diagonal cells are patients who
migrated. Above the diagonal = upstaged, below = downstaged.

---

## 2. Analysis Types

The `analysisType` option controls the scope of the analysis. There are four levels:

- **basic** -- Migration matrices and distribution tables only
- **standard** -- Adds C-index comparison and NRI
- **comprehensive** -- All statistical methods (default)
- **publication** -- Optimized output formatting for manuscripts

```{r standard-analysis}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "standard",
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE,
  showStageDistribution = TRUE,
  showStatisticalComparison = TRUE,
  showMigrationSummary = TRUE
)
```

The **stage distribution comparison** shows how patient counts shift between
systems. The **statistical comparison** provides C-index values for each staging
system so you can see which one discriminates better.

---

## 3. Net Reclassification Improvement (NRI) and Integrated Discrimination Improvement (IDI)

NRI quantifies whether the new staging system correctly reclassifies patients --
moving event patients to higher-risk stages and non-event patients to lower-risk
stages. IDI measures the integrated improvement in predicted probabilities.

These are the gold-standard metrics for staging validation, recommended by Pencina
et al. (2008) and widely used in AJCC staging literature.

```{r nri-idi}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  calculateNRI = TRUE,
  nriTimePoints = "12, 24, 60",
  calculateIDI = TRUE,
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE
)
```

**Interpreting NRI:**

- NRI > 0: Net improvement in classification with the new system
- NRI > 0.20 (the default `nriClinicalThreshold`): Clinically meaningful improvement
- The event NRI and non-event NRI components tell you whether the improvement comes
  from better classification of patients who had events, those who did not, or both

**Interpreting IDI:**

- IDI > 0: Improved discrimination
- IDI represents the increase in the difference between mean predicted probabilities
  for events and non-events

---

## 4. C-index Comparison

Harrell's concordance index (C-index) measures the staging system's ability to rank
patients by their survival prognosis. A C-index of 0.5 means random discrimination;
1.0 means perfect discrimination. In staging validation, a clinically meaningful
improvement is typically 0.02 or greater (configurable via `clinicalSignificanceThreshold`).

```{r cindex}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "standard",
  showConcordanceComparison = TRUE,
  showStatisticalComparison = TRUE,
  includeEffectSizes = TRUE,
  clinicalSignificanceThreshold = 0.02
)
```

The concordance comparison table shows:

- C-index for the old staging system
- C-index for the new staging system
- The difference (Delta C) and its confidence interval
- Whether the improvement crosses the clinical significance threshold

---

## 5. Will Rogers Effect Detection

The Will Rogers phenomenon occurs when patient reclassification creates the illusion
of improvement. Named after Will Rogers' quip about Oklahomans moving to California
and raising the average intelligence of both states, it can make a new staging system
appear superior when it is not.

The module detects this by comparing survival within each stage between patients who
migrated and those who did not. If migrated patients have systematically different
survival than the non-migrated patients in their new stage, the Will Rogers effect
is operating.

```{r will-rogers}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  showWillRogersAnalysis = TRUE,
  showWillRogersVisualization = TRUE,
  advancedMigrationAnalysis = TRUE,
  showMigrationHeatmap = TRUE
)
```

The **Will Rogers analysis table** provides a multi-criteria evidence assessment
with a traffic-light grading system (PASS / BORDERLINE / CONCERN / FAIL). The
**visualization** shows how survival curves shift within stages when patients are
reclassified.

---

## 6. Bootstrap Validation

Internal validation using the bootstrap is essential for ensuring your findings are
not over-optimistic. The module performs optimism-corrected estimates of the C-index
difference, using Harrell's recommended bootstrap procedure.

```{r bootstrap}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  performBootstrap = TRUE,
  bootstrapReps = 200,
  useOptimismCorrection = TRUE,
  showStatisticalComparison = TRUE
)
```

With 200 bootstrap repetitions (use 1000 for publications), the output includes:

- Apparent C-index difference (what you see in the data)
- Optimism estimate (how much the apparent estimate is inflated)
- Optimism-corrected C-index difference (the honest estimate)
- Bootstrap confidence intervals

If the optimism-corrected estimate remains above your clinical significance
threshold, you have robust evidence that the new staging system is genuinely better.

---

## 7. Decision Curve Analysis (DCA)

Decision Curve Analysis moves beyond discrimination to clinical utility. It asks:
*At what range of decision thresholds does using the new staging system lead to
better clinical decisions than treating all patients, treating no patients, or using
the old staging system?*

```{r dca}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  performDCA = TRUE,
  showDecisionCurves = TRUE,
  showClinicalInterpretation = TRUE
)
```

The **decision curve plot** shows net benefit on the y-axis across threshold
probabilities on the x-axis. The new staging system is clinically useful at
thresholds where its curve lies above both the "treat all" and "treat none"
reference lines, and ideally above the old staging system curve.

---

## 8. Calibration Analysis

Calibration assesses whether the predicted survival probabilities from the staging
system match observed outcomes. A well-calibrated staging system not only ranks
patients correctly (discrimination) but also assigns accurate absolute risk estimates.

```{r calibration}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  performCalibration = TRUE,
  showCalibrationPlots = TRUE
)
```

The **calibration plots** compare predicted vs. observed survival. Points lying on
the 45-degree line indicate perfect calibration. Systematic deviation above the line
means the model overestimates survival; below the line means it underestimates.

---

## 9. Survival Curves

Kaplan-Meier curves stratified by stage are the most intuitive way to visualize
staging system performance. Well-separated curves with no crossing indicate good
prognostic discrimination.

```{r survival-curves}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "standard",
  showSurvivalCurves = TRUE,
  survivalPlotType = "separate",
  showConfidenceIntervals = TRUE,
  showRiskTables = TRUE,
  showForestPlot = TRUE
)
```

Available `survivalPlotType` options:

- **separate**: Individual KM plots for old and new staging systems
- **sidebyside**: Old and new systems plotted side by side for visual comparison
- **overlay**: Both systems overlaid on the same axes

The **forest plot** shows stage-specific hazard ratios with confidence intervals,
making it easy to compare the magnitude of between-stage separation in each system.

---

## 10. Clinical Presets

For users who do not want to manually configure dozens of options, the module
provides clinical presets that activate sensible combinations:

```{r preset-research}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  clinicalPreset = "research_study"
)
```

| Preset | What It Enables | Best For |
|--------|----------------|----------|
| `routine_clinical` | Migration matrix, C-index, basic recommendation | Daily clinical validation |
| `research_study` | + NRI, survival curves, Will Rogers, bootstrap | Academic research projects |
| `publication_ready` | All methods + all visualizations | Manuscript preparation |
| `custom` | Manual control of every option | Advanced users |

The `complexityMode` option works similarly but controls UI complexity in the
jamovi interface:

- **quick**: Essential outputs only (5-10 min analysis)
- **standard**: Common validation metrics (30-60 min)
- **comprehensive**: All methods enabled (1-2 hours for large datasets)
- **custom**: Full manual control

---

## 11. Cancer-Type-Specific Analysis

Different cancer types have different expected migration patterns and clinically
meaningful thresholds. Specifying the cancer type adjusts interpretation guidelines.

```{r cancer-type}
stagemigration(
  data = lung_df,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  cancerType = "lung",
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE,
  showClinicalInterpretation = TRUE,
  showExplanations = TRUE
)
```

Supported cancer types: `general`, `lung`, `breast`, `colorectal`, `prostate`,
`headneck`, `melanoma`, `other`. Each adjusts thresholds and interpretation text.

---

## 12. Competing Risks Analysis

In many oncology settings, patients face multiple possible events -- cancer-specific
death and death from other causes. Standard survival analysis treats competing events
as censored, which can bias estimates. The competing risks framework uses Fine-Gray
subdistribution hazard models and Cumulative Incidence Functions (CIF) to properly
handle this.

```{r competing-risks}
# Add a competing event variable for demonstration
cr_data <- combined_data
set.seed(42)
cr_data$event_type <- ifelse(
  cr_data$event == 1,
  sample(c("cancer_death", "other_death"), sum(cr_data$event == 1),
         replace = TRUE, prob = c(0.75, 0.25)),
  "censored"
)
cr_data$event_type <- factor(cr_data$event_type)

stagemigration(
  data = cr_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  performCompetingRisks = TRUE,
  competingEventVar = "event_type",
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE
)
```

The competing risks analysis provides:

- Cumulative incidence functions for each event type
- Fine-Gray subdistribution hazard ratios comparing staging systems
- Gray's test for equality of CIF across stages

---

## 13. Random Survival Forests [Experimental]

For a non-parametric comparison that makes no assumptions about proportional hazards,
the module can fit random survival forests (RSF) to both staging systems and compare
variable importance and prediction accuracy.

```{r random-forest}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  performRandomForestAnalysis = TRUE,
  forestModelType = "rsf",
  forestNTrees = 200,
  calculateVariableImportance = TRUE,
  forestDiscriminationMetrics = TRUE,
  forestStagingComparison = TRUE,
  showMigrationOverview = TRUE
)
```

RSF analysis is computationally expensive. For exploratory work, use `forestNTrees = 200`;
for publications, use 500-1000. The output includes:

- Variable importance rankings for each staging system
- C-index from RSF models vs. Cox models
- Comparison of staging systems in the non-parametric framework

---

## 14. Advanced Features

### Homogeneity and Trend Tests

These tests verify that the staging system satisfies the fundamental requirements:
patients within the same stage should have similar prognosis (homogeneity), and
higher stages should consistently have worse prognosis (monotonic trend).

```{r homogeneity-trend}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  performHomogeneityTests = TRUE,
  performTrendTests = TRUE,
  performLikelihoodTests = TRUE,
  calculatePseudoR2 = TRUE,
  showStatisticalSummary = TRUE,
  showMethodologyNotes = TRUE
)
```

### Multifactorial Analysis

When other prognostic variables (age, grade, biomarkers) are available, the
multifactorial analysis adjusts the staging comparison for these confounders.

```{r multifactorial}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  enableMultifactorialAnalysis = TRUE,
  continuousCovariates = "age",
  categoricalCovariates = "sex",
  multifactorialComparisonType = "comprehensive",
  showMultifactorialTables = TRUE,
  showAdjustedCIndexComparison = TRUE,
  showNestedModelTests = TRUE
)
```

### Time-Dependent ROC Analysis

Compares the discriminative ability of staging systems at specific time points. This
is especially useful when staging performance changes over time (e.g., a staging system
may discriminate well at 1 year but poorly at 5 years).

```{r roc-analysis}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  performROCAnalysis = TRUE,
  rocTimePoints = "12, 24, 60",
  showROCComparison = TRUE
)
```

### RMST and Stage Migration Effect

Restricted Mean Survival Time (RMST) provides a clinically interpretable metric
that does not depend on the proportional hazards assumption. The Stage Migration
Effect (SME) formula quantifies the cumulative survival difference across stages.

```{r rmst-sme}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  calculateRMST = TRUE,
  calculateSME = TRUE,
  showStatisticalComparison = TRUE
)
```

---

## 15. Visualizations

The module produces several publication-quality visualizations:

### Migration Heatmap and Sankey Diagram

```{r heatmap-sankey}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "basic",
  showMigrationHeatmap = TRUE,
  showSankeyDiagram = TRUE,
  showMigrationSurvivalComparison = TRUE
)
```

- **Heatmap**: Color-coded migration matrix; darker = more patients. The diagonal
  shows retention; off-diagonal shows migration flows.
- **Sankey diagram**: Flow visualization where band thickness represents patient count.
  Excellent for presentations.
- **Migration survival comparison**: KM curves showing how survival in each stage
  changes before and after reclassification.

---

## 16. Edge Cases and Small Samples

The module handles small samples gracefully, with appropriate warnings when sample
size is insufficient for certain analyses.

```{r small-sample}
stagemigration(
  data = small_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "basic",
  showMigrationOverview = TRUE,
  showMigrationMatrix = TRUE,
  showExplanations = TRUE
)
```

**Data requirements:**

- Minimum 30 patients (100+ recommended for standard analysis)
- At least 2 stage levels in both systems
- Event rate between 5% and 95%
- For bootstrap: 100+ patients recommended
- For NRI/IDI: adequate events at each time point

---

## 17. Reporting and Interpretation

### Executive Summary and Copy-Ready Reports

```{r executive-summary}
stagemigration(
  data = combined_data,
  oldStage = "old_stage",
  newStage = "new_stage",
  survivalTime = "survival_time",
  event = "event",
  eventLevel = "1",
  analysisType = "comprehensive",
  calculateNRI = TRUE,
  calculateIDI = TRUE,
  performBootstrap = TRUE,
  bootstrapReps = 200,
  generateExecutiveSummary = TRUE,
  generateCopyReadyReport = TRUE,
  showClinicalInterpretation = TRUE,
  showAbbreviationGlossary = TRUE
)
```

The **executive summary** condenses all findings into a structured overview with key
metrics and a staging system adoption recommendation. The **copy-ready report**
generates plain-language paragraphs suitable for direct inclusion in manuscripts or
clinical reports.

### Guided Mode

For users new to staging validation, `enableGuidedMode = TRUE` provides step-by-step
guidance, assumption checking, and clinical interpretation assistance throughout the
analysis.

---

## Complete Option Reference

### Core Variables

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `oldStage` | Variable | -- | Original TNM staging variable (factor) |
| `newStage` | Variable | -- | Revised TNM staging variable (factor) |
| `survivalTime` | Variable | -- | Follow-up time in months (numeric) |
| `event` | Variable | -- | Event status indicator (numeric or factor) |
| `eventLevel` | Level | -- | Level indicating event occurrence |

### Analysis Control

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `analysisType` | List | `comprehensive` | Scope: basic / standard / comprehensive / publication |
| `clinicalPreset` | List | `routine_clinical` | Preset configuration: routine / research / publication / custom |
| `complexityMode` | List | `quick` | UI complexity: quick / standard / comprehensive / custom |
| `confidenceLevel` | Number | `0.95` | Confidence level for CIs and tests (0.80-0.99) |
| `cancerType` | List | `general` | Cancer type for tailored thresholds |

### Statistical Methods

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `calculateNRI` | Bool | `FALSE` | Net Reclassification Improvement |
| `nriTimePoints` | String | `"12, 24, 60"` | Comma-separated NRI time points (months) |
| `calculateIDI` | Bool | `FALSE` | Integrated Discrimination Improvement |
| `performROCAnalysis` | Bool | `FALSE` | Time-dependent ROC analysis |
| `rocTimePoints` | String | `"12, 24, 36, 60"` | Comma-separated ROC time points |
| `performDCA` | Bool | `FALSE` | Decision Curve Analysis |
| `performCalibration` | Bool | `FALSE` | Calibration analysis |
| `performHomogeneityTests` | Bool | `FALSE` | Within-stage homogeneity tests |
| `performTrendTests` | Bool | `FALSE` | Monotonic trend across stages |
| `performLikelihoodTests` | Bool | `FALSE` | Likelihood ratio tests |
| `calculatePseudoR2` | Bool | `FALSE` | Pseudo R-squared measures |
| `calculateSME` | Bool | `FALSE` | Stage Migration Effect formula |
| `calculateRMST` | Bool | `FALSE` | Restricted Mean Survival Time |

### Validation

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `performBootstrap` | Bool | `FALSE` | Bootstrap internal validation |
| `bootstrapReps` | Number | `1000` | Bootstrap repetitions (100-2000) |
| `performCrossValidation` | Bool | `FALSE` | k-fold cross-validation |
| `cvFolds` | Number | `5` | Number of CV folds (3-10) |
| `useOptimismCorrection` | Bool | `FALSE` | Apply optimism correction |

### Clinical Thresholds

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `clinicalSignificanceThreshold` | Number | `0.02` | Minimum C-index improvement considered clinically significant |
| `nriClinicalThreshold` | Number | `0.20` | Minimum NRI for clinical meaningfulness |

### Table Display

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `showMigrationOverview` | Bool | `TRUE` | Overview table with key migration statistics |
| `showMigrationSummary` | Bool | `FALSE` | Statistical summary with Chi-square / Fisher tests |
| `showStageDistribution` | Bool | `FALSE` | Side-by-side stage distribution comparison |
| `showMigrationMatrix` | Bool | `TRUE` | Detailed cross-tabulation matrix |
| `showStatisticalComparison` | Bool | `FALSE` | C-index and other statistical metrics |
| `showConcordanceComparison` | Bool | `FALSE` | Detailed concordance comparison |
| `showWillRogersAnalysis` | Bool | `FALSE` | Will Rogers phenomenon analysis |
| `showClinicalInterpretation` | Bool | `FALSE` | Clinical interpretation guide |
| `showStatisticalSummary` | Bool | `FALSE` | Comprehensive statistical summary |
| `showMethodologyNotes` | Bool | `FALSE` | Detailed methodology documentation |
| `showExplanations` | Bool | `TRUE` | Explanatory text for results |
| `showAbbreviationGlossary` | Bool | `FALSE` | Glossary of abbreviations |
| `includeEffectSizes` | Bool | `FALSE` | Effect sizes for comparisons |
| `generateExecutiveSummary` | Bool | `FALSE` | Key findings and recommendations |

### Visualization

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `showMigrationHeatmap` | Bool | `FALSE` | Color-coded migration heatmap |
| `showSankeyDiagram` | Bool | `FALSE` | Patient flow diagram |
| `showROCComparison` | Bool | `FALSE` | Time-dependent ROC curves |
| `showCalibrationPlots` | Bool | `FALSE` | Calibration plots |
| `showDecisionCurves` | Bool | `FALSE` | Decision curve plots |
| `showForestPlot` | Bool | `FALSE` | Hazard ratio forest plot |
| `showWillRogersVisualization` | Bool | `FALSE` | Will Rogers effect visualization |
| `showMigrationSurvivalComparison` | Bool | `FALSE` | Before/after survival curves |
| `showSurvivalCurves` | Bool | `FALSE` | Kaplan-Meier survival curves |
| `survivalPlotType` | List | `separate` | Plot layout: separate / sidebyside / overlay |
| `showConfidenceIntervals` | Bool | `FALSE` | CIs on survival curves |
| `showRiskTables` | Bool | `FALSE` | At-risk tables below curves |
| `plotTimeRange` | String | `"auto"` | Maximum time for plots (months or "auto") |

### Multifactorial Analysis

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `enableMultifactorialAnalysis` | Bool | `FALSE` | Enable adjusted comparisons |
| `continuousCovariates` | Variables | `NULL` | Continuous covariates (e.g., age) |
| `categoricalCovariates` | Variables | `NULL` | Categorical covariates (e.g., sex) |
| `multifactorialComparisonType` | List | `comprehensive` | adjusted_cindex / nested_models / stepwise / comprehensive |
| `baselineModel` | List | `covariates_only` | Reference model for comparison |
| `performInteractionTests` | Bool | `FALSE` | Test stage-covariate interactions |
| `stratifiedAnalysis` | Bool | `FALSE` | Stratified subgroup analysis |

### Competing Risks

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `performCompetingRisks` | Bool | `FALSE` | Enable competing risks analysis |
| `competingEventVar` | Variable | `NULL` | Competing event indicator variable |
| `performCompetingRisksAdvanced` | Bool | `FALSE` | Advanced Fine-Gray analysis |
| `competingRisksMethod` | List | `comprehensive` | finegray / causespecific / comprehensive |

### Random Survival Forest

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `performRandomForestAnalysis` | Bool | `FALSE` | Enable RSF analysis |
| `forestModelType` | List | `rsf` | Forest model type |
| `forestNTrees` | Number | `500` | Number of trees (100-5000) |
| `calculateVariableImportance` | Bool | `FALSE` | Variable importance rankings |
| `forestDiscriminationMetrics` | Bool | `FALSE` | RSF-based C-index |

### User Experience

| Option | Type | Default | Description |
|--------|------|---------|-------------|
| `enableGuidedMode` | Bool | `FALSE` | Step-by-step analysis guidance |
| `generateCopyReadyReport` | Bool | `FALSE` | Manuscript-ready text output |
| `enableAccessibilityFeatures` | Bool | `FALSE` | Color-blind safe palettes |
| `preferredLanguage` | List | `en` | Output language: en / tr |
| `optimizeForLargeDatasets` | Bool | `FALSE` | Memory-efficient processing for N > 10,000 |

---

## Recommended Workflow

For a complete staging validation study, we recommend:

1. **Start with the basics**: Run `analysisType = "basic"` to examine migration
   patterns, migration rates, and the cross-tabulation matrix
2. **Assess discrimination**: Enable C-index comparison and NRI/IDI to quantify
   improvement
3. **Check for Will Rogers**: Always check for the Will Rogers phenomenon before
   claiming the new system is superior
4. **Validate internally**: Use bootstrap validation with optimism correction to
   confirm findings are not over-optimistic
5. **Assess clinical utility**: Decision curve analysis determines whether improved
   discrimination translates to better clinical decisions
6. **Generate the report**: Use `generateExecutiveSummary` and
   `generateCopyReadyReport` to produce manuscript-ready output

Or simply use `clinicalPreset = "publication_ready"` to enable all of the above in
one step.

---

## References

1. Pencina MJ, D'Agostino RB Sr, D'Agostino RB Jr, Vasan RS. Evaluating the added
   predictive ability of a new marker: from area under the ROC curve to
   reclassification and beyond. *Stat Med*. 2008;27(2):157-172.

2. Feinstein AR, Sosin DM, Wells CK. The Will Rogers phenomenon. Stage migration
   and new diagnostic techniques as a source of misleading statistics for survival
   in cancer. *N Engl J Med*. 1985;312(25):1604-1608.

3. Vickers AJ, Elkin EB. Decision curve analysis: a novel method for evaluating
   prediction models. *Med Decis Making*. 2006;26(6):565-574.

4. Harrell FE Jr, Lee KL, Mark DB. Multivariable prognostic models: issues in
   developing models, evaluating assumptions and adequacy, and measuring and
   reducing errors. *Stat Med*. 1996;15(4):361-387.

5. Amin MB, Edge SB, Greene FL, et al., eds. *AJCC Cancer Staging Manual*. 8th ed.
   Springer; 2017.

6. Fine JP, Gray RJ. A proportional hazards model for the subdistribution of a
   competing risk. *J Am Stat Assoc*. 1999;94(446):496-509.

7. Royston P, Altman DG. External validation of a Cox prognostic model: principles
   and methods. *BMC Med Res Methodol*. 2013;13:33.
