---
title: "Cure Models for Long-Term Survivors - Comprehensive Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cure Models for Long-Term Survivors - Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all(".")
```

> **Note:** The `curemodels()` function is designed for use within **jamovi's GUI**.
> The code examples below show the R syntax for reference. To run interactively,
> use `devtools::load_all()` and call the R6 class directly:
> `curemodelsClass$new(options = curemodelsOptions$new(...), data = mydata)`.

# Cure Models for Long-Term Survivors

## Overview

In many solid tumours, a proportion of patients experience long-term disease-free
survival that is effectively indistinguishable from the general population. Standard
survival models (Cox, Kaplan-Meier) assume every patient will eventually experience
the event, which misrepresents prognosis for cancers with a curable subset. Cure
models address this by partitioning the population into two groups: patients who are
statistically cured and patients who remain at risk (uncured). This yields two
clinically actionable quantities: the **cure fraction** (probability of long-term
cure) and the **latency distribution** (survival among uncured patients).

The `curemodels` module provides four complementary modelling frameworks:

| Model | R Package | Approach | Best For |
|-------|-----------|----------|----------|
| Mixture cure | `smcure` | Separate logistic (cure) + survival (latency) | Standard clinical cure rate estimation |
| Non-mixture cure | `flexsurvcure` | Cure fraction embedded in parametric survival | Parametric model comparison across distributions |
| Background-mortality cure | `cuRe` | Excess hazard above population mortality | Long follow-up where background mortality matters |
| Nonparametric cure | `npcure` | Kernel smoothing, no distributional assumptions | Exploratory analysis without parametric constraints |

---

## Dataset

```{r data-load}
data("curemodels_test")
```

```{r data-overview}
str(curemodels_test)
```

```{r data-summary}
summary(curemodels_test[, c("FollowUpMonths", "Recurrence", "Age",
                             "Treatment", "Stage", "TumorSize")])
```

| Variable | Type | Description |
|----------|------|-------------|
| PatientID | Integer | Unique patient identifier |
| FollowUpMonths | Numeric | Follow-up time in months |
| Recurrence | Numeric (0/1) | Event indicator: 0=censored, 1=recurrence |
| RecurrenceFactor | Factor | Factor version of Recurrence |
| Age | Numeric | Patient age at diagnosis |
| Sex | Factor | Male / Female |
| Treatment | Factor | Surgery Only / Surgery+Adjuvant |
| Stage | Factor | I / II / III |
| TumorSize | Numeric | Tumor size in cm |
| PerformanceStatus | Integer | ECOG PS: 0, 1, or 2 |
| BackgroundHazard | Numeric | Expected population hazard rate |

```{r data-event-summary}
cat("Total patients:", nrow(curemodels_test), "\n")
cat("Events (recurrences):", sum(curemodels_test$Recurrence == 1), "\n")
cat("Censored:", sum(curemodels_test$Recurrence == 0), "\n")
cat("Event rate:", round(mean(curemodels_test$Recurrence) * 100, 1), "%\n")
table(curemodels_test$Treatment)
table(curemodels_test$Stage)
```

---

## 1. Package Availability Check

The cure models depend on several specialised R packages. Before running the
examples, verify they are available.

```{r check-packages}
required_pkgs <- c("smcure", "flexsurvcure", "cuRe", "npcure")
available <- sapply(required_pkgs, function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
})
cat("Package availability:\n")
for (i in seq_along(required_pkgs)) {
  cat(sprintf("  %s: %s\n", required_pkgs[i],
              ifelse(available[i], "installed", "NOT FOUND")))
}
```

---

## 2. Mixture Cure Model (smcure)

### 2.1 Basic Mixture Cure Model

The simplest case: estimate the overall cure fraction without covariates. The
mixture cure model decomposes the survival function as:

S(t) = pi + (1 - pi) * S_u(t)

where pi is the cure fraction and S_u(t) is the survival function of uncured patients.

```{r mixture-basic}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture"
)
```

### 2.2 Mixture Cure Model with Predictors (PH)

Adding covariates allows estimation of how treatment and stage affect both the
probability of cure (incidence component) and survival among uncured patients
(latency component).

```{r mixture-predictors-ph}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = c("Treatment", "Stage"),
  model_type = "mixture",
  smcure_model_type = "ph",
  cure_link = "logit"
)
```

The `modelTable` shows two sets of coefficients:

- **Cure component** (`$b`): logistic regression coefficients for cure probability.
  Positive values indicate higher cure probability.
- **Survival component** (`$beta`): Cox PH coefficients for uncured patients.
  Positive values indicate higher hazard (worse survival).

### 2.3 Mixture Cure Model with AFT

The accelerated failure time (AFT) formulation models the latency distribution
on the log-time scale rather than the hazard scale.

```{r mixture-aft}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = c("Treatment", "Stage"),
  model_type = "mixture",
  smcure_model_type = "aft"
)
```

In the AFT model, survival coefficients represent acceleration factors:
positive values mean longer survival (deceleration of time-to-event).

---

## 3. Non-Mixture Cure Model (flexsurvcure)

### 3.1 Weibull Distribution

The non-mixture cure model incorporates the cure fraction directly into the
survival function rather than as a separate mixture component. The Weibull
distribution is the most flexible standard parametric form.

```{r nonmixture-weibull}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "nonmixture",
  survival_dist = "weibull"
)
```

### 3.2 Distribution Comparison

Different parametric distributions make different assumptions about the hazard
shape. Comparing them helps identify the best-fitting latency model.

```{r nonmixture-exponential}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "nonmixture",
  survival_dist = "exponential"
)
```

```{r nonmixture-lognormal}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "nonmixture",
  survival_dist = "lognormal"
)
```

```{r nonmixture-loglogistic}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "nonmixture",
  survival_dist = "loglogistic"
)
```

| Distribution | Hazard Shape | When to Use |
|-------------|-------------|-------------|
| Weibull | Monotone increasing or decreasing | Default; most flexible standard choice |
| Exponential | Constant | When hazard is stable over time |
| Lognormal | Non-monotone (rises then falls) | When risk peaks then declines |
| Log-logistic | Non-monotone (rises then falls) | Similar to lognormal, proportional odds model |

---

## 4. Bootstrap Confidence Intervals

Point estimates of the cure fraction are useful, but clinical decisions require
uncertainty quantification. Bootstrap resampling provides confidence intervals
for the cure fraction and other parameters.

```{r bootstrap-ci}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "mixture",
  bootstrap_ci = TRUE,
  n_bootstrap = 200
)
```

For the `smcure` package, bootstrap is performed internally via `Var=TRUE`. For
`flexsurvcure` and `cuRe`, bootstrap is implemented by resampling the data and
refitting the model. Use `n_bootstrap = 1000` for publication-quality results;
`n_bootstrap = 100-200` is sufficient for exploratory analysis.

---

## 5. Cure Threshold and Sensitivity Analysis

### 5.1 Custom Cure Threshold

The cure threshold defines the minimum follow-up time (in months) beyond which
surviving patients are considered potentially cured. The default is 60 months
(5 years), which is standard for many solid tumours.

```{r cure-threshold-default}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture",
  cure_threshold = 60
)
```

For cancers with later recurrence patterns (e.g., breast cancer ER+), a longer
threshold may be appropriate:

```{r cure-threshold-long}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture",
  cure_threshold = 120
)
```

### 5.2 Sensitivity Analysis

Sensitivity analysis systematically varies the cure threshold to assess whether
the estimated cure fraction is robust or sensitive to this assumption.

```{r sensitivity-analysis}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture",
  sensitivity_analysis = TRUE,
  cure_threshold = 60
)
```

A stable cure fraction across thresholds (e.g., 36 to 84 months) increases
confidence in the estimate. Large variation suggests the cure plateau has not
been reached and longer follow-up may be needed.

---

## 6. Plots

### 6.1 Cure Fraction Plot

A bar chart showing estimated cure fractions by predictor group. Uses a
colorblind-safe palette (blue/orange).

```{r plot-cure-fraction}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "mixture",
  plot_cure_fraction = TRUE
)
```

When `bootstrap_ci = TRUE`, error bars are added to show the confidence interval
for each group's cure fraction.

```{r plot-cure-fraction-bootstrap}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "mixture",
  plot_cure_fraction = TRUE,
  bootstrap_ci = TRUE,
  n_bootstrap = 200
)
```

### 6.2 Survival Curves

Survival curves overlay the Kaplan-Meier estimate with the fitted cure model
curve, showing how well the model captures the observed survival pattern.

```{r plot-survival}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "mixture",
  plot_survival = TRUE
)
```

The fitted curve should approach the estimated cure fraction asymptotically:
it levels off rather than dropping to zero, reflecting the cured subpopulation.

---

## 7. Goodness of Fit

Goodness-of-fit tests assess whether the cure model provides a significantly
better fit than a standard survival model (which assumes no cure fraction).

```{r goodness-of-fit}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "mixture",
  goodness_of_fit = TRUE
)
```

The `goodnessOfFit` table reports:

- **Likelihood ratio test**: Compares the cure model to a nested model without
  the cure component. A significant p-value supports the existence of a cure
  fraction.
- **AIC/BIC**: Lower values indicate better fit, penalised for model complexity.
- **Interpretation**: Plain-language summary of the test outcome.

---

## 8. Model Comparison (All Types)

Running all four model types simultaneously provides a comprehensive comparison.
The `modelComparison` table ranks parametric models by AIC/BIC (npcure is
nonparametric and excluded from information criteria).

```{r model-comparison-all}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "all",
  npcure_covariate = "Age",
  npcure_bandwidth = "auto",
  npcure_time_points = 100,
  use_background_mortality = TRUE,
  background_hazard_var = "BackgroundHazard",
  plot_cure_fraction = TRUE,
  plot_survival = TRUE,
  goodness_of_fit = TRUE
)
```

### Interpreting the Comparison

- **Lowest AIC**: Best predictive model (balances fit and complexity)
- **Lowest BIC**: Most parsimonious model (stronger penalty for complexity)
- **Cure fractions**: Should be broadly consistent across models if the cure
  assumption is appropriate; large discrepancies suggest model misspecification
- **npcure**: Provides a nonparametric benchmark; if parametric models diverge
  substantially from npcure, the parametric assumptions may be wrong

---

## 9. Edge Cases

### 9.1 No Predictors (Intercept-Only Model)

```{r edge-no-predictors}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture"
)
```

### 9.2 cuRe Model Without Background Mortality

```{r edge-cure-no-background}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = "Treatment",
  model_type = "cure",
  use_background_mortality = FALSE
)
```

### 9.3 cuRe Model with Missing Background Hazard Variable

```{r edge-cure-missing-bhazard}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "cure",
  use_background_mortality = TRUE
)
```

This should produce an informative error about requiring the `background_hazard_var`.

### 9.4 npcure Without Covariate

```{r edge-npcure-no-covariate}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "npcure"
)
```

This should produce an informative error about requiring `npcure_covariate`.

### 9.5 Small Sample

```{r edge-small-sample}
small_data <- curemodels_test[1:20, ]
curemodels(
  data = small_data,
  time = "FollowUpMonths",
  status = "Recurrence",
  model_type = "mixture"
)
```

Expect warnings about small sample size or potential convergence issues.

---

## 10. Clinical Interpretation

The `interpretation` HTML output provides automatically generated clinical
context for the cure model results. It includes:

1. A statement of the estimated cure fraction with confidence interval (when
   bootstrap is enabled)
2. The median survival of uncured patients
3. Context based on the cure threshold
4. A copy-ready sentence for the Results section of a manuscript
5. Clinical recommendations for patient counselling

```{r clinical-interpretation}
curemodels(
  data = curemodels_test,
  time = "FollowUpMonths",
  status = "Recurrence",
  predictors = c("Treatment", "Stage"),
  model_type = "mixture",
  smcure_model_type = "ph",
  cure_link = "logit",
  cure_threshold = 60,
  bootstrap_ci = TRUE,
  n_bootstrap = 200,
  plot_cure_fraction = TRUE,
  plot_survival = TRUE,
  goodness_of_fit = TRUE,
  sensitivity_analysis = TRUE
)
```

---

## 11. References

1. Berkson J, Gage RP. Survival curve for cancer patients following treatment.
   *J Am Stat Assoc*. 1952;47(259):501-515.

2. Boag JW. Maximum likelihood estimates of the proportion of patients cured
   by cancer therapy. *J R Stat Soc Series B*. 1949;11(1):15-53.

3. Cai C, Zou Y, Peng Y, Zhang J. smcure: An R-package for estimating
   semiparametric mixture cure models. *Comput Methods Programs Biomed*.
   2012;108(3):1255-1260.

4. Peng Y, Dear KBG. A nonparametric mixture model for cure rate estimation.
   *Biometrics*. 2000;56(1):237-243.

5. Andersson TML, Dickman PW, Eloranta S, Lambert PC. Estimating and modelling
   cure in population-based cancer studies within the framework of flexible
   parametric survival models. *BMC Med Res Methodol*. 2011;11:96.

6. Lopez-Cheda A, Cao R, Jacome MA, Van Keilegom I. Nonparametric incidence
   estimation and bootstrap bandwidth selection in mixture cure models.
   *Comput Stat Data Anal*. 2017;105:144-165.

7. Lambert PC, Thompson JR, Weston CL, Dickman PW. Estimating and modeling
   the cure fraction in population-based cancer survival analysis.
   *Biostatistics*. 2007;8(3):576-594.

8. Yu B, Tiwari RC, Cronin KA, Feuer EJ. Cure fraction estimation from the
   mixture cure models for grouped survival data. *Stat Med*.
   2004;23(11):1733-1747.

9. De Angelis R, Capocaccia R, Hakulinen T, Soderman B, Verdecchia A.
   Mixture models for cancer survival analysis: application to
   population-based data with covariates. *Stat Med*. 1999;18(4):441-454.

10. Sy JP, Taylor JMG. Estimation in a Cox proportional hazards cure model.
    *Biometrics*. 2000;56(1):227-236.
