---
title: "Quick Reference Guide for jjstatsplot"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Quick Reference Guide for jjstatsplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, eval = FALSE)
```

# jjstatsplot Quick Reference

This quick reference provides essential information for using jjstatsplot efficiently in both jamovi and R.

## Function Overview

| Function | Purpose | Data Type | jamovi Location |
|----------|---------|-----------|-----------------|
| `jjhistostats()` | Distribution analysis | 1 continuous | Continuous → Histogram |
| `jjscatterstats()` | Bivariate relationships | 2 continuous | Continuous vs Continuous → Scatter Plot |
| `jjcorrmat()` | Multiple correlations | 2+ continuous | Continuous vs Continuous → Correlation Matrix |
| `jjbetweenstats()` | Group comparisons | 1 continuous + 1 categorical | Categorical vs Continuous → Box-Violin (Between) |
| `jjdotplotstats()` | Group means | 1 continuous + 1 categorical | Categorical vs Continuous → Dot Chart |
| `jjwithinstats()` | Paired comparisons | 2+ continuous (paired) | Categorical vs Continuous → Box-Violin (Within) |
| `jjbarstats()` | Categorical associations | 1-2 categorical | Categorical vs Categorical → Bar Charts |
| `jjpiestats()` | Proportional data | 1 categorical | Categorical vs Categorical → Pie Charts |
| `waffle()` | Distribution visualization | 1 categorical | Distribution → Waffle Charts |

## Quick Start Templates

### jamovi Workflow
```
1. Data → Open file
2. JJStatsPlot → [Category] → [Analysis]
3. Drag variables to appropriate boxes
4. Adjust options as needed
5. Export → Copy/Save plot
```

### R Workflow
```{r r-workflow}
library(ClinicoPath)

# Basic pattern
result <- jj[function](data = your_data, dep = "variable")
result$plot  # Main plot

# With grouping
result <- jj[function](data = your_data, dep = "variable", grvar = "group")
result$plot2  # Grouped plots
```

## Essential R Code Snippets

### Data Preparation
```{r data-prep}
# Load and check data
library(ClinicoPath)
data(mtcars)
str(mtcars)

# Convert to factors
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am, labels = c("Auto", "Manual"))

# Check missing data
summary(mtcars)
```

### Single Variable Analysis
```{r single-var}
# Histogram
hist_result <- jjhistostats(data = mtcars, dep = "mpg")
hist_result$plot

# Multiple variables
multi_hist <- jjhistostats(data = mtcars, dep = c("mpg", "hp", "wt"))
multi_hist$plot
```

### Bivariate Analysis
```{r bivariate}
# Scatter plot
scatter_result <- jjscatterstats(data = mtcars, dep = "mpg", group = "hp")
scatter_result$plot

# Correlation matrix
corr_result <- jjcorrmat(data = mtcars, dep = c("mpg", "hp", "wt"))
corr_result$plot
```

### Group Comparisons
```{r group-comparisons}
# Between groups
between_result <- jjbetweenstats(data = mtcars, dep = "mpg", group = "cyl")
between_result$plot

# Dot chart
dot_result <- jjdotplotstats(data = mtcars, dep = "mpg", group = "cyl")
dot_result$plot
```

### Categorical Analysis
```{r categorical}
# Bar chart
bar_result <- jjbarstats(data = mtcars, dep = "cyl", group = "am")
bar_result$plot

# Pie chart
pie_result <- jjpiestats(data = mtcars, dep = "cyl")
pie_result$plot
```

### Grouped Analysis
```{r grouped}
# Any function with grvar parameter
grouped_result <- jjhistostats(
  data = mtcars, 
  dep = "mpg", 
  grvar = "cyl"
)
grouped_result$plot2  # Note: plot2 for grouped plots
```

## Common Options and Customizations

### jamovi Interface Elements
- **Dependent Variable(s)**: Main variables to analyze
- **Grouping Variable**: Split data by categories
- **Grouping Variable for Plots**: Create separate plots
- **Statistical Options**: Test types, confidence levels
- **Plot Options**: Themes, colors, labels

### R Function Parameters
```{r parameters}
# Common parameters across functions
jj[function](
  data = your_data,           # Data frame
  dep = "variable",           # Dependent variable(s)
  group = "grouping_var",     # Grouping variable
  grvar = "plot_grouping",    # Separate plots by group
  # Function-specific options vary
)
```

## Statistical Tests by Function

| Function | Default Test | Alternative Tests |
|----------|--------------|-------------------|
| `jjhistostats()` | Shapiro-Wilk normality | Anderson-Darling, Kolmogorov-Smirnov |
| `jjscatterstats()` | Pearson correlation | Spearman, robust correlation |
| `jjbetweenstats()` | Welch's ANOVA | Student's t-test, Kruskal-Wallis |
| `jjdotplotstats()` | Similar to between-stats | Non-parametric alternatives |
| `jjwithinstats()` | Paired t-test | Wilcoxon signed-rank |
| `jjbarstats()` | Chi-square test | Fisher's exact test |
| `jjcorrmat()` | Pearson correlations | Spearman, partial correlations |

## Troubleshooting Checklist

### Before Analysis
- [ ] Check variable types (`str(data)`)
- [ ] Verify sufficient sample sizes (`table(group_var)`)
- [ ] Handle missing values (`summary(data)`)
- [ ] Ensure appropriate data format

### Common Errors
- **"No data to plot"**: Check variable selection and missing values
- **"Object not found"**: Verify variable names (case-sensitive)
- **Blank plots**: Use `result$plot2` for grouped analyses
- **Statistical test failures**: Check assumptions and sample sizes

### Performance Issues
- Large datasets: Sample or subset data
- Complex grouping: Reduce number of groups
- Memory problems: Clean workspace with `rm()` and `gc()`

## Export and Reporting

### jamovi Export
```
Right-click plot → Copy
File → Export → [Format]
Results panel → Copy statistical output
```

### R Export
```{r export}
# Save plot
result <- jjhistostats(data = mtcars, dep = "mpg")
ggsave("histogram.png", result$plot, width = 8, height = 6, dpi = 300)

# Extract for further modification
my_plot <- result$plot + 
  ggplot2::labs(title = "Custom Title") +
  ggplot2::theme_minimal()
```

## Statistical Reporting Template

When reporting results, include:

1. **Sample size**: n = X
2. **Test type**: "Welch's ANOVA was conducted..."
3. **Test statistic**: F(df1, df2) = X.XX
4. **p-value**: p < .001 or exact value
5. **Effect size**: η² = .XX or Cohen's d = X.XX
6. **Confidence intervals**: 95% CI [X.XX, X.XX]

Example: "A Welch's ANOVA revealed a significant difference in fuel efficiency across cylinder groups, F(2, 18.03) = 39.70, p < .001, η² = .70, suggesting a large effect."

## Documentation Links

- **Complete jamovi Guide**: `vignette("jamovi-user-guide", package = "jjstatsplot")`
- **R Programming Guide**: `vignette("r-programming-guide", package = "jjstatsplot")`
- **Analysis Gallery**: `vignette("analysis-gallery", package = "jjstatsplot")`
- **Troubleshooting**: `vignette("troubleshooting-faq", package = "jjstatsplot")`
- **ggstatsplot Documentation**: [indrajeetpatil.github.io/ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/)

## Getting Help

### Quick Checks
1. Try with built-in data (e.g., `mtcars`, `iris`)
2. Simplify analysis (remove grouping variables)
3. Check for recent updates
4. Restart jamovi/R session

### Report Issues
- Include data example, error message, and steps
- GitHub: [sbalci/ClinicoPathJamoviModule/issues](https://github.com/sbalci/ClinicoPathJamoviModule/issues)
- Email: drserdarbalci@gmail.com

This quick reference covers the essentials for efficient jjstatsplot usage. For detailed examples and advanced usage, consult the comprehensive vignettes listed above.