---
title: "ROC Curve Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ROC Curve Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette shows how to perform Receiver Operating Characteristic
(ROC) analysis with **meddecide** using the `psychopdaroc()` function.

## Loading the Data

```{r, eval=FALSE}
df_roc <- read.csv(system.file("extdata", "roc_example.csv", package = "meddecide"))
head(df_roc)
```

## Creating the ROC Curve

```{r, eval=FALSE}
roc_res <- psychopdaroc(data = df_roc, class = df_roc$class, value = df_roc$value)
roc_res$plot
```

The resulting plot shows the ROC curve along with the area under the curve (AUC).
You can extract the AUC value and other statistics from the result object.

```{r, eval=FALSE}
roc_res$AUC
```

## Bootstrapping and Confidence Intervals

`auc_ci()` and `bootstrap_ci()` provide convenience wrappers for computing
confidence intervals for diagnostic metrics.

```{r, eval=FALSE}
auc_ci(roc_res)
bootstrap_ci(roc_res)
```

These utilities allow you to assess the stability of your ROC analysis.
