---
title: "Use ClinicoPathDescriptives in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use ClinicoPathDescriptives in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.height = 4,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```




```{r data histopathology}
data("histopathology")
```

```{r}
#| glimpse histopathology

dplyr::glimpse(histopathology)
```



```{r}
ClinicoPath::agepyramid(
    data = histopathology,
    age = Age,
    gender = Sex,
    female = "Female")
```



```{r}
age_pyramid <-
ClinicoPath::agepyramid(
    data = histopathology,
    age = Age,
    gender = Sex,
    female = "Female")
```



```{r}
age_pyramid$pyramidTable
```


```{r}
age_pyramid$pyramidTable$asDF
```


```{r}
age_pyramid$plot
```



```{r}
age_pyramid$plot$state
```



```{r}
age_pyramid$plot$plot
```


```{r}
temp_file_png <- tempfile(fileext = ".png")
age_pyramid$plot$saveAs(paste0(temp_file_png, ".png"))

age_pyramid_plot <- png::readPNG(paste0(temp_file_png, ".png"))

temp_file_tiff <- tempfile(fileext = ".tiff")

tiff::writeTIFF(what = age_pyramid_plot,
                where = paste0(temp_file_tiff, ".tiff"),
                compression = "LZW")
```




```{r}
htmlTable::htmlTable(
  kableExtra::kable(
    age_pyramid$pyramidTable$asDF
    ))
```



Age Pyramid

agepyramidClass
	

Age Pyramid

alluvial()
	

Alluvial Diagrams

alluvialClass
	

Alluvial Plot

benford()
	

Benford Analysis

crosstable()
	

Cross Tables

crosstableClass
	

Cross Table

histopathology
	

histopathology

reportcat()
	

Summary of Categorical Variables

reportcatClass
	

Summary of Categorical Variables

summarydata()
	

Summary of Continuous Variables

summarydataClass
	

Summary of Continuous Variables

tableone()
	

Table One

tableoneClass
	

Table One

venn()
	

Venn Diagram

vennClass 
