---
title: "Use jjstatsplot in R"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Use jjstatsplot in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```


All functions

jjbarstats()
	

Bar Charts

jjbarstatsClass
	

Bar Charts

jjbetweenstats()
	

Box-Violin Plots to Compare Between Groups

jjbetweenstatsClass
	

Violin Plots to Compare Between Groups

jjcorrmat()
	

Correlation Matrix

jjcorrmatClass
	

Correlation Matrix

jjdotplotstats()
	

Dot Chart

jjdotplotstatsClass
	

Dot Chart

jjhistostats()
	

Histogram

jjhistostatsClass
	

Histogram

jjpiestats()
	

Pie Charts

jjpiestatsClass
	

Pie Charts

jjscatterstats()
	

Scatter Plot

jjscatterstatsClass
	

Scatter Plot

