---
title: "Use meddecide in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use meddecide in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ClinicoPath)
```


agreement()
	

Interrater Reliability

agreementClass
	

Interrater Reliability Analysis

decision()
	

Medical Decision

decisionClass
	

Medical Decision Making

decisioncalculator()
	

Medical Decision Calculator

decisioncalculatorClass
	

Decision Calculator

histopathology
	

histopathology

kappaSizeCI()
	

Confidence Interval Approach for the Number of Subjects Required

kappaSizeCIClass
	

Confidence Interval Approach for the Number of Subjects Required

kappaSizeFixedN()
	

Lowest Expected Value for a fixed sample size

kappaSizeFixedNClass
	

Lowest Expected Value for a fixed sample size

kappaSizePower()
	

Power Approach for the Number of Subjects Required

kappaSizePowerClass 

