---
title: "Visualization Gallery for Clinical Research"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Visualization Gallery for Clinical Research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Effective data visualization is crucial for communicating clinical research findings. This vignette showcases the comprehensive visualization capabilities of ClinicoPathDescriptives, demonstrating how to create publication-ready figures for various types of clinical data analysis.

## Visualization Philosophy

### Design Principles

1. **Clarity**: Clear, unambiguous representation of data
2. **Accuracy**: Truthful representation without distortion
3. **Efficiency**: Maximum information with minimum ink
4. **Aesthetics**: Professional appearance suitable for publication
5. **Context**: Appropriate for clinical research audience

```{r setup}
library(ClinicoPath)
library(dplyr)
library(ggplot2)
library(knitr)

# Load datasets
data(histopathology)
data(treatmentResponse)

# Prepare enhanced dataset for visualization examples
viz_data <- histopathology %>%
  mutate(
    Age_Group = case_when(
      Age < 40 ~ "< 40 years",
      Age >= 40 & Age < 60 ~ "40-59 years", 
      Age >= 60 ~ "≥ 60 years"
    ),
    Outcome_Status = case_when(
      Outcome == 1 ~ "Event Occurred",
      Outcome == 0 ~ "Event-Free",
      TRUE ~ "Unknown"
    ),
    Risk_Profile = case_when(
      Grade == "3" & LymphNodeMetastasis == "Present" ~ "High Risk",
      Grade == "3" | LymphNodeMetastasis == "Present" ~ "Intermediate Risk",
      TRUE ~ "Low Risk"
    ),
    Treatment_Response = case_when(
      Group == "Treatment" & Outcome == 0 ~ "Treatment Success",
      Group == "Treatment" & Outcome == 1 ~ "Treatment Failure", 
      Group == "Control" & Outcome == 0 ~ "Control Success",
      TRUE ~ "Control Failure"
    )
  )

cat("Enhanced dataset prepared with", nrow(viz_data), "patients and", ncol(viz_data), "variables\n")
```

## Section 1: Demographic Visualizations

### Age Pyramid - Population Structure

```{r}
# Classic age pyramid showing population structure
agepyramid(
  data = viz_data,
  age = "Age",
  sex = "Sex",
  title = "Study Population Age and Sex Distribution",
  subtitle = "Demographic profile of enrolled patients (N=250)"
)
```

### Age Pyramid by Treatment Group

```{r}
# Age pyramid comparing treatment groups
agepyramid(
  data = viz_data,
  age = "Age", 
  sex = "Sex",
  group = "Group",
  title = "Age and Sex Distribution by Treatment Group",
  subtitle = "Comparison of demographic balance between study arms"
)
```

### Age Pyramid by Risk Profile

```{r}
# Age pyramid by risk stratification
agepyramid(
  data = viz_data,
  age = "Age",
  sex = "Sex",
  group = "Risk_Profile", 
  title = "Demographics by Risk Stratification",
  subtitle = "Age and sex distribution across risk categories"
)
```

## Section 2: Categorical Relationship Visualizations

### Alluvial Diagrams - Treatment Pathway Analysis

```{r}
# Basic treatment pathway flow
alluvial(
  data = viz_data,
  vars = c("Group", "Grade", "LymphNodeMetastasis", "Outcome_Status"),
  title = "Treatment Pathway and Outcome Flow",
  subtitle = "Patient flow from treatment assignment to outcome"
)
```

### Complex Pathological Progression

```{r}
# Detailed pathological progression analysis
alluvial(
  data = viz_data,
  vars = c("Grade", "TStage", "LVI", "PNI", "LymphNodeMetastasis"),
  title = "Pathological Feature Progression",
  subtitle = "Relationships between tumor characteristics"
)
```

### Treatment Response Flow

```{r}
# Treatment response pathway
alluvial(
  data = viz_data,
  vars = c("Age_Group", "Risk_Profile", "Group", "Treatment_Response"),
  title = "Age, Risk, and Treatment Response Relationships", 
  subtitle = "Multi-dimensional treatment outcome analysis"
)
```

## Section 3: Set Relationship Visualizations

### Venn Diagrams - Biomarker Overlap

```{r}
# Prepare binary indicators for Venn analysis
venn_data <- viz_data %>%
  mutate(
    LVI_Present = ifelse(LVI == "Present", 1, 0),
    PNI_Present = ifelse(PNI == "Present", 1, 0),
    LN_Positive = ifelse(LymphNodeMetastasis == "Present", 1, 0),
    High_Grade = ifelse(Grade == "3", 1, 0)
  )

# Three-way Venn diagram of invasion markers
venn(
  data = venn_data,
  vars = c("LVI_Present", "PNI_Present", "LN_Positive"),
  names = c("Lymphovascular\nInvasion", "Perineural\nInvasion", "Lymph Node\nMetastasis"),
  title = "Invasion Marker Overlap Analysis",
  subtitle = "Co-occurrence of pathological invasion features"
)
```

### Four-way Set Analysis

```{r}
# Four-way analysis including grade
venn(
  data = venn_data,
  vars = c("LVI_Present", "PNI_Present", "LN_Positive", "High_Grade"),
  names = c("LVI", "PNI", "LN+", "Grade 3"),
  title = "Comprehensive Pathological Feature Overlap",
  subtitle = "Four-way analysis of high-risk features"
)
```

## Section 4: Hierarchical Data Visualization

### Variable Tree - Data Structure Overview

```{r}
# Comprehensive variable tree
vartree(
  data = viz_data,
  vars = c("Group", "Age_Group", "Sex", "Risk_Profile", "Grade", "TStage", 
           "LVI", "PNI", "LymphNodeMetastasis", "Outcome_Status"),
  title = "Complete Dataset Structure Overview",
  subtitle = "Hierarchical view of study variables and relationships"
)
```

### Outcome-focused Tree

```{r}
# Tree focused on outcome predictors
vartree(
  data = viz_data,
  vars = c("Treatment_Response", "Risk_Profile", "Age_Group", "Sex"),
  title = "Treatment Response Predictor Structure",
  subtitle = "Key variables associated with treatment outcomes"
)
```

## Section 5: Treatment Response Visualizations

### Waterfall Plot - Individual Response

```{r}
# Prepare response data with enhanced categorization
response_enhanced <- treatmentResponse %>%
  mutate(
    RECIST_Response = case_when(
      is.na(ResponseValue) ~ "Not Evaluable",
      ResponseValue <= -30 ~ "Partial Response",
      ResponseValue > -30 & ResponseValue < 20 ~ "Stable Disease", 
      ResponseValue >= 20 ~ "Progressive Disease"
    ),
    Response_Magnitude = case_when(
      ResponseValue <= -50 ~ "Major Response",
      ResponseValue <= -30 ~ "Partial Response",
      ResponseValue > -30 & ResponseValue < 20 ~ "Stable Disease",
      ResponseValue >= 20 & ResponseValue < 50 ~ "Progression",
      ResponseValue >= 50 ~ "Rapid Progression",
      TRUE ~ "Not Evaluable"
    )
  ) %>%
  arrange(ResponseValue)

# Standard waterfall plot
waterfall(
  data = response_enhanced,
  patient_id = "PatientID",
  response = "ResponseValue",
  title = "Individual Patient Treatment Responses",
  subtitle = "Tumor size change from baseline (RECIST criteria)"
)
```

### Waterfall with Response Categories

```{r}
# Waterfall colored by response category
waterfall(
  data = response_enhanced,
  patient_id = "PatientID",
  response = "ResponseValue",
  group = "RECIST_Response",
  title = "Treatment Response by RECIST Categories",
  subtitle = "Color-coded by standard response criteria"
)
```

## Section 6: Timeline Visualizations

### Swimmer Plot - Patient Timelines

```{r}
# Prepare swimmer plot data
swimmer_data <- viz_data %>%
  filter(ID <= 30) %>%  # Limit for clarity
  mutate(
    PatientID = paste0("PT", sprintf("%03d", ID)),
    Treatment_Duration = pmax(1, OverallTime * runif(n(), 0.5, 1.0)),
    Event_Time = ifelse(Outcome == 1, OverallTime, NA),
    Response_Time = Treatment_Duration * runif(n(), 0.2, 0.6)
  ) %>%
  arrange(desc(OverallTime))

# Basic swimmer plot
swimmerplot(
  data = swimmer_data,
  patient_id = "PatientID", 
  treatment_duration = "Treatment_Duration",
  overall_time = "OverallTime",
  title = "Patient Treatment Timelines",
  subtitle = "Individual follow-up duration and treatment exposure (n=30)"
)
```

### Swimmer Plot by Treatment Group

```{r}
# Swimmer plot with grouping
swimmerplot(
  data = swimmer_data,
  patient_id = "PatientID",
  treatment_duration = "Treatment_Duration", 
  overall_time = "OverallTime",
  group = "Group",
  title = "Treatment Timeline Comparison",
  subtitle = "Patient timelines stratified by treatment group"
)
```

## Section 7: Data Quality Visualizations

### Benford's Law Analysis

```{r}
# Benford's law for measurement data
benford(
  data = viz_data,
  var = "MeasurementA",
  title = "Data Quality Assessment - Measurement A",
  subtitle = "Benford's Law analysis for fraud detection"
)
```

### Multi-variable Benford Analysis

```{r}
# Benford's law for follow-up time
benford(
  data = viz_data,
  var = "OverallTime",
  title = "Follow-up Time Distribution Analysis", 
  subtitle = "Natural pattern assessment for temporal data"
)
```

## Section 8: Advanced Composite Visualizations

### Multi-dimensional Analysis

```{r}
# Complex alluvial showing multiple relationships
alluvial(
  data = viz_data %>% sample_n(200), # Sample for clarity
  vars = c("Age_Group", "Sex", "Risk_Profile", "Group", "Outcome_Status"),
  title = "Comprehensive Patient Journey Analysis",
  subtitle = "Demographics → Risk → Treatment → Outcome pathway"
)
```

### Biomarker Integration Visualization

```{r}
# Create biomarker categories for visualization
biomarker_viz <- viz_data %>%
  mutate(
    Measurement_Profile = case_when(
      MeasurementA > median(MeasurementA, na.rm = TRUE) & 
      MeasurementB > median(MeasurementB, na.rm = TRUE) ~ "High-High",
      MeasurementA > median(MeasurementA, na.rm = TRUE) & 
      MeasurementB <= median(MeasurementB, na.rm = TRUE) ~ "High-Low",
      MeasurementA <= median(MeasurementA, na.rm = TRUE) & 
      MeasurementB > median(MeasurementB, na.rm = TRUE) ~ "Low-High",
      TRUE ~ "Low-Low"
    )
  )

# Biomarker-outcome relationship
alluvial(
  data = biomarker_viz,
  vars = c("Measurement_Profile", "Risk_Profile", "Treatment_Response"),
  title = "Biomarker Profile and Treatment Response",
  subtitle = "Integration of molecular markers with clinical outcomes"
)
```

## Section 9: Publication-Ready Customization

### Color Scheme Guidelines

```{r}
# Demonstrate color considerations for publications
cat("Recommended color schemes for clinical research:\n\n")
cat("• Treatment groups: Blue (#2E86AB) vs Orange (#F24236)\n")
cat("• Risk levels: Green (#00CC66) → Yellow (#FFCC00) → Red (#FF3300)\n") 
cat("• Outcomes: Success (#4CAF50) vs Failure (#F44336)\n")
cat("• Demographics: Standard accessible palette\n")
cat("• Pathology: Sequential color scales for severity\n\n")
cat("All visualizations should be colorblind-friendly and print well in grayscale.\n")
```

### Figure Legend Best Practices

```{r}
# Example of comprehensive figure annotation
example_figure_legend <- "
Figure 1. Treatment Response Analysis in Cancer Patients (N=250)

Panel A: Waterfall plot showing individual patient responses to treatment, 
measured as percentage change in tumor size from baseline. Each bar represents 
one patient, ordered by response magnitude. Horizontal lines indicate RECIST 
criteria thresholds: -30% (partial response) and +20% (progressive disease).

Panel B: Alluvial diagram depicting patient flow from baseline characteristics 
through treatment assignment to clinical outcomes. Line thickness represents 
number of patients; connections show relationship patterns between variables.

Abbreviations: PR, partial response; SD, stable disease; PD, progressive disease; 
RECIST, Response Evaluation Criteria in Solid Tumors.
"

cat(example_figure_legend)
```

## Section 10: Workflow Integration

### Complete Analysis Pipeline Visualization

```{r}
# Demonstrate integrated workflow visualization
workflow_demo <- viz_data %>%
  filter(!is.na(Risk_Profile) & !is.na(Treatment_Response))

# Patient journey from enrollment to outcome  
alluvial(
  data = workflow_demo,
  vars = c("Sex", "Age_Group", "Risk_Profile", "Group", "Treatment_Response"),
  title = "Complete Clinical Trial Patient Journey",
  subtitle = "Integrated view from demographics to treatment outcomes (N=247)"
)
```

### Multi-panel Figure Concept

```{r}
cat("Multi-panel figure recommendations for manuscripts:\n\n")
cat("Panel A: Demographics (Age pyramid by treatment group)\n")
cat("Panel B: Baseline characteristics (Alluvial: risk → treatment assignment)\n") 
cat("Panel C: Treatment response (Waterfall plot with RECIST categories)\n")
cat("Panel D: Outcome relationships (Venn diagram of risk factors)\n")
cat("Panel E: Timeline analysis (Swimmer plot by response group)\n\n")
cat("This approach provides comprehensive visual narrative of study results.\n")
```

## Implementation in jamovi

All visualizations are available through jamovi's interface:

### Navigation Path
- **Demographics**: `Exploration > ClinicoPath Descriptive Plots > Age Pyramid`
- **Relationships**: `Exploration > ClinicoPath Descriptive Plots > Alluvial Diagrams`
- **Set Analysis**: `Exploration > ClinicoPath Descriptive Plots > Venn Diagram`
- **Data Structure**: `Exploration > ClinicoPath Descriptive Plots > Variable Tree`
- **Treatment Response**: `Exploration > Patient Follow-Up Plots > Treatment Response Analysis`
- **Timelines**: `Exploration > Patient Follow-Up Plots > Swimmer Plot`
- **Quality Assessment**: `Exploration > ClinicoPath Descriptives > Benford Analysis`

### jamovi Advantages for Visualization

1. **Interactive Parameter Adjustment**: Real-time preview of visualization changes
2. **Export Options**: Multiple formats for manuscripts and presentations
3. **No Programming Required**: Point-and-click creation of complex visualizations
4. **Consistent Styling**: Automatic application of publication-ready themes
5. **Reproducible Output**: Saved analyses for consistent figure generation

## Best Practices Summary

### Design Guidelines

1. **Choose appropriate visualization types** for your data structure
2. **Maintain consistent color schemes** across related figures
3. **Include comprehensive legends** and annotations
4. **Optimize for your target audience** (clinical vs. research)
5. **Test readability** in grayscale and for colorblind viewers

### Technical Considerations

1. **Resolution**: Use vector formats (SVG, PDF) for scalable figures
2. **Size**: Plan for journal column widths and presentation screens
3. **Accessibility**: Ensure compliance with accessibility standards
4. **File Formats**: Match journal requirements for submission

### Clinical Research Specific

1. **Patient Privacy**: Never include identifiable information in visualizations
2. **Statistical Accuracy**: Ensure visualizations accurately represent data
3. **Regulatory Compliance**: Follow guidelines for clinical trial reporting
4. **Peer Review**: Design with peer review process in mind

This visualization gallery demonstrates the full range of clinical research visualization capabilities available in ClinicoPathDescriptives, providing templates and examples for creating impactful, publication-ready figures that effectively communicate research findings.