---
title: "Joint Modeling Tutorial: Step-by-Step Analysis Guide"
author: "ClinicoPath"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: bootstrap
    code-folding: show
    fig-width: 8
    fig-height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

# Quick Start Guide

This tutorial walks you through a complete joint modeling analysis using the ClinicoPath module. We'll use the simple cancer dataset to demonstrate all the key steps.

## What You'll Learn

- How to prepare data for joint modeling
- Step-by-step analysis in jamovi
- How to interpret results
- Common troubleshooting tips

## Prerequisites

- Basic understanding of longitudinal data
- Familiarity with survival analysis concepts
- jamovi with ClinicoPath module installed

# Data Preparation

## Understanding the Data Format

Joint models require **long-format** data where:
- Each row represents one biomarker measurement
- Multiple rows per patient (one per visit)
- Survival information repeated for each patient

```{r data-format}
library(ClinicoPath)
library(dplyr)
library(ggplot2)
library(knitr)

# Load the simple cancer dataset
data(simple_cancer_data)

# Show data structure
head(simple_cancer_data, 12) %>%
  kable(caption = "Example of Long-Format Data (first 12 rows)")

# Data summary
cat("Dataset Overview:\n")
cat("- Patients:", length(unique(simple_cancer_data$patient_id)), "\n")
cat("- Total observations:", nrow(simple_cancer_data), "\n")
cat("- Average visits per patient:", round(nrow(simple_cancer_data) / length(unique(simple_cancer_data$patient_id)), 1), "\n")
cat("- Follow-up range:", range(simple_cancer_data$visit_time), "months\n")
cat("- Event rate:", round(mean(simple_cancer_data[!duplicated(simple_cancer_data$patient_id), "progression_status"]) * 100, 1), "%\n")
```

## Data Quality Checks

Before analysis, always check your data quality:

```{r data-checks}
# 1. Check for missing values
missing_summary <- simple_cancer_data %>%
  summarise(
    missing_id = sum(is.na(patient_id)),
    missing_time = sum(is.na(visit_time)),
    missing_marker = sum(is.na(tumor_marker)),
    missing_surv_time = sum(is.na(survival_time)),
    missing_status = sum(is.na(progression_status))
  )

kable(missing_summary, caption = "Missing Values Check")

# 2. Check visit patterns
visit_counts <- simple_cancer_data %>%
  count(patient_id) %>%
  count(n, name = "patients") %>%
  mutate(proportion = round(patients / sum(patients) * 100, 1))

kable(visit_counts, 
      col.names = c("Visits per Patient", "Number of Patients", "Proportion (%)"),
      caption = "Visit Distribution")

# 3. Check time consistency
time_issues <- simple_cancer_data %>%
  group_by(patient_id) %>%
  summarise(
    max_visit_time = max(visit_time),
    survival_time = first(survival_time),
    time_consistent = max_visit_time <= survival_time,
    .groups = 'drop'
  ) %>%
  filter(!time_consistent)

if(nrow(time_issues) > 0) {
  cat("⚠️ Warning: Some patients have visit times after survival time\n")
  print(time_issues)
} else {
  cat("✅ Time consistency check passed\n")
}
```

## Visualizing the Data

Understanding your data visually is crucial:

```{r data-visualization}
# Individual trajectories (sample)
set.seed(123)
sample_patients <- sample(unique(simple_cancer_data$patient_id), 12)
sample_data <- simple_cancer_data %>% filter(patient_id %in% sample_patients)

p1 <- ggplot(sample_data, aes(x = visit_time, y = tumor_marker, color = treatment)) +
  geom_line(aes(group = patient_id), alpha = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~patient_id, scales = "free_y", ncol = 4) +
  labs(title = "Individual Tumor Marker Trajectories (Sample of 12 Patients)",
       x = "Time (months)", y = "Tumor Marker") +
  theme_bw() +
  theme(strip.text = element_text(size = 8))

# Population-level trends
p2 <- ggplot(simple_cancer_data, aes(x = visit_time, y = tumor_marker, color = treatment)) +
  geom_smooth(method = "loess", se = TRUE, size = 1.2) +
  labs(title = "Population-Level Tumor Marker Trajectories",
       subtitle = "Loess smooth with 95% confidence intervals",
       x = "Time (months)", y = "Tumor Marker") +
  theme_bw()

print(p1)
print(p2)
```

# Step-by-Step jamovi Analysis

## Step 1: Load Data into jamovi

1. Open jamovi
2. Import your dataset (File → Open)
3. Ensure variables have correct data types:
   - Patient ID: ID or Nominal
   - Time variables: Continuous
   - Biomarker: Continuous  
   - Status: Nominal (0/1 coding)

## Step 2: Access Joint Modeling Module

1. Navigate to **Analyses** menu
2. Find **SurvivalT** section
3. Select **Advanced Longitudinal**
4. Choose **Joint Longitudinal-Survival Models**

## Step 3: Variable Selection

```{r variable-setup, echo=FALSE}
cat("Variable Selection Guide:\n\n")
cat("REQUIRED VARIABLES:\n")
cat("✓ Patient ID: patient_id\n")
cat("✓ Longitudinal Time: visit_time\n") 
cat("✓ Biomarker Variable: tumor_marker\n")
cat("✓ Survival Time: survival_time\n")
cat("✓ Survival Status: progression_status (0=censored, 1=event)\n\n")

cat("OPTIONAL VARIABLES:\n")
cat("• Covariates: age, treatment\n")
cat("  (Variables that might affect biomarker trajectory or survival)\n\n")

cat("IMPORTANT NOTES:\n")
cat("• Use the SAME survival time and status for all rows of each patient\n")
cat("• Biomarker measurements should vary across visits\n")
cat("• Time should start at 0 (baseline) for each patient\n")
```

## Step 4: Model Configuration

### Basic Settings (Recommended for First Analysis)

```{r basic-settings, echo=FALSE}
cat("LONGITUDINAL MODEL OPTIONS:\n")
cat("• Biomarker Trajectory: Linear (start here)\n")
cat("• Random Effects: Intercept + Slope (default)\n")
cat("• Error Structure: Independent (default)\n\n")

cat("SURVIVAL MODEL OPTIONS:\n")
cat("• Survival Model Type: Cox Proportional Hazards (default)\n")
cat("• Biomarker-Survival Association: Current Value (default)\n\n")

cat("ANALYSIS OPTIONS:\n")
cat("• Estimation Method: Bayesian (JMbayes2) - recommended\n")
cat("• MCMC Chains: 3 (default)\n")
cat("• MCMC Iterations: 12000 (default)\n")
cat("• Burn-in Period: 2000 (default)\n")
```

### Advanced Settings (After Initial Success)

```{r advanced-settings, echo=FALSE}
cat("ADVANCED TRAJECTORY OPTIONS:\n")
cat("• Quadratic: If biomarker shows curved pattern\n")
cat("• Splines: For very flexible, complex trajectories\n\n")

cat("ADVANCED ASSOCIATION OPTIONS:\n")
cat("• Current Slope: If rate of change matters more than level\n")
cat("• Current Value + Slope: Both level and slope affect survival\n")
cat("• Area Under Curve: Cumulative biomarker exposure\n\n")

cat("VALIDATION OPTIONS:\n")
cat("• Dynamic Risk Prediction: Enable for personalized predictions\n")
cat("• Internal Validation: Enable cross-validation\n")
cat("• Discrimination Metrics: Calculate C-index and time-dependent AUC\n")
```

## Step 5: Running the Analysis

1. **Check Settings**: Review all variable selections and options
2. **Run Analysis**: Click the ▶️ button or press Ctrl+R
3. **Monitor Progress**: Watch the real-time progress indicators
4. **Wait for Completion**: Bayesian analysis takes 2-10 minutes

### Expected Analysis Timeline

```{r timeline, echo=FALSE}
cat("TYPICAL ANALYSIS TIMELINE:\n")
cat("⏱️ Data Preparation: 30 seconds\n")
cat("⏱️ Longitudinal Model: 1 minute\n") 
cat("⏱️ Survival Model: 30 seconds\n")
cat("⏱️ Joint Model (Bayesian): 5-8 minutes\n")
cat("⏱️ Diagnostics & Validation: 1 minute\n")
cat("⏱️ Total Time: 8-11 minutes\n\n")

cat("FACTORS AFFECTING SPEED:\n")
cat("• Sample size (more patients = longer time)\n")
cat("• Number of visits per patient\n")
cat("• MCMC iterations (more = longer but better)\n")
cat("• Computer specifications\n")
```

# Interpreting Results

## Understanding the Output Sections

The joint modeling analysis produces several result sections:

### 1. Data Summary
- Number of patients and observations  
- Visit patterns and missing data
- Event rates and follow-up time
- Basic data quality checks

### 2. Longitudinal Model Results
- **Fixed Effects**: Population-average trajectory parameters
- **Random Effects**: Individual variation around population average
- **Model Fit Statistics**: AIC, BIC for model comparison

### 3. Survival Model Results  
- **Covariate Effects**: Hazard ratios for baseline covariates
- **Model Diagnostics**: Concordance index, likelihood ratio tests

### 4. Joint Model Results
- **Association Parameter**: Key parameter linking biomarker to survival
- **Convergence Diagnostics**: MCMC chain quality assessment
- **Model Fit**: Overall joint model performance

## Key Parameters to Focus On

```{r interpretation-focus, echo=FALSE}
cat("🎯 MOST IMPORTANT RESULTS:\n\n")

cat("1. ASSOCIATION PARAMETER (α)\n")
cat("   • This is the key finding of joint modeling\n")
cat("   • Measures how biomarker level affects survival hazard\n")
cat("   • Positive α: Higher biomarker = Higher death risk\n")
cat("   • Negative α: Higher biomarker = Lower death risk\n")
cat("   • |α| > 0.5: Strong association\n")
cat("   • |α| = 0.1-0.5: Moderate association\n")
cat("   • |α| < 0.1: Weak association\n\n")

cat("2. TRAJECTORY PARAMETERS\n")
cat("   • Intercept: Average biomarker level at baseline\n")
cat("   • Slope: Average rate of change over time\n")
cat("   • Random effects SD: How much individuals vary\n\n")

cat("3. MODEL VALIDATION\n")
cat("   • C-index > 0.7: Good discriminative ability\n")
cat("   • Time-dependent AUC: Classification accuracy over time\n")
cat("   • Cross-validation results: Internal validity\n")
```

## Clinical Interpretation Examples

```{r clinical-examples, echo=FALSE}
cat("💡 CLINICAL INTERPRETATION EXAMPLES:\n\n")

cat("EXAMPLE 1: Positive Association (α = 0.3)\n")
cat("→ 'Each 1-unit increase in tumor marker is associated with \n")
cat("   a 35% increase in hazard of progression (exp(0.3) = 1.35)'\n")
cat("→ Clinical meaning: Rising marker levels indicate worse prognosis\n")
cat("→ Action: Monitor patients with increasing markers more closely\n\n")

cat("EXAMPLE 2: Negative Association (α = -0.2)\n") 
cat("→ 'Each 1-unit increase in CD4 count is associated with\n")
cat("   an 18% decrease in hazard of AIDS progression (exp(-0.2) = 0.82)'\n")
cat("→ Clinical meaning: Higher CD4 counts are protective\n")
cat("→ Action: Focus interventions on patients with declining CD4\n\n")

cat("EXAMPLE 3: Weak Association (α = 0.05)\n")
cat("→ 'Biomarker level shows weak association with survival'\n") 
cat("→ Clinical meaning: Biomarker may not be clinically useful\n")
cat("→ Action: Consider other biomarkers or longer follow-up\n")
```

# Dynamic Predictions

One of the most powerful features of joint models is dynamic prediction - updating individual risk estimates as new biomarker values become available.

## Understanding Dynamic Predictions

```{r dynamic-predictions, echo=FALSE}
cat("🎯 DYNAMIC PREDICTION CONCEPT:\n\n")

cat("Traditional Survival Analysis:\n")
cat("• Uses only baseline information\n")
cat("• Risk estimate stays constant over time\n")
cat("• Cannot incorporate new biomarker data\n\n")

cat("Joint Model Dynamic Predictions:\n")
cat("• Uses ALL biomarker measurements up to current time\n")
cat("• Risk estimates update with each new measurement\n")
cat("• Accounts for individual biomarker trajectory\n")
cat("• Provides personalized, time-updated predictions\n\n")

cat("CLINICAL APPLICATIONS:\n")
cat("• Personalized treatment timing\n")
cat("• Dynamic monitoring schedules\n") 
cat("• Real-time clinical decision support\n")
cat("• Patient counseling with updated prognosis\n")
```

## Interpreting Dynamic Predictions

When dynamic predictions are enabled, you'll see:

1. **Prediction Horizons**: Risk estimates at 6 months, 1 year, 2 years, etc.
2. **Individual Profiles**: Risk trajectories for sample patients
3. **Uncertainty Intervals**: Confidence bounds around predictions
4. **Risk Stratification**: Classifications based on predicted risk

# Troubleshooting Common Issues

## Issue 1: Analysis Won't Start

**Problem**: Error message appears immediately

**Solutions**:
```{r trouble1, echo=FALSE}
cat("✅ SOLUTIONS:\n")
cat("1. Check variable types:\n")
cat("   • Patient ID: Should be ID, Factor, or Numeric\n")
cat("   • Time variables: Must be Continuous/Numeric\n")
cat("   • Biomarker: Must be Continuous/Numeric\n")
cat("   • Status: Should be Nominal (0/1) or Numeric\n\n")

cat("2. Check data format:\n")
cat("   • Must be in long format (multiple rows per patient)\n")
cat("   • Each patient needs at least 2 biomarker measurements\n")
cat("   • Survival time and status should be identical for all rows of same patient\n\n")

cat("3. Check for missing data:\n")
cat("   • Complete case analysis - remove rows with missing biomarker\n")
cat("   • At least 20 patients with events recommended\n")
```

## Issue 2: Long Computation Time

**Problem**: Analysis takes too long or appears stuck

**Solutions**:
```{r trouble2, echo=FALSE}
cat("✅ SOLUTIONS:\n")
cat("1. Reduce MCMC settings for initial run:\n")
cat("   • Iterations: 5000 (instead of 12000)\n")
cat("   • Chains: 2 (instead of 3)\n")
cat("   • Burn-in: 1000 (instead of 2000)\n\n")

cat("2. Try simpler model:\n")
cat("   • Linear trajectory (instead of splines)\n")
cat("   • Current Value association (default)\n")
cat("   • Remove some covariates initially\n\n")

cat("3. Alternative estimation:\n")
cat("   • Switch to Two-stage method (joineR)\n")
cat("   • Faster but less accurate than Bayesian\n")
```

## Issue 3: Convergence Problems

**Problem**: Warning about convergence or poor mixing

**Solutions**:
```{r trouble3, echo=FALSE}
cat("✅ SOLUTIONS:\n")
cat("1. Increase MCMC iterations:\n")
cat("   • Try 20000 iterations\n")
cat("   • Increase burn-in to 4000\n\n")

cat("2. Check data quality:\n")
cat("   • Remove outliers in biomarker values\n")
cat("   • Check for data entry errors\n")
cat("   • Ensure adequate sample size\n\n")

cat("3. Simplify model:\n")
cat("   • Start with linear trajectory\n")
cat("   • Use fewer covariates\n")
cat("   • Try different association structure\n")
```

## Issue 4: Unexpected Results

**Problem**: Results don't match clinical expectations

**Solutions**:
```{r trouble4, echo=FALSE}
cat("✅ SOLUTIONS:\n")
cat("1. Check data coding:\n")
cat("   • Event status: 1 = event, 0 = censored\n")
cat("   • Time units consistent (all in months or years)\n")
cat("   • Biomarker units and scaling\n\n")

cat("2. Examine data patterns:\n")
cat("   • Plot individual trajectories\n")
cat("   • Check for data quality issues\n")
cat("   • Verify biological plausibility\n\n")

cat("3. Model diagnostics:\n")
cat("   • Check convergence diagnostics\n")
cat("   • Examine residual plots\n")
cat("   • Compare with separate analyses\n")
```

# Best Practices Summary

## Before Analysis
- ✅ Verify data is in long format
- ✅ Check for missing values and outliers  
- ✅ Visualize biomarker trajectories
- ✅ Ensure adequate sample size (≥50 patients, ≥20 events)
- ✅ Plan analysis with clinical collaborators

## During Analysis
- ✅ Start with simple models (linear trajectory)
- ✅ Monitor progress indicators
- ✅ Be patient with Bayesian estimation
- ✅ Save intermediate results
- ✅ Check convergence diagnostics

## After Analysis  
- ✅ Interpret association parameter clinically
- ✅ Examine individual predicted trajectories
- ✅ Validate results with cross-validation
- ✅ Compare with separate longitudinal/survival analyses
- ✅ Discuss implications with clinical team

## Reporting Results
- ✅ Report association parameter with confidence interval
- ✅ Describe model specifications clearly
- ✅ Include convergence diagnostics
- ✅ Show example dynamic predictions
- ✅ Discuss clinical implications and limitations

---

## Next Steps

After mastering basic joint modeling:

1. **Try More Complex Trajectories**: Experiment with quadratic or spline models
2. **Multiple Biomarkers**: Analyze correlated longitudinal outcomes
3. **Competing Risks**: Handle multiple types of events  
4. **Time-Varying Effects**: Allow association to change over time
5. **External Validation**: Test model performance on new datasets

The joint modeling framework in ClinicoPath provides a powerful platform for longitudinal biomarker analysis. Start simple, build complexity gradually, and always prioritize clinical interpretability over statistical sophistication.