---
title: "ToDo"
---

# Packages


Imports: 
    jmvcore (>= 0.8.5),
    R6,
    dplyr,
    survival,
    survminer,
    finalfit,
    arsenal,
    purrr,
    glue,
    janitor,
    ggplot2,
    forcats,
    ggstatsplot,
    tableone,
    explore,
    tangram,
    irr,
    rlang,
    tidyselect,
    knitr



Remotes:
    easystats/correlation,
    neuropsychology/psycho.R@0.4.0


Suggests:
    rmarkdown,
    remotes,
    devtools,
    lubridate,
    broom, 
    GGally, 
    gridExtra, 
    Hmisc, 
    lme4, 
    magrittr, 
    mice, 
    pillar, 
    pROC, 
    scales, 
    stringr, 
    tibble, 
    tidyr, 
    covr, 
    cmprsk, 
    readr, 
    rstan, 
    survey, 
    testthat, 
    backports, 
    generics, 
    assertthat, 
    pkgconfig, 
    Rcpp, 
    BH, 
    plogr, 
    ellipsis, 
    gtable, 
    progress, 
    RColorBrewer, 
    reshape, 
    digest, 
    lazyeval, 
    viridisLite, 
    withr, 
    Formula, 
    latticeExtra, 
    acepack, 
    data.table, 
    htmlTable, 
    viridis, 
    htmltools, 
    base64enc, 
    minqa, 
    nloptr, 
    RcppEigen, 
    mitml, 
    cli, 
    crayon, 
    fansi, 
    utf8, 
    vctrs, 
    farver, 
    labeling, 
    munsell, 
    lifecycle, 
    stringi, 
    ggpubr, 
    maxstat, 
    survMisc, 
    jsonlite, 
    rex, 
    evaluate, 
    highr, 
    markdown, 
    xfun, 
    hms, 
    clipr, 
    mime, 
    tinytex, 
    StanHeaders, 
    inline, 
    loo, 
    pkgbuild, 
    numDeriv, 
    mitools, 
    pkgload, 
    praise, 
    zeallot, 
    colorspace, 
    prettyunits, 
    checkmate, 
    htmlwidgets, 
    pan, 
    jomo, 
    ordinal, 
    ucminf, 
    ggrepel,
    ggsci, 
    cowplot, 
    ggsignif,
    polynom, 
    exactRankTests,
    mvtnorm, 
    KMsurv, 
    zoo, 
    km.ci,
    xtable,
    curl, 
    openssl, 
    askpass, 
    sys, 
    matrixStats,
    callr, 
    desc,
    rprojroot,
    processx, 
    ps, 
    DBI,
    png, 
    jpeg, 
    boot, 
    grid, 
    snakecase,
    caret,
    iterators,
    timeDate,
    foreach,
    plyr,
    ModelMetrics,
    nlme,
    reshape2,
    recipes,
    BradleyTerry2,
    e1071,
    earth,
    fastICA,
    gam,
    ipred,
    kernlab,
    klaR,
    MASS,
    ellipse,
    mda,
    mgcv,
    mlbench,
    MLmetrics,
    nnet,
    party,
    pls,
    proxy,
    randomForest,
    RANN,
    spls,
    subselect,
    pamr,
    superpc,
    Cubist,
    rpart,
    qgraph,
    nFactors,
    ppcor,
    rstanarm,
    MuMIn,
    blavaan,
    
    
    
    
    









---

# Develop



```{r library}
# install.packages('jmvtools', repos=c('https://repo.jamovi.org', 'https://cran.r-project.org'))

# jmvtools::check("C://Program Files//jamovi//bin")

# jmvtools::install(home = "C://Program Files//jamovi//bin")
# 
# devtools::build(path = "C:\\ClinicoPathOutput")

# .libPaths(new = "C:\\ClinicoPathLibrary")

# devtools::build(path = "C:\\ClinicoPathOutput", binary = TRUE, args = c('--preclean'))

Sys.setenv(TZ="Europe/Istanbul")

library("jmvtools")
```


```{r check, include=FALSE}
jmvtools::check()

# rhub::check_on_macos()

# rhub::check_for_cran()

# codemetar::write_codemeta()


devtools::check()

```


```{r pkgdown build, include=FALSE}
rmarkdown::render('/Users/serdarbalciold/histopathRprojects/ClinicoPath/README.Rmd',  encoding = 'UTF-8', knit_root_dir = '~/histopathRprojects/ClinicoPath', quiet = TRUE)

devtools::document()

pkgdown::build_site()
```


```{r git force push, include=FALSE}
# gitUpdateCommitPush
CommitMessage <- paste("updated on ", Sys.time(), sep = "")
wd <- getwd()
gitCommand <- paste("cd ", wd, " \n git add . \n git commit --message '", CommitMessage, "' \n git push origin master \n", sep = "")
# gitCommand <- paste("cd ", wd, " \n git add . \n git commit --no-verify --message '", CommitMessage, "' \n git push origin master \n", sep = "")
system(command = gitCommand, intern = TRUE)

```



```{r add analysis, eval=FALSE, include=FALSE}

# jmvtools::install()
# 
# jmvtools::create('SuperAwesome')
# 
# jmvtools::addAnalysis(name='ttest', title='Independent Samples T-Test')
# 
# jmvtools::addAnalysis(name='survival', title='survival')
# 
# jmvtools::addAnalysis(name='correlation', title='correlation')
# 
# jmvtools::addAnalysis(name='tableone', title='TableOne')
# 
# jmvtools::addAnalysis(name='crosstable', title='CrossTable')
# 
# 
# jmvtools::addAnalysis(name='writesummary', title='WriteSummary')

# jmvtools::addAnalysis(name='finalfit', title='FinalFit')

# jmvtools::addAnalysis(name='multisurvival', title='FinalFit Multivariate Survival')

# jmvtools::addAnalysis(name='report', title='Report General Features')

# jmvtools::addAnalysis(name='frequencies', title='Frequencies')

# jmvtools::addAnalysis(name='statsplot', title='GGStatsPlot')

# jmvtools::addAnalysis(name='statsplot2', title='GGStatsPlot2')

# jmvtools::addAnalysis(name='scat2', title='scat2')

# jmvtools::addAnalysis(name='decisioncalculator', title='Decision Calculator')

# jmvtools::addAnalysis(name='agreement', title='Interrater Intrarater Reliability')

# jmvtools::addAnalysis(name='cluster', title='Cluster Analysis')

# jmvtools::addAnalysis(name='tree', title='Decision Tree')

```






```{r devtools install, eval=FALSE, include=FALSE}
devtools::install()
```


```{r jmvtools install, eval=FALSE, include=FALSE}
# jmvtools::check()
jmvtools::install()
```



```{r construct}
formula <- jmvcore::constructFormula(terms = c("A", "B", "C"), dep = "D")

jmvcore::constructFormula(terms = list("A", "B", c("C", "D")), dep = "E")


jmvcore::constructFormula(terms = list("A", "B", "C"))

vars <- jmvcore::decomposeFormula(formula = formula) 

unlist(vars)

cformula <- jmvcore::composeTerm(components = formula)

jmvcore::composeTerm("A")

jmvcore::composeTerm(components = c("A", "B", "C"))

jmvcore::decomposeTerm(term = c("A", "B", "C"))

jmvcore::decomposeTerm(term = formula)

jmvcore::decomposeTerm(term = cformula)



composeTerm <- jmvcore::composeTerm(components = c("A", "B", "C"))

jmvcore::decomposeTerm(term = composeTerm)


```









---

# Example


```{r read data}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
```


```{r writesummary, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

# library("ClinicoPath")

deneme$Age <- as.numeric(as.character(deneme$Age))

ClinicoPath::writesummary(data = deneme, vars = Age)

ggstatsplot::normality_message(deneme$Age, "Age")


ClinicoPath::writesummary(
    data = deneme,
    vars = Age)


```



```{r finalfit, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
library(dplyr)
library(survival)
library(finalfit)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ClinicoPath::finalfit(data = deneme,
                      explanatory = Sex,
                      outcome = Outcome,
                      overalltime = OverallTime)
```



```{r decision, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

ClinicoPath::decision(
    data = deneme,
    gold = Outcome,
    goldPositive = "1",
    newtest = Smoker,
    testPositive = "TRUE")

ClinicoPath::decision(
    data = deneme,
    gold = LVI,
    goldPositive = "Present",
    newtest = PNI,
    testPositive = "Present")
```

```{r eval=FALSE, include=FALSE}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ggstatsplot::ggbetweenstats(data = deneme, 
                            x = LVI,
                            y = Age)

```





```{r statsplot, eval=FALSE, include=FALSE}
devtools::install(upgrade = FALSE, quick = TRUE)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))
ClinicoPath::statsplot(
    data = deneme,
    dep = Age,
    group = Smoker)
```




```{r decision 2, eval=FALSE, include=FALSE}
mytable <- table(deneme$Outcome, deneme$Smoker)

caret::confusionMatrix(mytable)
confusionMatrix(pred, truth)
confusionMatrix(xtab, prevalence = 0.25)

levels(deneme$Outcome)

mytable[1,2]

d <- "0"

mytable[d, "FALSE"]

mytable[[0]]

```











```{r construct formula, eval=FALSE, include=FALSE}
formula <- jmvcore::constructFormula(terms = c("A", "B", "C"))

jmvcore::constructFormula(terms = list("A", "B", "C"))

vars <- jmvcore::decomposeFormula(formula = formula) 

vars <- jmvcore::decomposeTerms(vars)


vars <- unlist(vars)

formula <- as.formula(formula)


my_group <- "lvi"
my_dep <- "age"

formula <- paste0('x = ', group, 'y = ', dep)
myformula <- as.formula(formula)

myformula <- glueformula::gf({my_group}, {my_dep})

myformula <- glue::glue( 'x = ' , {my_group}, ', y = ' , {my_dep})

myformula <- jmvcore::composeTerm(myformula)


```


```{r}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

library(survival)
km_fit <- survfit(Surv(OverallTime, Outcome) ~ LVI, data = deneme)

library(dplyr)
km_fit_median_df <- summary(km_fit)
km_fit_median_df <- as.data.frame(km_fit_median_df$table) %>%
                janitor::clean_names(dat = ., case = "snake") %>%
                tibble::rownames_to_column(.data = ., var = "LVI")

```



```{r construct formula 2, eval=FALSE, include=FALSE}
library(dplyr)
library(survival)

deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

myoveralltime <- deneme$OverallTime
myoutcome <- deneme$Outcome
myexplanatory <- deneme$LVI

class(myoveralltime)
class(myoutcome)
typeof(myexplanatory)

is.ordered(myexplanatory)

formula2 <- jmvcore::constructFormula(terms = "myexplanatory")
# formula2 <- jmvcore::decomposeFormula(formula = formula2)
# formula2 <- paste("", formula2)
# formula2 <- as.formula(formula2)
formula2 <- jmvcore::composeTerm(formula2)


formulaL <- jmvcore::constructFormula(terms = "myoveralltime")
# formulaL <- jmvcore::decomposeFormula(formula = formulaL)

formulaR <- jmvcore::constructFormula(terms = "myoutcome")
# formulaR <- jmvcore::decomposeFormula(formula = formulaR)

formula <- paste("Surv(", formulaL, ",", formulaR, ")")
# formula <- jmvcore::composeTerm(formula)
# formula <- as.formula(formula)
# jmvcore::constructFormula(terms = c(formula, formula2))

deneme %>%
  finalfit::finalfit(formula, formula2) -> tUni

tUni
```


```{r eval=FALSE, include=FALSE}
library(dplyr)
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))

results <- deneme %>%
                ggstatsplot::ggbetweenstats(LVI, Age)
results

mydep <- deneme$Age
mygroup <- deneme$LVI


mygroup <- jmvcore::constructFormula(terms = "mygroup")
mygroup <- jmvcore::composeTerm(mygroup)

mydep <- jmvcore::constructFormula(terms = "mydep")
mydep <- jmvcore::composeTerm(mydep)


# not working
# eval(mygroup)
# rlang::eval_tidy(mygroup)
# !!mygroup
# {{mygroup}}
# sym(mygroup)
# quote(mygroup)
# enexpr(mygroup)

mygroup <- jmvcore::constructFormula(terms = "mygroup")
mydep <- jmvcore::constructFormula(terms = "mydep")

formula1 <- paste(mydep)
formula1 <- jmvcore::composeTerm(formula1)


mygroup <- paste(mygroup)
mygroup <- jmvcore::composeTerm(mygroup)

mydep <- deneme$Age
mygroup <- deneme$LVI

mydep <- jmvcore::resolveQuo(jmvcore::enquo(mydep))
mygroup <- jmvcore::resolveQuo(jmvcore::enquo(mygroup))

mydata2 <- data.frame(mygroup=mygroup, mydep=mydep)

results <- mydata2 %>%
                ggstatsplot::ggbetweenstats(
x = mygroup, y = mydep  )

results



myformula <- glue::glue('x = ', {mygroup}, ', y = ' , {mydep})

myformula <- jmvcore::composeTerm(myformula)

myformula <- as.formula(myformula)
















mydep2 <- quote(mydep)
mygroup2 <- quote(mygroup)


results <- deneme %>%
                ggstatsplot::ggbetweenstats(!!mygroup2, !!mydep2)
results
```








```{r construct formula 3, eval=FALSE, include=FALSE}
formula <- jmvcore::constructFormula(terms = c("myoveralltime", "myoutcome"))

vars <- jmvcore::decomposeFormula(formula = formula) 


explanatory <- jmvcore::constructFormula(terms = c("explanatory"))    

explanatory <- jmvcore::decomposeFormula(formula = explanatory)

explanatory <- unlist(explanatory)

myformula <- paste("Surv(", vars[1], ", ", vars[2], ")")



deneme %>%
finalfit::finalfit(myformula, explanatory) -> tUni

```


```{r table tangram}
deneme <- readxl::read_xlsx(path = here::here("tododata", "histopathology-template2019-11-25.xlsx"))


table3 <-
  tangram::html5(
    tangram::tangram(
      "Death ~ LVI + PNI + Age", deneme),
    fragment=TRUE,
    inline="nejm.css",
    caption = "HTML5 Table NEJM Style",
    id="tbl3")

table3

mydep <- deneme$Age
mygroup <- deneme$Death


formulaR <- jmvcore::constructFormula(terms =  c("LVI", "PNI", "Age"))

formulaL <- jmvcore::constructFormula(terms = "Death")

formula <- paste(formulaL, '~', formulaR)

formula <- as.formula(formula)


table <- tangram::html5(
    tangram::tangram(formula, deneme
                     ))

table


```






---

# Other Codes

## force git

```{r force git, eval=FALSE, include=FALSE}
# gitUpdateCommitPush
CommitMessage <- paste("updated on ", Sys.time(), sep = "")
wd <- getwd()
gitCommand <- paste("cd ", wd, " \n git add . \n git commit --message '", CommitMessage, "' \n git push origin master \n", sep = "")
system(command = gitCommand, intern = TRUE)

```


## arsenal

```{r arsenal, results='asis'}

tab1 <- arsenal::tableby(~ Age + Sex, data = deneme)

results <- summary(tab1)


# results$object
# results$control
# results$totals
# results$hasStrata
# results$text
# results$pfootnote
# results$term.name
# 
# tab1$Call
# 
# tab1$control

tab1$tables # this is where results lie


```


## define survival time

```{r define survival time, eval=FALSE, include=FALSE}
mydata$int <- lubridate::interval(
  lubridate::ymd(mydata$SurgeryDate),
  lubridate::ymd(mydata$LastFollowUpDate)
  )
mydata$OverallTime <- lubridate::time_length(mydata$int, "month")
mydata$OverallTime <- round(mydata$OverallTime, digits = 1)
```




## Multivariate Analysis Survival


```{r Multivariate Analysis, eval=FALSE, include=FALSE}
library(finalfit)
library(survival)
explanatoryMultivariate <- explanatoryKM
dependentMultivariate <- dependentKM

mydata %>%
  finalfit(dependentMultivariate, explanatoryMultivariate) -> tMultivariate

knitr::kable(tMultivariate, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```


---

```{r}
# Find arguments in yaml

list_of_yaml <- c(
    list.files(path = "~/histopathRprojects/ClinicoPath-Jamovi--prep/jmv",
               pattern = "\\.yaml$",
               full.names = TRUE,
               all.files = TRUE,
               include.dirs = TRUE,
               recursive = TRUE
    )
)


text_of_yaml_yml <- purrr::map(
    .x = list_of_yaml,
    .f = readLines
)

text_of_yaml_yml <- as.vector(unlist(text_of_yaml_yml)) 

arglist <-
    stringr::str_extract(
        string = text_of_yaml_yml, 
        pattern = 
            "([[:alnum:]]*):"
    )

arglist <- arglist[!is.na(arglist)]
arglist <- unique(arglist)
arglist <- gsub(pattern = ":", # remove some characters
                    replacement = "",
                    x = arglist)
arglist <- trimws(arglist) # remove whitespace

cat(arglist, sep = "\n")

```






<!-- # -->
<!-- #                 # tUni_df_descr <- paste0("When ", -->
<!-- #                 #                         tUni_df$dependent_surv_overall_time_outcome[1], -->
<!-- #                 #                         " is ", -->
<!-- #                 #                         tUni_df$x[2], -->
<!-- #                 #                         ", there is ", -->
<!-- #                 #                         tUni_df$hr_univariable[2], -->
<!-- #                 #                         " times risk than ", -->
<!-- #                 #                         "when ", -->
<!-- #                 #                         tUni_df$dependent_surv_overall_time_outcome[1], -->
<!-- #                 #                         " is ", -->
<!-- #                 #                         tUni_df$x[1], -->
<!-- #                 #                         "." -->
<!-- #                 # ) -->
<!-- # -->
<!-- #                 # results5 <- tUni_df_descr -->


```{r}
boot::melanoma
rio::export(x = boot::melanoma, file = "data/melanoma.csv")

survival::colon
rio::export(x = survival::colon, file = "data/colon.csv")

# BreastCancerData <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data"
# 
# BreastCancerNames <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names"
# 
# BreastCancerData <- read.csv(file = BreastCancerData, header = FALSE,
#                 col.names = c("id","CT", "UCSize", "UCShape", "MA", "SECS", "BN", "BC", "NN","M", "diagnosis") )

library(mlbench)

data("BreastCancer")
BreastCancer

rio::export(x = BreastCancer, file = "data/BreastCancer.csv")



```



