---
title: "Swimmer Plot Function: Example Datasets Guide"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Swimmer Plot Function: Example Datasets Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
ggswim_available <- FALSE

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
  
)
```


```{r setup 2, include = FALSE}
# Load required packages
library(ClinicoPath)

# Check if ggswim is available
ggswim_available <- requireNamespace("ggswim", quietly = TRUE)
if (!ggswim_available) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Note: ggswim package not available. Code examples shown but not executed.")
}
```

# Introduction

This vignette demonstrates the example datasets created specifically for the swimmer plot function (`swimmerplot`). Each dataset showcases different features and use cases, from basic timeline visualization to complex clinical trial analysis.

## Available Datasets

The ClinicoPath package includes five specialized datasets for swimmer plot demonstrations:

1. **`swimmer_unified_basic`** - Simple swimmer plot with basic features
2. **`swimmer_unified_comprehensive`** - Advanced features with multiple milestones
3. **`swimmer_unified_datetime`** - Date/time handling demonstration
4. **`swimmer_unified_events`** - Event markers and longitudinal data
5. **`swimmer_unified_oncology`** - Realistic oncology clinical trial data

# Basic Swimmer Plot Dataset

## swimmer_unified_basic

This dataset demonstrates fundamental swimmer plot functionality with clean, simple data.

```{r basic-dataset}
# Load the basic dataset
data(swimmer_unified_basic)

# Examine the structure
str(swimmer_unified_basic)
head(swimmer_unified_basic)
```

```{r basic-plot, eval=ggswim_available}
# Create basic swimmer plot
basic_result <- swimmerplot(
  data = swimmer_unified_basic,
  patientID = "PatientID",
  startTime = "StartTime",
  endTime = "EndTime",
  responseVar = "Response",
  timeUnit = "months",
  plotTheme = "ggswim",
  showLegend = TRUE,
  eventVar = NULL,
  milestone1Date = NULL,
  milestone2Date = NULL,
  milestone3Date = NULL,
  milestone4Date = NULL,
  milestone5Date = NULL,
  eventTimeVar = NULL,
  sortVariable = NULL
)

print(basic_result)
```

**Key Features Demonstrated:**
- Basic patient timeline visualization
- Response-based color coding
- Simple time intervals
- Clean, professional styling

# Comprehensive Dataset with Milestones

## swimmer_unified_comprehensive

This dataset showcases advanced swimmer plot features including multiple clinical milestones and patient characteristics.

```{r comprehensive-dataset}
# Load the comprehensive dataset
data(swimmer_unified_comprehensive)

# Examine the structure
str(swimmer_unified_comprehensive)
head(swimmer_unified_comprehensive)
```

```{r comprehensive-plot, eval=ggswim_available}
# Create comprehensive swimmer plot with milestones
comprehensive_result <- swimmerplot(
  data = swimmer_unified_comprehensive,
  patientID = "PatientID",
  startTime = "StartTime",
  endTime = "EndTime",
  responseVar = "BestResponse",
  milestone1Name = "Surgery",
  milestone1Date = "Surgery",
  milestone2Name = "First Response",
  milestone2Date = "FirstResponse",
  milestone3Name = "Progression",
  milestone3Date = "Progression",
  milestone4Name = "Death/Last FU",
  milestone4Date = "DeathLastFU",
  referenceLines = "median",
  showInterpretation = TRUE,
  personTimeAnalysis = TRUE,
  responseAnalysis = TRUE,
  plotTheme = "ggswim",
  sortOrder = "duration_desc"
)

print(comprehensive_result)
```

**Key Features Demonstrated:**
- Multiple milestone events (up to 4 milestones)
- Patient demographics integration
- Person-time analysis
- Clinical interpretation
- Reference lines for context
- Advanced sorting options

# DateTime Handling Dataset

## swimmer_unified_datetime

This dataset demonstrates how to work with actual calendar dates and datetime formats.

```{r datetime-dataset}
# Load the datetime dataset
data(swimmer_unified_datetime)

# Examine the structure
str(swimmer_unified_datetime)
head(swimmer_unified_datetime)
```

```{r datetime-plot, eval=ggswim_available}
# Create datetime swimmer plot with relative display
datetime_result <- swimmerplot(
  data = swimmer_unified_datetime,
  patientID = "PatientID",
  startTime = "StartDate",
  endTime = "EndDate",
  responseVar = "BestResponse",
  timeType = "datetime",
  dateFormat = "ymd",
  timeUnit = "months",
  timeDisplay = "relative",
  milestone1Name = "Surgery",
  milestone1Date = "Surgery",
  referenceLines = "protocol",
  plotTheme = "ggswim"
)

print(datetime_result)
```

**Key Features Demonstrated:**
- Real calendar date handling
- Date format specification (YYYY-MM-DD)
- Relative vs. absolute time display
- Protocol reference lines
- Multi-site clinical trial data

# Event Markers Dataset

## swimmer_unified_events

This longitudinal dataset shows how to display clinical events along patient timelines.

```{r events-dataset}
# Load the events dataset
data(swimmer_unified_events)

# Examine the structure
str(swimmer_unified_events)
head(swimmer_unified_events, 10)
```

```{r events-plot, eval=ggswim_available}
# Create event markers swimmer plot
events_result <- swimmerplot(
  data = swimmer_unified_events,
  patientID = "PatientID",
  startTime = "StartTime",
  endTime = "EndTime",
  responseVar = "Response",
  showEventMarkers = TRUE,
  eventVar = "EventType",
  eventTimeVar = "EventTime",
  markerSize = 6,
  laneWidth = 4,
  plotTheme = "ggswim",
  showLegend = TRUE
)

print(events_result)
```

**Key Features Demonstrated:**
- Event markers along timelines
- Multiple events per patient
- Treatment cycle tracking
- Adverse event severity coding
- Longitudinal data handling

# Oncology Clinical Trial Dataset

## swimmer_unified_oncology

This realistic dataset represents a comprehensive oncology clinical trial with detailed clinical milestones and biomarker information.

```{r oncology-dataset}
# Load the oncology dataset
data(swimmer_unified_oncology)

# Examine the structure
str(swimmer_unified_oncology)
head(swimmer_unified_oncology)

# Summary of key variables
table(swimmer_unified_oncology$BestResponse)
table(swimmer_unified_oncology$TumorType)
table(swimmer_unified_oncology$Stage)
```

```{r oncology-plot, eval=ggswim_available}
# Create comprehensive oncology swimmer plot
oncology_result <- swimmerplot(
  data = swimmer_unified_oncology,
  patientID = "PatientID",
  startTime = "StartTime",
  endTime = "EndTime",
  responseVar = "BestResponse",
  milestone1Name = "Surgery",
  milestone1Date = "Surgery",
  milestone2Name = "Biopsy",
  milestone2Date = "Biopsy",
  milestone3Name = "First Response",
  milestone3Date = "FirstResponse",
  milestone4Name = "Progression",
  milestone4Date = "Progression",
  milestone5Name = "Death/Last FU",
  milestone5Date = "DeathLastFU",
  referenceLines = "protocol",
  showInterpretation = TRUE,
  personTimeAnalysis = TRUE,
  responseAnalysis = TRUE,
  plotTheme = "ggswim",
  sortOrder = "duration_desc"
)

print(oncology_result)
```

**Key Features Demonstrated:**
- Realistic oncology trial structure
- Multiple tumor types and stages
- Biomarker integration
- Complete milestone tracking
- Regulatory-ready analysis
- Publication-quality visualization

# Dataset Comparison Summary

## Data Structure Overview

```{r dataset-comparison, eval=ggswim_available}
# Compare dataset characteristics
datasets <- list(
  "Basic" = swimmer_unified_basic,
  "Comprehensive" = swimmer_unified_comprehensive,
  "DateTime" = swimmer_unified_datetime,
  "Events" = swimmer_unified_events,
  "Oncology" = swimmer_unified_oncology
)

comparison <- data.frame(
  Dataset = names(datasets),
  Rows = sapply(datasets, nrow),
  Columns = sapply(datasets, ncol),
  Patients = sapply(datasets, function(x) length(unique(x[[1]]))),
  Has_Milestones = c("No", "Yes", "Yes", "No", "Yes"),
  Has_Events = c("No", "No", "No", "Yes", "No"),
  Time_Type = c("Numeric", "Numeric", "DateTime", "Numeric", "Numeric"),
  Clinical_Focus = c("General", "Advanced", "Multi-site", "Longitudinal", "Oncology"),
  stringsAsFactors = FALSE
)

print(comparison)
```

## Use Case Recommendations

### For Learning and Testing
- **swimmer_unified_basic**: Start here for initial exploration
- **swimmer_unified_comprehensive**: Learn advanced features

### For Development and Validation
- **swimmer_unified_events**: Test event marker functionality
- **swimmer_unified_datetime**: Validate date/time handling

### For Clinical Applications
- **swimmer_unified_oncology**: Template for oncology trials
- **swimmer_unified_comprehensive**: General clinical research

# Best Practices with Example Datasets

## Data Preparation Tips

1. **Column Naming**: Use consistent, descriptive names
2. **Missing Values**: Handle NA values appropriately
3. **Data Types**: Ensure proper variable types
4. **Date Formats**: Use standard ISO formats (YYYY-MM-DD)

## Visualization Guidelines

1. **Color Coding**: Use standard clinical response codes
2. **Reference Lines**: Add clinical context
3. **Sorting**: Choose appropriate patient ordering
4. **Themes**: Select professional styling

## Analysis Features

1. **Person-Time**: Enable for epidemiological insights
2. **Interpretation**: Use for automated clinical insights
3. **Export**: Generate data for regulatory submissions

# Extending the Examples

## Creating Custom Datasets

You can create your own datasets following these patterns:

```r
# Template for custom swimmer plot data
custom_data <- data.frame(
  PatientID = paste0("PT", sprintf("%03d", 1:n)),
  StartTime = rep(0, n),  # or actual start times
  EndTime = sample(6:36, n, replace = TRUE),
  Response = sample(c("CR", "PR", "SD", "PD"), n, replace = TRUE),
  # Add your specific variables here
  stringsAsFactors = FALSE
)
```

## Integration with Clinical Systems

These datasets serve as templates for integrating swimmer plots with:

- Electronic Health Records (EHR)
- Clinical Trial Management Systems (CTMS)
- Regulatory submission packages
- Publication manuscripts

# Conclusion

The swimmer plot unified function example datasets provide comprehensive coverage of clinical timeline visualization needs. From basic exploration to regulatory submissions, these datasets demonstrate the full capabilities of the ClinicoPath swimmer plot functionality.

## Next Steps

1. **Explore**: Try each dataset with different parameters
2. **Adapt**: Modify datasets for your specific use case
3. **Implement**: Apply to your clinical data
4. **Share**: Contribute improvements back to the package

---

*For more information about ClinicoPath swimmer plot functionality, see the main vignette and package documentation.*
